### 2013-07-14 Compiled/preface/preface.tex
adding preface

In the chapter on Internet Topology Research Redux 
we revisit some of the interesting properties of the critical
infrastructure at router, AS and PoP level which today underpin how
the world is knot together.

On top of this ever evolving topology,
traffic sources are regulated by end-to-end protocols such as TCP -
these too see continual changes, and the chapter on
Recent Advances in Transport Protocols covers some of the ways TCP is
being enhanced to cope with multihoming, and take advantage of multiple
active routes between ends, whether in the data center or in the
pocket.

We can put together the traffic sources, and the topology
under one framework, that of Internet Traffic Matrices. The next
chapter introduces a primer to this topic.

One general framework for understanding dynamic traffic management in
the Internet is to think of the system as one of continual
optimization. The chapter on Optimizing and Modeling Dynamics in Networks
revisits the goals of fairness and the control problem and its
solution space.

It has been said that all is fair in love and war, but the only
certainties in the world are death and taxes. Life may not be fair,
but we can enforce a different kind of fairness on the Internet
through traffic pricing. The next chapter discusses
Smart Data Pricing (SDP) and covers a range of 
Economic Solutions to Network Congestion, bringing game theory to bear
on the problem where the previous chapter engaged with the weapons of
optimization.

At a coarser grain (in time and space) we can control traffic by
partitioning our network into VPNs - the next chapter looks at the
practical tools available to the operator to manage a set of such
disjoint systems and their capacity, by focusing on MPLS and Virtual
Private Networks.

For some time, the capacity of the net has been consumed largely by
people downloading or live streaming content. Content Distribution
Networks are overlays that allow management of the load. A less
rigorous but nearly as popular family of tools exist based on the
famous peer-to-peer paradigm. The next chapter looks at how
the world has moved on from fighting, to embracing
Collaboration Opportunities for Content Providers and Network
Infrastructures.

The true explosion in Internet access in the developing world (and now
dominant form of access in the developed world too) has been through
mobile devices (smart phones, tablets and the rest). The next chapter
looks at the Design Space of Network Mobility, and how seamlessness is
achieved (or at least as close to seamless as we can get today).

Finally, there have been sporadic attempts to build provider-less
networks that are build out of mobile wireless nodes only. It is quite
a challenge, and a number of breakthroughs over the last few years have
led to engineering solutions for Enabling Multihop Communication in
Spontaneous Wireless Networks, which are starting to look like a
viable alternative to managed wireless networks in some niche areas.

This is an exciting time to be learning about communications and
building and extending systems for communications. Planet Earth is not
far away from being 100\% connected, and the capacity and functionality
that has been achieved over the last few decades is quite astounding.
It does not look like we have hit any fundamental limits, nor
will for some time to come!

Read this book by some of the world's leading lights in the area 
of communications and enjoy.

### 2013-07-19 Compiled/intro/board.tex
editorial board and reviewers

\documentclass[a4paper,16pt]{article} 
\usepackage{a4wide,times,latexsym}
\usepackage[compact,tiny]{titlesec}
\usepackage{url}
\usepackage{times}
\usepackage{graphics}
\usepackage{color}
%\usepackage[pdftex]{hyperref}
\usepackage{multirow}

\pagestyle{empty}
\begin{document}

\begin{center}
{\Huge \textbf{Editorial Board}}

\Large 
\vspace{1in}
\begin{tabular}{ll}
Ernst Biersack &	Eurecom \\
Olivier Bonaventure &	Universit$\acute{e}$ catholique de Louvain \\
Jon Crowcroft	&  University of Cambridge \\
Walid Dabbous	  & INRIA Sophia Antipolis \\
Bruce Davie	& VMware \\
Anja Feldmann	& Technische Universit$\ddot{a}$t Berlin\\
Timothy Griffin	& University of Cambridge \\
Hamed Haddadi &	Queen Mary University of London\\
Ramesh Johari	 & Stanford University\\
Srinivasan Keshav &	University of Waterloo\\
Jean-Yves Le Boudec	& $\acute{E}$cole polytechnique f$\acute{e}$d$\acute{e}$rale de Lausanne\\
Jennifer Rexford &	Princeton University\\
David Wetherall	& University of Washington\\
Walter Willinger	& AT\&T Labs Research\\
\end{tabular}
\end{center}

\newpage

\begin{center}
{\Huge \textbf{Review Board}}

\Large 
\vspace{1in}

\begin{tabular}{ll}
Mark Allman &	International Computer Science Institute \\
Emmanuel Baccelli    &	INRIA Saclay\\
Saleem Bhatti	& University of St Andrews \\
Olivier Bonaventure &	Universit$\acute{e}$ catholique de Louvain \\
Gonzalo Camarillo  &  Ericsson Research   \\
Dah Ming Chiu   &   The Chinese University of Hong Kong \\
Costas Courcoubetis  & Athens University of Economics and Business \\
Mark Crovella & Boston University \\
Jon Crowcroft	&  University of Cambridge \\
Bruce Davie	& VMware \\
Benoit Donnet  &   Universit$\acute{e}$ de Li$\grave{e}$ge  \\
Damien Fay & Bournemouth University  \\
Paul Francis & Max Planck Institute for Software Systems \\
Hamed Haddadi &	Queen Mary University of London\\
Luigi Iannone &  TELECOM ParisTech \\
Srinivasan Keshav    &	University of Waterloo\\
Olaf Maennel  & Loughborough University \\
Konstantinos Nikitopoulos & University College London \\
Costin Raiciu &  Universitatea Politehnica Bucuresti  \\
Jennifer Rexford &	Princeton University\\
Miguel Rio & University College London \\ 
Matthew Roughan & University of Adelaide \\
Jean-Louis Rougier  &  TELECOM ParisTech \\
Damien Saucez  & INRIA Sophia Antipolis \\
Aman Shaikh  & AT\&T Labs Research \\
Steve Uhlig & Queen Mary University of London\\
\end{tabular}
\end{center}

\end{document}

### 2013-07-19 Compiled/intro/intro.tex
first stab

Professors often rely on textbooks to teach undergraduate and graduate networking courses. While there are many good introductory textbooks, there are very few books on advanced networking topics that could be suitable to graduate courses in networking. To fill this gap, SIGCOMM Education Committee has launched a community project to develop a high-quality, open-source, edited eBook on “Recent Advances in Networking”. This eBook will be distributed online via the SIGCOMM website.

This eBook is composed of nine chapters chosen after a highly selective review process by the editorial board. The selected chapters cover advanced networking topics and accompanying teaching material (slides and exercises). All the source code of the eBook and the teaching material are kept on a version controlled repository that will be accessible for the entire SIGCOMM community. We expect that releasing such high quality teaching material will be beneficial for a large number of students and professors. The teaching material will be updated on a regular basis to reflect new advances in our field. We will also be adding new chapters on more emerging topics in the future volumes.

We wish to thank all there authors for providing such high quality chapters. We are also grateful to the reviewers and the editorial board for spending many hours on each chapter to ensure a coherent level amongst all the chapters. We hope you enjoy reading this eBook and find it a useful resource.

Hamed Haddadi, Olivier Bonaventure

August 2013

### 2013-07-21 Compiled/intro/chapterlist.tex
add chapters

\documentclass[a4paper,16pt]{article} 
\usepackage{a4wide,times,latexsym}
\usepackage[compact,tiny]{titlesec}
\usepackage{url}
\usepackage{times}
\usepackage{graphics}
\usepackage{color}
%\usepackage[pdftex]{hyperref}
\usepackage{multirow}

\pagestyle{empty}
\begin{document}

\begin{center}
{\Huge \textbf{List of Chapters}}
\end{center}

\vspace{1cm}

\large

\begin{enumerate}
\item \textbf{Internet Topology Research Redux}\\
Walter Willinger, Matthew Roughan

\item \textbf{Recent Advances in Transport Protocols}\\
Olivier Bonaventure, Janardhan Iyengar, Costin Raiciu

\item \textbf{Internet Traffic Matrices: A Primer} \\
Paul Tune, Matthew Roughan

\item \textbf{Optimizing and Modeling Dynamics in Networks } \\
Ibrahim Matta

\item \textbf{Smart Data Pricing (SDP): Economic Solutions to Network Congestion} \\
Soumya Sen, Carlee Joe-Wong, Sangtae Ha, Mung Chiang

\item \textbf{MPLS Virtual Private Networks } \\
Luca Cittadini, Giuseppe Di Battista, Maurizio Patrignani

\item \textbf{Collaboration Opportunities for Content Delivery and Network \\ Infrastructures} \\
Benjamin Frank, Ingmar Poese, Georgios Smaragdakis, Anja Feldmann, Bruce M. Maggs, Steve Uhlig, Vinay Aggarwal, and Fabian Schneider

\item \textbf{The Design Space of Network Mobility} \\
Pamela Zave, Jennifer Rexford

\item \textbf{Enabling Multihop Communication in Spontaneous Wireless Networks} \\
Juan Antonio Cordero, Jiazi Yi, Thomas Clausen, Emmanuel Baccelli

\end{enumerate}

\end{document}

### 2013-07-22 Compiled/chapter1/ebook-sec3.tex
all files compile correctly

% This work is licensed under
% http://creativecommon.org/licenses/by/3.0/
\section{Two patterns for implementing mobility}
\label{sec:sec3}

In this section we show that there are two completely different patterns
for implementing mobility.
They differ in where the change of attachment appears with respect to the
implementing layer,
in which algorithms and protocols of the implementing layer 
are involved in implementing mobility, and in which parts of the
shared state of the implementing layer are altered.
They also differ
in their detailed design
decisions, and in their cost, performance, and scalability issues.

Not only are these patterns non-overlapping, they also completely
cover all implementations of mobility, in the sense that each
implementation either follows one pattern or is clearly a composition
of the two patterns.

\subsection{Dynamic-routing mobility}
\label{sec:drm}

Figure~\ref{fig:drm} has two stages depicting the effect of mobility
on an inter-layer channel.
Recall that the channel is a {\it link}
in the state of the layer that uses it, and a {\it session} in the
state of the layer that implements it; its {\it higher endpoints}
are members in the user layer, while its {\it lower endpoints} are
members in
the implementing layer.

\begin{figure}
\centering
\includegraphics[scale=0.95]{figures/drm.pdf}
\caption{Two stages in an instance of dynamic-routing mobility.}
\label{fig:drm}
\end{figure}

The precise site of mobility here is the lower endpoint {\it A}.
In Stage 1 {\it A} is registered at {\it a1} in
Underlay 1.
{\it a1} and {\it A} are connected to the rest of their layers
through Links 1 and 2, respectively.  
Link 2 is implemented by Underlay 1.

Between Stage 1 and Stage 2
Link 1 stops working, possibly because the machine on which
{\it A} and {\it a1} reside has
been unplugged from a wired subnetwork, or
has moved out of range of a wireless subnetwork.
In a cascading sequence of events, 
Link 1 is destroyed,
Link 2 is destroyed, and the registration
of {\it A} at {\it a1} is destroyed. 
{\it A} is now disconnected from the rest of its layer.

Eventually the mobile machine may become plugged into another
wired subnetwork
or enter the range of another wireless subnetwork, as shown in Stage 2.
In a cascading sequence of events, member {\it a2} (which is
the mobile machine's member in the new Underlay 2)
connects to the rest of its layer through Link 3,
{\it A} becomes attached to new location {\it a2}, and new Link 4 is
created in the mobility layer and implemented by Underlay 2.
Note that {\it A} is now linked to {\it C} rather than {\it B};
this change is necessary because {\it C} is attached to Underlay 2 and
{\it B} is not.

Between Stages 1 and 2 there may be an interval during which {\it A}
has no connection with the rest of its layer.
There may also be an interval in which Stages 1 and 2 overlap, so
that {\it A} is temporarily attached to both underlays.

The hard problem to be solved in Figure~\ref{fig:drm} is that 
even after {\it A} is again reachable by other members of its layer
such as {\it D} and {\it E}, they do not know how to find it because
the routes to it are obsolete.
{\it Dynamic-routing mobility} relies on the routing algorithm of
the layer, which must learn about new links, recompute routes,
and update forwarding tables.
After this is accomplished, {\it D} will know that it can reach
{\it A} by forwarding to {\it C}.

There are three ways in which actual dynamic-routing
mobility can differ from the
example in Figure~\ref{fig:drm}.
Fortunately, none of them affect what the implementation has to do,
so none of them need be discussed separately.
First, the new attachment {\it a2} could be in the same layer as 
{\it a1}, rather than in a different layer.
Because {\it a1} and {\it a2} are different locations,
after the move {\it A} is probably linked to
a different member of its own layer, even though the new link is
implemented by the same lower layer as before.

Second, in Figure~\ref{fig:drm} the mobile member {\it A} has only
one attachment and one necessary link.
As shown in Figure~\ref{fig:scope}, members such as gateways have
multiple simultaneous attachments to different underlays.
Because each such attachment is necessary for the gateway's
purpose and supports its own link or links,
the mobility of each attachment is a separate problem to be solved.

Third, occasionally a layer implements sessions for the benefit of
its own members, rather than as a service to a higher user layer.
In this case there is no {\bf A} or {\bf E}, and the beneficiaries of
the mobility implementation are {\it A} and {\it E}.

A {\it router} is a member of a layer that receives and forwards messages
not destined for itself, whether it sends and receives messages on its
own behalf or not.
A {\it forwarding table} is a distributed copy of some of the
{\it routes} state component of a layer.
Implementations of dynamic-routing mobility incur four kinds of
resource cost:
\begin{itemize}
\item
{\it storage cost} is the cost of storing routes to mobile members,
in the forwarding tables of all the routers that need them;
\item
{\it update cost} is the cost of updating the stored routes as mobile
members move;
\item
{\it path cost} is the cost of longer or more congested message paths
due to mobility;
\item
{\it handoff latency} is the message delay caused by a move.
\end{itemize}
These costs will be discussed further in 
Section~\ref{sec:majordifferences}.

The primary issue in implementing dynamic-routing mobility (DRM) is
that large layers such as the classic Internet core achieve scalability
through a hierarchical name space.
In the Internet core, names (IP addresses) are organized into a
hierarchy based on geographical, topological, and administrative
factors.
A layer member is assigned a name based on its location in this hierarchy.
Subtrees in the hierarchy correspond to blocks of names, and routing
scales because it operates on aggregated
blocks rather than individual names.
Mobility violates the rules of this scheme, because a mobile member
retains its name as it moves across the boundaries of the hierarchy.
If implemented naively, it would require a large number of entries
in the forwarding table of each IP router for individual mobile machines.

This issue is so important that the design decisions made to implement
DRM are completely different in hierarchical and non-hierarchical
layers.
For that reason, we have divided examples of DRM into two sections
(Sections~\ref{sec:sec4} and \ref{sec:sec5}).

\subsection{Session-location mobility}
\label{sec:slm}

Figure~\ref{fig:slm} has the same two stages as Figure~\ref{fig:drm}.
The most important
difference is that {\bf A}'s location in the implementing
layer changes from {\it A1} to {\it A2}, rather than staying the same
as it did in Figure~\ref{fig:drm}.
In geomorphic terms, the mobile machine's representative in the
implementing layer (with name {\it A1}) has died, and has been reborn
as a member of the implementing layer with name {\it A2}.

\begin{figure}
\centering
\includegraphics[scale=0.95]{figures/slm.pdf}
\caption{Two stages in an instance of session-location mobility.}
\label{fig:slm}
\end{figure}

This is a natural occurrence in a layer with a hierarchical name
space.
It should be 
familiar from observing what happens
when a laptop with an IP address {\it A1} moves to a new subnetwork
of the Internet, and gets a new IP address {\it A2} from DHCP.
The laptop cannot continue to use {\it A1} in the new subnetwork,
because {\it A1} is not in the subnetwork's address block.

DHCP alone is not sufficient to implement mobility, however.
As explained in Section~\ref{sec:layers.and.mobility}
and shown in Figure~\ref{fig:slm},
the strongest form of mobility requires preserving the 
communication channel in the user layer.
The bulk of the work of implementing session-location mobility lies in 
ensuring that {\bf A}'s correspondents know that it is now located at
{\it A2} rather than {\it A1}.
Each lower endpoint that was participating in a session with {\it A1}
on behalf of {\bf A} must be informed that it should now be corresponding
with {\it A2} instead.

As explained in Section~\ref{sec:layercomponents},
when an underlay is implementing a channel for an overlay, the initiating
lower endpoint must be able to look up the location of the 
accepting higher endpoint in the underlay, so that it can send messages
to it.
This means that there must be a globally accessible copy of the
{\it locations} mapping in the layer.
Session-location mobility also requires updating this mapping when
a higher endpoint moves.

Generally the fastest handoffs are achieved when a new lower endpoint
sends updates directly to all its correspondent lower endpoints
(in addition to updating the {\it locations} mapping).
This requires, of course, that the new lower endpoint have the 
correct name of the lower endpoint at the other end of each session.

Interesting behavior arises if both ends of a session move 
concurrently.
Neither lower endpoint will know the new name of the far endpoint,
so neither can send an update to the other.
In this simultaneous handoff
scenario a mobile endpoint, finding that it cannot reach a far
endpoint to update it, will suspect that the far endpoint has moved also.
Both endpoints must fall back on lookup from the {\it locations}
mapping to get the new location of the far endpoint.

As with Figure~\ref{fig:drm}, the two stages in Figure~\ref{fig:slm}
might have a gap between them or might overlap.
If they overlap, there will be an interval during which {\it A}
has two attachments in the same layer.

In Figure~\ref{fig:slm} the underlays are not shown,
although they probably look similar to those in Figure~\ref{fig:drm}.
Most likely there is an underlay member
{\it a1} that is destroyed, 
and an underlay member {\it a2} that is created.
There is no mobility observable at this level, however, because 
{\it A1} is attached to {\it a1} in Underlay 1 throughout its lifetime, and
{\it A2} is attached to {\it a2} in Underlay 2 throughout its lifetime.
The only mobility that is observable is {\bf A}'s change of attachment
from {\it A1} to {\it A2}.

Strictly speaking some dynamic routing could be involved in
session-location mobility, because
{\it A2} is a new member of the layer and there must be routes to it.
In practice this is rarely an issue, because the name
{\it A2} is part of some
larger block to which routes already exist.

Like DRM, session-location mobility (SLM) has storage costs,
update costs, and handoff latency.
The storage costs are the costs of maintaining a 
scalable implementation of {\it locations}.
The update costs are the costs of updating {\it locations} and
current correspondents when a member moves.

Implementations of SLM vary in a number of ways
(see Section~\ref{sec:sec6}),
although no one variation is as important as the hierarchical {\it versus}
non-hierarchical
variation for DRM.

\subsection{Major differences between the patterns}
\label{sec:majordifferences}

There are obvious structural differences between the two patterns:
\begin{itemize}
\item
In DRM the change of attachment appears between the implementing layer
and the level 
below it, while in SLM the change of attachment appears between
the user layer and the implementing layer 
(see Figures~\ref{fig:drm} and \ref{fig:slm}).
\item
In DRM the bulk of the work is performed by the routing algorithm,
while in SLM the bulk of the work is performed by the session protocol
and location algorithm (see Figure~\ref{fig:table}).
\item
In DRM the major state components that change are {\it attachments, links,}
and {\it routes} 
(see Figure~\ref{fig:table}).
In SLM the major state components that change are {\it locations}
and {\it sessions}. 
\end{itemize}
These structural differences prove that the two patterns are
fundamentally different.

In attempting to understand mobility mechanisms, people are sometimes
confused by the fact that {\it routes} (changed by DRM) and 
{\it locations} (changed by SLM) are both mappings.
The {\it locations} mapping is usually implemented by a shared global
data structure called a
{\it directory}.
The {\it routes} mapping is usually distributed across the forwarding
tables of the routers, but is occasionally implemented as a directory.
The result is that directories are sometimes used in both DRM and SLM
implementations.

This similarity is superficial because it does not tell
us the most important thing about
these mappings, which is what they mean in terms of
network architecture.
The mappings used in DRM and SLM are always fundamentally different, and
can always be distinguished from one another.
As mentioned in Section~\ref{sec:drm},
{\it routes} is a peer-to-peer or intra-layer mapping:
at each router, entries in the forwarding
table map each destination name to a member, link, or path {\it in the
same layer}. 
{\it Locations}, on the other hand, is always an inter-layer mapping,
mapping names in a higher layer to names in a lower layer.

In describing mobility mechanisms, people often focus on the
``identifier-locator split.''
This may be useful intuition, but should be interpreted carefully.
In an episode of mobility there is always a layer member that retains
its identity (the ``identifier''), and two members at a lower level,
where the attachment of the identifier moves from one to the other
(the ``locators'').
The identifier-locator split
does not distinguish DRM from SLM, although in the two patterns
the identifiers and locators appear at different levels.
In addition, it is important to remember that these terms are
relative, as mobility can occur anywhere in a layer hierarchy.

On the surface, it may seem that DRM should be called ``in-network
mobility'' or the like, while SLM should be called ``end-to-end mobility''
or the like.
This reflects a misunderstanding of how general the patterns are,
and how freely they can be applied at different levels.
For one example, consider an application layer whose members run
only on Internet
hosts.
The members include user clients and named services.
The layer could have its own dynamic,
application-specific routing to services,
which allows services to be reached even though they move from server
to server.
This instance of DRM is not ``in network'' from most peoples' perspective.
For another example, an Internet router might itself be mobile,
and might have some of its links to other Internet routers
preserved as it moves by session-location
mobility at a lower level.
This instance of SLM does not involve any endpoints according to most
peoples' perspective.

Obviously a quantitative comparison between two mobility implementations
cannot be made without implementation details and a profile of the
expected load.
Nevertheless, it is possible to make some general comparisons
between the two patterns based on their potential strengths and 
weaknesses.
We say ``potential'' because any characteristic, whether positive or
negative, can be irrelevant in some situations.

The greatest potential
weakness of DRM is its storage, update, and path costs.
Normally routing information is different in different places,
so there is a lot of it, it is spread widely across a layer, and it
is expensive to update.
Attempts to economize on storage and update costs can lead to high
path costs (see Section~\ref{sec:sec5}), as messages travel further
to be routed successfully.
Path costs must be weighted heavily because {\it every}
message that travels on a
channel is affected by its path cost, if any.

Locations are very different from routes because the result of a location
query is usually the same no matter which member is querying (in
contrast to a route, which is different depending on where it is
starting from), and because a location query is needed only at the
beginning of a session and possibly after a move (in contrast to routes,
which are consulted on every hop of every message).
As a result, locations can be stored and updated much more cheaply
than routes.
For example, even a centralized directory would perform adequately
in many contexts.
And even if lookup of a location is slow, we do not count it as a
path cost because the cost is incurred a few times for each channel
rather than 
being built into the cost of transmitting each message on the channel.

The greatest potential
weakness of SLM is that it must be implemented with
the participation of session endpoints.
This means that deployment of an SLM mechanism requires new or
upgraded mobile devices that run the SLM protocol for sending and
receiving location updates.
Full interoperation with legacy endpoints calls for expensive middleboxes.
Security is a concern because endpoint devices can initiate
updates of the global layer state.

Concerns such as software upgrading and security have attracted less
attention with respect to DRM.
This is because a layer can, in principle, be designed so
that its members are partitioned into endpoints and routers, and
only the routers need be aware of or participate in an implementation
of DRM.
In reality these concerns are ubiquitous in distributed computing, and
can apply to routers as well.

### 2013-07-22 Compiled/chapter1/ebook-sec5.tex
all files compile correctly

% This work is licensed under
% http://creativecommon.org/licenses/by/3.0/
\section{Examples of dynamic-routing mobility in hierarchical layers}
\label{sec:sec5}

Both of the designs in this section are intended for mobility within
the Internet core.  For this reason, both must grapple with the
problem of a hierarchical name space as explained in
Section~\ref{sec:drm}.
To reduce overhead, both solutions
significantly limit the number of routers in the Internet that must
store and update state concerning how to reach each mobile node.

\subsection{Mobile IPv4}
\label{sec:mipv4}

Mobile IPv4 \cite{mobile-IP,rfc3344} drastically reduces storage and update
costs by reducing the number of routers that must have a current route
to a particular mobile host to one or two.
Also, because each router is responsible for only a limited number of
mobile hosts, no router is over-burdened by mobility.

Figure~\ref{fig:mipv4} shows the path of a message from correspondent
host {\it C}
to mobile host {\it M} in an Internet core layer with Mobile IPv4.
Router {\it HA} is the {\it home agent} of {\it M}, and is supposed to
have a route to it at all times.
Router {\it FA} is the {\it foreign agent} of {\it M}, meaning that
it is local to the subnetwork where {\it M} is now attached,
and currently knows a route to {\it M} through the subnetwork.

\begin{figure}
\centering
\includegraphics[scale=0.90]{figures/mipv4.pdf}
\caption{The path of a message to mobile host {\it M} with Mobile IPv4.
Special links are drawn with heavier lines.
Only the links employed in the path are shown.}
\label{fig:mipv4}
\end{figure}

The IP address {\it M} is in an aggregated routing block such that
all messages destined for the block are routed to {\it HA}.
Thus this router need only be the home agent for mobile hosts with
IP addresses in its
block.
The message from {\it C} arrives at {\it HA} by means of normal IP
routing through router {\it R1}.
The subnetwork of {\it HA} is {\it M}'s home subnetwork, so when {\it M}
is at home {\it HA} has a local route to it.

When {\it M} is not at home and becomes
attached to the subnetwork of {\it FA}, it gets a local ``care-of''
IP address {\it CO}
in that subnetwork.
{\it M} informs {\it FA},
which informs {\it HA} that it is the current foreign agent of {\it M}.
To forward messages to {\it M}, however, {\it HA} cannot merely forward
them toward {\it FA}.
It they were sent out on normal IP links, normal IP routing would send
them back to {\it HA}!
Messages to {\it M} from {\it HA} and {\it FA} must be forwarded on
special links that are separate from normal IP links.

As shown in Figure~\ref{fig:mipv4}, the special links in the Internet
core are implemented by a tunneling layer below the core layer.
The home agents, foreign agents, and mobile hosts of the Internet core
are all registered at members of the tunneling layer.
Home agents and foreign agents have the same names in both layers,
while
{\it M} is attached to member {\it CO} in the tunneling layer.
To forward a message for {\it M}
on its special link, {\it HA} in the core layer encapsulates the
message in another message destined for {\it FA}, and passes the message
to member {\it HA} in the tunneling layer.

Although the tunneling layer resembles the core layer (see below), 
its state differs from that of the core layer in several important 
respects:
\begin{itemize}
\item
{\it Routes:}
In the core layer, 
at {\it HA} messages for {\it M} are forwarded to {\it FA} on a special 
link,
at {\it FA} messages for {\it M} are forwarded to {\it M} on a special 
link,
and everywhere else messages for {\it M} are forwarded to {\it HA} on 
a normal link.
In the tunneling layer {\it M} does not exist.
\item
{\it Attachments:}
Some
members of the core layer are attached to members of the tunneling layer.
\item
{\it Locations:}
The core layer has no {\it locations} state, at least not related to
Mobile IPv4.
Although the tunneling layer need not maintain explicit
{\it locations} state for
mobile routers because they have the same names in both layers,
it must maintain explicit {\it locations} state for mobile hosts from the
core layer.
This state, which supplies the current local IP address of a mobile host,
is stored in the foreign agent to which it is relevant.
\end{itemize}

In Mobile IPv4, mobile hosts such as {\it M} send messages to their
correspondents such as {\it C} though normal IP links.
This often creates problems because IP address {\it M} is not part of
the normal routing block of the subnetwork at {\it FA}.
If there is ingress filtering for security in or near this subnet,
messages with a source address of {\it M} will be thrown away.
In Section~\ref{sec:sec7} we shall see how Mobile IPv6 eliminates this
problem.

Overall this is an interesting architecture because 
the Internet core layer and the tunneling layer
are mostly identical, and share the same
implementation.
Home agents, foreign agents, and mobile hosts are all aware of the
differences between the layers and aware of their dual membership and
dual roles.
The shared implementation works because none of the other members
of the layers need to be self-aware in that way.
They always behave the same, without knowing that sometimes their
actions contribute to the core layer, while other times their actions
contribute to the tunneling layer.

By distinguishing clearly between the two layers,
we make it possible to check the correctness of the software for each.
It also becomes possible to make further distinctions if advantageous.
For instance, implementation of a
link between {\it HA} and {\it R2} can be shared by
both layers, but it might make sense to distinguish links in the two
layers for reasons of performance or accounting.

\subsection{MSM-IP}

MSM-IP \cite{multicast} is a proposal for using IP multicast
to implement mobility.
A mobile host gets an IP address {\it M}
in the distinguished multicast block.
When the mobile host
attaches to a new subnetwork using local IP address {\it L},
{\it L} joins the multicast group for {\it M}, and the previous
local address used by {\it M} resigns from the group.

With IP multicast there is a distinguished set of multicast routers,
which are globally distributed and 
are responsible for routing messages destined for a multicast
address to all members of the address's current multicast group.
These routers exchange information and forward messages to each other
through special links,
exactly as the routers participating
in Mobile IP do.
The special links are implemented by a tunneling layer, exactly as the
special links in Mobile IP are.

With MSM-IP, every subnetwork that supports either mobile hosts or
their correspondents must have a multicast router.
Messages to mobile hosts (or true multicast groups) are recognized
by their distinguished addresses and sent to their local multicast
router, where they enter the special multicast routing 
system.

\subsection{Comparative resource costs}

The costs of dynamic-routing mobility depend greatly on the number
of routers that must have a current route to each mobile host.
More routers incur more storage and update costs.
Storage and update costs are much greater for MSM-IP than for Mobile IPv4,
because an entire network of multicast routers must be updated on each
move.

Using fewer routers, on the other hand, incurs more path cost.
With MSM-IP path cost is minimal, because a message travels from the
multicast router in the sender's subnetwork, along an optimal path 
through the distributed multicast routers, to the multicast router in
the receiver's subnetwork.
With Mobile IPv4 path cost can be high, because each message to a mobile
host must pass through the home agent, regardless of where the sender
is and where the mobile host is.
This problem of path cost or ``triangular routing'' is the reason why
the designers of Mobile IPv4 decided to send messages from mobile hosts
through normal IP links.
They incur no path cost, but they do run afoul of security
filtering.\footnote{Messages from MSM-IP mobile hosts do not have
problems with security filtering because multicast IP addresses are
recognizable as belonging to a special category.}

In Section~\ref{sec:majordifferences} we said that dynamic-routing
mobility does not in principle require participation of the endpoints.
Mobile hosts in Mobile IPv4 and MSM-IP do not have this advantage.
The reason that they must have special behavior is that both designs
use special routing mechanisms, separate from normal IP
routing, to find mobile hosts.
Because the routing mechanism is special, it is not necessary to
update every IP router when a mobile host moves.
But also, because the routing mechanism is special, mobile hosts must also
behave differently to interact with it in the correct way.

### 2013-07-22 Compiled/chapter1/ebook-sec6.tex
all files compile correctly

% This work is licensed under
% http://creativecommon.org/licenses/by/3.0/
\section{Examples of session-location mobility}
\label{sec:sec6}

In this section we compare four proposals for session-location
mobility: the
Host Identity Protocol (HIP)~\cite{RFC-4423,HIP}, 
the Identifier-Locator Network Protocol (ILNP) \cite{ILNP1,ILNP2},
the
Locator/Identifier Separation Protocol (LISP) Mobile Node~\cite{LISP-MN}, 
and the ``route optimization'' mechanism of Mobile 
IPv6 \cite{mipv6old,mipv6new}
(Section~\ref{sec:mipv6} will explain how route optimization fits into
Mobile IPv6 overall).
All but ILNP are IETF standards, and ILNP has resulted in many
IETF documents with ``experimental'' status.

These proposals have many similarities, as
they all provide mobility by splitting 
the Internet core layer (shown in Figure~\ref{fig:scope}) into two
layers.
These two layers 
are shown in Figure~\ref{fig:identloc}, and also correspond to
the two layers in Figure~\ref{fig:slm}.
SLM supports the persistence of inter-layer channels that are
links in the upper layer and sessions in the lower
layer.\footnote{Note that the members of the
identifier layer are {\it hosts}, while the members of the locator
layer are {\it interfaces}.
This distinction can safely be ignored in this section, but it is
important for multihoming as discussed in Section~\ref{sec:multihoming}.}

\begin{figure}
\centering
\includegraphics[scale=0.80]{figures/identloc.pdf}
\caption{The Internet core layer splits into two layers for
well-known examples of session-location mobility.  
In the identifier layer, session protocols such as TCP run
largely unmodified.
In the locator layer, the session protocol implements mobility.}
\label{fig:identloc}
\end{figure}

In an attempt to use a widely acceptable common terminology,
we call the upper layer 
the {\it identifier layer}, and the lower layer
the {\it locator layer}.
Names in the two layers are referred to as {\it identifiers} and
{\it locators}, respectively.
This common terminology does not necessarily match the terminology
typically used to explain each specific protocol.

Note that
Figure~\ref{fig:identloc} is an approximation of the real
implementations of these standards, in which the split between layers may
be implicit or incomplete.
In the geomorphic view, two separate session protocols are employed.
In the identifier layer, a largely unmodified TCP implementation provides
the usual TCP service as if identifiers were IP addresses.
(UDP and other service protocols operate here 
as well.)
In the locator layer, the only purpose of the session protocol is
to implement SLM.

\subsection{Names}

Table~\ref{tab:slm} compares the four proposals on their choices of
names.
They differ most on identifiers, which must be globally unique and
persistent, but have no other necessary constraints.

\begin{table}
\begin{center}
\footnotesize
\begin{tabular}{|l|l|l|} \hline
\multicolumn{1}{|c|}{\bf Protocol} & 
  \multicolumn{1}{c|}{\bf Identifier} & 
  \multicolumn{1}{c|}{\bf Locator} \\ \hline
HIP & (hash of) public key & IPv4 or IPv6 address \\ \hline
ILNPv6 & 64-bit IPv6 suffix & IPv6 address \\ \hline
LISP Mobile Node & IPv4/IPv6 address (called EID) & IPv4/IPv6 address (called RLoc) \\ \hline
Mobile IPv6 & IPv6 address & IPv6 address \\ \hline
\end{tabular}
\end{center}
\caption{Comparison of SLM proposals on the basis of names.}
\label{tab:slm}
\end{table}

HIP places a great emphasis on building in security, so the identifier
of a host is the host's
public cryptographic key.
With the use of keys as identifiers, 
messages can have self-authenticating source information.
Self-authentication provides security within the SLM locator
update protocol (see Section~\ref{sec:updateprotocols}), while
the guaranteed presence of a public key makes it easy to protect
the channel data with encryption.
Identifiers can also be hashes of public keys, which allows for shorter
identifiers without sacrificing self-authentication.

ILNPv6 is the IPv6 version of ILNP.
Its identifiers are 64 bits long; a host usually chooses a unique
identifier for itself
by taking the 48-bit MAC address of one of its hardware
interfaces and using a standard algorithm to extend it to 64 bits.

The significance of 64 bits is that in IPv6 routing based on 
hierarchical names and aggregation, the longest possible prefix is
64 bits.
This means that at its finest-grained, IPv6 routing examines the
first 64 bits of an IPv6 address and points to a
subnetwork.
The IPv6 address still has a 64-bit suffix to map to an IPv6 interface
attached to the subnetwork.
In ILNPv6, 
identifiers are carried in the 64-bit suffixes of IPv6 addresses.
In other words, an ILNPv6 locator is derived from an ILNPv6 identifier
by prefixing 64 bits that indicate a subnetwork where the identified
host can be found.\footnote{This is different from ILNP terminology,
in which the 64-bit prefix is itself the ``locator.''} 
This scheme is very efficient in its use of address bits.

By basing identifiers on MAC addresses, which are globally unique,
ILNP ensures that hosts have unique identifiers without relying on any
administrative authority.
ILNP requires that, when a host joins a new subnetwork, it is allowed
to choose its own 64-bit address suffix.

LISP Mobile Node and Mobile IPv6 are less interesting.
In both cases, identifiers are normal routable IP addresses.

Naming choices have the biggest effect on the deployment opportunities
of a design.
Mobile IPv6 requires the deployment of IPv6.
HIP and ILNP require more changes to TCP because their identifiers
are not IP addresses.
Deployment of new protocols is usually incremental, which means
that upgraded hosts and subnetworks must interoperate with legacy
hosts and subnetworks.
This raises the following interesting question: if an IP-based
SLM protocol is interoperating with ordinary IP, does the ordinary Internet
layer coincide with the identifier layer or the locator layer?
Interoperation will work best if the ordinary Internet layer 
coincides with the identifier layer.
This composition of layers (making the SLM identifier layer and Internet
layer into one) will work best if SLM identifiers look like ordinary
IP addresses.

\subsection{Directories}

An implementation of SLM requires a globally accessible
implementation of {\it locations} in the locator layer, mapping 
identifiers to locators.
Table~\ref{tab:slm2} compares the four standards on their choices of
location directory or other mapping implementation.

\begin{table}
\begin{center}
\footnotesize
\begin{tabular}{|l|l|} \hline
\multicolumn{1}{|c|}{\bf Protocol} & 
  \multicolumn{1}{c|}{\bf Location Directory} \\ \hline 
HIP & DNS \\ \hline
ILNP & DNS \\ \hline
LISP Mobile Node & LISP subsystem \\ \hline
Mobile IPv6 & home agent for each host has its locator  \\ \hline
\end{tabular}
\end{center}
\caption{Comparison of SLM proposals on the basis of directories.}
\label{tab:slm2}
\end{table}

LISP Mobile Node inherits its directory mechanism from LISP \cite{LISP},
which is an IETF standard designed for a different purpose 
(multihoming of large-scale enterprise subnetworks), and not
originally intended for the support of mobility.
The directory mechanism is a special-purpose distributed subsystem
of directory servers.
While this requires a substantial initial investment, it does give
the deployer maximum freedom.
For example, different deployments could use almost any name space
as the set of identifiers.

Both HIP and ILNP usually make use of the Domain Name System (DNS) as a
scalable, highly available directory subsystem.
When they do, note that their use of DNS is different from the
ordinary use of DNS, which is to map application-level names
(domain names) to IP addresses (which, as we have seen, can sometimes
be interpreted as identifiers or locators).
An IP-oriented SLM mechanism requires a directory or the equivalent to map
IP-oriented
identifiers to IP-oriented
locators, which has nothing inherently
to do with application-level names.

DNS relies on the hierarchical structure of domain names both
for scalability of lookups and to manage the distributed
administration of DNS servers.
Consequently, a DNS lookup must begin with a domain name.
Thus when HIP and ILNP use DNS as their directory subsystem,
every mobile host must have a domain name that serves as a key for 
finding its current location, even though the domain name is not
in the name space of either of the relevant layers, and may not be
needed for any other reason.
For example, in the use of a client-server service, 
the server usually has a domain name while the client does not.
But if the client is mobile with HIP or ILNP, it must have a domain
name, known to the server, for this purpose alone.

For both HIP and ILNP
it is necessary to add new record types to those stored
by DNS servers, because the value being looked up is not always an
IP address.
Finally, the DNS server with 
the authoritative copy of a locator must send it out with
a small time-to-live, preferably zero.
Otherwise other DNS servers 
will cache the information for longer times, 
impeding responsiveness to changes of location.

The route optimization (SLM) 
mechanism of Mobile IPv6 is an adjunct to the Mobile IPv6
DRM implementation (see Sections~\ref{sec:mipv4} and \ref{sec:mipv6}).
Because the DRM implementation uses home agents, the SLM implementation
uses them also.
The current locator of an identifier can always be obtained from its
home agent.
Mobile IPv6 may be less reliable than other designs because 
a home agents is a single point of failure with respect to its
mobile hosts.
Home agents do not necessarily have the built-in redundancy and
high availability that the directories of the other designs have.

\subsection{Locator update protocols}
\label{sec:updateprotocols}

An implementation of SLM must have a protocol through which 
mobile nodes update the directory and their correspondents after
a move.
The protocol must have security to prevent updates from unauthorized
hosts.

It would take far too much space to report on how each proposal
meets these requirements.
Also, many standards provide a menu of implementation alternatives,
some of them better-documented than others.
In lieu of this detail, 
we will merely touch on a few of the design issues for SLM
protocols.

Even without the problem of simultaneous handoff (as introduced in
Section~\ref{sec:slm}), an update protocol can suffer from lost
or re-ordered messages.
If a correspondent node or directory
receives two different update messages from a
mobile host in the wrong order, it could retain an obsolete locator
for the mobile node.
If a correspondent node or directory
determines from sequence numbers that an
update message has been lost, it might wait forever for a retransmission
that will not come because the mobile node is somewhere else and will
not receive the retransmission request.
These bugs have been discovered in real SLM protocols \cite{serval-icnp}.
In general, the two techniques to rely on are (1) version numbers
rather than sequence numbers, and (2) some form of protocol verification
to insure against otherwise-almost-inevitable mistakes.

If an endpoint loses track of the session's other endpoint because
of simultaneous handoff, loss of update messages, or a protocol bug,
it can always get the current locator by making a new lookup
in the directory.
In general, an SLM protocol can be made more robust by having mobile
nodes report their locators to the directory frequently,
and having correspondent nodes refresh their cached locators
from the directory frequently.
This robustness comes at the cost of increased overhead in the
form of message traffic.

HIP uses a different method to solve the problem of simultaneous handoff.
When a mobile host moves, its old locator is adopted by a
``rendezvous server'' that keeps track of its new locator.
The rendezvous server intercepts control messages destined for the
old locator, and forwards them to the new locator.
Even when both endpoints of a session move at the same time,
their update messages will reach each other through rendezvous servers.
As always, data messages travel directly between hosts.

\subsection{Encapsulation}

Like Table~\ref{tab:overlays},
Table~\ref{tab:slm3} compares the standards on the basis of
how they encapsulate overlay messages as they travel through an
underlay.
Note that there are two versions of the Mobile IPv6 standard which
differ in this respect (the newer \cite{mipv6new} supersedes the 
older \cite{mipv6old}, so this
comparison is of academic interest only).

\begin{table}
\begin{center}
\footnotesize
\begin{tabular}{|l|l|} \hline
\multicolumn{1}{|c|}{\bf Protocol} & 
  \multicolumn{1}{c|}{\bf Encapsulation} \\ \hline 
HIP & encapsulation with IPSec Encapsulating Security Payload \\ \hline
ILNPv6 & none, identifier is extracted from locator \\ \hline
LISP Mobile Node & simple encapsulation \\ \hline
Mobile IPv6 (RFC 3775) & simple encapsulation  \\ \hline
Mobile IPv6 (RFC 6275) & Home Address destination option and Type 2 Header  \\ \hline
\end{tabular}
\end{center}
\caption{Comparison of SLM proposals on the basis of encapsulation.}
\label{tab:slm3}
\end{table}

Consider a message being sent from left to right in 
Figure~\ref{fig:identloc}, at a time when identifier
{\it I1} has locator {\it L3} and identifier {\it I2} has
locator {\it L2}.
In the simplest implementation, the message consists of a message
with source {\it I1} and destination {\it I2}, encapsulated
in a message with 
source {\it L3} and destination {\it L2}.
There are other possibilities, however, motivated by the desire
to conserve space in message headers.
This is a serious concern in IPv6, where each of the four address
fields is 128 bits long.

LISP Mobile Node and the original version of Mobile IPv6 use simple
encapsulation as above.
HIP does also, with the additional proviso that the message body
containing the identifiers is protected with IPsec.
In ILNP each identifier is a suffix of its current locator, so it
need not be sent separately.

In the revised Mobile IPv6 standard, there is an optimization 
apparently based on the observation that, most of the time, only
one of the endpoints of a session will be mobile.
For a stationary node, the identifier and locator are always the
same, and need not be sent twice.
So a message from the mobile node needs a source identifier and
not a destination identifier, and a message to a mobile node
needs a destination identifier and not a source identifier.

Now let us assume that {\it I1} is mobile and {\it I2} is stationary.
For messages from {\it I2} to {\it I1}, a destination identifier is 
needed.
The revised Mobile IPv6 standard uses
a special Type~2 header.
This is a kind of ``source routing'' header, allowing the source to
provide a list of destination
addresses through which the message must be routed.
The messages have destination list ({\it L3; I1}), where the second
hop from {\it L3} to {\it I1} is internal to the mobile host.
For messages from {\it I1} to {\it I2}, a source identifier is needed.
The extra source address {\it I1} is inserted using a special
``Home Address destination option.''
This option is an extension to IPv6 allowing an extra field in a
message header.
In this way, the extra identifiers can be added to messages only when
needed.

One might speculate that such a complex optimization would cause
trouble in the form of further, cascading complexities,
and this is indeed the case.
There are elaborate rules in \cite{mipv6new} for processing
messages so that IPsec works correctly:
each message must be processed partially with {\it I1} in the ordinary
source or destination field, and partially with {\it I1} moved to the
Type 2 header or Home Address destination option field.
Note that this is an interaction that has been explicitly recognized
and accommodated in the standard.
No one knows how many problematic interactions with other protocols,
caused by this optimization,
will be discovered if Mobile IPv6 comes into widespread use.

The SLM implementation of Mobile IPv6 is constrained by the need to
compose with the DRM implementation of Mobile
IPv6 (see Section~\ref{sec:mipv6}).
If there were not so many constraints,
the size of
headers could be reduced without violating the principle of separation
of concerns.
For example, Figure~\ref{fig:identloc2} revises
Figure~\ref{fig:identloc} in an obvious way as suggested by the
geomorphic view.
The session protocol in the locator layer now performs both TCP and SLM
functions.

\begin{figure}
\centering
\includegraphics[scale=0.80]{figures/identloc2.pdf}
\caption{A more efficient version of Figure~\ref{fig:identloc}.}
\label{fig:identloc2}
\end{figure}

In the identifier layer, the link between {\it I1} and {\it I2}
is uniquely identified at one end by {\it I1, p1}, where {\it p1}
is a port number, and uniquely identified at the other end by
{\it I2, p2}.
Note that
the identifier layer is more like an application layer than an IP
layer in that it has no forwarding---only direct links between pairs
of communicating endpoints.
Consequently, once the link has been set up, there is absolutely
no need to transmit {\it I1} and {\it I2} in data messages.
Each endpoint simply uses the port number to pass messages unambiguously
across the layer boundary.
Figure~\ref{fig:identloc2} has many similarities with TCP Migrate
\cite{tcp-migrate} and Serval \cite{serval,serval-icnp}, which are other
proposals for SLM.

### 2013-07-22 Compiled/chapter1/ebook-sec7.tex
all files compile correctly

% This work is licensed under
% http://creativecommon.org/licenses/by/3.0/
\section{Composition of the patterns}
\label{sec:sec7}

\subsection{Structural modeling}
\label{sec:structural}

The geomorphic view of networking is rigorously defined and can
be formalized.
We have formalized various aspects of the geomorphic view in
Alloy, which is the modeling language of the Alloy 
Analyzer \cite{alloy-book},
and in Promela, which is the modeling language of the Spin model
checker \cite{Spin}.
These are {\it structural models}.

Networking researchers and practitioners are accustomed to {\it analytical models},
which are also formal, but quantitative rather than structural.
One can assign numbers to some symbols in an analytical model,
give the numbers and model to a suitable evaluator, and receive
numbers for other symbols in the model.

Structural models are similar, 
except that evaluation is logical rather than quantitative.
We assume that some formulas in a model are true,
give the assumptions and the model to a suitable evaluator such
as the Alloy Analyzer or Spin,
and receive information about the truth of other formulas.
We either learn that a formula is true, or get a counterexample
showing why it is not true.

Sections~\ref{sec:sec3} through \ref{sec:sec6} should have made
clear why the term we use to describe these models is
{\it structural}.
We use them to describe hardware and software structures within
networks,
and to compare mechanisms based on where their structures are
similar and different.
We also use them to show how structural decisions constrain other
decisions and affect important properties such as scalability
and interoperability.
In the next subsection, we will mention some additional knowledge
gained with the help of structural modeling.

\subsection{Generating the design space of mobility}
\label{sec:design}

An instance of mobility is an isolated
episode in which one layer member changes
its attachment from one underlay member to another.
One of our goals is to give network architects the freedom to handle
any instance of mobility with any mobility pattern at any level of
the layer hierarchy.
This should enhance efficiency and scalability by
allowing solutions that are finely tuned to the characteristics of
the problem they are solving.

The first step was to identify the two possible implementation patterns
and to provide sufficiently abstract versions of them
(Section~\ref{sec:sec3}).
The next step, taken in this section, is to show that any instance of
mobility can be implemented with either pattern at almost any level of
the layer hierarchy.

In the left column of Figure~\ref{fig:lift}, top half, we see a
fundamental instance of mobility in which the old and new locations
are in the same layer at Level 0.
As notated, the channel at Level 1 can be preserved by session-location
mobility (SLM) at Level 0.
In the left column, bottom half, we see a
fundamental instance of mobility in which the old and new locations
are in different layers at Level 0.
As notated, a channel at Level 2 can be preserved by dynamic routing
mobility (DRM) at Level 1.

\begin{figure}
\label{sec:fig:lift}
\centering
\includegraphics[scale=0.75]{figures/lift.pdf}
\caption{Generating the design space of mobility.}
\label{fig:lift}
\end{figure}

The middle column of the figure shows the effects of a ``lifting''
transformation in which each mobility implementation is moved up a level
in the hierarchy.
The purpose is to show that mobility can be implemented in many different
places, if the current architecture allows it or the designer has
control of the content and design of relevant layers.
In each case member {\it m} at Level 1 is replaced by two members
{\it m1} and {\it m2}.
Neither {\it m1} nor {\it m2} is mobile, as each has a stationary
registration in Level 0 throughout its lifetime.
Now member {\it m'} at Level 2 is mobile.
As shown in the figure (top), a channel in Level 2
with {\it m'} as its higher endpoint
can be preserved by SLM at Level 1. 
Or, as shown at the bottom, a channel in Level 3 with {\it m''} as
its higher endpoint and {\it m'} as its lower endpoint can be preserved
by DRM at Level 2.

The right column of the figure shows where one implementation pattern
can be replaced by the other.
To replace SLM by DRM (top right), it is necessary to lift the channel
up one level.
To replace DRM by SLM (bottom right), the channel can stay at the same
level, but the mobility must be lifted up a level.

Figure~\ref{fig:lift} illustrates the crucial point that mobility
is about name spaces and member identities in individual layers,
and about the mappings between these concepts in adjacent layers.
Identity is a fluid concept in software systems, which is why mobility
is fluid, and can be pushed around an architecture to appear in
different places.

If mobility is fluid, and different implementations are used for
different purposes at different levels of an architecture,
it follows that any layer could include implementations of either or
both mobility patterns.
Is this a problem?
We have proved that it is not, at least for implementations of the
patterns as modeled in Alloy.
Verification with the Alloy Analyzer shows that the two patterns
can be freely composed, in the same layer or
different layers of the same hierarchy.
They will work together without interference or other
undesirable interactions \cite{cnm}.

The limitation of this theoretical result is that,
to benefit from proven
compositionality,
implementations must maintain the minimal separation of concerns
inherent in our model of the geomorphic view.
If real implementations have state dependencies or interfering actions
that are not represented in our model, they are not necessarily
compositional even if the theorem says they are.
In effect the model is presenting sufficient
conditions for compositionality, which can be used as design guidelines
for real systems.

\subsection{Composition in Mobile IPv6}
\label{sec:mipv6}

As we saw in Section~\ref{sec:sec5}, Mobile IPv4 is an instance of DRM.
Mobile IPv6 \cite{mipv6old,mipv6new} uses a similar DRM mechanism, and
also composes it with the SLM mechanism described in 
Section~\ref{sec:sec6}.
We first consider the DRM mechanism.

Figure~\ref{fig:mipv6drm} is the Mobile IPv6 version of 
Figure~\ref{fig:mipv4}.
Note that Mobile IPv6 has no foreign agents, as their functions are
performed by the mobile hosts themselves.
In the figure, both the old attachment of {\it M} at {\it CO1}
and its new attachment at {\it CO2} are shown.
Normal routers are not shown, being replaced by ellipses in the
paths consisting of normal links.

\begin{figure}
\centering
\includegraphics[scale=1.00]{figures/mipv6drm.pdf}
\caption{The paths of messages to and from mobile host {\it M} with
the dynamic-routing mobility mechanism of Mobile IPv6.
Special links are drawn with heavier lines.
Only the links employed in the path are shown.}
\label{fig:mipv6drm}
\end{figure}

Figure~\ref{fig:mipv6drm} also differs from Figure~\ref{fig:mipv4}
in showing the source and destination addresses of the messages on
every link (source on top, destination below).
Thus a message in the core layer with source {\it C} and
destination {\it M} is forwarded on a special link from {\it HA} to
{\it M}.
The implementation of this special link in the tunneling layer 
encapsulates the message in another message with source {\it HA}
and destination {\it CO2}, and sends it through normal IP links and
routers.

Figure~\ref{fig:mipv6drm} also differs from Figure~\ref{fig:mipv4}
in showing the paths of return messages from {\it M} to {\it C}.
In contrast to MIPv4, return messages from {\it M} travel on a
special link as far as {\it HA}.
At {\it HA} they enter the realm of normal links and routers.
There is no problem with ingress filtering because {\it M} belongs to
the address block of {\it HA}'s subnetwork, so {\it M} is a normal
source address at that location.

The big problem with Mobile IP is the path cost of routing every
message through a mobile host's home agent.
Path cost is even worse in Mobile IPv6 than in Mobile IPv4, because
it is incurred by messages {\it from} a mobile host as well as
messages {\it to} it.
To reduce this problem, Mobile IPv6 standardizes a version of
SLM called ``route optimization,'' as already presented in
Section~\ref{sec:sec6}.
SLM is used only after the session between {\it C} and {\it M}
is established, and only if both endpoints have the protocol capability.

Figure~\ref{fig:mipv6slm} shows the SLM mechanism of Mobile IPv6 in
the same context as its DRM mechanism in Figure~\ref{fig:mipv6drm}.
For simplicity, this figure uses simple encapsulation as in 
\cite{mipv6old}.
After the endpoints have exchanged messages to set up SLM,
they send messages on special links implemented in the tunneling layer.
Note that these are {\it different} special links than those used
by DRM (Figure~\ref{fig:mipv6drm}).
The DRM special links involve {\it HA} and will change when {\it M}
moves.
The SLM special links do not involve {\it HA}, and will not change
from the perspective of the core layer when {\it M} moves.
With SLM, the only role played by {\it HA} in the tunneling layer is to
store the directory entry for {\it M}.
It is not needed originally because when SLM begins the two endpoints
are already connected and know each other's locations in the tunneling
layer, but it may be needed in case of simultaneous handoff.

\begin{figure}
\centering
\includegraphics[scale=1.00]{figures/mipv6slm.pdf}
\caption{The paths of messages to and from mobile host {\it M} with
the session-location mobility mechanism of Mobile IPv6.
Special links are drawn with heavier lines.}
\label{fig:mipv6slm}
\end{figure}

Figure~\ref{fig:mipv6slm} also shows the SLM mechanism of Mobile IPv6
in the same context as Figure~\ref{fig:identloc}.
The ``Internet core with Mobile IPv6'' layer in
Figure~\ref{fig:mipv6slm} is the same as the identifier layer in
Figure~\ref{fig:identloc}.
The tunneling layer in
Figure~\ref{fig:mipv6slm} is the same as the locator layer in
Figure~\ref{fig:identloc}.

So far our discussion of the Internet core/identifier
layer has been limited
to its links.
We have seen that {\it C} and {\it M} may have a choice of links
over which to send their messages.
Because the network path associated with each link is different,
its performance may be different.
It is the job of TCP in the Internet core/identifier layer to
smooth over any difficulties caused by diverse paths, for example
by ensuring that messages are delivered to an application layer in
FIFO order.

The design space generated in
Section~\ref{sec:design} is intended primarily for compositions
in which different instances of mobility are managed by different
mechanisms or at different levels of the architecture.
Composition in Mobile IPv6 is a little different because the exact same
instance of mobility---pictured in the figures as {\it M}'s change
of attachment from {\it CO1} to {\it CO2}---is being handled simultaneously
by both forms of mobility.
The DRM implementation is in the core/identifier layer, while the SLM
implementation is in the tunneling/locator layer.
DRM is the default mechanism, because SLM can only be used if both
endpoints are SLM-enabled.

### 2013-07-22 Compiled/chapter1/ebook-sec8.tex
all files compile correctly

% This work is licensed under
% http://creativecommon.org/licenses/by/3.0/
\section{Design considerations related to mobility}
\label{sec:sec8}

Mobility mechanisms lie close to the heart of networking, so they are
related to many other communication services and aspects of
networking.  In this section, we touch briefly on several other topics
closely related to mobility.

\subsection{Multihoming}
\label{sec:multihoming}

Increasingly, mobile devices connect to the Internet
via multiple interfaces ({\it e.g.,}
a laptop with WiFi and wired Ethernet
interfaces, a smartphone with WiFi and cellular interfaces, or a
virtual machine running on a physical server with multiple wired
Ethernet interfaces).  
Since these interfaces usually connect to
different administrative domains ({\it e.g.,}
a campus WiFi network and a
commercial cellular provider), they must have different IP addresses in
different address blocks.
There is no name for the Internet host itself, and no way to
route to the host itself rather than a specific one of its interfaces.

{\it Sequential multihoming} is the use by a host of multiple
interfaces, one after the other, during the lifetime of a channel.
A mobility mechanism can be used to implement sequential multihoming;
in principle either of the mobility patterns
can be used to provide it.

If we look at the specific real mobility protocols 
in Sections~\ref{sec:sec4} through \ref{sec:sec6}, however,
we see that the DRM protocols in Sections~\ref{sec:sec4} and 
\ref{sec:sec5} work with a single IP address per host, while the
SLM protocols in Section~\ref{sec:sec6} work with multiple IP addresses
per host.
This is an artifact of where these protocols sit in the IP stack rather
than an inherent property of the implementation patterns,
but it does mean that these SLM implementations are currently a better
match for multihoming than these DRM implementations.

{\it Simultaneous multihoming} allows a host to use its multiple
interfaces to contribute bandwidth to the same channel simultaneously.
Simultaneous multihoming
is closely related to sequential multihoming and therefore
to mobility, but not strictly the same because the multiple
interfaces of a host might not be allowed to {\it change} during
the lifetime of a channel.
Of course, what we really want is simultaneous {\it and} sequential
multihoming, which is a
little more complex than sequential multihoming because it requires
protocol extensions so that a layer member can
have and use more than one attachment at a time. 
All of HIP, ILNP, LISP Mobile Node, and Serval
already have session protocols capable of simultaneous multihoming.

Because there is such a close association between host multihoming
and mobility, the well-known multihoming protocols
Multipath TCP~\cite{RFC-6824} 
and the Stream Control Transmission Protocol (SCTP)~\cite{RFC-4960}
are worth studying.
They have been used enough to gather experience on the performance
aspects of switching from one interface to another, and on how performance
affects the buffering and rate control in real transport (session)
protocols such as TCP.
This experience is as relevant to mobility as it is to multihoming.

\subsection{Anycast}

Increasingly, the Internet is a platform for users to
access services hosted on multiple servers in multiple locations.
The appropriate network abstraction for their requirements
is {\it anycast},
in which a service has an {\it anycast name} that corresponds
to a group of servers offering the service.
A request for a communication channel to the service can result in a
channel to any member of the group of servers.

For simple query-response services like DNS,
all server replicas can share a single locator ({\it i.e.,} an IP address),
and rely on IP routing to direct client requests to one of the server
replicas.  
However, IP routing does not guarantee that multiple
messages sent to the same IP address would reach the same server
replica. 
In today's Internet, a domain name
for a geo-replicated service
can map to multiple IP addresses, one for each server replica.
This supports communication services with multiple message exchanges
between client and server.
It is different from mobility, however, because the
higher-level domain name has nothing to do with the channel after
the initial DNS lookup.

Alternatively, anycast could be combined with SLM, as it is in
Serval~\cite{serval,serval-icnp}.
A Serval identifier is an anycast name, and is registered as
located at a dynamic group of servers in the locator layer.
When a request for a channel to an identifier is 
handled by the Serval session protocol,
the protocol selects the locator of some member of the
group.
In contrast to the previous paragraph, the identifier remains
the name of the channel's higher endpoint throughout its lifetime.
The SLM session protocol can thus maintain the channel through both
mobility events and changes to the membership of the server group.

\subsection{Subnetwork mobility}

Mobility proposals typically focus on the
movement of a single mobile endpoint, like a mobile device or a virtual
machine.  However, in some scenarios a subnetwork serving
multiple endpoints can move from one location to another.  
For example, a
fast-moving bus, train, or plane may carry a LAN
that provides network
connectivity to a large collection of passengers.

Dynamic-routing
mobility in a hierarchical layer
naturally handles subnetwork mobility by updating the routes used
to reach the entire aggregated block of names.
For example, Boeing had an early in-flight WiFi service
that provided seamless mobility for airline passengers by associating
each international flight with an IP address block, and announcing the
block into the global routing system at different locations as the
plane moved~\cite{connexion}.  However, this solution required all
interdomain routers in the Internet to store and update fine-grained
routing information, leading to high overhead.

In our recent work~\cite{cnm}, we have shown that subnetwork
mobility is merely the mobility of the gateway's attachment to the
larger network, and is implemented with the same two patterns as
mobility of endpoints.
We identified several applications of the design
patterns that seem promising
for handling combinations of subnetwork mobility and endpoint
mobility. 
These solutions have the property that the mechanism for subnetwork
mobility (a bus moves its access point from one roadside LAN to another)
is completely independent of the mechanism for endpoint mobility
(a user with a laptop gets on and off the bus).
Nevertheless, it is not yet clear which solutions for
subnetwork mobility would be most viable in practice.

\subsection{Incremental deployment and interoperation}

Deploying new protocols that span
administrative domains is always challenging, since the Internet
is a federated
infrastructure and cannot easily have everyone upgrade to a new
protocol at the same time.
Most real deployments are DRM mechanisms that
operate within a single administrative domain ({\it e.g.,}
cellular networks, Ethernet LANs, or data-center networks),
or require support only from the mobile endpoint and a small number of
routers ({\it e.g.,} Mobile IPv4).

It is not surprising that most real mobility implementations use DRM,
because SLM entails many more deployment hurdles.
There can be a new set of identifiers, a new global directory service,
changes to both endpoints, and even
changes to the service interface that are visible to applications.
The early SLM protocol TCP Migrate \cite{tcp-migrate} is probably
the most deployable, requiring only changes to the operating system
at the participating endpoints, but even so it has not had significant
deployment.

Nevertheless, as noted in the introduction,
the pressure for better network mobility support is mounting.
Ubiquitous computing may be a particularly powerful motivator,
because an enormous number of sensors and actuators will require
network access.
This could accelerate the adoption of IPv6, enabling many
other changes in its wake.

For incremental deployment, an SLM-enabled host can interoperate
with a legacy host in a degraded mode in which mobility does not work
but other functions do.
For full mobility, an SLM-enabled host can interoperate with
a legacy host through a proxy or other middlebox.
This raises many new questions concerning how a middlebox is
introduced into the path between the hosts, and on the scalability
of stateful middleboxes.
These new questions must be added to the perennial list of old
interoperability questions, such as how to traverse
NAT boxes.
Identifying
effective ways to deploy these protocols incrementally remains an
active area of research and standards work.

\subsection{Security}

All mobility solutions raise important questions about
security.  
In DRM, who is authorized to announce
routing changes for an address or address block?  
In SLM, who is authorized to update the directory service and a
mobile endpoint's correspondents?  
Answering these questions
successfully requires unforgeable notions of identity, and secure
protocols for sending update messages to routers, directory servers,
and endpoints.  
Can accidental misconfigurations or malicious attacks
overload the routers, directory servers, or endpoints?  Preventing
denial-of-service attacks requires effective ways to limit the work
performed before recognizing that messages are unauthorized.

As
mentioned in Section~\ref{sec:majordifferences}, some people believe that
SLM protocols face greater security challenges
because arbitrary endpoints can initiate updates to global layer state.
On the other hand, SLM protocols provide for persistent identifiers
which can be used as the basis for authentication of hosts, a valuable
assistance to security.
SLM protocols that use DNS as the directory
service can update DNS
records with an existing secure protocol~\cite{RFC-2137}.  
Some protocols, like HIP and Serval, embed
an endpoint's public key (or a hash of the key) in the identifier; 
these identifiers support secure communication as well as
authentication.
Still, security is a rich and
important topic warranting a much deeper treatment, especially since
new protocols can easily introduce unforeseen vulnerabilities and new
threats.

### 2013-07-22 Compiled/chapter1/ebook-sec9.tex
all files compile correctly

% This work is licensed under
% http://creativecommon.org/licenses/by/3.0/
\section{Conclusion}
\label{sec:sec9}

In this chapter we have presented an abstract framework for describing, 
understanding, and comparing approaches to network mobility.  
As illustrations, we have covered several mobility protocols 
in some detail.
We believe the geomorphic model provides a clear and precise way to understand the considerable similarities between different mobility proposals,
allowing discussions to focus on their meaningful distinctions rather than artificial differences in terminology.

We have compared mobility proposals on both qualitative
(deployment constraints, security) and quantitative (resource costs,
latency) criteria.
The basis for making comparisons has been completely {\it structural},
in the sense of {\it structural modeling} as defined in 
Section~\ref{sec:structural}.
This is important because structural comparisons are vastly easier
to obtain than comparisons based on simulation, and should always
be the first step in any evaluation project.

In the interest of brevity, our discussions of quantitative
criteria have merely suggested trends and trade-offs, rather than
providing a more substantive analysis.
A true understanding of metrics such as
storage cost, update cost, and path cost requires a more 
detailed characterization of the proposals, including
supporting technologies such as routing protocols and directory services. 
Scalability depends on how these costs grow as the size of a network
grows within the expected range.

Equally important, different mobility mechanisms can be composed.
Even today it would not be surprising to see
dynamic-routing mobility used 
within an administrative domain, while
session-location mobility is used simultaneously
across administrative boundaries.
The ultimate goal would be to compose performance models along with
the mechanisms they are modeling, so that the performance of
a composed solution could be derived from the performance of its
components.

Today, many Internet applications that could benefit from mobility
use work-arounds instead, satisfying the need for session continuity
with {\it ad hoc} application-specific mechanisms, or simply doing
without \cite{handley}.
This is both an effect and a cause of scant deployment of mobility
mechanisms.
Most existing mobility protocols operate at fairly low levels
in a network architecture, 
specifically the link, network, and transport levels
of the classic Internet stack.
At these levels they are expensive, difficult to deploy, or both.

Many of these limitations are unnecessary, as
the essence of mobility is simply a dynamic binding of more abstract
names to more concrete names.  
As such, mobility can be easily implemented in middleware as a service to
even higher-level application layers.  
We believe that this is a fruitful avenue for further exploration,
particularly because it might be easier to optimize narrowly-targeted
implementations of mobility.

More generally, we believe that the geomorphic view promotes
common terminology, modularity,
separation of concerns, discovery of design patterns,
composition,
rigorous reasoning,
and code reuse in networking.
While widely appreciated by software engineers, 
these concepts have been less central to the study of networking.  
We believe that the geomorphic view should be extended to understand
other important aspects of networking.
We also believe that an appreciation of these concepts would be 
valuable for networking researchers and practitioners alike, 
far beyond the treatment of any one subject like network mobility.

### 2013-07-22 Compiled/chapter2/chapter2.tex
all files compile correctly

\newcommand{\CLASSINPUTinnersidemargin}{0.7in}
\newcommand{\CLASSINPUToutersidemargin}{0.7in}
\documentclass{article}

\usepackage{graphicx}
\usepackage{multirow}

\usepackage{url} % to include URLs in bib entries

\title{Optimizing and Modeling Dynamics in Networks}
\author{{\sc Ibrahim Matta} \\
Computer Science Department \\
Boston University \\
Boston, MA 02215, USA}
\date{\today}							% Activate to display a given date or no date

\def\pd#1#2{\frac{\partial#1}{\partial#2}}   % my own macros
\def\lg{{\rm log}}
\def\L{{\cal L}}
\def\sn{{\rm sin}}
\def\cs{{\rm cos}}
\def\lm{{\rm lim}}
\newcommand{\Conv}{\mathop{\scalebox{1.5}{\raisebox{-0.2ex}{$\ast$}}}}%
\def\junk#1{}

\begin{document}
\maketitle

\tableofcontents

%%%%%
\section{Introduction}

The Internet has grown very large. No one knows exactly how large, but rough estimates indicate billions of users (around 1.8B in 2009, according to eTForecasts \cite{eTForecasts}), hundreds of millions of web sites (over 200M in February 2009, according to Netcraft \cite{Netcraft}), and hundreds of billions of web pages (over 240B, according to the Internet archive \cite{Internet-archive}).
The Internet is also very dynamic --- users log in and out, new services get added, routing policies change, normal traffic gets mixed with denial-of-service (DoS) attack traffic, etc.

An important question is: {\em How do we manage such a huge and highly dynamic structure like the Internet?}
As a corollary, how can we build a network of the future unless we understand the {\em steady-state} and {\em dynamics} of what we build?

In this chapter, we resort to two mathematical frameworks: {\em optimization theory} to study optimal steady states of networks, and {\em control theory} to study the dynamic behavior of networks as they evolve toward steady state. Our emphasis will be on {\em congestion control} using the notion of {\em prices} to model the level of congestion, such as delays and losses, observed by users or traffic sources.

\paragraph{Expected Background:} We assume basic background in calculus and algebra.
We also assume basic knowledge of systems modeling, optimization theory, Laplace transforms, and control theory --- Keshav's textbook \cite{Keshav:2010} provides an excellent source for these mathematical foundations,
in particular, chapters 4, 5, and 8.
Basic knowledge of Internet's Transmission Control Protocol (TCP) \cite{tcp:1988},
namely Reno and Vegas \cite{vegas:1995} versions,
as well as queue management schemes, namely Random Early Drop (RED) \cite{red:1993},
should be helpful.
This chapter briefly covers needed background material to serve as a refresher or quick reference.
The material of this chapter has been used at Boston University in a second (advanced) networking course
taken by senior undergraduate and graduate students.

\paragraph{Contribution and Outline:}
The purpose of this chapter is to make the application of optimization and control theory to congestion control
more accessible through intuitive explanations and simple control applications, 
using examples from Internet's protocols.
This chapter has been largely influenced by the work of Frank Kelly \cite{Kelly-2001}, 
which introduces the notion of ``prices" and ``user utility",
the work by R. Srikant \cite{Srikant:2004},
which discusses the dynamics of user (traffic source) and network adaptations,
and control theory texts and notes (e.g., \cite{Ogata:2010,Lu:2001}). 
The exposition here attempts to tie these various mathematical models and techniques
through simple running examples and illustrations, 
modeling the dynamics of both flow control and routing.

We start by motivating the network control problem using analogy to 
the problem of producing, pricing, and consuming gas/oil (Section~\ref{sec:optimization}).
We introduce several examples of optimally allocating resources (link bandwidth) to users (traffic sources),
resulting in different notions of fairness. 
We then introduce dynamic equations that model source and link adaptation algorithms (Section~\ref{sec:control}).
Since these are generally non-linear equations,
we review the technique of linearization and how classical (linear) control theory can be used
to study stability and transient performance (Section~\ref{sec:linear}).
We use as a running example, a Vegas-like system controlled using different types of well-established controllers. 
Using linear approximation around a certain operating point, 
we can only study so-called {\em local} stability.

To study general ({\em global}) stability of non-linear models, 
we introduce the control-theoretic Lyapunov method (Section~\ref{sec:nonlinear}).
We also show how the control-theoretic Nyquist stability method can be applied
to the linearized model to study the impact of delay in feedback (i.e., measurements of the current state of the system).
The material on the Nyquist method is a bit more advanced and can be skipped on a first reading.
We generalize the notion of stability by introducing the concept of contractive mapping,
and extend its application to routing dynamics (Section~\ref{sec:routing}).

Finally, 
we provide two case studies that apply control-theoretic techniques introduced in this chapter:
the first study investigates stability under class-based scheduling
of rate-adaptive traffic flows (Section~\ref{sec:case1}),
and the second study investigates stability of data transfer over 
a dynamic number of rate-adaptive transport connections (Section~\ref{sec:case2}).
These case studies can be skipped on a first reading.

The chapter concludes with a set of exercises (Section~\ref{sec:exercises})
and their solutions  (Section~\ref{sec:solutions}).

%%%%%
\section{Network Control as an Optimization Problem} 
\label{sec:optimization}

In this section, we describe Frank Kelly's optimization framework \cite{Kelly-2001}, which models users' expectations (requirements) with utility functions, and network congestion signals (e.g., loss, delay) as prices. The network is shown to allocate transmission rates (throughputs) to users (flows) in such a way as to meet some {\em fairness} objective.

The objective of a user, or what makes the user happy, can be mathematically modeled as a {\em utility function}. For example, drivers observe the``price" of transportation and make one of many possible decisions: drive, take the subway instead, walk, bike, or stay home. The decision may involve several factors like the price of gas, convenience, travel time, etc. 
For example, if it rains, you might decide to drive to work, or you might decide to walk to work to save money and can then afford to go to the movies later in the week. Of course, how much driving a person does, is affected by all sorts of factors and user priorities are unknown to the system of gas stations and oil companies. But, each driver has her own utility!

Figure \ref{fig:gas-control-loop} illustrates with a block diagram the closed-loop relationship between drivers (users), gas stations (where gas is sold to and consumed by users), and the market (which represents OPEC\footnote{OPEC is the Organization of the Petroleum Exporting Countries.}, the government, and oil companies that collectively produce gas and set market prices based on user demand). Drivers set the total demand by observing gas prices. Notice that the gas price includes at-the-pump gas price, and possibly other ``exogenous" prices like tips for full service, fees for credit card payment, or additional local taxes. Observe also that prices observed by users are {\em delayed} and do not typically represent the exact current state of the market given inherent delays in gas production, refinement, transportation, etc.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering 
   \includegraphics[width=4in]{gas-control-loop} 
   \caption{The gas control loop}
   \label{fig:gas-control-loop}
\end{figure}

This kind of block diagram is typical of many closed-loop (feedback) control systems where the system is said to reach {\em equilibrium} if the demand (for gas by drivers) matches the supply (of gas in the market). In data networks, users drive the demand on the network and have different utilities (expectations) when downloading music, playing games, making skype (voice/video) calls, or denying others service by launching a denial-of-service (DoS) attack! In turn, the network observes that user demand and sets ``prices", where the price could be real money, or it could be some measure (indication) of congestion (e.g., delay, loss), or it could represent additional resources that need to be allocated to avoid congestion.

An important question is: {\em What is the goal of network design?} Is it to make users happy? You hope so! Then, mathematically, we say the goal of the network is to maximize the sum of utilities for all its users \cite{elastic:2006}.\footnote{We note that maximizing the sum of utilities involves interpersonal utility comparison, which needs to be ameliorated by linking to a common good, such as money.} Figure \ref{fig:net-control-loop} illustrates the data network equivalent of the gas control loop shown in Figure \ref{fig:gas-control-loop}. We next consider the modeling of user utility and network behavior (resource allocation), before introducing the optimization framework to study the (optimal) steady state for the users and network.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{net-control-loop.jpg} 
   \caption{The network control loop}
   \label{fig:net-control-loop}
\end{figure}

\subsection{Modeling the User}

Users typically have different utilities, i.e. different applications may perform differently based on the level of service (e.g., loss, delay) they get from the network. But, generally speaking, an application should perform better, the higher the rate (throughput) it is able to send at over the network. It is also generally the case that the gain (level of ``happiness") from higher throughput (i.e., {\em marginal utility}) diminishes as the throughput increases.

Figure \ref{fig:concave-utility} shows such a utility function that is typical of what is called 
{\em elastic traffic} \cite{elastic:2006}. 
Formally, user $r$ has utility $U_r(x_r)$ when allocated rate $x_r > 0$. 
$U_r(x_r)$ is an increasing, strictly concave function of $x_r$ (see Figure \ref{fig:concave-function}).
And the derivative $\dot{U}_r(x_r) \rightarrow \infty$ as $x_r \rightarrow 0$,
and  $\dot{U}_r(x_r) \rightarrow 0$ as $x_r \rightarrow \infty$.
Throughout this chapter we assume strictly concave utilities.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2in]{concave-utility.jpg} 
   \caption{Concave utility function}
   \label{fig:concave-utility}
\end{figure}

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2in]{concave-function.jpg} 
   \caption{Concave function. 
   A function $f(.)$ is said to be concave if $f(\alpha x_1 + (1-\alpha)x_2) \geq \alpha f(x_1) + (1-\alpha) f(x_2)$,
   i.e., for any two points $x_1$ and $x_2$, the straight line  that connects $f(x_1)$ and $f(x_2)$ is always below
   or equal to the function $f(.)$ itself. 
   Note that a differentiable concave function has a maximum value at some point $x_{max}$, and 
   that the derivative $\dot{f}(x_{max}) = 0.$
   A strictly concave function would have a strict inequality, whereas a convex function has a cup-like shape and
    has a minimum instead.}
   \label{fig:concave-function}
\end{figure}

\subsection{Modeling the Network}

We consider a network of $J$ resources, e.g., transmission links as they are typically considered the bottleneck. 
We denote by $R$ the set of all possible routes, and we assume that each user (source-destination traffic flow) is assigned to exactly one route $r$ (i.e., static single-path routing)\footnote{Given each user has one flow over a single path, we use the terms ``user", ``flow", and ``route" interchangeably.}.
We then define a 0-1 routing matrix $A$ such that:
$$
a_{jr} = \left\{ \begin{array}{rl} 
              1 & \mbox{ if resource $j$ is on route $r$} \\
              0 & \mbox{ otherwise}
              \end{array} \right.
$$
Figure \ref{fig:net-model} shows an example with three users  over a network of seven links: 
the first user (``blue" flow) uses the route made of links 1, 4, and 6;
the second user (``red" flow) uses the route made of links 2, 5, and 7;
and the third user (``green" flow) uses the route made of links 1, 4, and 7.
So the routing matrix has seven rows and three columns.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=5in]{net-model.jpg} 
   \caption{A network model}
   \label{fig:net-model}
\end{figure}

\subsection{The Optimization Problem}

Now we are ready to formulate a (centralized) optimization problem that allows the network to allocate rates to users so that the sum of their utilities is maximized \cite{Kelly-2001}. We refer to this problem as $SYSTEM(U, A, C)$ where the inputs are the user utility functions $U_r(.)$, the routing matrix $A$, and the (column) vector of link capacities $C$, and the output is the (column) vector of allocated rates $x$. \\

$SYSTEM(U, A, C):$
\begin{eqnarray*}
{\rm max} \sum_{r \in R} U_r(x_r) \ \\
{\rm subject~to}\ A x \leq C \ \\
{\rm over}\  x \geq 0
\end{eqnarray*}

For such an optimization problem, it is known that there exists a unique solution. This is the case because the function to optimize is strictly concave and the link capacity inequality constraints $A x \leq C$ form a so-called {\em convex set} (see Figure \ref{fig:convex-set}.)

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=1.5in]{convex-set.jpg} 
   \caption{A convex set. A convex set intuitively means that any linear combination of any two points located on the boundary of the region, which is formed by the linear inequalities, lies within the region itself. }
   \label{fig:convex-set}
\end{figure}

The practical challenge in solving this problem however is that the network does not know the utilities of its users, let alone its centralized nature makes it computationally expensive to solve!

To address these challenges, we start by decomposing the problem into $R$ problems, one for each user $r \in R$, and one problem for the network (we will later decompose this network problem further into individual resource problems). The network will present each user with a ``price" $\lambda_r$ (\$/bit).
Through these prices, the network attempts to infer user utilities.
Specifically, observing $\lambda_r$,
user $r$ will then choose an amount to pay $w_r$ (\$/second) for the service (that maximizes the user's utility), 
which in turn determines how much rate $x_r$ (bits/second) the user would get ($x_r = w_r/\lambda_r$).  
The network sets its prices $\lambda_r$ based on the load $x_r$, $\forall r$.

\subsection{Introducing Prices}

The decomposed optimization problem can then be stated in terms of the following user optimization problem, and network optimization problem. \\

$USER_r(U_r, \lambda_r):$
\begin{eqnarray*}
{\rm max}\ U_r(\frac{w_r}{\lambda_r}) \ \\
{\rm over}\ w_r \geq 0 
\end{eqnarray*}
where $\frac{w_r}{\lambda_r} = x_r $. Given the network price $\lambda_r$ and its own {\em private} utility function $U_r$, user $r$ determines how much it is willing to pay $w_r$ so as to maximize her own utility.

Knowing the vector $W=\{ w_r, \forall r \}$, its routing and capacity matrices, the network allocates user rates $x_r$ by optimizing some network function $f(x,W)$. Once $x_r$'s are obtained, 
prices are obtained as $\lambda_r = \frac{w_r}{x_r}$.\\

$NETWORK(A, C, W):$
\begin{eqnarray*}
{\rm max}\ \sum_{r \in R} f(x_r, w_r) \ \\
{\rm subject~to}\ A x \leq C \ \\
{\rm over}\ x \geq 0
\end{eqnarray*}

\subsection{Network Optimization}

The choice of the network function $f(x,W)$ determines how the capacity of the network gets allocated to users, and so how {\em fair} we might consider this allocation to be! For example, consider the following function:
\begin{eqnarray*}
f = \sum_{r \in R} w_r\; x_r
%\label{eq:greedy-net}
\end{eqnarray*}

Maximizing this function results in maximizing the total weighted throughput for all users. As a special case, for unit weights, the network optimization problem maximizes the total throughput through the network. This might seem to fly in the face of what we think is fair! Consider the following simple example (see Figure~\ref{fig:greedy-alloc}):
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{greedy-alloc.jpg} 
   \caption{Greedy network allocation}
   \label{fig:greedy-alloc}
\end{figure}
given both links have capacities of 6 units, the total throughput allocated to all users is the total network capacity of 12 units. This can be achieved by allocating 6 units of capacity to each of the 1-link flows (users): the ``red" user and the ``blue" user, leaving the 2-link (``green") flow with no capacity allocated to its user. That does not seem ``fair"! A different function $f$ would allocate rates to users differently and so it would provide a different notion of fairness.

But, the big question is: {\em how do (should) we define fairness?} The research literature introduces many notions of fairness, most notably the so-called {\em max-min fairness}.

\subsubsection{Max-min Fairness}

Intuitively, max-min fairness means that  resources (link capacities) are allocated to users (flows) so that
the allocation is: 
\begin{enumerate}
\item {\em fair:} all users get equal share of a link, as long as users have demand that would fully consume their share, and
\item {\em efficient:} each link is utilized to the maximum load possible.
\end{enumerate}
In other words, if a user cannot consume its equal share of a link, then the excess capacity must be (recursively) allocated equally among high-demanding users. So, the final outcome is that low-demanding users get exactly what they need, while high-demanding users get equal allocations. Consider the following multi-link network example 
(see Figure~\ref{fig:max-min-alloc}):
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{max-min-fair.jpg} 
   \caption{Max-min fair capacity allocation}
   \label{fig:max-min-alloc}
\end{figure}
all links have capacities of 150 units and we assume elastic traffic sources, i.e., sources that would consume all what they can get. We start with the first (left-most) link since it is used by most users so it is the most loaded one. Each flow using that link gets allocated an equal share of $150/3=50$ units. Proceeding to the next loaded link, the middle one, each of its two flows should get an equal share of 75, however flow $F_3$ is limited by its first link to 50 units of throughput. Thus, flow $F_4$ gets the left-over from $F_3$ to a total allocation of $75+25=100$. The right-most link, at capacity of 150, does not limit the throughput  of $F_4$, which ends up using only 100 units of that link, leaving 50 unused. At the end of this process, we say that the max-min fair allocation  vector is $x^{\rm T} =(50,\; 50,\; 50,\; 100)$.

Mathematically, max-min fairness is achieved when the network maximizes the following function:
\begin{eqnarray*}
f = {\rm min}_{r \in R} \ x_r
%\label{eq:prop-fair}
\end{eqnarray*}
Intuitively, maximizing the minimum of allocated rates results in equalizing these rates, as long as users have enough demand that will consume these rates over the network.

\subsubsection{Proportional Fairness}

Another equally popular fairness definition is the so-called {\em (weighted) proportional fairness}. This notion of fairness is achieved when the network maximizes the following function:
\begin{eqnarray*}
f = \sum_{r \in R} w_r \; \lg(x_r)
\label{eq:prop-fair}
\end{eqnarray*}
Note that the log function is a concave, and strictly increasing function. Thus, given optimal rate allocation solution $x^*$, that is feasible, i.e.,\ $x^* \geq 0$ and $A\ x^* \leq C$, any other feasible solution $x$ will cause the aggregate proportional change $\sum_{r \in R} w_r \frac{x_r - x^*_r}{x^*_r}$ to be less than or equal zero.  To show this, for simplicity, assume one user and unit weight, so $f(x) = \lg(x)$. Expanding $f(x)$ into its first-order (linear) Taylor's approximation around $x^*$, we obtain:
\begin{eqnarray*}
f(x) \approx f(x^*) + (x - x^*) \dot{f}(x^*)
\end{eqnarray*}
Given the derivative $\dot{f}(x^*) = \frac{1}{x^*}$, we have:
\begin{eqnarray*}
f(x) \approx f(x^*) + \frac{(x - x^*)}{x^*} 
\end{eqnarray*}
Since $f$ is maximized at $x^*$, $f(x^*) \ge f(x)$ and so the proportional fairness condition must hold: 
\begin{eqnarray*}
\frac{x - x^*}{x^*} \leq 0 
\end{eqnarray*}
Note that the presence of weight $w_r$ intuitively means that user (flow) $r$ is equivalent to $w_r$ users with unit weight each.

\subsubsection{General Parameterized Utility}

If the network function $f(x)$ is a function of the utilities of its users $U(x)$, then the network is in fact maximizing a function of user utilities. Assuming each user $r$ has unit weight $w_r$, $U_r(x_r)$ can be generalized as \cite{Mo:2000}:
\begin{eqnarray*}
U_r(x_r) = \frac{x_r^{1-\alpha}}{1-\alpha}
\end{eqnarray*}
where $\alpha$ is a parameter that determines the fairness criterion of the network. More specifically, if $\alpha \rightarrow 0$, then a user's utility is linear in its allocated rate and the network is effectively maximizing the sum of user utilities $\sum_{r \in R} U_r(x_r) = \sum_{r \in R} x_r$, which in turn yields a greedy allocation that maximizes the total throughput over the network.

On the other hand, if $\alpha \rightarrow 1$, then this is equivalent to a log utility, yielding proportional fair allocation. To see this, let's take the derivative of $U_r(x_r)$:
\begin{eqnarray*}
\dot{U}_r(x_r) = \frac{(1-\alpha) x_r^{-\alpha}}{1-\alpha} \rightarrow \frac{1}{x_r}  \ {\rm as}\  \alpha \rightarrow 1
\end{eqnarray*}
By integrating $\dot{U}_r(x_r)$, we get back $U_r(x_r) = \lg(x_r)$.

Similarly, it can be shown that $\alpha \rightarrow \infty$ corresponds to a minimum utility, yielding a max-min fair allocation.

\subsection{Solution to Optimization Problem}
\label{sec:sol-opt}

Consider the case where the network is maximizing the weighted sum of the log of user rates, i.e., the network is trying to solve the following optimization problem that would yield a weighted proportional fairness allocation: \\

$NETWORK(A, C, W):$
\begin{eqnarray*}
{\rm max}\ \sum_{r \in R} w_r \; \lg(x_r) \\
{\rm subject~to}\ A x \leq C \ \\
{\rm over}\ x \geq 0
\end{eqnarray*}

We can solve this problem using the theory of constrained convex optimization using the Lagrangian technique. Specifically, we move the constraints into the objective function that we want to optimize, thus making the optimization problem effectively unconstrained. We do so by introducing so-called ``Lagrangian multipliers" into the new objective (Lagrangian) function $L$:
\begin{eqnarray*}
{\rm max}\ L = \sum_{r \in R} w_r \; \lg(x_r) + \lambda^T(C - Ax)\\
\end{eqnarray*}
The (row) vector $\lambda^T$ is a Lagrangian vector with a variable $\lambda_j$ for each link $j$ in the network.
Note that $L$ is a strictly concave function, thus a solution exists at which the derivatives of $L$ with respect to each $x_r$ and each $\lambda_j$ are equal to zero:
\begin{eqnarray}
  \frac{\partial L}{\partial x_r} & =   & \frac{w_r}{x_r} - \sum_{j \in r} \lambda_j
  \label{eqn:kkt1}
\end{eqnarray}
\begin{eqnarray}
  \frac{\partial L}{\partial \lambda_j} & =  & (C_j - \sum_{r \in j} x_r)  
  \label{eqn:kkt2}
\end{eqnarray}
The notation $j \in r$ indicates all links $j$ used by user (flow/route) $r$,
whereas $r \in j$ denotes all flows $r$ using link $j$, i.e. the total load on link $j$.

By equating the first set of equations (\ref{eqn:kkt1}) to zero, we obtain the (weighted proportionally fair) solution:
\begin{eqnarray}
x_r = \frac{w_r}{\sum_{j \in r} \lambda_j} 
\label{eqn:opt-x}
\end{eqnarray}
We obtain $\lambda_j$ by also equating the second set of equations (\ref{eqn:kkt2}) to zero. 
Note that $\lambda_j$ and $(C_j - \sum_{r \in j} x_r)$ must be greater than or equal to zero since negative values do not maximize the objective function $L$!  Furthermore, 
$(C_j - \sum_{r \in j} x_r) \geq 0$ ensures that the link capacity constraints $\sum_{r \in j} x_r \leq C_j$ are automatically satisfied. 
If $(C_j - \sum_{r \in j} x_r) = 0$ then $\lambda_j$ can be greater than zero. On the other hand,
if  $\lambda_j = 0$, then the associated link may not be fully utilized, i.e.\ $\sum_{r \in j} x_r < C_j$. 
Intuitively, $\lambda_j$ represents the ``cost" associated with link $j$, so it is zero if the link is under-utilized,
and positive if the link is allocated to capacity.

\paragraph{Example:}
Consider the example in Figure~\ref{fig:greedy-alloc} but now assume the network's objective of proportionally allocating its capacity, i.e., 
\begin{eqnarray*}
{\rm max}\  f = \lg(x_0) + \lg(x_1) + \lg(x_2) 
\end{eqnarray*}
subject to:
\begin{eqnarray*}
x_0 + x_1 \leq 6 \\
x_0 + x_2 \leq 6 \\
x_0, x_1, x_2 \geq 0
\end{eqnarray*}
where $x_0$, $x_1$, and $x_2$ are the rates allocated to the two-link flow (user), the first-link flow, and the second-link flow, respectively. \footnote{Note that since the objective (log) function is strictly increasing, then the $x_i$'s should be as large as possible to consume the total capacity of the links, so the two inequalities on link capacities could be turned into equalities. }

Using the Lagrangian's solution method, we obtain:
\begin{eqnarray*}
{\rm max}\ L = \lg(x_0) + \lg(x_1) + \lg(x_2) + \lambda_1(6- (x_0+x_1)) + \lambda_2(6- (x_0+x_2))
\end{eqnarray*}
Taking derivatives, we obtain:
\[
\begin{array}{ccc}
  \frac{\partial L}{\partial x_0} & =   & \frac{1}{x_0} - (\lambda_1+ \lambda_2)   \\ \\
  \frac{\partial L}{\partial x_1} & =   & \frac{1}{x_1} - \lambda_1   \\ \\
  \frac{\partial L}{\partial x_2} & =   & \frac{1}{x_2} - \lambda_2   \\ \\
  \frac{\partial L}{\partial \lambda_1} & =  & 6 -   (x_0+x_1)\\ \\
  \frac{\partial L}{\partial \lambda_2} & =  & 6 -   (x_0+x_2) \\
\end{array}
\]
Equating these derivatives to zero, the last two equations show full utilization of the link capacities and that $x_1=x_2$, while the first three equations give the following values of $x_i$'s: 
\begin{eqnarray*}
x_1 = x_2 = \frac{1}{\lambda_1} = \frac{1}{\lambda_2} = \frac{1}{\lambda} \\ \\
x_0 = \frac{1}{2 \lambda}
\end{eqnarray*}
Substituting in the capacity equations, we obtain the price of each link $\lambda$:
\begin{eqnarray*}
\frac{1}{2\lambda} + \frac{1}{\lambda} = 6 \\ 
\end{eqnarray*}
Thus, $\lambda = \frac{1}{4}$, and so $x_0 = 2$, and $x_1 = x_2 = 4$. 
Note that in this optimal case, each link is fully utilized to capacity, and the flow that traverses two links is charged twice for each link it traverses and so it gets allocated a lower rate.\footnote{As we will later see,
this proportional rate allocation is what TCP Vegas \cite{vegas:1995} provides.}
{\bf End Example.}\\

If the utility of each user $r$ is a log function in its allocated rate $x_r$, then the (weighted proportionally fair) network solution 
$x_r = \frac{w_r}{\sum_{j \in r} \lambda_j}$ is in fact, a solution to the whole system optimization problem that includes the network, as well as all users possibly trying to independently (in a distributed way) maximize their own log utilities.
However, in a distributed setting,
as noted earlier, even if the network knows the user utility functions,
the network allocates user rates based on their willingness to pay, $w_r$, which might be unknown to the network.
This lack of knowledge can be overcome by observing the demand behavior of the user $x_r$ and 
the price $\lambda_r = \sum_{j \in r} \lambda_j$, and so $w_r$ is computed as $w_r = x_r \; \lambda_r$.
Otherwise, the network can just assign some weights $w_r$ to users based on some preference policy.

The moral of the story is that in practice,
there is no central network controller that knows $W$ and can then allocate rates to users.
Each user and each resource (link) might have its own individual controller that will operate
independently and so we need to study the collective behavior of such composite system
and answer questions such as:
Would the system converge (stabilize) to a solution in the long term (i.e., reaching steady state)?
If so, 
is this solution unique and how far is it from the target (optimal) operating point?
In general, if the system gets perturbed, is it stable, i.e.
does it converge back to steady state,
and how long does it take to converge and how smooth or rough was that?
In control-theoretic terminology,
we refer to the response to such perturbation until steady state is reached
as the {\em transient response} of the system.
We refer to how far the system is from being unstable,
or the magnitude of perturbation that renders the system unstable, as
{\em stability margin}.

To formally address these questions, we will resort to the modeling of user and network dynamic behaviors,
in the form of differential (or difference) equations, then use well-known control-theoretic techniques
to study the overall transient and steady-state behavior of the system.

\section{The Control Problem}
\label{sec:control}

The basic control problem is to control the output of a system given a certain input.
For example, we want to control the user demand (sending rate) given the observed network price (e.g., packet loss or delay).
Similarly, we want to control the price advertised by a network resource given the demand (rates) of its users.

There is basically two kinds of control: open-loop control, and closed-loop (feedback) control.
In open-loop control systems, there is no feedback about the state of the system and the output of the system is controlled directly by the input signal. This type of control is thus simple, but not as common as closed-loop control.
An example of open-loop control system is a microwave that heats food for the input (specified) duration.

Feedback (closed-loop) control is more interesting and multiple controllers may be present in the same control loop. See Figure~\ref{fig:net-control-loop} where a user controller is present to control demand based on price, and a resource controller is also present to control price based on demand. Feedback control makes it possible to control the system well even if we can't observe or know everything, or if we make errors in our estimation (modeling) of the current state of the system, or if things change. This is because we can continually measure and correct (adapt) to what we observe (i.e., feedback signal). For example, in a congestion control system, we do not need to {\em exactly} know the number of users, the arrival rate of connections, or the service rate of the bottleneck resource, since each user would adapt its demand based on its own observed (measured, fed back) price, which reflects the current overall congestion of the bottleneck resource.

Associated with feedback control is a delay to observe the feedback (measured) signal, which is referred to as {\em feedback delay}. More precisely, feedback delay refers to the time taken from the generation of a control signal (e.g., updated user demand) until the process/system reacts to it (e.g., demand is routed over the network), this reaction takes effect at each resource (e.g., load is observed on each link), and this effect is fed back to the controller (e.g., price is observed by the user).

\subsection{System Models}

Models of controlled systems can be classified along four dimensions:
\begin{itemize}
\item {\em Deterministic versus stochastic models.} The latter models capture stochastic effects like noise and uncertainties.

\item {\em Time-invariant versus time-varying models.} The latter models contain system parameters that change over time.

\item {\em Continuous-time versus discrete-time models.} In the latter models, time is divided into discrete-time steps.

\item {\em Linear versus non-linear models.} The latter models contain non-linear dynamics.

\end{itemize}
In most of our treatment, we consider the simplest kind of models that are deterministic, time-invariant, continuous-time, and linear. In modeling a controlled system, we characterize the relationships among system variables as a function of time, i.e., dynamic equations. See Figure~\ref{fig:system-model} where functions $f$ and $h$ are generally non-linear functions. 
The function $f$ models the evolution of the state of the system $x$ as 
a function of the current state and system's input $u$.
The function $h$ yields system's output $y$ as a function of the current state and input values.
As we will see later, for mathematical tractability, we often linearize dynamic non-linear models or 
we only consider operation in a linear regime, for example, we ignore non-linearity when 
the buffer goes empty or hits full capacity.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=3in]{system-model.jpg} % requires the graphicx package
   \caption{Typical system model}
   \label{fig:system-model}
\end{figure}

\subsection{Modeling Source and Network Dynamics}

Consider a source $r$ with log utility, i.e. $U_r(t) = w_r \; \lg(x_r)$, and 
a network that allocates rates in a weighted proportional fashion.
We saw earlier (cf.\ Section~\ref{sec:sol-opt}) that in steady state, the (optimal) solution (Equation \ref{eqn:opt-x}) is:
\begin{eqnarray}
x_r = \frac{w_r}{\lambda_r}
\label{eqn:kkt}
\end{eqnarray}
This can be re-written as $w_r - x_r \lambda_r = 0$.
Also, we saw that the optimal solution ensures that each link $l$ is fully utilized,
i.e. the load (total input rate) on link $l$, denoted by $y_l = \sum_{s: l \in s} x_s$, equals the link capacity $c_l$.

The dynamics of the sources and links can then be modeled such that these steady-state user rates and link loads are achieved. Specifically, we can write the dynamic (time-dependent) source algorithm as:
\begin{eqnarray}
\dot{x}_r (t) = k [w_r - x_r(t) \lambda_r(t)]
\label{eqn:source-adapt}
\end{eqnarray}
where $k$ is a proportionality factor.
Note that $w_r$ represents how much user $r$ is willing to pay,
whereas $x_r(t) \lambda_r(t)$ represents the cost (price) of sending at that rate.
Intuitively, the user sending rate increases (decreases) when the difference between these two quantities is positive (negative). And in steady state, $\dot{x}_r(\infty) \rightarrow 0$, and so we obtain the steady-state solution  $x_r = \frac{w_r}{\lambda_r}$ (as expected).

Given that the derivative of  $U_r(t)$, $\dot{U}_r(t) = \frac{w_r}{x_r}$, 
the source rate adaptation algorithm can be re-written as:
\begin{eqnarray*}
\dot{x}_r (t) = k x_r(t) [\dot{U}_r(t) - \lambda_r(t)]
\end{eqnarray*}
%
\begin{eqnarray}
\dot{x}_r (t) = K(t) [\dot{U}_r(t) - \lambda_r(t)]
\label{eqn:source-alg}
\end{eqnarray}
Intuitively, the user increases its sending rate if the marginal utility (satisfaction) is higher than the price that the user will pay, otherwise the user decreases its sending rate.

We can also write a dynamic equation for the adaptation in the link price $\lambda^l(t)$, called the link pricing algorithm:
\begin{eqnarray}
\dot{\lambda}^l(t) = h (y_l(t) - c_l)
\label{eqn:link-alg}
\end{eqnarray}
where $h$ is a proportionality factor, and the total price, $\lambda_r(t)$, for user $r$,
is the sum of the link prices along the user's route, i.e. $\lambda_r(t) =  \sum_{l: l \in r} \lambda^l(t)$.
Intuitively, the link price increases if the link is over-utilized (i.e. $y_l(t) > c_l$), otherwise
the link price decreases.
Note that at steady state, $\dot{\lambda}^l(\infty) \rightarrow 0$, and 
we obtain the steady-state optimal solution $y_l = c_l$ (as expected).

It turns out that the source and link algorithms, Equations \ref{eqn:source-alg} and \ref{eqn:link-alg},
represent general user and resource adaptation algorithms that collectively determine 
the transient and steady behavior of the whole system.
In what follows, we use the form of Equation \ref{eqn:source-alg} to reserve engineer different versions
of TCP and deduce the utility function that the TCP source tries to maximize.

\subsection{TCP and RED}

Many analytical studies considered the network system composed of TCP sources over a network of queues that employ a certain queue management policy. Examples of TCP variants include Reno, SACK \cite{Fall:1996}, NewReno, Vegas \cite{vegas:1995}, FAST \cite{fast:2006}, etc. Examples of queue management policies include Drop Tail, RED \cite{red:1993}, REM \cite{REM:2001}, PI \cite{PI:2000}, etc. One of the most widely studied instantiation is that of TCP sources over a RED bottleneck queue --- see Figure~\ref{fig:TCP-RED-diagram}.
We start by modeling the dynamic behavior of a TCP source, i.e., 
the time-dependent relationship between its transmission window (or sending rate) and 
its observed loss rate or delay (price).
We do so for both TCP Reno and Vegas versions, and also deduce their utilities.
Then, we model the buffering process inside the network (more precisely, the bottleneck queue),
assuming a linear regime (i.e., ignoring non-linearities due to the buffer becoming empty or full).
Using the average buffer length, we model the dynamic behavior of RED and how 
it generates packet losses (or markings) as indication of congestion (price).
This overall model represents the closed-loop feedback system shown in 
the block diagram of Figure~\ref{fig:TCP-RED-diagram}, 
which can then be analyzed using control-theoretic techniques.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=4in]{tcp-red.jpg} % requires the graphicx package
   \caption{TCP Reno over RED feedback control system}
   \label{fig:TCP-RED-diagram}
\end{figure}

\paragraph{Modeling TCP Reno:} First, consider the modeling of TCP Reno, where the congestion window {\em cwnd} is increased by $1/cwnd$ for every acknowledged TCP segment / non-loss, i.e., it is (roughly) increased by 1 every round-trip time, and {\em cwnd} is decreased by half for every loss. Thus, we can write the following equation for changes in the congestion window of a single TCP flow, where $p$ is the segment loss probability:

\begin{eqnarray*}
\Delta cwnd = \frac{1}{cwnd} (1-p) - \frac{cwnd}{2} p
\end{eqnarray*}
Let $x$ denote the sending rate, and $T$ the round-trip time, thus $x = \frac{cwnd}{T}$.
Assuming acknowledgments (ACKs) come equally spaced, 
the time between ACKs (or lack thereof) is given by $\frac{T}{cwnd}$.
Thus, we can re-write the above equation in terms of change in {\em rate} as:
\begin{eqnarray*}
\frac{d}{dt}  cwnd(t) = \frac{\frac{1}{cwnd(t)} (1-p(t)) - \frac{cwnd(t)}{2} p(t)}{\frac{T}{cwnd(t)}}
\end{eqnarray*}
Dividing both sides by $T$, we get:
\begin{eqnarray*}
\frac{d}{dt} x(t) = \frac{\frac{1}{x(t) T^2} (1-p(t)) - \frac{x(t)}{2} p(t)}{\frac{1}{x(t)}}
\end{eqnarray*}

\begin{eqnarray}
\frac{d}{dt} x(t) = \frac{1}{T^2} (1-p(t)) - \frac{{x(t)}^2}{2} p(t)
\label{eqn:reno}
\end{eqnarray}

%\begin{eqnarray}
%\frac{d}{dt} x(t) = \frac{1}{T^2}  - (\frac{1}{T^2} + \frac{{x(t)}^2}{2}) p(t)
%\label{eqn:reno}
%\end{eqnarray}

Let's denote the loss probability $p(t)$ of TCP connection $r$ as $p_r(t)$.
$p_r(t)$ depends on the current load on path $r$,
and can be approximated by the sum of loss probabilities experienced on individual links $j \in r$ 
along the connection's path.
More specifically,
\begin{eqnarray*}
p_r(t) = \sum_{j \in r} p_j(\sum_{s: j \in s} x_s(t))
\end{eqnarray*}

Assuming small $p$ such that $(1-p) \approx 1$, we can re-write Equation~\ref{eqn:reno} as follows:
\begin{eqnarray*}
\frac{d}{dt} x(t) = \frac{1}{T^2}  - \frac{{x(t)}^2}{2} p(t)
\end{eqnarray*}

\begin{eqnarray}
\frac{d}{dt} x(t) = \frac{{x(t)}^2}{2} [\frac{2}{T^2 x(t)^2}  - p(t)]
\label{eqn:reno-util}
\end{eqnarray}

Comparing Equation \ref{eqn:reno-util} with Equation \ref{eqn:source-alg},
we can deduce the utility function of a TCP Reno source:
\begin{eqnarray*}
\dot{U}(x) = \frac{2}{T^2 x^2} 
\end{eqnarray*}
Integrating $\dot{U}(x)$ we get:
\begin{eqnarray*}
U(x) =  \frac{-2}{T^2 x} 
\end{eqnarray*}
Observe that maximizing Reno's utility results in minimizing the quantity $\frac{1}{x}$, which can be viewed as the ``potential delay" as it is inversely proportional to the allocated rate $x$. 
Thus, a network allocation based on such utility is referred to as {\em minimum potential delay fair allocation}.

\paragraph*{Example:} 
Revisting the example in Figure~\ref{fig:greedy-alloc} but now assume the network's objective is to allocate its capacity according to the minimum potential delay fair allocation, i.e., 
\begin{eqnarray*}
{\rm max} \ f = \frac{-1}{x_0} + \frac{-1}{x_1} + \frac{-1}{x_2} 
\end{eqnarray*}
subject to:
\begin{eqnarray*}
x_0 + x_1 \leq 6 \\
x_0 + x_2 \leq 6 \\
x_0, x_1, x_2 \geq 0
\end{eqnarray*}
where $x_0$, $x_1$, and $x_2$ are the rates allocated to the two-link flow (user), the first-link flow, and the second-link flow, respectively.

Using the Lagrangian's solution method, we obtain:
\begin{eqnarray*}
{\rm max}\ L =  \frac{-1}{x_0} + \frac{-1}{x_1} + \frac{-1}{x_2} + \lambda_1(6- (x_0+x_1)) + \lambda_2(6- (x_0+x_2))
\end{eqnarray*}
Taking derivatives, we obtain:
\[
\begin{array}{ccc}
  \frac{\partial L}{\partial x_0} & =   & \frac{1}{x_0^2} - (\lambda_1+ \lambda_2)   \\ \\
  \frac{\partial L}{\partial x_1} & =   & \frac{1}{x_1^2} - \lambda_1   \\ \\
  \frac{\partial L}{\partial x_2} & =   & \frac{1}{x_2^2} - \lambda_2   \\ \\
  \frac{\partial L}{\partial \lambda_1} & =  & 6 -   (x_0+x_1)\\ \\
  \frac{\partial L}{\partial \lambda_2} & =  & 6 -   (x_0+x_2) \\
\end{array}
\]
Equating these derivatives to zero, the last two equations show full utilization of the link capacities and that $x_1=x_2$, while the first three equations give the following values of $x_i$'s: 
\begin{eqnarray*}
x_1 = x_2 = \frac{1}{\sqrt{\lambda_1}} = \frac{1}{\sqrt{\lambda_2}} = \frac{1}{\sqrt{\lambda}} \\ \\
x_0 = \frac{1}{\sqrt{2 \lambda}}
\end{eqnarray*}
Substituting in the capacity equations, we obtain the price of each link $\lambda = 0.08$,
and so $x_0 \approx 2.5$, and $x_1 = x_2 \approx 3.5 $.

Note that in this optimal case, each link is fully utilized to capacity, and 
the rate allocated to a flow is inversely proportional to the square-root of the price it observes along its path.

Note also that this captures the well-known {\em steady-state} relationship between the throughput of a TCP Reno source and the inverse of the square-root of the {\em loss probability} observed by the TCP source \cite{tcp-thru:2000}.
A TCP Reno source adapting based on Equation \ref{eqn:reno} would converge to such steady-state throughput value.
{\bf End Example.}\\

\paragraph{Modeling TCP Vegas:} 
Now, let us consider the modeling of another version of TCP --- TCP Vegas \cite{vegas:1995}.
This version, unlike Reno, tries to {\em avoid} congestion, rather than induce loss and 
and then adapt the transmission (congestion) window to it.  
The basic idea behind Vegas is to calculate the {\em actual} throughput of the connection
as $\frac{w(t)}{T(t)}$, where $w(t)$ is the current window size,
and $T(t)$ is the measured round-trip time (RTT) over the connection's path.
This RTT includes queueing delay, as well as propagation delay $D$.
Ideally, with no congestion, the ideal throughput can be computed by the source
as $\frac{w(t)}{D}$, where $D$ is estimated using the minimum RTT recently observed by the source.
To ensure high utilization of the network, we want some queueing, i.e. 
the actual throughput is lower than the ideal one, but not too low to start causing congestion (i.e. buffer overflow
at the bottleneck link resulting in losses).
Vegas then adapts $w(t)$ based on some target difference, $\alpha$, 
between the actual throughput and the ideal one.
More specifically,
the window increases if $(\frac{w(t)}{D} - \frac{w(t)}{T(t)}) < \alpha$,
decreases if $(\frac{w(t)}{D} - \frac{w(t)}{T(t)}) > \alpha$,
and stays the same otherwise. This dynamic source behavior,
i.e. change in window over time, can be modeled as:
\begin{eqnarray*}
\frac{dw(t)}{dt} = k [\alpha - (\frac{w(t)}{D} - \frac{w(t)}{T(t)})]
\end{eqnarray*}
This can be re-written as:
\begin{eqnarray*}
\frac{dw(t)}{dt} = \frac{k}{D} [\alpha D - (w(t) - \frac{w(t)}{T(t)}D)]
\end{eqnarray*}
Denoting the sending rate (throughput) by $x(t) = \frac{w(t)}{T(t)}$, 
and $\gamma = \frac{k}{D}$, we have:
\begin{eqnarray*}
\frac{dw(t)}{dt} = \gamma [\alpha D - (w(t) - x(t)D)]
\end{eqnarray*}
At steady state, as $\dot{w}(\infty) \rightarrow 0$, we have:
\begin{eqnarray*}
w - x D = \alpha D
\end{eqnarray*}
Observe that the left-hand side represents the difference between the window size of packets
injected by the source, and the number of packets in flight / propagating along the path (represented by
the product of throughput and propagation delay). Thus,
the left-hand side represents the number of packets in the bottleneck queue,
and $\alpha D$ denotes the {\em target} queue occupancy of the bottleneck link.
Intuitively, Vegas tries to maintain a small number of $\alpha D$ packets
(i.e., 1-2 packets)
in the bottleneck queue to maintain both small delay and high (100\%) utilization.
Section~\ref{sec:linear} uses control theory to analyze a Vegas-like transmission model.

Given that $x = w/T$, we get:
\begin{eqnarray*}
x T - x D = \alpha D
\end{eqnarray*}
Denoting the queueing delay by $Q$, we have $T = Q + D$, and so: 
\begin{eqnarray*}
x Q = \alpha D
\end{eqnarray*}
\begin{eqnarray*}
x = \frac{\alpha D}{Q}
\end{eqnarray*}
Comparing with Equation \ref{eqn:kkt}, we can deduce that
the willingness to pay $w_r$ for a Vegas user $r$ is $\alpha D$
and that the price $\lambda_r$ experienced by the user is the queueing delay $Q$.

Now, to deduce the utility function that a Vegas user tries to maximize,
let us write its rate adaptation equation following Equation \ref{eqn:source-adapt}:
\begin{eqnarray*}
\dot{x}_r(t) = k [\alpha D - x_r(t) Q(t)]
\end{eqnarray*}
\begin{eqnarray*}
\dot{x}_r(t) = K(t) [\frac{\alpha D}{x_r(t)} - Q(t)]
\end{eqnarray*}
Thus, comparing with Equation \ref{eqn:source-alg}, we deduce:
\begin{eqnarray*}
\dot{U}_r(t) = \frac{\alpha D}{x_r(t)} 
\end{eqnarray*}
Integrating, we obtain:
\begin{eqnarray*}
U_r(t) = \alpha D \; \lg(x_r(t)) 
\end{eqnarray*}
Recall that maximizing such user utilities results in a weighted proportional fair allocation.

\paragraph{Modeling RED:}
Let us now consider the modeling of the buffer and associated RED queue management algorithm \cite{red:1993}.
Figure~\ref{fig:red} shows how RED tries to avoid congestion by dropping (or marking) packets with probability $p_c$ 
as a (non-linear) function of the average queue length $v$.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=2in]{red.jpg} % requires the graphicx package
   \caption{RED dropping (or marking) function}
   \label{fig:red}
\end{figure}
First, we model the evolution of the queue length $b(t)$ as 
a function of the total input rate, $y(t) = \sum x_s(t)$, and (bottleneck) link capacity, $C$:
\begin{eqnarray*}
\dot{b}(t) =  y(t) - C
\end{eqnarray*}
Denoting by $v(t)$, the Exponential Weighted Moving Average (EWMA) of the queue length:
\begin{eqnarray*}
v(t+\delta) = (1 - \alpha) v(t) + \alpha b(t)
\end{eqnarray*}
\begin{eqnarray*}
v(t+\delta) - v(t) = \alpha (b(t) - v(t))
\end{eqnarray*}
Given $v(t)$ gets updated at the link rate, i.e. $\delta = \frac{1}{C}$,
and $\dot{v}(t) \approx \frac{v(t+\delta) - v(t)}{\delta}$, we have:
\begin{eqnarray*}
\dot{v}(t) = \alpha C \; (b(t) - v(t))
\end{eqnarray*}
This last equation represents the dynamic model of RED averaging, which in turn determines the price $p_c(t)$ that users experience.

To simplify the model and gain insight, let us ignore the (hard) non-linearities of the RED function
and consider only the linear region:
\begin{eqnarray*}
p_c(t) = \sigma v(t) + \eta = \sigma \int \dot{v}(t) dt + \eta = \sigma \int \alpha C(b(t) - v(t)) dt + \eta
\end{eqnarray*}
where $\sigma = p_m/(B_{max}-B_{min})$, and $\eta= -p_m\;B_{min}/(B_{max}-B_{min})$.

To gain more insight, let us further ignore the RED averaging, 
assuming that the price is set in proportion to the actual queue length,
$B_{min}=0$ and $p_m = 1$, then we have:
\begin{eqnarray*}
p_c(t) = \frac{1}{B_{max}} \; b(t)
\end{eqnarray*}
Differentiating both sides, we obtain:
\begin{eqnarray*}
\dot{p}_c(t) = h \; \dot{b}(t) = h(y(t) - C)
\end{eqnarray*}
where $h = \frac{1}{B_{max}}$.
Comparing with Equation \ref{eqn:link-alg}, 
the packet dropping (congestion marking) probability, $p_c(t)$,
represents the ``price", i.e. Lagrangian multiplier, observed by users of this buffer.
Note that at steady state, $\dot{p}_c(\infty) \rightarrow 0$,
and so $y=C$, i.e. the link is fully utilized at steady state.

\subsection{Solving the Feedback Control System}

We have developed dynamic (time-dependent) models for users (sources), e.g. TCP, and the network (links), e.g. RED,
and the interaction between them through prices.
The next step is to solve for the transient and steady-state performance of such system.
Solving such systems is challenging because of inherent non-linearilities,
e.g. the ``hard" non-linearities (discontinuities) in the RED pricing function,
or the ``soft" non-linearity of TCP where the sending rate changes quadratically in the current rate.
Non-linear control theory becomes a useful tool as it deals directly with non-linear differential equations.
Specifically, a method called {\em Lyapunov} \cite{Ogata:2010} allows one to study convergence (stability) by showing that the value of some positive function of the state  of the system continuously decreases as the system evolves over time.
Finding such a Lyapunov function can be challenging, and 
transient performance can often only be obtained by solving the system equations numerically.

To this end, a technique called {\em linearization} can prove more tractable
where the non-linear system is approximated by a set of {\em linear} equations around a single operating point (state).
See Figure \ref{fig:linearize}.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=3in]{linearize.jpg} % requires the graphicx package
   \caption{Linearization}
   \label{fig:linearize}
\end{figure}
With linearization, we become concerned with {\em local stability} and study perturbations around the operating point
using standard (linear) control theory. 
By local stability, we mean that if the system is perturbed within a small region around the operating point 
then the system will converge and stabilize back to that point.
This is in contrast to {\em global stability} where the original (non-linear) system is shown to 
converge from {\em any} starting state.
To linearize the non-linear system around an operating point,
the basic idea is to expand the non-linear differential equation into 
a Taylor series around that point and then ignore high-order terms.

In what follows, we briefly review some basics of classical control theory for linear systems,
then we introduce non-linear control theory. We also show examples of control theoretic analysis for 
the dynamic models introduced above.
For more detailed background on control theory,
we refer the reader to \cite{Ogata:2010,Keshav:2010,Lu:2001}.

\section{Linear Control Theory}  
\label{sec:linear}

In linear control theory, we transform differential equations in the {\em time domain} to algebraic equations in the so-called {\em frequency} or {\em Laplace} domains. Once this Laplace transformation is done, we use simple algebra to study the performance of the system without the need for going back to the (complicated) time domain. Specifically, we can transform a function $f(t)$ to an algebraic function $F(s)$, referred to as the Laplace transform of $f(t)$, as follows:
\begin{eqnarray*}
F(s) = \int_0^{\infty} f(t) e^{-st} dt
\end{eqnarray*}
where $s$ is a complex variable: $s = \sigma + j \omega$, $\sigma$ is the real part of $s$, denoted by $Re(s)$,
and $\omega$ is the imaginary part of $s$, denoted by $Im(s)$.

\paragraph{Example (Unit step function):}
The Laplace transform of a unit step function $u(t)$, where $u(t) = 1$ if $t > 0$, and $u(t) = 0$ otherwise, is given by:
\begin{eqnarray*}
U(s) = \int_0^{\infty} 1 . e^{-st} dt = \frac{1}{s}
\end{eqnarray*}

\paragraph{Example (Impulse function):}
The Laplace transform of a unit impulse function $\delta(t)$, where $\delta(t) = 1$ if $t = 0$, and $\delta(t) = 0$ otherwise, is given by:
\begin{eqnarray*}
U(s) = \int_0^{\infty} 1 . e^{-st} dt = e^{0} = 1
\end{eqnarray*}

Table~\ref{tab:transforms} lists basic Laplace transforms.
\begin{table}[htdp]
\caption{Basic Laplace transforms}
\begin{center}
\begin{tabular}{|c|l|} \hline
Impulse input:	$f(t) = \delta(t)$ & $F(s) = 1$ \\ [2ex]
Step input: 	$f(t) = a.1(t)$	& $F(s) = a/s$ \\[2ex]
Ramp input: 	$f(t) = a.t$		& $F(s) = a/s^2$ \\[2ex]
Exponential:	$f(t) = e^{at}$	& $F(s) = 1/(s-a)$ \\[2ex]
Sinusoid input:	$f(t) = \sn(at)$	& $F(s) = a/(s^2+a^2)$ \\ [2ex]\hline
\end{tabular}
\end{center}
\label{tab:transforms}
\end{table}%

Table~\ref{tab:composition} lists basic composition rules, 
where $\L[f(t)]$ denotes the Laplace transform of $f(t)$, i.e. $F(s)$.
\begin{table}[htdp]
\caption{Composition rules}
\begin{center}
\begin{tabular}{|c|} \hline 
Linearity:	$\L[a\ f(t) + b\ g(t)] = a F(s) + b G(s)$ \\[2ex]
Differentiation: $\L[df(t)/dt] =  sF(s) - f(0) = sF(s)$ if $f(0) = 0$ \\[2ex]
Integration: $\L[\int f(\tau) d\tau] = F(s)/s$ \\[2ex]
Convolution: $y(t) = g(t) \Conv u(t) = \int_0^{t} g(t-\tau) u(\tau) d \tau \Rightarrow Y(s) = G(s) U(s)$ \\ [2ex] \hline
\end{tabular}
\end{center}
\label{default}
\label{tab:composition} 
\end{table}%

\paragraph{Example:}
Consider the following second-order linear, time-invariant differential equation, where $y(t)$ represents the output of a system, and $u(t)$ represents the input:
\begin{eqnarray*}
a_2 \ddot{y}(t) + a_1 \dot{y}(t) + a_0 y(t) = b_1 \dot{u}(t) + b_0 u(t)
\end{eqnarray*}
In the time domain, if we represent the system by $g(t)$,
then $y(t)$ can be obtained by convolving $u(t)$ with $g(t)$, i.e. $y(t)=g(t) \Conv u(t)$.
This involves a complicated integration over the system responses,
$g(t-\tau)$, to impulse inputs of magnitude $u(\tau)$, for all $0 < \tau < t$.

Assuming $y(0)=u(0)=0$, taking the Laplace transform of both sides, we obtain:
\begin{eqnarray*}
a_2 s^2 Y(s) + a_1 s Y(s) + a_0 Y(s) = b_1 s U(s) + b_0 U(s)
\end{eqnarray*}
\begin{eqnarray*}
Y(s) = \frac{(b_1 s  + b_0)}{(a_2 s^2 + a_1 s  + a_0)}  U(s) = G(s) U(s)
\end{eqnarray*}
Thus, in the Laplace domain, the output $Y(s)$ can be obtained by simply multiplying 
$G(s)$, called the {\em transfer function} of the system, with $U(s)$.
We can then take the inverse Laplace transform, $\L^{-1}[Y(s)]$, to obtain $y(t)$,
or as we will later see, we can simply analyze the stability of the system by examining 
the roots of the denominator of the transfer function $G(s)$ and their location in the complex s-domain.

Note that because $Y(s) =  G(s)$ for an impulse input, i.e. $U(s)=1$,
the transfer function $G(s)$ is also called {\em impulse response function}.

\subsection{Modeling a Vegas-like System}

Consider the system in Figure~\ref{fig:vegas-control} where a controller $G_c$ is used to match
the queue length $b(t)$ to a target $B_r$ by controlling the input window or sending rate $x(t)$.
The output rate from the queue is denoted by $c(t)$.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=2.5in]{vegas-control.jpg} % requires the graphicx package
   \caption{Vegas-like system}
   \label{fig:vegas-control}
\end{figure}
The goal is to first write down the differential equations that model the different components of the system,
then instead of solving the equations in the time domain, we will transform them to the Laplace domain and
analyze the stability of the system algebraically.

We start by describing the buffer evolution as:
\begin{eqnarray*}
\frac{d}{dt} b(t) = x(t) - c(t)
\end{eqnarray*}
Then, $x(t)$ is the output of convolving the error $e(t) = B_r - b(t)$ with the controller function $G_c(t)$,
 i.e. 
\begin{eqnarray*}
x(t) = G_c(t) \Conv e(t)
\end{eqnarray*}

Now, taking the Laplace transforms, we obtain:
\begin{eqnarray*}
s B(s) = X(s) - C(s)  \Rightarrow  B(s) = \frac{X(s) - C(s)}{s}
\end{eqnarray*}
\begin{eqnarray*}
X(s) = G_c(s) E(s) = G_c(s) (B_r(s) - B(s))
\end{eqnarray*}

Figure~\ref{fig:vegas-BD} shows the system using its transfer functions and their input/output flows,
where $G_0 = \frac{1}{s}$.
This is called the {\em block diagram} and provides a powerful pictorial tool.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=4in]{vegas-BD.jpg} % requires the graphicx package
   \caption{Block diagram of Vegas-like system}
   \label{fig:vegas-BD}
\end{figure} 
From the block diagram, one can write the algebraic equation of the output in terms of the input(s).
Dropping the ``s" parameter for convenience:
\begin{eqnarray*}
\frac{(B_r - B) G_c - C}{s} = B
\end{eqnarray*}
Rearranging, we get:
\begin{eqnarray}
B = \frac{G_c}{s + G_c} B_r - \frac{1}{s + G_c} C
\label{eqn:vegas-TF}
\end{eqnarray}
Note that the system has two inputs: $B_r(s)$ and $C(s)$, subjected to two transfer functions,
$\frac{G_c(s)}{s + G_c(s)}$ and $\frac{-1}{s + G_c(s)}$, respectively, and 
adding their responses we obtain the output $B(s)$.

\subsection{Proportional Control and Stability of Vegas-like System}

One basic controller $G_c$ is referred to as {\em Proportional (P)} controller, where
the controlled variable $x(t)$ is simply set in proportion to the error signal,
i.e. $x(t) = K_p\ e(t)$.
In this case, $G_c(s)$ is simply the constant $K_p$.

Substituting in Equation \ref{eqn:vegas-TF}, we have:
\begin{eqnarray}
B(s) = \frac{K_p}{s + K_p} B_r(s) - \frac{1}{s + K_p} C(s)
\label{eqn:P-ess}
\end{eqnarray}

An important question is: {\em does the P-controller make the system stable?}
More precisely, if we subject the system to impulse input(s),
does the system converge back to a quiescent state?
Control theory gives a systematic way to answer such stability question by
examining the roots of the denominator of the system's transfer function,
called the {\em characteristic equation}. 
In this case,
the characteristic equation is:
\begin{eqnarray*}
s + K_p = 0 \Rightarrow s = - K_p
\end{eqnarray*}
The system is stable if the roots (also called {\em poles}) lie in the left-hand side of the complex s-plane.
Thus this system is stable if $- K_p < 0 \Rightarrow K_p > 0$.

Note that we did not need to go back to the time domain to analyze the stability of the system.
But let's do that here to understand why poles on the  left-hand side of the s-plane makes the system stable.
Taking the inverse Laplace transform of  Equation \ref{eqn:P-ess}, and assuming impulse inputs,
i.e. $B_r(s) = C(s) = 1$, we get:
\begin{eqnarray*}
b(t) = K_p e^{-K_p t} - e^{-K_p t}
\end{eqnarray*}
We can then see that $b(t)$ decays exponentially over time, starting from $b(0)=(K_p-1)$.
We say that the system is stable or exhibits {\em overdamped response}.

We can also analyze transient performance by noting that
$b(0)=(K_p-1)$ represents an {\em overshoot} response to the impulse input,
and that this overshoot is lower for lower $K_p$.
See Figure~\ref{fig:overdamped}.
So by controlling $K_p$, referred to as the {\em controller gain},
we can also control the system's transient response.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=1.6in]{overdamped.jpg} 
   \caption{Overdamped response}
   \label{fig:overdamped}
\end{figure}

\subsection{Proportional Integral Control and Stability of Vegas-like System}

Another type of controller is known as {\em Proportional Integral (PI)} controller where
the controlled variable $x(t)$ is set in proportion to the integral of the error signal,
i.e. $x(t) = K_i\ \int e(t) \; dt$.
In this case, 
taking the Laplace transform, $G_c(s) = \frac{K_i}{s}$. 
Note that the integration means that the history of the error is used to control $x(t)$.

Substituting in Equation \ref{eqn:vegas-TF}, we have:
\begin{eqnarray}
B(s) = \frac{K_i}{s^2 + K_i} B_r(s) - \frac{s}{s^2 + K_i} C(s)
\label{eqn:PI-ess}
\end{eqnarray}

To analyze stability, we again examine the poles of the characteristic equation:
\begin{eqnarray*}
s^2 + K_i = 0  \Rightarrow s = \stackrel{+}{-} j \sqrt{K_i}
\end{eqnarray*}
Given $K_i > 0$, the two imaginary conjugate poles lie in the left-hand side of the complex s-plane,
and so the system is stable, though {\em critically} stable as we explain next.

To convince ourselves, let us go back to the time domain by taking the inverse Laplace transform:
\begin{eqnarray*}
\L^{-1}[\frac{K_i}{s^2 + K_i}] = \L^{-1}[\frac{K_i}{(s - j \sqrt{K_i})(s + j \sqrt{K_i})}] =
\L^{-1}[\frac{A}{(s - j \sqrt{K_i})} + \frac{B}{(s + j \sqrt{K_i})}] 
\end{eqnarray*}
And for some values of $A$ and $B$, this yields:
\begin{eqnarray*}
A e^{j \sqrt{K_i}t} + B e^{- j \sqrt{K_i}t}
\end{eqnarray*}
Given the fact that $e^{j \theta} =  \cs \; \theta + j \ \sn \; \theta$,
the function in the time domain oscillates in a sinusoidal fashion.
Although the time function does not decay over time,
it does {\em not} diverge, i.e. it is not unstable! 
So, we consider such a system to have bounded oscillations in response to
impulse input and we say that it is {\em critically (or marginally) stable} or 
the system exhibits  {\em undamped oscillatory response}.
Note that a higher value of $K_i$ results in more oscillatory behavior.
See Figure~\ref{fig:undamped}.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2.5in]{undamped.jpg} 
   \caption{Undamped response}
   \label{fig:undamped}
\end{figure}

\subsection{Stability and Pole Placement}

More formally, a linear time-invariant system is stable if
all poles of its transfer function are in the left-hand side of the s-plane,
i.e. the real part of all poles is negative.
Figure \ref{fig:pole-loc} shows the time response given the location 
of the poles.

Note that if the poles are complex conjugates and strictly in the left-hand side 
of the s-plane, the system is stable as oscillations in response to impulse input decay over time,
and we say that the system exhibits {\em underdamped response}.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=4in]{pole-loc.jpg} % requires the graphicx package
   \caption{Time response depending on pole location}
   \label{fig:pole-loc}
\end{figure}

\subsection{Transient Performance and Steady-state Error}
\label{sec:transient}

Besides stability, there are other performance metrics of interest that
characterize the transient performance of the system and the quality 
of the steady state.
Figure \ref{fig:transient} shows several of these metrics,
including the time for the controlled variable to reach its peak (maximum) value,
the time to reach the target, 
the maximum overshoot over the steady-state value,
and the error that remains at steady state
when the system stabilizes away from the desired target value.

\begin{figure}[htbp]
   \centering
   \includegraphics[width=4in]{perf-spec.jpg} % requires the graphicx package
   \caption{Performance Metrics}
   \label{fig:transient}
\end{figure}

For our Vegas-like system,
the controlled variable is the window size, i.e. number of packets
allowed into the system. The response is the queue length, which
we measure and compare to the target  buffer size.
A ``good" system is one that converges quickly to the desired target
with minimum oscillations (i.e., overshoots and undershoots) and
with almost zero steady-state error.

\subsection{Steady-state Error}

In control theory, the steady-state error of a stable system is defined as:
\begin{eqnarray*}
e_{ss} = \lm_{t \rightarrow \infty}\ e(t) = \lm_{t \rightarrow \infty}\  (r(t) - b(t))
\end{eqnarray*}
where $r(t)$ is the reference input, and $b(t)$ is the system output (response).
This error reflects how accurately the system can achieve the desired target,
which is chosen to be a step input.

We state without proof the Final Value Theorem \cite{Ogata:2010}:
\begin{eqnarray*}
e_{ss} = \lm_{t \rightarrow \infty}\ e(t) = \lm_{s \rightarrow 0}\ s\ E(s)
\end{eqnarray*}
This theorem allows us to calculate $e_{ss}$ algebraically in the Laplace domain.

\paragraph{Example (P-control of Vegas-like system):}
\begin{eqnarray*}
E(s) = B_r(s) - B(s)
\end{eqnarray*}
Substituting for $B(s)$ from Equation \ref{eqn:P-ess} and using the Final Value Theorem, we obtain:
\begin{eqnarray*}
e_{ss} = \lm_{s \rightarrow 0}\ s\ [B_r(s) - \frac{K_p}{s + K_p} B_r(s) + \frac{1}{s + K_p} C(s)]
\end{eqnarray*}

Assuming step inputs, i.e. $B_r(s) = \frac{B_r}{s}$ and $D(s) = \frac{D}{s}$, we have:
\begin{eqnarray*}
e_{ss} = \lm_{s \rightarrow 0}\ [B_r - \frac{K_p}{s + K_p} B_r + \frac{1}{s + K_p} C] = \frac{C}{K_p}
\end{eqnarray*}
Recall that under the P-controller, the system is (overdamped) stable, i.e. 
$b(t)$ approaches the target $B_r$ without oscillations, however,
at steady state, $b(t)$ misses the target by $\frac{C}{K_p}$ and 
stabilizes at a value lower than $B_r$.
Notice that the higher the service capacity $C$ is, the larger the steady-state error.
So, to decrease the steady-state error, the controller gain $K_p$ could be increased.
However, increasing $K_p$ increases the overshoot.
A tradeoff clearly exists between transient performance and steady-state performance,
and one has to choose $K_p$ to balance the two and meet desired operation goals.
 {\bf  End Example.}

\paragraph{Example (PI-control of Vegas-like system):}
\begin{eqnarray*}
E(s) = B_r(s) - B(s)
\end{eqnarray*}
Substituting for $B(s)$ from Equation \ref{eqn:PI-ess} and using the Final Value Theorem, we obtain:
\begin{eqnarray*}
e_{ss} = \lm_{s \rightarrow 0}\ s\ [B_r(s) - \frac{K_i}{s^2 + K_i} B_r(s) + \frac{s}{s^2 + K_i} C(s)]
\end{eqnarray*}

Assuming step inputs, i.e. $B_r(s) = \frac{B_r}{s}$ and $C(s) = \frac{C}{s}$, we have:
\begin{eqnarray*}
e_{ss} = \lm_{s \rightarrow 0}\ [B_r - \frac{K_i}{s^2 + K_i} B_r + \frac{s}{s^2 + K_i} C] = 0
\end{eqnarray*}
Although the steady-state error is zero under the PI-controller, 
recall that the system is critically stable, i.e. it converges to the target while
oscillating. Decreasing the controller gain $K_i$ decreases these oscillations,
however at the expense of longer time to reach steady state.
This illustrates again the inherent tradeoff between transient performance
and the quality of the steady state.

\section{Analyzing the Stability of Non-linear Systems}
\label{sec:nonlinear}

As we have seen, linear control theory can be applied to non-linear systems if we assume
a small range of operation around which the system behavior is linear
or approximately linear. This linear analysis
is simple to use, and the system, if stable, has a unique equilibrium point.

On the other hand, most control systems are non-linear, and operate over a wide range of
parameters, and multiple equilibrium points may exist. In this case, non-linear control theory 
could be more complex to use.

In what follows, we first consider a non-linear model of the adaptation of sources and network,
and use a non-linear control-theoretic stability analysis method, 
called {\em Lyapunov method} \cite{Ogata:2010}.
Then, we linearize the system and illustrate the application of linear control-theoretic analysis.

\subsection{Solving Non-linear Differential Equations}

Recall Vegas-like source adaptations from Equation \ref{eqn:source-adapt}:
\begin{eqnarray*}
\frac{d x_r(t)}{dt} = k [w_r - x_r(t) p_r(t)]
\end{eqnarray*}
where $p_r(t)$ represents the total price observed by user $r$ along its path.
Note that this differential equation is non-linear since $p_r(t)$ is a function of the rates $x_s(t)$:
\begin{eqnarray*}
p_r(t) = \sum_{{\rm link}\; l \in {\rm route}\; r}\ p_l(t) = \sum_{l \in r}\ p_l(\sum_{s: l \in s} x_s(t))
\end{eqnarray*}
We assume that the pricing function $p_l(y)$ is monotonically increasing in the load $y$.

At steady state, if the system stabilizes, setting the derivatives to 0, we obtain the steady-state solution:
\begin{eqnarray*}
 k [w_r - x_r\; p_r] = 0 \Rightarrow x_r = \frac{w_r}{p_r} = \frac{w_r}{\sum_{l \in r}\; p_l}
\end{eqnarray*}

To prove stability, we use the non-linear method of Lyapunov.
The basic idea is to find a positive scalar function $V(x(t))$, we call the Lyapunov function,
and show that the function monotonically increases (or decreases)
over time, approaching the steady-state solution.

Define $V(x)$ as follows:
\begin{eqnarray*}
V(x) = \sum_{r \in R} w_r \; \lg(x_r) - \sum_{j \in J} \int_{0}^{\sum_{s: j \in s} x_s} p_j(y) dy
\end{eqnarray*}
Finding a suitable Lyapunov function that shows stability is tricky and more of an art!
If you can't find one, it does {\em not} mean that the system is not stable.
Note that this $V(x)$ has some special meaning:
the first term represents the utility gain from making users happy,
while the second term represents the cost in terms of price.
So $V(x)$ represents the net gain.
Also, note that since the first term is concave because of the log function,
and the second term is assumed to be monotonically increasing,
then the resulting $V(x)$ is concave, i.e. it has a maximum value.

To show that $V(x(t))$ is strictly convergent, we want to show that $\frac{dV(x(t))}{dt} > 0$,
which implies that $V(x(t))$ strictly increases (i.e. the net gain keeps increasing over time), 
until the system stabilizes and reaches steady state when $\frac{dV(x(t))}{dt} = 0$ 
(i.e. the net gain $V(x)$ reaches its maximum value).

First, we note:
\begin{eqnarray*}
\frac{\partial V(x)}{\partial x_r} = \frac{w_r}{x_r} - \sum_{j \in r} p_j(\sum_{s: j \in s} x_s)
\end{eqnarray*}

Then:
\begin{eqnarray*}
\frac{V(x(t))}{dt} = \sum_{r \in R} \frac{\partial V(x(t))}{\partial x_r} \; \frac{d x_r(t)}{dt}
\end{eqnarray*}

\begin{eqnarray*}
\frac{V(x(t))}{dt} = \sum_{r \in R} [\frac{w_r}{x_r} - \sum_{j \in r} p_j(\sum_{s: j \in s} x_s(t))] \; k [w_r - x_r(t) p_r(t)]
\end{eqnarray*}

\begin{eqnarray*}
\frac{V(x(t))}{dt} = \sum_{r \in R} k \;x_r(t) [\frac{w_r}{x_r} - \sum_{j \in r} p_j(\sum_{s: j \in s} x_s(t))]^2 > 0
\end{eqnarray*}

Observe that this non-linear stability analysis shows that the system is stable, no matter what 
the initial state $x(0)$ is. This property is referred to as {\em global stability},
which is in contract to {\em local stability} that we prove when the system is linearized locally
around a certain operating point as we will see next.

\subsection{Linearizing and Solving Linear Differential Equations}

As noted earlier, finding Lyapunov functions to prove global stability of non-linear control systems,
even for simple models, is challenging. For example, consider more sophisticated models
with feedback delay, different regions of TCP operation (e.g., timeouts, slow-start),
queue management with different operating regions (e.g., RED), and
challenging or adversarial environments (e.g., exogenous losses over wireless links or due to DoS attacks).

Using linearization, we can separately study simpler linear models around the different points (regions)
of operation. More specifically, we linearize the system around a single operating point $x^*$ and
study perturbations around $x^*$, i.e. if we perturb the system away from $x^*$ to a point $x(0)$ such that the initial perturbation $\delta x(0) = x(0) - x^*$, we want to show that $\delta x(t)$ diminishes over time and the system returns to its original state $x^*$, i.e. $\delta x(t) \rightarrow 0$. In this case, we say that the system is {\em locally stable} around $x^*$.

Let's consider again the Vegas-like source adaptation and assume, for simplicity, a single user over a single resource:
\begin{eqnarray*}
\frac{d x(t)}{dt} = k [w - x(t) p(x(t))]
\end{eqnarray*}

Define the perturbation $\delta x(t) = x(t) - x^*$. Then we can write:
\begin{eqnarray*}
\frac{d (\delta x(t) + x^*)}{dt} = k [w - (\delta x(t) + x^*) p((\delta x(t) + x^*))]
\end{eqnarray*}

Expanding the non-linear term $p((\delta x(t) + x^*))$ into its first-order Taylor series:
\begin{eqnarray*}
p((\delta x(t) + x^*)) \approx p(x^*) + \dot{p}(x^*) \delta x(t)
\end{eqnarray*}

Substituting with this linear approximation, we get:
\begin{eqnarray*}
\frac{d \delta x(t)}{dt} = k [w - (\delta x(t) + x^*) \{p(x^*) + \dot{p}(x^*) \delta x(t)\}]
\end{eqnarray*}

\begin{eqnarray*}
\frac{d \delta x(t)}{dt} = k [w - \delta x(t) p(x^*) - x^* p(x^*) - \dot{p}(x^*) \delta^2 x(t) - \dot{p}(x^*) x^* \delta x(t)]
\end{eqnarray*}

If $x^*$ is the optimal steady-state point, we know that $w - x^* p(x^*) = 0$ (cf.\ Equation~\ref{eqn:kkt}). 
Also, given small perturbation $\delta x(t)$, $\delta^2 x(t) \approx 0$.
Then, we have:
\begin{eqnarray*}
\frac{d \delta x(t)}{dt} = k [ - \delta x(t) p(x^*)  - \dot{p}(x^*) x^* \delta x(t)]
\end{eqnarray*}

\begin{eqnarray*}
\frac{d \delta x(t)}{dt} = -k [p(x^*)  + x^* \dot{p}(x^*) ]\; \delta x(t)
\end{eqnarray*}

Let $k[p(x^*)  + x^* \dot{p}(x^*)] = \gamma$, we have:
\begin{eqnarray}
\frac{d \delta x(t)}{dt} = - \gamma\ \delta x(t)
\label{eqn:linearized}
\end{eqnarray}

This is now a linear differential equation, which unlike the original non-linear differential equation,
we can easily study using linear control-theoretic techniques, or
in this simple case, solve by straightforward integration:
\begin{eqnarray*}
\int_{0}^{t} \frac{d\; \delta x(t)}{\delta x(t)} = - \gamma\ \int_{0}^{t} dt
\end{eqnarray*}

\begin{eqnarray*}
\lg (\delta x(t)) -  \lg (\delta x(0)) = - \gamma t
\end{eqnarray*}

\begin{eqnarray*}
\lg (\frac{\delta x(t)}{\delta x(0)})  = - \gamma t  
\end{eqnarray*}

\begin{eqnarray*}
\delta x(t)  = \delta x(0)\; e^{- \gamma t}
\end{eqnarray*}

Note that from this time-domain analysis, the system is shown to be stable, i.e. the perturbation vanishes over time and the system returns to its original state $x^*$. We also observe that the system response decays exponentially from its original perturbation $\delta x(0)$, i.e. without oscillations, and so the response is classified as {\em overdamped}.

If the linearized differential equation modeling the system were more complicated, it is much easier to transform it into the Laplace domain and analyze the system algebraically. 
Denoting $\delta x(t)$ by $u(t)$, the Laplace transform of $\delta x(t)$ by $U(s)$, and taking the Laplace transform of Equation \ref{eqn:linearized}, we get:
\begin{eqnarray*}
s\; U(s) - u(0) = - \gamma\ U(s)
\end{eqnarray*}

\begin{eqnarray*}
U(s)  = \frac{u(0)}{s + \gamma }  
\end{eqnarray*}

For stability analysis, we examine the location of the poles (roots) of the characteristic equation $s + \gamma = 0$, yielding the pole $s = -\gamma$. Since the pole is strictly in the left-side of the s-plane, given $\gamma > 0$, the system is stable and its response is overdamped.

To evaluate the steady-state error, we define the error as $e(t) = u(0) - u(t)$, and applying the Final Value Theorem with an impulse perturbation of magnitude $u(0)$, i.e. $U(0)=u(0)$, we obtain:
\begin{eqnarray*}
e_{ss} = \lm_{s \rightarrow 0}\ s E(s) = \lm_{s \rightarrow 0}\ s [u(0) - \frac{u(0)}{s + \gamma } ] = 0
\end{eqnarray*}
So, there is no steady-state error.

\subsubsection{Effect of Feedback Delay and Nyquist Stability Criterion}

As we just noted above, the power of solving the linearized model in the Laplace domain comes when the model is even slightly more complicated. For example, let us consider a feedback delay $T$ such that Equation \ref{eqn:linearized} looks like:
\begin{eqnarray*}
\frac{d u(t)}{dt} = - \gamma\ u(t - T)
\end{eqnarray*}

Taking the Laplace transform, and noting that the Laplace transform of a delayed signal $u(t-T)$ is
$e^{-sT} U(s)$, we obtain:
\begin{eqnarray*}
s U(s) - u(0) = - \gamma\ e^{-sT} U(s)
\end{eqnarray*}

\begin{eqnarray*}
U(s) = \frac{u(0)}{s + \gamma\ e^{-sT}}
\end{eqnarray*}

Then, the characteristic equation is:
\begin{eqnarray}
s + \gamma\ e^{-sT} = 0
\label{eqn:delayed}
\end{eqnarray}
which we need to solve to locate the poles and determine the stability of the system.

To solve such characteristic equation, we resort to another control-theoretic method called the {\em Nyquist stability criterion} \cite{Ogata:2010}.  To this end, we introduce, without proof,
the {\em Cauchy's principle} \cite{Ogata:2010}, 
which states that given $F(s)$, and we plot $F(s)$ as we vary $s$ along a certain contour (trajectory) in the s-plane --- see Figure~\ref{fig:cauchy} --- and denote the following:
\begin{itemize}
\item $Z$: the number of {\em zeros} of $F(s)$, i.e. the roots of the numerator of $F(s)$, inside the contour.

\item $P$: the number of {\em poles} of $F(s)$, i.e. the roots of the denominator of $F(s)$, inside the contour.

\item $N$: the number of times that the plot of $F(s)$ encircles the origin in the $F(s)$-plane, 
such that an encirclement is negative if it is in the opposite direction of the s-contour.\footnote{
The example in Figure~\ref{fig:cauchy} shows that $N=-3$.
The value of $N$ is negative because the s-contour that is mapped to $F(s)$ is clockwise,
whereas the resulting $F(s)$ plot is counter-clockwise.
Moreover, the $F(s)$ path encircles the origin in the  $F(s)$-plane three times --
drawing a line emanating from  the origin intersects the $F(s)$ path three times.
}
\end{itemize}

Then the following relationship holds:
\begin{eqnarray*}
Z = P + N
\end{eqnarray*}

\begin{figure}[htbp]
   \centering
   \includegraphics[width=4in]{cauchy.jpg} % requires the graphicx package
   \caption{Cauchy's principle}
   \label{fig:cauchy}
\end{figure}

The Nyquist method applies the Cauchy's principle as follows.
Let's say we want to analyze the stability of a closed-loop control system
whose forward transfer function is $G(s)$ and its feedback transfer function is $H(s)$---see Figure \ref{fig:closed-loop}.
Then, the closed-loop transfer function is given by $\frac{G(s)}{1+G(s) H(s)}$,
where $G(s)H(s)$ is referred to as the {\em open-loop} transfer function.
The characteristic equation is given by: $F(s) = 1 + G(s) H(s) = 0$.
Observe that the zeros of $F(s)$ are the closed-loop poles,
and the poles of $F(s)$ are the poles of $G(s)H(s)$ (so-called ``open-loop" poles).
\begin{figure}[htbp]
   \centering
   \includegraphics[width=3in]{closed-loop.jpg} % requires the graphicx package
   \caption{Typical closed-loop control system}
   \label{fig:closed-loop}
\end{figure}

By taking the s-contour to be around the right-side (i.e. unstable side) of the s-plane
(see Figure \ref{fig:s-contour}),
and noting the number of {\em unstable} open-loop poles $P$
and the number of encirclements $N$ around the origin in the $F(s)$-plane, we determine 
the number of {\em unstable} zeros $Z$ of $F(s)$, i.e. number of unstable closed-loop poles,
using Cauchy's relationship: $Z = P + N$.
If $P = 0$ and $N = 0$, then $Z = 0$ implies that there are no {\em unstable} closed-loop poles
and so the closed-loop system is stable.\footnote{
Note that a pole on the imaginary axis is not considered unstable and
the s-contour avoids such a pole and we show it as a small circle around it.
}
\begin{figure}[htbp]
   \centering
   \includegraphics[width=3in]{s-contour.jpg} % requires the graphicx package
   \caption{Contour around the unstable right-side of the s-plane}
   \label{fig:s-contour}
\end{figure}

This process can be slightly simplified if instead of plotting $F(s)$,
we instead plot the open-loop transfer function: $G(s)H(s)$ and 
observe its encirclements of the $(-1, j 0)$ point in the $G(s)H(s)$-plane,
instead of  the origin $(0, j 0)$ in the $F(s)$-plane.
Given there are no poles of $G(s)H(s)$ in the right-side of the s-plane, i.e. $P=0$,
in order for the closed-loop system to be stable,
the plot of  $G(s)H(s)$ should not encircle -1 as we vary $s$ along the contour enclosing
the right-side of the s-plane.
We are mostly interested in varying $s$ along the imaginary axis,
i.e. $s = j \omega$ where $\omega$ varies from 0 to $\infty$.
This is because the plot for $\omega$ from $- \infty$ to 0 is symmetric,
and the plot for the semi-circle as $s \rightarrow \infty$ maps to the origin in the $G(s)H(s)$-plane.
Thus, we are interested in plotting $G(j \omega) H(j \omega)$
as $\omega$ varies from 0 to $\infty$.

\paragraph{Example:}
Let's go back to the characteristic equation in Equation \ref{eqn:delayed}:
\begin{eqnarray*}
s + \gamma\ e^{-sT} = 0  \Rightarrow F(s) = 1 + \frac{\gamma}{s} e^{-sT} \Rightarrow G(s)H(s) = \frac{\gamma}{s} e^{-sT}
\end{eqnarray*}
Note that $G(s)H(s)$ does not have any unstable poles, i.e. $P=0$. 
In particular, $s=0$ is considered a (critically) stable pole.

Ignoring the constant factor $\gamma$ for now, we want to plot:
\begin{eqnarray*}
\frac{e^{-j \omega T}}{j \omega} \ \ \ \ \ \  \omega: 0 \rightarrow \infty
\end{eqnarray*}

Noting that $e^{j \theta} = \cs \; \theta + j\; \sn \; \theta$, we have:
\begin{eqnarray*}
e^{-j \omega T} = \cs(\omega T) - j\; \sn(\omega T) 
\end{eqnarray*}
Then,
\begin{eqnarray*}
\frac{e^{-j \omega T}}{j \omega} = - j \frac{\cs(\omega T)}{\omega} - \frac{\sn(\omega T)}{\omega} 
\end{eqnarray*}

Since we are interested in determining intercepts with the real-axis of $G(j \omega) H(j \omega)$
and whether they occur to the right or left of -1 (see Figure~\ref{fig:ex-nyquist}),
we want to determine the values of $\omega$ 
for which the imaginary part of $G(j \omega) H(j \omega)$, i.e. $- \frac{\cs(\omega T)}{\omega}$, is zero.
Such intercepts occur when $\omega T = \frac{\pi}{2}, \frac{3 \pi}{2}, \cdots$, when the cosine value is zero.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=3in]{ex-nyquist.jpg} 
   \caption{Example showing the effect of feedback delay}
   \label{fig:ex-nyquist}
\end{figure}

Now, at these values of $\omega T$, we can determine the points of interception along the real-axis, i.e.
the magnitude $|G(j \omega) H(j \omega)|$ when the plot intercepts the real-axis:
\begin{eqnarray*}
 - \frac{\sn(\omega T)}{\omega} = - \frac{1}{\frac{\pi}{2T}}, +\frac{1}{\frac{3 \pi}{2T}}, \cdots
\end{eqnarray*}
For the system to be stable, $|G(j \omega) H(j \omega)|$ at these intercepts must be less than 1,
so the $G(s) H(s)$ plot does not encircle -1.
This is the case if after restoring the constant factor $\gamma$ we initially ignored, 
the following condition holds:
\begin{eqnarray*}
\gamma \frac{2T}{\pi} < 1
\end{eqnarray*}  {\bf End Example.}

Observe that $T$ is the feedback delay, so as $T$ gets larger, it becomes harder to satisfy the stability condition.
Intuitively, this makes sense since a larger feedback delay results in outdated feedback (measurements) and 
it becomes impossible to stabilize the system. 
This is the fundamental reason why TCP over long-delay paths does not work, and
architecturally, control has to be broken up into smaller control loops.

\section{Routing Dynamics}
\label{sec:routing}

So far, we assumed routes taken by flows to be static. In general, routes may also be adapted based on feedback on link prices (reflecting load, delay, etc.), 
albeit over a longer timescale of minutes, hours or even days compared to that of milliseconds for sending rate adaptation.
Figure \ref{fig:block-routing} shows a block diagram that includes both route and sending rate adaptation.
\begin{figure}[htbp]
   \centering
   \includegraphics[width=5in]{block-routing.jpg} % requires the graphicx package
   \caption{Block diagram with both flow and routing control}
   \label{fig:block-routing}
\end{figure}

Figure \ref{fig:routing-conv} illustrates the general process of adaptation.
Flow or routing control determines the amount of load directed to a particular link based
on the link's observed price --- 
relative to that of other possible links on alternate routes in the case of routing.
We call this mapping from link price $\lambda$ to link load $x$, 
the {\em response function} $G(\lambda)$.
Given link load, a certain price is observed for the link.
We call such load-to-price mapping, the {\em pricing (feedback) function} $H(x)$.
The process of adaptation is then an iterative process:
\begin{eqnarray*}
\lambda = H(x) \\
x = G(\lambda)
\end{eqnarray*}

\begin{figure}[htbp]
   \centering
   \includegraphics[width=4in]{routing-conv.jpg} % requires the graphicx package
   \caption{Convergence}
   \label{fig:routing-conv}
\end{figure}

We can then write:
\begin{eqnarray*}
\lambda = H(G(\lambda))  = F(\lambda)
\end{eqnarray*}
where $F(\lambda)$ is an {\em iterative} function whose stable (fixed) point $\lambda^*$ is the intersection of
the response function and the pricing function.
Figure \ref{fig:contractive} illustrates convergence to a fixed point.
Starting from an initial $\lambda_0$,
we find $F(\lambda_0)$,
then projecting on the $45^{o}$ line we obtain $\lambda_1 = F(\lambda_0)$,
which we use to find $F(\lambda_1)$,
and this iterative process continues until we reach the fixed point.

\begin{figure}[htbp]
   \centering
   \includegraphics[width=3in]{contractive.jpg} % requires the graphicx package
   \caption{Contractive mapping}
   \label{fig:contractive}
\end{figure}

In order to converge to that fixed point, $F(\lambda)$ must be a so-called {\em contractive mapping}.
Intuitively,
$F(\lambda)$ is contractive iff its slope is less than 1, i.e.
$|F(\lambda_2) - F(\lambda_1)| < \alpha |\lambda_2 - \lambda_1|$, $\alpha < 1$.
Figure \ref{fig:non-contractive} illustrates a mapping that results in divergence.

\begin{figure}[htbp]
   \centering
   \includegraphics[width=3in]{non-contractive.jpg} % requires the graphicx package
   \caption{Divergent mapping}
   \label{fig:non-contractive}
\end{figure}

Intuitively, the use of Lyapunov functions to prove convergence tests whether the iterative process
describing the evolution of the system over time is a contractive mapping, i.e. 
the distance to the fixed point keeps shrinking at every iteration.

\paragraph{Example:}
Consider the adaptive routing of $N > 0$ unit-rate flows over two possible paths whose prices are given by
{\em monotonically increasing} functions
$p_1(x)$ and $p_2(N-x)$, where $x$ represents the number of flows (or load) routed on the first path.
Note that $x$ completely defines the state of the system.
Also, assume that routing to the least-loaded path is done gradually, to avoid wild oscillations,
where $0 < \alpha < 1$ of the flows are re-routed.
Using a discrete-time model where routes are adapted at discrete-time steps, 
we can write the following difference equations:
$$
x(t+1) = \left\{ \begin{array}{rl} 
               x(t) + \alpha \; [N - x(t)]  & \mbox{if  $p_1(x(t)) \leq p_2(N -  x(t))$ } \\
               x(t) - \alpha \; x(t) & \mbox{otherwise}
                        \end{array} \right.
$$

At steady state, this system might converge to one of two possible stable (fixed) points.
One possibility is obtained when substituting with $x(t) \rightarrow x^*$ in the first difference equation: 
$x(t) \rightarrow x^* \Rightarrow x^* = x^* + \alpha [N - x^*] \Rightarrow x^* = N$, 
so all traffic will end up getting routed on the first path. 
A {\em necessary} condition to reach that $x^* = N$ fixed point is that $p_1(N) \leq p_2(0)$, i.e. 
the first path is least loaded (priced) even when all $N$ flows are on it.

Another possibility is obtained when substituting with $x(t) \rightarrow x^*$ in the second difference equation:
$x(t) \rightarrow x^* \Rightarrow x^* = x^* - \alpha x^* \Rightarrow x^* = 0$, 
so all traffic will end up getting routed on the second path. 
A {\em necessary} condition to reach that $x^* = 0$ fixed point is that $p_1(0) > p_2(N)$, i.e. 
the second path is least loaded (priced) even when all $N$ flows are on it.

We can show convergence to one of these fixed points depending on which necessary condition holds: 
$p_1(N) \leq p_2(0)$ or $p_1(0) > p_2(N)$.

Let's assume $p_1(N) \leq p_2(0)$ holds. We want to define a Lyapunov function $V(x) \geq 0$ and show that
$V(x(t+1)) \leq V(x(t))$ for some or all starting state $x(0)$, 
i.e. $V(x)$ monotonically decreases toward the $x^* = N$ fixed point
where equality holds.
If there are only certain values of the starting state $x(0)$ for which the system converges then
such conditions must hold, in addition to the necessary condition,
for convergence to happen. 
In this case, we say that the necessary condition by itself is not sufficient for convergence.

Define $V(x) = N - x$. Note that $V(x) \geq 0$ because $0 \leq x \leq N$, and $V(x) = 0$ when $x = N$, i.e. at the fixed point. 
So, under convergence, we expect $V(x)$ to monotonically decrease toward zero. 
Substituting for $x(t+1)$ in $V(x)$, we obtain:
\begin{eqnarray*}
V(x(t+1)) = N - x(t+1) 
\end{eqnarray*}

Given the pricing functions are monotonically increasing with load,
$p_1(N) \leq p_2(0) \Rightarrow p_1(x(t)) \leq p_2(N -  x(t))$, $\forall x(t)$, 
and we can only use the first difference equation to substitute for $x(t+1)$:
\begin{eqnarray*}
V(x(t+1)) = N - (x(t) + \alpha [N - x(t)]) = (1 - \alpha)(N - x(t)) 
\end{eqnarray*}
\begin{eqnarray*}
V(x(t+1)) =   (1 - \alpha) V(x(t)) \leq V(x(t))
\end{eqnarray*}
So, we can conclude that the system is convergent regardless of the starting state $x(0)$ as long as $0 < \alpha < 1$.

Thus, $0 < \alpha < 1$, along with the necessary condition $p_1(N) \leq p_2(0)$,
represent necessary and sufficient conditions for convergence.

A similar convergence proof can be obtained if on the other hand, the necessary condition $p_1(0) > p_2(N)$ holds.
{\bf End Example.}

\section{Case Study: Class-based Scheduling of Elastic Flows}
\label{sec:case1}

In this and the following section, 
we consider the modeling and control-theoretic analysis of two traffic control case studies. 
This first case study \cite{MattaGuo:icnp00,GuoMatta:icnp01} 
concerns the performance of elastic  flows, i.e., rate-adaptive flows similar to TCP.
The goal is to investigate the effect of class-based scheduling
that isolates elastic flows into two classes (service queues) based on different characteristics,
for example based on their lifetime (transfer size), 
or burstiness of their arrivals/departures and
sending rate (window) dynamics.
We want to show the benefits of isolation,
in terms of better predictability and fairness,
over traditional shared queueing systems.

We formulate 
two control models.
In the first model (Section~\ref{sec:aggregate-control}),
each flow controls its input traffic rate
based on the {\em aggregate} state of the network
due to all $N$ flows.
In the second model (Section~\ref{sec:individual-control}),
each flow  (or class of
homogeneous flows) controls its rate based
on its own {\em individual} state within the network.
We assume that the flows use PI control for adapting their sending rate.

In the aggregate control model,
the packet sending rate of flow $i$,
denoted by $x_i(t)$, is adapted based
on the difference between a target {\em total} buffer space,
denoted by $B$, 
and the current {\em total} number of outstanding packets,
denoted by $q(t)$.
In the individual control model,
$x_i(t)$ is adapted based on
flow (or class) $i$'s target, denoted by $B_i$,
and its current number of outstanding packets,
denoted by $q_i(t)$.
We denote by $c(t)$
the total packet service rate,
and by $c_i(t)$
the packet service rate of flow/class $i$.
In what follows,
for each control model,
we determine conditions under which
the system stabilizes.
We then solve for the values of the state variables
at equilibrium, and show whether fairness
(or a form of weighted resource sharing) can be achieved.
Table~\ref{tab:model} lists all system variables
along with their meanings.

{\small
\begin{table}[htbp]
\begin{center}
\caption{Table defining system variables}
\label{tab:model}
\begin{tabular}{|l|l|} \hline
Variable & Meaning \\\hline \hline
$N$ & total number of flows (or classes of
homogeneous flows) \\
$x_i(t)$  & packet sending rate of flow/class $i$  \\
$q_i(t)$  & number of outstanding packets of flow/class $i$  \\
$c_i(t)$  & packet service rate of flow/class $i$  \\
$q(t)$    & total number of outstanding packets  \\
$c(t)$    & total packet service rate \\ 
$B$      & target total buffer space \\
$B_i$    & target buffer space allocated to flow/class $i$ \\
$\alpha_i$ & parameter controlling increase and 
	decrease rate of $x_i(t)$ \\ \hline 
\end{tabular}
\end{center}
\end{table}
}

\subsection{Aggregate Control or Sharing}
\label{sec:aggregate-control}

Under aggregate PI control,
the evolution of the system state is described by the following
differential equations:
\begin{eqnarray}
\dot{x}_i(t) &=& \alpha_i (B-q(t)) \nonumber \\ 
\dot{q}(t) &=& \sum_{j=1}^{N} x_j(t) - c(t) 
\label{eqn:aggregate-state}
\end{eqnarray}

\paragraph{Stability Condition:}
Without loss of generality,
assume a constant packet service rate
(i.e.\ $c(t) = C$ for all $t$),
all flows start with the same initial input state
(i.e.\ $x_i(0)$ is the same for all $i$),
and that all flows adapt at the same rate
(i.e.\ $\alpha_i = \alpha$ for all $i$).
Then, Equations~(\ref{eqn:aggregate-state}) can be re-written as:
\begin{eqnarray}
\dot{x}_i(t) &=& \alpha (B-q(t))  \nonumber \\ 
\dot{q}(t) &=& \sum_{j=1}^{N} x_j(t) - C 
\label{app:aggregate-state}
\end{eqnarray}
Since flows adapt their $x_i(t)$ at the same rate,
then $x_i(t) = \frac{\sum_{j=1}^{N} x_j(t)}{N}$ for all $i$.
Denote by $e(t)$ the error at time $t$,
i.e.\ $e(t) = B-q(t)$,
and let $y(t) =  \sum_{j=1}^{N} x_j(t) - C$.
Equations~(\ref{app:aggregate-state}) can then be re-written as:
\begin{eqnarray}
\frac{\dot{y}(t)}{N} &=& \alpha\ e(t)  \nonumber \\ 
\dot{q}(t) &=& y(t)
\label{app:aggregate-state-2}
\end{eqnarray}
Taking the Laplace transform 
of Equations~(\ref{app:aggregate-state-2}), 
and assuming the buffer is initially empty (i.e. $q(0)=0$),
we get:
\begin{eqnarray}
\frac{1}{N} (s Y(s) - y(0))  &=& \alpha\ E(s)  \nonumber \\ 
s\ Q(s) &=& Y(s) \nonumber \\
E(s) &=& B - Q(s) 
\label{app:aggregate-state-Z}
\end{eqnarray}

Solving Equations~(\ref{app:aggregate-state-Z}), we obtain the closed-loop system's characteristic equation
(see Figure~\ref{fig:case1-block} for the system's block diagram):
\begin{eqnarray}
s^2 + \alpha\; N = 0 \Rightarrow s = \stackrel{+}{-} j \sqrt{\alpha\; N}
\label{eqn:poles-aggr}
\end{eqnarray}

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{case1-block.jpg} 
   \caption{Block diagram of the link sharing model}
   \label{fig:case1-block}
\end{figure}

For $\alpha > 0$, this system is marginally stable.
However,  the magnitude of oscillations increases for higher $\alpha$ and/or higher $N$.

This indicates that 
the existence of flows that
rapidly change their sending rates
through high values of $\alpha_i$ 
can cause the system to have high oscillations.
This suggests that elastic flows
that aggressively change their sending rates,
may affect the stability of other flows that
change their sending rates cautiously,
in a system that mixes both kinds of flows.
Furthermore, in such a system,
the value of $N$ may be high so as 
to cause high oscillations.

We now derive the values of the state variables
at equilibrium.
Denote by $x_i$ and
$q$ the steady-state values
of $x_i(t)$ and $q(t)$, respectively.
Then, at equilibrium, we have from Equations~(\ref{app:aggregate-state}):
\begin{eqnarray*}
0 &=&  \alpha (B - q)  \nonumber \\ 
0 &=&  \sum_{j=1}^{N} x_j - C 
\end{eqnarray*}
Thus, at equilibrium, $q = B$ and $\sum_{j=1}^{N} x_j = C$.
In other words,
the system converges to a state where
the total input rate matches the total service rate,
and the target total buffer space is met.

Note that if 
$\alpha_i = \alpha$ for all $i$,
then $x_i(t)$ changes at the same rate
for every flow $i$.
Consequently, if we start the evolution of
the system with $x_i(0)$ being the same
for all flows,
only then we have equal sharing of the network
at steady-state, i.e.\ $x_i = \frac{C}{N}$,
regardless of the initial value $q(0)$.
However, in general,
when $x_i(0)$ are not equal
for all flows,
the system converges to an {\em unfair} state,
more precisely, to a state where
\begin{eqnarray*}
x_i = x_i(0) + \frac{C - \sum_{j=1}^{N} x_j(0)}{N}
\end{eqnarray*}

To summarize,
controlling several flows 
by observing the resulting aggregate state of
the network may lead to high oscillations 
due to either
the existence of flows which
are rapidly adjusting their sending rates,
or a high number of flows competing
for the same {\em shared} resource.
Furthermore, 
the system is highly likely to converge to
an unfair state where flows receive
unequal shares of the resource.

\subsection{Individual Control or Isolation}
\label{sec:individual-control}

Under individual PI control,
the evolution of the system state is described by the following
differential equations:
\begin{eqnarray}
\dot{x}_i(t) &=&  \alpha_i (B_i-q_i(t))  \nonumber \\ 
\dot{q}_i(t) &=&  x_i(t) - c_i(t) 
\label{eqn:individual-state}
\end{eqnarray}

Recall that under individual control,
flow/class $i$ regulates its input, $x_i(t)$,
based on its {\em own} number of outstanding packets.
For simplicity,
assume a constant packet service rate,
i.e.\ $c_i(t) = C_i$ for all $t$.
Following the same stability analysis as aggregate control,
it is easy to see that flow/class $i$ stabilizes and
the poles of the closed-loop system are:\footnote{
We set $N = 1$ in Equation~(\ref{eqn:poles-aggr}).}
\begin{eqnarray*}
	s = \stackrel{+}{-} j \sqrt{\alpha_i}
\label{eqn:individual-stability}
\end{eqnarray*}
Observe that, unlike aggregate control,
flows/classes are {\em isolated} from each other.
Therefore, 
the existence of flows/classes that
rapidly change their sending rates
through high values of $\alpha_i$, 
does not affect the stability of other flows.
This isolation can be implemented using,
for example, a class-based queueing (CBQ) discipline \cite{cbq:1995}.
In such CBQ system,
each class of homogeneous flows can be allocated
its own buffer space and service capacity.

We now derive the values of the state variables
of flow/class $i$
at equilibrium.
Denote by $x_i$ and
$q_i$ the steady-state values
of $x_i(t)$ and $q_i(t)$, respectively.
Then, at equilibrium, we have from Equations~(\ref{eqn:individual-state}):
\begin{eqnarray*}
0 &=& \alpha_i (B_i-q_i) \nonumber \\ 
0 &=&  x_i - C_i
\end{eqnarray*}
Thus, at equilibrium, $q_i = B_i$ and $x_i = C_i$.
In other words,
each flow/class $i$ converges to a state where
its input rate matches its allocated service rate,
and its target buffer space is met.
We note that 
if the allocated buffers $B_i$ and 
service capacities $C_i$ are equal,
then every flow receives an equal share
of the resources,
{\em regardless of} the initial values 
$x_i(0)$ and $q_i(0)$.
One can also achieve a weighted resource sharing
by assigning different $B_i$ and $C_i$ allocations.
Thus, a flow/class with higher priority
(e.g., short interactive TCP flows operating aggressively in slow start)
can be allocated more resources,
so as to receive better throughput/delay service
than other flows (e.g., long TCP flows operating cautiously in congestion avoidance).

To summarize,
controlling each flow (or class of homogeneous flows) 
by observing its own individual state 
within the network 
provides isolation between them.
Thus, stability can be achieved 
for a flow/class regardless of
the behavior and number of other flows/classes.
Furthermore, 
the system can converge to
a fair state where flows/classes receive
a weighted share of the resources.

\section{Case Study: Elastic Transport Tunnel}
\label{sec:case2}

  Consider $n$ regular user connections between sending and receiving
  end-hosts, all passing through two ``gateways" --- let's call them a source gateway and
  a destination gateway.  Our main goal
  is to provide a ``soft bandwidth-guaranteed" tunnel \cite{MinaBestavrosMattaX:iscc04} for these user
  flows over an Internet path of bottleneck capacity $C$, which is
  also shared by another set of $x$ flows, representing cross traffic
  (see Figure~\ref{fig:case2-tunnel}).
  By ``soft" guarantee we mean that there is no explicit resource reservation.
  Consider that user and cross-traffic connections
  are all rate-adaptive connections (similar to TCP). 
  These $x$ cross-traffic connections present a
  challenge: as $x$ keeps changing, the bandwidth allocation for the
  $n$ user flows keeps changing in tandem.  So an important
  question is whether it is possible to ``counter'' the change in $x$
  so as to ensure that the $n$ user flows are able to maintain a
  desirable bandwidth.

  \begin{figure}[htbp] %  figure placement: here, top, bottom, or page
     \centering
     \includegraphics[width=4in]{case2-tunnel.jpg} 
     \caption{Soft bandwidth-guaranteed tunnel}
     \label{fig:case2-tunnel}
  \end{figure}

Clearly without the intervention of the two gateways, the answer to the above
  question is {\em no}. When different flows share a link, the effect
  of each individual flow (or an aggregate of flows) affects the rest
  since all are competing for a fixed amount of resources.  However,
  if the gateways dynamically maintain a number $m$ of open rate-adaptive (e.g., TCP)
  connections between them, they can increase $m$ to provide a positive pressure that
  would equalize the pressure caused by the cross-traffic connections,
  if the latter occurs.  Since $m$ will be changing over time, we
  describe the gateway-to-gateway tunnel, made of the $m$ connections, as {\em elastic}.  Note that the
  source gateway can decide to reduce $m$ (i.e.\ relieve pressure) if $x$
  goes down---the reason is that as long as the tunnel is achieving
  its target bandwidth, releasing extra bandwidth should improve the
  performance of cross-traffic connections, which is in the spirit of
  best-effort networking.

To illustrate this scenario and  the issues involved,
  consider a gateway-to-gateway tunnel going through a single bottleneck
  link.  Assuming long-lived TCP-like load, the behavior of the bottleneck can be
  approximated by Generalized Processor Sharing (GPS) \cite{GPS:1994}, 
  i.e.\ each connection receives the same fair share of resources \cite{Chiu:1989}.
  Thus, each connection ends up with
  $\frac{C}{m+x}$ bandwidth.  This, in turn, gives the $m$ gateway-to-gateway rate-adaptive
  flows, or collectively the elastic gateway-to-gateway tunnel, a bandwidth of
  $\frac{Cm}{m+x}$.  As the source gateway increases $m$ by opening more
  rate-adaptive connections to the destination gateway, the tunnel can grab more
  bandwidth.  If $x$ increases, and the gateways measure a tunnel's
  bandwidth below a target value (say $B^*$), then $m$ is increased to
  push back cross-traffic connections.  If $x$ decreases, and the gateways
  measure a tunnel's bandwidth above $B^*$, then $m$ is decreased for
  the good of cross-traffic connections.  It is important to note that
  the source gateway should refrain from unnecessarily increasing $m$,
  thus achieving a tunnel's bandwidth above $B^*$, since an
  unnecessary increase in the total number of competing rate-adaptive flows
  reduces the share of each connection and may cause flows to
  timeout leading to inefficiency and unfairness.  
  The source gateway also has the responsibility
  of scheduling user packets coming on the $n$ user connections over
  the tunnel, i.e.\ the $m$ gateway-to-gateway connections.
  In this case study, we do not focus on scheduling but the control theoretic 
  modeling and analysis of the tunnel's bandwidth adaptation.
 We study the effect of different types of controllers employed at the source gateway.
 Such controller determines the degree of
elasticity of the gateway-to-gateway rate-adaptive tunnel,
thus it determines
the transient and steady-state behavior of
the soft bandwidth-guaranteed service. \\

\noindent{\bf{Na\"{\i}ve Control:}}
This na\"{\i}ve controller 
measures the bandwidth $b'$ grabbed by 
the current $m'$ gateway-rate-adaptive connections.
Then,
it directly computes the quiescent number $\hat{m}$
of gateway-rate-adaptive  connections that should be open as:
\begin{eqnarray*} 
 \label{eqn:q_m} 
 \hat{m} &=& \frac{B^*}{b'}m'
\end{eqnarray*}

\noindent
Clearly, this controller na\"{\i}vely relies on
the previously measured bandwidth $b'$
and adapts without regard to delays in
measurements and possible changes in network conditions,
e.g.\ changes in the amount of cross traffic.
We thus investigate general well-known controllers
which judiciously zoom-in toward the target bandwidth value.
To that end,
let us develop a flow-level model
of the system dynamics. 
The change in
the bandwidth grabbed $b(t)$ by the $m(t)$ 
gateway-rate-adaptive flows (constituting the elastic gateway-to-gateway tunnel) 
can be described as:
\begin{eqnarray*} 
 \label{eqn:b_dot}    
   \dot b(t) &=&   \alpha [ (C-B^*) m(t) - B^* x(t)]
\end{eqnarray*}

\noindent 
This dynamic equation captures what we want to model:
$b(t)$ increases with $m(t)$, and decreases
as the number of cross-connections $x(t)$ increases. 
$\alpha$ is a constant that represents the degree of multiplexing of
flows and we choose it here to be the steady-state connection's fair share
ratio of the bottleneck capacity. 
At steady-state,
$\dot b(t)$ equals zero, which yields (as expected):
\begin{eqnarray*} 
 \label{eqn:m_x_ss}    
   B^* &=&   \frac{C \hat m} {(\hat x+\hat m)}
\end{eqnarray*}
where $\hat m$ and $\hat x$ represent the steady-state values for the
number of gateway-rate-adaptive flows and cross-traffic flows, respectively.
\noindent
Based of the current bandwidth allocation $b(t)$ and the target
bandwidth $B^*$, an error signal $e(t)$ can be obtained as:
\begin{eqnarray*} 
 \label{eqn:e}    
   e(t) &=&   B^* - b(t)
\end{eqnarray*}

\noindent{\bf{P and PI Control:}}
A controller would adjust $m(t)$ based on the value
of $e(t)$. 
For a simple
Proportional controller (P-type), such adjustment is described by:
\begin{eqnarray} 
 \label{eqn:p}    
   m(t) &=&  K_p e(t)
\end{eqnarray}
\noindent
Recall that P-type controllers are known to result in a non-zero steady-state error.
To exactly achieve the target $B^*$ (i.e.\ with zero steady-state error),
a Proportional-Integral (PI-type) controller can be used:
\begin{eqnarray} 
 \label{eqn:pi}    
   m(t) &=& K_p e(t)+ K_i \int e(t) \ dt
\end{eqnarray}

\noindent
Figure \ref{fig:block1} shows the block diagram 
of this elastic-tunnel model.
In the Laplace domain,
denoting the controller transfer
function by $G_c(s)$, 
the output $B(s)$ is given by:
{\small
\begin{eqnarray} 
 \label{eqn:tf} B(s)= \frac{G_c(s)G_1(s)}{1+G_c(s)G_1(s)} B^*(s) +
   \frac{G_2(s)}{1+G_c(s)G_1(s)} X(s)
\end{eqnarray}
}
where $G_1(s)$ is given by:
\begin{eqnarray*} 
 \label{eqn:g1}    
   G_1(s) &=& \frac{\beta}{s}
\end{eqnarray*}
where $\beta = \alpha (C-B^*)$.
$G_2(s)$ is given by:
\begin{eqnarray*} 
 \label{eqn:g2}    
   G_2(s) &=& \frac{\gamma}{s}
\end{eqnarray*}
where $\gamma = -\alpha B^*$.
\noindent
For the P-controller, 
from Equation~(\ref{eqn:p}), $G_c(s)$ is simply $K_p$.
For the PI-controller, from Equation~(\ref{eqn:pi}),
$C(s)$
equals  $K_p + \frac{K_i}{s}$.
Thus,
the transfer function $\frac{B(s)}{B^*}$ 
in the presence of a  P-controller
is given by:
\begin{eqnarray*} 
 \label{eqn:tfp}    
    \frac{B(s)}{B^*} &=& \frac{K_p \beta}{s + K_p \beta}
\end{eqnarray*}
The system with P-controller is always stable
since the root of the characteristic equation 
(i.e.\ the denominator of the transfer function) is
negative, given by $-K_p \beta$ for $\beta > 0$ and $B^* < C$.
\noindent
In the presence of a PI-controller,
the transfer function $\frac{B(s)}{B^*}$ 
is given by: 
\begin{eqnarray*} 
 \label{eqn:tfpi} 
 \frac{B(s)}{B^*} &=& \frac{K_p\beta s + K_i\beta}
{s^2 + K_p\beta s + K_i\beta}
\end{eqnarray*}
One can choose the PI-controller parameters
$K_p$ and $K_i$ to achieve a certain convergence
behavior to the target bandwidth $B^*$.
We next define transient performance measures
to assess such convergence behavior.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{tunnel-block1} 
\caption{Block diagram of the elastic-tunnel model} 
\label{fig:block1}
\end{figure}

\subsection{Transient Performance}
Transient behavior represents the system's response 
which decays with
time. 
In the design of reliable systems,
it is extremely important 
that transient response meets certain requirements, 
e.g.\ reasonable settling time and overshoot (cf.\ Section~\ref{sec:transient}). 
Often times, the transient response is
obtained by subjecting the system to an impulse or a step input and
observing the output(s).
One has to guarantee that the response of the
system to specific inputs does not render the system
unstable or pushes it away from its intended target.

Figure~\ref{fig:trans1} shows
 the step response of the
transfer function given in
Equation~(\ref{eqn:tf}).
The left column 
shows  the response to a unit step change
in the target bandwidth, while the right column 
shows the
response to a unit step change in the cross-traffic. 
Figure~\ref{fig:trans1}(a),
for the P-controller,
shows that while the response could be acceptable due to
a step change in the reference bandwidth (i.e., the response $b(t)$ achieves the unit target), 
it fails to remove the steady-state
error (i.e., non-zero $b(t)$) obtained from the step change in the cross-traffic. 
Figures~\ref{fig:trans1}(b) and~(c)
show the response due to the PI-controller. 
One can
see that through a careful choice of $K_p$ and $K_i$, the transient
response can be adjusted.  
Notice that with a PI-controller,
the elastic-tunneling system can reach the target bandwidth with
zero steady-state error in response to a step change
in cross-traffic.

\begin{figure*}[thbp]
\begin{center} 
\begin{minipage}{0.4\textwidth} 
   \includegraphics[width=2.5in]{plots/trans/P_step_B_01.pdf} 
\centerline{Target Bandwidth Step Response}
\end{minipage}
\hspace{1.2cm}
\begin{minipage}{0.4\textwidth} 
   \includegraphics[width=2.5in]{plots/trans/P_step_x_01.pdf} 
\centerline{Cross-traffic Step Response}
\end{minipage}
\centerline{(a) Proportional controller with $K_p$ = 0.1}\\

\begin{minipage}{0.4\textwidth} 
   \includegraphics[width=2.5in]{plots/trans/PI_step_B_02_1.pdf} 
\centerline{Target Bandwidth Step Response}
\end{minipage}
\hspace{1.2cm}
\begin{minipage}{0.4\textwidth} 
   \includegraphics[width=2.5in]{plots/trans/PI_step_x_02_1.pdf} 
\centerline{Cross-traffic Step Response}
\end{minipage}
\centerline{(b) Proportional Integral controller with $K_p$ = 0.2 and $K_i$ = 1}\\

\begin{minipage}{0.4\textwidth} 
   \includegraphics[width=2.5in]{plots/trans/PI_step_B_1_05.pdf} 
\centerline{Target Bandwidth Step Response}
\end{minipage}
\hspace{1.2cm}
\begin{minipage}{0.4\textwidth} 
   \includegraphics[width=2.5in]{plots/trans/PI_step_x_1_05.pdf} 
\centerline{Cross-traffic Step Response}
\end{minipage}
\centerline{(c) Proportional Integral controller with $K_p$ = 1 and $K_i$ =
0.5}\\
\caption{Transient analysis of the elastic-tunnel model}
\label{fig:trans1} 
\end{center}
\end{figure*}

\begin{figure}[tbp]
   \includegraphics[width=5in]{tunnel-block} 
\caption{Elastic-tunnel model (with feedback delay)} 
\label{fig:block2}
\end{figure}

\subsection{Effect of Feedback Delay}
So far in our analysis, we have ignored the feedback delay which is
inherent in the design of any control system 
that tries to adjust its signal through
a delayed feedback loop.

Figure~\ref{fig:block2} augments the block diagram
of Figure~\ref{fig:block1}
with feedback delay denoted by $H(s)$. 
This feedback
delay arises either due to delayed
mesurements of bandwidth and/or
delayed response of the system as a result of
applying new control.
For example, when 
a new gateway-rate-adaptive connection is opened, 
it doesn't get its steady-state throughput instantaneously,
rather after some delay (say $\tau$). Thus, $H(s)$ is given by:
\begin{eqnarray*} 
 \label{eqn:hs}    
   H(s) &=& e^{-\tau s}
\end{eqnarray*}
where $\tau$ represents the feedback time delay.
For small $\tau$, the above
equation can be approximated by:
\begin{eqnarray*} 
 \label{eqn:ahs}    
   H(s) &\approx& 1 - \tau s
\end{eqnarray*}
If we are using a PI-controller, the characteristic equation in 
the presence
of feedback delay becomes:
\begin{eqnarray*} 
 \label{eqn:cch}    
    s^2(1-\beta \tau K_p) + s(K_p\beta - \beta \tau K_i) + \beta K_i = 0
\end{eqnarray*}

Figure~\ref{fig:trans2} shows the response
of the PI-controller to a unit step change in the target bandwidth.
As the feedback delay $\tau$ increases,
the system may not converge to the target bandwidth.

\begin{figure*}[htbp]
\begin{center} 
\begin{minipage}{0.4\textwidth} 
   \includegraphics[width=2.5in]{plots/trans/PI_step_B_001_002.pdf} 
\centerline{Target Bandwidth Step Response}
\end{minipage}
\hspace{1.2cm}
\begin{minipage}{0.4\textwidth} 
   \includegraphics[width=2.5in]{plots/trans/PI_step_B_001_002_delay_05.pdf}  
\centerline{Response with Feedback Delay}
\end{minipage}
\centerline{Proportional Integral controller with $K_p$ = 0.01 and
$K_i$ = 0.02}\\
\caption{Transient analysis of the elastic-tunnel model (with feedback delay)}
\label{fig:trans2} 
\end{center}
\end{figure*}

\section{Exercises}
\label{sec:exercises}

\begin{enumerate}

\item Let $r$ be a max-min fair rate vector corresponding to a given network and set of flows. This max-min fair allocation maximizes the allocation of each flow $i$ subject to the constraint that an incremental increase in $i$'s allocation does not cause a decrease in some other flow's allocation that is already as small as $i$'s or smaller.

\begin{itemize}
\item[(a)] Suppose that some of the flows are eliminated and let $\bar{r}$ be a corresponding max-min fair rate vector. Show by example that we may have $\bar{r}_p < r_p$ for some of the remaining flows $p$.

\item[(b)] Suppose that some of the link capacities are increased and let $\bar{r}$ be a corresponding max-min fair rate vector. Show by example that we may have  $\bar{r}_p < r_p$ for some flows $p$.
\end{itemize}

\item Consider two network links in tandem (one after the other) of capacities 6 Mbps each. For two different sets of elastic flows (i.e. the utility function of each flow/user is a log function of its allocated rate), you are asked to write down the corresponding network optimization problem, where the network tries to maximize the sum of flow utilities subject to link capacity constraints. For each set of flows, described in parts (a) and (b) below, rewrite the constrained optimization problem as an unconstrained optimization problem: write down the Lagrangian function and corresponding equations to solve for the optimal rate allocation. What are the optimal rates allocated to different flows for each one of these two settings?
\begin{itemize}
\item[(a)] Consider two flows: one flow using the first link only and another flow using both links.

\item[(b)] Consider the same two flows from part (a), as well as a third flow using the second link only.
\end{itemize}

\item  Consider two network links in tandem (one after the other) of capacities 6 Mbps each. 
Assume three flows: one flow using the first link only, another flow  using the second link only,
and a third flow using both links.
Assume the utility function of each flow/user is a log function of its allocated rate, 
and that the two-link flow is given a weight of 2, while the two one-link flows are given a weight of 1.

You are asked to write down the corresponding network optimization problem, where the network tries to maximize the sum of the weighted flow utilities subject to link capacity constraints.  Rewrite the constrained optimization problem as an unconstrained optimization problem: write down the Lagrangian function and corresponding equations to solve for the optimal rate allocation. What are the optimal rates allocated to each flow?

\item Consider a source adapting its sending rate $x(t)$ so the buffer size of  its path's bottleneck $b(t)$ stays at a certain target value $T$. Denote the error signal by  $e(t) = T-b(t)$. The sending rate is adapted according to one of the following three controllers:
\begin{enumerate}
\item $x(t)=K_p  e(t)$

\item $x(t)= K_p  e(t)+ K_i \int_0^t e(t)dt$

\item $x(t)= K_p  e(t)+ K_i \int_0^t e(t)dt+ K_d  \frac{d}{dt} e(t)$  
\end{enumerate}
where  $K_p, K_i, K_d$   are constant parameters of the rate controllers.

\begin{itemize}
\item[(a)] Assume $c(t)$ is the capacity available to the source at time $t$, write down the differential equation for $b(t)$.

\item[(b)] By transforming the system to the Laplace domain, determine the conditions under which the system is stable for each type of controller, i.e. does $b(t)$ converge to a given $T$ for certain values of $K_p, K_i, K_d$. Do this by examining the roots (poles) of the characteristic equation of the system's transfer function. Draw the block diagram of the system that shows the relationships between the system variables in the (Laplace) s-domain for each type of controller.

\item[(c)] Again by examining the roots (poles) of the characteristic equation of the system's transfer function and using the Final Value Theorem, compare the transient and steady-state performance under each type of controller.

\item[(d)] Support your answers above by numerically solving the system's equations over time for each type of controller. Assume a small time step $\Delta$, say $\Delta=1$, and solve the discretized version of the equations at these time steps -- you can then approximate the differentiation $\frac{d}{dt} b(t)$ by $b(t)-b(t-1)$.
\end{itemize}

\item Consider an adaptation of a transmission window $w$ whose goal is to reach a target window $T$ as follows:
\begin{eqnarray*}
w_{(k+1)} =  w_k+ \alpha(T-w_k)
\end{eqnarray*}
where  $0 < \alpha < 1$. 
\begin{itemize}
\item[(a)] Derive the necessary condition (if any) for convergence to the target window size.
\item[(b)] Use the Lyapunov method to show whether the system converges regardless of the initial window size.
\end{itemize}

\item Given a system with the following adaptation rules: $x(t)$, representing a sending rate, is adapted using an AIMD policy, whereas $p(t)$, representing the price, is adapted in proportion to how far $x(t)$ is from a target capacity $c$:

\begin{eqnarray*}
\frac{d x(t)}{dt} = 1- x(t) p(x(t)) 
\end{eqnarray*}

\begin{eqnarray*}
p(x(t))= \alpha (x(t)-c)
\end{eqnarray*}

\begin{itemize}
\item[(a)] Why is that system non-linear?

\item[(b)] Linearize the system around a certain operating point $x_0$.

\item[(c)] By transforming the linearized system to the Laplace domain, obtain the condition on $x_0$ under which the linearized system is stable.
\end{itemize}

\item

This question is based on the paper {\em The Revised ARPANET
Routing Metric, by Zinky and Khanna} \cite{Khanna:1989}.
A unified way to model adaptive resource
management---whether it is TCP adaptive to RED or routing adaptive
to changing link costs or other examples---is through two
functions: a feedback (pricing) function such as that of RED or
link utilization metric, and an adaptation function such as that
of TCP or utilization-based routing. Consider a resource that
generates prices $p$ and users that adapt their load $\lambda$
based on the currently reported $p$. Assume the following pricing
and load adaptation functions: \begin{center} $\lambda = 1 - p$
\end{center} \[ p = \left\{ \begin{array}{ll}
       0.1 & \mbox{if $ 0 \leq \lambda < 0.4$} \\
       0.2  & \mbox{if $ 0.4 \leq \lambda \leq 0.8$} \\
       1.0  & \mbox{if $ 0.8 < \lambda \leq 1.0$}
       \end{array}
\right. \]

Show the two functions on the ($\lambda, p$) plane. Trace one
trajectory showing convergence to a fixed point, and another
trajectory showing oscillations. {\em Hint:} consider initial
values of $\lambda$ = 0.2 and 0.6.

\end{enumerate}

\section{Solutions to Exercises}
\label{sec:solutions}

\begin{enumerate}

\item

\begin{itemize}
\item[(a)] Refer to Figure \ref{fig:max-min-alloc}. Using max-min fairness the original
flow assignment is: (F1=50, F2=50, F3=50, F4=100). Assume F2 is
removed. The new assignment is: (F1=75, F2=0, F3=75, F4=75). F1
and F3 will share the extra capacity, while F4 will decrease its
rate. This is only possible because F4's original assigned rate
was not less than or equal to F3's rate, before F2 was removed,
thus allowing F3 to increase its rate.

\item[(b)] Refer to Figure \ref{fig:max-min-alloc}. Using max-min fairness the original
flow assignment is: (F1=50, F2=50, F3=50, F4=100). If the capacity
of link 1 is increased to 225 then the new allocated rates will
be: (F1=75, F2=75, F3=75, F4=75). The extra capacity on link 1
will be shared by F1, F2 and F3. Again, this is only possible
because F4's original assigned rate is not less than or equal to
F3's rate, before the capacity of link 1 was increased, thus
allowing F3 to increase its rate. 
\end{itemize}

\item

Note that earlier in these notes, we did not explicitly cover the technique of ``slack",
which we use here to denote residual link capacity when solving the dual Lagrangian problem.
Intuitively, if the link is fully utilized, i.e. it has zero slack,
then its price (Lagrangian multiplier) is non-zero.
On the other hand, a non-bottleneck link will have a non-zero slack
and so a price of zero.

\begin{itemize}
\item[(a)] Let $z_1^2$ and $z_2^2$ denote the (non-negative) slack on links 1 and
2, respectively. Let $x_1$ and $x_2$ denote the rates assigned to
flows 1 and 2, respectively. Let $f(x_1, x_2)$ denote the function
to be maximized.

\vspace{-3mm}

\begin{eqnarray*}
f(x_1, x_2) = \lg ~x_1 + \lg ~x_2
\end{eqnarray*}

\noindent Constraints on links 1 and 2, respectively, are as
follows:

\vspace{-5mm}

\begin{eqnarray*}
x_1 + x_2 + z_1^2 - c_1 = 0
\end{eqnarray*}

\vspace{-5mm}

\begin{eqnarray*}
x_2 + z_2^2 - c_2 = 0
\end{eqnarray*}

\noindent Replacing $c_1$ and $c_2$ with their values we get:

\vspace{-5mm}

\begin{eqnarray*}
x_1 + x_2 + z_1^2 - 6 = 0
\end{eqnarray*}

\vspace{-5mm}

\begin{eqnarray*}
x_2 + z_2^2 - 6 = 0
\end{eqnarray*}

\noindent The Lagrangian function to be differentiated is as
follows:

\vspace{-3mm}

\begin{eqnarray*}
F(x_1, x_2) = \lg ~x_1 + \lg ~x_2 - \lambda (x_1 + x_2 + z_1^2 -
6) - \mu(x_2 + z_2^2-6)
\end{eqnarray*}

\noindent Taking the partial derivative with respect to each
variable we get the following equations:

\begin{eqnarray*}
\pd{F}{x_1} = {\frac {1}{x_1}} - \lambda = 0
\end{eqnarray*}

\begin{eqnarray*}
\pd{F}{x_2} = {\frac {1}{x_2}} - \lambda - \mu = 0
\end{eqnarray*}

\begin{eqnarray*}
\pd{F}{\lambda} = -(x_1 + x_2 + z_1^2 - 6) = 0
\end{eqnarray*}

\begin{eqnarray*}
\pd{F}{\mu} = -(x_2 + z_2^2 - 6) = 0
\end{eqnarray*}

\begin{eqnarray*}
\pd{F}{z_1} = -2z_1\lambda = 0
\end{eqnarray*}

\begin{eqnarray*}
\pd{F}{z_2} = -2z_2\mu = 0
\end{eqnarray*}

\noindent
Since F1 can utilize any left over capacity (slack) on
link 1, $z_1 = 0$, which implies that $\lambda \ne 0$.
Conversely, since F2 is limited by link 1, $z_2 \ne 0$, which
implies that $\mu = 0$. \bigbreak

\noindent Replacing $z_1 = 0$ and $\mu = 0$ into the derived
equations, we get:

\vspace{-3mm}

\begin{eqnarray*}
{\frac{1}{x_1}} - \lambda = 0
\end{eqnarray*}

\vspace{-5mm}

\begin{eqnarray*}
{\frac{1}{x_2}} - \lambda - 0 = 0
\end{eqnarray*}

\vspace{-5mm}

\begin{eqnarray*}
x_1 + x_2 + 0 -6 = 0
\end{eqnarray*}

\noindent Solving for $x_1$ we get $x_1 =
{\frac{1}{\lambda}}$

\noindent Solving for $x_2$ we get $x_2 =
{\frac{1}{\lambda}}$. Hence, $x_1 = x_2$.

\noindent Replacing $x_2$ by $x_1$ in the capacity equation we get:

\vspace{-5mm}

\begin{eqnarray*}
x_1 + x_1 = 6
\end{eqnarray*}

\noindent Hence, $x_1 = x_2 = 3.$ \bigbreak

%%%%%%%%%%%%%%%%

\item[(b)] Let $z_1^2$ and $z_2^2$ denote the slack on links 1 and
2, respectively. Let $x_1$, $x_2$ and $x_3$ denote the rates
assigned to flows 1, 2 and 3, respectively. Let $f(x_1, x_2, x_3)$
denote the function to be maximized.

\vspace{-3mm}

\begin{eqnarray*}
f(x_1, x_2, x_3) = \lg ~x_1 + \lg ~x_2 + \lg ~x_3
\end{eqnarray*}

\noindent Constraints on links 1 and 2, respectively, are as
follows:

\vspace{-5mm}

\begin{eqnarray*}
x_1 + x_2 + z_1^2 - c_1 = 0
\end{eqnarray*}

\vspace{-5mm}

\begin{eqnarray*}
x_3 + x_2 + z_2^2 - c_2 = 0
\end{eqnarray*}

\noindent Replacing $c_1$ and $c_2$ with their values we get:

\vspace{-5mm}

\begin{eqnarray*}
x_1 + x_2 + z_1^2 - 6 = 0
\end{eqnarray*}

\vspace{-5mm}

\begin{eqnarray*}
x_3 + x_2 + z_2^2 - 6 = 0
\end{eqnarray*}

\noindent The Lagrangian function to be differentiated is as
follows:

\vspace{-5mm}

\begin{eqnarray*}
F(x_1, x_2, x_3) = \lg ~x_1 + \lg ~x_2 + \lg ~x_3 - \lambda (x_1 +
x_2 + z_1^2 - 6) - \mu(x_3 + x_2 + z_2^2 - 6)
\end{eqnarray*}

\noindent Taking the partial derivative with respect to each
variable we get the following equations:

\begin{eqnarray*}
\pd{F}{x_1} = {\frac {1}{x_1}} - \lambda = 0
\end{eqnarray*}

\begin{eqnarray*}
\pd{F}{x_2} = {\frac {1}{x_2}} - \lambda - \mu = 0
\end{eqnarray*}

\begin{eqnarray*}
\pd{F}{x_3} = {\frac {1}{x_3}} - \mu = 0
\end{eqnarray*}

\begin{eqnarray*}
\pd{F}{\lambda} = -(x_1 + x_2 + z_1^2 - 6) = 0
\end{eqnarray*}

\begin{eqnarray*}
\pd{F}{\mu} = -(x_3 + x_2 + z_2^2 - 6) = 0
\end{eqnarray*}

\begin{eqnarray*}
\pd{F}{z_1} = -2z_1\lambda = 0
\end{eqnarray*}

\begin{eqnarray*}
\pd{F}{z_2} = -2z_2\mu = 0
\end{eqnarray*}

\noindent From the equations above we can deduce the following:

\vspace{-3mm}

\begin{eqnarray*}
x_1 = {\frac{1}{\lambda}}
\end{eqnarray*}

\vspace{-3mm}

\begin{eqnarray*}
x_2 = {\frac{1}{\lambda + \mu}}
\end{eqnarray*}

\vspace{-3mm}

\begin{eqnarray*}
x_3 = {\frac{1}{\mu}}
\end{eqnarray*}

\noindent Since F1 and F3 can utilize any left over capacity
(slack) on links 1 and 2, $z_1 = z_2 = 0$. This implies that
$\lambda \ne 0$ and $\mu \ne 0$. \bigbreak

\noindent Replacing $z_1 = 0$ and $z_2 = 0$ into the derived
equations we get:

\vspace{-3mm}

\begin{eqnarray*}
x_1 + x_2 = 6
\end{eqnarray*}

\vspace{-3mm}

\begin{eqnarray*}
x_2 + x_3 = 6
\end{eqnarray*}

\noindent Replacing $x_1$ by ${\frac{1}{\lambda}}$, $x_2$ by
${\frac{1}{\lambda + \mu}}$ and $x_3$ by ${\frac{1}{\mu}}$ into
the equations above, we get:

\vspace{-3mm}

\begin{eqnarray*}
{\frac{1}{\lambda}} + {\frac{1}{\lambda + \mu}}  = 6
\end{eqnarray*}

\vspace{-3mm}

\begin{eqnarray*}
{\frac{1}{\lambda + \mu}} + {\frac{1}{\mu}}  = 6
\end{eqnarray*}

\noindent Multiplying the first equation by (-1) and adding it to
second equation, results in the following:

\vspace{-3mm}

\begin{eqnarray*}
{\frac{1}{\mu}} - {\frac{1}{\lambda}}  = 0
\end{eqnarray*}

\noindent Hence, $\lambda = \mu$ \bigbreak

\noindent Thus, $x_1 = x_3 = {\frac{1}{\lambda}}$ and $x_2 =
{\frac{1}{2\lambda}} = {\frac{x_1}{2}} = {\frac{x_3}{2}}$

\noindent Finally we get:

\vspace{-3mm}

\begin{eqnarray*}
x_1 + {\frac{x_1}{2}} = 6
\end{eqnarray*}

\noindent Hence, $x_1 = x_3 = 4$ and $x_2 = 2.$

\end{itemize}

%%%%%% answer to exercise 3
\item The objective function we want to maximize is:

$F(x) = \lg\ x_1 + \lg\ x_2 + 2\ \lg\ x_3$

\noindent subject to the capacity constraints:

$x_1 + x_3 \leq 6$ \\
$x_2 + x_3 \leq 6$ \\
$x_1, x_2, x_3 \geq 0$

The Lagrangian (unconstrained) function that we want to maximize is:

$L(x) = \lg\ x_1 + \lg\ x_2 + 2\ \lg\ x_3 - \lambda_1(x_1 + x_3 - 6) - \lambda_2(x_2 + x_3 - 6)$

Note that we do not explicitly include the slacks for the link capacities here, since both links should be fully utilized as the one-link flows are not limited by any other link, so the slack values are zero.

Taking the partial derivatives of $L(.)$ we obtain:

$\pd{L}{x_1} = \frac{1}{x_1} - \lambda_1 = 0 \Rightarrow x_1 = \frac{1}{\lambda_1} $

$\pd{L}{x_2} = \frac{1}{x_2} - \lambda_2 = 0 \Rightarrow x_2 = \frac{1}{\lambda_2} $

$\pd{L}{x_3} = \frac{2}{x_3} - (\lambda_1 + \lambda_2) = 0 \Rightarrow x_3 = \frac{2}{\lambda_1 + \lambda_2} $

$\pd{L}{\lambda_1} = 0  \Rightarrow x_1 + x_3 - 6 = 0 \Rightarrow x_1 + x_3 = 6 $

$\pd{L}{\lambda_2} = 0  \Rightarrow x_2 + x_3 - 6 = 0 \Rightarrow x_2 + x_3 = 6 $

The last two equations yield $x_1 = x_2  \Rightarrow \lambda_1 =  \lambda_2 = \lambda$

From the capacity equation, we have:

$x_1 + x_3 = \frac{1}{\lambda} + \frac{2}{2\lambda} = 6 \Rightarrow \lambda = \frac{1}{3}$, thus

$x_1 = x_2 = 3$, and $x_3 = 6 - 3 = 3. $

%%%%%%%

\item \noindent We use the following notation: \smallbreak

\noindent In the time domain we have:\\
\noindent $b(t)$, buffer size at time $t$\\
\noindent $x(t)$, sending rate at time $t$\\
\noindent $c(t)$, service rate at time $t$\\
\noindent $T$, target buffer size\\
\noindent $e(t)$, error (difference between current buffer size
and the target) at time $t$ \bigbreak

\noindent In the frequency domain we have:\\
\noindent $B(s)$, current buffer size\\
\noindent $X(s)$, sending rate\\
\noindent $C(s)$, service rate\\
\noindent $T(s)$, target buffer size\\
\noindent $E(s)$, error signal\\
\noindent $D(s)$, controller (based on error signal computes
sending rate) \bigbreak

\noindent We will assume that the target buffer size and the
buffer's service rate are constant values.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Question One Part a
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bigbreak \bigbreak

\begin{itemize}

\item[(a)] \smallbreak

\noindent $e(t) = T - b(t)$\\
\noindent ${\frac{d}{dt}} b(t) = x(t) - c(t)$

%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Question One Part b
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bigbreak \bigbreak

\item[(b)] \smallbreak

\begin{figure*}[htp]
\centerline{
  % Requires \usepackage{graphicx}
  \includegraphics[width=0.7\textwidth]{blockDiagram}
  }
   \caption{System block diagram}
   \label{fig:sol-block}
\end{figure*}

\noindent The system's block diagram is depicted in Figure~\ref{fig:sol-block}.
Using the block diagram, we can formulate the following equation:
\bigbreak

\noindent $~( ~[ ~(T(s) - B(s)) ~D(s)~] - C(s)~) ~({\frac{1}{s}}) = B(s)$\\

\noindent $~( ~[T(s)D(s) - B(s)D(s)] - C(s)~) ~({\frac{1}{s}}) = B(s)$\\

\noindent $~(T(s)D(s) - B(s)D(s) - C(s)) ~({\frac{1}{s}}) = B(s)$\\

\noindent $[{\frac{T(s) D(s)}{s}} - {\frac{B(s) D(s))}{s}} -
{\frac{C(s)}{s}}] = B(s)$\\

\noindent $B(s) + {\frac{B(s) D(s)}{s}} = {\frac{T(s) D(s)}{s}} -
{\frac{C(s)}{s}}$\\

\noindent $B(s)[1 + {\frac{D(s)}{s}}] = {\frac{T(s) D(s)}{s}} -
{\frac{C(s)}{s}}$\\

\noindent $B(s)[{\frac{s + D(s)}{s}}] = {\frac{T(s) D(s)}{s}} -
{\frac{C(s)}{s}}$\\

\noindent $B(s) = {\frac{T(s) D(s)}{s + D(s)}} - {\frac{C(s)}{s +
D(s)}}$

%%%%%%%%%%%%%%%%%%%%%%%%%
%Question One Part b --- P-Controller
%%%%%%%%%%%%%%%%%%
\bigbreak \bigbreak

\noindent {\bf P-Controller}: \smallbreak

\noindent $x(t) = K_p e(t)$ \bigbreak

\noindent By taking the Laplace transform we get: \smallbreak

\noindent $X(s) = K_p E(s) \Rightarrow D(s) = {\frac{X(s)}{E(s)}}
= K_p$ \bigbreak

\noindent Replacing $D(s)$ into the equation for $B(s)$ we get:
\smallbreak

\noindent $B(s) = {\frac{T(s) K_p}{s + K_p}} - {\frac{C(s)}{s +
K_p}}$ \bigbreak

\noindent Roots of the system's characteristic equation are:
\smallbreak

\noindent $s + K_p = 0 \Rightarrow s = -K_p < 0$\\
\noindent Thus, system is stable for all values of $K_p > 0$

%%%%%%%%%%%%%%%%%%%%%%%%
%Question One Part b --- PI-Controller
%%%%%%%%%%%%%%%%%%%%%%%%
\bigbreak \bigbreak

\noindent {\bf PI-Controller}: \smallbreak

\noindent $x(t) = K_p e(t) + K_i \int^t_0 e(t) dt$ \bigbreak

\noindent By taking the Laplace transform we get: \smallbreak

\noindent $X(s) = K_p E(s) + {\frac{K_i}{s}} E(s) \Rightarrow D(s)
= {\frac{X(s)}{E(s)}} = K_p + {\frac{K_i}{s}}$ \bigbreak

\noindent Replacing $D(s)$ into the equation for $B(s)$ we get:
\smallbreak

\noindent $B(s) = {\frac{(K_p + {\frac{K_i}{s}}) T(s)}{s + (K_p +
{\frac{K_i}{s}})}} - {\frac{C(s)}{s + (K_p + {\frac{K_i}{s}})}}$
\bigbreak

\noindent Roots of the system's characteristic equation are:
\smallbreak

\noindent $s + K_p + {\frac{K_i}{s}} = 0$\\
\noindent $s^2 + K_ps + K_i = 0$\\
\noindent $a = 1$, $b = K_p$, $c = K_i$\\
\noindent $\Delta = b^2 - 4ac = K_p^2 - 4(1)(K_i) = K_p^2 - 4 K_i$
\bigbreak

\noindent $s = {\frac{-b \pm \sqrt{\Delta}}{2a}}$\\

\noindent $s = {\frac{-K_p \pm \sqrt{K_p^2 - 4K_i}}{2}}$\\

\noindent $s = -{\frac{1}{2}}K_p \pm {\frac{1}{2}} \sqrt{K_p^2 -
4K_i}$ \bigbreak

\noindent System will be stable with two real roots (i.e., overdamped) if the
following conditions are true:\\
\noindent $K_p > 0$\\
\noindent $K_p^2 - 4K_i > 0 \Rightarrow K_p^2 > 4K_i$\\
\noindent $\sqrt{K_p^2 - 4K_i} < K_p \Rightarrow K_p^2 - 4K_i <
K_p^2 \Rightarrow K_i > 0$ \bigbreak

\noindent System will be stable with two complex roots (i.e., underdamped) if the
following conditions are true:\\
\noindent $K_p > 0$\\
\noindent $K_p^2 - 4K_i < 0 \Rightarrow K_p^2 < 4K_i$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Question One Part b --- PID-Controller
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bigbreak \bigbreak

\noindent {\bf PID-Controller}: \smallbreak

\noindent $x(t) = K_p e(t) + K_i \int^t_0 e(t) dt + K_d
{\frac{d}{dt}} e(t)$ \bigbreak

\noindent By taking the Laplace transform we get: \smallbreak

\noindent $X(s) = K_p E(s) + {\frac{K_i}{s}} E(s) + K_d s E(s)
\Rightarrow D(s) = {\frac{X(s)}{E(s)}} = K_p + {\frac{K_i}{s}} +
K_d s$ \bigbreak

\noindent Replacing $D(s)$ into the equation for $B(s)$ we get:
\smallbreak

\noindent $B(s) = {\frac{(K_p + {\frac{K_i}{s}} + K_d s) T(s)}{s +
(K_p + {\frac{K_i}{s}} + K_d s)}} - {\frac{C(s)}{s + (K_p +
{\frac{K_i}{s}} + K_d s)}}$ \bigbreak

\noindent Roots of the system's characteristic equation are:
\smallbreak

\noindent $s + K_p + {\frac{K_i}{s}} + K_d s = 0$\\
\noindent $s^2 + K_ps + K_i + K_d s^2 = 0$\\
\noindent $(1+K_d)s^2 + K_p s + K_i = 0$\\
\noindent $a = 1+K_d$, $b = K_p$, $c = K_i$\\
\noindent $\Delta = b^2 - 4ac = K_p^2 - 4(1+K_d)(K_i) = K_p^2 - 4
K_i(1+K_d)$ \bigbreak

\noindent $s = {\frac{-b \pm \sqrt{\Delta}}{2a}}$\\

\noindent $s = {\frac{-K_p \pm \sqrt{K_p^2 - 4K_i(1+K_d)}}{2(1+K_d)}}$\\

\noindent $s = -{\frac{K_p}{2(1+K_d)}} \pm {\frac{1}{2(1+K_d)}}
\sqrt{K_p^2 - 4K_i(1+K_d)}$ \bigbreak

\noindent System will be stable with two real roots if the
following conditions are true:\\
\noindent ${\frac{K_p}{1+K_d}} > 0$\\
\noindent $K_p^2 - 4K_i(1+K_d) > 0 \Rightarrow K_p^2 > 4K_i(1+K_d)$\\
\noindent $\sqrt{K_p^2 - 4K_i(1+K_d)} < K_p \Rightarrow K_p^2 -
4K_i(1+K_d) < K_p^2 \Rightarrow 4K_i(1+K_d) > 0$ \bigbreak

\noindent System will be stable with two complex roots if the
following conditions are true:\\
\noindent ${\frac{K_p}{1+K_d}} > 0$\\
\noindent $K_p^2 - 4K_i(1+K_d) < 0 \Rightarrow K_p^2 <
4K_i(1+K_d)$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Question One Part c
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bigbreak \bigbreak

\item[(c)] \smallbreak

\noindent Final Value Theorem\\
\noindent $\lim_{t \to \infty} f(t) = \lim_{s \to 0} sF(s)$
\bigbreak

\noindent Thus, we have:\\
\noindent $\lim_{t \to \infty} b(t) = \lim_{s \to 0} sB(s)$
\bigbreak

\noindent Note that,\\
\noindent $B(s) = {\frac{T(s) D(s)}{s + D(s)}} - {\frac{C(s)}{s +
D(s)}}$ \bigbreak

\noindent Let the service rate be constant and equal to $C$. We
therefore have, $C(s) = {\frac{C}{s}}$. Similarly, let the target
buffer size be a constant and equal to $T$. We therefore have,
$T(s) = {\frac{T}{s}}$. Replacing these two values into the
equation for $B(s)$, we get:\\
\noindent $B(s) = {\frac{{\frac{T}{s}} D(s)}{s + D(s)}} -
{\frac{\frac{C}{s}}{s + D(s)}}$ \bigbreak

\noindent Multiplying by $s$ to get $sB(s)$, we get:\\
\noindent $sB(s) = {\frac{T D(s)}{s + D(s)}} - {\frac{C}{s +
D(s)}}$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Question One Part c --- P-Controller
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bigbreak \bigbreak

\noindent {\bf P-Controller}:\\
\noindent $D(s) = K_p$\\

\noindent $sB(s) = {\frac{T K_p}{s + K_p}} - {\frac{C}{s +
K_p}}$\\

\noindent $\lim_{s \to 0} sB(s) = \lim_{s \to 0} [{\frac{T K_p}{s
+ K_p}} - {\frac{C}{s + K_p}}] = T - {\frac{C}{K_p}}$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Question One Part c --- PI-Controller
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bigbreak \bigbreak

\noindent {\bf PI-Controller}:\\
\noindent $D(s) = K_p + {\frac{K_i}{s}}$\\

\noindent $sB(s) = {\frac{T (K_p + {\frac{K_i}{s}})}{s + (K_p +
{\frac{K_i}{s}})}} - {\frac{C}{s + (K_p + {\frac{K_i}{s}})}} =
[{\frac{T (K_p s + K_i)}{s^2 + K_p s + K_i}} - {\frac{Cs}{s^2 + K_p s + K_i}}]$\\

\noindent $\lim_{s \to 0} sB(s) = \lim_{s \to 0} [{\frac{T (K_p s
+ K_i)}{s^2 + K_p s + K_i}} - {\frac{Cs}{s^2 + K_p s + K_i}}] = T$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Question One Part c --- PID-Controller
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bigbreak \bigbreak

\noindent {\bf PID-Controller}:\\
\noindent $D(s) = K_p + {\frac{K_i}{s}} + K_d s$\\

\noindent $sB(s) = {\frac{T (K_p + {\frac{K_i}{s}} + K_d s)}{s +
(K_p + {\frac{K_i}{s}} + K_d s)}} - {\frac{C}{s + (K_p +
{\frac{K_i}{s}} + K_d s)}} = {\frac{T (K_p s + K_i + K_d s^2)}{s^2
+ K_p s + K_i + K_d s^2}} - {\frac{Cs}{s^2 +
K_p s + K_i + K_d s^2}}$\\

\noindent $sB(s) = {\frac{T (K_p s + K_i + K_d s^2)}{(1+K_d)s^2 +
K_p s + K_i}} - {\frac{Cs}{(1+K_d)s^2 +
K_p s + K_i}}$\\

\noindent $\lim_{s \to 0} sB(s) = \lim_{s \to 0} [{\frac{T (K_p s
+ K_i + K_d s^2)}{(1+K_d)s^2 + K_p s + K_i}} -
{\frac{Cs}{(1+K_d)s^2 + K_p s + K_i}}] = T$

\item[(d)] \smallbreak

We leave it to the reader to show the plots for $b(t)$.

\end{itemize}

\item

\begin{itemize}
\item[(a)] \smallbreak

\noindent At steady state, $w_k \rightarrow w^*$. This implies
that $w^* = w^* + \alpha(T - w^*)$. Thus, $w^* = T$. This means
that there is no necessary condition, since the system will always
converge to the target window size. \bigbreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Question Two Part b
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item[(b)] \smallbreak

\noindent We have $w_{k+1} = w_k + \alpha(T - w_k)$.
Re-writing this equation we get, $w_{k+1} = w_k(1-\alpha) + \alpha T$. 
\noindent 
Assume $w_0 < T$, let the
Lyapunov function be $L(w) = T - w > 0$.

%\begin{center}
\bigbreak
\noindent $L(w_{k+1}) = T - w_{k+1}$\\
\noindent $L(w_{k+1}) = T - [w_k + \alpha(T - w_k)]$\\
\noindent $L(w_{k+1}) = T - w_k - \alpha T + \alpha w_k$\\
\noindent $L(w_{k+1}) = (1-\alpha)T - (1-\alpha)w_k$\\
\noindent $L(w_{k+1}) = (1-\alpha)(T-w_k)$\\
\noindent $L(w_{k+1}) = (1-\alpha)L(w_k) < L(w_k)$ \bigbreak
%\end{center}

\noindent This is true since $(1-\alpha) < 0$. In other words,
$L(w)$ is indeed a decreasing function. \bigbreak

\noindent If instead we assume that $w_0 > T$, then let us define the Lyapunov
function as $L(w) = w - T > 0$. Similarly we have,

%\begin{center}
\bigbreak
\noindent $L(w_{k+1}) = w_{k+1} - T$\\
\noindent $L(w_{k+1}) = [w_k(1-\alpha) + \alpha T] - T$\\
\noindent $L(w_{k+1}) = w_k(1-\alpha) - (T - \alpha T)$\\
\noindent $L(w_{k+1}) = w_k(1-\alpha) - T(1-\alpha)$\\
\noindent $L(w_{k+1}) = (1-\alpha)(w_k-T)$\\
\noindent $L(w_{k+1}) = (1-\alpha)L(w_k) < L(w_k)$ \bigbreak
%\end{center}

\noindent This is true since $(1-\alpha) < 0$. In other words,
$L(w)$ is indeed a decreasing function. \bigbreak
\end{itemize}

\item

\begin{itemize}
\item[(a)] \smallbreak

\noindent ${\frac{d}{dt}}x(t) = 1 - x(t)p(x(t))$\\
\noindent $p(x(t)) = \alpha(x(t) - c)$\\
\noindent ${\frac{d}{dt}}x(t) = 1 - x(t)[\alpha (x(t) - c)]$\\
\noindent ${\frac{d}{dt}}x(t) = 1 - x(t)[\alpha x(t) - \alpha c)]$\\
\noindent ${\frac{d}{dt}}x(t) = 1 -\alpha x^2(t) + \alpha cx(t)$
\bigbreak

\noindent The equation above is non-linear because it has an
$x^2(t)$ term. \bigbreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Question Three Part b
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item[(b)] \smallbreak

\noindent Let $f(x) = 1 -\alpha x^2(t) + \alpha cx(t)$\\
\noindent $f'(x_0) = -2 \alpha x_0 + \alpha c$  \\ 
(Note that one may choose to linearize the $x^2$ term only. Here we choose to linearize the whole $f(x)$.) \\
\noindent $\Delta f = f'(x_0) \Delta x$\\
\noindent ${\frac{d}{dt}} \Delta x = f'(x_0) \Delta x$\\
\noindent Let $y(t) = \Delta x$\\
\noindent ${\frac{d}{dt}} y(t) = f'(x_0) y(t)$\\
\noindent ${\frac{d}{dt}} y(t) = [-2 \alpha x_0 + \alpha c]y(t)$\\
\noindent Let $\beta = [-2 \alpha x_0 + \alpha c]$\\
\noindent ${\frac{d}{dt}} y(t) = \beta y(t)$ \bigbreak

\item[(c)] \smallbreak

\noindent Transforming this equation to the s-domain, we get:
\smallbreak

\noindent $sY(s) -y(0) = \beta Y(s)$\\
\noindent $sY(s) - \beta Y(s) = y(0)$\\
\noindent $Y(s)[s - \beta] = y(0)$\\
\noindent $Y(s) = {\frac{y(0)}{s - \beta}}$ \bigbreak

\noindent The root of the characteristic equation $s - \beta$ is
$s = \beta$. For the system to be stable, $\beta$ must be less
than zero. Assuming $\alpha > 0$, we need the following to be
true: \bigbreak

\noindent $\beta < 0$\\
\noindent $-2 \alpha x_0 + \alpha c < 0$\\
\noindent $-2 \alpha x_0 < -\alpha c$\\
\noindent $2 \alpha x_0 > \alpha c$\\
\noindent $x_0 > {\frac{c}{2}}$ \bigbreak

\end{itemize}

\item We leave it to the reader to produce the plot. Notice that the fixed point is $(\lambda=0.8, p=0.2)$  where the pricing curve and the load curve intersect. Also the system diverges for starting $\lambda=0.2$, whereas it converges for $\lambda=0.6.$ Specifically, in the latter case, $\lambda=0.6$ yields $p=0.2$, which in turn yields $\lambda=1-p=0.8$ and the system stabilizes at the fixed point.

\end{enumerate}

\bibliography{dyna}
\bibliographystyle{plain}

\end{document}

### 2013-07-22 Compiled/chapter3/20Q.tex
all files compile correctly

\section{20 Open Questions and Future Directions}\label{sec:20Q}

Current trends and future directions in smart pricing practices aim to make proposed pricing plans economically viable. For instance, substantial research has been done on day-ahead pricing, including the development of carefully designed user interfaces to display price and usage data as shown in Figure \ref{fig:tube_gui}. Incorporating human factors into the engineering and design phase along with economic models can provide a holistic approach in solving the challenges of network congestion.
%More generally, client-side architectures may ease the user burden of dynamic pricing plans; however, the design of such components remains an open problem, and this conjecture has not yet been thoroughly tested on real users.

In addition to the pricing plans proposed above, new pricing plans have recently been proposed that have been rarely studied in the academic literature. Some promising directions include the following:

\begin{figure}
\centering
\begin{subfigure}[b]{0.23\textwidth}
	\includegraphics[width = \textwidth]{Figures/Main.png}
	\caption{Home screen.}
	%\vspace{0.05in}
	\label{fig:datawiz_main}
	\end{subfigure}
	\begin{subfigure}[b]{0.23\textwidth}
	\includegraphics[width = \textwidth]{Figures/Apps_Mobile.png}
	\caption{Cellular usage per app.}
	%\vspace{0.05in}
	\label{fig:datawiz_main_graph}
	\end{subfigure}
	\begin{subfigure}[b]{0.23\textwidth}
	\includegraphics[width = \textwidth]{Figures/Maps.png}
	\caption{Usage locations.}
	%\vspace{0.05in}
	\label{fig:datawiz_predict}
	\end{subfigure}
	\begin{subfigure}[b]{0.23\textwidth}
	\includegraphics[width = \textwidth]{Figures/Members.png}
	\caption{Shared usage.}
	%\vspace{0.05in}
	\label{fig:datawiz_map2}
	\end{subfigure}

	\vspace{0.3in}

	\begin{subfigure}[b]{0.23\textwidth}
	\includegraphics[width = \textwidth]{Figures/Usage.png}
	\caption{Individual usage.}
	%\vspace{-0.05in}
	\label{fig:datawiz_indv}
	\end{subfigure}
	\begin{subfigure}[b]{0.23\textwidth}
	\includegraphics[width = \textwidth]{Figures/Line_Graph.png}
	\caption{Usage over time.}
	%\vspace{-0.05in}
	\label{fig:datawiz_time}
	\end{subfigure}
	\begin{subfigure}[b]{0.23\textwidth}
	\includegraphics[width = \textwidth]{Figures/Pie_Chart.png}
	\caption{Usage distribution.}
	%\vspace{-0.05in}
	\label{fig:datawiz_pie}
	\end{subfigure}
	\begin{subfigure}[b]{0.23\textwidth}
	\includegraphics[width = \textwidth]{Figures/Devices.png}
	\caption{Usage of devices.}
	%\vspace{-0.05in}
	\label{fig:datawiz_device}
	\end{subfigure}
\caption{Screenshots of the DataWiz app with shared data plans. A user can add other users or devices to her shared data plan and can keep track of the shared data usage as well as individual usage. When a user is requested to be a part of the shared data plan, a push notification is sent to the user and the user can either accept or reject this invitation.}
\label{fig:datawiz}
\end{figure}

\begin{description}
\item[Shared data plans:] AT\&T and Verizon in the United States recently introduced shared data plans, in which several devices share a common data cap. While some studies of shared data plans have been made \cite{WITS}, the effects of such plans on user behavior, and how such a data cap can be shared fairly and efficiently among users, remain to be studied in detail.

One practical challenge for users on shared data plans is keeping track of their usage across multiple devices--while many apps exist to track usage on individual devices; e.g., Onavo Count, My Data Manager, and DataWiz; none track the usage of multiple devices. In fact, tracking usage on multiple devices requires not only modifications to user interface design but also integration with a common server that can send usage information of other user devices to each device on a shared plan. A login and agreement process is also necessary to ensure that users can see each others' usage only once both users have given permission.

Screenshots of a prototype shared data plan app based on the DataWiz usage monitoring application are shown in Figure \ref{fig:datawiz}. The app lets users view their monthly, weekly, and daily usage as compared to their monthly data cap (Figure \ref{fig:datawiz_main}), as well as per-app and per-location usage (Figures \ref{fig:datawiz_main_graph} and \ref{fig:datawiz_predict}). The small pie chart indicator on the top left corner of the phone allows users to quickly see their monthly usage; the chart is filled according to how much of his or her monthly cap the user has consumed.
%
In addition to these basic features, users can add and remove other users and devices to their data plan; if a user is added to someone else's data plan, the invitation recipient receives a push notification and can accept or reject the invitation. Upon acceptance, both users may view each other's usage over time (Figures \ref{fig:datawiz_indv} and \ref{fig:datawiz_time}) and on a pie chart (Figure \ref{fig:datawiz_pie}). Users may add multiple devices to their accounts; each user's total usage is then the combined usage from all of these devices (Figure \ref{fig:datawiz_device}). A push notification is sent if a user is removed from a shared data plan.

\item[Fair throttling:] Instead of merely charging users more over a certain cap, ISPs may forcibly limit usage by throttling users to a limited bandwidth rate. However, researchers have still not thoroughly examined how these bandwidth limits should be set, how they should vary over time, and what their implications are in terms of fairness across different users.

\item[Heterogeneous networks:] Many ISPs are turning to supplementary networks such as WiFi and femtocells to offload traffic from congested cellular networks. While access to such networks is often free, in the future they may wish to implement more systematic access pricing to influence the adoption of such technologies and distribution of the user population across complementary networks like WiFi and 3G \cite{wifi-3g-infocom,Im}.

\item[Sponsored content:] A major question in pricing is ``whom to charge" for delivering traffic. In a two-sided pricing model (like in the credit card business) of the Internet, the price of connectivity is shared between content providers (CPs) and end users (EUs). ISPs become the middle man (or platform) proving the connectivity between CPs and EUs. This \emph{clearing house} or data traffic exchange market would extend the existing 1-800 model of phone-call services in the USA, which charges the callee rather than the caller. The tradeoffs in the resulting economic benefit between CPs and EUs remains to be quantified. Intuitively, end-users' access prices are subsidized by third party sponsors (e.g., advertisers, content providers etc.) and the ISPs have an additional source of revenue. Perhaps more surprisingly, content providers may also stand to gain from two-sided pricing if subsidizing connectivity to end-users translates into a net revenue gain through a larger amount of consumption. However, the gain to content providers depends on the extent to which content-provider payment translates into end-users' subsidy, and on the demand elasticities of the consumers. The precise gains to the three entities will therefore depend on the respective bargaining powers stemming from their contributions and price sensitivities.

Another special case of sponsored content includes \emph{zero-rating} and \emph{1-800 reverse billing} policies for data traffic. Under zero-rating, an ISP makes certain types of application traffic available to the users for free. This kind of policy, although contentious from a net neutrality viewpoint, is a major step in app-based pricing and has been practiced in some parts of Europe (e.g., Mobistar introduced a `zero-rated' plan for Facebook, Twitter, and Netlog). Understanding the impact of such pricing plans on the network ecosystem and its neutrality are important active research directions in the area of network economics.

% \item[Time-dependent pricing:] Day-ahead pricing for mobile data has been found to be both acceptable to users and effective at reducing the network's peak load in a small pilot trial \cite{ha2012tube}. Yet it remains to be seen whether these results hold for a larger sample of users.
\end{description}

\subsection{Static Pricing}

Usage-based static pricing has traditionally been offered by ISPs around the world, and is in some sense the simplest and least controversial form of SDP. Yet even simple caps on monthly usage require a means to communicate those caps to users and, on the ISP side, accounting infrastructure to keep track of users' remaining quotas. Pricing plans like token bucket pricing or negotiated contracts require even more interaction with end users, leading to questions that include:
\begin{enumerate}
\item
How can users use their quota efficiently and keep track of a monthly usage quota?
\item
If users choose different QoS levels or times to receive better QoS, e.g., in Paris metro or token bucket pricing, how can they do so without much technical knowledge of what ``QoS''  means? How can the ISP's infrastructure keep track of users' choices and offer the appropriate QoS?
\item
Without such technical knowledge, how can users negotiate contracts (e.g., cumulus pricing) with ISPs? How can ISPs enforce these contracts?
\item
If the ISP offered some form of personalized (e.g., app-based) pricing, how would it measure the usage of different applications for each user? Where in the network should such measurements take place (i.e., client devices or the network core)? From a regulatory perspective, does this violate privacy or network neutrality concerns?
\item
How will users share the monthly data quotas imposed by shared data plans among different devices?
\end{enumerate}
%Many ISPs and third party developers, including a team led by the authors, have already released mobile apps that tell users how much of their monthly quota is left \cite{datawiz}. Yet the effects of such applications are still unclear, and little research has been done on the \emph{design} of such apps. Similarly, there has been relatively little work on automating QoS choices or negotiations for users, and the feasibility of app-based pricing or other personalized data plans remains in question.

\subsection{Dynamic Pricing}
While static pricing offers some challenges in communicating between ISPs and end users, dynamic pricing introduces even more complications as the user must be informed of changes in price. Deployment questions unique to dynamic pricing include:
\begin{enumerate}
\setcounter{enumi}{5}
\item
How often should the prices change? Should they change with the network congestion, or should they change only after a fixed time interval (e.g., one hour)?
\item
Should users be told the prices in advance? Will they accept or respond to prices that change in real time?
\item
The answer to the previous question can be more broadly phrased as follows--how can users be appropriately informed of the changing prices (e.g., with an app on their mobile devices)? What kind of design is optimal for such an app? Going further, what mechanisms can be developed to help users adjust their behavior in response to the prices?
\item
In the context of mobile data, network bottlenecks are generally highly location-dependent. Should the prices vary by location as well as by time? How will this affect users who move from one location to another?
\item
How can the prices be computed efficiently? Should this computation be done online or offline? What usage monitoring must take place, and how real time does it need to be?
\item
In addition to efficient usage monitoring, how can the ISP anticipate user reactions to the prices so as to set the ``optimal" prices? How can these change over time? Does the measurement process adequately protect user privacy?
\item
Should dynamic pricing be coupled to QoS? If so, how?
\end{enumerate}
%A recent trial of day-ahead time-dependent pricing for mobile data addressed some of these questions by implementing hourly changes in prices, which are fixed one day in advance \cite{ha2012tube}. The prices were calculated based on aggregated usage and observed changes in aggregate usage for different pries offered, so as to make the computation scalable and protect individual users' privacy. The trial included the implementation of a client-side app that showed users their price and usage history and automatically scheduled some applications to cheaper times of the day so as to keep users' monthly expenses below a maximum budget. Yet this trial involved relatively few participants, and it did not address such questions as location-dependent bottlenecks or QoS coupling.

\subsection{Sponsored Content}

% While they look quite different in practice, both sponsored content and fair throttling naturally accommodate with content providers and government regulators, perhaps more than other forms of SDP. Thus, several unique questions arise for these particular pricing plans:
Sponsored content pricing, in which content providers and advertisers subsidize users' spending on data, has not been widely deployed, partially due to the network neutrality implications of content provider subsidies. As a relatively new type of pricing, many questions remain to be answered:
\begin{enumerate}
\setcounter{enumi}{12}
\item
What is the preferred mode of ``sponsoring'' in sponsored content/access? Should it be based on increasing the user's data cap, monetary discounts, or improved speed (e.g., less throttling)?
\item
Will content providers sponsor content on a per-transaction basis? If so, how should these transactions be metered, and how much should they charge?
\item
How can ISPs measure the cost of each transaction and develop accounting systems to keep track of content providers' sponsorship?
\item
Does the idea of ``sponsored content'' violate network neutrality? Or can it be structured in a net-neutral way, e.g., sponsoring some data usage but not specifying the application?
\end{enumerate}

\subsection{Fair Throttling and Heterogeneous Networks}

Other solutions to network congestion that do not explicitly use SDP include fair throttling and deployment of heterogeneous networks to offload traffic. Fair throttling has not been widely deployed in practice--while many ISPs do throttle users who exceed a certain usage cap, such measures are fairly crude and do not take into account users' full profiles. More sophisticated throttling, e.g., Comcast's throttling of Netflix traffic in 2007, has been controversial. In contrast, many ISPs have begun offering WiFi hotspots, but it remains unclear how effective they are in relieving congestion on mobile networks. Thus, interesting theoretical and implementation questions remain for both these types of pricing, including the following:
\begin{enumerate}
\setcounter{enumi}{16}
\item
What criteria should the ISP consider when performing ``fair'' throttling? Does measuring these criteria violate user privacy or network neutrality (e.g., throttling based on the usage of specific application types)?
\item
Should users be directly involved in prioritizing different types of traffic? How can their preferences be incorporated into the throttling algorithm without the act of declaring such preferences becoming onerous to the user?
\item
How much traffic can be offloaded to other heterogeneous networks (e.g., 4G traffic to WiFi)? How cost effective is deploying such networks as a solution to network congestion? How can ISPs estimate the monetary and spectral benefits achieved through such traffic offloading or demand shifting?
\item
If ISPs were to charge for bundled access to supplementary networks like WiFi hotspots, how would such pricing plans affect users' adoption and the overall network congestion?
\end{enumerate}

These 20 questions are only some of the key questions that arise in deploying SDP and can help researchers identify interesting topics for exploration. In the coming years, as the Internet evolves further, answering these questions and others that emerge will help determine how we access (and pay for) the Internet in a highly connected, data-driven world.

### 2013-07-22 Compiled/chapter3/Appendix.tex
all files compile correctly

\appendix

\section{Data Plans around the World}\label{Dataplans}

In most countries, the dominant forms of mobile data pricing are still usage-based or tiered. However, the actual cost of mobile data can vary greatly from country to country, depending on such factors as the country's network infrastructure, market competition, and population density. One common distinction is between pre- and postpaid plans.\footnote{Prepaid plans are those in which a consumer pays for his/her data usage beforehand, while for postpaid plans, a consumer is billed for his/her monthly usage at the end of the billing cycle.} Castells et al. \cite{castells2007mobile} found a strong correlation between the availability of prepaid plans for voice calls and adoption rates of mobile telephone subscribers, which is corroborated by Hauge et al. \cite{hauge2009whose}.  In this Appendix, we provide an overview of prepaid and postpaid plan adoption in different parts of the world and its relationship to consumer market demographics.

\begin{figure}
\centering
\includegraphics[width=0.75\textwidth]{Figures/World.pdf}
\caption{Prepaid and postpaid mobile broadband data plans around the world (data from 2011).  Data were taken from operator websites as well as \cite{chinamobile,SAfrica,cia,kenyatotal,egypttotal,singapore,itu,italy,india,docomo,oecd,chinatotal,russiatotal,braziltotal,korea,vietnamtotal,worldbank}.}
\label{fig:world}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.75\textwidth]{Figures/GNI.pdf}
\caption{GNI per capita vs. monthly cost of 1GB of mobile data (data sources as in Figure \ref{fig:world}).}
\label{fig:gni}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.75\textwidth]{Figures/Density_PPP.pdf}
\caption{Population density vs. monthly cost of 1GB of mobile data (data sources as in Figure \ref{fig:world}).}
\label{fig:density}
\end{figure}

Figure \ref{fig:world} shows whether prepaid or postpaid plans are the dominant form of data plans in several countries of the world, as of 2011. To determine this data, we examined the standard mobile subscriptions for the largest mobile operator in each country and plotted the monthly cost (in \$ after accounting for purchasing power parity) of 1GB of data usage.  We note that the data plans in different countries have different broadband caps, which are not reflected in the graph, but may be found in \cite{itu}. The cost of 1GB of data/month has been plotted against the number of mobile broadband subscribers in each country; we observe that there is a large variation in cost. Additionally, we see that the countries of Africa (e.g., Kenya, South Africa), the Middle East (e.g., Turkey, Egypt), India and China tend to prefer prepaid plans, while postpaid plans dominate in Europe and the Americas. There is a weakly positive correlation between the price for mobile data and the number of broadband subscribers in the country.

Figure \ref{fig:gni} more explicitly shows the correlation between prepaid mobile broadband data plans and lower per capita gross national incomes (GNI): we see that countries with a lower GNI per capita tend to have prepaid plans as the dominant subscription mode, with the exception of Australia.  Moreover, two of the three countries offering both pre- and postpaid plans (Italy and Vietnam) have GNI per capita below the wealthiest 9 countries. Wealthier countries tend to have lower prices for mobile data, even accounting for purchasing power parity. In lower-GNI countries, it is possible that only a wealthy subset who can afford higher prices generally purchase mobile data plans; another explanation is that these countries have lower investment costs due to their relative wealth.

Infrastructure development is also related to population density: in denser countries, networks need to cover a smaller geographical area to reach the same number of people, resulting in lower investment costs. In the rural United States, small carriers often cite wireline backhaul over long distances as a major source of wireless network costs \cite{neca}. Figure \ref{fig:density} shows a negative correlation between population density and the price for 1GB of mobile data, consistent with our hypothesis that denser countries have lower infrastructure costs. Prepaid data plans are somewhat more popular in less dense countries, with India a notable exception: 83\% of countries with prepaid plans have lower population density than 36\% of countries with high population density and postpaid data plans. These countries also tended to have lower GNIs, which along with their population density suggest that their network infrastructure is less comprehensive than that of denser, wealthier countries.

% Within the U.S., prepaid data plans have gained popularity due to their lower, shorter-term financial commitment and relative simplicity compared to postpaid usage-based pricing plans with data caps.  Moreover, operators are offering more attractive prepaid devices with data-centric plans that appeal to young, price-sensitive users \cite{pcworld}.

%\section{Coping with Usage-based Pricing}
%
%While usage-based pricing is one of the simplest pricing plans, besides charging flat fees for unlimited usage, a major limitation to usage pricing is that many people do not understand how much data their applications use. Thus, even tiered data plans with monthly caps can be hard for them to manage. Several user apps have been developed to

### 2013-07-22 Compiled/chapter3/Congestion.tex
all files compile correctly

\section{A Review of Smart Data Pricing}\label{sec:congestion}

Smart data pricing encompasses a wide variety of different pricing algorithms and proposals. In this section, we briefly discuss some of these ideas, following the taxonomy given in Figure \ref{fig:taxonomy}. We include a brief overview of related pricing plans in the electricity and transportation industries, which can help yield insights into the feasibility of various forms of SDP for data, as well as ideas for new pricing plans. Other, more thorough reviews may be found in \cite{CourcoubetisWeber,sen2013survey,Songhurst}.

\begin{figure}
\centering
\includegraphics[width = 0.8\textwidth]{Figures/Taxonomy.pdf}
\caption{Examples of broadband pricing plans proposed in the research literature.}
\label{fig:taxonomy}
\end{figure}

A primary goal of SDP is to create the right incentives (or price points) for users to modify their usage behavior so as to help ISPs with better resource allocation and utilization. But creating these incentives requires ISPs to account for users' responses to the prices offered. Of particular relevance is the timescale associated with the pricing mechanism -- do the prices continually change as the network load changes? If so, how frequently and by how much? How to balance the trade-offs between users' reluctance to real time dynamic pricing and the inability of static pricing to exploit the time elasticity of demand of different applications in congested times? How to balance the trade-offs between the users' need for transparency and control over her usage and the need for automation in dynamic pricing scenarios?

\emph{Static pricing} plans are those that change prices on a relatively longer timescale, e.g., months or years: the offered prices do not vary with immediate changes in the network congestion level. The popularity of these plans arises from the certainty they provide to a user's expected monthly bill. For instance, tiered data plans with pre-specified rates are prevalent in the United States, and several European and Asian ISPs offer usage-based pricing in which users are charged in proportion to their usage volume. But such usage-based pricing leaves a timescale mismatch: ISP revenue is based on monthly usage, but peak-hour congestion dominates its cost structure (e.g., network provisioning costs increase with the peak-hour traffic). Another well-known pricing plan is time-of-day (ToD) pricing, in which users are charged higher prices during certain ``peak'' hours of the day. But even with ToD pricing, the hours deemed as ``peak'' are fixed, which results in two challenges. First, traffic peaks arise in different parts of the network at different times, which can be hard to predict in advance and could end up creating two peaks during the day--one during peak periods, for traffic that cannot wait for several hours for lower-price periods, and another peak during discounted ``off-peak'' periods for time-insensitive traffic. Such patterns have been observed in dynamic pricing for voice calls in operational networks \cite{Economist}. We discuss several of these existing static pricing plans and proposals in greater detail in Section \ref{sec:static}.

\emph{Dynamic pricing} takes the ToD idea further in that it does not pre-classify peak and off-peak periods, instead adjusting prices at a finer timescale or by location in response to the network congestion. However, prices that vary depending on the current network load can be sometimes inconvenient for users.  Hence, dynamic pricing variants for SDP, such as automated ``smart market" \cite{MacKie-Mason,Murphy-Murphy}, raffle-based pricing \cite{loiseau2011incentive}, and day-ahead pricing \cite{ha2012tube}, have been proposed to guarantee the prices a day in advance to give users some certainty about the future prices on offer. Each day, new prices are computed for different times (e.g., hours) of the next day, based on predicted congestion levels.  A detailed discussion on these dynamic pricing proposals will be provided in Section \ref{sec:dynamic}.

\subsection{Static Pricing}\label{sec:static}

Due to the fixed nature of their prices, static data plans do not generally allow ISPs to adapt to real-time congestion conditions. In particular, the ISP cannot prevent or alleviate network congestion at peak times by manipulating the prices. On the other hand, static pricing tends to be more acceptable to users, as it offers more certainty and is simpler than dynamically changing prices. Indeed, the most basic form of static pricing, \emph{flat pricing}, is also the most simple for users, though it does not impose any sort of usage incentives \cite{shakkotai}. Some other important examples of static pricing include the following:

\textbf{\emph{Usage-based:}}
In its purest form, usage-based pricing charges users in proportion to the amount of data that they consume, without regard to the type of data (e.g., application) or time of consumption. The principal advantage of such a pricing plan lies in its relative simplicity: it imposes a monetary penalty on heavy (i.e., high-usage) users to reduce congestion \cite{hande,Li}, but also penalizes users even when the network is lightly loaded. Moreover, usage-based pricing requires users to keep close track of their usage in order to determine how much they have spent on data \cite{walrand}.

\textbf{\emph{Tiered:}}
A more common variant of pure usage-based pricing is tiered pricing, in which users pay a fixed amount of money for a monthly \emph{data cap} (e.g., \$30 for 3GB). This fixed fee covers usage up to the cap, after which users may pay another fixed fee to increase the cap by a discrete amount, e.g., \$10 per extra GB. Thus, tiered or capped pricing can be viewed as a discretization of usage-based pricing. Many ISPs have adopted such a pricing plan or another variant in which the data cap is shared across several devices (i.e., a \emph{shared data plan}). Like usage-based pricing, tiered pricing is simple for users to understand and penalizes heavy usage. %; however, it also puts less burden on users to keep track of their usage in order to calculate their monthly bills. % Yet neither usage-based nor tiered pricing prevents users from using the network at the same time, and thus do not significantly help reduce the peak congestion.

\textbf{\emph{Quality of Service (QoS) classes:}}
Some static pricing plans offer multiple traffic classes with different qualities of service (QoS). A simple differentiated pricing plan is \emph{Paris metro pricing} (PMP), which is named after an actual pricing practice on the Paris metro in the 1900s \cite{Odlyzko}. In Paris metro pricing, the ISP separates data traffic into different logical traffic classes and charges different prices for logically separate traffic classes (i.e., each class is identical to the others in their treatment of data packets). Only users willing to pay a higher price will adopt this traffic class, which leads to a better QoS due to fewer users. But one of the key issue of academic debate related to PMP has been its viability, namely,  whether it is a mechanism to increase the profits for service providers, or whether it achieves higher social welfare. As pointed out in \cite{Chiu2010}, the conclusions of this debate depend on how users react to the congestion externality of the underlying system. Other researchers have investigated more direct forms of QoS pricing, in which users can indicate their desired QoS in their packets and are charged a higher per-byte fee for higher QoS \cite{Cocchi1,Marbach,M3I}. %Determining users' optimal QoS decisions and the per-byte fees to charge are non-trivial research questions.

%each of which is identical in its treatment of data packets but charges users differently.  Thus, users willing to pay more will select the more expensive, and hence less congested, logical traffic class.

Another form of QoS pricing is \emph{token pricing}, in which users receive tokens at a fixed rate (e.g., 1 per minute) \cite{Token}. Users can then spend these tokens to send some of their traffic at a premium QoS; users can choose the timing of these premium sessions, e.g., to coincide with their individual priorities and preferences. %One can then examine users' spending decisions and their implications for congestion reduction.

\textbf{\emph{Negotiated contracts:}}
In these types of pricing schemes, users pre-negotiate contracts with the ISP regarding the price of sending traffic over the network. The main research question for such contracts is then characterizing this user-ISP interaction and both parties' optimal decisions. For instance, in \emph{reservation-based pricing}, users specify a monthly budget for data; the ISP can then accept or reject users' connections based on users' remaining budget and the real-time network congestion \cite{Delgrossi,PKF, ParrisF}.

In \emph{expected capacity pricing}, Clark proposed a mechanism in which users similarly negotiate a price in advance based on an ``expected'' quality of service (e.g., file transfer time), so that at congested times the ISP can freely allocate network resources based on whether a given packet lies ``within'' a user's purchased traffic profile \cite{DDClark}. The goal of this pricing scheme is to ``\emph{provide additional explicit mechanisms to allow users to specify different service needs, with the presumption that they will be differentially priced} \cite{DDClark}." Expected capacity pricing allows users to explicitly specify their service expectation (e.g., file transfer time), while accounting for differences in applications' data volume and delay tolerance. The idea is that by entering into profile contracts for expected capacity with the operator, different users should receive different shares of network resources when the network gets congested \cite{Songhurst}.  
One specific proposal to realize this service involved traffic flagging (i.e., each packet is marked as being \emph{in} or \emph{out} of the user's purchased profile, irrespective of network congestion level) by a traffic meter at access points where the user's traffic enters the network. This is followed by congestion management at the switches and routers where packets marked as \emph{out} are preferentially dropped during congested periods, but are treated in an equal best-effort manner at all other times.  The expected capacity is thus not a capacity guarantee from the network to the user, but rather a notion of the capacity that a user expects to be available and a set of mechanisms that allow the user to obtain a different share of the resource at congested times.  
%This pricing can be simply enforced at the router and switches of the network, and allows service providers to have more stable estimates of the future necessary capacity based on the total expected capacity sold, rather than the sum of peak rates of all users' access links.  A dynamic pricing version of the scheme has also been explored \cite{DDClark}. However, in order to implement this pricing scheme, the assignment of price values to expected capacity profiles requires further study.

An ISP offers similar contracts under \emph{cumulus pricing}, but users can re-negotiate the price after passing ``cumulus'' usage points \cite{Hayel}. Cumulus pricing consist of three stages: specification, monitoring, and negotiation. A service provider initially offers a flat-rate contract to the user for a specified period based on the user's estimate of resource requirements. During this time the provider monitors the user's actual usage and provides periodic feedback to the user (by reporting on ``cumulus points'' accumulated from their usage) to indicate whether the user has exceeded the specified resource requirements. Once the cumulative score of a user exceeds a predefined threshold, the contract is renegotiated.

\textbf{\emph{App-based and sponsored content:}}
Different applications consume different amounts of data traffic (e.g., streaming video consumes much more data than retrieving emails). Some researchers have thus proposed app-based pricing, in which users are charged different rates for different apps \cite{Ericsson}. Such pricing plans also include ``zero-rated'' apps, whose traffic is free for the user. A variant of such pricing schemes is ``sponsored content'', in which a third-party (advertiser, content provider, or the ISP itself) ``sponsors'' some part of the traffic in return for accessing specific content or using data at less congested times.

App-based plans have been offered in Europe, largely on a promotional basis. However, app-based pricing presents technical challenges for ISPs-- ISPs need to identify and track how much data each user consumes on specific applications, which may raise privacy concerns. Moreover, some apps open links in separate apps (e.g., links in Flipboard may open a separate Internet browser), creating confusion among users as to the app to which some traffic belongs, and whether this traffic counts towards the sponsored volume or not. Even in academia, sponsored content research is relatively sparse, though a few initial models have been developed \cite{andrewsSDP,hande2009network}.

\textbf{\emph{Time-of-day (ToD):}} ToD pricing charges users different usage-based rates at different times of the day (e.g., peak and off-peak hours) \cite{ParrisF}. The free nighttime minutes offered for voice calls by most US ISPs before 2013 are one simple form of ToD pricing. However, as the peak times and rates are fixed in advance, ToD pricing can end up creating two peaks, one during the ``peak'' period and one in the ``off-peak'' period; indeed, this phenomenon was observed in Africa when MTN Uganda offered discounted prices for voice calls made at night.

Some ISPs offer two-period ToD pricing plans with different charging rates at day and night times.  For example, BSNL in India offers unlimited night time (2-8 am) downloads on monthly data plans of Rs 500 (\$10) and above. Other variations of ToD pricing are offered elsewhere; for instance, the European operator Orange has a ``Dolphin Plan''  for $\pounds$15 (\$23.50 USD) per month that allows unlimited web access during a ``happy hour'' corresponding to users' morning commute (8-9 am), lunch break (12-1 pm), late afternoon break (4-5 pm), or late night (10-11 pm). The underlying idea is to allow consumers to \emph{self-select} themselves into ``time-buckets" with QoE guarantees, with the hope of exploiting the variance in consumers' time-of-day preference to spread out demand more evenly over the day.

%Though the prices do change with time in ToD pricing, we list ToD pricing under static pricing due to the fact that the prices remain the same from day to day.

\subsection{Dynamic Pricing}\label{sec:dynamic}
Dynamic pricing allows prices to be changed in (near) real-time, which unlike static pricing allows an ISP to adjust its prices in response to observed network congestion. However, in doing so the ISP significantly complicates its pricing, making it much harder for users to understand. Thus, implementing and offering dynamic pricing plans requires ISPs to account for human factors that can make real-time changes in price more amenable to users. Some of the proposed dynamic pricing plans are discussed below:

\textbf{\emph{Real-time congestion:}} If ISPs can monitor their network for real-time signs of congestion, they can increase prices when congestion is observed, and decrease them when the traffic load is relatively light. Thus, there is a \emph{feedback loop} between ISPs offering prices and users correspondingly adjusting their usage \cite{Ganesh,Paschalidis}. This \emph{responsive pricing} sets prices so as to keep user demand under a certain level; if an ISP further chooses the prices so as to optimize a proportional fairness criterion on the amount of bandwidth allocated to different users, we obtain \emph{proportional fairness pricing} \cite{Courcoubetis,kelly1998rate,gibbens-kelly}. Many variants of responsive pricing have been proposed in the literature, principally as a congestion control mechanism; in practice, it would be impractical for users to manually respond to the prices offered for each Internet connection. Hence, automation of client devices (or agents) to intelligently adapt their data consumption will be necessary to realize such real-time pricing. But recent HCI studies \cite{chetty2010s,sigchi} have revealed complex patterns of household politics and user opinion regarding such decision-making about bandwidth consumption. In particular, there is a reluctance among users to delegate such bidding or scheduling to automated agents that stems from a conflict between the psychological assurance of manual control and the convenience of automation, which in turn depends on the perceived trust-worthiness of the underlying system. Many of the findings reported later in this chapter on user behavior and user interface design may serve as guidelines in designing user-friendly client-side agents to enable such pricing plans.

Another form of congestion pricing, \emph{effective bandwidth pricing}, incorporates a form of QoS by charging users based on their connection's peak and mean rates \cite{Kelly-eff}. One can also explicitly incorporate different QoS by using \emph{priority pricing}, in which users can pay less by accepting a longer delay at congested times \cite{Gupta}. If the prices are chosen correctly, the system reaches an equilibrium, in which each user's packets are processed within the delay paid for.

\textbf{\emph{Auction-based:}} One disadvantage of real-time congestion pricing is that in practice, the ISP must set the prices (just) before observing user behavior. Since user demand can change with time, the ISP may end up setting non-optimal prices due to outdated assumptions of user demand. ``Smart market'' pricing addresses this slight delay with an auction-like scheme, in which users attach a bid to their packets that signifies their willingness to pay \cite{MacKie-Mason,Murphy-Murphy}. ISPs then admit a limited number of packets in descending order of the bids so as to limit network congestion. Users are charged the lowest bid admitted, which represents the ``cost of congestion.'' While smart market pricing allows true real-time pricing, it also requires automated agents on user devices to make bids as necessary and keep track of the final amount charged.

\textbf{\emph{Raffle-based:}} This is a variation of dynamic time-dependent pricing inspired by lottery reward mechanism. Under \emph{raffle-based pricing}, the exact price that users pay is determined after-the-fact, i.e., in a probabilistic manner that depends on the amount of data consumed by a user \cite{loiseau2011incentive}. Users have a chance to receive a monetary reward during congested times if they agree to shift their demand to less-congested times. They are entered into a lottery for a fixed reward, where the probability of winning the lottery depends on the user's contribution to the total amount of traffic shifted. While such a pricing plan is attractive to ISPs in that the total reward offered is fixed, users may be less willing to shift their traffic because of the uncertainty in winning the lottery and the reward amount, which depend on external factors like the behavior of other users.

\begin{figure}
\centering
\includegraphics[width = 0.6\textwidth]{Figures/Loop_TUBE.pdf}
\vspace{-0.1in}
\caption{Feedback loop schematic of day-ahead pricing.}
\label{fig:tube-loop}
% \vspace{-0.05in}
\end{figure}

\textbf{\emph{Day-ahead time-dependent:}} In an effort to increase user certainty of the prices, ISPs can guarantee their time-dependent prices one day in advance, and continue to compute new prices to maintain this sliding one-day window of known prices \cite{ha2012tube,Carlee}. Users can then plan their usage in advance, while ISPs can adapt their estimates of user behavior and usage volume in calculating the prices for subsequent days. Day-ahead pricing thus strikes a balance between user convenience and ISP adaptability. This has been a successful pricing mechanism in the electricity market, and hence can be adapted to broadband networks with careful consideration. A schematic of the resulting feedback loop is shown in Figure \ref{fig:tube-loop}. In the next section, we examine a prototype of day-ahead pricing for mobile data in order to illustrate the ``end-to-end'' nature of an SDP deployment.

We pause to briefly compare day-ahead TDP with the other types of dynamic pricing discussed above. Real-time pricing and Smart Market mechanisms require users to delegate some control and operate in an automated mode as the time-scale is too short for user-mediated choices. On the other hand, simple 2-period (day \& night) time-of-day pricing has time scale that is too long to take advantage of any spare capacity availability and time elasticity of demand which vary at much shorter time-scale. Auction-based mechanisms will require modifications to the network equipments (e.g., agents to recognize bid amounts and perform admission control) and client side agents for automated bidding. Raffle-based pricing creates uncertainty in rewards and unless the time-varying prices are known in advance, users may be reluctant to adopt such data plans. A day-ahead dynamic time-dependent pricing plans solves many of these issues by providing guarantees on the future prices in advance, takes advantage of demand elasticities at shorter time-scales, and provides ISPs with a mechanism to optimize the prices they offer.

But what are the challenges of realizing dynamic day-ahead time-dependent pricing?
\begin{itemize}
\item How to develop an economic model for dynamic day-ahead TDP which computes optimized prices that accounts for users' time elasticity of demand in maximizing the total revenue of the network provider? The price computation needs to consider (a) the cost incurred in offering price discounts, (b) savings from shifting some traffic from peak to off-peak hours, (c) the increase in baseline demand in discounted periods due to potential ``sales day" effect.
\item How to engineer a system that enables this pricing by developing both provider and client-side modules (in particular, the user interfaces needed for users to react to the offered prices)?
\item How can researchers carry out field trials of such pricing plans by interposing themselves as a ``bandwidth reseller" between the network providers and its real consumers?
We will address these questions in Sections \ref{sec:dayahead}, \ref{sec:psychology}, \ref{sec:engg}, and \ref{sec:TDP}.  
\end{itemize}

\subsection{Comparison with other Markets: Similarities and Differences }

Let us now take a look at what forms of time-dependent pricing have been already field tested and exist in the real world in networks that suffer from congestion problems to identify differences and opportunities for innovating TDP plans for broadband networks.   
Much like today's data networks, the electricity and transportation markets have both experienced a capacity shortage over the past decade and have developed new pricing plans to cope with the resulting shortfall. By comparing electricity usage and road traffic to data traffic, we see that these industries are quite similar to data networks, and that their pricing plans may inform SDP for mobile data. Indeed, both industries observe a highly variable demand throughout the day, allowing for both static and dynamic pricing plans.
%Over the past decade, the electricity market has experienced a capacity shortage much like that anticipated for wireless ISPs in the near future.  
In particular, time-of-day road tolls have been offered in many transportation networks, and many electricity utilities have both trial-ed and deployed time-of-day pricing. We give an overview of such pricing plans in this section, with the aim of highlighting the unique challenges posed by refining such pricing plans to accommodate broadband data networks. Figure \ref{fig:compare} gives an overview of the analogies between pricing plans proposed for the transportation, broadband, and electricity industries.

The similarities and differences between these pricing plans reflect the different industries for which they are designed. In particular, we observe the following distinctions:
\begin{enumerate}
\item
\textbf{\emph{Real-time communication:}} User devices on data networks, e.g., smartphones, are capable of real-time communication with the ISP network, for instance if the prices change in real time. But such real-time feedback for price (toll) changes in road networks is harder to realize and will require additional infrastructural support. In electricity markets, new smart grid interfaces have been developed that can display real-time prices, but individual devices, e.g., air conditioners or vacuum cleaners, generally cannot interact directly with the provider smart grid and require a smart energy controller to schedule their energy consumption.
\item
\textbf{\emph{Elasticity of demand:}} Smartphones' ability to communicate with the ISP network in real time is complemented by users' ability to easily control their usage on individual devices and applications. For instance, a user could simply stop streaming a video if the price increases; such measures could also be automated within the device. The users' decisions will reflect the large variance in the demand elasticity of different types of applications (some of which, such as software downloading, P2P, file backup may not even require user participation and can be completed in small chunks whenever low prices are available). In contrast, devices on electricity networks typically consume energy constantly as long as they are active. There is little opportunity for many devices (e.g., washer, dryer, lights) to complete their activities in an intermittent manner without requiring active user engagement. In road networks, the contrast is even more stark; users in the network (e.g., already driving) cannot easily exit or postpone their activity.
\item
\textbf{\emph{Long-term volatility:}} Most people do not have a concrete idea of how much data they consume each month, partly because most data plans charged a flat fee for unlimited access until recently. Moreover, an individual's data usage can vary greatly from day to day, as relatively casual actions such as streaming a video can have a large impact on total data consumption. In contrast, most people have a relatively good idea of how much they drive per day, and the distance traveled, and road toll fees. Thus, people may be more able to plan ahead by buying permits (e.g., EZ pass) or carpooling during congested hours. In electricity markets, household demand similarly does not vary much from day to day. Consumption of electricity is largely driven by user \emph{needs}, rather than the more volatile \emph{preferences} that drive demand for Internet data.
\end{enumerate}
% For instance, user devices on data networks, such as smartphones, are capable of real-time communication with the ISP network, and allow a user to interact with pricing policies in real time. Moreover, users are able to quickly adjust their behavior, e.g., stopping a video streaming session. With transportation networks, users cannot view real-time price adjustments or leave the network easily once they have begun driving. New smart grid technologies enable some real-time price updates, but these are not nearly as widespread as smartphones. Users can often plan ahead more readily for their transportation and electricity needs, e.g., buying permits from secondary markets at congested times. A user's data rate, however, is highly variable even on a long timescale, and most users are not aware of how much data they actually consume. Most people, however, do have an idea of how many miles they drive at what times of the day, which makes planning for road pricing particularly easy.
\begin{figure}
\centering
\includegraphics[width = 0.6\textwidth]{Figures/Relationships.pdf}
\caption{Comparison of pricing plans in the transportation, broadband, and electricity industries.}
\label{fig:compare}
\end{figure}
% These trials and the accompanying literature suggest that similar plans may be adopted by network operators. 
% Indeed, the electricity and broadband networks are quite similar in that data traffic can be compared to electricity usage, while different data applications are analogous to different energy-consuming household appliances.  In light of this analogy, the discussion below gives a brief overview of the existing literature on pricing for electricity industries, focusing on time-of-day and dynamic pricing.  A summary of all these pricing policies and their related literature is provided in Table \ref{references}.

\subsubsection{Static Pricing}

Traditional road pricing has been simple flat-rate cordon pricing, analogous to flat pricing of data. Pricing by vehicle type, analogous to app-based pricing for data, has also been proposed, e.g., charging trucks more than passenger vehicles \cite{ushighways}. Forms of flat-rate priority pricing have also been implemented, most obviously in the Paris metro's pricing scheme from which data networks' Paris metro pricing takes its name. High-occupancy vehicle or ``carpool'' lanes can also been seen as analogous to priority pricing, in that users can self-select to take advantage of less-congested HOV lanes by paying the higher ``price'' of carpooling with other passengers.

In a common variation on flat-rate tolls in road networks, the flat-rate toll can vary depending on the time of the day \cite{RoadPricing}, for a pricing plan analogous to time-of-day pricing. However, such charges are still flat-rate, i.e., they do not depend on the distance traveled over the road network. Distance-traveled pricing, analogous to usage-based pricing in broadband networks in that users' charge is proportional to the distance traveled, has also been proposed for transportation networks, and has been offered in Taiwan and the U.S. \cite{holguin2006comparative,wen2005traveler}. In fact, the Taiwanese implementation varies the distance-traveled price depending on the time of the day; it is thus a form of time-of-day pricing.

Time-of-day pricing is the major form of static pricing practiced in the electricity industry. Most trials of time-of-day pricing for electricity markets have focused on peak/off-peak pricing, as electricity demand generally follows a less variable pattern than data demand, with extremely low demand at night and higher demand during the day. For instance, one major source of electricity consumption is air conditioning in the summer, which follows a fairly regular pattern of being on during the day and off at night. Indeed, many trials have shown time-of-day pricing to be effective in reducing excess demand during peak hours. One popular variant that has also been trial-ed is \emph{critical peak pricing}, in which certain days are designated as ``critical,'' e.g., especially hot days during the summer. On these critical days, the peak price goes up to increase users' incentives to reduce demand. Some studies with California consumers have shown that critical-peak pricing is much more effective than simple peak/off-peak pricing \cite{CharlesRiver,herter2007residential}. In this trial, users with ``smart devices'' that automatically reduce energy consumption reduced their usage almost twice as much as other users, indicating that user interfaces for interacting with prices are critical to the success of dynamic or time-of-day pricing plans.

\subsubsection{Dynamic Pricing}

Congestion-based pricing has been proposed in both the transportation and electricity industries. One form of congestion pricing in road networks charges users at a price-per-mile rate that is based on their average speed. However, though considered in Cambridge, U.K., this pricing plan was never implemented \cite{RoadPricing}. A more complex pricing plan proposed using several dynamic origin-destination models to compute effective route costs depending on real-time congestion conditions in the road network \cite{Joksimovic}. Drivers would then be able to take shorter routes for higher prices; however, computing these prices is highly non-trivial, and it would be difficult to communicate the prices of different routes to drivers in the network.

One variation on dynamic pricing for road networks involves a \emph{secondary market}, in which governments can sell permits to pass through congested areas. Users can then form a market to sell these permits \cite{starkie1986efficient}. However, similar pricing schemes have not yet been proposed for data networks, likely due to the difficulty in setting up a secondary market among users. Moreover, the increasingly ubiquitous nature of data connectivity has made it more impractical to ask users to completely refrain from consuming data at congested times.

Some electricity pricing researchers have argued that dynamic pricing can lead to significant gains over simple ToD pricing \cite{borenstein2005long}. Both congestion pricing and auction pricing have been proposed for electricity markets; however, such works often have a more consumer-focused outlook than do pricing proposals for data.
%
In an auction-based electricity market, electricity distributors can make dynamic offers to users (i.e., households) who respond with real-time electricity purchases.  Auction schemes have been proposed that take into account varying electricity capacity, which can significantly improve market efficiency \cite{vytelingum2010trading}.

Many papers have studied responsive dynamic pricing from a user's perspective of predicting future prices and scheduling devices accordingly.  A game-theoretic framework can be used to model users' scheduling of energy usage as a cooperative game; if users cooperate, the total demand on a network can then be reduced, enhancing efficiency \cite{caron2010incentive}. Other works propose algorithms to predict prices in advance \cite{Du,mohsenian2010optimalres} and schedule user devices accordingly; users thus try to anticipate electricity providers' real-time pricing. This price prediction is not necessary with day-ahead pricing, though day-ahead pricing offers electricity providers less flexibility \cite{joe2012optimized}. However, such prediction and scheduling algorithms, which have received relatively little attention for data usage, might help make dynamic congestion pricing for data more amenable to users.

% Mohsenian-Rad et. al.'s related paper \citeyear{mohsenian2010optimal} considers the same problem, but with an emphasis on several users sharing a power source and simultaneously scheduling energy consumption in a distributed manner.  More recently, Du and Lu \citeyear{Du} introduced an appliance commitment algorithm that schedules thermostatically controlled household loads based on price and consumption forecasts to meet an optimization objective.

Other papers consider users' actions in conjunction with the provider's price determination \cite{borenstein2002dynamic}. Such approaches can facilitate a study of social welfare, and may incorporate uncertainty in supply and demand \cite{brunekreeft2000price,chao1983peak,samadi2010optimal}. One may also consider a feedback loop between users and an electricity provider, which can yield real-time pricing algorithms analogous to those for dynamic congestion control in data networks \cite{roozbehani2010dynamic}.
Some works have also considered appliance-specific models of user demand, analogous to different applications having different demands for data \cite{li2011optimal}. A unique feature of these models is the ability to store electricity, e.g., in batteries, for use in later congested periods. Thus, from the provider's perspective, the user can effectively shift his or her energy consumption to less congested times, even though from the user's perspective nothing has been shifted.

### 2013-07-22 Compiled/chapter3/Economics.tex
all files compile correctly

\section{Economics of SDP}\label{sec:econ}

% The main goal of smart data pricing is to relieve network congestion by incentivizing users to alter their behavior. Using pricing as an incentive mechanism is not a new idea--indeed, it might be said that most pricing plans exist to incentivize users! Regarding mobile data usage specifically, one can divide the prior literature into two categories: \emph{static pricing}, in which the offered prices are essentially fixed and change on a timescale of months or years, and \emph{dynamic pricing}, in which the prices offered change on a timescale of seconds or hours. In this chapter, we use one example of static and one example of dynamic pricing to elucidate key economic concepts useful for smart data pricing research.

Given the wide variety of SDP pricing algorithms presented in Section \ref{sec:congestion}, a thorough discussion of the theory behind each one is impractical for a book chapter. In this section, we instead select four representative scenarios to illustrate some of the key economic principles often used in formulating different types of pricing algorithms. We first consider static pricing on a single link, and then consider both real-time dynamic pricing and day-ahead time-dependent pricing. Readers familiar with network economics may wish to skip this section.

\subsection{Usage-Based Pricing: A Single Link Example}\label{sec:single}

An operator generally sets its mobile data prices so as to achieve a certain objective, e.g., maximizing profit. In this section, we review some standard economic concepts that are often used in formulating such objective functions. We consider two agents: end users and ISPs.\footnote{Sponsored content and app-based pricing models may also include content providers as a separate type of agent.} For simplicity, we consider only one ISP with a given set of customers, and we suppose that the ISP wishes to build a last-mile access link in its network. The ISP wishes to determine both the \emph{capacity to provision} on this link, as well as the \emph{price per unit bandwidth} to charge its users on the link. This is a standard monopolist profit maximization that we discuss below. We denote the capacity with the variable $x$, and the price by the variable $p$. The ISP-user interaction is summarized in Figure \ref{fig:supply-demand}.

\begin{figure}
\centering
\includegraphics[width = 0.8\textwidth]{Figures/Supply_Demand.eps}
\caption{User-ISP interaction in a mobile data marketplace.}
\label{fig:supply-demand}
\end{figure}

We first consider users' decisions to purchase certain amounts of bandwidth on the ISP's new access link. In modeling this user behavior, we suppose that each user acts so as to maximize his or her \emph{consumer surplus function}, denoted by $U_j(x_j,p)$ for each customer $j = 1,2,\ldots, J$. The function $U_j$ is the net benefit to a consumer from the utility received in purchasing $x_j$ amount of bandwidth for a price $p$ per unit bandwidth.\footnote{This function may be additively decomposed into the form $U_j(x_j,p) = V_j(x_j) - px_j$, i.e., a utility term $V_j$ and the price paid $px_j$. In this scenario, $V_j(x_j)$ is often called the utility, and $U_j(x_j,p)$ the net benefit received by the user. Additively incorporating the price can also be interpreted as incorporating user budget constraints through Lagrange multipliers; more details can be found in Section \ref{sec:realtime}.} Thus, given a price $p$, if $U_j(y_j, p) > U_j(x_j,p)$, user $j$ prefers to purchase $y_j$ units of bandwidth, rather than $x_j$ units. Since the ISP chooses the value of $p$, each user $j$ takes the price as given and chooses the quantity of bandwidth to purchase $\left(x_j\right)$ so as to maximize the utility $U_j(x_j, p)$. We denote this utility-maximizing quantity as $x_j^\ast(p)$.\footnote{The argument $p$ emphasizes the fact that this optimal bandwidth $x_j^\ast$ depends on the price $p$ offered by the ISP.} These functions $x_j^\ast(p)$ are called users' \emph{demand functions}; adding them up, we obtain the \emph{aggregate demand function}, $D(p) = \sum_j x_j^\ast(p)$.

We now consider the ISP's problem of choosing a link capacity $x$ and price $p$. Given a price $p$ and assuming full utilization of the link capacity, the ISP chooses $x$ so as to maximize its utility function. Usually, the ISP's utility is simply its \emph{profit}, but other functions can be used. We write the ISP profit as $px - c(x)$, where $px$ is the ISP revenue and the function $c(x)$ denotes the cost of building a link of capacity $x$. Given $p$, the ISP can then find $x^\ast(p)$, the optimal link capacity as a function of the price $p$. We use $S(p) = x^\ast(p)$ to denote this supply side function.

When the user and ISP are at a \emph{market equilibrium}, supply equals demand: $D(p) = S(p)$. At such a price $p^\ast$ satisfying this relation, each user maximizes his or her own utility by purchasing $x_j^\ast\left(p^\ast\right)$ amount of bandwidth, and the ISP maximizes its utility by providing just enough capacity $x^\ast\left(p^\ast\right) = \sum_j x_j^\ast\left(p^\ast\right)$ to support those users' demands.
%
One often-analyzed property of this equilibrium is the \emph{social welfare}, defined as the sum of the utility received by all users $j$ and the ISP:
\begin{equation*}
\sum_j U_j\left(x_j^\ast,p^\ast\right) + p^\ast\sum_j x_j^\ast\ - c\left(\sum_j x_j^\ast\right),
\end{equation*}
where $x_j^\ast$ is understood to be evaluated at the equilibrium price $p^\ast$. This social welfare can be divided into two portions: the \emph{user surplus}, or the sum of user utilities, and the \emph{ISP surplus}, or the utility (here, profit) obtained by the ISP. Depending on the utility functions $U_j$ and the cost function $c$, the total social welfare may change, and the users and ISP may receive different portions of the overall social welfare.

Before moving on, we pause to discuss some of the more common extensions of the simple problem above. One is to introduce \emph{budget constraints} on each user's utility maximization problem: the user may not want to spend more than a certain amount $B_j$, in which case each user $j$ maximizes the utility $U_j(x_j,p)$ subject to the constraint $px_j \leq B_j$. We may also consider a situation in which users impose \emph{externalities} on each other, i.e., a given user $j$'s utility is affected by the capacity allocated to other users $i\neq j$. For instance, there may be a positive externality in which user $j$'s utility increases as other users send traffic over the link in order to interact with user $j$. On the other hand, one could also observe negative externalities, in which congestion from other users' traffic diminishes a particular user's utility, e.g., by increasing delay.

When solving for the market equilibrium above, we initially took the price $p$ as fixed for the end users and ISP, and then found the equilibrium market price $p^\ast$. In fact, one can obtain this equilibrium price by only examining the optimal behavior of end users and ISPs, i.e., without explicitly considering market equilibrium. Suppose that the ISP, knowing users' demand functions $x_j^\ast(p)$, calculates its revenue as a function of price to be $p\sum x_j^\ast(p)$ (the price, multiplied by the user demand as a function of price). The ISP can then choose both $p$ and $x$ so as to maximize its profit $p\sum_j x_j^\ast(p) - c(x)$, subject to the constraint that the link capacity be able to accommodate users' total demand $\sum_j x_j^\ast$, i.e., that $x \geq \sum_j x_j^\ast$. It is easy to see that (assuming the cost $c(x)$ is increasing in the capacity $x$), at the optimum, $x = \sum_j x_j^\ast$. The ISP then chooses the optimal price $p$ so as to maximize $p\sum_j x_j^\ast(p) - c\left(\sum_j x_j^\ast(p)\right)$. One can show that the resulting optimal price, which we will call $\overline{p}$, is the same equilibrium price $p^\ast$ obtained above: at $\overline{p}$, each user $j$ demands $x_j^\ast(\overline{p})$, and the ISP chooses its optimal capacity $x^\ast(\overline{p})$. This is exactly the point at which the supply and demand curves intersect, i.e., $p^\ast$.

The above reasoning, in which an ISP chooses a price to offer subject to users' behavior as a function of the price chosen, is a simple example of a \emph{game} between users and ISPs. In such a game, several players interact with each other, and each player acts to maximize his or her own utility, which may be influenced by other players' decisions. For instance, in this scenario, users interact with the ISP by utilizing the access link in its network and paying some price. Their decisions on how much capacity to utilize (i.e., choosing $x_j^\ast$) are influenced by the ISP's choice of the price $p$. This interpretation of the single-link example leads us to next consider some basic principles of \emph{game theory} in relation to SDP.

\subsection{Incentive Compatibility: Game-Theoretic Principles}\label{sec:gametheory}

To illustrate some of the basics of game theory, we again consider the single link example above. The user-ISP interaction in such a scenario is an example of a \emph{Stackelberg game}, in which one player, the ``leader,'' makes a decision (e.g., the ISP sets a unit price $p$ for link capacity) and the remaining players, or ``followers,'' then make their own decisions based on the leader's actions. In SDP, this framework reflects the need to consider users' and ISPs' optimal actions when choosing pricing policies that incentivize particular types of resource allocations. In this example, users choose their demands $x_j^\ast(p)$, given the ISP's price $p$. Stackelberg games, which often arise in user-ISP interactions, may be solved using \emph{backwards induction}: first, one computes the followers' actions as a function of the leader's decision (in our example, we compute the functions $x_j^\ast(p)$). The followers' actions are sometimes called a \emph{best response} to the leader. The leader then takes these actions into account and makes his or her own decision (given that users' demands are $x_j^\ast(p)$, the ISP chooses the optimal price $p$). This decision is then the best response to the followers.

The backwards induction process leads to a \emph{subgame perfect equilibrium} in the Stackelberg game: at this equilibrium, each player is maximizing his or her own utility, and no player has an incentive to change his or her behavior. To formalize this definition, we will need to first explain the concept of a \emph{Nash equilibrium}. Consider a general game with $n$ users, each of whom can take an action, e.g., by choosing the value of a variable $y_j$; $j = 1,2,\ldots,n$; and suppose that each user $j$'s utility $V_j$ is a function of all of the $y_j$ variables, i.e., $V_j = V_j\left(y_1,y_2,\ldots,y_n\right)$. Then a set of actions $z_1,\ldots,z_n$ is a Nash equilibrium if $V_j\left(z_1,\ldots,z_j,\ldots,z_n\right) \geq V_j\left(z_1,\ldots,y_j,\ldots,z_n\right)$ for any $y_j \neq z_j$. In other words, assuming that all the other players take actions $z_i$, player $j$'s action $z_j$ optimizes its utility $V_j$.

We may generalize the concept of a Nash equilibrium to a Stackelberg game's subgame-perfect equilibrium by considering \emph{subgames} of the Stackelberg game. We do not give the general definition of a subgame here, but it may be understood by envisioning the Stackelberg game as a dynamic game with different levels defined by the time of decision: on the first level, users make their decisions, and on the second, ISPs make their decisions. A subgame encompasses a group of players who mutually interact, but do not directly interact with other players at their level. In our scenario, a subgame would be a subset of users and the ISP. A subgame-perfect equilibrium of the full Stackelberg game is then a set of actions that comprise a Nash equilibrium in each subgame of the full game. It can be shown that any equilibrium found from backwards induction is a subgame-perfect equilibrium; one can easily check that this is the case in our example scenario. Nash and subgame-perfect equilibria are considered stable in that once they have been achieved, no user has an incentive to change their behavior. (Unfortunately, one cannot in general guarantee that such an equilibrium will be achieved in the first place, and a game may have multiple Nash equilibria.)

Another type of game that often arises in SDP is that of competing service providers. For instance, we may have an \emph{oligopoly} of a few companies who dominate the market for mobile data, e.g., AT\&T and Verizon in the United States are the dominant market players. Each of these companies then competes for customers (i.e., market share) and revenue with the others. This competition defines their interactions, and each company can try to make strategic decisions that optimize its market share. Given a mathematical model of the companies' actions, one can then try to study the corresponding game, e.g., by computing possible Nash equilibria.

While certainly useful for explicit pricing problems like that considered above, game theory can also be applied to more general resource allocation problems, just as SDP allows ISPs to incentivize users to consume data so as to realize particular resource allocations. To illustrate these uses, we again consider the single link example, but we now suppose that the link's capacity is fixed and that the ISP wishes to allocate this fixed amount of capacity $x$ among its $n$ users.

If users selfishly maximize their individual utilities (i.e., choose demands $x_j^\ast(p)$), then the ISP can set a virtual price $p$ to force an allocation in which $\sum_j x_j^\ast(p) = x$, i.e., all of the available capacity is utilized, and each user maximizes his or her utility. This price serves as a signal through which the ISP can control users' demands. However, such an allocation may be unfair: very price-sensitive users may be able to afford significantly less capacity than others. Since revenue is no longer involved, the ISP can afford to care about other objectives like fairness. Indeed, a vast literature exists on just such a problem; we will not go into fairness theory here, but we will present one approach inspired by game theory.

In the Stackelberg game discussed above, users did not cooperate: each user maximized only his or her own utility, subject to the ISP's offered price. Yet if users do cooperate, they may reach a better decision. We can study this problem by first defining individual users' utilities $U_j(y_j)$; given a capacity amount $y_j$, each user $j$ derives utility $U_j(y_j)$. For instance, users could jointly choose their demands $y_j$, subject to the capacity constraint $\sum_j y_j \leq x$, so as to maximize an overall utility function $U\left(U_1(y_1),\ldots,U_n(y_n)\right)$. Depending on the choice of $U$, of course, one would obtain different allocations $y_j^\ast$. We use $y_j^\ast$ to denote the $y_j$ that jointly maximize $U$. Nash proposed that the $y_j^\ast$ satisfy the following four axioms:
\begin{enumerate}
\item
\emph{Invariant to affine transformations:} For each user $j$, define the utility function $V_j(y_j) = \alpha_j U_j(y_j) + \beta_j$ for some constants $\alpha_j > 0$, $\beta_j$. Then the allocation $\left\{z_j^\ast\right\}$ maximizing $U\left(V_1(z_1),\ldots,V_n(z_j)\right)$ satisfies $V_j\left(z_j^\ast\right) = \alpha_jU_j\left(y_j^\ast\right) + \beta_j$ for each user $j$, where the allocation $\left\{y_j^\ast\right\}$ maximizes $U\left(U_1(y_1),\ldots,U_n(y_n)\right)$. An affine transformation of the utility functions $U_j$ does not change the utility received at the optimal allocation.
\item
\emph{Pareto-optimality:} An allocation $\left\{y_1^\ast,\ldots,y_n^\ast\right\}$ is Pareto-optimal if for any user $j$, any feasible allocation $\left\{z_1,\ldots,z_n\right\}$ with $U_j(z_j) > U_j\left(y_j^\ast\right)$ satisfies $U_i(z_i) < U_i\left(y_i^\ast\right)$ for some user $i$. In other words, no user can be made better off without making another worse off.
\item
\emph{Independence of irrelevant alternatives:} Suppose that $U\left(U_1(y_1),\ldots,U_n(y_n)\right) > U\left(U_1(z_1),\ldots,U_n(z_n)\right)$ for two feasible allocations $\left\{y_j\right\}$ and $\left\{z_j\right\}$. Then if the problem constraints are relaxed to allow new feasible allocations, we still have $U\left(U_1(y_1),\ldots,U_n(y_n)\right) > U\left(U_1(z_1),\ldots,U_n(z_n)\right)$.
\item
\emph{Symmetry:} Suppose that $\left\{y_1,\ldots,y_n\right\}$ and $\left\{z_1,\ldots,z_n\right\}$ are feasible capacity allocations with $U_{j_1}(y_{j_1}) = U_{j_2}(z_{j_2})$ for some users $j_1$ and $j_2$, $U_{j_2}(y_{j_2}) = U_{j_1}(z_{j_1})$, and $U_j(y_j) = U_j(z_j)$ for all $j\neq j_1, j_2$. Then $U\left(U_1(y_1),\ldots,U_n(y_n)\right) = U\left(U_1(z_1),\ldots,U_n(z_n)\right)$. In other words, switching the order of the utilities received does not change the overall utility $U$.
\end{enumerate}
An allocation satisfying these four axioms is said to be a \emph{Nash bargaining solution}. One can show that if $U$ is taken to be $\prod_j U_j(y_j)$, then the resulting $y_j^\ast$ is a Nash bargaining solution. Taking the logarithm, we see that this is equivalent to maximizing $\sum_j \log\left(U_j(y_j)\right)$. In other words, users choose their demands to maximize the sum of the \emph{logarithms} of their utilities $U_j$. Since the logarithm is sub-linear for large $U_j(y_j)$, the optimal allocation $\left\{y_j^\ast\right\}$ will penalize large values of $U_j$ relative to smaller ones, yielding a ``more equal'' allocation $U_j\left(y_j^\ast\right)$ than simply maximizing the sum of utilities $\sum_j U_j(y_j)$.

% One way to perform this allocation is to use \emph{virtual prices}, in which the ISP offers a price $p$, just as in our example above. However, the prices are not monetary ones; users do not need to actually pay the ISP. The prices are simply a way for the ISP to (indirectly) allocate capacity to users. simply to impose the requirement that users' total demand equal the capacity $x$. Using the same utility functions $U_j$ and demands $x_j^\ast$ from before, we have $\sum_j x_j^\ast(p) = x$, which may be solved for the optimal price $p$. Yet such an allocation may be unfair: very price-sensitive users may be able to afford significantly less capacity than others. Thus, if an ISP wishes to \emph{directly} allocate

\subsection{Real-Time Dynamic Pricing}\label{sec:realtime}

So far, we have focused on pricing and bandwidth allocation of a single access link. However, in reality an ISP's network does not consist of single bandwidth links: it is, in fact, a network, with multiple nodes and links between them. Data traffic between two nodes, e.g., between a user and a content provider, flows across a subset of the network links. Since different links may experience different types of congestion at different times, an ISP may want to adjust the prices charged based on how much congestion is experienced by a particular user at a given time. It is this philosophy that lies behind dynamic pricing for congestion control.

To illustrate the basic concepts of congestion control, we consider a relatively simple example given in Kelly et al.'s seminal paper on the subject \cite{kelly1998rate}. Consider a set of nodes, indexed by $n = 1,2,\ldots,N$, and a set of links indexed by $l = 1,2,\ldots,L$ that connect different nodes together. We suppose that each node $n$ wishes to communicate with another node, and we use $R_n$ to denote the subset of links traversed by node $n$'s traffic.\footnote{Choosing the optimal routes for each node $n$ is a non-trivial problem in itself; for simplicity, we assume here that all of the routes $R_n$ are fixed.} The ISP's goal is then to set a traffic rate $x_n$ for each node $n$, such that 1) the total amount of traffic on any link $l$ lies below link $l$'s capacity $c_l$, and 2) all users are as satisfied as possible. To accomplish this, each link can set a unit \emph{congestion price} for traffic on the link. By prescribing the evolution of these prices in time, the ISP can satisfy its two objectives in a distributed manner.

We first define a \emph{routing matrix} to summarize the routes taken by different nodes' traffic over the network: let $R$ be an $L\times N$ matrix, and set $R_{ln} = 1$ if $l\in R_n$, i.e., node $n$'s traffic travels over link $l$, and $R_{ln} = 0$ otherwise. If we concatenate nodes' traffic rates $x_n$ into an $N\times 1$ vector $\vec{x}$, we see that $\vec{y} = R\vec{x}$ yields a vector of length $L$. Each entry $y_l$ of $\vec{y}$ equals the total volume of traffic on link $l$. Letting $\vec{c}$ be an $L\times 1$ vector of the capacities of each link $l$, we then have the capacity constraint $R\vec{x} \leq \vec{c}$: the total amount of traffic on each link $l$ cannot exceed the link's capacity.\footnote{This inequality is to be interpreted component-wise, i.e., each component of the left-hand and right-hand side vectors should satisfy it.} This constraint ensures that the ISP's first objective is satisfied.

The ISP's second objective is that each user be ``as satisfied as possible.'' We define satisfaction by defining utility functions $U_n(x_n)$ for each node $n$; the ISP is then assumed to assign source rates $x_n$ so as to maximize the total sum of utilities, $\sum_n U_n(x_n)$, subject to the constraint $R\vec{x} \leq \vec{c}$. To solve this problem, we next make the assumption that each utility function $U_n$ is concave. Such an assumption is consistent with the economic principle of \emph{diminishing marginal utility}, i.e., that the extra utility received from an additional unit of bandwidth decreases as the user receives more and more bandwidth. Under this assumption, the ISP's objective function $\sum_n U_n(x_n)$ is concave. Since the constraints $R\vec{x}\leq \vec{c}$ are linear, the overall optimization problem is a convex optimization.

We now follow standard optimization theory and introduce a $L\times 1$ vector of Lagrange multipliers $\vec{p}$, with each $p_l$ corresponding to link $l$'s capacity constraint in the component-wise inequality $R\vec{x}\leq \vec{c}$. These multipliers $\vec{p}$ will eventually become the congestion prices set by the links $l$. The ISP's optimization problem is then equivalent to solving
\begin{equation}
\min_{\vec{p}\geq 0}\max_{\vec{x}}\left(\sum_{n = 1}^N U_n(x_n) + \vec{p}^T\left(\vec{c} - R\vec{x}\right)\right).
\label{eq:opt_dyn}
\end{equation}
Solving (\ref{eq:opt_dyn}) centrally can be done relatively easily; it is not hard to see that we have the solution
\begin{equation*}
x_n^\ast = {U_n'}^{-1}\left(q_n\right),\quad p_l^\ast = \begin{cases} 0 &{\rm if}\;c_l - y_l > 0 \\ > 0 &{\rm if} c_l - y_l = 0 \end{cases},
\end{equation*}
where we define $\vec{y} = R\vec{x}$ and the $n$th entry of the $N\times 1$ vector $\vec{q} = R^T\vec{p}$ equals the sum of the congestion prices for the links traversed by node $n$'s prices; $q_n$ thus represents the total price paid by user $n$. However, our goal is to develop a \emph{distributed} solution, in which nodes adjust their rates $x_n$ and links adjust their prices $p_l$ so as to converge to the optimal solution. The ISP can drive these dynamics with the link prices--i.e., links change their prices $p_l$, and nodes respond by adjusting their rates $x_n$ according to the solution above. An example of a price-driven algorithm is \cite{low1999optimization}
\begin{equation}
p_l(t + 1) = \left[p_l(t) - \gamma \left(y_l(t) - c_l\right)\right]^+,\quad x_n(t + 1) = {U_n'}^{-1}\left(q_n(t + 1)\right),
\label{eq:dyn}
\end{equation}
where the argument $t$ denotes the value of a variable at time $t$, and we consider discretized times $t = 1,2,\ldots$ The constant $\gamma > 0$ is a stepsize parameter, and the $+$ superscript $\left[\alpha\right]^+$ denotes the maximum of a quantity $\alpha$ and 0.\footnote{This modification ensures that the prices are nonnegative.} We note that each link $l$ evolves its price $p_l$ using only the traffic on that link $y_l(t)$ and its capacity $c_l$, both of which are known without communicating with other nodes or links. Similarly, each node $n$ adjusts its rate $x_n$ based only its price $q_n$, a quantity that can be carried with node $n$'s traffic and is known without node $n$ communicating with other nodes or links. One can show that these dynamics converge to the optimal prices and rates if the stepsize $\gamma$ is sufficiently small. Moreover, as user utilities $U_j$ or link capacities $c_l$ change over time, following the dynamics (\ref{eq:dyn}) will reposition the rates and prices to their new optimal values. Thus, real-time dynamic pricing can adapt to network congestion levels and keep ISP traffic from exceeding the network capacity. We explore some variations on this dynamic pricing model in the next section.

\subsection{Dynamic Day-ahead Time-Dependent Pricing}\label{sec:dayahead}

One limitation of real-time dynamic pricing is that it requires users to respond to price changes by adjusting their demands in real time. Yet to consciously involve users in adapting their demand to price changes, a longer timescale is preferable.\footnote{There is of course a tradeoff in choosing the timescale: while longer timescales are simpler for users to understand, shorter timescales allow prices to more accurately reflect congestion.} In this section, we again consider the case of a single access link for longer timescale time-dependent pricing. Our goal is to develop an alternative model that is simple enough to be practical as a real pricing plan, yet also helps to alleviate congestion on ISP networks. We aim to offer prices to users in advance, so that users can have more certainty in planning changes to their usage behavior. Indeed, previous studies showed that prices which change in real time can discourage changes in usage, as users are not sure whether the price will decrease in the future \cite{shih2001pricing}. Additionally, in computing the optimized prices (discounts) that maximize network provider's profit, the model needs to consider (a) the cost incurred from offering price discounts, (b) the savings from shifting some traffic from peak to off-peak hours, and (c) the increase in baseline demand in discounted periods due to potential ``sales day" effect.

We divide one day into $T$ time periods--e.g., $T = 24$ periods of one hour each--and suppose that the ISP can offer a different price at each time $t = 1,2,\ldots,T$. We suppose that the ISP has an existing link of capacity $C$, and that it may accommodate additional demand at a marginal rate $\gamma$. This additional cost may represent the increased cost of handling customer complaints once usage exceeds capacity $C$, or an additional investment cost necessary for increasing the network capacity. We let $X_t$, $t = 1,2,\ldots,T$, denote user demand at each time $t$. Then the ISP's cost of accommodating these demands is
\begin{equation}
\sum_{t = 1}^T \gamma\max\left(X_t - C, 0\right).
\label{eq:capacity}
\end{equation}
Given that accommodating demand in the peak periods is more expensive than demand during less-congested periods in which $X_t < C$, the ISP has an incentive to offer lower prices in less-congested periods, thus inducing users to shift some demand into those periods. This is the core idea of time-dependent pricing: by offering lower prices during less congested times, the ISP can even out its demand over the day, resulting in lower capacity costs. Our treatment here follows that in \cite{ha2012tube}.

To formalize this argument, we next need to develop a mathematical model for users' shifting of traffic in response to the prices offered. We let $p_t$ denote the price offered by the ISP at each time $t$. We suppose that the ISP can offer a maximum price that is normalized to 1, e.g., if the ISP currently offers a usage-based price of 1 without time-dependent pricing. We can then define the discount offered at each time $t$ as $d_t = 1 - p_t$. Users' willingness to shift some traffic from one period $t_1$ to another period $t_2$ then depends on the additional discount offered in period $t_2$, i.e., $d_{t_2} - d_{t_1}$. If $d_{t_2} >> d_{t_1}$, then the user will be able to save more money and thus will be more willing to shift his or her traffic.\footnote{One could also consider functions such as the ratio of discounts in the two periods, but $d_{t_2} - d_{t_1}$ has a reasonable interpretation as the amount of money a user can save by shifting traffic.} However, users' willingness to shift their usage does not just depend on the discounts offered: it also depends on the time shifted $t_2 - t_1$. For instance, a user may be very willing to shift some traffic by half an hour, but much less willing to shift his or her usage by more than an hour, even with the same discounts.\footnote{In this section, we suppose that users only delay their traffic, i.e., traffic is shifted only to later, not earlier, times. In practice both types of shifting can occur, but we make the reasonable assumption that it is more natural for users to delay some traffic than it is to anticipate future usage.}

The discounts offered and time shifted are not the only factors determining how willing users are to shift their data traffic: the \emph{type of traffic} also matters. Software downloads, for instance, may be readily delayed by several hours, but users are often much less willing to delay urgent apps like email retrieval. For simplicity, in the following discussion we assume only one type of traffic, but our model can be easily extended to multiple traffic types by adding up the amount of traffic shifted for each traffic class. Considering multiple traffic classes is necessary for ISPs to accurately take user behavior into account: since each user will have some traffic that can be delayed more than other traffic, the fact that we consider the simultaneous behavior of multiple users need not mitigate the need for multiple traffic classes.\footnote{The amount of traffic corresponding to each traffic class can be treated as a model parameter; along with the other model parameters, it can be estimated from observed aggregate usage data.} We use $w(d,t)$ to denote users' probability (i.e., willingness) to delay their traffic by an amount of time $t$, in exchange for an additional discount $d$. We suppose that $w$ is increasing in the discount $d$ and decreasing in time $t$, and that the value of $w$ lies between 0 and 1, so that it may be interpreted as a probability. Many functions $w$ meet these criteria, e.g., the functions
\begin{equation*}
w(d,t) = \frac{\max(d,0)}{\Gamma(t + 1)^\beta},
\end{equation*}
where $\Gamma$ is a normalizing factor and $\beta \geq 0$ is a model parameter that controls the rate at which willingness to shift decreases with the time shifted $t$. For larger values of $\beta$, the willingness to shift decays faster with time, denoting increased impatience.

The general idea behind our model is to use the $w(d,t)$ functions to calculate how much traffic is shifted to different times of the day, relative to a baseline traffic pattern without time-dependent discounts. Our model thus attempts to reflect users' thought processes in looking at a set of future prices and deciding whether to delay their traffic or not. While a more traditional model would use utility functions (cf. Section \ref{sec:single}) to describe users' behavior, user utility functions are difficult to derive from user behavior, since ``utility'' attempts to quantify the relatively vague idea of ``user benefit.'' The probability of shifting traffic, on the other hand, can be calculated directly from observing the amount of traffic shifted in response to discounts offered to users.

Given the functions $w$, the expected amount of traffic shifted from time $t_1$ to time $t_2$ is
\begin{equation*}
w\left(d_{t_2} - d_{t_1}, \left|t_2 - t_1\right|_T\right),
\end{equation*}
where $\left|t_2 - t_1\right|_T$ denotes the number of periods between time $t_2$ and period $t_1$, modulo the number of periods $T$ (e.g., if $t_2 < t_1$, then $\left|t_2 - t_1\right|_T$ is the number of periods between period $t_1$ and period $t_2$ on the next day). Given an initial amount of traffic $Y_t$ in each period $t$, we calculate that $Y_{t_1}w\left(d_{t_2} - d_{t_1}, \left|t_2 - t_1\right|_T\right)$ amount of traffic is shifted from time $t_1$ to time $t_2$. The ISP then loses $\left(d_{t_2} - d_{t_1}\right)Y_{t_1}w\left(d_{t_2} - d_{t_1}, \left|t_2 - t_1\right|_T\right)$ amount of revenue due to the traffic shifted from time $t_1$ to $t_2$. Some additional revenue is lost from the unshifted traffic in each period, for a total revenue loss of
\begin{equation}
\sum_{t = 1}^T \left[Y_td_t + \sum_{s\neq t} Y_s\left(d_t - d_s\right)w\left(d_t - d_s, \left|t - s\right|_T\right)\right].
\label{eq:loss}
\end{equation}
In addition to this revenue loss, offering discounts at some times may induce users to increase their overall usage during those time periods. This increase is independent of any usage shifted from more expensive times and constitutes a psychological ``sales day effect,'' which was observed during time-dependent pricing trials \cite{ha2012tube,sigchi}. It reflects the qualitative insight that user demand increases as the price charged decreases: users see that they are saving money, relative to usage during high-price periods, and are thus encouraged to spend and use more. The larger the discount offered $d_t$, the larger this increase will be. We can model this increase with a power law: given an initial amount of traffic $Y_t$ in period $t$, the amount of traffic after a discount $d_t$ is offered (neglecting any traffic shifted to other periods) is $Y_t\left(1 + d_t\right)^\alpha$ for some positive model parameter $\alpha$. We then have the desirable property that demand does not change if no discount is offered $\left(d_t = 0\right)$; if $\alpha = 0$, the total demand does not depend on the discount at all. The ISP thus earns additional revenue
\begin{equation}
Y_t\left(\left(1 + d_t\right)^\alpha - 1\right)(1 - d_t)
\label{eq:revenue}
\end{equation}
due to this additional demand in period $t$. We can then add the ISP's revenue loss from discounts offered (\ref{eq:loss}), less the revenue gain (\ref{eq:revenue}) from additional traffic, to the ISP's cost of capacity (\ref{eq:capacity}) to obtain the objective function
\begin{equation*}
\sum_{t = 1}^T \left[Y_td_t - Y_t\left(\left(1 + d_t\right)^\alpha - 1\right)(1 - d_t) + \sum_{s\neq t} Y_s\left(d_t - d_s\right)w\left(d_t - d_s, \left|t - s\right|_T\right) + \gamma\max\left(X_t - C,0\right)\right],
\end{equation*}
where the total traffic at each time $t$ is
\begin{equation*}
X_t = Y_t\left(1 + d_t\right)^\alpha + \sum_{s\neq t} Y_s w\left(d_t - d_s, \left|t - s\right|_T\right) - \sum_{s\neq t} Y_t w\left(d_s - d_t, \left|s - t\right|_T\right).
\end{equation*}
The first term represents the increase in traffic due to the discount, while the second is the amount of traffic deferred into period $t$, and the third term the amount of traffic deferred out of period $t$. The ISP then chooses the discounts $d_t$ (equivalently, the prices $p_t = 1 - d_t$) to minimize its objective function. Under reasonable conditions on $\alpha$, $\gamma$, and $w$, this is a convex optimization problem, which may be rapidly solved.

By solving this optimization problem, the ISP can obtain a set of prices for one day; it can then offer day-ahead pricing by running an optimization in each period that determines the optimal discount to offer one day from the current time. Moreover, the ISP can observe the traffic consumed in each period once these discounts are offered. Comparing this data to the traffic observed without discounts, the ISP can estimate the parameters of its user behavior models (i.e., $\alpha$ and $w$ above) from the observed changes in usage, given the prices offered. These estimates can be periodically updated and used to calculate new prices, completing a feedback loop (cf. Figure \ref{fig:tube-loop}) between users and the ISP \cite{ha2012tube}.

### 2013-07-22 Compiled/chapter3/Exercises.tex
all files compile correctly

\section{Exercises}

\begin{enumerate}
\item \emph{Nash Bargaining}

Consider a single access link and suppose that its bandwidth capacity $C$ is shared by $n$ users. Let $x_i$ denote the amount of bandwidth received by each user $i = 1,2,\ldots,n$, and suppose for simplicity that each user receives utility $x_i$ from $x_i$ amount of bandwidth. Suppose that the ISP allocates bandwidth to users so as to maximize
\begin{equation}
\prod_{i = 1}^n x_i\quad{\rm s.t.}\;\sum_{i = 1}^n x_i \leq C.
\end{equation}
Show that the resulting allocation $\left\{x_i^\ast\right\}$ satisfies the four Nash bargaining axioms presented in Section \ref{sec:gametheory}.

\item \emph{Time-of-Day Smart Grid Pricing}

Taken from M. Chiang, \emph{Networked Life: 20 Questions and Answers}, Cambridge University Press, 2012.
\\ \\
Smart grid electricity providers often set time-dependent prices for energy usage.  This problem considers a simplified example with two periods, the day-time and the night-time.  The provider can set different prices for the two periods, and wishes to shift some night usage to the day.  The energy provider always offers the full price during the night, and offers a reward of \$$p$/kWh during the day.

Suppose that with uniform (not time-dependent) prices, customers vacuum at night, using 0.2 kWh, and also watch TV, using 0.5 kWh, and do laundry, using 2 kWh.  During the day, customers use 1 kWh.  The probability of users shifting vacuum usage from the night to the day is
\begin{equation}
1 - \exp\left(-\frac{p}{p_V}\right),
\end{equation}
where $p_V = 2$, and the probability of shifting laundry to the daytime is
\begin{equation}
1 - \exp\left(-\frac{p}{p_L}\right),
\end{equation}
where $p_L = 3$.  Users never shift their TV watching from the night to the day.

Suppose that the provider has a capacity of 2 kWh during the night and 1.5 kWh during the day.  The marginal cost of exceeding this capacity is \$1/kWh.  Assume that energy costs nothing to produce until the capacity is exceeded.

\begin{enumerate}
\item
Compute the expected amount of vacuum and laundry energy usage (in kWh) that is shifted from the night to the day, as a function of $p$.
\item
Find (to the nearest cent) the reward $p$ which maximizes the energy provider's profit.
\item
Suppose that if vacuum or laundry usage is shifted from the night to the day, it is shifted by 12 hours.  Compute the expected time shifted of vacuum and laundry using $p = p^\ast$, the optimal reward found above.
\end{enumerate}

\item \emph{Paris Metro Pricing}

Taken from M. Chiang, \emph{Networked Life: 20 Questions and Answers}, Cambridge University Press, 2012.
\\ \\
Consider a metro system where two kinds of services are provided: service class $1$ and service class $2$. Let $p_1$, $p_2$ be the one-off fees charged per user when accessing service classes 1 and 2 respectively. Suppose each user is characterized by a valuation parameter $\theta\in [0,1]$ such that its utility of using service class $i$ is
$$U_{\theta}(i)=(V-\theta K(Q_i,C_i))-p_{i},$$
where $V$ is the maximum utility of accessing the service, $K(Q_i,C_i)$ measures the amount of congestion of service class $i$, given $Q_i\geq 0$ as the proportion of users accessing service class $i$ (with $\sum_i Q_i=1$), and $C_i\geq0$ is the proportion of capacity allocated to service class $i$ (with $\sum_i C_i=1$).

At the equilibrium, i.e., no user switches from his selection, $U_{\theta}(i)$ is merely a linear function of $\theta$. Suppose the equilibrium is illustrated as in Figure \ref{fig:equilibrium}.\\

\begin{figure}
\centering
\includegraphics[width=8cm]{Figures/Equilibrium.pdf} 
\caption{\label{fig:equilibrium}Illustration of equilibrium in PMP.}
\end{figure}

Let $\theta_1$ be the $\theta$ of the user who is indifferent to joining the first service class or opting out of all the services, let $\theta_2$ be that of the user who is indifferent to joining the first service class or the second service class, and let $F(\theta)$ be the cumulative distribution function of $\theta$.
\begin{enumerate}
\item
Show that
\begin{subequations}
\begin{align*}
Q_1=&F(\theta_1)-F(\theta_2),\\
Q_2=&F(\theta_2),\\
V-p_1=&\theta_1 K(Q_1,C_1),\\
p_1-p_2=&\theta_2 (K(Q_2,C_2)-K(Q_1,C_1)).\\
\end{align*}
\end{subequations}

\item
Assume that $\theta$ is uniformly distributed, i.e., $F(\theta)=\theta$, and that the congestion function is defined as
$$K(Q,C)=\frac{Q}{C}.$$
Solve for $\theta_1$ and $\theta_2$ as functions of $V$, $p_1$, and $p_2$. 
\end{enumerate}

(Hint: Try $\frac{p_1-p_2}{V-p_1}$.)\\

(For details, see C. K. Chau, Q. Wang, and D. M. Chiu, ``On the Viability of Paris Metro Pricing for Communication and Service Networks," \emph{Proc. IEEE INFOCOM}, 2010.)\\

\item \emph{Two-Sided Pricing}

Taken from M. Chiang, \emph{Networked Life: 20 Questions and Answers}, Cambridge University Press, 2012.
\\ \\
Suppose an ISP charges a content provider (CP) the usage price $h_{CP}$ and flat price $g_{CP}$ and charges an end user (EU) the usage price $h_{EU}$ and flat price $g_{EU}$. For simplicity, we assume zero flat prices $\left(g_{CP}=g_{EU}=0\right)$. Let $\mu$ be the unit cost of provisioning capacity. The demand functions of the CP and EU, denoted as $D_{CP}$ and $D_{EU}$ respectively, are given as follows:\\
\begin{subequations}
\begin{align*}
D_{EU}(h_{EU})=&\left\{
\begin{array}{ll}
x_{EU,max}(1-\frac{h_{EU}}{h_{EU,max}})&\mbox{, if } 0\leq h_{EU}\leq h_{EU,max}\\
0,&\mbox{, if } h_{EU}>h_{EU,max}
\end{array}\right.\\
D_{CP}(h_{CP})=&\left\{
\begin{array}{ll}
x_{CP,max}(1-\frac{h_{CP}}{h_{CP,max}})&\mbox{, if } 0\leq h_{CP}\leq h_{CP,max}\\
0,&\mbox{, if } h_{CP}>h_{CP,max}.
\end{array}\right.\\
\end{align*}
\end{subequations}
The parameters are specified as follows:
\begin{subequations}
\begin{align*}
h_{CP,max}&=2.0\mu,\\
h_{EU,max}&=1.5\mu,\\
x_{CP,max}&=1.0,\\
x_{EU,max}&=2.0.\\
\end{align*}
\end{subequations}
The ISP maximizes its profit by solving the following maximization problem
\begin{equation}
\begin{array}{ll}
\mbox{maximize}	&(h_{CP}+h_{EU}-\mu)x\\
\mbox{subject to}	&x\leq \mbox{min}\{D_{CP}(h_{CP}),D_{EU}(h_{EU})\}\\
\mbox{variables} 	&x\geq 0, h_{CP}\geq 0, h_{EU}\geq 0.
\end{array}
\end{equation}

Find the optimal $x^\star, h_{CP}^\star, h_{EU}^\star$.\\

\item \emph{Monitoring Mobile Data Usage}

Many commercial mobile applications have been developed to help users keep track of their mobile data usage. Some examples include 3GWatchdog Pro (Android), DataWiz (iOS and Android), MyDataManager (Android), and Onavo Count (Android and iOS).
\begin{enumerate}
\item
Visit two or three app websites and list their features (e.g., showing usage by application, forecasting future usage, alerts when you approach your monthly data quota). Are there any significant differences between the apps? Can you identify any consistent differences between iOS and Android apps?
\item
What visual elements are used in the app designs? For instance, do the apps use bars or pie charts to represent usage? How are these displays different on different apps?
\item
Based on your answers to the above questions, try to design your own app for tracking mobile data usage. What screens would you implement? What features would you offer?
\end{enumerate}
\end{enumerate}

### 2013-07-22 Compiled/chapter3/Factors.tex
all files compile correctly

\section{Driving Factors of Network Congestion}\label{sec:factors}

With mobile devices becoming smarter, smaller, and ubiquitous, consumers are embracing the technology and driving up the demand for mobile data.  According to Cisco's VNI \cite{CiscoVNImobile}, in 2012, global mobile data traffic grew more than 70 percent year over year, to 855 petabytes a month. The growth rate varied by regions, with 44\% growth in Western Europe, and about 101\% in the Middle East and Africa and a 95\% growth rate in Asia Pacific. This section identifies some of the key factors that are expected to drive this growth in demand for mobile data (ref. Figure \ref{fig:driving_factors}).

\begin{figure}[h]
\centering
\includegraphics[width = 0.9\textwidth]{Figures/Driving_factors.pdf}
\vspace{-0.1in}
\caption{Factors driving the demand for mobile data.}
\label{fig:driving_factors}
\vspace{-0.05in}
\end{figure}

{\bf\emph{Cloud Services and M2M Applications:}}
Cloud-based services that synchronize data across multiple mobile devices, such as iCloud, Dropbox, and Amazon’s Cloud Drive, can be a significant factor in traffic growth for ISPs \cite{icloud}. Similarly, machine-to-machine (M2M) applications that generate data intermittently (e.g., sensors and actuators, smart meters) or continuously (e.g., video surveillance) often load the network with large signaling overhead \cite{BellLabs}.  However, these traffic types also have some intrinsic time elasticities that create opportunities for intelligently shifting them to low-congestion times through pricing incentives.

{\bf\emph{Mobile Videos:}}
Video has also been a major contributor to mobile data traffic growth, accounting for 51 percent of global mobile data traffic at the end of 2012. It is expected to account for 66 percent of global mobile data traffic by 2017 \cite{CiscoVNImobile}. A study by Gartner \cite{Gartner2012} states that the worldwide mobile video market had 429 million mobile video users in 2011, projected to grow exponentially to 2.4 billion users by 2016. Smartphones and tablet sales will contribute 440 million new mobile video users during the forecast period. The report also forecasts that the worldwide share of mobile video connections on 3G/4G will increase from 18\% in 2011 to 43\% in 2015 \cite{Goog-mobile}. These growth rates are being further fueled by mobile video content delivery via mobile-optimized websites and video advertisements.

{\bf\emph{Capacity-Hungry Applications:}}
The popularity of handheld devices has led to rapid growth in the development of other bandwidth-hungry applications for social networking, music, personalized online magazines, etc. in addition to file downloads and video streaming. Virgin Media Business reports that the average smartphone software uses 10.7 MB per hour, with the highest-usage app, Tap Zoo, consuming up to 115 MB/hour. In the current ecosystem, app developers do not have enough incentives to account for network conditions, and consequently many smartphone apps are not optimized for bandwidth consumption.

{\bf\emph{Bandwidth-Hungry Devices:}}
The widespread adoption of handheld devices, equipped with powerful processors, high-resolution cameras, and larger displays, has made it convenient for users to stream high-quality videos and exchange large volumes of data.  Data from laptops with 3G dongles and netbooks with wireless high-speed data access contributes the most to wireless network congestion \cite{BellLabs}.  As for smartphones, Cisco projects that the average monthly data usage will rise from 150 MB in 2011 to 2.6 GB in 2016 \cite{CiscoVNImobile}.  New features like Siri on the iPhone 4S, which has doubled Apple users' data consumption, are driving this growth \cite{siri}.

The key takeaways from the above discussion of the various factors contributing to network congestion are:
\begin{itemize}
\item Different types of applications and services have different levels of time elasticity of demand (e.g., cloud backup versus financial applications).
\item There is a great variance in the rate of growth of different types of applications (e.g., demand for mobile videos is growing fast).
\item Different applications consume bandwidth at different rates and have a large variance in QoE requirements (e.g., many apps are not optimized for bandwidth while some well-designed apps can adapt to available network QoS). 
\item There is a growing variance in the bandwidth requirements of different smart devices (e.g., iPad versus iPhone versus feature phones).
\end{itemize}
These factors contribute to the need for smarter data plans that can account for the variances across users' QoE needs, time elasticity of demand, application traffic characteristics and their willingness to pay for the service.

%These developments have led ISPs to view pricing as their ultimate congestion management tool, paving the way for the adoption of Smart Data Pricing (SDP) measures, such as time-dependent and usage-based pricing.
Before delving deeper into SDP's promise in addressing congestion issues \cite{SDPForum}, in the next section we first explore how these trends are impacting the various stakeholders of the network ecosystem, i.e., network operators, consumers, app developers, and content providers.

### 2013-07-22 Compiled/chapter3/Intro.tex
all files compile correctly

\section{Introduction}\label{intro}

Advances in Internet technologies have resulted in an unprecedented growth in demand for data. In particular, demand in the mobile Internet sector is doubling every year \cite{CiscoVNI}. Given the limited wireless spectrum availability, the rate of growth in the supply of wireless capacity (per dollar of investment) is unlikely to match the rate of growth in demand in the long run. Internet Service Providers (ISPs) are therefore turning to new pricing and penalty schemes in an effort to manage the demand on their network, while also matching their prices to cost. But changes in pricing and accounting mechanisms, if not done carefully, can have significant consequences for the entire network ecosystem. Multiple stakeholders in this ecosystem, including operators, consumers, regulators, content providers, hardware and software developers, and architects of network technologies, have all been tackling these issues of charging and allocating limited network resources. Even back in 1974, while writing about the future challenges of computer communication networks, Leonard Kleinrock \cite{Kleinrock} noted: 
\begin{quote}
\emph{[H]ow does one introduce an equitable charging and accounting scheme in such a mixed network system? In fact, the general question of accounting, privacy, security and resource control and allocation are really unsolved questions which require a sophisticated set of tools.}
\end{quote}
While much progress has been made on developing technical solutions, methods, and tools to address these issues, continued growth of the network ecosystem requires developing a better understanding of the underlying economic and policy perspectives. The broader area of \emph{network economics}, which deals with the interplay between technological and economic factors of networks, is therefore receiving more attention from engineers and researchers today. Economic factors like pricing, costs, incentive mechanisms and externalities\footnote{Network externality is the notion that the cost or value of being a part of a network for an individual user depends on the number of other users using that network. For example, the value of a network grows as more users adopt and positive externalities are realized from being able to communicate with other users on the network. Similarly, when many users start to content for limited resources of a bottleneck link of a network, negative externalities from congestion diminish a user's utility from accessing the network.} affect the  adoption outcomes (i.e., success or failure of network technologies) and stability \cite{Sen2010,Joseph2007,wifi-3g-infocom}, influence network design choices \cite{Sen2011-MVF,Sen2009-SVS}, and impact service innovation \cite{Zhang2010}. Conversely, technological limitations and regulatory constraints determine which kind of economic models are most suited to analyze a particular network scenario. This interplay between technology, economics, and regulatory issues is perhaps most easily observed in the case of broadband access pricing, for example, in evaluating the merits of ``flat-rate" versus ``usage-based" pricing or the neutrality of ``volume-based" versus ``app-based" accounting, etc. In this chapter we discuss the current trends in access pricing among service operators, factors that affect these decisions, analytical models and related considerations. In particular, we observe that \emph{Smart Data Pricing}\footnote{SDP is the broad set of ideas and principles that goes beyond the traditional flat-rate or byte-counting models and instead considers pricing as a network management solution. See http://www.smartdatapricing.org.} (SDP) is likely to emerge as an effective way to cope with increased network congestion. These smarter ways to count and treat data traffic illustrate three shifts in the principles of network management:

\begin{enumerate}
\item {\bf \emph{Pricing for end-user Quality of experience (QoE) and not just byte-counting}}: Simple policies like usage-based pricing (byte-counting) (a) force users to pay the same amount per unit of bandwidth consumed irrespective of the congestion levels on the network,\footnote{In 1997, David Clark wrote \cite{DDClark} that ``The fundamental problem with simple usage fees is that they impose usage costs on users regardless of whether the network is congested or not."} and (b) fail to account for the fact that different applications have different bandwidth requirements to attain a certain QoE for the user. SDP should try to match the cost of delivering application-specific desired QoE requirements of the user to the ISP's congestion cost at the time of delivery.     
\item {\bf \emph{Application layer control to impact physical layer resource management}}:  Today's smart devices with their easy to use graphical user interfaces can potentially enable consumer-specified choice for access quality. Whether done manually or in an automated mode, users' specifications of their willingness to pay for their desired QoE of different applications can be taken in as inputs at the APP layer and used to control PHY layer resource allocation and media selection (e.g., WiFi offloading versus 3G). But enabling this requires consumer trials to understand how to design incentives and create interfaces that can be effective in modifying end-user behavior.
\item {\bf \emph{Incorporating edge devices as a part of network management system}}: Instead of managing traffic only in the network core, SDP explores ways to make edge devices (e.g., smart mobile devices and customer-premise equipments like gateways) a part of the network resource allocation and management system. For example, instead of throttling traffic in the network core using the policy charging and rules function (PCRF), the edge devices (e.g., home gateways) themselves could locally regulate demand based on a user's budget, QoE requirements, and network load or available prices. Such measures to push control from the network core out to the end-users, while preserving the end-to-end principles of the Internet, have been gaining attention among networking research groups \cite{M3I}. 
\end{enumerate}

But before delving deeper into pricing ideas, let us pause to address some common misconceptions often encountered in public discourse. First, many believe that the Internet's development cost was borne by the United States Government, and hence that taxpayers have already paid for it. In reality, by 1994 the National Science Foundation supported less than 10\% of the Internet and by 1996 huge commercial investments were being made worldwide \cite{MIT}.

Second, users often do not realize that the Internet is not free \cite{MIT, chetty2011my} and think its cost structure is the same as that of information goods. In contrast to \emph{information goods}, which tend to have zero marginal costs,\footnote{Marginal cost is the change in the total cost that arises when the quantity produced changes by one unit, e.g., the cost of adding one more unit of bandwidth.} Internet operators incur considerable network management operation and billing costs. MacKie-Mason and Varian \cite{MM-Varian} have shown that while the marginal cost of some Internet traffic can be zero because of \emph{statistical multiplexing}, congestion costs can be quite significant. 
In regard to delivery of bits, it is worthwhile to note that there are some important factors at play: 
\begin{enumerate}[(a)]
\item There is a large and growing variance in the QoE requirements of the different types of applications that consumers are using today, and
\item The network operator's cost of delivery per bit for a given QoE level also has significant variance, ranging from essentially zero marginal cost in uncongested times to very high in congested times.
\item There is also a variance in user's willingness to pay for different types of traffic and QoE levels.
\end{enumerate}

So why not match the right pairs? Most SDP ideas aim to do exactly that, i.e., match the operator's cost of delivering bits to the consumer's QoE needs for different application types at the amount they are willing to spend.

Third, there is a popular misconception that network costs are high because billing costs account for 50\% of telephony costs. Although true for running costs, it is only 4-6\% when depreciation of sunk costs is added \cite{M3I}. Another important cost for wireless operators today is the cost of acquiring new spectrum to support the growing bandwidth needs of the customers. However, spectrum is limited and expensive, and even auction-based spectrum reallocation schemes are projected to fall short of the demand for spectrum.

Fourth, the belief that better technologies like 4G and offloading mechanisms will solve the problems is already being questioned -- ``The reasons are two-fold: The amount of spectrum made available to U.S. wireless companies is limited, but the carriers have also been sluggish in buying up enough backhaul to support their capacity requirements. There is only so much data that can be crammed into wireless spectrum -- and only so much spectrum available to wireless networks. Thanks to rising mobile data demands, a current wireless spectrum surplus of 225 MHz will become a deficit of 275 MHz by 2014, according to the FCC \cite{CNN-4G}."

%Higher access speeds lead to both higher demand and applications that consume higher amount of bandwidth.

Fifth, users fear that changes in pricing policy will increase their access fees. This need not always be the case, as one can design incentive mechanisms that reward good behavior (e.g., price discounts in off-peak hours to incentivize shifting of usage demand from peak times). In other words, smarter pricing mechanisms can \emph{increase consumer choices} by empowering users to take better control of how they spend their monthly budget. For example, under time-dependent usage-based pricing \cite{comm-mag,ha2012tube}, users have better control over their monthly bills by choosing not only \emph{how much} they want to consume, but also \emph{when} they do so). Smart data pricing also has to be smart in its implementation and in its user interface design, with careful study of user psychology and human-computer interaction aspects, as we will illustrate in later sections and case studies.

Lastly, we also need to remember that pricing is related to the market competition and user population density. For the interested reader, an overview of access fees in different parts of the world is provided in Section \ref{Dataplans}.

The following questions provide a useful way to think about SDP: 
\begin{enumerate}[(I)]
\item Why do we need SDP? Isn't network pricing an untouchable legacy?\\
Section \ref{sec:factors} provides an overview of the driving factors behind network congestion, and the challenges that it poses to various stakeholders of the network ecosystem are discussed in Section \ref{sec:threats}. We also discuss the rapid evolution in pricing among network operators and highlight in Section \ref{SDP} how Smart Data Pricing ideas will be useful in finding solutions that can work in today's networks.
\item Haven't other fields already used pricing innovations? What are the key SDP ideas relevant to communication networks?\\
We provide an overview of Internet pricing ideas in the existing literature in Section \ref{sec:congestion}, including some pricing plans from the electricity and transportation industries that can be applied to broadband pricing. Section \ref{sec:econ} provides an overview of a few examples and analytical models of known pricing mechanisms to illustrate key economic concepts relevant to the SDP literature. We also highlight many crucial differences between SDP in communication or data networks and pricing innovations in other industries.
\item Isn't SDP too complex to implement in the real world?\\
Section \ref{sec:TDP} provides a case study of a field trial of ``day-ahead time-dependent pricing" and discusses the model, system design, and user interface design considerations for realizing this plan. It serves to demonstrate both the feasibility of creating such SDP plans for real deployment while also pointing out the design issues that should be kept in mind. The discussion highlights the end-to-end nature of an SDP deployment, which requires developing pricing algorithms, understanding user psychology, designing an effective interface for communicating those prices to users, and implementing an effective system to communicate between the users and ISPs.
\item What are the outstanding problems in enabling SDP for the Internet?\\
SDP is an active area of research in the network economics community and a set of 20 questions and future directions are provided in Section \ref{sec:20Q} for researchers and graduate students to explore. Many of these research questions have been discussed at various industry-academia forums and workshops on SDP \cite{SDPForum,SDP2013}.  
\end{enumerate}

%This chapter is organized as follows: We first provide an overview of the driving factors behind network congestion in Section \ref{sec:factors} and discuss in Section \ref{sec:threats}, the various congestion related issues in the network ecosystem. We then discuss Smart Data Pricing principles and ideas in Section \ref{SDP}. An overview of pricing ideas in the existing literature is given in Section \ref{sec:congestion}, including some pricing plans from the electricity and transportation industries that can be applied to broadband pricing. Section \ref{sec:econ} provides an overview of a few examples and analytical models of known pricing mechanisms to illustrate key economic concepts essential for understanding the SDP literature. We then delve into one such pricing plan in Section \ref{sec:TDP}, namely, ``day-ahead time-dependent pricing,'' in order to illustrate the end-to-end nature of an SDP deployment, which requires developing pricing algorithms, designing an effective interface for communicating those prices to users, and implementing an effective system to communicate between an ISP and its users. Finally, in Section \ref{sec:20Q} we point to some (of many) research questions for broadband access pricing that are still open for further research in this area.

### 2013-07-22 Compiled/chapter3/SDP.tex
all files compile correctly

\section{Smart Data Pricing}\label{SDP}

%The demand for data is growing rapidly in both wired and wireless networks. This growth is primarily driven my the rise of mobile video, bandwidth-hungry apps, bulk transfer to the cloud, and user-friendly interfaces of smart devices. The Cisco Visual Networking Index predicts that data from mobile devices will grow to a volume of 10.8 exabytes per month in 2016, while wired network traffic will increase at a compound annual growth rate of 28\% per year \cite{CiscoVNI}.

%The predicted doubling of demand for mobile data every year also means that technical solutions and spectrum reallocation are going to be insufficient to meet the growing needs of the users. ISPs' (Internet service providers) are therefore turning to new pricing plans in an effort to manage their demand for data and match their prices to cost \cite{comm-mag}. Such policies include charging based on usage volume, throttling users' bandwidth, and imposing caps on monthly usage.  Figure \ref{fig:timeline} provides an overview of the rapidly evolving pricing and demand control practices of the major U.S. ISPs since 2008. In fact, over the past few years, ISPs around the world have started to offer innovative pricing plans, including usage-based and app-based pricing to tackle the problem of network congestion \cite{computing-survey}.

 \begin{figure}
\centering
\includegraphics[width = 0.85\textwidth]{Figures/Timeline.pdf}
\vspace{-0.1in}
\caption{Broadband pricing plans offered by major U.S. ISPs, 2008 - 2013.}
\label{fig:timeline}
\vspace{-0.05in}
\end{figure}

Broadband access pricing and demand control practices have rapidly evolved among U.S. ISPs since 2008, as seen in Figure \ref{fig:timeline}. Over the past few years, ISPs around the world have started to offer innovative pricing plans, including usage-based and app-based pricing to tackle the problem of network congestion \cite{computing-survey}. Smart Data Pricing (SDP) \cite{SDP2013} is an umbrella term for a suite of pricing and policy practices that have    
been proposed in the past or are being explored as access pricing options by operators instead of the traditional flat-rate model. Such SDP models can include any or of the following mechanisms or a combination, which will be discussed later in the chapter: (a) Usage-based pricing/metering/throttling/capping, (b) Time/location/congestion-dependent pricing, (c) App based pricing/sponsored access, (d) Paris metro pricing, (e) Quota-aware content distribution, (f) Reverse billing or Sponsored Content. SDP does not even need to be an explicit pricing mechanism; it can be another form of innovative congestion management like WiFi offloading or ``fair-throttling''\footnote{Fair throttling involves accounting for user's usage history of contributing to congestion in determining what share of available bandwidth the user should receive in a congested time.}.

The basic ideas of congestion pricing have received much attention as a research topic both in computer networks and information systems literature, and are once again getting a fresh look from academics in recent years. Given the change in the economic and regulatory environment of Internet pricing, it is likely that some of the ideas will be realized in future data plans. However, research in the design of such smart data pricing plans should account for some new factors: (i) the growth in traffic with high time-elasticity of demand (e.g., downloads, P2P, cloud backup, M2M) and the ability to schedule such traffic to a less congested time without user-intervention, (ii) revisiting the issue of dividing the elements of a congestion control-feedback loop between the network backend and the smart end-user devices, (iii) development of new system architectures to deploy these pricing ideas and demonstration of their potential benefits through field trials. In other words, it requires understanding both the \emph{economic theory} of pricing models as well as the \emph{systems engineering} and \emph{human-computer interaction} aspects of realizing such data plans. These require a multi-disciplinary approach in SDP research that bridges theory, systems, and user trials by drawing on economic theory, network engineering and user behavioral studies in a collaborative environment, as shown in Figure \ref{fig:theme}.

 \begin{figure}
\centering
\includegraphics[width = 0.5\textwidth]{Figures/Sdp-theme.pdf}
\vspace{-0.1in}
\caption{Smart Data Pricing research components}
\label{fig:theme}
\vspace{-0.05in}
\end{figure}

% Unfortunately, most academic materials presently available online tend to only provide just enough background on congestion pricing models from a theoretical perspective, rather than taking a more holistic approach.  In particular, most of the materials on economic aspects of the Internet available to students and faculties are either technical papers from conferences such as ACM SIGCOMM, NetEcon, CoNEXT, INFOCOM or edited books of workshop proceedings \cite{MIT}. It is arguably fair to say that most networking textbooks do not cover this important topic in enough depth, and this lack of background materials at the graduate level is indeed a serious issue that the community is now reckoning with. There is a growing realization in the network research community about the need to include network economics as an integral part of engineering curriculum given the complex interplay between technological and economic factors. To this end, we propose to bridge this gap by contributing a chapter that exclusively focuses on the network economics of smart pricing practices for the Internet.

%The key elements of this chapter will be: (i) a detailed overview of pricing proposals, (ii) system design considerations in smart pricing, and (iii) human factors and results of pricing trials of SDP plans. % The authors have been deeply involved in studying pricing plans, published survey articles and research papers on SDP, and have conducted real trials of SDP with ISPs in US and India, which will inform the contents presented in this contribution.   Additionally, the authors will seek inputs from their collaborators and peers  with whom they have been working closely for network economics research,  especially Prof. Andrew Odlyzko of the University of Minnesota and Prof. Roch Guerin of the University of Pennsylvania, both of whom were invited keynote speakers at the first SDP workshop hosted by the authors in Princeton University in 2012 \cite{SDPForum}. We would also provide a summary of open questions and challenging directions in network pricing, along with supplementary background materials, that can benefit graduate students in identifying new research directions to pursue.

% In the following sections, we first provide an overview of the driving factors behind network congestion in Section \ref{sec:factors}. This rising demand for bandwidth has led to new threats to the current Internet ecosystem: ISPs are increasingly feeling a capacity crunch on their networks, leading to new pricing plans that have raised concerns among consumers hit with these new forms of pricing. Content providers, whose videos and other content make up the increased traffic volume that consumers are demanding, also have a stake in this evolving ecosystem, as do application developers and government regulators attempting to find a balance between ISP capacity concerns and Internet availability. We discuss these different perspectives in Section \ref{sec:threats}. In Section \ref{sec:econ}, we consider a few examples of SDP to illustrate key economic concepts essential for understanding the SDP literature. Section \ref{sec:congestion} gives an overview of this literature, including some pricing plans from the electricity and transportation industries that can be applied to broadband pricing. We then delve into one such pricing plan, day-ahead time-dependent pricing, in order to illustrate the end-to-end nature of an SDP deployment, which requires developing pricing algorithms, designing an effective interface for communicating those prices to users, and implementing an effective system to communicate between an ISP and its users. Finally, we point to some (of many) research questions for broadband access pricing that have yet to be answered.

### 2013-07-22 Compiled/chapter3/TDP.tex
all files compile correctly

\section{Case Study of SDP: A Trial of Day-ahead TDP}\label{sec:TDP}

While pricing algorithms are essential to SDP research, in practice such algorithms must be able to function within an ISP network. In this section, we discuss the design and results of a trial of Section \ref{sec:dayahead}'s day-ahead time-dependent pricing (TDP) to highlight some ways in which implementability concerns can influence the development of pricing algorithms. We examine some important system and user interface design principles that were used in developing the prototype of this system, called TUBE, and finally present some trial results that illustrate how these elements can come together in practice. While some SDP trials have been conducted in the past, e.g., the Berkeley INDEX project in the 1990s \cite{INDEX}, the design of this TUBE pilot trial illustrates the way new factors such as smartphones' computing capabilities affect SDP's feasibility.

%Economic models for and system design for enabling dynamic pricing, such as time-dependent pricing (TDP), smart market auctions, sponsored access. Developing a control-feedback loop between the network core and end-devices (i.e., demand-response). Need for new usage accounting and signaling protocols. User interface design factors for smart pricing. Results from field trials of TDP systems.

\subsection{Model Considerations}

Most forms of dynamic pricing, in which prices must be determined in (near) real-time, require the prices to adapt based on users' behavior. For instance, users' perception of the prices offered may change over time, and demographically distinct user populations may react differently to the same set of prices (e.g., teenagers versus businessmen). Offering dynamic pricing thus requires that the ISP first estimate its users' behavior and then use this to inform its choice of prices. In the case of time-dependent pricing, such estimates are particularly important. The basic philosophy of TDP is that by offering lower prices at less congested times, an ISP incentivizes users to shift some of their usage from more expensive, congested times to less congested times. Users' demand over the day is thus even-ed out, with peak usage decreasing; this decrease in peak usage then reduces ISPs' need to over-provision capacity for their peak demand. While lower prices would effectively encourage users to shift their demand, thus reducing costs, ISPs would also lose a large amount of revenue if the prices were too low. Moreover, users might shift their usage too much, and end up creating a new peak period during the discounted times.

In practice, these estimates of user behavior must take into account the available information that the ISP can collect from its users. For instance, TUBE's TDP algorithms, discussed in Section \ref{sec:dayahead}, use only \emph{aggregate usage data}, that is, the total usage volume on the network at different times, in order to estimate user behavior and calculate the prices. This approach has the following benefits:
\begin{itemize}
\item
\emph{Scalability:} Since only aggregate usage is recorded and used in the algorithms \cite{ha2012tube}, we can scale up the user behavior and price computations to multiple users and multiple applications. The algorithm complexity does not increase with the number of users contributing to the aggregate usage totals.
\item
\emph{User privacy:} The amount of traffic that an individual consumes for different applications can be sensitive information (e.g., unusually large amounts of streaming video might reveal a movie buff). The TUBE algorithm does not consider application-specific usage, so the ISP need not receive or record such sensitive information.
%\item
%\emph{Application-specific modeling must be estimated}: Different mobile applications respond differently to TDP: users may be much more willing to delay a software update, for instance, than to delay watching a video. Since application-specific usage data, however, our algorithm must estimate the delay sensitivity of different application types.
\item
\emph{Utility function estimation:} Utility function estimation is usually a hard problem. When temporal considerations are involved, it can potentially become even more complicated, as the utility of consuming data at any given time depends on the prices offered at all times of the day. %But the TDP approach proposed in \cite{ha2012tube}  we do not employ utility functions in modeling users' reactions to prices.
\item
\emph{Empirical observations:} Instead of using utility functions, we can model users' willingness to shift their demand from one period to another, depending on the time elapsed between these periods and the price difference. Such usage shifts can be directly observed by comparing the amount used at different times and prices, and the model can then adapt as these observed shifts change over time.
\end{itemize}
By following these principles, we develop a scalable price calculation and user behavior estimation algorithm (see Section \ref{sec:dayahead} and \cite{ha2012tube} for details) that can be feasibly deployed in a real system.

\subsection{System Design}

A core feature of SDP, and time-dependent pricing in particular, is that it involves both end users and ISPs. Thus, the system design must have components both on ISP servers and on user devices. Figure 
\ref{fig:tdp_arch} shows this division of functionality and the requisite communication channels between the user device and ISP server. In order to make the system practical, we follow three basic principles:

\begin{figure}
\centering
\includegraphics[width = 0.6\textwidth]{Figures/TDP_Architecture.pdf}
\caption{User-ISP functionality division and feedback communication in time-dependent pricing.}
\label{fig:tdp_arch}
\end{figure}

\begin{itemize}
\item
\emph{Functionality separation:} Users and ISPs have different roles in an SDP system: while users respond to the prices offered, an ISP must set the prices. TUBE utilizes individual user devices to facilitate not only displaying prices to users, but also helping them respond to the prices offered via an \emph{autopilot mode} that automatically schedules apps to lower price periods. Since such computations need not involve the ISP, this functionality is located on users' devices.
\item
\emph{A feedback loop:} In order to successfully adapt prices to user behavior, the ISP needs to monitor usage in its network.\footnote{Such usage monitoring also allows the ISP to calculate the amount spent by individual users.} Thus, users must periodically send their usage to the ISP server. Similarly, the ISP must periodically update the prices displayed on users' devices as new prices are calculated. This mutual communication forms a feedback loop.
\item
\emph{An open API:} An ISP's users may have many different devices with different operating systems--for instance, iOS, Android, and Windows phones and tablets. Each of these devices must therefore communicate with the ISP server. To ensure consistency across different device types, TUBE offers an open API for transmission of usage and prices between the ISP and users.
\end{itemize}

\subsection{User Interfaces}

SDP depends not just on pricing algorithms and system design, but ultimately on \emph{whether users respond to the prices offered}. Thus, careful user interface design is necessary to ensure that users understand the prices being offered and to encourage them to respond accordingly. In some cases, interface design goes beyond displaying prices; users' devices can automatically adjust data usage based on the prices offered and user preferences. TUBE's user interface components can be grouped into three different categories:
\begin{itemize}
\item
\emph{Information displays:} Since TUBE offers day-ahead TDP, the prices for the next day should be displayed to users. But users may also find it helpful to track their spending by viewing how much usage they have consumed in the past. TUBE thus shows users both the price and usage for several past hours, so that users can understand how they usually respond to the prices offered and how this affects their spending on data. TUBE also shows the amount used for the five apps with the highest data usage, so that users can see which applications consume more data. Figure \ref{fig:tube_gui}abc shows some sample screenshots of these features. 
\item
\emph{Out-of-app indicators:} Most users find checking a mobile application too onerous for keeping track of current or future prices. A more convenient way to display the prices is to show a color-coded price indicator on the device home screen to (qualitatively) signal the current price to the user, without requiring any special action on the user's part. Such color-coding can also be helpful for visualizing the future prices within the app, so that users can quickly decide whether to wait for lower prices.
\item
\emph{Automation:} Many users prefer not to manually schedule different applications due to the complications involved in tracking future prices. TUBE thus offers an \emph{autopilot mode} that takes into account users' delay sensitivity for different applications and monthly budget to automatically schedule some apps to lower-price times. The autopilot mode utilizes users' past spending to forecast how much the user will spend over a month. If this amount exceeds a user's monthly budget, delay-tolerant apps can be scheduled to lower-price periods; as users' spending further exceeds their budget, apps with lower delay tolerances will be scheduled to lower-price periods. However, such algorithms need to be optimized so as to be as non-intrusive as possible; in user interviews after the TUBE trial, many trial participants expressed concern over an automated algorithm controlling their data usage \cite{sigchi}. One way to accommodate these concerns is to allow users to override the autopilot scheduling and to configure algorithm parameters, e.g., changing the delay tolerances of different apps (Figure \ref{fig:tube_gui}d).
\end{itemize}
% The current and future prices, as well as corresponding usage volumes, are shown; the prices are color-coded to aid visual comprehension. Users can also schedule apps for future, lower-price times.
\begin{figure}[t]
% \vspace{-0.05in}
\begin{center}
\begin{subfigure}[b]{0.22\textwidth}
	\includegraphics[width=\textwidth]{Figures/iphone_ui1.pdf}
	\caption{Price display.}
\end{subfigure}
%
\begin{subfigure}[b]{0.22\textwidth}
	\includegraphics[width=\textwidth]{Figures/iphone_ui2.pdf}
	\caption{Price and usage.}
\end{subfigure}
%
\begin{subfigure}[b]{0.22\textwidth}
	\includegraphics[width=\textwidth]{Figures/iphone_ui3.pdf}
	\caption{Top 5 applications.}
\end{subfigure}
%
\begin{subfigure}[b]{0.22\textwidth}
	\includegraphics[width=\textwidth]{Figures/iphone_ui5.pdf}
	\caption{App scheduling.}
\end{subfigure}
\end{center}
\vspace{-0.25in}
\caption{Screenshots of user interfaces for time-dependent pricing. Users can (a) check the prices for next 24 hours, (b) view their price and usage history, (c) identify the top 5 apps by bandwidth usage, and (d) schedule their apps at different times of the day.}
\label{fig:tube_gui}
\vspace{-0.05in}
\end{figure}

\subsection{Trial Results}

Recently, the authors of the present work developed a prototype of the above pricing algorithms, system components, and user interfaces and trial-ed it with 50 end users \cite{ha2012tube}. We here present some results from this TUBE trial, which illustrate both the importance of user interface design and the effectiveness of optimized TDP. % We also interviewed trial participants after the trial was complete. Not only did participants express appreciation for TDP's ability to offer price discounts, they also agreed in general with SDP's philosophy of more intelligent pricing for data to help users save money and ISPs reduce congestion. Users' openness to these new pricing plans signals that, with the right pricing algorithms and effective system and UI designs, new forms of data pricing can be feasible in practice.

An initial phase of the TUBE trial offered alternating high (10\% discount) and low (40\% discount) time-dependent prices in different hours. After two weeks of following the high-low-low price pattern, the prices changed, repeating the pattern of a 9\% discount at midnight, followed by 28\%, 30\%, 28\%, 9\%, and 30\% discounts in subsequent hours. The home screen price indicator was green for discounts over 30\%, orange for 10--29\% discounts, and red for discounts below 10\%. % Besides being color coded, the indicator also showed the numerical discount value.

Usage in different hours with these pricing patterns can be compared to assess the effect of the indicator color and numerical discount: hours deemed as Type 1 periods offered a 10\% discount in the first stage of the experiment and 28\% discount in the second stage, with the indicator remaining orange despite this increase in the discount.  Type 2 periods offered a 10\% (orange) discount in the first stage and 30\% (green) discount in the second stage, while Type 3 periods offered a 10\% discount in the first and 9\% discount in the second stage of the experiment (the indicator was orange in both periods).  Table \ref{periodtypes} summarizes the combinations of discounts and colors used in the two stages that characterize each type of period.
\begin{table}
\renewcommand{\arraystretch}{1.1}
\centering
\caption{Period types in the color experiment.}
\vspace{-0.1in}
\begin{tabular}{|c|c|cc|cc|}
\hline
Type & Periods & \multicolumn{2}{c|}{First Stage} & \multicolumn{2}{c|}{Second Stage} \\
 & & Color & Disc. & Color & Disc. \\ \hline
1 & 2, 8, 14, 20 & Orange & 10\% & Orange & 28\% \\ \hline
2 & 3, 6, \ldots, 24 & Orange & 10\% & Green & 30\% \\ \hline
3 & 5, 11, 17, 23 & Orange & 10\% & Orange & 9\% \\ \hline
%\st{4} & \st{1, 7, 13, 19} & \st{Green} & \st{40\%} & \st{Orange} & \st{9\%} \\ \hline
%\st{5} & \st{4, 10, 16, 22} & \st{Green} & \st{40\%} & \st{Orange} & \st{28\%} \\
%\hline
\end{tabular}
\label{periodtypes}
\end{table}

To analyze the trial results, the percentage changes in usage for each type of period were computed, relative to usage without time-dependent prices. These changes showed that \emph{users responded more to changes in the price indicator color than changes in the numeric value of the TDP discounts.} In post-trial interviews, nearly all of the trial participants indicated that they relied on the price indicator colors to know the current prices, rather than opening the TUBE app.

Figure \ref{fig:OHP_OLP_OHP} compares the usage changes observed in different period types. Each data point represents one user's average change in each period type, with the size of the data point indicating the volume of usage in the second stage of the experiment.  The reference line represents equal changes in both period types considered. Figure \ref{fig:OHP_OLP_OHP}a shows the average change in usage for each user in Type 1 periods versus Type 3 periods.  For both period types, the color did not change, but the discount in Type 1 periods increased significantly.  Thus, if users had reacted to the numerical prices, usage should increase in Type 1 and decrease in Type 3 periods: users' data points should lie above the reference line.  Figure \ref{fig:OHP_OLP_OHP}a shows that this is the case with only half of the users.  Since the indicator color did not change, users were mostly agnostic to the numerical values of the discounts.
\begin{figure*}
\centering
	\begin{subfigure}[b]{0.51\textwidth}
		\includegraphics[width = \textwidth]{Figures/OHP_OLP_OHP.eps}
		\caption{Period types 1 and 3.}
	\end{subfigure}
	\hspace{-0.05\textwidth}
	\begin{subfigure}[b]{0.51\textwidth}
		\includegraphics[width = \textwidth]{Figures/OHP_GLP_OLP.eps}
		\caption{Period types 1 and 2.}
	\end{subfigure}
\vspace{-0.1in}
\caption{Average percent changes in usage for the period types in Table \ref{periodtypes}. Users' usage behavior is (a) not affected by the prices when only the numerical discounts, but not the indicator color changes. When (b) both the color and numerical discount change, users increase their usage behavior more in low-price periods.}
\label{fig:OHP_OLP_OHP}
\vspace{-0.1in}
\end{figure*}
Figure \ref{fig:OHP_OLP_OHP}b plots the average change in usage in Type 2 versus Type 1 periods.  The discounts in both periods increased by comparable amounts, but the indicator color changed from orange to green only in Type 2 periods.  Most users' data points lie above the reference line, indicating that usage increased more (or decreased less) in Type 2 as compared to Type 1 periods. Thus, users responded to the indicator color despite the comparable numerical discounts. In fact, 80\% of our participants admitted to this behavior when asked in post-trial interviews whether they paid attention to the indicator color, numerical discounts, or both.

The final stages of the trial offered optimized time-dependent prices, with initial user behavior estimates based on the usage observed in previous stages of the trial with non-optimized prices.  The reduction in peak traffic was measured by the \emph{peak-to-average ratio} (PAR), i.e., the ratio of usage in the peak period to average per-period usage, for each day.  Comparing the PARs from before and after optimized TDP reveals that \emph{optimized time-dependent prices reduce the peak-to-average ratio} from usage before time-dependent prices were offered (time-independent pricing, or TIP).  Moreover, \emph{overall usage significantly increased} after TDP was introduced, partially because people used more in the discounted valley periods. % Valley filling and peak reduction together contributed to lower peak-to-average ratios with TDP.
% Part of this increase is likely due to different times of the year, but we conjecture that the discounts offered with TDP induced people to spend more.

Figure \ref{fig:par}a shows the distribution of daily PARs both before and after TDP was introduced.  The maximum PAR decreases by 30\% with TDP, and approximately 20\% of the PARs before TDP are larger than the maximum PAR with TDP.  Thus, TDP significantly reduced the peak-to-average ratio, flattening demand over the day. Moreover, this decrease in PAR is not due to a net loss of traffic.  Figure \ref{fig:par}b shows the average per-user daily usage observed before and after TDP.  The overall volume of usage after TDP is greater than that before TDP; in fact, across all users, the average change in usage from TIP to TDP is a 130\% increase.  Part of this increase may be due to the time of year--TIP usage was measured from July to September, and the TDP usage in January. TDP, however, is likely a major factor: the discounts during off-peak periods allowed users to consume more data while still spending less money and decreasing the PAR. In fact, in post-trial interviews 30\% of the trial participants admitted to consciously using more data in the heavily discounted periods, with one explicitly comparing the situation to shopping at a clothing sale in department stores.
\begin{figure*}
\centering
	\begin{subfigure}[b]{0.42\textwidth}
		\includegraphics[width = \textwidth]{Figures/PAR.eps}
		\caption{PARs.}
	\end{subfigure}
	\hspace{-0.03\textwidth}
	\begin{subfigure}[b]{0.42\textwidth}
		\includegraphics[width = \textwidth]{Figures/Avg.eps}
		\caption{Average daily traffic.}
	\end{subfigure}
\vspace{-0.1in}
\caption{Usage statistics in the TIP (time-independent pricing) and optimized TDP phases of the TUBE trial. When optimized TDP is offered, (a) the ISP's peak-to-average ratio generally decreases, while (b) the average daily traffic per user increases.}
\label{fig:par}
\vspace{-0.1in}
\end{figure*}

### 2013-07-22 Compiled/chapter3/Threats.tex
all files compile correctly

\section{Impact on the Network Ecosystem}\label{sec:threats}

%To understand the implications of these factors, we need to first look at the impact that this growth is having on the various stakeholders of the Internet ecosystem before delving deeper into the promises that \emph{Smart Data Pricing} (SDP) holds \cite{SDPForum}. In the following discussion, we explore the current challenges and threats from the perspective of carriers, consumers, and content providers.

%The Internet ecosystem today suffers from threats to both its \emph{sustainability} and \emph{economic viability}. The sustainability concern is due to the near doubling of demand for mobile data every year  \cite{CiscoVNI}, which brings into question the feasibility of finding any purely technical solutions to this problem. Additionally, changes in the policy to make more spectrum available, although useful, is unlikely to provide any long-term solution.  For example, the US Government's initiative in releasing an additional 50 MHz of spectrum into the market for six carriers to compete for is widely viewed as being quite insufficient \cite{tollfree}.

%The latter concern regarding the economic viability of Internet access is due to the large overage charges and throttling that ISPs (e.g., AT\&T and Verizon) are now imposing on their customers to penalize demand. To understand the implications of these factors, we need to first look at the impact that this growth is having on the various stakeholders of the Internet ecosystem before delving deeper into the promises that \emph{Smart Data Pricing} (SDP) holds \cite{SDPForum}. In the following discussion, we explore the current challenges and threats from the perspective of carriers, consumers, and content providers.

\subsection{ISPs' Traffic Growth}

%The Cisco Visual Networking Index \cite{CiscoVNI} predicted in 2012 that mobile data traffic\footnote{``Mobile data traffic'' includes handset-based data traffic, such as text messaging, multimedia messaging, and handset video services, as well as traffic generated by notebook cards and mobile broadband gateways.} will grow at a compound annual growth rate (CAGR) of 78\% between 2011 and 2016, reaching a volume of 10.8 exabytes per month in 2016. By that time, Cisco predicted that the average smartphone user will consume 2.6 GB per month, as opposed to 150 MB/month in 2011 \cite{CiscoVNImobile}. This growth in per-device data consumption is fueled by the demand for bandwidth-intensive apps: for example, Allot Communications reports that usage of Skype's video calling service grew by 87\% in 2010, while usage of Facebook's mobile app grew by 267\% \cite{allot}. Devices themselves are also contributing to the increase in traffic volume; Apple's iPad 3 quadrupled the screen resolution from the iPad 2, allowing videos of higher quality to be streamed to the device \cite{apple}.

%Cisco predicted that by 2016, Wi-Fi and mobile data will comprise 61\% of all Internet traffic, with wired traffic comprising the remaining 39\% \cite{CiscoVNI}. Although the faster growth rate in mobile data traffic is a major concern, growing congestion on wired networks both at the edge and in the networks' middle mile also poses a significant problem. 
%-- consumer traffic from fixed IPs is expected to grow at a CAGR of 28\% per year from 2011-2016, while managed IP traffic is expected to sustain a 21\% CAGR \cite{CiscoVNI}.\footnote{Cisco's definition of ``consumer'' includes fixed IP traffic generated by households, university populations, and Internet cafes, `while ``managed IP'' includes corporate IP WAN traffic and IP transport of TV and VoD.})

By 2016, ISPs are expected to carry 18.1 petabytes per month in managed IP traffic.\footnote{Cisco's definition of ``managed IP'' includes traffic from both corporate IP wide area networks and IP transport of television and video-on-demand.} But this growth is causing concern among ISPs, as seen during Comcast's initiative to cap their wired network users to 300 GB per month \cite{Comcast-cap}. Even back in 2008, Comcast made headlines with their decision (since reversed) to throttle Netflix as a way to curb network congestion \cite{Comcast-Level3}. Video streaming from services like Netflix, Youtube, and Hulu, are a major contributor to wired network traffic. In fact Cisco predicts that by 2016 fixed IPs will generate 40.5 petabytes of Internet video per month \cite{CiscoVNI}.

Rural local exchange carriers (RLECs) are also facing congestion in their wired networks due to the persistence of the middle-mile problem for RLECs. Although the cost of middle mile bandwidth has declined over the years (because of an increase in the DSL demand needed to fill the middle mile), the bandwidth requirements of home users have increased quite sharply \cite{vglass-2}. Still, the average speed provided to rural customers today fails to meet the Federal Communications Commission's (FCC) broadband target rate of 4 Mbps downstream speed for home users. The cost of middle mile upgrades to meet this target speed will be substantial and is a barrier to digital expansion in the rural areas \cite{vglass-2}. Research on access pricing as a mechanism to bring down middle mile investment costs by reducing the RLEC's peak capacity and over-provisioning needs can therefore also help in bridging the digital divide.

\subsection{Consumers' Cost Increase}

Network operators have begun to pass some of their network costs to consumers through various penalty mechanisms (e.g., overage fees) and increasing the cost of Internet subscriptions. For instance, when Verizon announced in July 2012 that they were offering shared data plans for all new consumers and discontinuing their old plans, many consumers ended up with higher monthly bills \cite{nyt-shared}. To remain within monthly data caps, consumers are increasingly relying on usage-tracking and data compression apps (e.g., Onavo, WatchDogPro, DataWiz) \cite{datawiz} that help to avoid overage fees. Such trends are common in many parts of the world; in South Africa, for instance, consumers use ISP-provided usage-tracking tools \cite{chetty2012you} to stay within the data caps. Similarly in the U.S., research on in-home Internet usage has shown that many users are concerned about their wired Internet bills and would welcome applications for tracking their data usage and controlling bandwidth rates on in-home wired networks \cite{chetty2011my,netlimiter}. Empowering users to monitor their data usage and control their spending has led to a new area of research that considers economic incentives and human-computer interaction (HCI) aspects in a holistic manner \cite{sigchi}.

\subsection{Application Developers' Perspective}

Introducing pricing schemes that create a feedback-control loop between the client side device and network backend devices requires new mobile applications that will support such functionalities. However, most mobile platforms in use today (e.g., iOS, Android, and Windows) have different levels of platform openness.  The iOS platform for iPhones and iPads has several restrictions: it strictly specifies what kind of applications can run in the background and further prevents any access other than the standard application programming interfaces (APIs). For example, obtaining an individual application's usage and running a pricing app in the background are prohibited. By contrast, the Android and Windows platforms allow these features, e.g., introducing an API to report individual applications' usage to third-party apps.

An interesting direction to overcome these limitations is to initiate the creation of open APIs between user devices and an ISP's billing systems. For example, this can allow the user devices connected to the ISP's network to easily fetch current pricing, billing, and usage information from the network operator, while also allowing the ISP to easily test and deploy new pricing schemes through the standardized interface. Such an API would foster innovations in pricing for both consumers and providers.

Additionally, new pricing plans create an opportunity for developers to optimize their app according to changing pricing conditions. For instance, some apps that require preloading content, such as magazine apps, might time these preloading downloads so as to coincide with lower-price times, thus saving users money \cite{sigchi}. This sensitivity to price might even improve users' experience, as lower prices generally occur during times of lower congestion and higher throughput. Shifting usage so as to save money could be especially significant for video apps, as these tend to have higher usage volumes.\footnote{Some video apps cannot shift usage due to legal restrictions on caching content. However, many apps like YouTube own the rights to their video content.} Such adaptation would also require an API allowing apps to access the network prices in real time.

% Wireless ISPs' current billing systems (including 2G, 3G, and 4G) heavily depend on the RADIUS (Remote Authentication Dial In User Service) protocol, which supports centralized Authentication, Authorization, and Accounting (AAA) for users or devices to use a network service \cite{rfc2865}. In particular, RADIUS accounting \cite{rfc2886} is well suited to support usage-based pricing, since it can keep track of the usage of individual sessions belonging to each user. Interim-update messages to each session can be sent periodically to update the usage information. However, RADIUS accounting lacks support for dynamic pricing plans, which require time-of-day usage at various time scales\footnote{Note that interim update messages are sent periodically when a session joins the system, and hence, the time interval for interim updates should be kept low to support sending time-of-day usage, which may introduce significant control overhead.} Therefore, extending these protocols to support new pricing mechanisms, standardizing interfaces, and the creation of open APIs between network operators and application developers will be interesting directions for future research in this area.

%Additionally, empowering users to monitor and control of their data usage requires understanding of the intersection between the economic incentives and human-computer interaction (HCI) aspects \cite{sigchi}.

\subsection{Software/Hardware Limitations}

Wireless ISPs' current billing systems (including 2G, 3G, and 4G) heavily depend on the RADIUS (Remote Authentication Dial In User Service) protocol, which supports centralized Authentication, Authorization, and Accounting (AAA) for users or devices to use a network service \cite{rfc2865}. In particular, RADIUS accounting \cite{rfc2886} is well suited to support usage-based pricing, since it can keep track of the usage of individual sessions belonging to each user. Yet individual session lengths are often quite long, making it difficult to retrieve usage at the smaller timescales needed for dynamic pricing.

RADIUS account sessions are initiated by the Network Access Server (NAS) when a user first attempts to connect to the network: the NAS sends a user's login credentials to the RADIUS server, which compares the credentials to a secure database. The RADIUS server then \emph{authenticates} the session and \emph{authorizes} it to access different functionalities on the network. Once this session has been initiated, a start record is created in the RADIUS logs. Interim \emph{accounting} request messages can be sent periodically to update the usage information. When the end user terminates the connection, the NAS sends a stop message to the RADIUS server and a stop record is created that stores the total usage volume of that session. Since these RADIUS sessions can have very long durations, up to several hours, RADIUS logs cannot be used to calculate usage at smaller timescales.\footnote{Note that interim update messages are sent periodically when a session joins the system, and hence, the time interval for interim updates should be kept low to support sending time-of-day usage, which may introduce significant control overhead.} Moreover, the RADIUS log has no information on the type of application(s) corresponding to each session. While one session may encompass usage from multiple apps used in parallel, in some cases individual apps initiate new sessions; thus, the concept of a ``session'' cannot be definitively tied to an individual app \cite{rfc2886}.

Changing the RADIUS protocol to deliver usage estimates at a smaller time granularity would require significant overhead in both control signaling and storing RADIUS records. A perhaps easier alternative would be to record usage at the network edge, i.e., on client devices--such functionality already exists, but this approach would be vulnerable to users' deliberately falsifying the usage recorded on their device. Similarly, RADIUS logs do not contain any information on per-application usage, but client devices can easily obtain this information. Thus, application-specific pricing could also benefit from usage tracking functionalities on the end user devices. Some verification procedures could be implemented to guard against user tampering, e.g., comparing the total monthly usage measured by RADIUS servers and client devices, but would require careful design and might not be completely secure.

% Another interesting direction is the creation of an open API between user devices and an ISP's billing systems. The open API will foster innovations in pricing for both consumers and providers. For example, the user devices connected to the ISP's network can easily fetch their pricing, billing, and usage information from the network, and the ISP can also easily test and deploy new pricing schemes through the standard interface.

%\emph{Reducing App Bandwidth:} In light of the growing demand for mobile data, a natural question is whether more efficient app development can reduce apps' traffic volume, and more fundamentally whether bandwidth consumption is a consideration during app development. A small online survey of mobile app developers revealed divided opinions \cite{sigchi}. For example, one iOS developer with 4 years of experience and 12 apps told us “No, most app developers do what must be done. Only experienced app developers will look for ways to reduce bandwidth usage” (D1). Developer D2 with 5 years of experience echoed this view: “User experience and interface seem to be the main need in the App Store.” But another iOS developer with 2 years of experience and 3 apps opined that apps do consider data usage: “Bandwidth overhead is very important to the user and should thus be important to the developer too” (D3). However, he added that the “previous version of Facebook app was - because of all UIWebViews - very slow on mobile connections. Reducing the bandwith by using native UI interfaces gave a huge speed bump.” When asked whether TDP plans will hurt app development, developer D1 felt that developers will adapt to new plans, while D3 felt that it won’t have any effect: “I don't think that developers will change their design just because of changes in data plans.”
%
%The developers also provided examples of apps that can actually benefit from TDP: “applications that can cache websites and newsfeeds to read later” (D3), “Dropbox or any file sharing and backup app” (D2), and apps like YouTube that can choose “different resolutions depending on network availability” (D1). But, developer D3 opined: “The Internet’s greatest advantages above all other media is, that it's always up-to-date. It is the fastest media and adding such plans will cut off this advantage. The users want to be up-to-date.”

\subsection{Content Delivery Issues}

Any change in access pricing has to be studied in the larger context of Internet's net-neutrality and openness. These discussions center around the issues of (a) who should pay the price of congestion (i.e., content providers or consumers) and (b) how such pricing schemes should be implemented (i.e., time-of-day, app-based bundles, etc.).  The major concern with policy change is the possibility of paid prioritization of certain content providers' traffic, price discrimination across consumers, and promoting anti-competitive behavior in bundled offerings of access plus content.  While such developments can indeed hurt the network ecosystem, one aspect that should receive more attention is the threat to data usage even under simple usage-based or tiered data plans.  As Internet users become more cautious about their data consumption \cite{BostonGlobe}, content providers are providing new options to downgrade the quality of experience (QoE) for their users to help them save money.  For instance, Netflix has started allowing ``\emph{users to dial down the quality of streaming videos to avoid hitting bandwidth caps}''  \cite{Newman}. Additionally, it is ``\emph{giving its iPhone customers the option of turning off cellular access to Netflix completely and instead relying on old-fashioned Wi-Fi to deliver their movies and TV shows}'' \cite{Fitchard}.  Thus, the ecosystem today is being driven by an attitude of penalizing demand and lessening consumption through content quality degradation.

Network researchers are investigating these issues broadly along two lines of work: (i) opportunistic content caching, forwarding, and scheduling, and (ii) budget-aware online video adaptation. Opportunistic content delivery involves the smart utilization of unused resources to deliver higher QoE; for example, to alleviate the high cost of bulk data transfers, Marcon et al. \cite{marcon2012netex} proposed utilizing excess bandwidth (e.g., at times of low network traffic) to transmit low-priority data. Since this data transmission does not require additional investment from ISPs, they can offer this service at a discount, relieving data transfer costs for clients. While utilizing excess bandwidth introduces some technical issues (e.g., the potential for resource fluctuations), a prototype implementation has shown that they are not insurmountable \cite{laoutaris2011inter}. The second stream of works on online video adaptation systems, such as Quota Aware Video Adaptation (QAVA) \cite{Jiasi}, have focused on sustaining a user's QoE over time by predicting her usage behavior and leveraging the compressibility of videos to keep the user within the available data quota or her monthly budget.  The basic idea here is that the video quality can be degraded by non-noticeable amounts from the beginning of a billing cycle based on the user's predicted usage so as to avoid a sudden drop in QoE due to throttling or overage penalties when the monthly quota is exceeded. This relates to the SDP theme of enabling self-censorship of usage and QoE on the client side device through user-specified choices. %A common feature of these new research directions is that they address the concerns of consumers and content providers by accounting for both technological and economic factors.

\subsection{Regulatory Concerns}

Pricing in data networks has remained a politically charged issue, particularly for pricing mechanisms that could potentially create incentives for price discrimination, non-neutrality, and other anti-competitive behavior through app-based pricing or bundling of access and content.  Academics have already cautioned that the ongoing debate on network neutrality in the U.S. often overlooks service providers' need for flexibility in exploring different pricing regimes \cite{Yoo}:   
\begin{quote}
\emph{Restricting network providers' ability to experiment with different protocols may also reduce innovation by foreclosing applications and content that depend on a different network architecture and by dampening the price signals needed to stimulate investment in new applications and content.}
\end{quote}

But faced with the growing problem of network congestion, there has been a monumental shift in the regulatory perspective in the US and other parts of the world.  This sentiment was highlighted in FCC Chairman J. Genachowski's 1 December 2010 statement \cite{FCC}, which recognizes \emph{``the importance of business innovation to promote network investment and efficient use of networks, including measures to match price to cost."}

### 2013-07-22 Compiled/chapter3/engg.tex
all files compile correctly

\section{Engineering Aspects of SDP}\label{sec:engg}

Following the works of congestion pricing by Kelly, Gibbens, and others \cite{Kelly-eff,gibbens-kelly,kelly1998rate,kelly2000}, researchers have focused on developing the basic idea of pricing bandwidth resources according to congestion marks (i.e., price ``pollution" instead of just volume). Congestion marks on packets can be used as a mechanism to indicate to the end-users or end-systems (or ``agents") that they may need to pay more to avail a certain level of QoS when the network gets congested. A team of researchers, led by Briscoe,\footnote{http://bobbriscoe.net/pubs.html} today are developing these ideas on the accounting architecture, feedback aspects of congestion signaling, incentive compliance, security etc. in conjunction with working groups like WG CONEX in the IETF\footnote{http://datatracker.ietf.org/wg/conex/charter/} that focus on implementation aspects. One such EU collaborative project, the \emph{M3I} (Market Managed Multi-service Internet) has proposed an architecture for market management of the Internet's QoS and demonstrated new business models that it can enable. The core of the QoS problem tackled by the M3I project is to solve the fast control problem to avoid QoS degradation during short-term congestion. In doing so, the M3I architecture requires network providers to deploy ECN (Explicit Congestion Notification) on all routers so that the congestion experienced field in the IP packet header can be set with a probability related to the current network load, allowing prices to adapt to the congestion conditions in the network. The M3I solution can be realized in different ways and in different scenarios, which have been extensively discussed in \cite{M3I}. One solution uses these pricing feedback signals to encourage self-admission control. In effect this is real time pricing (RTP) version of time-of-day pricing. A second solution synthesizes admission control at the network edge from a dynamically priced wholesale service.\footnote{http://bobbriscoe.net/projects/m3i/dissem/cc02/m31\_cc02.pdf} The M3I technology has also implemented several business models over different QoS mechanisms as discussed in \cite{M3I-business}.  In a similar vein, \cite{Chiu2008} has used game-theoretic model to explore the uplink pricing as a way to provide differential pricing to P2P and regular users in a competitive market.

Like M3I, much SDP research realizes the vision of pushing control out of the network into the hands of the end-users. However, giving users more usage control has to be introduced carefully so that the users are not overwhelmed with choices and network providers do not lose revenue. Users typically prefer to have some certainty about their monthly bills and hence as opposed to real-time dynamic pricing, some SDP researchers have looked at day-ahead time-dependent pricing. But day-ahead pricing requires network providers to be able to optimize for the future prices they are willing to offer based on past usage patterns and their prediction of user elasticity of demand for various types of traffic. On the user-side, it requires developing interfaces through which users can provide their feedback to these prices.

The key subsystems required in realizing the basic feedback-control architecture envisioned in M3I \cite{M3I} and SDP are:
\begin{enumerate}
\item \emph{Tariff Communication} is used to distribute pricing policy to other subsystems. The Price Communication Protocol (aka Tariff Distribution Protocol \cite{protocol}) is a flexible protocol that can use different transport mechanisms like UDP multicast, HTTP and RSVP to distribute tariffs between the ISP's management systems and to customers. 
\item Charging, Authentication, and Accounting module applies the chosen tariff plan to measured usage data for billing purposes.
\item Price Computation calculates optimal prices to offer to the user. In M3I, this price adapts to the real time network load while in SDP the prices are computed in advance to provide guarantees to the user's expenses. 
\item Optimization and Prediction Modules are used by the provider to predict the future congestion levels and estimate the users' patience across different traffic classes at different types of traffic based on the history of traffic volumes and the observed changes to it in reaction to offered prices. The optimization module then computes the optimal prices (or discounts) to offer to the end-users in the SDP architecture.
\item Data Gathering/Usage Metering is used by the provider as inputs to the charging and accounting systems and for price calculation.
\item Mediation is used to aggregate gathered data and do format conversions as needed.
\item Client Scheduler is used in the SDP framework to create an ``autopilot" mode of operation for the end-user which schedules the various types of application traffic based on user's specified delay elasticities and offered future prices so as to minimize user expenses.
\item Charge Reaction allows customers to react to the offered prices which gets fed back into the feedback-control loop between the ISP and its customers. In some scenarios such function is performed by a software agent on the end-user device (e.g., ``autopilot" mode in SDP or the Dynamic Price Handler in M3I), while in others, as in day-ahead TDP, user interfaces need to be provided for users to react to the offered prices.  
\end{enumerate}

Next we discuss a case study of a particular realization of SDP --dynamic day-ahead time-dependent pricing; the model considerations for it, system architecture, HCI design aspects and results from a field trail.

### 2013-07-22 Compiled/chapter3/psychology.tex
all files compile correctly

\section{Psychological Aspects of SDP}\label{sec:psychology}

Arguably, the foremost factor in realizing a successful data plan is its adoptability by end users. But this depends not only on good economic models and system capabilities, but also on understanding consumer behavior and psychological aspects, and in particular, how client-side interfaces need to be designed to make them effective. These concerns require the networking researchers to interact and work closely with the ongoing efforts in the human-computer interaction (HCI) community. This section covers some of the basic themes emerging from the HCI research on psychology of Internet users.

As early as 2005, HCI researchers introduced networking, particularly user behavior considerations in home networks, as an important aspect of HCI studies \cite{Grinter2005}. In the context of broadband networks, few HCI works have developed human-facing systems to manage network usage.  The \emph{Eden} system \cite{Eden} modified a home router to provide users with an intuitive interface for managing their ``home network experience." The main focus of such home-networking tools has been on designing GUIs to help users understand the physical location of different devices in their home and be able to perform basic administrative functionalities like perform membership management, access control, network monitoring, etc. In a similar vein, the \emph{Homework project} \cite{Mortier2012} modifies the handling of protocols and services at the home router to monitor data usage, prioritize different devices, and monitor other users’ data consumption (usually in the context of parental control) in order to reflect the interactive needs of the home. Other studies have focused on understanding the impact of monitoring and sharing bandwidth speed in a wired home network \cite{chetty2011my}. The authors carried out a field trial with 10 households using a visual network probe designed to help educate and empower consumers by making broadband speeds and sources of slow-downs more visible. In a separate work, Chetty et al. \cite{chetty2010s} also carried out a field trial of their \emph{Home Watcher} bandwidth management tool to study the effect of viewing others’ bandwidth usage on social dynamics in the household. They showed that when resource contention amongst different household members is made visible, users have better understanding of bandwidth their usage and allocation, revealing new dimensions in household politics. Yet while these works addressed several crucial elements of network intervention (e.g., throttling, capping, parental control) and its related visualization tools, they have been limited to either modifying the network stack within the OS \cite{chetty2010s} or deploying a custom-built access point \cite{Mortier2012}. In contrast, the need today is to focus on understanding the following issues: (a) how economic incentives (i.e., pricing) can impact and modify mobile user behavior, (b) how researchers can carry out field trials (almost seamlessly from an end-user perspective) by interposing between the ISP and its real customers (i.e., participants).

In the context of mobile networks, \cite{Roto2006} considers the need for mechanisms to help users track their spending on mobile data. This line of research on user interface designs needs more attention to realize more dynamic pricing plans and study user response to such plans. However, realizing such data plans requires assumptions of rational behavior to be realized in practice, i.e., that people perceive the pricing signals and change their behavior. The Berkeley INDEX project \cite{INDEX} in the 1990s for wired Internet suggested that users should be able to view prices and select desired QoS levels (i.e., bandwidth speed), with similar results reported by the M3I project \cite{M3I}. 
Similar field trials on the HCI aspects of time-dependent pricing for mobile data have been reported in \cite{sigchi}, and the main themes emerging from these studies will be discussed below.

Recently, there have been calls to investigate HCI aspects of time-dependent pricing in the context of ecological sustainability and energy consumption \cite{Pierce2012}. Even without economic incentives, good UI designs of energy monitors have been shown to be effective in changing consumer behavior \cite{Froelich2010,Pierce2012}. These investigations have ranged from a large-scale media art installation visualizing energy consumption in an office building, to power strips that change color to show the energy used by individual electrical sockets \cite{Heller2011,Holmes2007}. One of the usual tradeoffs in visualizing energy usage is the use of pictorial versus numerical usage amounts. For instance, the iPhone application \emph{WattBot} allows users to monitor their home energy usage, with colors indicating usage amounts \cite{Petersen2009}. While the colors enabled users to quickly grasp their qualitative energy usage, users also wanted to track their evolving usage behavior by viewing their usage history \cite{Chetty2008}. In addition to the ``manner" of presenting usage data, users expressed concern over the ``convenience" of checking their usage. For instance, researchers testing a desktop widget that showed computer energy efficiency found that users appreciated the inconspicuous, easy-access nature of the widget \cite{Kim2010}. Many of these design insights were incorporated in the client-side UI design of the TDP trial for mobile users reported in \cite{sigchi}. Another key consideration in designing such mobile applications for smarter data plans is that the GUIs need to be account for the form-factor, presentation, and convenience of use on mobile devices and platforms. Incorporating features like parental control and usage history have been found to be desirable to users, even in the TDP paradigm \cite{sigchi,datawiz}.

The main themes emerging from these research on HCI aspects of networking are that:
\begin{enumerate}
\item Consumers are very concerned about the increasing cost of data plans but are not fully aware of their monthly usage. Therefore mechanisms that help them to better monitor and control their own usage and allocate bandwidth among household members (or in shared data plans for mobile) are perceived to be useful by users. Moreover, given the right economic incentives, many are willing to wait for some form of a feedback signal from the network on the congestion levels and change their behavior, e.g., choose higher QoS service for critical applications in smart market scenarios \cite{M3I}, or wait for discounted periods for non-critical applications in case of time-dependent data plans \cite{sigchi}. This is therefore a promising direction in the evolution of Internet pricing, and follows similar trends in other markets like electricity and transport networks. 
\item The key factor in enabling smart data plans is designing the user interfaces that are effective and understandable to users. Previous research has shown that well-designed user-side applications can not only help users make decisions on deferring usage, but also can become a tool that helps them to self-educate, monitor, and control their usage and spending. Among the features found to be effective as means of communicating useful information to users were color-coding\footnote{While color-coding has been used and appeared effective in several scenarios, care must be taken to provide a secondary signal for persons suffering from color-blindness.} of price/discounts, information indicators (e.g., current price) on the home screen of devices, usage history, recommendations on usage deferrals, etc.
\item Past research also revealed interesting insights on the user psychology of controlling their usage. First, trial participants viewed parental control at the granularity of apps as being useful in managing their usage. Second, they showed a desire to control their usage manually instead of delegating control to an autopilot mode. When coupled with the desire for parental control, we find users want to take charge of their consumption behaviors, for themselves and their families, in ways that require transparency and flexibility. This tradeoff between users' desire for transparency and control will be an important design considerations for enabling smarter data plans.
\item Consumers prefer a certain degree of certainty in the prices, which explains the popularity of flat-rate plans. Therefore for dynamic and smarter data plans to succeed, the time granularity of changes in prices have to be carefully accounted for. Users want to know and have certainty about these prices/discounts in advance, and therefore naive dynamic pricing plans may not work. In particular, real-time pricing that provides price signals based on current congestion levels may demand a higher degree of user engagement (or automation) than users are willing to adopt. Plans that have been explored in energy markets, such as ``day-ahead'' pricing \cite{ha2012tube,sigchi}, therefore may be more acceptable to users.
\item As both networking and HCI address increasingly complex socio-technical ecosystems, e.g., mobile Internet, cloud computing, smart grids, etc., incorporating economic analysis as a part of user behavior studies is important. There is also a need to understand how researchers can create new frameworks for realistic experiments and field trials on the HCI aspects of network economics. Various research ideas have taken different systems approaches, from modifying the network stack within the OS \cite{chetty2010s} to deploying custom-built access points \cite{Mortier2012}, to the TUBE project \cite{ha2012tube} which developed a system for researchers to act as a resale ISP by interposing between ISPs and their customers, and provided these two sides with client-side mobile applications and ISP-side incentive computation modules.  
\end{enumerate}

% \cite{Roto2006} 	Roto, V., Geisler, R., Kaikkonen, A., Popescu, A. and Vartiainen, E. Data traffic costs and mobile browsing user experience. In Proc. Workshop on Empowering the Mobile Web, in conjunction with WWW 2006.
%\cite{Grinter2005}	Grinter, R.E., Edwards, K. W. and Newman, M. W. The work to make a home network work. In Proc. ECSCW, Springer (2005): pp. 469 – 488.
%\cite{Eden}	Yang, J., Edwards, W. K. and Haslem, D. Eden: Supporting home network management through interactive visual tools. In Proc. UIST, ACM Press (2010): pp. 109 – 118.  
%\cite{Mortier2012} Mortier, R., Rodden, T., Tolmie, P., Lodge, T., Spencer, R., Crabtree, A., Koliousis, A. and Sventek, J. Homework: Putting interaction into the infrastructure. In Proc. UIST, ACM Press (2012): pp. 197– 206.
%\cite{Chetty2011} Chetty, M., Haslem, D., Baird, A., Ofoha, U., Sumner, B. and Grinter, R. Why is my Internet slow?: Making network speeds visible. In Proc. CHI 2011, ACM Press (2011): 1889 – 1898.
%\cite{Chetty2010} Chetty, M., Banks, R., Harper, R., Regan, T., Sellen, A., Gkantsidis, C., Karagiannis, T. and Key, P. Who’s hogging the bandwidth?: The consequences of revealing the invisible in the home. In Proc. CHI 2010, ACM Press (2010): pp. 659 – 668.
%\cite{Pierce2012} Pierce, J. and Paulos, E. Beyond energy monitors: Interaction, energy, and emerging energy systems. In Proc. CHI 2012, ACM Press (2012): 665 – 674.
%\cite{Froelich2010} Froelich, J., Findlater, L. and Landay, J. The design of eco-feedback technology. In Proc. CHI 2010, ACM Press (2010): 1999 – 2008.
%\cite{Heller2011} Heller, F. and Borchers, J. PowerSocket: Towards on-outlet power consumption visualization. In Ext. Abstracts CHI 2011, ACM Press (2011): 1981 – 1986.
%\cite{Holmes2007}Holmes, T. Eco-visualization: Combining art and technology to reduce energy consumption. In Proc. C&C 2007, ACM Press (2007): 153 – 162.
%\cite{Petersen2009} Petersen, D., Steele, J. and Wilkinson, J. WattBot: A residential electricity monitoring and feedback system. In Ext. Abstracts CHI, ACM Press (2009): 2847 – 2852.
%\cite{Chetty2008} Chetty, M., Tran, D. and Grinter, R. Getting to green: Understanding resource consumption in the home. In Proc. UbiComp 2008, ACM Press (2008): 242 – 251.
%\cite{Kim2010} Kim, T., Hong, H. and Magerko, B. Design requirements for ambient display that support sustainable lifestyle. In Proc. DIS 2010, ACM Press (2010): 103 – 112.

### 2013-07-22 Compiled/chapter6/main.tex
all files compile correctly

\newcommand{\CLASSINPUTinnersidemargin}{0.7in}
\newcommand{\CLASSINPUToutersidemargin}{0.7in}

\documentclass{article}
\usepackage{amsfonts,amsmath,amssymb}
\interdisplaylinepenalty=2500

\usepackage{graphicx}
\usepackage{multirow}
\usepackage{array}
\usepackage{fancyvrb}
\usepackage{boxedminipage}
\usepackage{color}
\usepackage{paralist}
\usepackage{algorithmic}
\usepackage{slashbox}
\usepackage{xspace}
\usepackage{url}
\usepackage{subfigure}
\usepackage{marginnote}

%\usepackage{fancybox}
\usepackage{framed}
\definecolor{shadecolor}{rgb}{1,1,0.7}

\newenvironment{codice}
{
\noindent
\begin{minipage}[b]{\columnwidth}
\definecolor{shadecolor}{rgb}{1,1,1}
\begin{framed}}
{
\vspace{-0.4cm}
\end{framed}
\end{minipage}
%\vspace{0.5cm}
\definecolor{shadecolor}{rgb}{1,1,0.7}
}

\newcommand{\ed}[1]{\textsf{\textcolor{red}{[#1]}}}
\newcommand{\remove}[1]{}

\newtheorem{problem}{Problem}
\newtheorem{property}{Property}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{observation}{Observation}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{proposition}

\begin{document}

%\title{A Tutorial on MPLS Virtual Private Networks}
\title{MPLS Virtual Private Networks}

\author{Luca Cittadini \and Giuseppe Di Battista \and Maurizio Patrignani}

\date{~}

\maketitle

\section*{Summary}
This chapter is devoted to Virtual Private Networks (VPNs) designed with Multi 
Protocol Label Switching (MPLS) \cite{rfc4364,rfc3031,rfc5036}, one of the most 
elusive protocols of the network stack. Saying that MPLS is ``elusive'' is not 
overemphasizing: starting from its arduous fitting within the ISO/OSI protocol 
stack, continuing with its entangled relationships with several other routing 
and forwarding protocols (IP, OSPF, MP-BGP, just to name a few), and ending with 
the complex technicalities involved in its configuration, MPLS defies 
classifications and challenges easy descriptions.

On the other hand, and in a seemingly contradictory way, the configuration of 
VPNs with MPLS is rather simple and elegant, despite the complexity of the 
underlying architecture. Also, MPLS flexibility and maintenance ease 
make it a powerful tool, and account for its ubiquity in Internet Service 
Providers' networks.

The chapter is organized as follows. Section~\ref{se:intro} gives a brief introduction and
motivation behind the concept of Virtual Private Network and explains why Layer 3 MPLS VPNs
are by far the most popular widespread kind of VPNs deployed today.

In Section~\ref{se:background} we introduce the reader to basic concept and terminology
about Label Switching (also known as Label Swapping) and Virtual Private Networks.

Section~\ref{se:overview} gives a high-level step-by-step description of an MPLS VPN. This is based on three main ingredients: an any-to-any IP connectivity inside the
network, a signalling mechanism to announce customer IP prefixes, and an encapsulation
mechanism, based on MPLS, to transport packets across the network.

Section~\ref{se:details} explores in detail the complex interplay between IP and MPLS that
is at the basis of MPLS VPNs.

More technical details about dynamic routing and connecting to the Internet, 
advanced usage of routing, and preserving IP-specific per-hop behavior are 
provided in Section~\ref{se:advanced}.

Strengths and limitations of MPLS VPNs are discussed in Section~\ref{sec:summary}. The
same section proposes further readings on the subject.

The reader who is interested in getting only a high-level understanding on how MPLS VPNs work can
read Sections~\ref{se:intro}, \ref{se:background}, and~\ref{se:overview}. An indepth view
of MPLS VPNs can be gained by reading Sections~\ref{se:details} and~\ref{se:advanced}.

\newpage

\tableofcontents

\newpage

%%%%%%%%%
%%%%%%%%%
%%%%%%%%%
%%%%%%%%%
\section{Virtual Private Networks}\label{se:intro}

After giving a brief introduction and motivation behind the concept of Virtual 
Private Network, this section explains why Layer 3 MPLS VPNs are by far the most 
popular widespread kind of VPNs deployed today.

\subsection{The Need for Virtual Private Networks}
The concept of Virtual Private Networks (VPNs) is essential in today's 
networks and will probably become paramount in tomorrow's networks, yet it is 
sometimes considered too advanced to be covered in a networking course. This 
apparently contrasts with the simplicity of the concept of a VPN: in its most 
generic form, a VPN is a closed (``Private'') group of nodes
that want to be connected in a network (``Network'') and are willing to use 
virtual connections, or \emph{pseudowires} (``Virtual'') instead of physical 
connections.

Such a definition captures the essence of a VPN from the perspective of the 
customer. A network provider has a slightly different abstraction about a VPN, 
mostly because she has a different interpretation of the keyword ``Network'': 
within the graph that represents her own network, she needs to provide 
connectivity to a subset of the nodes. Despite being seemingly very easy, each 
of the other two keywords that appear in the definition hides a fair amount of 
complexity that is not obvious at first glance. 
\begin{description}
 \item[Virtual] Where in the ISO/OSI stack does virtualisation happen?
 \item[Private] Is there any authentication mechanism? Does the VPN
need to preserve confidentiality of the messages?
%  \item[Network] What does the network topology look like?
% Does it support one-to-many traffic?
 \end{description}

Each of these questions has many possible answers, which is the reason why there 
are so many different types of VPNs in today's networks. For example, a 
peer-to-peer network can be seen as a VPN where pseudowires are transport 
sessions, there is no authentication amongst nodes and no traffic encryption, 
and the topology of the network is defined by a dynamic algorithm. At the 
opposite side of the spectrum we have optical networks, which can be seen as 
VPNs where pseudowires are light paths through optical switching devices, there 
is no authentication and no encryption, and the network topology is defined by 
simply configuring arbitrary pseudowires among the nodes.

In the context of VPNs, the term ``virtualisation'' indicates the technology 
that is used to multiplex traffic from multiple VPNs on the same underlying 
network topology. The most important feature of a VPN technology is what 
multiplexing technique is used and at which layer of the protocol stack. In 
general, pushing the multiplexing component down to the lower layers of the 
protocol stack (e.g., the physical or data-link layer) implies a higher 
implementation cost compared to the higher layers (e.g., the transport or 
application layer). For example, deploying an optical network to be able to run 
arbitrary pseudowires between two computers is several orders of magnitude more 
expensive than connecting those two computers to the Internet and writing a 
software that establishes a tunnel between them. On the other hand, 
multiplexing is transparent to upper layer protocols: for this reason, 
multiplexing at lower layers in the stack allows us to support a wider fraction 
of the protocol stack.

The most common layers where multiplexing happens are layer 2 and layer~3. 
A layer 2 VPN (L2VPN) transports packets of a specific layer 2 protocol and 
hence, thanks to the layered architecture of the protocol stack, is capable of 
supporting any kind of layer 3 protocol. L2VPN technologies join the nodes 
belonging to the same VPN within the same broadcast domain. For example, with a 
L2VPN, all nodes in the VPN could participate in the same VLAN and exchange 
Ethernet packets. We refer the reader to~\cite{rfc4665} for a detailed 
discussion of requirements for L2VPNs, and to~\cite{rfc4664} for a reference 
model of L2VPNs and a discussion of the main functional components.
%
Analogously, a layer 3 VPN (L3VPN) transports packets of a specific layer 3 
protocol and hence is capable of supporting any kind of layer 4 protocol. Nodes 
belonging to the same L3VPN can exchange IP packets that are routed through a 
provider network. We refer the reader to~\cite{rfc4031} for a detailed 
discussion of requirements for L3VPNs, and to~\cite{rfc4110} for a reference 
model of L3VPNs and a discussion of the main functional components.

% For example, suppose we 
% have a distributed application running on $1,000$ heterogeneous hosts which 
% exchange data using TCP sessions over the Internet. If we decide to replace 
% the Internet connection with a VPN (e.g., for security reasons), then deploying 
% an L3VPN has the advantage of being completely transparent to the hosts.

\subsection{Layer 3 VPNs and MPLS} \label{sec:intro-goals}
Layer 3 VPNs are by far the most popular of VPNs deployed today.  One reason is that layer 3 offers a good trade-off 
between deployment cost and transparency to end hosts. Another, perhaps stronger 
reason is that, as the Internet converged towards today's everything-over-IP 
scheme, it seemed natural to place the multiplexing component at the highest 
layer that supports transporting IP packets\footnote{We are recently observing a 
similar convergence trend at layer 2 with Ethernet: consequently, in the last 
few years there has been a significant increase in the demand of virtual layer 2 
services.}.

Despite a variety of technologies to realize virtual layer 3 services, most 
L3VPNs are based on the Multi Protocol Label Switching protocol (MPLS). The 
popularity of an L3VPN technology strongly depends on its ability to meet the 
demands of customers, providers, and vendors:
\begin{description}

\item[Customers' needs:]

Typical VPN customers (e.g., private companies, public administrations, etc.) have several 
geographically distributed sites and would like to have a unique IP network 
connecting all of them. Besides mere connectivity, they have other 
requirements: (i) they want to keep their own IP addressing plan for all 
the sites;  (ii) they want their traffic to be logically separated from the 
traffic of other customers that happen to use the same shared infrastructure; 
and (iii) they want guaranteed quality of service.

\item[Providers' targets:]

Providers have invested lots of resources in building their own network 
backbone. Since they have an existing infrastructure with many distributed PoPs 
(Points of Presence) connected to the backbone, they would prefer to sell 
pseudowires rather than physical connections to their customers.  Among multiple techniques to implement 
pseudowires, providers prefer those that involve lower configuration efforts, 
which usually implies lower maintenance costs. Moreover, they want the 
implementation to be scalable with respect to the number of customers: the 
amount of state to keep in the core of the network should only depend on the 
network topology, not on the number of customer VPNs.

\item[Vendors' strategies:]

No network technology can be easily deployed without meeting the strategies of 
network device producers and vendors, whose immediate aim is to sell many 
machines (possibly expensive carrier-grade routers) and, in the long run, to 
drive the shift from a variety of old technologies for VPNs (e.g., ATM or 
Frame Relay) to new technologies that are simpler to manage and hence have the 
potential to grow the vendor's market share.
\end{description}

After having introduced MPLS terminology and having given an overview of its
main building blocks, in Section~\ref{sec:summary} we discuss the extent to 
which MPLS is able to meet the above requirements.

\begin{figure}
\centering
% trim=1cm 5cm 2cm 4cm, clip=true, (left bottom right top)
%\frame{\includegraphics[trim=0cm 1cm 0cm 3cm, clip=true, width=0.7\columnwidth]{figures/mpls-slides-0}}
 \includegraphics[trim=0cm 1.5cm 0cm 1.5cm, clip=true, width=0.7\columnwidth]{figures/mpls-slides-0}
 \caption{The sample network used throughout this chapter.}
 \label{fig:mpls-slides-0}
\end{figure}

\begin{shaded}
\noindent
Throughout this chapter we will refer to a very simple scenario (see
Fig.~\ref{fig:mpls-slides-0}) where a provider has a network infrastructure with three
PoPs (in Turin, Milan, and Rome) and offers connectivity to two customers. Customer 1 has
two sites and has an IP addressing plan that allocates the $200.1.6.0/24$ to its site
in Milan and the $10.0.0.0/24$ to its site in Rome. Customer 2 has two sites too
and has an IP addressing plan that allocates the $10.0.0.0/24$ to its site in Turin
and the $193.1.1.0/24$ to its site in Rome. Observe that the two customers have overlapping IP address space.

We will use the sample scenario to illustrate most of the concepts introduced in 
this chapter. The text that describes and refers to the sample scenario will be 
framed into shaded boxes like the one that encloses this paragraph. The configuration
language is that of a leading router vendor and references can be found
in~\cite{de-ghein}.
\end{shaded}
%%%%%%%%%
%%%%%%%%%
%%%%%%%%%
%%%%%%%%%
\section{Background and Terminology}\label{se:background}

In this section we introduce the reader to basic concept and terminology about 
Label Switching (also known as Label Swapping) and Virtual Private Networks.

Throughout the chapter, we extensively refer to two tightly related yet 
distinct concepts: forwarding and routing. \emph{Forwarding} is the process of receiving 
a packet from a network interface and deciding on which interface that packet 
should be sent. Usually, in order to minimize the latency of traversing a 
router, the decision about where to forward a packet is taken based on some 
pre-computed data structure. 
This is usually referred to as the \emph{forwarding table} because a table is 
the simplest logical structure to accommodate forwarding information. A table 
can be used as a physical data structure if addresses can be matched exactly. 
However, IP addresses must be forwarded based on the longest matching 
prefix~\cite{rfc4632}. This implies that efficient IP lookups need a more 
sophisticated data structure than a table. We refer the reader 
to~\cite{varghese} for a discussion on various data structures and algorithms to 
speed up prefix-match lookups, which exceeds the scope of this chapter. In the 
following, we simply refer to the logical forwarding table, irrespective of the 
actual physical data structure.

\emph{Routing} is the process by which each router builds its forwarding table 
and adapts it as the network topology changes over time.

Correspondingly, we have \emph{forwarding and routing protocols}, where the 
formers describe the formatting rules for network packets and the conventions 
that routers and hosts have to follow in order to exchange them, while the 
latter describe packet formats and conventions used to exchange routing 
information among routers. The information that routing protocols provide is 
used by each router to populate its forwarding table.

Finally, standard network terminology distinguishes between the corresponding 
router's software layers. Namely, the layer where the forwarding process takes 
place is called \emph{data plane} or \emph{forwarding plane}, while the layer 
where the routing process is managed is called \emph{control plane}.

%
%%%
%%%%%
%%%
%
\subsection{Label Switching}
In this section we introduce the concept of label switching as a 
forwarding paradigm. After having described the fundamentals characteristics of 
label switching in general, we move to MPLS-specific details in the following 
sections.
Traditionally, there are two different approaches to packet forwarding, each mapping to a
specific structure of the forwarding table. They are called \emph{forwarding by network
address} and \emph{label switching}.

The most intuitive approach is \emph{forwarding by network address}, that is the 
approach of IP. When a packet arrives at a router, the router parses the 
destination address from the packet header and looks it up in its forwarding 
table. The forwarding table has a simple 2-column structure where each row maps 
a destination address to the egress interface that the packet should be 
forwarded to (see Table~\ref{tab:fib-destination}). For scalability and 
efficiency reasons, it is possible to aggregate several destination prefixes 
into a single row, provided that they can be numerically aggregated and that 
they share the same egress interface.

\begin{table}
 \centering
 \begin{tabular}{|c|c|}
 \hline
 \textbf{Destination address} & \textbf{Egress interface} \\
 \hline
 10.100.100.0/24 & en2 \\
 \hline
 10.100.200.128/25 & en1 \\
 \hline  
 22.30.100.0/24 & en2 \\
 \hline \end{tabular}
 \caption{Structure of the forwarding table in the ``forwarding by network 
address'' approach.}
 \label{tab:fib-destination}
\end{table}

An alternative approach is known as \emph{label switching}\remove{, that is the approach of MPLS}. 
Essentially, while forwarding by network address requires that the egress 
interface be chosen based on the \emph{destination} of the packet, label 
switching requires that such an interface be chosen based on the \emph{flow} 
the packet belongs to, where a flow corresponds to an instance of transmission, i.e., a set of
packets, from a source to a destination and is identified by a tag (called \emph{label})
attached to each packet of the flow.

\begin{figure}
\centering  % left bottom right top
 \includegraphics[trim=1cm 7cm 2cm 4cm, clip=true,width=0.7\columnwidth]{figures/reti-slides-0}
% \frame{\includegraphics[trim=1cm 7cm 2cm 4cm, clip=true,width=0.7\columnwidth]{figures/reti-slides-0}}
 \caption{A label switching network (where labels are not swapped at each hop).}
 \label{fig:reti-slides-0}
\end{figure}

\begin{figure}
\centering
 \includegraphics[trim=1cm 7cm 2cm 4cm, clip=true,width=0.7\columnwidth]{figures/reti-slides-1}
 \caption{A label switching network (where labels are swapped at each hop).}
 \label{fig:reti-slides-1}
\end{figure}

\begin{table}
 \centering
 \begin{tabular}{|c|c|c|c|}
 \hline
 \textbf{Incoming interface} & \textbf{Incoming label} & \textbf{Egress 
interface} &  \textbf{Egress label} \\
 \hline
 en2 & 101 & en5 & 218 \\
 \hline
 \end{tabular}
 \caption{Structure of the forwarding table in the ``forwarding by label 
swapping'' approach with per-interface label scope.}
 \label{tab:fib-label-interface}
\end{table}

\begin{table}
 \centering
 \begin{tabular}{|c|c|c|}
 \hline
 \textbf{Incoming label} & \textbf{Egress interface} &  \textbf{Egress label} \\
 \hline
 101 & en5 & 218 \\
 \hline
 \end{tabular}
 \caption{Structure of the forwarding table in the ``forwarding by label 
swapping'' approach with per-router label scope.}
 \label{tab:fib-label-router}
\end{table}

As an example, Fig.~\ref{fig:reti-slides-0} shows a label switching network 
where each flow has an associated label (labels are represented with colors). 
Packets with a red label belong to the flow from router A to router B of Fig.~\ref{fig:reti-slides-0}.
If a packet with a red label enters interface 1 of router E, it will exit from interface 2 with the same label.

If the label switching technologies 
followed the approach of Fig.~\ref{fig:reti-slides-0} they would have the 
advantage that labels do not have to be changed at each hop. On the other hand, if they 
did they would have the big drawback of 
requiring a centralized control of the assigned labels, as labels should be unique for the entire network. 
Fig.~\ref{fig:reti-slides-1} shows what actually happens in label switching 
networks, where labels are swapped at each hop. This choice requires that labels 
are unique for each router or for each interface only and does not need a 
centralized control. Fig.~\ref{fig:reti-slides-2} illustrates the forwarding 
table of a router. Independent of whether labels are swapped at 
each hop or not, the forwarding paths towards the same destination node 
typically form a tree rooted at the destination node itself, even though this is 
not mandatory.

More formally, the operations performed by a label switching router can be summarized as follows.
When the packet arrives at the router, the router extracts (\emph{pops}) the label 
from the header, looks the label value up in its forwarding table, and finds
\begin{inparaenum}[(i)]
 \item the egress interface the packet should be forwarded to, and
 \item a new label to apply (\emph{push}) to the packet.
\end{inparaenum}

A forwarding process based on labels rather than destination addresses poses 
challenges to the corresponding routing protocols. In fact, the instances of 
flow traversing the network might be much more volatile than the addressing scheme 
used to identify their destinations. Before transmitting a new flow, a route 
from its source to its destination has to be computed and a new label has to be 
assigned to each leg of the route. As observed before, in order to facilitate the task of
picking a 
new, unused, label, labels are not required to be unique for the entire network but are
required to be unique for each router or for 
each interface only. This is why they have to be changed at each hop.      %
Depending on whether labels have a per-interface or per-router scope, the 
forwarding table is structured as in Table~\ref{tab:fib-label-interface} or 
Table~\ref{tab:fib-label-router}, respectively. 
Observe that such a simple structure for forwarding tables 
allows efficient lookups (e.g., by using a hash table or a direct access table).

Label switching is not a unique feature of MPLS and it is not necessarily 
implemented at the network level of the protocol stack: other protocols, notably 
ATM and Frame Relay, traditionally adopt the same forwarding mechanism. 
Initially, the reason to prefer label switching was performance: looking up a 
label value in the forwarding table was much faster than looking up an IP 
address. Besides the fact that labels can take values in a much smaller range 
than IP addresses, label values can be looked up exactly, while IP addresses 
need to be looked up by the longest matching prefix. However, modern routers use 
extremely specialized hardware (e.g., content-addressable memories) and very 
efficient data structures (e.g., tries) to implement their forwarding tables, in 
such a way that the performance gain of label switching over forwarding by 
destination address is now believed to be no longer an argument.

\begin{figure}
\centering
% \includegraphics[width=0.7\columnwidth]{figures/reti-slides-2}
% left bottom right top
% \frame{\includegraphics[trim=1cm 5cm 2cm 4cm, clip=true,width=0.7\columnwidth]{figures/reti-slides-2}}
 \includegraphics[trim=1cm 5cm 2cm 4cm, clip=true,width=0.7\columnwidth]{figures/reti-slides-2}
 \caption{A label switching network where the forwarding table of a router is shown.}
 \label{fig:reti-slides-2}
\end{figure}

% A few points have to be observed with respect to the relationship among a label 
% switching protocol and other protocols of the same network:
% \begin{itemize}
% 
% \item While, in theory, it would be possibile to fill in forwarding tables 
% manually, any practical use of the label switching approach computes them 
% automatically. The easiest strategy is that of following, hop by hop, the 
% routes computed by some other network protocol, usually with a forwarding by 
% destination address approach. This is why a label switching protocol usually 
% does not replace more traditional protocols, but rather relies on them. MPLS, 
% for example, uses IP routing tables to compute a route from source to 
% destination. 
% \item Traffic in a label switching network may be heterogeneous: labeled and 
% non-labeled packets may be received at the interfaces and forwarded based on 
% their first header. This is particularly important in view of the previous 
% observation. 
%Hence, when the header containing the label is placed directly 
% after the level 2 packet, as it is the case for MPLS, from the point of view 
% of 
% the forwarding process it may be considered as an alternative network 
% protocol. 
% \end{itemize}

%
%%%
%%%%%
%%%
%
\subsection{MPLS header and terminology}

The MPLS protocol brings the label switching forwarding paradigm to the extreme,
managing, instead of a single label, a whole stack of labels, where the external one
determines the egress interface.
It does not fit the ISO/OSI model very well. The MPLS header is transported 
over L2 packets and can encapsulate L3 packets as well as L2 packets. 
% However, it relies 
% on information from L3 to build the router's forwarding tables. 
Since MPLS does 
not fit the definition of either L2 protocols nor L3 protocols, it is frequently 
referred to as a ``layer 2.5'' protocol, emphasizing the fact that it requires 
L2 connectivity and can encapsulate IP packets.

\begin{figure}
 \centering
 \begin{tabular}{l |c|}
 \cline{2-2}
  Layer 3 & IP \\ \cline{2-2}
  Layer 2.5 & MPLS \\ \cline{2-2}
  Layer 2 & Ethernet, Frame relay, ATM, PPP, etc \\ \cline{2-2}
  Layer 1 & Physical layer \\ \cline{2-2}	
 \end{tabular}
 \caption{MPLS and ISO/OSI network layers.}
 \label{fig:mpls-iso-osi}
\end{figure}

When an IP packet from a 
router needs to be transported over an MPLS 
backbone, the first MPLS-enabled router in the network pushes an MPLS header in 
between the Ethernet header and the IP header. The resulting packet layout is 
depicted in Fig.~\ref{fig:mpls-iso-osi}.

The MPLS header consists of a \emph{stack} of 4-byte records where each record 
has the following structure (depicted in Fig.~\ref{fig:mpls-header}): 
\begin{itemize}
 \item a \textbf{label} field (20 bits), which carries the label value;
 \item a \textbf{ToS} field (3 bits) which is used to discriminate different 
levels of quality of service (QoS) and to carry explicit congestion 
notifications (ECN);
 \item a \textbf{bottom-of-stack} field (1 bit) which is set to 1 when the 
record is the last record in the stack; and
 \item a \textbf{TTL} field (8 bits) which is decremented at each hop, 
similarly to the TTL field in the IP header.
\end{itemize}

\begin{figure}
 \centering
 \includegraphics[width=0.7\columnwidth]{figures/mpls-header}
 \caption{Structure of a record in an MPLS header.}
 \label{fig:mpls-header}
\end{figure}

\begin{figure}[tb]
\centering
 %\frame{
 \includegraphics[trim=0cm 7cm 0cm 7cm, clip=true,
width=0.9\columnwidth]{figures/mpls-slides-28}
 %}
 \caption{The evolution of the MPLS label stack as a packet traverses several routers that perfom random push and pop operations.}
 \label{fig:mpls-slides-28}
\end{figure}

When an MPLS-enabled router receives a packet, it can perform three different 
operations:
\begin{inparaenum}[(i)]
 \item \emph{push} a label onto a (possibly empty) stack,
 \item \emph{pop} a label from the stack (possibly resulting in an empty 
stack), or
 \item \emph{swap} the top label of the stack, which can be seen as a pop 
operation followed by a push operation.
\end{inparaenum}
% 
Figure~\ref{fig:mpls-slides-28} shows the evolution of the MPLS label stack as a packet traverses several routers that perform  random push and pop operations.

MPLS-VPN terminology uses specific names to distinguish routers that do not understand 
labels at all, routers that push (or pop) labels, and routers that simply swap 
labels. 
%
Routers belonging to the first group are called \emph{customer edge} (CE) 
routers because they are not MPLS-enabled. Typically those are the customer's 
routers that need to be interconnected via an L3VPN. CE routers can only handle 
IP packets and are not aware of the MPLS layer which is used to implement the 
VPN.

Routers belonging to the second group are called \emph{provider edge} (PE) 
routers, or \emph{label edge routers} (LERs). They are placed at the edge of the 
MPLS backbone of the provider, have direct connectivity to the CE routers, and 
act as the access point for the customer to the VPN. While they need to perform label swap operations because they are part 
of the backbone, they spend most of their time pushing labels (when an IP packet 
comes from a CE router) and popping labels (when an MPLS packet needs to be 
forwarded to a CE router).

Routers belonging to the third group are called \emph{provider} (P) routers, or 
\emph{label switching routers} (LSRs). They are in the core of the MPLS network. 
Since they do not interact directly with non-MPLS routers, they mainly perform 
label swapping operations in order to forward packets to other P or PE routers.

MPLS groups destinations into Forwarding Equivalence Classes (FEC). 
Packets that need to be forwarded to the same CE using the same path and with 
the same quality of service belong to the same FEC.

\begin{figure}
\centering
 \includegraphics[trim=0cm 1.5cm 0cm 1.5cm, clip=true, width=0.7\columnwidth]{figures/mpls-slides-1}
 \caption{Inside the provider's infrastructure.}
 \label{fig:mpls-slides-1}
\end{figure}

% \begin{figure}
% \centering
%  \includegraphics[width=0.7\columnwidth]{figures/mpls-slides-2}
%  \caption{CE -- Customer Edge Routers.}
%  \label{fig:mpls-slides-2}
% \end{figure}
% 
% \begin{figure}
% \centering
%  \includegraphics[width=0.7\columnwidth]{figures/mpls-slides-3}
%  \caption{PE -- Provider Edge Routers.}
%  \label{fig:mpls-slides-3}
% \end{figure}
% 
% \begin{figure}
% \centering
%  \includegraphics[width=0.7\columnwidth]{figures/mpls-slides-4}
%  \caption{P -- Provider Routers.}
%  \label{fig:mpls-slides-4}
% \end{figure}

\begin{shaded}
\noindent
Fig.~\ref{fig:mpls-slides-1} shows some details of the provider's infrastructure of our
scenario. It is
both an MPLS network and an IP network (it has an MPLS data plane and an IP data
plane).

If we look at it from the MPLS point of view, we can distinguish CE, PE, and P 
routers. The small, red routers placed at the customer premise in the corners 
of Fig.~\ref{fig:mpls-slides-1} are CE routers. CE routers are directly 
attached to the blue routers at the edge of the provider premise, which are the 
PE routers (or LERs). Finally, the grey routers in the core of the provider 
network are the P routers (or LSRs).
% The red arrows in Figs.~\ref{fig:mpls-slides-2}, \ref{fig:mpls-slides-3},
% and~\ref{fig:mpls-slides-4} show the CE routers, the PE routers, and the P-routers of
% our
% example network, respectively. Also, routers are labeled, according to the MPLS
% terminology, as LERs or LSRs.

Since the provider network is also an IP network an IP address is given to the interfaces.
To do this, our provider exploits prefix $80.0.0.0/8$. This prefix will not be announced
outside the provider's network. The reason for the presence of label AS100
in the provider network will be explained soon.

The two CE routers serving Customer $1$ are connected through a VPN called VPN1 
(on the right side of Fig.~\ref{fig:mpls-slides-1}). The two CE routers serving 
Customer $2$ are connected through a VPN called VPN2 (on the left side). 
\end{shaded}

%%%%%%%%%
%%%%%%%%%
%%%%%%%%%
%%%%%%%%%
\section{Checkmate VPNs in Three Moves}\label{se:overview}

In this section we give a high-level description of an MPLS VPN. Such a description is
based on three main ingredients that we call ``moves''. We claim that a reader that
understands these three moves will be able to checkmate this complex matter.

From the perspective of the customer, an MPLS VPN 
%is nothing more than a cloud which is transparent to IP packets: as if the customer's CE routers were 
%connected by a \emph{pseudowire} which traverses the cloud. 
simply routes IP packets among customers CE routers, as if they were connected by a \emph{pseudowire}.
Observe that, as customers may have overlapping address spaces, their packets cannot be simply routed through the provider network. Instead, they need to be encapsulated.  
It is tempting to implement such a pseudowire using a tunnel (e.g., GRE, IP-in-IP or IPSec) 
between PE routers where the customer packets travel across the provider network 
\emph{encapsulated} into IP or IPSec packets. However, as the number of 
interconnected sites grows, manually managing configured tunnels and maintaining 
forwarding tables becomes excessively complex. For example, if we were to use 
tunnels to implement an L3VPN over 5 customer sites, a full-mesh topology would 
translate to 20 manually configured tunnels. Moreover, if the customer adds a 
new subnet to one of its sites, we need to update the forwarding tables of all 
our 5 PE routers. Observe that the complexity
of managing tunnels can be eased by automatic setup mechanisms. However, such 
mechanisms are out of the scope of this chapter, therefore we refer the 
interested reader to~\cite{rfc4023,rfc5512,cisco-preso}.

The intrinsic problem with tunnels is that they rely on a pre-determined 
endpoint which is configured at tunnel setup time. Ideally, we would like to 
take advantage of the benefits of encapsulation without dealing with the 
issue of knowing the tunnel endpoint in advance. Namely, we would like packets 
to be encapsulated at the ingress PE and decapsulated at egress PE. We can 
split this goal into three high-level steps that we call moves:

\begin{enumerate}
\item[Move 1:] Achieve any-to-any IP connectivity among PEs,
\item[Move 2:] Define a signalling mechanism to distribute customer prefixes among PEs, 
and
\item[Move 3:] Define an encapsulation mechanism to transport packets from one 
PE 
to another across the network.
\end{enumerate}

One of the key benefits of using encapsulation (Move 3) is that the complexity 
of configuring L3VPNs for customers is confined to PEs. The core of the network 
(i.e., P routers) does not need to know anything about customer prefixes: it 
simply needs to know how to transport packets from one PE to another (Move 1). 
This means that the size of the forwarding table of P routers depends on the 
number of PE routers rather than on the number of customer prefixes. Finally, if PE 
routers use a signalling mechanism to dynamically synchronize the list of 
customer prefixes, the only pieces of information that need to be manually 
configured at each PE are the L3VPN identifier and the IP address of the CE 
router.

In the following we elaborate each move in more detail.

\subsection{Move 1 -- Any-to-any IP connectivity among PEs}
The first move is actually quite simple. It is nothing more than what any 
Internal Gateway Protocol (IGP) is designed to achieve: seamless, redundant and 
dynamic IP-level any-to-any connectivity. Since PEs are our encapsulation 
endpoints, we want them to be reachable independent of the availability of 
specific network interfaces. In other words, we do not want to use the IP 
address of physical interfaces for PEs, but loopback addresses. A loopback address is an address associated with a virtual 
interface of the router. Since it is virtual, a loopback interface it is active 
independent of the status of physical network interfaces. To fulfill Move 1, we 
simply assign a loopback address to each PE router and use an IGP (e.g. OSPF or 
IS-IS) 
%to ensure any-to-any connectivity among loopback addresses.
to announce these addresses as /32 prefixes in order to ensure any-to-any connectivity among them.

\begin{figure}
\centering
 \includegraphics[trim=0cm 1.5cm 0cm 1.5cm, clip=true, width=0.7\columnwidth]{figures/mpls-slides-5}
 \caption{Loopbacks of PEs.}
 \label{fig:mpls-slides-5}
\end{figure}

\begin{figure}
\centering
 \includegraphics[trim=0cm 1.5cm 0cm 1.5cm, clip=true, width=0.7\columnwidth]{figures/mpls-slides-8}
 \caption{IP connectivity for LER1.}
 \label{fig:mpls-slides-8}
\end{figure}

\begin{figure}
\centering
 \includegraphics[trim=0cm 1.5cm 0cm 1.5cm, clip=true, width=0.7\columnwidth]{figures/mpls-slides-9}
 \caption{IP connectivity for LSR2.}
 \label{fig:mpls-slides-9}
\end{figure}

\begin{figure}
\centering
 \includegraphics[trim=0cm 1.5cm 0cm 1.5cm, clip=true, width=0.7\columnwidth]{figures/mpls-slides-10}
 \caption{The OSPF weight of a link.}
 \label{fig:mpls-slides-10}
\end{figure}

\begin{shaded}
\noindent
Fig.~\ref{fig:mpls-slides-5}
shows the loopback addresses assigned to the PEs of our example network. Also, we assume
that routers use OSPF to propagate reachability information of loopbacks of
routers.

Configuring routers to fulfill Move 1 is straightforward. In our sample network, 
the configuration of LER1 for Move 1 is as follows.
% (throughout the chapter we show
% configuration snippets in a Cisco-like language; other vendors' languages accept very
% similar constructs).

\begin{codice}
\begin{verbatim}
  interface Loopback0
    ip address 80.80.80.1 255.255.255.255
  interface GigabitEthernet1/0
    ip address 80.1.1.1 255.255.255.252
  router ospf 10
    network 80.0.0.0 0.255.255.255 area 0
\end{verbatim}
\end{codice}

The first two lines assign an IP address to interface \texttt{loopback0}. The second pair
of lines assign an IP address to interface \texttt{GigabitEthernet1/0} that connects LER1
with LSR1. The last two lines activate OSPF protocol.

Fig.~\ref{fig:mpls-slides-8} shows the result of command {\tt show ip route} performed on
router LER1. Fig.~\ref{fig:mpls-slides-9} shows the result of command {\tt show ip route} 
performed on router LSR2. Command {\tt show ip route} has the effect of showing the
control plane routing table of routers.

In order to force a more interesting routing in
the following part of the example, we set OSPF weight $500$ for a specific link,
discouraging the use of that link by the IGP routing
protocol, as shown in Fig.~\ref{fig:mpls-slides-10}. 
\end{shaded}

\subsection{Move 2 -- Use BGP to distribute customer prefixes}
In order to distribute reachability information about customer prefixes, MPLS 
relies on a variant of BGP called Multi-Protocol BGP (MP-BGP)\cite{rfc2858}. 
Whereas BGP advertises reachability information for IPv4 addresses only, MP-BGP 
supports multiple \emph{address families} (e.g., IPv4 and IPv6). Since 
advertising VPN addresses implies exchanging not only IPv4 prefixes, but also 
additional information to identify the VPN, MP-BGP treats VPNs as a separate 
address family. PE routers establish a full-mesh of iBGP peerings and each PE 
announces to all the other PEs the customer prefixes that it can reach via the 
CE router it is connected to. The Multi-Protocol extension to BGP is needed to 
introduce the concept of the ``customer'' (i.e., the ``L3VPN identifier'') which 
does not exist in plain BGP.

Compared with any ad-hoc signalling mechanism that could have been designed 
specifically for MPLS, the choice of using BGP has the advantage of relying on a 
well-known protocol and thus making the learning curve smoother for 
practitioners. Moreover, BGP has built-in mechanisms (e.g., route reflection) to 
scale as the number of PE routers increases.

\begin{figure}
\centering
 \includegraphics[trim=0cm 1.5cm 0cm 1.5cm, clip=true, width=0.7\columnwidth]{figures/mpls-slides-6}
 \caption{Use of BGP to distribute customer prefixes.}
 \label{fig:mpls-slides-6}
\end{figure}

\begin{shaded}
\noindent
Fig.~\ref{fig:mpls-slides-6} shows a high-level illustration 
of how the BGP peerings with LER3 and LER2 can be used by LER1 to announce 
customer prefixes.
\end{shaded}

\vspace{0.1cm}

\begin{shaded}
\noindent
Configuring MP-BGP peerings is very similar to configuring plain iBGP
peerings. 
Consider the following snippet from the configuration of router LER1:
\begin{codice}
\begin{verbatim}
  router bgp 100
    neighbor 80.80.80.4 remote-as 100
    neighbor 80.80.80.4 update-source Loopback0
    neighbor 80.80.80.5 remote-as 100
    neighbor 80.80.80.5 update-source Loopback0
    !
    address-family vpnv4
      neighbor 80.80.80.4 activate
      neighbor 80.80.80.5 activate
    exit-address-family
\end{verbatim}
\end{codice}

The first line starts the BGP configuration and states that the router belongs to AS100.
Observe that all the routers are supposed to belong to Autonomous System (AS) 100. This
AS number will not be necessarily propagated outside the provider's network and is only
needed to establish peerings between PEs.

The following lines specify the BGP peerings. The presence of the ``\texttt{vpnv4}''
address family identifies LER2 and LER3 as MP-BGP neighbors of LER1.
\end{shaded}

\subsection{Move 3 -- Use MPLS encapsulation among PEs}
Having performed Move 1 and Move 2, a PE router $r$ is able to select the PE 
router $r'$ that is connected to a given customer prefix (by Move 2). Also, $r$ is 
able to forward IP packets to $r'$ (by Move 1). The only piece missing is an 
encapsulation mechanism to transport IP packets from $r$ to $r'$. One such 
encapsulation mechanism is MPLS: the PE router $r$ encapsulates the IP packet by 
pushing two MPLS labels. The label at the top of the stack (\emph{outer} label) 
is switched by P routers in order to deliver the packet to router $r'$. The 
label at the bottom of the stack (\emph{inner} label) is left untouched and it 
is used by the egress PE $r'$ to identify the correct L3VPN. Observe that the 
inner label is necessary because $r$ and $r'$ could serve a variety of 
customers, and address spaces might be overlapping. For instance, routers $r$ 
and $r'$ could be serving two distinct VPNs for two customers, both using 
addresses in the RFC 1918 space.

Let us briefly recap how a packet is delivered across an MPLS cloud. When PE 
router $r$ receives a packet from a CE router, it picks the VPN identifier and 
the destination address and, based on information contained in its MP-BGP 
Routing Information Base (RIB), it finds the PE router the packet should be 
delivered to. The MP-BGP RIB also contains the inner label that should be used. 
In our running example, the egress PE router is $r'$. Then, $r$ pushes the inner 
MPLS label and an outer label which is guaranteed to deliver the packet to 
$r'$.

How does $r$ pick this outer label? The outer label that maps to router $r'$ is 
determined by the Label Forwarding Information Base (LFIB) of $r$, which is the 
forwarding table for MPLS.

The task of distributing labels and maintaining the LFIB of label switch routers 
is performed by the Label Distribution Protocol 
(LDP)\cite{rfc5036}\footnote{Alternative protocols such as RSVP and BGP can also 
serve the same purpose, but are out of the scope of this chapter.}. LDP is able 
to setup a Label Switch Path (LSP) from one PE to another. In its simplest form, 
LDP maps an address prefix FEC (i.e., a FEC which represents an IP prefix) to a 
label. Each router receives LDP mappings from all its neighbors. In order to 
populate the LFIB, each router inspects its own forwarding table to determine 
the IP nexthop for the FEC, and picks the corresponding label. This way, LDP 
effectively creates a forwarding tree rooted at the egress point of the FEC, by 
simply importing the nexthop from the IP data plane (remember Move 1) at each 
intermediate hop.

\begin{shaded}
\noindent
It is extremely simple to configure a router to fulfill Move 3, because the LDP 
protocol can be safely run in the default configuration, and enabling MPLS 
encapsulation on specific interfaces is a single command. The configuration of 
LER1 for Move 3 is as simple as the following.

\begin{codice}
\begin{verbatim}
mpls label protocol ldp
interface GigabitEthernet1/0
  ip address 80.1.1.1 255.255.255.252
  mpls ip
\end{verbatim}
\end{codice}
\end{shaded}
%%%%%%%%%
%%%%%%%%%
%%%%%%%%%
%%%%%%%%%
%%%%%%%%%
\section{An In-Depth View of MPLS VPNs}\label{se:details}

We have seen that the architecture of MPLS VPNs builds upon three building 
blocks: a working IP data plane that is capable of interconnecting the loopback 
addresses of PE routers, a BGP-based control plane to distribute reachability 
information about customer prefixes, and MPLS encapsulation among PEs.

While the first building blocks might seem straightforward at a first glance, 
there are a number of details which complicate the big picture but 
nevertheless are important in order to grasp the internals of MPLS 
VPNs.

%
%%%
%%%%%
%%%
%
\subsection{IP Data Plane}
IP connectivity between PE routers is easy to achieve using any suitable routing 
protocol. However,
PE routers might be attached to a number of CEs of different 
customers, and must ensure that each CE is mapped to the correct VPN. A 
traditional IP data plane is unfit for this purpose since the IP address space 
of customers can overlap. Hence, a PE router must be able to route packets based 
on both the IP address and the specific VPN the packet belongs to. To accomplish 
this task, MPLS VPNs exploit a technique called Virtual Routing and Forwarding 
(VRF) which allows a router to have multiple (virtual) routing tables, 
potentially a separate virtual routing table for each network interface (either 
physical or logical). With this technique, mapping a CE to the correct VPN is as 
easy as configuring the corresponding interface within a specific VRF table. 
An MPLS inner label actually identifies a VRF instance.

One way to implement VRF while still maintaining a single forwarding table is 
using the ingress interface as an additional input parameter in the forwarding 
table. In such an implementation, the organization of the forwarding table of a 
router would be the one illustrated in Table~\ref{tab:fib-destination-vrf}. 
Observe that only the PE router needs to support VRF. The CE router is 
configured with a plain eBGP configuration and is completely unaware of the VRF 
implemented on the provider side.

\begin{table}
 \centering
 \begin{tabular}{|c|c|c|c|}
 \hline
 \textbf{VRF id} & \textbf{Ingress interface} & \textbf{Destination address} & 
\textbf{Egress interface} \\
 \hline
 VRF-3 & en5 & 10.100.200.32 & en2 \\
 \hline
 \end{tabular}
 \caption{Structure of the forwarding table in the ``forwarding by network 
address'' approach with Virtual Routing and Forwarding (VRF).}
 \label{tab:fib-destination-vrf}
\end{table}

\begin{shaded}
\noindent
Assigning an interface to a specific VRF instance is straightforward. In our 
sample network, we configure router LER1 as follows.\\

\begin{codice}
\begin{verbatim}
  interface GigabitEthernet2/0
    ip vrf forwarding VPN1
    ip address 10.0.0.1 255.255.255.0
  interface GigabitEthernet3/0
    ip vrf forwarding VPN2
    ip address 193.1.1.1 255.255.255.0
\end{verbatim}
\end{codice}

Address \texttt{10.0.0.1} is the address assigned to the interface that connects LER1
to Customer 1, while address \texttt{193.1.1.1} is assigned to the interface that
connects LER1 to Customer 2.
\end{shaded}

%
%%%
%%%%%
%%%
%
\subsection{MP-BGP Control Plane}
Multi-protocol extensions to BGP allow us to segregate each L3VPN in a different 
\emph{namespace}, identified by a proper VPN identifier. Separating namespaces 
is important because an IP prefix is not guaranteed to be unique across multiple 
VPNs: in practice, customers might want to use their own private IP address 
spaces, possibly overlapping with the address space of other customers. MP-BGP 
solves this issue by introducing the concept of VPN-IP addresses, that is, IP 
addresses tagged with a 8-byte VPN identifier which is called \emph{route 
distinguisher} (RD). A VPN-IP address is nothing more than the concatenation of 
the RD and the IP prefix. By imposing that different VPNs be assigned distinct 
RD values, the uniqueness of VPN-IP addresses is guaranteed even in the presence 
of overlapping IP address space among customers. A special RD value consisting 
of 8 NULL bytes represent the default VPN, which allows MP-BGP to distribute 
information regarding pure IP routes alongside information about VPN-IP prefixes.

Observe that, while a VPN-IP prefix uniquely identifies a destination, it 
provides no information about the reachability of that destination. For this 
reason, MP-BGP messages associate VPN-IP prefixes with the MPLS labels that 
should be used for forwarding.

\begin{figure}
\centering
 \includegraphics[trim=0cm 1.5cm 0cm 1.5cm, clip=true, width=0.7\columnwidth]{figures/mpls-slides-12}
 \caption{How MP-BGP can distribute per-VPN reachability information. Observe that the Local Preference attribute is not meaningful for locally originated prefixes, which are automatically preferred by the BGP decision process.}
 \label{fig:mpls-slides-12}
\end{figure}

\begin{shaded}
\noindent
It is easy to assign an RD value to a single VRF instance:\\

\begin{codice}
\begin{verbatim}
ip vrf VPN1
  rd 100:11
ip vrf VPN2
  rd 100:22
\end{verbatim}
\end{codice}

Fig.~\ref{fig:mpls-slides-12} shows the output of command {\tt show ip bgp vpnv4 
all} on router LER1. This command has the same effect of {\tt show ip bgp} but 
it shows the routing entries related to IPv4 VPNs. In this case the output highlights 
that LER1, in addition to its locally originated prefixes 10.0.0.0/24 with Route Distinguisher 100:1 and 193.1.1.0/24 with Route Distinguisher 100:2, knows two remote prefixes. Namely, it knows 200.1.6.0 with 
Route Distinguisher 100:11 and 193.1.1.0/24 with Route Distinguisher 100:22.
\end{shaded}

Tagging IP prefixes with a VPN identifier is an easy solution, but it is 
suboptimal in a specific use case which has seen increasing popularity 
recently: the so-called \emph{extranets}. In its simple definition, an extranet 
is simply a connection between two different VPNs that are guaranteed to have 
non-overlapping IP address spaces. A realistic example might be a specific 
site of one customer that needs to connect to another specific site of 
another customer. A naive implementation of extranets would define an ad-hoc 
VPN and assign it a new RD value. However, this solution is undesirable because 
it creates multiple VPNs that have duplicate entries, yielding a waste of 
router memory (to store the entries) and a waste of router's CPU time (to 
process update messages that are identical but for the RD value).

In order to overcome such limitations, MPLS decouples the concept of route 
distinguisher, which is used to segregate the address space in multiple 
namespaces, from the concept of \emph{route target} (RT) which is another tag 
that is used to control which routes are imported in a given VPN and, 
similarly, which routes are exported from a given VPN. The route target is 
transported by MP-BGP using extended communities. More precisely, by 
exporting a route from a VPN we attach a user-defined RT community to all 
VPN-IP prefixes belonging to that VPN. On the other hand, by importing a 
given RT into a VPN we accept that every route having that RT value will 
be visible from the devices in that VPN.

\begin{shaded}
\noindent
Each VRF instance can be configured to import or export routes labelled with a 
specific Route Target value. In our simple example, assuming that no extranet 
connectivity is required between Customer 1 and Customer 2, each VRF instance 
can simply import a single RT value, as the following configuration snippet of 
LER1 shows:

\begin{codice}
\begin{verbatim}
  ip vrf VPN1
    rd 100:11
    route-target export 100:1000
    route-target import 100:1000
  ip vrf VPN2
    rd 100:22
    route-target export 100:2000
    route-target import 100:2000
\end{verbatim}
\end{codice}

This means that all the prefixes of VPN1 announced via MP-BGP by LER1 to any other PE
are tagged with RT 100:1000. Also, any prefix that is tagged 100:1000 and is announced to
LER1 via MP-BGP is imported into the VRF of VPN1. The configuration for VPN2 is similar.
\end{shaded}

\begin{shaded}
\noindent
Route Targets provide network operators with the flexibility of leaking specific 
routes into specific VRF instances, easing the deployment of extranets. Route 
Targets are transported in MP-BGP messages as extended BGP communities. For 
this reason, the configuration of MP-BGP peers needs to specify that the peer 
supports extended communities (which are disabled by default).

\begin{codice}
\begin{verbatim}
  router bgp 100
    address-family vpnv4
      neighbor 80.80.80.4 activate
      neighbor 80.80.80.4 send-community both
      neighbor 80.80.80.5 activate
      neighbor 80.80.80.5 send-community both
    exit-address-family
\end{verbatim}
\end{codice}
\end{shaded}

\begin{shaded}
\noindent
To better understand the interplay between MP-BGP and the Route Targets, let us look 
at the content of an MP-BGP packet captured in our network (see
Fig.~\ref{fig:mpls-slides-26}). Observe how the route target in the blue frame is
contained in the extended communities.

The announcements tells to the MP-BGP peer receiving it that the packets that will be
received with the inner MPLS label 24 (red frame in the picture) will refer to the
specified route target and the specified route distinguisher (green frame in the picture).
\end{shaded}

\begin{figure}
\centering
 \frame{\includegraphics[trim=0cm 2.5cm 0cm 2.5cm, clip=true,
width=0.9\columnwidth]{figures/mpls-slides-26}}
 \caption{An MP-BGP signaling packet captured over the network.}
 \label{fig:mpls-slides-26}
\end{figure}

%
%%%
%%%%%
%%%
%
\subsection{MPLS Control and Data Plane}
The task of MPLS control plane is simply to establish Label Switched Paths 
(LSPs) between the loopback addresses of PE routers. LDP is in charge of 
populating and maintaining routers' LFIBs that implement the LSPs. In its most 
popular distribution mode, called \emph{unsolicited downstream}, LDP works in 
the following way. Each router creates a label for locally originated prefixes 
(e.g., the loopback address). The binding between a label and a locally 
originated prefix is called a \emph{local} binding. By contrast, a \emph{remote} 
binding is a binding between a label and a remotely originated prefix. Each 
router advertises its local bindings to its neighbors. When a neighboring router 
receives a binding for prefix $p_1$ and label $l_1$ on interface $i_1$, it looks 
up its IP forwarding table to check whether the advertised prefix is routed on 
interface $i_1$. If this is the case, it picks another label $l_2$ and starts 
announcing a binding for $p_1$ and $l_2$. Meanwhile, it updates its LFIB with 
the tuple $\langle l_2, l_1, p_1, i_1 \rangle$. This means that when a packet arrives that is 
labelled $l_2$, the LFIB will swap $l_2$ with $l_1$ and deliver it via interface 
$i_1$\footnote{This explanation assumes per-router label scope, which is the 
default for most router vendors. An alternative is per-interface label scope, 
when the router can advertise different labels for each interface, which is 
used mostly on ATM interfaces.}.

Regarding MPLS data plane, we have already seen that the ingress PE router looks 
up its MP-BGP RIB to find the loopback address of the egress PE router, looks up 
its LFIB to select the outer label, and then encapsulates the received IP packet by 
pushing the inner and the outer MPLS labels. The packet is then label-switched 
across the MPLS network to the egress PE router using the bindings found in the 
LFIB of each router. As an optimization, the penultimate router, i.e., the 
router that receives a local binding from the egress PE, can pop the outer 
label, in such a way that the egress PE router only receives the inner label and 
therefore performs a single lookup in its LFIB.

\begin{figure}
\centering
 \includegraphics[trim=0cm 1.5cm 0cm 1.5cm, clip=true, width=0.7\columnwidth]{figures/mpls-slides-14}
 \caption{The MPLS forwarding table of LER1.}
 \label{fig:mpls-slides-14}
\end{figure}

\begin{figure}
\centering
 \includegraphics[trim=0cm 1.5cm 0cm 1.5cm, clip=true, width=0.7\columnwidth]{figures/mpls-slides-15}
 \caption{The MPLS forwarding table of LSR1.}
 \label{fig:mpls-slides-15}
\end{figure}

\begin{figure}
\centering
 \includegraphics[trim=0cm 1.5cm 0cm 1.5cm, clip=true, width=0.7\columnwidth]{figures/mpls-slides-16}
 \caption{The MPLS forwarding table of LSR2.}
 \label{fig:mpls-slides-16}
\end{figure}

\begin{shaded}
\noindent
Figs.~\ref{fig:mpls-slides-14}, \ref{fig:mpls-slides-15}, and \ref{fig:mpls-slides-16}
show the MPLS forwarding tables of some routers of our network.
\end{shaded}

\begin{figure}
\centering
 \includegraphics[trim=0cm 1.5cm 0cm 1.5cm, clip=true, width=0.7\columnwidth]{figures/mpls-slides-19}
 \caption{An IP packet originated by the Rome site of Customer 1 reaches PE 
router LER1.}
 \label{fig:mpls-slides-19}
\end{figure}

\begin{figure}
\centering
 \includegraphics[trim=0cm 1.5cm 0cm 1.5cm, clip=true, width=0.7\columnwidth]{figures/mpls-slides-20}
 \caption{An MPLS packet reaches P router LSR1.}
 \label{fig:mpls-slides-20}
\end{figure}

\begin{figure}
\centering
 \includegraphics[trim=0cm 1.5cm 0cm 1.5cm, clip=true, width=0.7\columnwidth]{figures/mpls-slides-21}
 \caption{An MPLS packet reaches P router LSR2.}
 \label{fig:mpls-slides-21}
\end{figure}

\begin{figure}
\centering
 \includegraphics[trim=0cm 1.5cm 0cm 1.5cm, clip=true, width=0.7\columnwidth]{figures/mpls-slides-22}
 \caption{An MPLS packet traveling to PE router LER2.}
 \label{fig:mpls-slides-22}
\end{figure}

\begin{figure}
\centering
 \includegraphics[trim=0cm 1.5cm 0cm 1.5cm, clip=true, width=0.7\columnwidth]{figures/mpls-slides-23}
 \caption{An MPLS packet reaches PE router LER2.}
 \label{fig:mpls-slides-23}
\end{figure}

\begin{figure}
\centering
 \includegraphics[trim=0cm 1.5cm 0cm 1.5cm, clip=true, width=0.7\columnwidth]{figures/mpls-slides-24}
 \caption{An IP packet for Customer 1.}
 \label{fig:mpls-slides-24}
\end{figure}

\begin{shaded}
\noindent
Figs.~\ref{fig:mpls-slides-19}--\ref{fig:mpls-slides-24} illustrate the travel of a
packet through our network.

First, Fig.~\ref{fig:mpls-slides-19} shows what happens when an IP packet 
originated by the Rome site of Customer 1 reaches the PE called LER1. Namely, it 
is encapsulated into an MPLS packet with two labels and then sent to LSR1. The 
inner label (yellow) identifies the VRF while the outer label (red) is the label 
used for the forwarding process.

Second (Fig.~\ref{fig:mpls-slides-20}), the MPLS packet reaches LSR1, its outer 
red label is replaced with a blue label, and the packet is forwarded to LSR2.

Third (Figs.~\ref{fig:mpls-slides-21} and~\ref{fig:mpls-slides-22}), the MPLS 
packet reaches LSR2. LDP makes LSR2 aware that it is the penultimate hop in the 
LSP. For this reason, LSR2 simply pops the outer label and forwards the packet 
to LER2.

Fourth (Fig.~\ref{fig:mpls-slides-23}), the MPLS packet reaches LER2. LER2 
notices that there is only one MPLS label (which used to be the inner label), so 
the packet is meant to be forwarded via IP in one of the VPNs that LER2 serves. 
LER2 uses the VRF label to identify the VRF instance, then looks up the IP 
destination address in the VRF forwarding table.

Fifth (Fig.~\ref{fig:mpls-slides-24}), the IP packet is delivered to its final
destination.
\end{shaded}

%%%%%%%%%
%%%%%%%%%
%%%%%%%%%
%%%%%%%%%
%%%%%%%%%
\section{Advanced Topics}\label{se:advanced}

In this section we give more technical details about dynamic routing and connecting to the
Internet, creating complex VPN topologies, and dealing with IP-specific 
features (e.g., MTU) that need extra care when encapsulation is involved.

%
%%%
%%%%%
%%%
%
\subsection{Dynamic Routing and Connecting to the Internet}\label{sse:internet}

So far we have not yet discussed how the PE router can learn the prefixes that 
are served by its directly attached CE router. Of course, it is trivial to 
configure static routes on the PE, however this creates an undesirable 
coupling between the provider and the customer: whenever the customer wants to 
add a different IP subnet, it has to bother the provider to configure static 
routes before that IP subnet is reachable from other customer sites in the same 
VPN.

The solution is to have the CE and the PE establish an eBGP peering where the CE 
announces its local networks, while the PE announces all the networks that it 
learns in the same VPN. Observe that the, contrary to the MP-BGP peerings among 
PEs, peerings between CEs and PEs are pure eBGP peering: the CE does not know 
anything about VPNs and route distinguishers. It is the MP-BGP process on the PE 
router that takes care of processing the reachability information learned from 
the CE and updating the VPN reachability information accordingly.

Setting up an eBGP session in order to use the MPLS-VPN service may discourage 
those customers who do not have a strong BGP expertise. In such cases, usually 
the providers also offer to their customers CE management and configuration.

A BGP peering between the CE and the PE also allows a CE in a VPN to announce a 
default route, causing all other sites in the same VPN to route Internet traffic 
via that CE router. This might be advantageous if the customer's policy forces 
Internet traffic to pass through a centralized checkpoint (e.g., a firewall or a 
proxy). However, this is not the only way to connect a VPN to the Internet. For 
example, a PE might be configured to forward natively all the packets from a CE 
which do not match any VPN route. Alternatively, the default route might be 
given its own route target, and whenever a VPN site needs Internet access the PE 
simply imports that route target in the corresponding VPN routing table. We 
refer the reader to~\cite{rfc4364} for a discussion of alternatives to get 
Internet access within a VPN.

\begin{shaded}
In our sample network, customer 1 would like to be able to create a new IP 
subnet in Rome, advertise the new subnet to LER1 via an eBGP peering, and 
automatically make the customer site in Milan able to access it. LER1 should 
receive the new route via eBGP, tag the route with the correct RD value, and 
advertise it in MP-BGP. In order to do this, it suffices to configure an eBGP 
peering in the context of a VRF instance, as the following configuration snippet 
of LER1 shows.

\begin{codice}
\begin{verbatim}
router bgp 100
  address-family ipv4 vrf VPN1
  neighbor 10.0.0.2 remote-as 65001
  exit-address-family
  !
  address-family ipv4 vrf VPN2
  neighbor 193.1.1.2 remote-as 65002
  exit-address-family 
\end{verbatim}
\end{codice}

\end{shaded}

%
%%%
%%%%%
%%%
%
\subsection{Designing Complex VPNs}
So far we have assumed any-to-any connectivity within a VPN, i.e., all sites of 
a VPN communicate with all other sites. Sometimes more sophisticated 
configurations are needed. For example we might have a VPN where not all pairs 
of sites are allowed to exchange packets. A typical situation is the so called 
hub-and-spoke configuration, where a customer has a main site and several 
peripheral sites and the peripheral sites can communicate only through the main 
site.

How to do this is illustrated in the following example.

\begin{figure}
\centering
 \includegraphics[trim=0cm 1.5cm 0cm 1.5cm, clip=true, width=0.7\columnwidth]{figures/mpls-slides-27}
 \caption{A configuration where a single customer has four sites: Site 2, 3, and 4 are
only allowed to exchange traffic with  Site 1 in Rome.}
 \label{fig:mpls-slides-27}
\end{figure}

\begin{shaded}
\noindent
A suitable use of Route Distinguishers and Route Targets allows sophisticated
configurations like the one shown in Fig.~\ref{fig:mpls-slides-27} where Rome is
the main site and Turin, Milan, and Naples are the peripheral sites.

%Suppose we choose the following Route Distinguishers for the four sites, where Rome is
%the main site and Turin, Milan, and Naples are the peripheral sites:
%\begin{enumerate}
%\item[Turin:] RD \texttt{100:1}
%\item[Milan:] RD \texttt{100:2}
%\item[Rome:] RD \texttt{100:3}
%\item[Naples:] RD \texttt{100:4}
%\end{enumerate}

We can choose  Route Distinguisher \texttt{100:1} for all four sites and split the customer VPN into three VPNs. VPN1 is used to connect Turin with Rome,
VPN2 is used to connect Milan with Rome, and VPN3 is used to connect Naples with Rome.

For each VPN we define a distinct Route Target:
\begin{enumerate}
\item[VPN1:] \texttt{100:1000}
\item[VPN2:] \texttt{100:2000}
\item[VPN3:] \texttt{100:3000}
\end{enumerate}

The configuration of peripheral sites, like for example Turin, is as
follows:\\
\begin{codice}
\begin{verbatim}
ip vrf siteTurin
        rd 100:1
        route-target import 100:1000
        route target export 100:1000
\end{verbatim}
\end{codice}

Rome's PE configuration (the hub) is as follows:\\
\begin{codice}
\begin{verbatim}
ip vrf siteRome
        rd 100:1
        route-target import 100:1000
        route target export 100:1000
        route-target import 100:2000
        route target export 100:2000
        route-target import 100:3000
        route target export 100:3000
\end{verbatim}
\end{codice}

In this way Rome imports all the Route Targets and exports all the Route Targets and is
hence able to communicate with all sites. On the other hand a peripheral site like Turin
imports and exports Route targets only with respect to Rome and hence is able to communicate 
with Rome only.
\end{shaded}

%
%%%
%%%%%
%%%
%
\subsection{ToS, TTL, and MTU}
Whenever encapsulation of IP packets happens, there are three main questions 
that arise:
\begin{enumerate}
 \item what happens to the ToS / DSCP information in the IP header that the 
customer might have set in order to properly prioritize traffic?
 \item what happens to the TTL field in the IP header and how does 
encapsulation cope with forwarding loops?
 \item how does encapsulation affect MTU for upper layer protocols?
\end{enumerate}

Luckily, MPLS has an easy answer for the first two questions. Recall from 
Fig.~\ref{fig:mpls-header} that MPLS has dedicated fields for ToS and TTL. When 
the ingress PE router receives an IP packet from the CE router, it simply copies 
the values of ToS and of TTL in the MPLS header. More precisely, a push operation 
implies copying ToS and TTL from the IP header to the MPLS header. Conversely, a 
pop operation implies copying the TTL value from the MPLS header back to the IP 
header. This way, the TTL continues to serve as a hop count%
% 
\footnote{Observe that when the TTL in the MPLS header reaches $0$ (e.g. in a 
traceroute), a P router does not know how to send the corresponding ICMP error 
back to the sender, because it lacks information about VPNs. A naive yet 
effective solution is to generate the ICMP packet and label-switch it to the 
egress PE anyway. The egress PE (which has information about VPNs) will then 
send the ICMP packet back to the sender.} 
% 
even within the MPLS network, and P routers can honor the quality of service 
parameters related to the ToS field.

Regarding the third question, since an MPLS label takes 4 bytes and the PE 
router pushes two of them, the MTU within the MPLS network should be at least 8 
bytes larger than the MTU that the CE is aware of. Given that modern OSes tend 
to perform path MTU discovery by default, MTU is becoming less of an issue for 
MPLS deployments. Rewriting the Maximum Segment Size TCP options at the PE router is also a 
common solution, even though it does not support UDP traffic.

%%%%%%%%%
%%%%%%%%%
%%%%%%%%%
%%%%%%%%%
%%%%%%%%%
\section{Summary}\label{sec:summary}

\subsection{Strengths of MPLS VPNs}

After having described the details of MPLS VPNs, we are able to discuss the 
extent to which the goals that we stated in Section~\ref{sec:intro-goals} are 
met.

By using Route Distinguishers and label stacks within the provider cloud, 
customers can retain their IP address plan and the traffic belonging to 
different customers is properly segregated. Moreover, the configuration of CE 
routers is completely unaware of MPLS-specific details. Since MPLS transports 
QoS information by copying the ToS field from the IP header, MPLS VPNs can in 
principle provide different forwarding treatment to different packets. However, 
the architecture of MPLS VPNs does not inherently support QoS, because LSPs are 
simply built from the underlying IP plane. Other mechanisms (e.g., 
\cite{rfc3209}) can be employed to compute LSPs based on QoS features.

Providers are able to keep the configuration in the core of the network 
extremely simple and scalable: in fact, the configuration of P routers does not 
depend on the number of deployed VPNs. Since the backbone is only concerned with 
transporting packets from a PE to another, the sizeof the forwarding table of P routers only depends on the number of PEs and does not depend on the number of prefixes of VPNs.
%forwarding table of P routers only contains one entry for each loopback address, which makes forwarding performance independent of the number of VPNs. 
Configuring a new VPN implies 
modifying the configuration of the PE routers that are directly connected to the 
customer's sites. Moreover, such a configuration boils down to assigning a 
unique RD and RT and establishing eBGP peerings with the CE routers.

%%%%%%%
%%%%%%%
%%%%%%%
\subsection{Limitations of MPLS VPNs}

Virtual Private Networks designed with MPLS have also some known limitations. 
At least the following should be mentioned.

\begin{itemize}

 \item BGP know-how is needed to configure the customer CE. As discussed in Section~\ref{sse:internet}, this sometimes forces the carrier to provide also CE management.

 \item Customer CEs need to support BGP if eBGP peering are established with the PEs. This may not be the case for cheap entry-level router models. As an alternative, BGP can be replaced by static routes configured on PEs, sacrificing part of the flexibility that a dynamic routing protocol between CEs and PEs provides.

 \item The P in the MPLS acronym stands for ``Private''. However, MPLS VPNs are only private at routing level: no authentication, confidentiality, or integrity is provided by the architecture. For instance, the provider can inspect all customers' traffic in plaintext. Even worse, since the separation is enforced at the routing level, it turns out that the ability of guaranteeing isolation within the same VPN actually depends on the provider's topology~\cite{integrity}.

 \item The basic MPLS architecture lacks support for quality of service. QoS can usually be offered on top of MPLS, e.g., by establishing LSPs which reflect traffic engineering policies. However, adding traffic engineering and quality of service support comes at the cost of keeping more state in the network, hence posing scalability concerns~\cite{juniper-preso,rfc5439}.

 \item In most configurations, the customer and the provider share the job of maintaining a network, which potentially complicates debugging routing and connectivity problems.

\end{itemize}

%%%%%%%
%%%%%%%
%%%%%%%
\subsection{Further Readings}\label{sec:further-readings}

The interested reader could refer to the classical books about MPLS authored by Minei and Lucek~\cite{minei-lucek} and by De Ghein~\cite{de-ghein}.

Most of the technologies regarding MPLS are defined by RFCs. These include the main MPLS VPNs architecture~\cite{rfc3031,rfc4110}, label distribution via LDP~\cite{rfc5036}, Layer 2 VPNs~\cite{rfc4664}, BGP variants~\cite{rfc2858,rfc4364,rfc5512}, and RSVP-TE~\cite{rfc3209}.

Considerations about about MPLS VPNs integrity and scalability can be found in~\cite{integrity} and~\cite{juniper-preso,rfc5439}, respectively.

\subsection*{Acknowledgments}

We thank the anonymous reviewers for constructive criticism and for suggestions 
that helped us improve both the content and the presentation of this chapter. We 
also thank Mario Cola and Massimo Rimondini for their help and 
friendship.

\bibliographystyle{plain}
\bibliography{bibliography}

\end{document}

### 2013-07-22 Compiled/chapter7/as.tex
all files compile correctly

\clearpage
\section{AS-level topology}
\label{sec:as}

When trying to establish a precise meaning or interpretation of the use of 
``Internet topology,'' in much of the existing literature, we find that the 
phrase has often been taken to mean a virtual construct or graph created by 
the Border Gateway Protocol (BGP) routing protocol.  Commonly referred to as 
the inter-domain or Autonomous-System (AS) topology --- named after the logical 
blocks (ASes) that are used in BGP to designate the origin and path of routing 
announcements --- it is this particular connectivity structure that we focus 
on in this section, though we will see that the notion of the AS-topology is 
more slippery than commonly imagined. In particular, we will discuss some of 
the main issues that arise in the context of studying the Internet's AS topology 
(ranging from proper definitions and interpretations of this construct to 
measurements) and focus less on modeling-related aspects as they are still in
their infancy, especially when compared to the advances in router-topology
modeling described in \autoref{sec:router}.

\subsection{A look back}
\label{sec:as_hist}

As far as we know, the first researchers to use BGP-based measurements
in the form of route monitor data for topology-related work were
Govindan and Reddy~\cite{govindan97:_topology}, who introduced the
notion of the {\em inter-domain topology} defined as ``the graph of
domains and the inter-domain peering relationships.''  However,
although they were quite specific in regard to being interested in
routing, the concept was reused when
Faloutsos~\etal~\cite{faloutsos99:_power_law_relat_of_inter_topol}
coined the term ``Internet topology'', a paper that is more widely
cited (at least outside of the network research literature)
than~\cite{govindan97:_topology}.  The
paper~\cite{faloutsos99:_power_law_relat_of_inter_topol} is
responsible for advancing the alluring notion that the inter-domain
topology of the Internet is a well-defined object and can be {\em
  accurately} obtained and reconstructed from the available BGP route
monitor data. As we shall see, this is not at all the case, and it has
fed into a large subsequent scientific literature, already discussed
earlier, \eg \cite{barabasi99,barabasi02}.

The problems lie in the very definitions of the AS-topology and the
measurements that have been used to study this topology (we return to
the measurements in \autoref{sec:as_measurement} below).  In terms of
definitions, in the context of the AS-level Internet, it is tempting
to simply equate a node with an AS, but this begs the question what an
AS really is.  The term refers formally to the AS number (ASN)
allocated by IANA (Internet Assigned Numbers Authority) or the
Regional Internet Registries (RIRs). An ASN is tendered to enable
routing using BGP.  This is {\bf not} equivalent to the popular view
that associates an AS with a set of routers that appear to the outside
as if they formed a single coherent system with a 1:1 mapping between
it and some administering company.

For instance, an organization may often own a router which has at least one 
interface IP address belonging to another organization. In fact, many point-to-point 
IP links occur across a ``/30'' subnet. When the link joins two networks, this subnet 
must be allocated from the IP blocks of one or the other connecting network, and so 
most such connections result in IP addresses from neighboring ASes appearing locally.

Another problem arises from the fact that although an AS is often considered to 
correspond to a single technical administrative domain, \ie a network run by one 
organization, it is common practice for a single organization to manage multiple 
ASes, each with their own ASN~\cite{cai10:_AS_to_organ_map}. For instance, Verizon 
Business (formerly known as UUNET) uses ASNs 701, 702, 703 to separate its
E-BGP network into three geographic regions, but runs a single IGP instance throughout 
its whole network. In terms of defining nodes of a graph, these three networks are all 
under the same operational administrative control, and hence should be viewed as a 
single node. On the other hand, as far as ASNs are concerned, they are different and 
should be treated as three separate nodes.  The situation is actually more complex 
since corporations like Verizon Business own some 200+ ASNs~\cite{cai10:_AS_to_organ_map} 
(not all are actually used, though). In many of these cases, a clear boundary between 
these multiple ASes may not really exist, thus blurring the definition of
the meaning of a node in an AS graph. Similar problems can arise when a single AS 
is managed by multiple administrative authorities which consist of individuals 
from different corporations. For example, AS 2914 is run partially by NTT/America 
and partially by NTT/Asia.

All this presumes that an AS is a uniform, contiguous entity, but that
is not necessarily
true~\cite{muehlbauer06:_build_as,muehlbauer07:_in_search}. An AS may
very well announce different sets of prefixes at different exit points
of its network, or use BGP to balance traffic across overloaded links
(other reasons for heterogeneous configurations are reported
in~\cite{bush09:_inter_optom}). \autoref{fig:puzzle} illustrates the
problem. The AS-graph simplifies, in some cases grossly, the very
complicated structure of the entities involved, which are often
heterogeneous, and not necessarily even contiguous either
geographically or logically.

For all these reasons, it should be clear that modeling an AS as a single 
{\em atomic} node without internal (or external) structure is overly simplistic 
for most practical problems. Moreover, these issues cannot simply be addressed 
by moving towards graph representations that can account for some internal 
node structure (such as in~\cite{muehlbauer06:_build_as}), mainly because 
BGP is unlikely to reveal sufficient information to infer the internal structure 
for the purpose of faithful modeling.

\begin{figure}[!th] 
  \begin{center}
    \begin{subfigure}[b]{0.34\textwidth}
      \centering 
      \includegraphics[width=1\textwidth]{puzzle_1.pdf}
      \vspace{17mm}
      \caption{A simple section of the ``AS-graph''.}
    \end{subfigure}
    \hspace{0.05\textwidth}
    \begin{subfigure}[b]{0.54\textwidth}
      \centering
      \includegraphics[width=1\textwidth]{puzzle_2.pdf}
      \caption{A possible picture of the router-level connectivity.}
    \end{subfigure}
    \caption{An illustration of the obfuscation of the AS-graph
      (in the vein of \cite{griffin02:_under_border_gatew_protoc_bgp}). The
      graph may appear simple, but hides heterogeneous, non-atomic,
      dis-contiguous entities and interconnects. At the minimum, this
      should illustrate the dangers of talking about the ``Internet''
      graph. \label{fig:puzzle}}
  \end{center}
\end{figure}

Moreover, the AS-graph treats ASes as nodes, with connecting edges,
but the real situation is much more complex. ASes are complex networks
in their own right, and are connected sometimes by multiple edges
(M{\'e}rindol~\etal~\cite{merindol09:_quant_ases} found that
over half of the ASes they studied were connected by multiple links),
and sometimes through Internet eXchange Points (IXPs) that connect
multiple ASes. In fact, the traditional approach of modeling the
AS-level Internet as a simple connected di-graph is an abstraction
incapable of capturing important facets of the rich semantics of
real-world inter-AS relationships, including different
interconnections for different policies and/or different
interconnection
points~\cite{muehlbauer06:_build_as,muehlbauer07:_in_search}. The
implications of such abstractions need to be recognized before
attributing network-specific meaning to findings derived from the
resulting models.

% In addition to these seemingly trivial, but actually subtle and
% important points of definition, we have those of the level of
% abstraction desired:
% \begin{enumerate}

% \item The commonly-used practice of abstracting ASes to generic atomic
%   nodes without any internal structure is an over-simplification that
%   severely limits our ability to capture critical features associated
%   with real-world ASes such as route diversity, policy diversity, or
%   multi-connectivity.

% \item The traditional approach of modeling the AS-level Internet as a
%   simple connected di-graph is an abstraction incapable of capturing
%   important facets of the rich semantics of real-world inter-AS
%   relationships, including different interconnections for different
%   policies and/or different interconnection points. The implications
%   of such abstractions need to be recognized before attributing
%   network-specific meaning to findings derived from the resulting
%   models.

% \item AS topology as an uninspiring and often meaningless abstract
%   graph towards an approach that views the AS Internet as an economic
%   construct that is constrained by socio-technological factors and is
%   driven by economic incentives and business decisions made by the
%   major players in this area (\eg service and content providers,
%   large corporations, governments).  Although this notion has been
%   advanced by the networking operator community for some
%   time~\cite{norton1,norton_paid_peering,huston99:_peering_i,huston99:_peering_ii},
%   the networking research community has been slow to react and to
%   distance itself from the popular graph view of the inter-domain
%   topology (examples of exceptions
%   include~\cite{chang95,dhamdhere:conext10,labovitz10:_inter_inter_domain_traff}).

% \end{enumerate}

\subsection{Know your measurements}
\label{sec:as_measurement}

In studying the AS-level Internet, there are some critical differences compared 
to looking at the router-level Internet:

\begin{itemize}

\item No-one ``owns'' the AS structure. There isn't anyone with the type of 
privileged view that a network operator has of its own network. There are 
tens of thousands of ASes, and so we can't reasonably expect to consult all 
of them to collate a picture either.

\item ASes are not ``nodes''. They are complex in their own right, so viewing 
the AS-level Internet as an AS-graph is a big abstraction of reality.

\item Routing between ASes is very different from routing within ASes and 
highlights the difference between graph representations that 
reflect ``reachability" vs. ``connectivity" information.

\end{itemize}

These differences create interesting problems and opportunities for
measurements, some with parallels to the router-level measurement
problems and others without any such parallels.

\subsubsection{Data-plane vs. control-plane measurements}

As discussed in \autoref{sec:router_measurements}, despite all its
deficiencies, traceroute has been the method-of-choice for obtaining
router-level measurements. As a prime example of an active measurement
tool that is confined to the data plane (\ie probe packets take the
same paths as generic data packets), traceroute has also been used to
obtain information about the AS topology but has additional problems
in this domain.

Apart from the already problematic issues (\eg load-balancing,
aliasing, missing data), IP addresses along traceroute paths must now
be mapped to ASes.  This mapping is even harder than the mapping to
routers, not just because the data for doing so is inaccurate or
incomplete (\eg IP to organization allocations may not work because
an organization does not directly correspond to an AS), but also
because the border of an AS is not well-defined in terms of IP
addresses. It is common for a link between two ASes to come from a
subnet allocated by one of the ASes, resulting in an interface in the
other network with an address that is not its
own~\cite{mao03:_as-tr,marchetta13:_detec_third_addres_tracer_traces}.
The problem is further complicated by variations such as anycast or
Multiple Origin ASes \cite{Zhao01:_MOAS}, which provide yet another
set of counter-examples to a straight-forward mapping between AS and
address space. Some work has concentrated on trying to improve the
mapping \cite{pansiot10:_extrac}, and these represent technical
advances, but it is important to understand that the fundamental
difficultly lies in the fact that the boundaries of the ``business''
are not equivalent to the AS boundaries.

The other major alternative to obtaining information about the AS topology is 
to collect control plane data in the form of directly measured routing information.  
The primary example of such control plane data are BGP-derived measurements. 
BGP is a path-vector routing protocol, and as such each node transmits
to its neighbors information about the {\em best} path that it knows to 
a destination. Each node then takes the information it has received about 
best paths, and computes its own best path, which it transmits to its neighbors.  
A route monitor receives this information as would any router, and from the 
transmitted path information, can infer links between ASes.  The two best known 
projects that rely on BGP route monitors, Oregon RouteViews~\cite{ORVdata}, 
and RIPE (R\'{e}seaux IP Europ\'{e}ens)'s Routing Information Service~\cite{RIPEdata} 
both use this approach, and each connects to a few dozen different ASes.

However, by its very design, BGP is an information-hiding rather than an 
information-revealing routing protocol. In addition, by its very design, 
BGP is all about reachability and {\bf not} connectivity.  Using it for 
mapping the Internet inter-domain topology is a ``hack'', and so it should 
come as no surprise that it has its own set of problems, including the following:

\begin{itemize}

\item The AS-path information in the announcements is primarily
included for loop detection and does not have to correspond to
reality. It is easy (and not uncommon) to insert additional ASes into 
a path for various purposes, \eg traffic engineering or
measurement~\cite{colitti06:_poison,bush09:_inter_optom}, and
moreover, the AS-path does not have to represent the data path.

\item Path-vector protocols do not transmit information on every path 
in the network. For instance, backup paths may never appear in any 
routing announcements (unless there is a failure), and so may not be 
seen by a route monitor.

\item Path-vector protocols only transmit ``best'' paths, and so there 
is a large loss of visibility from any one viewpoint. It is sometimes 
argued that a large number of viewpoints would
alleviate this, but the viewpoint locations are highly biased
towards larger networks, and this known ``vantage point problem" 
severely biases the possible views of the network~\cite{roughan08:_bigfoot}.

\end{itemize}

The BGP measurement data being provided by RIPE and RouteViews was originally 
intended to help debug networks, not for mapping. While this data collections 
have been invaluable for that intended original purpose, it is unsurprising 
that it is inadequate when used for a rather different purpose such as mapping 
the AS Internet. However, when this aspect is carefully taken into account, 
good work can be done but requires a critical evaluation of the data.  Problems 
arise primarily when this data is used uncritically.  Other useful sources of 
AS-level measurements such as looking glass servers and route registries suffer 
from similar problems~\cite{Mahadevan06,he07}, and do so for similar reasons: 
they weren't intended to draw a map of the AS Internet.

\subsubsection{Attribute discovery}

The AS topology may be interesting to scientists in itself, but to be useful 
to network engineers, the routing policies that accompany it should also be known.  
It has been common to approximate the range of policies between ASes by a simple 
set of three relationships: (a) customer-provider, (b) peer-peer, and (c) siblings.  
This reduction was at least in part motivated by Huston~\cite{huston99:_peering_i,huston99:_peering_ii} and has been used in various places
\cite{subramanian02:_charac_inter,wang03:_infer,xia04:_as}.  While many 
relationships fall into these three categories, there are frequent exceptions~\cite{hyun03:_tracer_and_bgp,muehlbauer06:_build_as,Qiu07}, for instance, in the form 
of partial transit in a particular region~\cite{maz:partial-transit,norton1}.

Forgetting for the moment the simplification in assuming all policies fit this 
model and the simplifications the AS-graph itself makes, the relationships can 
be represented in the graph by providing simple labels for each edge. Typically, 
the next step after inferring network topology is to infer policies between ASes. 
The most common approach to this problem is to assume the universality of the peer-peer, 
customer-provider, sibling-sibling model, and to infer the policies by finding an 
allocation of policies consistent with the observed
routing~\cite{gao_inferring,xia04:_as,wang03:_infer,battista_inferring:2003,dimitropoulos07:_as_relat}.

Once relationships are established, a seemingly reasonable next step
is to estimate the hierarchical structure as in
\cite{subramanian02:_charac_inter}. However, the effect of large
numbers of (biased) missing links has not really been considered in
these algorithms.  In fact, the tier structure of the Internet seems
to be largely an illusion. Recent work has shown that there is little
value in the model at
present~\cite{gill08:_flatt_inter_topol,labovitz10:_inter_inter_domain_traff};
but, in contrast to the claims of these papers, there is no strong
evidence that the situation has actually changed or that the tier
model was ever a good model (except maybe in the early stage of the
``public" Internet in the latter 20th century) particularly in light
of the problems in the data.

Alternatively, we can infer a generic set of policies consistent with
routing observations using a more detailed set of routing
measurements~\cite{path_inference_sigmetrics:2005,muehlbauer06:_build_as}
and estimate performance by comparing predicted routes to real routes
(held back from the inference process).

\subsubsection{The ``missing link" problem: Extent and impact}

Perhaps the most obvious problem that results from relying on BGP measurement 
data to map the AS-level Internet is that there are many missing links in the 
resulting AS-graph.  To illustrate the extent of this problem, years of 
concentrated research efforts that relied on a combination of improved inference methods and additional data sources~\cite{chang04:_towards,zhang05:_collec_inter_as_topol,muehlbauer06:_build_as,dimitropoulos07:_as_relat,he07,he09:_lord_of_links,roughan08:_bigfoot,augustin09:_ixps,chen09:_sidewalk,oliveira10:_incompl_of_obser_inter,dhamdhere11:_twelv_years_evolut_inter_ecosy,sanchez13:_dasu}
have produced a picture of the Internet's AS topology that --- as of 
2011 --- consisted of some 35,000-40,000 ASes (nodes) and about 115,000-135,000 
edges (AS links), with about 80,000-90,000 of them being of the customer-provider type and 
35,000-45,000 of the peer-peer type.

More recently, this supposedly up-to-date and most complete view of
the AS-level Internet changed drastically thanks
to~\cite{ager12:_anatom_europ_ixp} that relied on ground truth data
from one of the largest IXPs in Europe (and worldwide) that had at the
time of this study almost 400 member ASes. The main finding of this
recent study is that in this single location, the number of actively
used AS links of the peer-peer type was more the 50,000 --- larger
than the number of all AS links of the peer-peer type in the entire
Internet known as of 2011. Moreover, being extremely conservative when
extrapolating from this IXP to the Internet as a whole,
\cite{ager12:_anatom_europ_ixp} shows that there are easily more than
200,000 AS links of the peer-peer type in the entire Internet, more
than twice the number of all AS links of the customer-provider type
Internet-wide.  Importantly, the main reason for this abundance of AS
links of the peer-peer type at IXPs is well understood --- many IXPs,
especially the larger ones, offer as free service to their member ASes
the use of their route server. This service greatly facilitates the
establishment of peer-peer links between the members of an IXP and has
become enormously popular with members that have an ``open" (as
compared to restrictive or selective) peering policy. Especially for
the larger IXPs, such networks typically constitute the vast majority
of IXP member ASes. \autoref{fig:ixp} provides an illustration of the
connectivity through this IXP and shows that a majority of its member
ASes have an open peering policy (some 300+ members) and establish AS
links of the peer-peer type with one another.

\begin{figure}[thbp] 
  \begin{center}
% pdfimages gets actual 'images' (bitmaps) out of the PDF, but not plots
%  pdfjam p163-ager.pdf  '7' --outfile p163-ager-p7.pdf 
%     extracts a particular page so I can use it here
    \includegraphics*[width=0.65\columnwidth,viewport=77mm 225mm 133mm 270mm]{p163-ager-p7.pdf}
%                                                    xleft ybottom xright ytop
%                                                 a4 = 0 0 210 297
    \caption{Scatter-plot of number of peers per member, based on a
      classification of the member ASes in the four business
      categories defined above: LISP (Large ISP), SISP (Small ISP),
      HCDN (Hosting/service and Content Distribution Network)), and
      AEN (Academic and Enterprise Networks), and by tier.  
      (Reprinted from \cite{ager12:_anatom_europ_ixp}; \copyright 2012 ACM, Inc. Included here by permission.)
      \label{fig:ixp}}
  \end{center}
\end{figure}

In short, for many years, researchers have worked with AS-graphs that are typically complete 
in terms of nodes, but easily miss more than half the edges. Importantly, these graphs 
have generally a 2:1 ratio of customer-provider type vs. peer-peer type links when 
a 1:3 ratio is much more likely to reflect Internet reality. Clearly, for gaining 
any economic-based understanding of the AS Internet, getting that ratio approximately 
correct is paramount because it is directly impacting how money flows in the 
Internet --- while in a customer-provider relationship, the former pays the latter 
for bandwidth, peer-peer relationships are typically settlement-free (\ie no 
money is exchanged between the involved parties).

Besides their immediate economic impact, the above missing edges cause also 
significant problems in inferring the AS graph. For instance, it is a requirement 
that a network be multi-homed to obtain an ASN. This means the AS needs to intent 
to connect to at least two upstream providers. In this sense a ``single-homed stub-AS''
does not exist.  Without any doubt, there are exceptions to this rule. However, the 
second link is often a backup link which is invisible to BGP outside of the immediate 
connection, because of BGP's information hiding\footnote{Note that complex BGP policies 
may play a role in this as well~\cite{cost_community,tim:wedgies}.}. Thus, it may 
appear as if a large number of ASes are single-homed stubs.

In \cite{oliveira10:_incompl_of_obser_inter}, the authors separate the missing 
links into {\em hidden} and {\em invisible}.  Whereas the latter are links that 
are missing from the data for structural reasons (\ie it is not just a question 
of quantity (\ie numbers of monitors) but quality (\ie location of monitor)), 
the {\em hidden} links may be found with enough measurements (over time, or 
multiple viewpoints). In \cite{roughan08:_bigfoot} the authors extend that by 
dividing links into a number of classes based on their observability.

The ``missing link" problem in the AS context is much more serious than if those 
links were ``missing at random''.  In particular, the bias in the type of links 
that are missing~\cite{roughan08:_bigfoot} is critical when calculating some 
metrics on the graph, such as distances, precisely because such links are often 
{\em designed} to cut down on the number of ASes traffic must traverse. The 
missing data is also crucial for understanding reliability: for instance, papers 
such as \cite{barabasi00} that argue that high-degree nodes create
vulnerabilities in the Internet ignore the backup links that are invisible in 
these dataset, but obviously crucial when studying the resilience of the network.

\subsection{The Internet's AS-level topologies}

Despite the limitations of measurements, there is a considerable
amount known about the AS-level topology of the Internet, and we talk
here about the issues in defining and modelling that topology.  We
have seen that the definition of an AS is fraught with
problems. Assuming for the time being that the concept of an AS is
well defined so that it makes sense to equate each AS with a node in a
graph, then what is the set of links?  Unfortunately, the question of
which ASes are ``adjacent'' also has no simple answer, and defining
the meaning of a ``link'' between two ASes requires further
consideration.

Does a link mean the ASes have a business relationship, physical
connectivity, connecting BGP session, or that they share traffic? All
the above are reasonable definitions, and none are equivalent. A
common definition is that two ASes are said to be connected (at a
particular time), if they can exchange routing data (and presumably IP
traffic) without the help of an intermediary AS that provides {\em
  transit}. However, this says little about the true business
relationships that are sometimes discussed as a matter of course when
the AS-graph is considered. Moreover, this abstraction looses
considerable information. In reality there are multiple topologies we
want to model, each with its own meaning, structure, potential
applications, and inference problems.

\begin{itemize}

\item {\em Business relationship graph:} in its simplest form this
  graph simply indicates (by an edge) that a business relationship
  exists between the corporations that own two ASNs. Edges could be
  usefully labelled by the type of business relationship, and we list
  a small subset of the possible relationships in
  \autoref{tab:as_graph_set}.

\item {\em Physical link-level graph:} this graph indicates whether
  two ASNs have a physical (layer 1) connection, and how many such
  connections they have. The multiple nature of such connections leads
  this to being a multigraph, as it is very common for two ASes to be
  connected by multiple links and in different geographic locations
  \cite{norton_paid_peering,liljenstam03:_devel_of_inter_backb_topol,spring03:_inflation}. 
  The idea is clearly illustrated by Figure~1 in
  \cite{liljenstam03:_devel_of_inter_backb_topol}, which shows a
  ``pancake'' diagram of the North American Internet backbone. Perhaps
  the reason this critical aspect of the topology is typically ignored
  is that it is very hard to measure---BGP monitor data is in general blind
  to this facet of the topology.
% Traceroutes can show more detail, but suffer from their own problems.
  In addition, this graph should really be a hypergraph. A single
  ``edge'' can connect multiple ASes, for example through an IXP
  \cite{hyun03:_tracer_and_bgp,xu04:_ixps,augustin09:_ixps}. One might
  argue that they are joined by a switch/router, each using
  point-to-point links, but in at least some cases, that switch has no
  place in a AS graph (\ie it has no ASN).  The graph's edges could
  be usefully annotated with link capacity and potentially other
  features such as geographic location.

\item {\em Connectivity graph:} this graph indicates that layer-2
  connectivity exists between two ASNs. In many cases the layer-2
  connectivity between ASNs would be congruent with the layer-1
  connectivity, but with recent advances in network virtualization this
  may not hold for long~\cite{vroom}.

\item {\em BGP routing graph:} the edges in this graph indicate pairs
  of ASes that have an active BGP session exchanging routing
  information (\ie a BGP session that is in the `established' state~\cite{RFC4271}).

\item {\em Policy graph:}  the edges in this graph are the same as
  those in the BGP routing graph, but include directed policy
  annotations~\cite{tim10:ssp}. We define this separately from the BGP routing graph
  because it may require a multigraph to allow for policy differences
  between different regions.

\item {\em Traffic graph:} it is the same as the BGP routing graph,
  but the edges are annotated with the amount of traffic exchanged
  between the corresponding ASes.

\end{itemize}

This is hardly a complete set of possibilities, but already we can see
the potential complexity here. Nevertheless, it appears unusual for studies to
even define precisely what graph they examine (exceptions being papers
such as \cite{govindan97:_topology,oliveira10:_incompl_of_obser_inter}
where the BGP routing graph is explicitly considered).  In
\autoref{tab:as_graph_set}, we list some of the possible graphs, and
their basic properties.  There is no clean 1:1 mapping between
``network'' and ``organization'' and ``AS''
\cite{hyun03:_tracer_and_bgp,cai10:_AS_to_organ_map}, and so it is
highly non-trivial to map between these graphs, and they are certainly
not equivalent.

\begin{table*}[t]
\begin{center}
  \begin{tabular}{|l|l|l|}
    \hline 
    Graph & Edge Annotation & Graph Type \\
    \hline 
    business relationship & subsidiary, partner, customer,... & directed graph \\
    physical link-level & link capacity & multi- hyper-graph \\
    connectivity graph & - & multigraph \\
    BGP routing graph & - & undirected graph \\
    policy graph & BGP policies & directed multigraph \\
    traffic graph & traffic volumes & directed graph \\
     \hline 
  \end{tabular}
  \caption{Example elements of the set of AS graphs.}
  \label{tab:as_graph_set}
\end{center}
\end{table*}

% It should also be clear by now that ASes are not atomic, they are
% geographically distributed entities \cite{rasti10:_eyebal_ases},
% comprised of multiple components distributed over some area of
% space. In principle (according to the RFCs) the AS should have one
% routing policy. Not only is it not exactly clear what this means, but
% it is clearly not
% true~\cite{hyun03:_tracer_and_bgp,muehlbauer06:_build_as,bush09:_inter_optom}.
% The components of an AS may not even be contiguous. An AS may rely on
% a provider AS for transit of its traffic between multiple otherwise
% disconnected components.

\subsection{A look ahead}

Given our list of problems described here, one might be tempted to
think that the AS-graph and routing data in general are useless until
these datasets are drastically improved. However, apart from their
operational utility, RouteViews and RIPE RIS have provided the
essential ingredients for many important studies that match those
services' goals~\cite{RISgoals}. A number of these studies have
improved the Internet significantly, and in the majority of such
successful papers there is no need to exploit the ``graph'' view of
the network. Examples include: (a) The discovery of slow convergence
and persistence oscillation in routing
protocols~\cite{varadhan96:_oscillation,labovitzSig97,labovitzInfo99,labovtizFTCS99,labovitzSig00,govindanNetworks00,griffin02:_analy_med_oscil_probl_bgp}. (b)
Understanding of the impacts (positive and negative) of route flap
dampening~\cite{mao02:_flap,pelsser:rfd}. (c) Determining how much
address space and how many ASNs are being actively
used~\cite{address_space_growth}. (d) Looking for routing ``Bogons''
often related to Internet address hijacking
%\footnote{Hijacking is becoming a potentially destructive problem. See the recent incident with China Telecom \cite{chinahijacking}.}
\cite{ramachandran06:_spam,cymru,hijacking,nick-bogon,chinahijacking}. (e) Debugging
network
problems~\cite{bush07:_testin,feldmann04:_bgp,roughan04:_SNMP_BGPb}.

On the measurement side, there have also been many advancements
towards improving our view of AS topology. For instance:

\begin{enumerate}

\item As BGP routing changes, often
  multiple potential paths are explored and these paths (which are
  unlikely to actually be used as a final choice) can show some of the
  alternative routes available in the
  network~\cite{zhang05:_collec_inter_as_topol}, and thus a more
  complete topology.

\item Missing edges can be found using additional datasets, \eg RIRs
  and looking glasses~\cite{chang04:_towards,zhang05:_collec_inter_as_topol,he07,he09:_lord_of_links},
  or IXP data~\cite{he07,he09:_lord_of_links,augustin09:_ixps,sanchez13:_dasu},
  though care must be exercised with any additional dataset.

\item A routing beacon~\cite{labovitzSig97,mao03:_bgp_beacon,bush09:_inter_optom}
  is just a router that advertises and withdraws
  certain prefixes on a regular schedule. Examination of the observed
  announcements and withdrawals by various route monitors then allows
  estimates of protocol behavior such as convergence time.

\item Route poisoning prevents announcement from reaching
  certain parts of the Internet. As with beacons, it allows one to
  examine the behavior of BGP in a more controlled manner. This is perhaps 
  the only way to see (some) backup paths, or to understand whether an ISP uses default
  routing~\cite{colitti06:_poison,bush09:_inter_optom}.

\item There are also attempts to not just estimate the topology
  but derive some quality measure for the resultant
  AS-graph~\cite{inter_as_level_topol_const_analy,winter09:_model_inter_routin_topol_with,roughan08:_bigfoot}.

\end{enumerate}

There is often an unfortunate side-effect to some of these types of measurement in
form of a Heisenberg-like uncertainty principle. That is, it is not clear
whether observed changes are due to the micro-phenomenon of path
exploration or macro-phenomena of link changes, new entrants, etc.
The longer we make observations, the more complete they may seem,
but we then do not know whether all of those links existed at the
same time. Such uncertainty principles appear to be present in a
number of Internet measurement contexts~\cite{roughan05:_fundamental} 
where we trade off ``accuracy'' of the measurements against ``time localization''.
In any case, this approach does not overcome the structural bias mentioned earlier.

At the same time, the above-mentioned and other advances on the measurement side
suggest that the missing link problem may be improved, providing
``more complete'' AS graphs. However, there is a profound need
(illustrated by the above) for better data accuracy measurements, and
better response to data quality issues from subsequent users of the
data. Obvious ways to improve are to conduct sensitivity analysis (of
results) to missing or incorrect input data.

In addition, it is to be hoped that more controlled experiments are
conducted (\ie experiments that have a ``control'' sample against
which the experimental data can be compared) in order to precisely
derive which factors of interest affect which variables.  Controls
allow one to discriminate alternative explanations for results, and
prevent the affects of one confounding factor drowning out the affects
of others (see \cite{mao03:_bgp_beacon,bush09:_inter_optom}). This is
basic tenet of the scientific method, but seems to have been ignored
in this area of research. Most studies have been ``observational'',
and while there is a valid role for such experiments, for instance in
epidemiology, they are intrinsically harder to interpret.

Lastly, another aspect of this richer set of AS topologies is that it should be
obvious by now that economic or commercial objectives by and large determine
and shape the structure and evolution of their real-world counterparts, and that 
these constructs are once again naturally expressed through optimization rather 
than random graph models, though in this case the optimization problems may come
from game theory or economics rather than mathematical programming.

\subsection{Notes}

The primary sources for the material presented in this section are

\begin{itemize}

\item[\cite{roughan11}] M. Roughan, W. Willinger, O. Maennel,
D. Perouli, and R. Bush. 10 Lessons from 10 Years of Measuring and Modeling the Internet's Autonomous Systems, in: {\em IEEE Journal on Selected Areas in Communications} 29(9):1810-1821, 2011.

\item[\cite{ager12:_anatom_europ_ixp}] B. Ager, N. Chatzis,
A. Feldmann, N. Sarrar, S. Uhlig, and W. Willinger. Anatomy of a
large European IXP, in: {\em Proc. ACM SIGCOMM'12, ACM Computer
Communication Review} 42(4), 2012.

\end{itemize}
and they contain lengthier discussions of many of the issues touched upon here.

For additional and more in-depth reading materials (in
addition to the references indicated throughout) we point to

\begin{itemize}

\item[\cite{chang-thesis06}] H. Chang. Modeling the Internet's
Inter-Domain Topology and Traffic Demand Based on Internet Business Characterization, PhD Thesis, University of Michigan, 2006.

\item[\cite{Donnet07}] B.Donnet and  T. Friedman, Internet Topology Discovery: A Survey, IEEE Communications Surveys \& Tutorials, 9(4), pp.56-69, 2007.

\item[\cite{dhamdhere-thesis08}] A. Dhamdhere. Understanding the
Evolution of the AS-level Internet Ecosystem, PhD Thesis, Georgia Institute of Technology, 2008.

\item[\cite{haddadi08:_networ_topol}] H. Haddadi, M. Rio,
G. Iannaccone, A. Moore and R. Mortier, Network topologies:
inference, modeling and generation, IEEE Communications Surveys,
10(2), 2008.

\item[\cite{dhamdhere11:_twelv_years_evolut_inter_ecosy}] A. Dhamdhere
  and C. Dovrolis. Twelve Years in the Evolution of the Internet
  Ecosystem, in: {\em IEEE/ACM Transactions on Networking} 19(5),
  2011.

\end{itemize}

### 2013-07-22 Compiled/chapter7/conclusion.tex
all files compile correctly

\clearpage
\section{Conclusion}
\label{sec:conclusion}

This chapter has aimed at clarifying the state of the art in Internet
topology measurement and modelling, and correcting a number of clear
and present flaws in reasoning. As we outlined in the introduction, we
can see a number of themes recurring at multiple levels of hierarchy
in topology modelling:

\begin{description}

\item[{\bf Theme 1:}] When studying highly-engineered systems such as
  the Internet, ``details'' in the form of protocols, architecture,
  functionality, and purpose matter.

\item[{\bf Theme 2:}] When analyzing Internet measurements, examining
  the ``hygiene'' of the available measurements (\ie an in-depth
  recounting of the potential pitfalls associated with producing the
  measurements in question) is critical.

\item[{\bf Theme 3:}] When validating proposed topology models, it is
  necessary to treat network modeling as an exercise in
  reverse-engineering and not as an exercise in model-fitting.

\item[{\bf Theme 4:}] When modeling highly-engineered systems such as
  the Internet, beware of M.L. Mencken's quote ``For every complex
  problem there is an answer that is clear, simple, and wrong.''

\end{description}

We have not tried to survey the entire literature in this area, and we
apologize to those whose work has not appeared here, but there are
other extant surveys mentioned at the relevant points throughout this
chapter, for specific components of the work. We also have not tried
to critique every model, but rather tried to provide general guidance
about modelling. It is intended that the readers could themselves
critique existing and new models based on the ideas presented here.

In addition, we do not claim to have covered every type of topology
associated with the Internet. Specifically, we have avoided topologies
at the applications layer, for instance those associated with the WWW
or online social networks.  We made this choice simply because these
topologies are (despite being ``Internet'' topologies) profoundly
different from the topologies we have included. They are almost purely
virtual whereas all of the networks considered here have a physical
component, which leads to the arguments for optimization as their
underlying construction.  An important open problem in this context is
the role that societal-related factors play over more economic- or
technology-based drivers in the formation and evolution of these
virtual topologies.

Finally, in each section, we have aimed at illuminating some of the
current problems and identifying hopefully fruitful directions for
future research in this area.

### 2013-07-22 Compiled/chapter7/intro.tex
all files compile correctly

\clearpage
\section{Introduction}

Internet topology research is concerned with the study of the various
types of connectivity structures that are enabled by the layered
architecture of the Internet. More than a decade of Internet topology
research has produced a number of high-profile "discoveries" that
continue to fascinate the scientific community, even though (or,
especially because) they have been simultaneously touted by different
segments of that community as either seminal, controversial, seriously
flawed, or simply wrong.  Among these highly-popularized discoveries
are the observed power-law relationships of the Internet topology, the
network's scale-free nature, and its extreme vulnerability to attacks
that target the highly-connected nodes in its core (\ie the
Achilles' heel of the Internet).

The purpose of this chapter is to bring order to the current state of
Internet topology research and separate ``the wheat from the chaff''.
In particular, by relying on carefully vetted data and readily
available domain knowledge, we re-examine the reported discoveries and
expose them to higher standards with respect to statistical inference
and model validation.  In the process, we reveal the superficial
nature of many of these discoveries and provide alternative solutions
that reflect networking reality and do not collapse under scrutiny
with high-quality data or when examined in detail by domain experts.

\subsection{The many facets of Internet connectivity}

Internet topology research is concerned with the study of the various
types of connectivity structures that are enabled by the layered
architecture of the Internet.  These structures include the inherently
physical components of the Internet's infrastructure (\eg routers
and switches and the fiber cables connecting them) as well as a wealth
of more logical topologies that can be defined and studied at the
higher layers of the Internet's TCP/IP protocol stack (\eg IP-level
graph, AS-level network, Web-graph, P2P networks, Online Social
Networks or OSNs).

As early as the ARPANET\index{ARPANET}, researchers were drawing maps
of the network representing its
connectivity~\cite{cerf90:_selected_maps}.  The earliest date to
1969. In those days, the entire network was simply enough to draw on
the back of an envelope\footnote{For instance see
  \url{http://personalpages.manchester.ac.uk/staff/m.dodge/cybergeography/atlas/roberts_arpanet_large.gif}},
and accurate maps could be drawn because every piece of equipment was
expensive, installation was a major task, and only a few people worked
on the network.

As the network grew, its complexity also grew, until the point where
no one person could draw such a map. At that point, automated
strategies started to arise for measuring the topology. The earliest
Internet topology studies date back to the time of the
NSFNET\index{NSFNET} and focused mainly on the network's physical
infrastructure consisting of routers, switches and the physical links
connecting them (\eg
see~\cite{Calvert97,pansiot98:_inter}). The decommissioning of
the NSFNET around 1995 led to a transition of the Internet from a
largely monolithic network structure (\ie NSFNET) to a genuinely
diverse "network of networks."  Also known as Autonomous Systems
(ASes), together these individual networks form what we now call the
"public Internet" and are owned by a diverse set of organizations
and companies that includes large and small Internet Service Providers
(ISPs), transit providers, network service providers, Fortune 500
companies and small businesses, academic and research organizations,
content providers, Content Distribution Networks (CDNs), Web hosting
companies, and cloud providers.

With this transition came an increasing fascination of the research
community with a largely economics-driven connectivity structure
commonly referred to as the Internet's AS-graph; that is, the logical
Internet topology where nodes represent individual ASes and edges
reflect observed relationships among the ASes (\eg
customer-provider, peer-peer, or sibling-sibling relationship). It is
important to note that the AS-graph\index{AS-graph} says little about
how two ASes connect with one another at the physical level; in
particular, it says nothing about if or how they exchange actual
traffic.  Nevertheless, starting shortly after 1995, this fascination
with the AS-graph has resulted in thousands of research publications
covering a range of aspects related to measuring, modeling, and
analyzing the AS-level topology of the Internet and its evolution over
time~\cite{govindan97:_topology,roughan11}.

At the application layer, the emergence of the World Wide Web (WWW) in
the late 1990 as a killer application generated general interest in
exploring the Web-graph, where nodes represent web pages and edges
denote hyperlinks~\cite{Broder:2000:GSW:346241.346290}. While this overlay network
or logical connectivity structure says nothing about how the servers
hosting the web pages are connected at the physical or AS level, its
scale and dynamics differ drastically from its physical-based or
economics-driven underlays -- a typical Web-graph has billions of
nodes and even more edges and is highly dynamic; a large ISP's
router-level topology consists of some thousands of routers, and
today's AS-level Internet is made up of some 30,000-40,000 actively
routed ASes and an order of magnitude more links.

Other applications that give rise to their own "overlay" or logical
connectivity structure and have attracted some attention among
researchers include email and various P2P systems such as Gnutella,
Kad, eDonkey, and BitTorrent\index{P2P}\index{BitTorrent}.  More
recently, the enormous popularity of Online Social Networks (OSNs) has
resulted in a staggering number of research papers dealing with all
different aspects of measuring, modeling, analyzing, and designing
OSNs\index{OSN}.  Data from large-scale crawls or, in rare
circumstances, OSN-provided data have been used to examine snapshots
of many real-world OSNs or OSN-type systems, where the snapshots are
generally simple graphs with nodes representing individual users and
edges denoting some implicit or explicit friendship relationship among
the users.

\subsection{Many interested parties with different objectives}

The above-mentioned list of possible connectivity structures that
exist in today's Internet is by no means complete, but illustrates how
these structures arise naturally within the Internet's layered
architecture. It also highlights the many different meanings of the
term ``Internet topology,'' and sensible use of this term requires
explicitly specifying which facet of Internet connectivity is
considered because the differences are critical.

The list also reflects the different motivations that different
researchers have for studying Internet-related graphs or networks.
For example, engineers are mainly concerned with the physical facets
of Internet connectivity, where technological issues generally
dominate over economic and social aspects.  However, the more
economics-minded researchers are particularly interested in the
Internet's AS-level structure where business considerations and market
forces mix with technological innovation and societal considerations
and shape the very structure and evolution of this logical topology.
Moreover, social scientists see in the application-level connectivity
structures that result from large-scale crawls of the various OSNs new
and exciting opportunities for studying different aspects of human
behavior and technology-enabled inter-personal communication at
previously unheard of scale.

In addition, mathematicians are interested in the different
connectivity structures mainly because of their many novel features
and properties that tend to require new and creative modeling and
analysis methodologies. From the perspective of many computer
scientists, the challenges posed by many of these intricate
connectivity structures are algorithmic in nature and arise from
trying to solve specific problems involving a particular topological
structure. For yet another motivation, many physicists turned
network scientists see the Internet as one of many examples of
large-scale complex networks that awaits the discovery of universal
properties that do not depend on system-specific details and advance
our understanding of these complex networks irrespective of the domain
in which they arose in the first place.

\subsection{More than a decade of Internet topology research}

When trying to assess the large body of literature in the area of
Internet topology research that has accumulated since about 1995 and
has experienced enormous growth especially during the last 10+ years,
the picture that emerges is at best murky.

On the one hand, there are high-volume datasets of detailed network
measurements that have been collected by domain experts.  These
datasets have been made publicly available so other researchers can
use them.  As a result, Internet topology research has become a prime
example of a measurement-driven research effort, where third-party
studies of the available datasets abound and have contributed to a
general excitement about the topic area, mainly because many of the
inferred connectivity structures have been reported to exhibit
surprising properties (\eg power-law relationships for inferred
quantities such as node
degree~\cite{faloutsos99:_power_law_relat_of_inter_topol}). In turn,
these surprising discoveries have led network scientists and
mathematicians alike to develop new network models that are provably
consistent with some of this highly-publicized empirical evidence.
Partly due to their simplicity and partly due to their strong
predictive power, these newly proposed network models have become very
popular within the larger scientific
community~\cite{barabasi00,barabasi09:_scale_free_networ,barabasi12}. For
example, they have resulted in claims about the Internet that have made
their way into standard textbooks on complex networks, where they are
also used to support the view that a bottom-up approach dominated by
domain-specific details and knowledge is largely doomed when trying to
match the insight and understanding that a top-down approach centered
around a general quest for "universality" promises to
provide~\cite{barabasi02:_linked,dorogovtsev03:_evolut_networ,pastor-satorras04:_evolut_struc_inter,newman10:_networ}.

On the other hand, there is a body of work within the networking
research literature that argues essentially just the opposite and
presents the necessary evidence in support of a inherently
engineering-oriented approach filled with domain-specific details and
knowledge~\cite{Li04,Alderson05,willinger09:_mathem_and_inter}. In contrast to being
measurement-driven, this approach is first and foremost concerned with
notions such as a network's purpose or functionality, the hard
technological constraints that the different devices used to build a
network's physical infrastructure have to obey, or the sources of
uncertainty in a network's "environment" with respect to which the
built network should be robust.  As for the measurements that have
been key to the top-down approach, the reliance on domain knowledge
reveals the data's sub-par quality and highlights how errors of
various forms occur and can add up to produce results and claims that
create excitement among non-experts but quickly collapse when
scrutinized or examined by domain experts. While there exist currently
no textbooks that document these failures of applying detail- and
domain knowledge-agnostic perspective to the Internet, there is an
increasing number of papers in the published networking research
literature that detail the various mis-steps and show why findings and
claims that look at first glance impressive and conclusive to a
science-minded reader turn out to be simply wrong or completely
meaningless when examined closely by domain
experts~\cite{willinger09:_mathem_and_inter,Alderson10,krishnamurthy11:_what}.

In short, a survey of the existing literature on Internet topology
research leaves one with the distinct impression that ``too many cooks
spoil the broth.''  We hope that in the not-too-distant future, this
impression will be replaced by ``many hand make light work'', and we
see this chapter as a first step towards achieving this goal.

\subsection{Themes}

In writing this chapter there are a number of themes that emerge, and
it is our intention to highlight them to bring out in the open the
main differences between a detail-oriented engineering approach to
Internet topology modeling versus an approach that has become a
hallmark of network science and aims at abstracting away as many
details as possible to uncover ``universal'' laws that govern the
behavior of large-scale complex networks irrespective of the domains
that specify those networks in the first place.

\vspace{5mm}

\begin{description}
\item[{\bf Theme 1:}] When studying highly-engineered systems such as
  the Internet, ``details'' in the form of protocols, architecture,
  functionality, and purpose matter.

\item[{\bf Theme 2:}] When analyzing Internet measurements, examining
  the ``hygiene'' of the available measurements (\ie an in-depth
  recounting of the potential pitfalls associated with producing the
  measurements in question) is critical.

\item[{\bf Theme 3:}] When validating proposed topology models, it is
  necessary to treat network modeling as an exercise in
  reverse-engineering and not as an exercise in model-fitting.

\item[{\bf Theme 4:}] When modeling highly-engineered systems such as
  the Internet, beware of M.L. Mencken's quote ``For every complex
  problem there is an answer that is clear, simple, and wrong.''

\end{description}

### 2013-07-22 Compiled/chapter7/pop.tex
all files compile correctly

\clearpage
\section{PoP-level topology}

When designing or reconfiguring the physical infrastructure of an ISP, network 
operators are often guided by a design principle that emphasizes 
{\em hierarchy}~\cite{Cisco05,Gill,Morris07}.  There are two main reasons 
for implementing hierarchical network designs: {\em scalability} and {\em simplicity}.  
Compared to non-hierarchical designs, hierarchical networks can often be built at scale, 
mainly because hierarchy makes a network easier to visualize --- a key feature towards 
making it easier to manage. The situation is analogous to modularity in programming 
languages --- ideally it allows consideration of network components in isolation.

A common form of hierarchy in IP networks is based on the concept of the PoP (or 
Point of Presence). A PoP is a loosely defined term. Some providers may use the 
term to mean a physical building (housing a group of routers, switches and other 
devices), whereas others mean a metropolitan area where service is provided. 
However it is defined, though, it is a useful construct because it describes 
the logical structure of the network as the designer intended, rather than its 
particular implementation in terms of individual routers. Moreover, irrespective 
of the meaning, PoPs have an explicit geography (\eg street address or 
city/metropolitan area).  This then leads to our third major category 
of ``Internet topology'' --- the PoP-level topology.

PoP-level topologies are ideal for understanding tradeoffs between connectivity 
and redundancy, and also provide the most essential information to competitors 
or customers (about where a network is based, or who has the best access network 
in a region).  Additional reasons why the PoP-level view of networks is interesting 
include

\begin{itemize}

\item Network maps are often drawn at this level because it is an easy level 
for humans to comprehend.

\item Network optimization is often conducted at this level because the problem 
size is generally reasonable (\eg dozens of PoPs as compared to potentially 
hundreds of routers) and because inter-PoP links are much more expensive than 
intra-PoP links.

\item The internal design of PoPs is almost completely determined by simple 
templates \cite{Cisco05,Gill,Morris07}.

\item Networks change less frequently at the PoP level than at the router 
level~\cite{Shavitt10}; and

\item The PoP level is the more interesting level for many activities because 
it is less dependent on the details of protocol implementations, router vendor 
and model, and other technological details.

\end{itemize}

\begin{figure}[!thbp] 
  \begin{center}
    \begin{subfigure}[b]{0.58\textwidth}
      \centering
      \includegraphics[width=1\textwidth]{Abilene.jpg}
    \end{subfigure}
    \hspace{0.02\textwidth}
    \begin{subfigure}[b]{0.38\textwidth}
      \centering
    \includegraphics[width=1\columnwidth]{Geant2012.jpg}
    \end{subfigure}
    \caption{PoP-level network topologies taken from \url{www.topology-zoo.org}.
      \label{fig:pop}}
  \end{center}
\end{figure}

The last point is subtle but important for modelling. For instance, when using a 
network as part of a simulation, one would like to have a network that is 
{\em invariant} to the method being tested. If a network designer might change 
his/her network in response to a new protocol, say a routing or traffic engineering 
algorithm, then the test will be ambiguous if it uses existing networks as models. 
PoP-level networks are less sensitive to these details than router-level networks, 
because routers impose physical and technological constraints that are almost 
completely dependent on the details of the router vendor, model and even the 
version of software running on the router.

Two examples of PoP-level topologies are depicted in \autoref{fig:pop}, 
showing the structure of two of the largest research networks (Abilene, 
and G\'{E}ANT) in the world.

\subsection{A look back}

The study of PoP-level topologies has a briefer history than the major 
alternatives. Although the concept has existed for almost as long as networks, 
the work on modelling and measurement has typically focussed on routers (or 
their equivalent) though it is noteworthy that in simple networks (with one 
router per PoP) the two are the same.

The first steps were taken when real data-networks were observed and
it was noted that they had structure in the form of hierarchy that was
not well represented in simple random graphs.  This observation led to
the development of \emph{structural topology
  generators}\cite{Zegura97}, based on the idea that a topology
generator should reflect the obvious hierarchical structures visible
in real networks (\eg the Georgia Tech Internetwork Topology Models
(GT-ITM)~\cite{GTITM} generator). However, this model was not
specifically aimed at modelling the PoP-level.  PoPs have been used in
more advanced structural topology generators such as
IGen~\cite{quoitin09:_igen}. IGen explicitly treats network design as
an optimization, rather than following simple stochastic rules, in
order to mimic the manner in which real networks are designed.  IGen
uses this not only for topology but also to create some of the other
important details of the network, such as the link metrics or iBGP
configurations.

The Rocketfuel project~\cite{rocketfuel_1} sought to measure (using traceroute 
with all the problems described earlier) individual ISPs, and as part of this, 
presented data and maps at the PoP level. The idea was extended by the iPlane 
project \cite{iPlane}, and by DIMES
\cite{Shavitt:2012:GIP:2238856.2238872}. However, the flaws in
traceroute make this data and the ensuing maps suspect from the start. 
However, \cite{rocketfuel_1} raised the bar with respect to validating the 
obtained PoP-level maps by trying to solicit feedback from the operators in 
charge of the mapped ISPs.

More recently, the concept of a PoP has been explicitly used to help
guide measurement approaches, in the hope to overcome some of the
limitations of traceroute~\cite{Feldman08,Yoshida09,Shavitt10}.
However, in the absence of strong validation and given traceroute's
many problems, it is likely that most of the known issues still exist
in these approaches.

Alternatively, Knight~\etal~\cite{Zoo} have collected a set of over 
200 maps published by ISPs themselves, and transcribed these into an open 
set of data available from \url{www.topology-zoo.org}. Much of this data is 
at the PoP-level, indicating the importance of this level of network 
representation to ISPs. For a similar but complementary effort, see~\cite{atlas}.

\subsection{Know your measurements}

The problems in measuring PoPs are essentially the same as in any
traceroute-based survey, though it is thought (or perhaps just hoped)
that mapping IP addresses to PoPs is more straight-forward than
mapping IPs to either routers or to ASes. To the best of our
knowledge, no rigorous testing of this hypothesis has been conducted
to date, but there are some indications (\eg see \cite{Zoo}) that the
PoP-level maps provided by traceroute are no better than their router-
or AS-level counter-parts.

The topology-zoo data~\cite{Zoo}, on the other hand, is provided by
ISPs themselves and should in principle be much more accurate.
However, this dataset must also be treated carefully (as should all
data) because of potential transcription errors, or mistakes or
approximations in the maps provided by the ISPs. While such errors are
much less frequent than those encountered in measured networks, an
added concern with respect to ISP-provided maps is stale data because
there is little incentive to provide up-to-date maps.

As mentioned earlier, another important component of many PoP-level
topologies is the geographic element. Such topologies are much easier
to visualize geographically \cite{knight12:_i}, and so a frequent
interest is geolocation of the PoPs. While this is not a chapter on
geolocation, it suffices to say that many research papers have been
written on the problems of accurately mapping IP addresses to their
geolocation (\eg see \cite{Poese:2011:IGD:1971162.1971171}). Moreover,
while the routers and switches of a PoP are typically located in a
single location, city, or metropolitan area, the ``eyeballs'' (\ie end
users) connected to the PoP will be spread over some
area~\cite{rasti10:_eyebal_ases}.  However, if the researcher is
willing to diligently mine various data sources, there is hope of at
least being able to geolocate PoPs as they house potentially hundreds
or thousands of IP addresses and reside in locations with known
physical addresses (\eg carrier hotels)~\cite{atlas}.

\subsection{The nature of the PoP-level graph}

There is a common meme in network modelling that the design of a US ISP backbone 
network involves simply selecting the NFL cities, and then joining them up with 
a few lines on a map. While the real process of network design is rarely so trite, 
the picture above isn't entirely unfair.

Most notably, PoPs are usually selected based on commercial criteria (\eg the 
desirability and size of the potential customer base in an area). So network 
engineers get little choice over the locations that they are connecting. They 
could be the NFL cities in the US, or the larger cities of another country, or 
the capitals of countries on a continent, and so on. 
Once the PoP locations have been selected, they need to be connected in some 
redundant fashion to ensure some degree of robustness to node or link failures. 
Historically, connecting these PoPs may have been done in a mostly {\em ad hoc} 
manner; see for instance \url{http://personalpages.manchester.ac.uk/staff/m.dodge/cybergeography/atlas/roberts_arpanet_large.gif}.

However, since the burst of the Internet bubble ({\em circa} 2000), capital 
investment has become harder to obtain, and network operators and engineers 
had to justify such investment more carefully. At this point, network capacity 
started to be more carefully planned, not always using formal mathematical 
optimization, but certainly using traffic measurements to ensure less wasted 
capacity. Much of this planning and optimization is performed at the PoP-level 
simply because the router-level is much more complicated (in size, and complexity of
constraints), and because inter- and intra-PoP link costs vary by a large margin.

We have discussed network design by constrained optimization extensively in 
\autoref{sec:router} (see also references such as \cite{Li04}), and so here 
we shall only consider the main differences for PoP-level design (apart from 
those already listed above).

Perhaps the most important difference is that the physical and engineering 
constraints on a router do not directly apply for a PoP. At least in theory, 
a PoP can use as many routers as needed to provide sufficient number of ports 
for any arbitrary node degree and sufficient throughput per port. Naturally, 
the constraints in this case will arise in the form of costs, and optimization-based 
formulations of the PoP-level network design problem will feature budget constraints 
to reflect this aspect.  As budget constraints can vary greatly among different 
companies, when we look at actual PoP-level ISP backbone networks, we see a wide 
variety of designs ranging from the meshy designs with high-degree nodes only at 
the edge predicted by the HOT model, to hub and spoke like networks~\cite{Zoo}. 
In fact, the sheer variety of network designs we observe in reality suggest that 
while some network operators seem to aim at optimizing performance (given some 
lenient cost constraints), others appear to be willing to sacrifice performance 
in order to keep costs low. Moreover, network operators in different countries 
can encounter very different link costs depending on the local geographic and 
commercial environment.

A critical but rarely discussed property of the PoP-level topology is that it 
provides the ``glue" between the more physical router-level topology on the one 
hand and the more logical AS-level topology on the other hand. Functioning as 
an ``intermediary" between these two topologies highlights the important aspect 
of the PoP-level topology that its granularity is ``just right" for many networking 
problems of practical interest --- not too coarse as to ignore important context 
(\eg the case with various AS topologies) but also not too fine as to be 
overwhelmed with unnecessary details (\eg the case with various physical 
topologies). We next discuss this property in more detail.

\subsubsection{From PoP-level to router-level connectivity}

Given a PoP-level network, there is an additional interesting
question: ``Can we map this network down to the router level?''  The
GT-ITM model addressed this through random generation of its
subnetworks, but in practice the design process of network engineers
in this case is a text-book application of repeating
patterns~\cite{Cisco05,Gill,Morris07} and hence anything but random.

The main reason for following this design process is that network 
designers often apply ``cookie cutter'' methods to design networks as a whole
or the internals of PoPs, though that term unnecessarily
trivializes the importance of repeated patterns. Repetition makes
network operations vastly simpler: the management of two PoPs requires
the same skills.  Equipment can be bulk purchased, debugging is
easier, and adding new PoPs is simpler. Finally, networks based on
templated design lead to simple design methodologies. For instance,
the inter-PoP level network topology can be optimized relatively
simply, as details such as redundancy will be supplied by the
provision of pairs of redundant routers in each PoP, with redundant
links between them.  Design often refers to the graph topology of
router interconnections, but templated design can extend to other
details, such as physical configuration within racks, connections with
external networks, or additional servers such as Domain Name Service
or Network Management Systems.

This type of design can be mathematically described using
graph-products, though for more details, we invite the reader to consult
\cite{parsonage11:_gener_graph_produc_networ_desig_analy}.

\subsubsection{From PoP-level to AS-level connectivity: The pancake-view of the Internet}

Until now we have only really discussed the PoP-level topology of a
single network. However, there is considerable interest in how these
networks interconnect.

The most prominent and commonly-accepted view of the Internet is as a
``network of networks'' or ASes in the AS-graph representation
discussed in \autoref{sec:as}.  A much neglected and rarely-mentioned
representation is the ``pancake view'' where we consider each network
to be a layer and where the different layers (networks) are stacked on
top of one another to form a pancake-like
structure~\cite{liljenstam03:_devel_of_inter_backb_topol}. To show
where the different networks inter-connect, we add links across
layers; intra-network connectivity is shown as links within each
layer.  For a set of ``peer'' networks, one advantage of this pancake
view is that these networks often cover similar geographic areas and
inter-connect in multiple locations, but at a limited set of cities
(determined either by where private interconnects are seen as
commercially viable, or where IXPs are available). Importantly,
depending on the types of networks, many of them host their PoPs in
one and the same commercial colocation facilities whose street
addresses are generally known\footnote{One problem in establishing
  such a view lies in the limitations of current IP geolocation
  services \cite{Poese:2011:IGD:1971162.1971171}.}.  As such, the
pancake view allows one to visualize not only this connectivity inside
and between such providers but also the geography of their PoP-level
topologies.

However, as far as we are aware, there has been no significant work
studying this pancake view together with the different inter-connections, 
other than noting that it exists. The dearth of studies and models perhaps 
stems from the problems in obtaining the measurements necessary for constructing
this view (see the discussions in \autoref{sec:as}), but it is
perhaps one of the most interesting areas for future Internet topology research.

\subsection{A look ahead}

We have focused in this section and the earlier sections mostly on traditional ISPs
or network service providers which operate networks that have more or
less pronounced backbones and cover geographic areas ranging from 
individual countries to entire geographic regions to the entire globe.
However, there are many other networks that are not ISPs and consist of PoPs
without their own physical infrastructures to connect them (\eg content
providers, CDNs, Web-hosting companies). The PoPs of these networks are
typically located in commercial colocation facilities or data centers
that are operated by third-party companies for the explicit purpose of 
interconnecting such infrastructureless networks among one another or 
with ISPs or network service providers.

The importance of the role of such dedicated Internet infrastructure
in the form of colocation facilities is best illustrated with a
concrete example.  As of December 2012, Equinix \url{www.equinix.com},
one of the leading companies in global interconnection and data
centers, owned and operated in 14 countries in 5 continents some 30
colocation facilities. In these 30+ colocation facilities that are
located in the major cities around the world, more than 4,000 networks
connected directly to their customers and partners.

Another largely under-researched topic concerns the fact that in this chapter,
we treated router-, PoP-, and AS-level topologies as static objects, when in
reality, they evolve over time.  In particular, it is rarely the case that
a network operator designs a new network from scratch.  Network design typically
has to include as important aspects awareness and knowledge of the 
existing network that the operator intends to change due to, for example,
drastic changes in the traffic demand that the original network design can
no longer handle efficiently.

In view of these an other added challenges, the PoP-level view promises
to be one of the more useful and interesting direction for future Internet
topology research.  In particular, measurements and models at this level have
considerable scope for the future, and extensions of HOT-like optimization
models may provide much more realistic synthetic networks than are
currently available. At the same time, work on interconnection of networks 
at this level also provides considerable scope, but will require significant
advances in our ability to measure and model the flow of traffic across the
different networks as it is the traffic that ultimately determines much of
the structure and evolution of the different topologies that we discussed 
in this chapter.

\subsection{Notes}

The primary source for the material presented in this section (and a
much lengthier discussion of many of the issues) is

\begin{itemize}
\item[\cite{Zoo}] S.Knight, H.Nguyen, N.Falkner, R.Bowden, M.Roughan,
  The Internet topology zoo. IEEE Journal on Selected Areas in
  Communications (JSAC) 29, 9, 1765-1775, October 2011.

\end{itemize}

\noindent For additional and more in-depth reading materials (in
addition to the references indicated throughout) we point to

\begin{itemize}

\item[\cite{Shavitt:2012:GIP:2238856.2238872}] Y.Shavitt and
  N.Zilberman, Geographical Internet PoP-level maps. In Proceedings of
  the 4th international conference on Traffic Monitoring and Analysis
  (Berlin, Heidelberg), TMA'12, Springer-Verlag, pp. 121-124, 2012.

\item[\cite{parsonage11:_gener_graph_produc_networ_desig_analy}]
  E.Parsonage, H.Nguyen, R.Bowden, K.Knight, N.Falkner, M.Roughan,
  Generalized graph products for network design and analysis. In 19th
  IEEE International Conference on Network Protocols (ICNP)
  (Vancouver, CA), October 2011.

\item[\cite{Shavitt10}] Y.Shavitt and N.Zilberman, A structural
  approach for PoP geo-location. In IEEE Infocom 2010.

\item[\cite{Yoshida09}] K.Yoshida, Y.Kikuchi, M.Yamamoto, Y.Fujii,
  K.Nagami, I.Nakagawa and H.Esaki, Inferring PoP-level ISP topology
  through end-to-end delay measurement. In PAM, pp. 35-44, 2009.

\end{itemize}

### 2013-07-22 Compiled/chapter7/primer.tex
all files compile correctly

\clearpage
\section{Primer}

We start first by defining some common ideas, motives, and problems
within the scope of network topology modelling.

\subsection{A Graph Primer}

In this context \define{topology} usually refers to the structure of
the \define{graph} representation of a network. That is, the common
notion used to describe network topology is the mathematical {\em
  graph}. \index{graph} A graph ${\cal G}$ is defined by a set of
nodes ${\cal N}$ (often called vertices) and edges (or links) ${\cal
  E} \subset {\cal N} \times {\cal N}$, so we usually write ${\cal G}
= ( {\cal N}, {\cal E})$.  Here, we shall denote the number of nodes
$N = |{\cal N}|$, and the number of edges $E = |{\cal E}|$.

Nodes are usually associated with some logical of physical structure
in a network: a router, switch, PoP, or AS. Edges are associated with
the appropriate type of logical or physical link between these nodes.

A graph describes connectivity between logical resources such as
routers, or IP address, but simple connectivity is rarely as useful as
when additional information such as names, capacity or distance are
attached to these abstract objects. Such can easily be included in
these descriptions by creating labelling functions of the node or edge
sets, in the form: $f: {\cal N} \rightarrow \R$ or $f: {\cal E}
\rightarrow \R$ in the case of real-valued labels. We could similarly
defined labelling functions with text labels, or integer or vector
values, and so on. However, it is naive to treat labels as an
``add-on'' as they carry semantics that can be important in the
network. For instance, when we define link distances (be these
geographic or semantic), that can change the notion of distance in the
network as a whole.

We can also define functions of groupings of nodes or edges, though in
this case it is not as conceptually obvious why we might. However, an
exemplary case is that of ``on-net'' where we might define a function
that classifies pairs of nodes as on the same subnet or not. Thus,
such functions can ascribe meaning to groupings of nodes.

Many of the Internet graphs have symmetric links (that is, if
$i\rightarrow j$ is a link, then $j \rightarrow i$ is also a link) and
so these networks are \define{undirected}, but we also need sometimes
to represent asymmetric links, and do so with a \define{directed
  graph} or \define{digraph}, and we call the links in such a digraph
\define{arcs}.

In the study of network topology we might come across the more
generalized graph concepts of the \define{multi-graph} and
\define{hyper-graph}. 
\begin{itemize}

  \item \define{hypergraph}: links connect more than two nodes
    \begin{itemize}
    \item \eg where you have a connective medium (rather than a
      wire), for instance in a wireless network.
    \end{itemize}

  \item \define{multigraph} or \define{pseudograph}: has multiple
    parallel links between two nodes
    \begin{itemize}
    \item \eg it is easy to have two links between two routers.
    \end{itemize}

\end{itemize}
We'll exclude these cases unless explicitly stated, but it is worth
noting that each of these do apply to particular aspects of the
Internet.

We say two nodes are \define{connected} if a path exists between them,
and that a graph is connected if all pairs of nodes are connected. A
graph is $k-$node connected if the graph remains connected after the
removal of any set of $k-1$ or fewer nodes (and corresponding links)
and $k-$edge connected if the graph remains connected after the
removal of any $k-1$ edges.

For an undirected graph $G$, define the \define{neighborhood} of node
$i$ by
     \[ N_i = \big\{ j \mid (i,j) \in E \big\}, \]
\ie the set of adjacent nodes to $i$, and we define the degree of
the node to be  the number of elements in the neighborhood to be 
     \[ k_i = |N_i |. \]
In a directed graph, we define two concepts: the \define{in-degree}
(the number of links connecting to the link) and the
\define{out-degree} (the number of links originating from it). 
\begin{eqnarray*}
  \mbox{ \rm in-degree}(i) & = & \left|  \{ (j,i) | (j,i) \in
                                     {\cal E}  \}   \right|, \\
  \mbox{ \rm out-degree}(i) & = & \left| \{ (i,j) | (i,j) \in
                                     {\cal E}  \}   \right|.
\end{eqnarray*}
We often consider statistics of the degree distribution $p_k$ (which
gives the probability that a node has degree $k$), the average node
degree being the most obvious such. It can be easily calculated from
the sum of degrees, which has the interesting property
    \[ \sum_{i \in N} k_i = 2 |E|, \]
generally referred to the Handshake lemma.

The node-degree distribution provides a common characterization of a
graph (though by no means a complete characterization).  It is
noteworthy, however, that although this distribution is frequently
discussed, the concept is somewhat ill-defined. It can be directly
measured for a real network, in which case $p_k$ is the probability
that a randomly selected node from the measured graph has degree
$k$. However, it is often used in the context of a set of simulated
graphs, where it is used to mean the probability that a node in the
ensemble of networks has degree $k$ with this probability. The
difference is subtle, but it is worth keeping track of such
discrepancies.

% Higher order statistics of the node degrees area also examined in some
% cases. In general, we might look at the ...

% However, there are some simple terms for some of these, perhaps the
% most noteworthy be \define{assortativity}, which is destined by 
% \[ r = \frac{\sum_{j,k} j k ( e_{j,k} - q_j q_k) }{\sigma^2_q},
% \]
% where $e_{i,j}$ is the fraction of edges that connect a vertex of type
% $i$ to one of type $j$, or the joint probability distribution of the
% remaining degree at either end of a randomly chosen link so
% $\sum_{j,k} e_{j,k} = 1$. We define $q_k$ by 
% \[ \sum_{j} e_{j,k} = q_k. 
% \] 
% $\sigma_q^2$

% For the statisticians, assortativity can be thought of as Pearson's
% correlation coefficient of remaining degrees at either ends of a
% random edge.

% Assortativity ranges from $-1$ to $1$ with the following meanings:
%   \begin{itemize}
%   \item $r$ near 1 -- high degree nodes often connect to high degree nodes,
%   \item $r$ near -1 -- high degree nodes often connect to low degree nodes.
%   \end{itemize}

% The above metrics are meaningful in the sense that they tell us the
% number of links per node, but often more important for engineers are
% distances.

There are many other graph {\em metrics}. For instance, the
\define{distance}\footnote{The graph distance has a long history. In
  mathematics, perhaps the best known example is the Erd\H{o}s number,
  which is the distance of a author from Erd\H{o}s in the
  co-authorship graph. In popular culture there is an equivalent: the
  Bacon number, or the distance between actors in the graph of
  co-appearances.} between two connected nodes in an unweighted graphs
is generally defined to be the number of edges in the shortest path
connecting them. We can then examine quantities such as the average
distance, and the \define{diameter} of the network (the maximum
distance).

% Erd\H{o}s numbers: if you wrote a paper with Erd\H{o}s,
%   your number is 1. If you wrote a paper with a direct co-author, your
%   number is two, and so on. Essentially it is the graph distance you
%   are from Erd\H{o}s in a co-authorship graph.\\[2mm]

% \url{http://en.wikipedia.org/wiki/Erdos_number}\\[2mm]

% My Erd\H{o}s number is 4 (through Charles Pearce, Gavin Brown, and
% Robert Tijdeman.)\\[2mm]

% \url{http://www.ams.org/mathscinet/collaborationDistance.html}

  % \begin{itemize}

  % \item radius of a graph is the minimum eccentricity of any vertex
  %   \[ \mbox{radius}(G(N,E)) = \min_{i \in N} \epsilon(i). \]

  % \item diameter of a graph is the maximum eccentricity of any vertex
  %   \[ \mbox{diameter}(G(N,E)) = \max_{i \in N} \epsilon(i). \]
  %   which is the maximum distance between any pair of nodes.

  % \item a peripheral vertex is one whose eccentricity achieves the
  %   diameter.

  % \end{itemize}

And there are many other metrics: assortativity, clustering
coefficient, centrality, and so on. They are all attempts to capture
the nature of a graph in a small set of measures, and as such provide
simpler, seemingly more intuitive ways to consider graphs.  For other
discussions of these, and comparisons in the context of Internet
topologies see \cite{jamakovic08,haddadi08:_networ_topol}.  We must be
wary though, as it should be clear that the potential for problems is
immediate. No small set of numbers can truly represent graphs. For
instance, consider the Hamiltonian cycle\footnote{A Hamiltonian cycle
  is a path (on the graph) that visits each node exactly once, and
  then returns to the start point.} problem. The problem of
determining if a network has such a path is well known to be
NP-complete, and as such no small set of statistics of the graph will
provide a characterization that is sufficient to consider this
problem. Thus, these simple statistics must miss important properties
of the network.

% centrality

%   \begin{itemize}

%   \item Measures the ``centrality'' of a vertex in a graph

%   \item Different measures
%     \begin{itemize}
%     \item degree centrality (normalized degree) of nodes 
%       \begin{itemize}
%       \item interpretation --- how likely to catch a disease
%       \item extension to a metric on a graph (maximized by star)
%       \end{itemize}

%     \item closeness centrality 
%       \begin{itemize}
%       \item reciprocal of mean geodesic distance between $v$ and other nodes
%       \end{itemize}

%     \item between centrality
%       \begin{itemize}
%       \item normalized measure of how many shortest-paths a vertex
%         appears on
%       \end{itemize}

%     \end{itemize}

%   \end{itemize}

% clustering coefficient

%   \begin{itemize}

%   \item global measure of whether nodes tend to cluster
%   \[ c = 3 t_1 / t_2 , \]
% where
% \begin{eqnarray*}
%   t_1 & = & \mbox{number of triangles} \\
%   t_2 & = & \mbox{number of connected triples}
% \end{eqnarray*}

%   \item local measure of how close a node and its neighbors are to
%     being a clique
% \vspace*{-1mm}
% \[ c_i = \frac{\left| \{ (j,k)\in E | j,k \in N_i  \} \right|}{k_i  (k_i-1) /2}, \]
% where $N_i$ is the neighborhood of $i$, and $k_i=|N_i|$.

%   \end{itemize}

% $c_i$ counts the fraction of links in the local neighborhood, as
% compared with a clique.\\[2mm]

% we can compute a network average clustering co-efficient using
% \[ \bar{C} = \frac{1}{n} \sum_{i=1}^{n} c_i . \]

It may be useful to the reader to consider some of the tools that are
available for working with graphs. They have different sets of
feature, but perhaps the most important is whether they are used in
conjunction with a programming language and which one, so we have
listed a (no doubt incomplete) set below with some very basic
information.
  \begin{itemize}
\setlength{\itemsep}{-1mm}
\setlength{\topsep}{-1mm}

  \item MatlabBGL \url{http://www.stanford.edu/~dgleich/programs/matlab_bgl/}
    \begin{itemize}
    \item Graph libraries for Matlab, 
    \item using Boost Graph Library (BGL) \\
 {\url{http://www.boost.org/doc/libs/1_42_0/libs/graph/doc/index.html}}
    \end{itemize}

  \item igraph \url{http://igraph.sourceforge.net/}
    \begin{itemize}
    \item Libraries for working with graphs in R or Python
    \end{itemize}

  \item GraphVis \url{http://www.graphviz.org/}
    \begin{itemize}
    \item Toolkit for visualization of graphs
    \end{itemize}

  \item NetworkX \url{http://networkx.lanl.gov/}
    \begin{itemize}
    \item Python toolkit for working with graphs
    \end{itemize}

  \item GDToolkit
     \url{http://www.dia.uniroma3.it/~gdt/gdt4/index.php}
     \begin{itemize}
     \item OO C++ library for handling and drawing graphs
     \end{itemize}

  \item JUNG \url{http://jung.sourceforge.net/}
    \begin{itemize}
    \item Java universal network/graph framework
    \end{itemize}

  \item IGen \url{http://informatique.umons.ac.be/networks/igen/}
    \begin{itemize}
    \item A toolkit for generating IP network topologies based on
      network design heuristics. 
    \end{itemize}

  \end{itemize}

\subsection{Motivations for network topology investigations}

Underlying the whole research area is often an only vague notion of
why we study topology. The motives for such studies are in fact quite
diverse, and the implications are important. Topology studies
motivated by network managers with operational imperatives have
profoundly different requirements to those of the scientific research
community. Broadly speaking, we can divide the motivations as follows:
\begin{itemize}

\item {\bf Scientific:} Despite the fact that computer networks are
  designed (rather than grown as organic networks), little is known
  about their generic properties. Such knowledge would be useful
  (apart from satisfying simple curiosity) because there are many
  future protocols (for instance multicast protocols), and
  network-engineering algorithms (for instance see \cite{fortz00})
  whose design could benefit from an understanding of typical
  networks, rather than the typically very small, and unrealistic test
  examples often employed.

\item {\bf Adversarial:} Some network operators (though not all
  \cite{Zoo}) believe that commercial rivals might gain advantage
  through obtaining proprietary information about their network
  design. Similarly, there is a general belief that such information
  might facilitate an attack on the network. For instance, knowledge
  of a competitor's customers might be used by an adversary to target
  Denial of Service (DoS) attacks. One possible motivation for
  topology discovery is for just such an adversary to gain information
  to target attacks.

\item {\bf Managerial:} It is often assumed that a network operator
  has a ``database of record'' that contains all the important
  information about their network. While this may be true in some
  cases, it is more often the case that such databases are hard to
  keep up-to-date and consistent with other data sources. This is
  actually a common problem in database
  management~\cite{dasu03:_data_mining}. Such databases must be
  maintained by humans, and as soon as they grow large, and complex
  (particularly when they are dynamic), it becomes exceedingly
  difficult to eradicate all human errors.  In addition, when the
  network undergoes change, particularly unplanned changes (failures),
  the database is unlikely to be accurate. Hence, for management of
  complex, dynamic networks, another source of data about the network
  is needed. It should be no surprise that obtaining this data from
  the network itself is the best solution for ensuring up-to-date,
  accurate records.

\item {\bf Informational:} This is a fairly general category of
  motivations, but differs from pure scientific curiosity in the
  immediacy of its application. For instance, customers of network
  operators often desire information about the networks to which they
  currently subscribe (in order to debug services, or obtain quality
  of service measurements), and also about networks to which they
  might subscribe (to help make choices about who could provide them
  with the most effective service). Often, a customer may not entirely
  trust its provider, or potential provider, and wish to verify
  information themselves. Hence, there is a need for such customers to
  be able to discover the networks to which they might subscribe, or
  where they should place services.

\end{itemize}
Each of these motivations imposes different requirements on our study
in terms of accuracy, immediacy, and the types of measurements
available. High degrees of accuracy are needed for network management;
and certainly the measurements used scientifically have rarely been
very accurate (though this is a significant problem with that
research).

\subsection{Type of network}

The other core aspect we should consider is the type of topology to be
examined, as it can have a drastic affect on both observations and
behavior of the network. Two obvious dimensions are
\begin{description}
\item[Physical vs virtual:] Physical networks are built from hardware:
  routers and cables (copper or optical fiber for the Internet),
  transformers and cables for electricity, or cities and roads for the
  road network. Virtual networks may have physical nodes, but virtual
  edges (as in a social network), or virtual nodes and edges (as in
  online social networks, or the WWW).

  The main difference is that there are usually large costs in
  building, or adding to a physical network. Virtual networks, on the
  other hand, have much smaller costs (an HTML link costs almost
  nothing to create). Costs have a profound affect on network designs,
  as we shall later see, but also on dynamic behavior. If it is
  easier to change a network, it can change more quickly.

  The other major issue for a physical network is that it is often
  bound by physical constraints, and this also profoundly affect their
  design.

  \autoref{fig:phys_v_virt} illustrates the differences, and also makes the point
  that it isn't really a binary difference. Networks lie on a spectrum.

\begin{figure}[htp]
  \begin{center}
    \includegraphics[width=1.1\figurewidth]{physical_vs_virtual.pdf}
    \caption{Physical vs virtual networks.\label{fig:phys_v_virt}}
  \end{center} 
\end{figure}

\item[Transport vs Information flows:] A more subtle differentiator
  is between what the network carries. Some networks physically
  transport some type of material (cars, water, ...) whereas the flows
  in other networks are (almost) pure information (the Internet,
  ...).

  The importance of this distinction for networks may be less
  immediately obvious, but it certainly does have implications. When
  physical transport is involved in a network, the constraints on that
  network are likely to be even more stringent, and the ability to
  change the network even more limited. Costs for changing the road
  network, for instance, are usually higher than changing the
  equivalent proportion of a IP network.

\begin{figure}[htp]
  \begin{center}
     \includegraphics[width=1.1\figurewidth]{physical_vs_information.pdf}

    \caption{Transport vs information flows.\label{fig:trans_v_inf}}
  \end{center}
\end{figure}

\end{description}
Within this chapter, we are primarily interested in the ``Internet'',
and that includes both physical (OSI layer 1-3) networks, and virtual
networks (MPLS, WWW, online social networks, etc.). However, all of
the networks considered here are information transportation networks.

There are other dimensions on which networks could be classified. For
instance, by the nature of the transport. Does it come in discrete
chunks (\eg cars, packets, or the post) or continuously (\eg water or
electricity)? Is the transport connection oriented (\eg the telephone
network) or packet oriented (\eg the Internet)?

And there are other general issues we need to deal with:
\begin{itemize}

  \item Physical networks are embedded in geography, but logical
    networks often aren't, and yet the same terminology is often
    applied to each.

  \item Connectivity often changes over time, with the time-scale
    varying depending on the type of network.

  \item The Internet is often said to be a ``network of networks''. It
    is often hard to consider one network in isolation, they have
    relationships, but the situations is even more complicated than
    often imagined.
    \begin{description}

    \item[peers] Networks may be connected to {\em peers}, \ie similar
      networks that may be competing or co-operating (or both in some
      cases), \eg two ISPs operating in the same region.

    \item[parents] Networks may have a parent-child relationship in
      the sense that one network controls the other, \eg the SS7
      network with respect to traditional telephone network.

    \item[layers] A single network may have multiple layers, each of
      which can be represented by a different graph, \eg the physical-
      vs the network-layers in the Internet.

    \item[external] There is substantial interaction between
      notionally separated networks, \eg the power grid and the
      Internet, both because the Internet uses electricity, but also
      because spikes in electricity demand could potentially be caused
      by network flash crowds (certainly TV programs have a very
      important impact on electricity usage).

    \end{description}

\end{itemize}

That brings us naturally to the particular object of discussion here
-- the Internet (and its topology).  The term ``Internet'' means
(many) different things to (many) different people.  Even within the
networking community, the term is often used ambiguously, leading to
misunderstandings and confusion and creating roadblocks for a
genuinely scientific treatment of an engineered system that has
revolutionized the way we live.

While mathematics in the form of graph theory has been equally
culpable in adopting the use of this vague nomenclature, the ``new
science of networks'' has popularized it to the point where phrases
like ``topology of the Internet'' or ``Internet graph'' have entered
the mainstream science literature, even though they are essentially
meaningless without precisely-stated definitions.  For one, ``Internet
topology'' could refer to the connectivity structures encountered in
any of layers in the protocol stack, or at various levels of
aggregation. Common examples are
\begin{enumerate}

\item {\bf Router-level (layer 3):} An often sought topology is the
  router level. Somewhat ambiguously, this may also be called the
  network level, or IP level, but ``network'' is a heavily overloaded
  term here, and the IP level can also be ambiguous. For instance, IP
  level could refer to the way IP addresses are connected, that is it
  could refer to the interfaces of one router as separate nodes
  \cite{broido01:_inter}, but that is rarely what is useful for
  network operations or research. We could also add at layer 3, in
  addition to {\em interface-level} topology described above, the {\em
    subnet-level} topology
  \cite{kardes12:_cheleb,gunes09:_resol_ip_inter,Tozal:2012:ENL:2342042.2342069,Tozal:tracenet:imc2010,broido01:_inter},
  describing the interconnectivity of logical subnets (often described
  by an IP-level prefix), but here we focus on the more commonly
  considered router level.

  The router-level graph shows a range of interesting implementation
  details of a network. This type of information is critical for
  network management applications, as much of Internet management
  rests at the IP layer, and it is of great importance for network
  adversaries. For instance, developing tools to measure network
  traffic requires an understanding of the router-layer topology, in
  order to match traffic to links. Similarly traffic engineering, and
  reliability analyses are carried out at this level. One complication
  of this layer is that we sometimes wish to obtain the topology
  extending out to end-hosts, which are not technically routers, but
  we shall include these in our definition of router-layer topology,
  unless otherwise specified.

\item {\bf Switch-level (layer 2):} A single IP layer logical link may
  hide several layer-2 devices (hubs and switches). The increasing
  prevalence of Ethernet, and the ability to provide redundancy at
  reasonable cost, has led to a proliferation of such devices, and
  most Local Area Networks (LANs) are based around such. Hence, very
  many networks which have trivial, or simple IP layer topologies have
  complex and interesting layer-2 topologies. Multi-Protocol Label
  Switching (MPLS) further complicates the situation by creating
  logical layer-2 networks without physical devices, often in the form
  of cliques. Measurements often see only one layer, creating
  misunderstandings of a network's true resilience and more general
  graph properties. For instance, layer-2 devices can connect large
  numbers of routers, making them appear to have higher degree at
  layer-3 \cite{Merindol:layer2:imc2010} (for more detailed discussion
  see \autoref{sec:mpls}).

\item {\bf Physical-level (layer 1):} Below the link layer (layer 2),
  lies the physical layer. Again, many physical devices may underly a
  single logical link. Discovery of this layer is of critical
  importance in network management tasks associated with
  reliability. In particular, the concept of Shared Risk Link Groups
  (SRLG) requires knowledge of which links are carried on which fibers
  (using Wavelength Division Multiplexing), in which conduits. If a
  backhoe digs up a single conduit, it will cause a bundle of fibers
  to fail, and so connections that are in the same SRLG will all fail
  simultaneously. Clearly redundant links need to be in different
  SRLG, and discovery of the physical topology is important to ensure
  that this is the case.

\item {\bf PoP-level:} A Point-of-Presence (PoP) is a loosely defined
  grouping of devices, often defined by a metropolitan area.  PoP
  level topologies are quite useful, essentially because these graphs
  describe the logical structure of the network as the designer
  intended, rather than its particular implementation in terms of
  individual routers. Such topologies are ideal for understanding
  tradeoffs between connectivity and redundancy, and also provide the
  most essential information to competitors or customers (about where
  a network is based, or who has the best access network in a
  region). Network maps are often drawn at this level because it is an
  easy level for humans to comprehend.

\item {\bf Application layer:} There has been significant interest in
  logical topologies of the application layer, \eg for the Web
  (using HTTP, and HTML), and the P2P applications. 
% We will not
%   consider such topologies in any detail here. Some approaches may be
%   similar, but by and large the approaches to determining these
%   topologies are very application dependent, \eg web crawlers for the
%   WWW.

\item {\bf AS-level:} AS topologies have generated much interest in
  the scientific literature~\cite{barabasi99,barabasi00,barabasi02},
  because they appear to show interesting properties (such as
  power-laws) in common with other un-engineered networks such as
  biological networks. Also, much data on AS topologies is publicly
  available. While of interest in the scientific literature, this
  data's use is confused by many myths and
  misunderstandings~\cite{roughan11}.  The data may provide mild
  competitive benefits, in allowing operators to determine who peers
  with who, but the measured data often comes without attributes that
  would make the data truly useful in this regard. Finally, it is hard
  to see how such data could be used in an attack, although much
  publicized reports such as \cite{barabasi02} suggest, incorrectly
  (see \cite{Li04}), that the observed structure of the AS graph may
  lead to an ``Achilles heel'' of the Internet.

  % \item {\bf Others:} There are other topologies that may be of
  % interest from time to time, for instance, that of a multicast
  % tree.

\end{enumerate}
The number of possible topologies we might wish to discover highlights
the complexity of this problem, and why discovery is so valuable for
network management. In this chapter we will consider the router-level
topology in detail, and then discuss some of the similarities and
differences with respect to AS- and PoP-level topologies.

In addition to understanding the Internet network as a simple graph,
there are many other features of the graph that one would also wish to
know, for instance, its routing, link capacities, and geographic
locations. We describe such qualities as graph attributes, and find
that most can either be attributed to edges of the graph, for instance
\begin{sitemize}

  \item link capacities,

  \item link length,

  \item routing weights (\eg for shortest-path routing),

  \item link utilizations,

  \item link performance (for example, bit-error-rate, delay, loss,
  jitter, reordering, buffer utilization),

  \item link status (up/down), and

  \item a link's lower layer properties (\eg number of physical hops),

\end{sitemize}
or to the nodes of the graph
\begin{sitemize}

  \item geographic location,

  \item type of node, \eg brand of router, or version of software,

  \item performance measures (\eg CPU utilization), and

  \item node status (up/down).

\end{sitemize}
We could further divide this list into \define{intrinsic} network
properties, such as node location, or link capacity (things that
cannot change easily), and \define{extrinsic} properties, such as
performance, or traffic related properties, which can change
dramatically despite there being no change in the underlying network.

% In a few cases, it is hard to attribute a quality to an individual
% component of the network. For instance, routing in a network is often
% of interest. In the case of shortest path routing, this can be
% completely characterized by simple link weight attributes for each
% link, but many networks use routing protocols to create more complex
% routings. This type of complexity can sometimes be simply described by
% simple attributes, but may in other cases may be rather more
% difficult. For example OSPF with areas is often used to create a type
% of hierarchy within an AS --- in this example we can still use link
% attributes because we can assign an OSPF area to each link, but the
% attributes may be more difficult to measure. In other cases, for
% example if BGP was used as an IGP to create policy based routing
% internally in a network, we cannot describe the routing with simple
% attributes (to some extent the routing could still be captured by the
% forwarding table, which is an extreme example of a router attribute,
% but we shall exclude such extreme cases of attributes for the
% moment.). In these cases, the attributes really belong to end-to-end
% paths through the network, keyed on IP addresses (source and
% destination). We shall not consider such attributes in this report,
% though they may be of interest.

%% As an example of how one might use attributes: given node locations, it
%% is trivial to build a PoP level topology from a router-level topology.

### 2013-07-22 Compiled/chapter7/router.tex
all files compile correctly

\clearpage
\section{Router-level topology}
\label{sec:router}

\subsection{A look back}
\label{sec:router_hist}

Since the early days of the ARPANET, networking researchers have been
interested in drawing the type of networks they
designed~\cite{heart78:_arpanet}.  An early map of the ARPANET is
reproduced in \autoref{fig:arpa_1971} and shows the network's logical
connectivity structures in April 1971, when the network as a whole
consisted of some 15 nodes and a node was little more than one or two
state-of-the-art computers connected via an Interface Message
Processor or IMP (the modern equivalent would be a router).  In this
context, logical structure refers to a detailed drawing of the
connectivity among those nodes and of node-specific details (\eg type
of machine), by and large ignoring geography.  In contrast, geographic
structure refers to a map of the US that includes the actual locations
of the network's physical nodes and shows the physical connectivity
among those nodes. Such accurate maps could be drawn because at that
time, each piece of equipment was expensive and needed to be accounted
for, only a few groups with a small number of researchers were
involved in the design and installation of the network, and the
network changed relatively slowly.

\begin{figure}[thbp]
  \begin{center}
    \begin{subfigure}[b]{0.75\textwidth} \centering
      \includegraphics[width=1\textwidth]{cerf_arpanet_1971_april.png}
    \end{subfigure}
    \caption{The ARPANET in 1971 (reprinted from
      \cite{cerf90:_selected_maps}; \copyright 1990 ACM, Inc. Included here by permission.)
      \label{fig:arpa_1971}}
% \url{http://www.cs.utexas.edu/users/chris/DIGITAL_ARCHIVE/ARPANET/DARPA4799.pdf}
% (page III-144, page III-81)
  \end{center}
\end{figure}

\begin{figure}[thbp]
  \begin{center}
    \includegraphics[width=0.9\textwidth]{cerf_arpanet_1984.png}
    \caption{The early ARPANET  (reprinted from
      \cite{cerf90:_selected_maps}; \copyright 1990 ACM, Inc. Included here by permission.)
      \label{fig:arpa_1984}}
  \end{center}
\end{figure}

The network quickly grew in size and complexity. For instance,
\autoref{fig:arpa_1984} shows the geographic counterpart from 1984 of
the ARPANET map depicted in \autoref{fig:arpa_1971}.  Manually
accounting for the increasing number of components quickly became
prohibitive and motivated the adoption of automatic strategies for
obtaining some of the available connectivity as well as traffic
information.  A prime example for effectively visualizing this
collected information is reproduced from
\cite{frazer:_merit_histor_nsfnet_backb_projec} and shown in
\autoref{fig:nsfnet}, which depicts a 3D rendering of the (US portion
of the) NSFNET around 1991, annotated with traffic-related
information. At that time, the NSFNET backbone consisted of some 14
nodes that were interconnected with T1 links as shown and, in turn,
connected to a number of different campus networks (\eg collections of
interconnected LANs). However, even though the internal structure of
the backbone nodes was well-known (\ie each node was composed of nine
IBM RTs linked by two token rings with an Ethernet interface to
attached networks), nobody had any longer access to the internals of
all the different campus networks and as a result, drawing the 1991
NSFNET equivalent of the ARPANET's logical connectivity structure
(\autoref{fig:arpa_1971}) was no longer possible.

\begin{figure}[thbp]
  \begin{center}
    \includegraphics[width=0.7\textwidth]{1024px-NSFNET-traffic-visualization-1991.jpg}
    \caption{A visualization of the NSFNET circa 1991 (by Donna Cox
and Robert Patterson,
\href{http://avl.ncsa.illinois.edu/project-archive/visualizing-the-early-internet}{National
Center for Supercomputing Applications, University of Illinois at
Urbana-Champaign}. See also
\url{http://en.wikipedia.org/wiki/File:NSFNET-traffic-visualization-1991.jpg}).
      \label{fig:nsfnet}}
  \end{center}
\end{figure}

With the decommissioning of the NSFNET in 1995 and the rise of the
"public Internet", the researchers' ability to obtain detailed
connectivity and component information about the internals of the
different networks that formed the emerging "network of networks"
further diminished and generated renewed interest in the development of
abstract, yet informed, models for router-topology evaluation and
generation. For example, the Waxman model \cite{Waxman88}, a variation
of the classical Erd\"{o}s-R\'{e}nyi random graph model \cite{erdos60}
was the first popular topology generator commonly-used for network
simulation studies at the router level. However, it was largely
abandoned in the late 1990s in favor of models that attempted to
explicitly account for non-random structure as part of the network
design. The arguments that favored structure over randomness were
largely empirical in nature and reflected the fact that the inspection
of real-world router-level ISP networks showed clear signs of
non-random structures in the form of the presence of backbones, the
appearance of hierarchical designs, and the importance of locality.
These arguments also favored the notion that a topology generator
should reflect the design principles in common use; \eg to achieve
some desired performance objectives, the physical networks must
satisfy certain connectivity and redundancy requirements, properties
which are generally not guaranteed in random network topologies.
These principles were, for example, advanced in
\cite{Zegura96,Calvert97,Zegura97} and were ultimately integrated into
the popular Georgia Tech Internetwork Topology Models (GT-ITM)
\cite{gt_itm}.

These more structure-oriented router topology generators were viewed
as the state-of-the-art until around 2000 when, in turn, they were
largely abandoned in favor of a new class of random graph models whose
trademark was the ability to reproduce the newly discovered power-law
relationship in the observed connectivity (\ie node degree) of
router-level graphs of the Internet. This discovery was originally
reported in the seminal paper by
Faloutsos~\etal~\cite{faloutsos99:_power_law_relat_of_inter_topol},
who used a router-level graph constructed from data that was collected
a few years earlier by Pansiot and Grad~\cite{pansiot98:_inter} for
the purpose of obtaining some experimental data on the actual shape of
multicast trees in the Internet. The Boston university Representative
Internet Topology gEnerator (BRITE) \cite{Medina01} became a popular
representative of this new class of models, in part also because it
combined the more structure-oriented perspective of the GT-ITM
generator with the new focus that emphasized the ability to reproduce
certain metrics or statistics of measured router topologies (\eg node
degree distribution).

One of the hallmarks of networks that have power-law degree
distributions and that are generated according to any of a number of
different probabilistic mechanisms (\eg preferential attachment
\cite{barabasi99}, random graphs with a given expected degree sequence
\cite{Chung04}, power-law random graphs 
\cite{Aiello00}) is that they can be shown to have
a few centrally located and highly connected {\em hubs} through which
essentially most traffic must flow.  When using these models to
represent the router-level topology of the Internet, the presence of
these highly connected central nodes has been touted the Internet's
``Achilles-heel'' because network connectivity is highly vulnerable to
attacks that target the high-degree hub nodes \cite{barabasi00}. It has been
similarly argued that these high-degree hubs are a primary reason for
the epidemic spread of computer worms and viruses \cite{berger05:_spread_of_viruses,pastor-satorras01:_epidem}.
Importantly, the presence of highly connected central nodes in a
network having a power-law degree distribution is the essence of the
so-called scale-free network models. They have been a highly popular
theme in the study of complex networks, particularly among researchers
inspired by statistical physics \cite{albert_barabasi02}, and have fuelled the rise of
a new scientific discipline that has become known as ``Network
Science" \cite{barabasi12}.  In the process, they have also seen wide-spread
use among Internet topology researchers.

However, as the general fascination with and popularity of network
science in general and scale free network modeling in particular grew,
so did the arguments that were voiced by Internet researchers
and questioned the appropriateness and relevance of the scale-free
modeling approach for studying highly-engineered systems such as the
Internet's router topology. In fact, at around 2010, when the number
of publications in the area of network science reached a new height,
the number of papers that were published in the networking research
literature and applied scale-free network models to describe or study
router-level topologies of the Internet was close to zero.  This begs
the question ``What happened?'', and the answer provided in the next
section is really a classic lesson in how errors of various forms
occur and can add up to produce results and claims that create
excitement among non-networking researchers, but quickly collapse when
scrutinized with real data or examined by domain experts.

\subsection{Know your measurements}
\label{sec:router_measurements}

Between 1990 and 2000, Internet topology research underwent a drastic change from being
a data-starved discipline to becoming a prime example of a largely measurement-driven research 
activity.  As described earlier, even though the development of abstract, yet informed,
models for network topology evaluation and generation has always been a give and take between
theoreticians and empiricists, for router topology modeling, the essential role that
measurements have started to play came into full focus in a sequence of three seminal papers that
appeared between 1998-2000.

\subsubsection{Three seminal papers on router topology modeling}
\label{sec:router_seminal}

The key papers that turned router topology modeling into a full-fledged measurement-driven
research activity cover the whole spectrum of modeling activities, from
measurement experiments to model construction and validation to graph-theoretical network
analysis, and are listed below:

\begin{itemize}
\item[(i)] ``On routes and multicast trees in the Internet'' by
  J.-J. Pansiot and D. Grad (1998) \cite{pansiot98:_inter} described
  the original measurement experiment that was performed in mid-1995
  and produced data on actual routes taken by packets in the
  Internet. This data was subsequently used to construct a router
  graph of the Internet.

\item[(ii)] ``On power-law relationships of the Internet topology'' by
  M. Faloutsos~\etal~(1999)
  \cite{faloutsos99:_power_law_relat_of_inter_topol} reported (among
  other observations) on the observed power-law relationship in the
  connectivity of the router-level topology of the Internet measured
  by Pansiot and Grad~\cite{pansiot98:_inter}.

\item[(iii)] ``Error and attack tolerance of complex networks'' by
  R. Albert~\etal~(2000) \cite{barabasi00} proposed a scale-free
  network model to describe the router topology of the Internet and
  argued for its validity on the basis of the latest findings by
  Faloutsos~\etal~\cite{faloutsos99:_power_law_relat_of_inter_topol}.
  It touted the new model's exemplary predictive power by reporting on
  the discovery of a fundamental weakness of the Internet (a property
  that was became known as the Internet's "Achilles' heel") that went
  apparently unnoticed by the engineers and researchers who have
  designed, deployed, and studied this large-scale, critical
  infrastructure, but followed directly from the newly proposed
  scale-free modeling approach.
\end{itemize}

\begin{figure}[!htp] 
  \begin{center}
    \includegraphics*[width=0.8\columnwidth,viewport=0mm 52mm 260mm 160mm]{ams-figs-final.pdf}
    \caption{A toy example of a scale-free network of the preferential
      attachment type (b) generated to match a power-law type node
      degree distribution (a).  (First published in Notices of the
      American Mathematical Society, Volume 56, No.3 (May 2009):
      586-599 \cite{willinger09:_mathem_and_inter}. Included here by
      permission.)
      \label{fig:hot_1}}
    % \url{http://www.ams.org/notices/200905/rtx090500586p.pdf}  (figure 1)
  \end{center}
\end{figure}

At first glance, the combination of these three papers appears to show
network modeling at its best -- firmly based on experimental data,
following modeling practices steeped in tradition, and discovering
surprisingly and previously unknown properties of the modeled network.
An example of a toy network resulting from taking the findings from
these seminal papers at face value is shown in \autoref{fig:hot_1}.
However, one of the beauties of studying man-made systems such as the
Internet is that -- because of their highly-engineered architectures,
a thorough understanding of its component technologies, and the
availability of extensive (but not necessarily very accurate)
measurement capabilities -- they provide a unique setting in which
most claims about their properties, structure, and functionality can
be unambiguously resolved, though perhaps not without substantial
efforts. In the remainder of this section, we will illustrate how in
the context of the Internet's router topology, applying readily
available domain knowledge in the form of original design principles,
existing technological constraints, and available measurement
methodologies reveals a drastically different picture from that
painted in these three seminal papers. In fact, we will expose the
specious nature of scale-free network models that may appeal to more
mathematically inclined researchers because of their simplicity or
generality, but besides having no bearing on the Internet's router
topology are also resulting in wrong claims about the Internet as a
whole.

\subsubsection{A first sanity check: Using publicly available information}

A first indication of apparent inconsistencies between the proposed
scale-free models for the Internet's router topology and the actual
Internet comes from the inspection of the router topologies of actual
networks that make the details of their network internals publicly
available. For example, networks such as Internet2~\cite{internet2} or
G\'{E}ANT~\cite{geant} show no evidence that there exist any centrally
located and highly connected ``hubs'' through which essentially most
traffic must flow.  Instead, what they typically show is the presence
of a more or less pronounced ``backbone'' network that is fed by
tree-like access networks, with additional connections at various
places to provide a degree of redundancy and robustness to components
failures\footnote{This is not a universal phenomena. For instance
  \cite{Zoo} notes that some networks do exhibit hub-like structure, but
  it is the lack of universality that is important here, as exhibited
  by these and other counter examples.}.

This design pattern is fully consistent with even just a cursory
reading of the most recent product catalogs or white papers published
by the main router vendors
\cite{cisco13:_build_cost_effec_scalab_core,cisco13:_converged,juniper13:_evolv_backb_networ_using_mpls_super}.
For one, the most expensive and fastest or highest-capacity pieces of
equipment are explicitly marketed as backbone routers. Moreover, due
to inherent technological limitations in how many packets or bytes a
router can handle in a given time interval, even the latest models of
backbone routers can support only a small number of very
high-bandwidth connections, typically to connect to other backbone
routers. At the same time, a wide range of cheaper, slower or
lower-capacity products are offered by the different router vendors
and are targeted primarily at to support network access.  On the
access side, a typical router will have many lower-bandwidth
connections for the purpose of aggregating customer traffic from the
network's edge and subsequently forwarding that traffic towards the
backbone. In short, even the latest models advertised by today's
router vendors are limited by existing technologies, and even for the
top-of-the-line backbone routers, it is technologically infeasible to
have hundreds or thousands of high-bandwidth connections. At the same
time, while technically feasible, deploying some of the most expensive
equipment and configuring it to support hundreds or thousands of
low-bandwidth connections would be considered an overall bad
engineering decision (\eg excessively costly, highly inefficient, and
causing serious bottlenecks in the network).

However, the root cause for these outward signs of a clear mismatch
between the modeled and actual
router topology of the Internet goes deeper and lies in the original design philosophy of the
Internet.  As detailed in \cite{clarke95:_design}, while the top level goal for the original DARPA Internet architecture was {\it ``to develop an effective technique for multiplexed utilization
of existing interconnected networks"}, the requirement that {\it ``Internet communication must
continue despite loss of networks or gateways"} topped the list of second level goals. To survive
in the face of components failing off, the architecture was to mask completely any transient 
failure, and to achieve this goal, state information which describes an existing connection must
be protected. To this end, the architecture adopted the ``fate-sharing" model that gathers this 
state information at the endpoints of connections, at the entities that are utilizing the 
service of the network. Under this model, it is acceptable to lose the state information 
associated with an entity if, at the same time, the entity itself is lost; that is, there 
exists no longer any physical path over which any sort of communication with that entity can
be achieved (\ie total partition). Ironically, these original design principles
outlined in \cite{clarke95:_design} favor precisely the opposite of what the scale-free modeling approach
yields -- no centrally located and highly connected ``hubs'' because their removal makes
partitioning the network easy.

\subsubsection{An in-depth look at traceroute: \\ Examining a popular measurement technique}
\label{sec:mpls}

While the above-mentioned empirical, technological, and architectural
arguments cast some serious doubts on the scale-free network modeling
approach for the router topology of the Internet, they say nothing
about the measurements that form the basis of this approach and has
given it a sense of legitimacy among scientists in general and
networking researchers in particular.  To appreciate the full role
that measurements play in this discussion, it is informative to
revisit the original paper by Pansiot and Grad~\cite{pansiot98:_inter}
that describes the measurement experiment, discusses the measurement
technique used, and provides a detailed account of the quality of the
data that form the basis of the scale-free approach towards modeling
the Internet's router topology.

In essence, \cite{pansiot98:_inter} describes the first (at that time)
large-scale traceroute campaign performed for the main purpose of
constructing a router graph of the Internet from actual Internet
routes. Although traceroute-based, the authors of
\cite{pansiot98:_inter} quickly point out that their purpose of using
the traceroute tool (\ie obtaining actual Internet routes to
construct a router graph) differed from what V. Jacobson
\cite{jacobson89:_tracer} had in mind when he originally designed the
tool (\ie tracing a route from a source to a destination for
diagnostic purposes).  As a result, a number of serious issues arise
that highlight why using the traceroute technique for the purpose of
constructing a router graph is little more than an "engineering hack"
and can certainly not be called a well-understood "measurement
methodology."

{\bf IP alias resolution problem:} One serious problem
explained in detail in \cite{pansiot98:_inter} with using traceroute-based data for
constructing router graphs is that the traceroute tool only returns
the IP addresses of the interface cards of the routers that the probe
packets encountered on their route from the source to their
destination.  However, most routers have many interface cards, and
despite many years of research efforts that have produced a series of
increasingly sophisticated heuristics \cite{bender08:_fixin_allys,gunes09:_resol_ip_inter,sherry10:_resol_ip_alias_presp_times}, the networking
community still lacks rigorous and accurate methods for resolving the
IP alias resolution problem; that is, determining whether two
different interface IP addresses belong to or can be mapped to the
same router. While the essence of this problem is illustrated in
\autoref{fig:hot_2}, the impact it can have when trying to map a
router topology of an actual network is shown in \autoref{fig:hot_3}.

\begin{figure}[!htp] 
  \begin{center}
% pdfimages gets actual 'images' (bitmaps) out of the PDF, but not plots
% pdfjam ams-figs-final.pdf '5' --outfile ams-figs-p5.pdf
%     extracts a particular page so I can use it here
    \includegraphics*[width=0.6\columnwidth,viewport=10mm 110mm 200mm 200mm]{ams-figs-p2.pdf}
%                                                    xleft ybottom xright ytop
%                                                 a4 = 0 0 210 297
    \caption{The IP alias resolution problem. Paraphrasing Fig. 4 of \cite{rocketfuel_1}, 
	 traceroute does not list routers (boxes) along paths but IP addresses of 
	 input interfaces (circles), and alias resolution refers to the correct 
	 mapping of interfaces to routers to reveal the actual topology. In the case
	 where interfaces 1 and 2 are aliases, (b) depicts the actual topology while 
	 (a) yields an ``inflated'' topology with more routers and links.
	 (First published in Notices of the American Mathematical Society, Volume 56, 
	 No.3 (May 2009): 586-599 \cite{willinger09:_mathem_and_inter}. Included here by permission.)
      % \cite[Figure~2]{willinger09:_mathem_and_inter}.
      \label{fig:hot_2}}
    % \url{http://www.ams.org/notices/200905/rtx090500586p.pdf}  (figure 2)
  \end{center}
\end{figure}

\begin{figure}[thbp] 
  \begin{center}
% pdfimages gets actual 'images' (bitmaps) out of the PDF, but not plots
% pdfjam ams-figs-final.pdf '5' --outfile ams-figs-p5.pdf
%     extracts a particular page so I can use it here
    \includegraphics*[width=1\columnwidth,viewport=0mm 70mm 210mm 240mm]{ams-figs-p3.pdf}
%                                                    xleft ybottom xright ytop
%                                                 a4 = 0 0 210 297
    \caption{The IP alias resolution problem in practice. Shown is a comparison 
	 between the Abilene/Internet2 topology inferred by Rocketfuel (left) and the 
	 actual topology (top right). Rectangles represent routers with interior ovals 
	 denoting interfaces. The histograms of the corresponding node degrees are 
	 shown in the bottom right plot. 
	 (Reprinted from \cite{sherwood08:_discar}; \copyright 2008
         ACM. Inc. Included here by permission.)
         \label{fig:hot_3}}
    % \url{http://www.ams.org/notices/200905/rtx090500586p.pdf}  (figure 3)
  \end{center}
\end{figure}

\clearpage
{\bf Lesson 1:} {\em Due to the absence of accurate and rigorous methods for solving the IP alias
resolution problem, the actual values of the connectivity of each router (\ie node degrees)
inferred from traceroute measurements cannot be taken at face
value.}

{\bf Opaque Layer-2 clouds:} Another serious issue with
using generic traceroute-based measurements for construction router
graphs is also discussed at length in \cite{pansiot98:_inter} and
illustrated in \autoref{fig:hot_4}.  Being strictly limited to IP or
layer-3, the problem with traceroute is that it is incapable of
tracing through opaque layer-2 clouds that feature circuit
technologies such as Asynchronous Transfer Mode (ATM) or Multiprotocol
Label Switching (MPLS). These technologies have the explicit and
intended purpose of hiding the network's physical infrastructure from
IP, so from the perspective of traceroute, a network that runs these
technologies will appear to provide direct connectivity between
routers that are separated by local, regional, national, or even
global physical network infrastructures. An example of using
traceroute to map a network that uses MPLS is depicted in
\autoref{fig:hot_4} and shows an essentially completely connected
graph at Layer 3 with multiple high-degree nodes, even though the
physical router topology is very sparse. Similarly, if traceroute
encounters an ATM cloud, it falsely ``discovers'' a high-degree node
that is really a logical entity -- often an entire network potentially
spanning many hosts or great distances -- rather than a physical node
of the Internet's router-level
topology. Donnet~\etal~\cite{benoit12:_reveal_mpls} found that at
least 30\% of the paths they tested traversed an MPLS tunnel.

Recent extensions of the ICMP protocol, using traceroute to trace
through opaque MPLS clouds have become technically
feasible~\cite{rfc4950}, but operators often configure their routers to
hide the MPLS tunnels by turning off this option
\cite{sommers11:_preval_charac_mpls_deploy_open_inter}. Even then it
may be possible to detect the MPLS tunnels
\cite{benoit12:_reveal_mpls}, but the inference techniques are not a
guarantee, and are quite particular to MPLS, which is not the only
technique for creating tunnels, so there may still be some opaque
networks to deal with.  More to the point, even where such inferences
are possible, most data sets do not contain this type of analysis, and
most subsequent analyses of the data have ignored the issue.

\begin{figure}[thbp] 
  \begin{center}
% pdfimages gets actual 'images' (bitmaps) out of the PDF, but not plots
% pdfjam ams-figs-final.pdf '5' --outfile ams-figs-p5.pdf
%     extracts a particular page so I can use it here
    \includegraphics*[width=1\columnwidth,viewport=17mm 70mm 193mm 240mm]{ams-figs-p4.pdf}
%                                                    xleft ybottom xright ytop
%                                                 a4 = 0 0 210 297
    \caption{How traceroute detects fictitious high-degree nodes in
      the network core. (a) The actual connectivity of an opaque
      layer-2 cloud, i.e., a router-level network running a technology
      such as ATM or MPLS (left) and the connectivity inferred by
      traceroute probes entering the network at the marked router
      (right).  (b) The Rocketfuel-inferred backbone topology of
      AS3356 (Level3), a Tier-1 Internet service provider and leader
      in the deployment of MPLS.  (Figure (b) reprinted from \cite{rocketfuel_1};
      \copyright 2002 ACM, Inc. Included here by permission.)
       % \cite[Figure 4]{willinger09:_mathem_and_inter}.
      \label{fig:hot_4}}
    % \url{http://www.ams.org/notices/200905/rtx090500586p.pdf}  (figure 4)
  \end{center}
\end{figure}

{\bf Lesson 2:} {\em Due to an inability of the generic traceroute
technique to trace through opaque Layer-2 clouds, or understand the
connectivity created by Layer-2
devices~\cite{Merindol:layer2:imc2010}, the inferred high-degree nodes
(\ie routers with a large number of connections) are typically
fictitious, an artifact of an imperfect measurement tool.}

{\bf Limited vantage points:} We have commented earlier
that since a router is fundamentally limited in terms of the number of
packets it can process in any time interval, there is an inherent
tradeoff in router configuration: it can support either a few
high-throughput connections or many low-throughput connections. Thus,
for any given router technology, a high-connectivity router in the
core reflects a poor design decision -- it will either have poor
performance due to its slow connections or be prohibitively expensive
relative to other options. Conversely, a good design choice is to
deploy cheap high-degree router near the edge of the network and rely
on the very technology that supports easy multiplexing of a large
number of relatively low-bandwidth links. Unfortunately, neither the
original traceroute-based study of Pansiot and
Grad~\cite{pansiot98:_inter} nor any of the larger-scale campaigns
that were subsequently performed by various network research groups
have the ability to detect those actual high-degree nodes. The simple
reason is that these campaigns lack access to a sufficient number of
vantage points (\ie sources for launching traceroute probes and
targets) in any local end-system to reveal these actual high
connectivity patterns at the network's edge.

{\bf Lesson 3:} {\em If there were high-degree nodes in the network,
existing router technology relegates them to the edge of the network
where no generic traceroute-based measurement campaigns is able to
detect them because of a lack of vantage points nearby.}

There are other issues with large-scale traceroute campaigns that
impact the quality of the resulting measurements and have received
some attention in the literature.  For example, the use of traceroute
has been shown to make experimental data susceptible to a type of
measurement bias in which some nodes of the network are oversampled,
while others are undersampled. However, while this feature has
received considerable attention
\cite{LakhinaByersCrovellaXie03,achlioptas05:_bias_of_tracer_sampl},
in the presence of systematic errors due to an inability to perform
accurate IP alias resolution or trace through opaque Layer-2 clouds,
this work is largely of theoretical interest and of little practical
relevance for modeling the Internet's router topology.

% \noindent {\bf Efficiency:} reducing the number of packets --

% "Efficient IP-level Network Topology Capture" [Paper]
% Bourgeau Thomas (UPMC Sorbonne Universities)
% Timur Friedman (UPMC Sorbonne Universities)
% PAM 2013

\subsubsection{Just the facts: power-law scaling and router-level topologies}

When applying lessons 1-3 to the main findings reported in the seminal
papers discussed in \autoref{sec:router_seminal} we are faced with the
following facts:
\begin{description}
\item[{\bf Fact 1:}] A very typical but largely ignored fact about Internet-related 
measurements in general and traceroute measurements in particular is that what we can 
measure in an Internet-like environment is generally not the same as what we really 
want to measure (or what we think we actually measure). This is mainly because as a 
decentralized and distributed system, the Internet lacks a central authority and does 
not support third-party measurements.

\item[{\bf Fact 2:}] A particularly ironic fact about traceroute is that the high-degree nodes 
it detects in the network core are necessarily fictitious and represent entire opaque 
layer-2 clouds, and if there are actual high-degree nodes in the network, existing technology
relegates them to the edge of the network where no generic traceroute-based measurement
experiment will ever detect them.

\item[{\bf Fact 3:}] In particular, due to the inherent inability of traceroute to (i) reveal 
unambiguously the actual connectivity (\ie node degree) of any router, and (ii) correctly 
identify even the mere absence or presence of high-degree nodes (let alone their actual values),
statistical statements such as those made in \cite{faloutsos99:_power_law_relat_of_inter_topol} claiming that the Internet's router 
connectivity is well described by  a power-law distribution (or, for that case, any other type of distribution) cannot be justified with any reasonable degree of statistical confidence.

\item[{\bf Fact 4:}] Since historical traceroute-based measurements cannot be taken at 
face value when (mis)using them for inferring router topologies and the inference results 
obtained from such data cannot be trusted, the claims that have been made about the 
(router-level) Internet in \cite{barabasi00} are without substance and collapse under careful scrutiny.
\end{description}

In short, after almost 15 years of examining the idiosyncrasies of the traceroute tool,
there exists overwhelming evidence that the sort of generic and raw traceroute measurements
that have been used to date to infer the Internet's router topology are seriously flawed to 
the point of being essentially of no use for performing scientifically sound inferences. 
Yet, the myth that started with 
\cite{faloutsos99:_power_law_relat_of_inter_topol}; \ie the router topology of the Internet exhibits 
power-law degree distributions persists and continues to be especially popular with researchers 
that typically work in the field of network science and show in general little interest in 
domain-specific "details" such as traceroute's idiosyncrasies.

At the same time, it is worthwhile pointing out that most of the
above-mentioned flaws and shortcomings of traceroute-based
measurements are neither new nor controversial with networking
researchers.  In fact, when discussing the use of the traceroute tool
as part of their original measurement experiment, the authors of  \cite{pansiot98:_inter} described many of the issues
discussed in this section in great detail and commented on the
possible implications that these inherently traceroute-related issues 
can have for constructing router graphs of the Internet. In this
sense, \cite{pansiot98:_inter} is an early example of an exemplary measurement paper,
but unfortunately, it has been largely ignored and essentially
forgotten.  For one,
\cite{faloutsos99:_power_law_relat_of_inter_topol}, which critically relies on the data
described in \cite{pansiot98:_inter} for their power law claim for the Internet's
router topology, fails to recognize the relevance of these issues and
does not even comment on them. Moreover, the majority of papers that
have appeared in this area after the publication of \cite{faloutsos99:_power_law_relat_of_inter_topol} typically
cite only \cite{faloutsos99:_power_law_relat_of_inter_topol} and don't even mention \cite{pansiot98:_inter}.

Traceroute-based measurements are not the only approach for obtaining
router-level topologies, just the most commonly presented in the
research literature. Network operators can obtain measurements of
their own networks using much more accurate methods: for instance,
from configuration files~\cite{feldmann00:_netscope}, or using route
monitors~\cite{shaikh01:_ospf}, but those techniques require
privileged access to the network, and so haven't been used widely for
research. More recently, the mrinfo tool~\cite{mrinfo} has been used
to measure topologies using IGMP (the Internet Group Management
Protocol)~\cite{merindol09:_quant_ases,pansiot10:_extrac}. IGMP has
the advantage that routers that respond provide much more complete
information on their interfaces than those responding to traceroutes
(so aliasing is less an issue), but there are still coverage problems
created by lack of support, or deliberate filtering or rate limiting
of responses to the protocol~\cite{marchetta12:_quant_igmp}.

\subsection{Network modeling: An exercise in reverse-engineering}
\label{sec:router_models}

The conclusion from the previous section that the available traceroute measurements 
are of insufficient quality to infer any statistical quantity of the data (including
node degree distribution) with sufficient statistical confidence is a show-stopper 
for traditional network modeling.  Indeed, given that the data cannot be trusted, 
relying on statistics of unknown accuracy (\eg by and large arbitrary node degrees) makes 
model selection precarious, and model validation in the sense of checking if
the selected model describes the data ``well" is an oxymoron -- providing a ``good'' fit
for statistical quantities of unknown accuracy is meaningless.

As such, the scale-free approach to modeling 
the Internet's router topology advanced in \cite{barabasi00} is an example of what can go wrong
if serious flaws of the underlying data are ignored and the available measurements 
are taken at face value.  It should therefore come as no surprise that the resulting
modeling framework and ensuing claims quickly collapse when scrutinized with readily
available domain knowledge or vetted against alternative and solid sources of information.
However, the lessons learned from this ill-fated approach to router topology modeling
rises the question: What are viable alternative approaches to modeling the Internet's
router topology that are by and large independent of the available but problematic
traceroute measurements?

\subsubsection{Router topology modeling as a constrained optimization problem}

Having dismissed traceroute-based data as a source for informing our approach to 
modeling the Internet's router topology, we turn to readily available domain 
knowledge as critical alternate information source.  To this end, we focus on 
the specific problem of modeling the physical infrastructure of a regional,
national, or global Internet Service Provider (ISP).

The first key ingredient of this ``first-principles'' approach is the
realization that ISPs design their physical infrastructures for a
purpose; that is, their decisions are driven by possibly ISP-specific
objectives and reflect trade-offs between what is feasible and what is
desirable. While in general it may be difficult if not impossible to
define or capture the precise meaning of a particular ISP's purpose
for designing its network, an objective that expresses a desire to
provide connectivity to the rest of the Internet for its end users and
an ability to carry an expected traffic demand efficiently and
effectively, subject to prevailing economic and technological
constraints, is unlikely to be far from the ``true'' purpose.

The second ingredient concerns the trade-offs an ISP has to made between what is
feasible (in terms of available products sold by the different router vendors) and
what is desirable (in terms of cost, performance, ease-of-management or other 
criteria for the built-out router topology). In particular, router technology 
constraints are a significant force shaping network connectivity at the router-level
and, in turn, router topology design. Due to hard physical limits, even the most 
expensive and highest-capacity router models available on the market in any given
year operate within a ``feasible region" and corresponding "efficiency frontier"
of possible bandwidth-degree combinations; that is, they can be configured to either
have only a few high-bandwidth connections and perform at their capacity or have
many low-bandwidth connections and tolerate a performance hit due to the overhead that
results from the increased connectivity.

Similarly, economic considerations also
affect network connectivity and router topology design. For example, the cost of 
installing and operating physical links in a network can often dominate the cost of 
the overall router infrastructure. In essence, this observation creates enormous 
practical incentives to design the physical plant of an ISP so as to keep the number
of links small and avoid whenever possible long-haul connections due to their high
cost.  These incentives to share costs via multiplexing impact and are impacted by
available router technologies and argue for a design principle for an ISP's router
topology that favors aggregating traffic at all levels of network hierarchy, from
its periphery all the way to its core.

The third and final key ingredient of the proposed first-principle alternative to 
router topology modeling is concerned with the role that randomness plays in this
approach. Recall that the traditional approach is typically graph theory-based where
randomness is explicit and appears in the form of a series of coin tosses (using 
potentially bias coins as in the case of scale-free networks of the preferential 
attachment type) that determine whether or not two nodes (\ie routers) are connected
by a physical link, irrespective of the type of routers involved or link considered.
In stark contrast, in our approach, randomness enters in a very different and
less explicit manner, namely in terms of the uncertainty that exists about the 
``environment'' (\ie the traffic demand that the network is expected to carry).
Moreover, irrespective of the model chosen for quantifying this uncertainty, the
resulting network design is expected to exhibit strong robustness properties with 
respect to changes in this environment.

When combining all three ingredients to formulate an ISP's router topology design 
problem, the mathematical modeling language that naturally reflects the objectives
of an ISP, its need to adhere to existing technology constraints and respect
economic considerations, and its desire to operate effectively and efficiently in
light of the uncertainly in the environment is {\it constrained optimization.}  
Thus, we have changed network modeling from an exercise in model fitting into an
exercise in reverse-engineering and seek a solution to a constrained optimization
problem formulation that captures by and large what the ISP can afford to build, 
operate, and manage (\ie economic considerations), satisfies the hard constraints 
that technology imposes on the network's physical entities (\ie routers and links),
and is robust to changes in the expected traffic that is supposed to handle

\subsubsection{Heuristically optimal router topologies}

In the process of formulating the design of an ISP's router topology as a constrained
optimization problem, we alluded to a synergy that exists between the technological
and economic design issues with respect to the network core and the network edge.
The all-important objective to multiplex traffic is supported by the types of 
routers available on the market. In turn, the use of these products re-enforces traffic 
aggregation everywhere in the network. Thus, the trade-offs that an ISP has to make
between what is technologically feasible versus economically sensible can be expected 
to yield router topologies where individual link capacities tend to increase while
the degree of connectivity tends to decrease as one moves from the network edge to 
its core.

This consistent picture with regard to the forces that by and large
govern the built-out and provisioning of an ISP's router topology and include aspects 
such as equipment constraints, link costs, and bandwidth demands suggests that the
following type of topology is a reasonably ``good'' design for a single ISP's physical
plant: (i) Construct a core as a loose mesh of expensive, high-capacity, 
low-connectivity routers which carry heavily aggregated traffic over high-bandwidth links.
(ii) Support this mesh-like core with hierarchical tree-like structures at the edge
of the network for the purpose of aggregating traffic from end users via cheaper,
lower-capacity, high-connectivity routers.  (iii) Augment the resulting structure
with additional connections at various selectively-chosen places to provide a degree
of redundancy and robustness to component failures.  The result is a topology that
has a more or less pronounced backbone, which is fed by tree-like access networks, 
with additional links added for redundancy and resilience. We refer to this design
as {\it heuristically optimal} to reflect its consistency with real design considerations
and call the resulting "solutions" {\it heuristically optimal topologies}, or HOT for short.
Note that such HOT models have been discussed earlier in the context of
highly organized/optimized tolerances/tradeoffs \cite{carlson02:_pnas,Fabrikant02}.

An important aspect of the proposed HOT models is that even though we have formulated 
the design of an ISP's router topology as a constrained optimization problem
that could in principle be solved optimally, we are typically not concerned 
with a network design that is ``optimal'' in a strictly mathematical sense and is also 
likely to be NP-hard. Instead, our interest is in solutions that are ``heuristically 
optimal'' in the sense that they result in ``good'' performance. The main reason
for not pursuing optimal solutions more aggressively is the imprecise nature of
essentially all ingredients of the constrained optimization problem of interest.
For one, it is unrealistic to expect that an ISP's true objective for building out
and provisioning its physical infrastructure can be fully expressed in mathematical
terms as an objective function. Furthermore, a bewildering number of different types of
routers and connections make it practically impossible to account for the nuances
of the relevant feasible regions or efficiency frontiers. Finally, any stochastic
model for describing the expected traffic demand is an approximation of reality 
or at best based on imprecise forecasts.  Given this approximate nature of the
underlying constrained optimization problem, we seek solutions that captures by and 
large what the ISP can afford to build, operate, and manage (\ie economic 
considerations), satisfy some of the more critical hard constraints that technology 
imposes on the network's physical entities (\ie routers and links), and exhibit
strong robustness properties with fluctuations in the expected traffic demands
(\ie insensitivity to changes in the uncertain environment).

\subsubsection{A toy example of a HOT router topology}

To illustrate the proposed HOT approach, we use a toy example that is
rich enough to highlight the key ingredients of the outlined
first-principles methodology and demonstrate its relevance for router
topology modeling as compared to the popular model-fitting approach.
It's toy nature is mainly due to a number of simplifying assumptions
we make that facilitate the problem formulation. For one, by simply
equating throughput with revenues, we select as our objective function
the maximum throughput that the network can achieve for a given
traffic demand and use it as a metric for quantifying the performance
of our solutions. Second, considering an arbitrary distribution of
end-user traffic demand $B_i$, we assume a gravity model for the
unknown traffic demand; that is, assuming shortest-path routing, the
demands are given by the traffic matrix element $x_{i,j} = \alpha B_i
B_j$ between routers $i$ and $j$ for some constant $\alpha$. Lastly,
we consider only one type of router and its associated technologically
feasible region; that is, (router degree, router capacity)-pairs that
are achievable with the considered router type, and implicitly avoid
long-haul connections due to their high cost.

The resulting constrained optimization problem can be written in the
form
\begin{equation}
   \max_{\rho} \sum x_{i,j} \mbox{~~~such~that~~~} A {\cal X} \leq C,
\end{equation}
where ${\cal X}$ is the vector obtained by stacking all the demands
$x_{i,j}$; $A$ is the routing matrix obtained by using standard
shortest path routing and defined by $A_{k,l} = 1$ or 0, depending on
whether or not demand $l$ passes through router $k$; and $C$ is the
vector consisting of the router degree-bandwidths constraints imposed
by the technologically feasible region of the router at hand.

% \begin{figure}[thbp] 
%   \begin{center}
% % pdfimages gets actual 'images' (bitmaps) out of the PDF, but not plots
% % pdfjam ams-figs-final.pdf '5' --outfile ams-figs-p5.pdf
% %     extracts a particular page so I can use it here
%     \includegraphics*[width=0.8\columnwidth,viewport=0mm 120mm 210mm 215mm]{ams-figs-p5.pdf}
% %                                                    xleft ybottom xright ytop
% %                                                 a4 = 0 0 210 297
%     \caption{Generating networks using constrained optimization. (a) Engineers 
% 	 view network structure as the solution to a design problem that measures performance 
% 	 in terms of the ability to satisfy traffic demand while adhering to node and arc 
% 	 capacity constraints. (b) A network resulting from heuristically optimized tradeoffs 
% 	 (HOT). This network has very different structural and behavioral properties, even when 
% 	 it has the same number of nodes, links, and degree distribution as a scale free 
% 	 network shown in Figure 9.
% 	 (First published in Notices of the American Mathematical Society, Volume 56, 
% 	 No.3 (May 2009): 586-599
%          \cite{willinger09:_mathem_and_inter}. Included here by
%          permission.)
%          \vspace{-4mm}
%       \label{fig:hot_5}}
%     % \url{http://www.ams.org/notices/200905/rtx090500586p.pdf}  (figure 5)
%   \end{center}
% \end{figure}

\begin{figure}[thbp] 
  \begin{center}
% pdfimages gets actual 'images' (bitmaps) out of the PDF, but not plots
% pdfjam ams-figs-final.pdf '5' --outfile ams-figs-p5.pdf
%     extracts a particular page so I can use it here
    \begin{subfigure}[b]{0.44\textwidth} \centering
      \includegraphics*[width=\textwidth,viewport=0mm 165mm 97mm 215mm]{ams-figs-p5.pdf}\\
      \[ \begin{array}{rcl}
            \displaystyle \max_{\alpha} \sum_{i,j} x_{ij} & = & \displaystyle \max_\alpha \alpha \sum_{i,j} B_i B_j \\
            \displaystyle s.t. \sum_{i,j: k\in r_{ij}} x_{ij} & \leq & C_k, \;\forall k 
        \end{array}
      \]
      \caption{Optimisation.}
    \end{subfigure}
    \begin{subfigure}[b]{0.52\textwidth} \centering
      \includegraphics*[width=\textwidth,viewport=97mm 130mm 210mm 215mm]{ams-figs-p5.pdf}
%                                                      xleft ybottom xright ytop
%                                                   a4 = 0 0 210 297
      \caption{Result.}
     \end{subfigure}
    \caption{Generating networks using constrained optimization. (a) Engineers 
	 view network structure as the solution to a design problem that measures performance 
	 in terms of the ability to satisfy traffic demand while adhering to node and arc 
	 capacity constraints. (b) A network resulting from heuristically optimized tradeoffs 
	 (HOT). This network has very different structural and behavioral properties, even when 
	 it has the same number of nodes, links, and degree distribution as a scale free 
	 network shown in Figure 9.
	 (First published in Notices of the American Mathematical Society, Volume 56, 
	 No.3 (May 2009): 586-599
         \cite{willinger09:_mathem_and_inter}. Included here by
         permission.)
         \vspace{-4mm}
      \label{fig:hot_5}}
    % \url{http://www.ams.org/notices/200905/rtx090500586p.pdf}  (figure 5)
  \end{center}
\end{figure}

While all the simplifying assumptions can easily be relaxed to allow
for more realistic objective functions, more heterogeneity in the
constraints, or more accurate descriptions of the uncertainty in the
environment, \autoref{fig:hot_5} illustrates the key characteristics
inherent in a heuristically optimal solution of such a problem. First,
the cost-effective handling of end user demands avoids long-haul
connections (due to their high cost) and is achieved through traffic
aggregation starting at the edge of the network via the use of
high-degree routers that support the multiplexing of many
low-bandwidth connections. Second, this aggregated traffic is then
sent toward the {\em backbone} that consists of the fastest or
highest-capacity routers (\ie having a small number of very
high-bandwidth connections) and that forms the network's mesh-like
core. The result is a network of the form described earlier: a more or
less explicit backbone representing the network core and tree-like
access networks surrounding this core, with additional connections
 as backup in case of failures or congestion.

 The realism of this reverse-engineering approach to router topology
 modeling is demonstrated in \autoref{fig:cenic_abilene} which shows
 the router topologies of two actual networks -- CENIC (circa 2004)
 and Abeline (circa 2003).

\begin{figure}[p] 
  \begin{center}
%  pdfjam topology-sigcomm04.pdf '5' --outfile topology-sigcomm04-p5.pdf

    \includegraphics*[width=\columnwidth,viewport=20mm 174mm 190mm 265mm]{topology-sigcomm04-p5.pdf}
 %                                                    xleft ybottom xright ytop
%                                                 a4 = 0 0 210 297

    \caption{CENIC and Abilene networks. (Left): CENIC backbone. The
      CENIC backbone is comprised of two backbone networks in
      parallel -- a high performance (HPR) network supporting the
      University of California system and other universities, and the
      digital California (DC) network supporting K-12 educational
      initiatives and local governments. Connectivity within each POP
      is provided by Layer-2 technologies, and connectivity to the
      network edge is not shown. (Right): Abilene network. Each node
      represents a router, and each link represents a physical
      connection between Abilene and another network. End user
      networks are represented in white, while peer networks (other
      backbones and exchange points) are represented in gray. Each
      router has only a few high bandwidth connections, however each
      physical connection can support many virtual connections that
      give the appearance of greater connectivity to higher levels of
      the Internet protocol stack. ESnet and G\'{E}ANT are other backbone
      networks. (Reprinted from \cite{Li04}; \copyright 2004 ACM, Inc. Included here by permission.)
     \label{fig:cenic_abilene}}
  \end{center}
    % \url{http://netlab.caltech.edu/publications/topology-sigcomm04.pdf} (figure 4)
\end{figure}

\subsubsection{On the (ir)relevance of node degree distributions}

The above description of our engineering-based first-principles
approach to router topology modeling shows that node degree
distributions in general and power low-type node degree distributions
in particular are clearly a non-issue and play no role whatsoever in
our formulation of an ISP router topology design as a constrained
optimization problem. Thus, we achieved our goal of developing a
network modeling approach that does not rely in any way on the type of
measurements that have informed previous network modeling approaches
but have been shown earlier to be of insufficient quality to be
trusted to form the basis of any scientifically rigorous modeling
pursuit.

However, even if the available traceroute measurements could be trusted and taken 
at face value, the popular approach to network modeling that views it as an exercise
in model fitting is by itself seriously flawed, unless it is accompanied by a rigorous
validation effort.  For example, assuming that the data can be trusted so that
a statistic like an inferred node degree distribution is indeed solid and reliable.
In this case, who is to say that a proposed model's ability to match this or any other 
commonly considered statistics of the data argues for its validity, which is in
essence the argument advanced by traditional approaches that treat network modeling
as an exercise in model fitting?  It is well known in the mathematics literature 
that there can be many different graph realizations for any particular node degree sequence
and there are often significant structural differences between graphs having
the same degree sequence. Thus, two models that match the data equally well with 
respect to some statistics can still be radically different in terms of other properties, 
their structures, or their functionality. A clear sign of the rather precarious
current state of network-related modeling that is rooted in the almost exclusive focus
on model fitting is that the same underlying data set can give rise to very
different, but apparently equally ``good'' models, which in turn can give rise to completely 
opposite scientific claims and theories concerning one and the same observed phenomenon.
Clearly, network modeling and especially model validation ought to mean more
than being able to match the data if we want to be confident that the results that we 
derive from our models are relevant in practice.

\begin{figure}[thbp] 
  \begin{center}
%  pdfjam topology-sigcomm04.pdf '5' --outfile topology-sigcomm04-p5.pdf

    \includegraphics*[width=\columnwidth,viewport=20mm 167mm 190mm 265mm]{topology-sigcomm04-p9.pdf}
 %                                                    xleft ybottom xright ytop
%                                                 a4 = 0 0 210 297

    \caption{
      Five networks having the same node degree distribution. (a) Common node degree
	 distribution (degree versus rank on log-log scale); (b) Network resulting from 
	 preferential attachment; (c) Network resulting from the GRG method; (d) Heuristically 
	 optimal topology; (e) Abilene-inspired topology; (f) Sub-optimally designed topology.
	 (Reprinted from \cite{Li04}; \copyright 2004 ACM, Inc. Included here by permission.)
\label{fig:node_degree}}
  \end{center}
    % \url{http://netlab.caltech.edu/publications/topology-sigcomm04.pdf} (figure 6)
\end{figure}

To illustrate these points, \autoref{fig:node_degree} depicts five
representative toy networks, constructed explicitly to have one and
the same node degree distribution. This distribution is shown in plot
(a) and happens to be the one of our HOT router topology example in
\autoref{fig:hot_5}. While plots (b) and (c) show two scale-free
networks constructed according to the preferential attachment method
and general random graph method, respectively, plots (d)-(f) are
three different HOT examples, including our earlier example in
\autoref{fig:hot_5} (plot (e)) and a sub-optimal or poorly-engineered
HOT topology in (f). While the differences among these five topologies
with identical node degree distributions are already apparent when
comparing their connectivity structures, they can be further
highlighted by considering both a performance-related and a
connectivity-only topology metric.  In particular, the
performance-related metric {\em Perf(g)} for a given network $g$ is
defined as {\em{Perf(g)}} $ = max_{\alpha} \sum x_{i,j}
\mbox{~~s.t.~~} RX \leq C$ and represents the maximum throughput with
gravity flows of the network $g$. In contrast, the connectivity-only
topology metric {\em S(g)} is the network likelihood of $g$ defined as
{\em S(g)} $ = \sum_{(i,j) \in E(g)} \omega_i \omega_j /s_{max}$ where
$\omega_i$ denotes the degree of node $i$, $E(g)$ is the set of all
edges in $g$, and $s_{max}$ is a normalization constant. For a
justification of using the $S(g)$ metric to differentiate between
random networks having one and the same node degree sequence, we refer
to \cite{Li_internet_math}.

While computing for each of the five networks their {\em Perf(g)}-value is straightforward, 
evaluating their network performance requires further care so as to ensure that the 
different network have the same total ``cost", where cost is measured in number of routers.
When simultaneously plotting network performance versus network likelihood for all
five networks models in \autoref{fig:perf_vs_likeli}, a striking contrast is observed. The well-engineered
HOT networks (d) and (e) have high performance and low likelihood while the random
degree-based networks (b) and (c) have high likelihood but low performance. To contrast,
network (f) has both low performance and low likelihood and is proof that networks can be
designed to have poor performance. The main reason
for the degree-based models to have such poor performance is exactly the presence of the 
highly connected ``hubs'' that create low-bandwidth bottlenecks. The two HOT models'
mesh-like cores, like real ISP router topologies, aggregate traffic and disperse it
across multiple high-bandwidth routers.

\begin{figure}[thbp] 
  \begin{center}
%  pdfjam topology-sigcomm04.pdf '10' --outfile topology-sigcomm04-p10.pdf     
    \includegraphics*[width=0.7\columnwidth,viewport=18mm 202mm 100mm 268mm]{topology-sigcomm04-p10.pdf}
%                                                    xleft ybottom xright ytop
%                                                 a4 = 0 0 210 297
     \caption{Performance vs. likelihood for each of the topologies in \autoref{fig:node_degree}, plus 
	 other networks (grey dots) having the same node degree
         distribution obtained by pairwise 	 random rewiring of
         links. 
	 (Reprinted from \cite{Li04}; \copyright 2004 ACM, Inc. Included here by
         permission.)
         \label{fig:perf_vs_likeli}}
  \end{center}
   % \url{http://netlab.caltech.edu/publications/topology-sigcomm04.pdf} (figure 8)
\end{figure}

The interpretation of this picture is that a careful design process
explicitly incorporating technological constraints can yield
high-performance topologies, but these are extremely rare from a
probabilistic graph point of view. In contrast, equivalent scale-free
networks constructed by generic degree-based probabilistic
constructions result in more likely, but poorly-performing
topologies. Consistent with this, the ``most likely'' network
(included in \autoref{fig:perf_vs_likeli}) has also sub-par performance. This picture can be
further enhanced when considering alternative performance measures
such as the distribution of end user bandwidths and router
utilization. As detailed in \cite{Li04}, the heuristically optimal
networks (d) and (e) achieve high utilization in their core routers
and support a wide range of end-user bandwidth requirements. In
contrast, the random degree-based networks (b) and (c) saturate only
their ``hub'' nodes and leave all other routers severely
underutilized, thus providing uniformly low bandwidth and poor
performance to their end-users. A main lesson from this comparison of
five different networks with identical node degree distributions for
network modeling is that {\it functionality (\eg performance) trumps structure (\eg connectivity)}. That is, connectivity-only metrics
are weak discriminators among all graph of a given size with the same
node degree distribution, and it requires appropriate
performance-related metrics to separate "the wheat from the chaff."

We explained earlier that on the basis of currently available traceroute
measurements, claims of power-law relationships have no substance as far as
the Internet's router topology is concerned.  However, by examining available
router technologies and models, we have also shown that it is certainly conceivable 
that the actual node degrees of deployed routers in an actual ISP can span a range 
of 2-3 orders of magnitude; that is, the corresponding node degree distribution exhibits 
high variability, without necessarily conforming to a power law-type distribution. 
At the same time, \autoref{fig:node_degree} illustrates that irrespective of the type of node degree
distribution, graphs with identical node degree distributions can be very different 
in their structure and differ even more drastically in terms of their functionality
(\eg performance). What is also true is that the same core network design can 
support many different end-user bandwidth distributions and that by and large, 
the variability in end-user bandwidth demands determines the variability of the 
node degrees in the resulting network. To illustrate, consider the simple example
presented in \autoref{fig:node_degree_dist}, where the same network core supports different types of 
variability in end user bandwidths at the edge (and thus yields different overall 
node degree distributions). The network in \autoref{fig:node_degree_dist}(a) provides uniformly high 
bandwidth to end users; the network in \autoref{fig:node_degree_dist}(b) supports end user bandwidth 
demands that are highly variable; and the network in \autoref{fig:node_degree_dist}(c) provides uniformly low bandwidth to end users. Thus, from an engineering perspective, not only 
is there not necessarily any implied relationship between a network node degree 
distribution and its core structure, there is also no implied relationship between 
a network's core structure and its overall degree distribution.

\begin{figure}[!bp] 
  \begin{center}
%  pdfjam topology-sigcomm04.pdf '11' --outfile topology-sigcomm04-p11.pdf     
    \includegraphics*[width=\columnwidth,viewport=20mm 213mm 190mm 268mm]{topology-sigcomm04-p11.pdf}
%                                                    xleft ybottom xright ytop
%                                                 a4 = 0 0 210 297
     \caption{Distribution of node degree and end-user bandwidths for
       several topologies having the same core structure: (a)
       uniformly high bandwidth end users, (b) highly variable
       bandwidth end users, (c) uniformly low bandwidth end
       users. 
       (Reprinted from  \cite{Li04}; \copyright 2004 ACM, Inc. Included here by permission.)
       \label{fig:node_degree_dist}}
  \end{center}
   % \url{http://netlab.caltech.edu/publications/topology-sigcomm04.pdf} (figure 9)
\end{figure}

Thus, the proposed engineering-based first-principles approach to
modeling the Internet router topology demystifies power law-type node
degree distributions altogether by identifying its root cause in the
form of high variability in end-user bandwidth demands. In view of
such a simple physical explanation of the origins of node degree
variability in the Internet's router-level topology, Strogatz'
question, paraphrasing Shakespeare's Macbeth, ``... power-law scaling,
full of sound and fury, signifying nothing?'' \cite{strogatz05:_roman}
has a resounding affirmative answer.

\subsection{A look ahead}

Late in the last century, when router-level topology modeling started
to turn into a measurement-driven research activity, the conventional
wisdom was to start with traceroute-based measurements, use them to
infer router-level connectivity, and argue for the validity of a
proposed model if it faithfully reproduces certain statistics of the
inferred connectivity structure (\eg node degree distribution).
However, the last decade of Internet topology research has shown that
this traditional and widely-used approach to router-topology modeling
is flawed in more than one way, and we have collected and presented
this gradually accumulating evidence in this section -- the underlying
measurements are highly ambiguous (\autoref{sec:router_measurements}),
the inferred connectivity structures are erroneous
(\autoref{sec:router_models}), and the resulting models are infeasible
and/or do not make sense from an engineering perspective because they
are either too costly, have extremely poor performance, or cannot be
built with from existing technology in the first place.

This section also describes and reports on an alternative design-based approach
to router-level topology modeling and generation that has come into focus during the
last 5-10 years and represents a clean break with tradition. The most visible sign 
of this break is the emergence of constrained optimization as new modeling language,
essentially replacing the traditional language of random graph theory.  While the 
latter treats nodes and links as largely generic objects and focuses almost exclusively 
on structural aspects such as connectivity, the former supports a much richer treatment 
of topologies -- nodes and links are components with their own structure, constraints, 
and functionality, and their assembly into a topology that is supposed to achieve a 
certain overall objective and should do so efficiently and effectively within the
given constraints on the individual components or the system as a whole is the essence 
of the constrained optimization-based network design approach.  In essence, this approach
echoes what was articulated some 15 years ago in~\cite{Doar96,Calvert97,Zegura97},
but it goes beyond this prior work in terms of empirical evidence, problem formulation,
and solution approach. As a result, the described design-based approach has by and large
put an end to graph theory-based router topology modeling for the Internet.

At the same time, the design-based approach that has been developed and reported
in bits and pieces in the existing literature and is presented in this section for
the benefit of the reader in one piece has also far-reaching implications for 
Internet topology research in general and router-level topology modeling, analysis, 
and generation in particular.  For one, it shows the largely superficial nature 
of router-level topology generators that are based on graph-theoretic models.  
As appealing as they may be to a user because of their simplicity (after all, 
all a user has to specify is in general the size of the graph), they are by and 
large of no use for any real application where details like traffic, routing, 
capacities, or functionality matter.

Second, while the design-based approach yields realistic router-level
topology models that are inherently generative in nature, it puts at
the same time an end to the popular request for a largely generic
black-box-type topology generator.  Users in real need for synthetic
router-level maps have to recognize that this need doesn't come for
free. Instead, it comes with the responsibility to provide detailed
input in terms of expected customers -- their geographic dispersion,
and the traffic matrix (see \cite{tune13:_sigcom_ebook} for more
details) -- design objectives and constraints, etc.  In addition, the
level of detail required of a generated ISP router-level topology (\eg
POP-, router-, interface card-level) depends critically on and cannot
be separated from the purpose for which these generated maps will be
used.  Again, this puts a considerable burden on the user of a
synthetically generated map and tests her understanding of the
relevant issues to a degree unheard of in Internet topology modeling
in the past.

Third, the explicit focus of the design-based approach on ISPs as crucial 
decision makers renders the commonly-expressed desire for synthetic router-level 
maps of the global Internet largely pointless. The Internet is a network of 
networks, with the sovereign entities being the autonomous systems (ASes). A subset
of these ASes that are in the business of providing network service to other ASes
or Internet access to end users are owning and operating their networks that 
together make up much of the physical infrastructure of the global Internet. 
As a result, a key first step in understanding the structure and temporal 
evolution of the Internet at the different physical and logical layers is 
to study the physical infrastructures of the service and access
providers' networks and how they react in response to changes in the
environment, technology, economy, etc.

Finally, once we have a more-or-less complete picture of the
router-level topology for the individual ISPs, we can start
interconnecting them at common locations, thereby bringing ISP
router-level and AS-level topology under one umbrella.  In the
process, it will be critical to collapse the detailed router-level
topologies into their corresponding PoP-level maps which are
essentially the geographic maps mentioned in the context of the
ARPANET in \autoref{sec:router_hist} and serve as glue between the
detailed router-level topologies and an appropriately defined and
constructed AS-level topology of the Internet.  For a faithful and
realistic modeling of this combined router-, POP-, and AS-level
structure of the Internet, it will be important to account for the
rich structure that exists in support of network interconnections in
practice.  This structure includes features such as third-party
colocation facilities that house the PoPs of multiple ASes in one and
the same physical building. It also includes components of the
Internet's infrastructure such as Internet eXchange Points (IXPs).
This existing structure is inherently non-random but is a reflection
of the incentives that exist, on the one hand, for network and access
providers to build, manage, and evolve their physical infrastructures
and, on the other hand, for content providers, CDNs, and cloud
providers to establish peering relationships with interested
parties. Importantly, neither such structures nor incentives precludes
an application of the described constrained optimization-based
approach to network design; they merely require being creative with
respect to formulating a proper objective function, identifying the
nature of the most critical constraints, and being able to pinpoint
the main sources of uncertainty in the environment.

\vspace{-2mm}
\subsection{Notes}
\vspace{-1mm}

The primary sources for the material presented in this section are:

\begin{itemize}

\item[\cite{Li04}] L. Li, D. Alderson, J.C. Doyle and W. Willinger.
A first principles approach to understanding the Internet's router-level topology, 
{\em Proc. ACM SIGCOMM'04, ACM Computer Communication Review} 34(4), 2004.

\item[\cite{Doyle05}] J. C. Doyle, D. L. Alderson, L. Li, S. Low, M. Roughan, S. Shalunov, 
R. Tanaka, and W. Willinger. 
The "robust yet fragile" nature of the Internet, 
{\em PNAS} 102(41), 2005.

\item[\cite{willinger09:_mathem_and_inter}] W. Willinger, D. Alderson, and J.C. Doyle. 
Mathematics and the Internet: A Source of Enormous Confusion and Great Potential, 
{\em Notices of the AMS} 56(5), 2009.

\end{itemize}

An excellent short treatment of the discussion in \autoref{sec:router_models} about network modeling
as an exercise in reverse-engineering vs. as an exercise in model fitting can be 
found in Chapter 10 of the recent book

\begin{itemize}

\item[\cite{chiang12:_networ_life}] M. Chiang.
Networked Life: 20 Questions and Answers.
Cambridge University Press, 2012.

\end{itemize}

For additional and more in-depth reading materials we  point to

\begin{itemize}

\item[\cite{Alderson05}] Alderson, D., Li, L., Willinger, W., and Doyle, J.C. 
Understanding Internet Topology: Principles, Models, and Validation, 
{\em IEEE/ACM Transactions on Networking} 13(6): 1205-1218, 2005.

\item[\cite{krishnamurthy11:_what}] B. Krishnamurthy and W. Willinger. 
What are our standards for validation of measurement-based networking research? 
{\em Computer Communications} 34, 2011.

\end{itemize}

For some "food for thought" regarding topics such as power law distributions and 
scale-free networks, and network science, we recommend 
\begin{itemize}

\item[\cite{Willinger04}] W. Willinger, D. Alderson, J.C. Doyle, and L. Li. 
More ``normal'' than normal: Scaling distributions and complex systems. 
In: R.G. Ingalls, M.D. Rossetti, J.S. Smith, and B.A. Peters (Editors).
IEEE. Proc. of the 2004 Winter Simulation Conference, Piscataway, NJ, 2004.

\item[\cite{Mitzenmacher01}] M. Mitzenmacher.
A Brief History of Generative Models for Power Law and Lognormal Distributions,
Internet Mathematics 1(2):226-251, 2004.

\item[\cite{keller05:_revis}] E. Fox Keller.
Revisiting ``scale-free'' networks,
BioEssays 27(1)):1060-1068, 2005.

\item[\cite{barabasi09:_scale_free_networ}] A.-L. Barabasi.
Scale-free Networks: A Decade and Beyond,
Science 325, 2009.

\item[\cite{barabasi12}] A.-L. Barabasi.
The network takeover,
Nature Physics 8, pp. 14-16, 2012.

\item[\cite{mitzenmacher06:_editor}] 
	M. Mitzenmacher.
	Editorial: The Future of Power Law Research.
	{\em Internet Mathematics}, vol. 2. no. 4, pp. 525-534, 2006.

\end{itemize}

% \subsection{Additional References}

% Heart, F., McKenzie, A., McQuillian, J., and Walden, D.
% ARPANET Completion Report, Bolt, Beranek and Newman, Burlington, MA, January 4, 1978. 
% \url{http://www.cs.utexas.edu/users/chris/DIGITAL_ARCHIVE/ARPANET/DARPA4799.pdf}
% \cite{heart78:_arpanet}

% The Internet: Changing the way we communicate
% \url{http://www.nsf.gov/about/history/nsf0050/pdf/internet.pdf}
% \cite{NSFinternet}

% K.D. Frazer, \cite{frazer:_merit_histor_nsfnet_backb_projec}
% Merit's History, The NSFNET Backbone Project 1987-1995, Merit Network, Inc.
% \url{http://www.livinginternet.com/doc/merit.edu/phenom.html}

% A.-L. Barabasi and R. Albert. \cite{barabasi99}
% Emergence of scaling in random networks.
% Science 286, 509-512, 1999.

% F. Chung and L. Lu. \cite{Chung04}
% The average distance in a random graph with given expected degrees.
% Internet Mathematics 1, 91-113, 2003.

% W. Aiello, F. Chung, and L. Lu. \cite{Aiello00}
% A Random Graph Model for Massive Graphs.
% Proc. STOC, 2000.

% R. Albert, H. Jeong, and A.-L. Barabasi. \cite{barabasi00}
% Attack and error tolerance of complex networks.
% Nature 406, 378-382, 2000.

% R. Pastor-Satorras and A. Vespignani. \cite{pastor-satorras01:_epidem}
% Epidemic spreading in scale-free networks.
% Physical Review Letters 86(14), pp. 3200-3203, 2001.

% M.E.J. Newman. \cite{Newman03}
% The Structure and Function of Complex Networks.
% SIAM Review 45, 167-256, 2003.

% A.-L. Barabasi.\cite{barabasi09:_scale_free_networ}
% Scale-Free Networks: A Decade and Beyond.
% Science 325, 412-413, 2009.

% D. Clark.\cite{clarke95:_design}
% The Design Philosophy of the DARPA Internet Protocols.
% Proc. ACM SIGCOMM'88, Stanford, CA, pp. 109-114, 1988.

% J. Sherry, E. Katz-Bassett, M. Pimenova, H.V. Madhyastha, T. Anderson, and A. Krishnamurthy.\cite{sherry10:_resol_ip_alias_presp_times}
% Resolving IP Aliases with Prespecified Timestamps.
% Proc. ACM SIGCOMM Internet Measurement Conference IMC'10, 2010.

% A. Bender, R. Sherwood, and N. Spring. \cite{bender08:_fixin_allys}
% Fixing Ally's growing pains with velocity modeling. 
% Proc. ACM SIGCOMM Internet Measurement Conference IMC'08, 2008.

% M. Gunes and K. Sarac. \cite{gunes09:_resol_ip_inter}
% Resolving IP aliases in building traceroute-based Internet maps.
% IEEE/ACM Transactions on Networking 17(6):1738-1751, 2009.

% J. Sommers, B. Eriksson, and P. Barford. \cite{sommers11:_preval_charac_mpls_deploy_open_inter}
% On the Prevalence and Characteristics of MPLS Deployments in the Open Internet.
% Proc. ACM SIGCOMM Internet Measurement Conference IMC'11, Berlin, Germany, 2011.

% A. Lakhina, J. W. Byers, M. Crovella, and P. Xie.\cite{LakhinaByersCrovellaXie03}
% Sampling Biases in IP topology Measurements.
% Proc. IEEE INFOCOM'03, 2003.

% D. Achlioptas, A. Clauset, D. Kempe, and C. Moore.\cite{achlioptas05:_bias_of_tracer_sampl}
% On the bias of traceroute sampling.
% Journal of the ACM 56(4), 2009.

### 2013-07-22 Compiled/chapter8/abstract.tex
all files compile correctly

\begin{abstract}
  The increasing demand of various services from the Internet has led
  to an exponential growth of Internet traffic in the last decade, and
  that growth is likely to continue. With this demand comes the
  increasing importance of network operations management, planning,
  provisioning and traffic engineering. A key input into these
  processes is the \textit{traffic matrix}, and this is the focus of
  this chapter.

  The traffic matrix represents the volumes of traffic from sources to
  destinations in a network. Here, we first explore the various issues
  involved in measuring and characterising these matrices. The
  insights obtained are used to develop models of the traffic,
  depending on the properties of traffic to be captured: temporal,
  spatial or spatio-temporal properties. The models are then used in
  various applications, such as the recovery of traffic matrices,
  network optimisation and engineering activities, anomaly detection
  and the synthesis of artificial traffic matrices for testing routing
  protocols. We conclude the chapter by summarising open questions in
  Internet traffic matrix research and providing a list resources
  useful for the researcher and practitioner.
\end{abstract}

### 2013-07-22 Compiled/chapter8/applications.tex
all files compile correctly

\clearpage
\section{Applications}
\label{sec:applications}

\subsection{Traffic Matrix Recovery}

As noted earlier it may be difficult to measure traffic matrices
directly, but we need to recover traffic matrices from measurements
before they can be used. The technique will obviously depend on the
available measurements, but there is a glut of works on the recovery
of traffic matrices, whether at the OD level, IE level or AS
level. Amongst these traffic matrices, IE traffic matrices are
relatively easier to recover, as measurements of these matrices are
available through SNMP link counts. Recovery of OD level traffic
matrices are fraught with challenges because at any point in time,
only a subset of IP traffic is seen by a network. There is no way to
know what goes on in the entire IPv4 address range, 
unless all measurements of the
global network were combined, but even so, the data from such an
endeavour would be massive and computationally intractable to
analyse, let alone accurate in the first place. Similarly, for AS level 
traffic matrices, the lack of measurements as well as error prone 
measurement tools lead to inaccurate recovery of these matrices.

The major challenge in recovering the IE traffic matrix from SNMP
measurements is that the problem is highly \emph{underconstrained}.
The set of linear equations \autoref{eq:observation} is
under-determined, \ie~there are many solutions to the
observations. Recall that the measurements themselves are subject to
error possibly due to poor data collection methods and poor vendor
implementation of SNMP polling.  They are also not fine-grained since
polling of the measurements is performed every 5 minutes (and the
polling intervals may not be perfectly synchronised across a whole
network). Clearly, any inference method is required to be robust
against these errors and uncertainties.

The underconstrainedness of the problem may be mitigated by active
measures. One is direct measurement, using dedicated monitors or
in-built measurement software on routers such as NetFlow
\cite{NetFlow}.  Direct measurements at even a single point of ingress
results in measurement of an entire row of the traffic matrix,
drastically reducing the number of missing matrix entries. Another
interesting proposal is to change the IGP (Interior Gateway Protocol)
link weights over several snapshots within the measurement interval to
provide fresh sets of observations, thereby resulting in a system of
linear equations with a unique solution (full rank) out of the SNMP
measurements \cite{Nucci04IGPchange,soule07:_estim_dynam_traff_matric_using}. 
Both these techniques may be
impractical, either being too costly in the case of direct
measurements, or requiring direct intervention by the network operator
for IGP weight changes. Most proposals simply avoid these by settling
on a passive approach of inferring the traffic matrix straight from
SNMP data.

There are two main approaches to traffic matrix inference. The first is the deterministic approach, where
$\by$ is assumed to provide hard constraints, rather than statistical
data. Goldschmidt~\cite{Goldschmidt00LP} formulated this as
a Linear Program (LP) where the objective was designed 
 to find bounds on traffic matrix elements. In simple terms,
the LP finds the traffic matrix with the worst case upper and lower
bound on the traffic demand subject to constraints. Recall the vectorised 
traffic matrix $\bx$ has size $N(N-1)$. For the upper
bound, the LP model is defined with the objective function 
\be
\max_{\bx} \sum_{j=1}^{N(N-1)} \omega_j x_j,
  \label{eq:lp}
\ee

\noindent where $\omega_j$ is a weight for an OD pair $j$, also called
the coefficient of demand. There are three constraints to satisfy,
namely,
\begin{enumerate}
\item \textbf{observation constraints}:
\be
\sum_{j=1}^{N(N-1)} A_{ij} x_j \le y_i, \ i=1,2,\cdots, L,
\label{eq:observation_constraint}
\ee
\item \textbf{flow conservation constraints}:
\be
\sum_{\underset{\ell_1\ne\ell_2}{\ell_1 = i, \ell_2 =j,}} y_{\ell_1} A_{\ell_2 k} 
- \sum_{\underset{\ell_1\ne\ell_2}{\ell_1 = j, \ell_2 = i,}} y_{\ell_1} A_{\ell_2 k} = 
\begin{cases}
x_k, & \text{if $j$ is the source of $k$},\\
-x_k, & \text{if $j$ is the destination of $k$},\\
0, & \text{otherwise}.
\end{cases}
\label{eq:flow_conservation}
\ee
\item \textbf{non-negativity constraints}: $x_j \ge 0$, $j=1,2,\cdots, N(N-1)$.
\end{enumerate} 
\medskip

\noindent Similarly, the lower bound is found by substituting the
maximisation operation in \eqref{eq:lp} with a minimisation operation.
The LP only produces a nontrivial solution if the lower bound and
upper bound on the traffic demand is greater than zero and less than
the observed total link count, \ie~$\sum_j y_j$.

Unfortunately, the utility of the LP is only restricted to small toy
problems. First, two linear programs have to be solved each time to
obtain the upper and lower bounds on traffic demands, which is
computationally expensive for large $N$.  Second, the LP was shown to have
terrible performance when tested on several types of traffic matrices
\cite{Medina02TMdirections}. Estimates of some traffic matrix entries
were in excess of 200$\%$, with most in excess of 100$\%$ error,
proving that while the LP may be useful for certain small topologies, in
general it is not considered a practical estimation method. The reason
for this is because the LP sets many estimated values to zero, resulting
in overcompensation for the rest of the estimated values in order to
meet the total traffic constraints. Third, there is a high sensitivity
of the solution to weight choices, which implies that different
solutions will be obtained depending on the chosen weights.

Instead, a more successful alternative is the use of statistical
models and regularisation, \ie\ treating the traffic matrix as a
realisation of a random process generated from a model. Regularisation
refers to the inference technique of imposing additional structural
assumptions on the problem to reduce
underconstrainedness. Regularisation methods are defined by four
components:
\begin{enumerate}
\item a \textbf{prior solution}, generated from a \emph{model}, 
\item a \textbf{model deviation} measure, used to compute the deviation of a feasible solution from the model, 
\item a \textbf{distortion} measure, used to compare the deviation of the model with the observations, and
\item an \textbf{adjustment step}, to ensure the constraints on the total traffic entering and exiting all ingress and egress nodes 
respectively, as well as non-negativity constraints, are satisfied.
\end{enumerate}
In terms of an optimisation procedure, solving the tomography problem is equivalent to
\begin{eqnarray}
\label{eq:opt_TMinverse}
\bx^\star = \argmin_{\bx \in \Real^{N(N-1)}} & R(\bx,\by) + \lambda d(\bx,\cM),
\end{eqnarray}
where $R(\cdot,\cdot)$ denotes the distortion measure,
$d(\cdot,\cdot)$ denotes the model deviation measure and $\lambda \ge
0$ is the penalty constant that amplifies the penalisation of a
feasible solution which strays too far away from the model\footnote
{Technically, $\bx$ comprises non-negative integers, but a relaxation
  to real numbers is used as it is easier to compute, especially when
  considering large traffic matrices.}. Typically, $R(\bx,\by) = \|\by
- \bA \bx\|_2$. Regularisation techniques are \emph{biased} to a
particular prior model. Thus, if the model is inconsistent, then the
estimator \eqref{eq:opt_TMinverse} would be inconsistent as
well. However, if the prior model chosen describes the final solution
somewhat accurately, then it is expected that the final estimate would
be fairly accurate.

As an example, suppose the prior
model, $\bx^{(0)}$, used is the gravity model, which can be derived
from link measurements by calculating the ingress and egress traffic
volumes (by summing link measures on the edge-links of the network). 
One proposed penalty~\cite{Zhang05InfoTh} is defined as 
\be d(\hat{\bx},\bx^{(0)}) = H(\hat{\bx}) + H(\bx^{(0)}) - H(\hat{\bx},\bx^{(0)}),
\label{eq:entropy_penalty}
\ee
where 
\be
H(\bx) = - \sum_{j=1}^{N(N-1)} \frac{x_j}{\sum_{k=1}^{N(N-1)} x_k} \log \frac{x_j}{\sum_{k=1}^{N(N-1)} x_k},
\label{eq:entropy}
\ee
is the empirical entropy, while 
\be
H(\bx,\bx^{(0)}) = \sum_{j=1}^{N(N-1)} \frac{x_j}{\sum_{k=1}^{N(N-1)} x_k} \log (\frac{x_j}{\sum_{k=1}^{N(N-1)} x_k} \Big/  
\frac{x^{(0)}_j}{\sum_{k=1}^{N(N-1)} x^{(0)}_k}),
\label{eq:joint_entropy}
\ee
is the joint empirical entropy, between the estimate and the prior model. The penalty function \eqref{eq:entropy_penalty} measures 
the uncertainty between the quantities $\bx$ and $\bx^{(0)}$, and is commonly known as the \emph{mutual information} 
\cite{Cover06InfoTheory}. The joint entropy term $H(\bx,\bx^{(0)})$ quantifies the uncertainty between $\bx$ and 
$\bx^{(0)}$. If $H(\bx,\bx^{(0)}) = 0$, then $\bx$ is statistically
independent of $\bx^{(0)}$.

The approach is highly flexible: it can deal with the generalised
gravity model simply by using a new prior model and constraints
\eqref{eq:conditional_independence} are added to account for the
different traffic classes (access and peering traffic).

The penalty can be rewritten and thought of as the Kullback-Leibler
distance \cite{Cover06InfoTheory} between the estimate $\hat{\bx}$ and
the prior model $\bx^{(0)}$, implying that the estimation objective
seeks to preserve as much prior information from $\bx^{(0)}$ as
possible, while minimising $R(\bx,\by)$. This can be used directly, or
approximated, for instance as a weighted quadratic
\cite{Zhang03InfoSIGCOMM,Zhang03Fast}.

Using suitable models, most of the existing inference methods can be
described in this framework (see \cite{Zhang05InfoTh} for details).
Or, other penalties can be used, such as the nuclear norm, given by
\be 
  d(\bX) =\|\bX\|_{\ast} = \sum_{i = 1}^r \sigma_i
  \label{eq:nuclear_penalty}
\ee
for low rank model recovery.

The solution of the optimisation procedure is often adjusted after
regularisation using Iterative Proportional Fitting (IPF)
\cite{Deming40IPF}, so as to satisfy the observed total traffic
constraints and non-negativity constraints (those that were not
included in the regularisation for computational reasons). In
practice, the IPF is a very simple algorithm, performing fast even on
large traffic matrices.

\begin{figure}[ht]
\Topology
\caption{Three example topologies where local traffic matrices provide
  benefits (motivated by the seminal figure in
  \cite{baran64:_distr}). \textbf{Left}: centralised, or star,
  topology, \textbf{Centre}: decentralised topology, \textbf{Right}:
  distributed topology.}
\label{fig:topology}
\end{figure}

% An important issue for regularisation methods is how much of an
% improvement regularisation would be over the initial ill-posed
% problem. Intuitively, if the model and constraints provide an
% information advantage, then it is easier to infer the traffic matrix.

Additional information can also be used, for instance, if some rows of
the matrix are known from measurements, then this eases the number of
variables to be estimated, making the problem a little simpler.
Another source of potential data is the collection of \textit{local
  traffic matrices} \cite{Varghese03Measure}, providing information on
traffic between interfaces of routers. We can see why this is useful
by considering the three network topologies in Figure
\ref{fig:topology}, with a centralised or star, a decentralised and a
distributed topology. If the network has a star topology, then the
entire traffic matrix is known if the local traffic matrix of the
router right in the centre is obtained. For the other two topologies,
collection of local traffic matrices in strategic places of the
topology is likely to reduce the underconstrainedness of the original
inference problem, though less (relative) information is provided the
more distributed the topology.  Local traffic matrices have been
demonstrated to provide a significant information boost in
\cite{Zhang03InfoSIGCOMM}, especially if the interfaces are
well-connected, and that is highly dependent on the underlying network
topology. If direct flow measurements from dedicated monitors are
available, they provide a huge boon as an entire row of an IE traffic
matrix would be revealed. In practice, however, these are generally
not available as they are deemed expensive. The advantage of the
regularisation method is that these additional information may be
incorporated easily via constraints.

Another issue is their computational tractability. Speed is an issue
for these algorithms, since traffic matrices are often large. Most
model deviation and distortion measures are chosen to be 
convex\footnote{Convex relaxations of a non-convex objective function is often
used as a substitute, for \eg, the nuclear norm is used in place of the rank function
to recover low rank matrices. Under these circumstances, it is crucial to 
note the assumptions of the model, so as to know when the recovered solution
is an excellent approximation of the true solution.}, with
linear constraints. In this way, problem \eqref{eq:opt_TMinverse} becomes a
convex optimisation problem, where many fast, scalable and efficient algorithms
have been developed to solve such problems \cite{Boyd04Opt}.

The discussions here only considered point-to-point traffic
matrices. For IE matrices, the point-to-multipoint matrix may be more
useful instead. Recall from the above that an ideal traffic matrix is
invariant to other network aspects to be useful for network
design. Unlike the point-to-point traffic matrix, the
point-to-multipoint matrix contains records on the amount of traffic
from one ingress point to a set of egress points. These sets are
chosen to preserve invariance under changes in the egress point, a
property much more useful for network planning. Inference of the
point-to-multipoint traffic matrices may be done in a similar fashion
to point-to-point IE matrices \cite{Zhang03InfoSIGCOMM}.

\subsection{Network Optimisation}

An ISP of a backbone network must ensure
each link in the network has adequate capacity. The consequences of
failure to provide  such capacity is congestion, and a resulting loss
of quality service, which in severe cases would result in loss of
customers. However, over-provisioning can be wasteful, and so
optimisation is used to strike the right balance between cost and
capacity.

Network optimisation and engineering 
\cite{Mitra05StochasticTE,Buriol03GAOSPF,Murphy02TE,Fortz02OSPF,Fortz03OSPF,Nucci07IGP,Uhlig04Implications,Feamster03BGPTE,Rexford06Route} involves several tasks with varying planning
horizons. Assuming node locations are fixed, in the long term, we plan
a network by considering the link locations, and capacity. We refer to
this as {\em network planning}. It may be categorised into two common
scenarios: \emph{incremental planning} on existing networks, or
\emph{green-fields planning} \cite{Roughan10Robust}. In the former,
the planning takes an evolutionary route, since the network designer
is constrained by the current existing network. Any upgrades to the
network are deliberately incremental, so as not to disrupt current
operations. Green-fields planning, as the name suggests, starts from
scratch: the entire network is designed from ground up. Shorter-term
network optimisation tasks include traffic
engineering~\cite{Roughan03TETM} and some potential routing schemes.

As an illustration of the use of a traffic matrix for testing routing schemes, 
consider a simple example of a topology with 
three nodes in \autoref{fig:demands}. The traffic matrix is given by
\ben
\bX = 
\left(
\begin{array}{ccc}
0 & 1 & 4\\
2 & 0 & 2\\
3 & 2 & 0
\end{array}
\right).
\een
The path distances from PoP 1 to 2, 1 to 3 and 2 to 3 is 1, 2 and 4 respectively. The goal is 
to find the shortest path routing based on the traffic matrix and the topology. Recall
that the link loads may be expressed as $\by = \bA \bx$, where 
$\bx = (1,4,1,2,3,2)^\T$ is the vectorised
form of $\bX$, without including self-traffic. The routing matrix $\bA$ 
here is size 3 by 6, since there are 3 links and 
6 flows, as we do not consider self-traffic. Each element is either 0 or 1 as the
flows are assumed to be delivered whole. The capacity of each link is assumed to be
infinite, or limitless.

Thus, finding the routes may be cast as the optimisation problem
(integer program) with a simple cost function based on path distance
\be
\begin{array}{ccc}
\bA^{\star} &= \displaystyle \argmin_{{A_{i,j} \in \{0,1\}, \forall i,j}} & y_{1,2} + 2y_{1,3} + 4y_{2,3}\\
& \text{ subject to} & \by = \bA \bx.
\end{array}
\label{eq:minKL}
\ee
The solution in this case is a simple one with 
\ben
\bA^\star = 
\left(
\begin{array}{cccccc}
1 & 0 & 1 & 1 & 0 & 1\\
0 & 1 & 0 & 1 & 1 & 1\\
0 & 0 & 0 & 0 & 0 & 0
\end{array} 
\right).
\een
The example presented is a simple, trivial one. In practice, networks have a larger
topology and additional costs may be included, such
as bandwidth utilisation of each link so as to minimise the maximum utilisation of the network,
QoS constraints, and constraints on the maximum capacity of each link. For instance, 
if the maximum capacity of the link from PoP 1 to 3 is at most 4 \ie~$y_{1,3} \le 4$, 
then the routing matrix above is no longer a 
feasible solution. The new solution must incorporate the link from PoP 2 to 3 to conform to 
constraints (assuming the link 2 to 3 has adequate capacity).

\begin{figure}
  \begin{center}
    \input{./Figures/demands_example}
    \caption{Example of a simple network of three PoPs and path distances of each link. The shortest path
    routes chosen are from PoP 1 to 2, and 
    from PoP 1 to 3, as these paths have the lowest distance (corresponding to the thick lines).
      \label{fig:demands}}    
  \end{center}
\end{figure}

In all of the above mentioned tasks, one of the key ingredients is the traffic
matrix. The reason these matrices are so useful comes down to {\em
  invariance}. The traffic on a particular link obviously varies as
the links in the network change. However, an ideal traffic matrix is
invariant to other aspects of the network such as the network topology
and underlying routing protocols. Invariance allows the design to be
varied without the inputs to the process changing.  To a certain
extent, the IE traffic matrix satisfies the invariance property, but
it is far from perfect for some tasks, most notably it is highly
sensitive to external routing changes and some internal changes
\cite{Teixeira05TMRoute,Teixeira04PotatoSIG}.  The OD matrix is in
some sense preferable \cite{Alderson06Topology}, but harder to measure
in most cases.  The point-to-multipoint IE matrix (discussed in the
previous section) is a useful compromise.

Furthermore, a network operator would require a prediction of the
traffic matrix out to the level of the planning horizon for a
task. Any forecast depends on the time scale involved and the
underlying model used. At short time scales, say minutes, stationarity
may be a reasonable approximation, and there are therefore many time
series approaches the problem. On time scales of hours to days to
weeks, the cyclostationary nature of the data must be included. The
temporal models presented earlier can provide such predictions.  For
instance, with the model \eqref{eq:temporal_model}, we can estimate
the average traffic at some time in the future by simply extrapolating
the mean. Longer-term prediction often focusses purely on the
large-scale trend $L(t)$ (see \autoref{ssec:temporal_modelling}), 
often captured using a simple growth model (linear 
or exponential) and regression. In all cases,
historical data is needed, usually several times as long as the
prediction interval.

In addition, whenever performing prediction we should provide an 
estimate of variances, or confidence intervals, though this component of the
problem has not been well-studied in the specific context of traffic
matrices.

% Current routing algorithms require some knowledge of the traffic
% matrix to determine the paths of the traffic flows in the network.
% Typically, link state routing protocols such as Open Shortest Path
% First (OSPF) or Intermediate System-Intermediate System (IS-IS) are
% used for intra-domain routing, while the Border Gateway Protocol (BGP)
% is utilised for inter-domain routing. Regardless, these algorithms
% depend on the network topology and the traffic matrix. For example,
% with BGP, the operator needs to know the best egress point to route
% the traffic out of the network. By knowing both the set of ingress
% nodes and their traffic demands, as well as the topology, the best BGP
% configuration may be determined. Tests of new routing protocols on
% network topologies also require knowledge of the traffic
% matrix. Furthermore, these protocols can be tested on artificially
% synthesised traffic matrices based on a current model of the traffic.

% In day-to-day tasks, the network operator requires a picture of what
% is happening in the network. If a link becomes congested, then the
% traffic matrix at the link level helps a network operator identify the
% blockage and take necessary steps to correct the situation. In this
% situation, routing information also plays vital role, as weight
% changes are required to re-route a portion of the traffic to less
% congested links. Inference algorithms play an important role here, as
% inferred traffic matrices from SNMP measurements may be used for these
% tasks. An empirical study \cite{Roughan03TETM} proved that despite the
% inaccuracies of the recovered traffic matrices, these estimated
% traffic matrices still yield remarkably good solutions for traffic
% engineering tasks. Since mosts traffic engineering tasks revolve
% around large aggregate flows, as long as inaccuracies of the large
% flows are kept controlled and minimal, these estimated traffic
% matrices are good enough for network optimisation.

\subsection{Reliability Analysis}

Traffic matrices may also be used to conduct reliability analyses,
where the effect (on traffic) or network failures is considered. A
basic task in most network design is to create redundant paths to
carry traffic in case of failure, but if high reliability is required,
then an operator should also ensure that there is sufficient capacity
in the network to carry this traffic along its alternate paths.  For
more details of this task see \cite{Roughan10Robust}.

\subsection{Anomaly Detection}

In reality, not all traffic of a network is
legitimate. Various attacks may be launched: DDoS attacks \cite{Patrikakis04DDoS}
or worm outbreaks such as the Nimda worm \cite{Nimda}. Non-malicious, but equally
violent spikes in traffic may be caused by a flash crowd or
implementation bugs. We call these surges \textit{anomalies}, and if they catch
a network operator by surprise they can congest networks, causing
untold damage to daily network activities. Other types of anomalies
may cause drops in traffic, again resulting in performance problems.

All these anomalies may be rare, but the potential damage can be
tremendous. It is for these reasons network operators strife to detect
anomalies, with the hope of protecting their networks from these
harmful effects.

Although network equipment vendors do provide some form of fast
detection and diagnosis mechanisms, these features are generally not
adequate for the problems listed above. Consequently, methods were
developed to counter anomalies. One approach is to use detailed
(packet level) traces and signature-based detection to detect known
attacks, but this does not help if the attack is unknown (in advance)
or if the necessary measurements are not available. Other techniques
infer statistical anomalies from the traffic flow data or SNMP
measurements. Traffic matrices play an important role in this respect,
since these matrices record traffic volumes across a whole network.

The basic principle of anomaly detection is to define a \emph{baseline
  operating condition} of the network, by establishing normal
conditions of the traffic. The baseline could be from a model, say a
gravity model, for example. There are many approaches, such as
entropy-based methods \cite{Brauckhoff09Anomaly,Gu05Entropy}, as
network anomaly detection itself is a vast topic, but here, the focus
is on the direct use of traffic matrices for anomaly detection.

Deviations from the baseline predictions of the traffic are quantified
with a chosen norm, and one is flagged as an anomaly if it exceeds a
predefined threshold. There are two sources of error: \emph{false
  positives}, when normal traffic is flagged as an anomaly, and
\emph{false negatives}, when a detector does not flag an anomaly.  The
latter type of error, where we miss a potential anomaly, are
apparently a more serious problem (given the serious nature of
anomalies). However, if too many false positives occur, then operators
can be overwhelmed, and will typically ignore the alarm system. The
false positive problem is exacerbated if the number of tests is large,
and in traffic matrix analysis (where we might conduct one test per
traffic matrix element, per time interval) that number can be very
large, requiring a very low false positive rate.

We consider the tradeoff between the two in an ROC (Receiver Operating
Characteristic) curve which shows the two types of errors plotted
against each other as a function of the chosen threshold (or other
suitable tuning parameter). However, proper assessment of an approach
requires ground-truth data, which is, by the nature of anomalies, hard
to obtain in the volumes required.

% http://en.wikipedia.org/wiki/Receiver_operating_characteristic

Models themselves can be modified to account for anomalous traffic. In
\autoref{ssec:temporal_modelling}, the model \eqref{eq:temporal_model}
itself has a term to account for sudden spikes in traffic, which was
demonstrated empirically to be useful in detecting large shifts in
traffic. Low rank models \cite{Zhang09TMCS} were shown to be highly
effective in detecting anomalies as well.

Many anomaly detection proposals may be broadly classified as methods
to preprocess measurement data via a linear transformation, in order
to separate normal traffic from anomalous traffic. This was observed
in \cite{Zhang05Anomography}.  Their \emph{anomography} (a portmanteau
of ``anomaly'' and ``tomography'') framework is easy to understand and
is aimed at providing a framework for discussing these types of
techniques. It proceeds as follows: start by assuming the routing
matrix is static in the entire duration of the measurements. Given a
series of SNMP measurements, $\bY = \bA \cX$, a new inference problem
is obtained by multiplying $\bY$ with a linear transform $\bT$ to
obtain $ \tilde \bY = \bA \tilde \cX$, which are the anomalous link
loads. Whether the focus is on spatial or temporal anomalies depends
on whether $\bY$ is pre- or post-multiplied with $\bT$:
\begin{enumerate}
\item \textit{spatial anomography}: pre-multiplication, \ie~$\tilde
  \bY = \bT \bY$, uses the spatial relationships between traffic at
  particular points in time to find traffic that is unusual with
  respect to other flows at the same time; and 
\item \textit{temporal anomography}: post-multiplication, \ie~$\tilde
  \bY =  \bY \bT$; uses the relationships between traffic at different
  times to determine if traffic is unusual for its point in time.
\end{enumerate}
The two have been combined to create spatio-temporal anomaly
detection~\cite{Zhang09TMCS}, though the full details of this go
beyond the scope of this chapter.

The above assumes the routing matrix is static over the series of
measurements. The models themselves have to be modified to account for
possible route changes. Some models are less amenable to modification,
requiring a large number of constraints that scale with the number of
measurements \cite{Zhang05Anomography}, which makes them undesirable
for practitioners.

Anomaly detection employing SNMP data took off with a series of papers
\cite{Lakhina04Anomaly,Lakhina04Diagnosing,Lakhina05Mining,Li06Anomaly},
where the low intrinsic spatial dimensionality of traffic matrices was
exploited via a PCA-based anomaly detector. In the spatial PCA-based
method, the principal components and axes of $\bY$ are computed from
its columns and ordered from most to least significant component, to
obtain a subspace $\bP = \lbrack\bv_1\, \bv_2\, \cdots
\bv_m\rbrack$. The traffic space is then divided to a \emph{normal
  subspace} and an \emph{anomalous subspace}. The traffic time series
is then projected on each principal axes, starting from $\bv_1$ and so
forth, and the projection magnitude is compared to a simple hard
threshold of three standard deviations from the mean. Once there
exists a projection exceeding this threshold, say at some $\bv_{K}$,
this component and subsequent components are classified as belonging
the anomalous subspace $\bP_A = \lbrack \bv_K\,\bv_{K+1}\,
\cdots\,\bv_m\rbrack$. The anomalous traffic is identified by
projecting the time series onto the anomalous subspace and projecting
the traffic back to obtain $\tilde \bY$.

Lakhina \textit{et al.}'s spatial PCA method fits in the framework since the last step
of extracting the anomalous traffic involves the projection $\tilde
\bY = (\bP_A^\T \bP_A) \bY$ so the linear transformation is $\bT =
(\bP_A^\T \bP_A)$.  However, its shortcomings have been the subject of
scrutiny. Implementation of spatial PCA to network traffic is likely
to be ineffective due to several drawbacks
\cite{Ringberg07PCA,Zhang05Anomography}. Spatial PCA can be
contaminated with a large anomaly\footnote{Though in fact this is a
  problem in general for anomaly detection, and has not received the
  attention it deserves.}, rendering it unable to detect the
anomaly. In fact, PCA has been known to flag an entire measurement
interval although there is only one anomaly present
\cite{Zhang05Anomography}. Additionally, PCA is very sensitive to the
underlying data: adequate measurements are required and there must be
a sufficient level of traffic aggregation before underlying trends can
be detected by PCA. It is not robust enough in practice, requiring
much fine tuning. Finally, there is a high computational cost involved
in computing the principal components of a traffic matrix.

Other alternatives exist: wavelet transformations
\cite{Barford02Anomaly,Papagiannaki05Long}, Fourier transformations, autoregressive
integrated moving average or ARIMA \cite{Zhang05Anomography}, and
temporal PCA \cite{Zhang05Anomography}, where PCA is applied to the
rows of $\bY$ (the temporal dimension) instead. In all these
techniques, the baseline traffic flows are assumed to follow the
prescribed model. In the Fourier model, baseline traffic is assumed to
be composed of low frequencies. High frequencies may potentially
indicated the presence of anomalies since these correspond to sudden
changes in the traffic. Thus, the transformation filters out low
frequencies and examines the remaining high frequencies to determine
if any of these frequencies exceed a predetermined threshold. A
similar rationale holds for the wavelet transform model. The ARIMA
model \cite{Brockwell02TimeSeries} is very well-known in time series
analysis, providing flexibility in the choice of parameters. The model
generalises popular models such as a model with a built-in Holt-Winters
smoothing, the random walk model and exponentially weighted moving 
average models. It also allows memory and long range dependency 
\cite{Veitch97Wavelet} to be built into the model via fractional 
ARIMA, as evidenced and used to great effect in 
\cite{Scherrer07Memory}\footnote{See
  \cite{Brockwell02TimeSeries} for a good introduction to time series
  analysis.}.

After $\tilde \bY$ is obtained, the anomalous traffic $\tilde \cX$ has
to be recovered. The choice of a particular inference algorithm would
depend on the model. The spatial PCA method uses a greedy algorithm to
find the largest anomaly in each time bin
\cite{Lakhina04Anomaly}. Other methods include the use of $\ell_1$
regularisation, inspired by compressive sensing
\cite{CandesTaoCS06,DonohoCompressed06}, which was shown, when coupled
with the ARIMA model, outperforms other methods, including PCA and
wavelet-based anomaly detection \cite{Zhang05Anomography}.

In short, the model of the traffic matrix matters in
anomaly detection. It serves as a baseline. However, it also needs to
consistently allow for anomalies.  One problem with approaches such as
PCA is the models implied by the approach are often left unstated
(implicit) and do not allow the anomalies to be separated as part of
estimation (thus they can pollute the estimation process). Good
techniques, going on into the future, need to be able to perform such
separation consistently.

\subsection{Traffic Matrix Synthesis}

Synthesis of the traffic matrix is an important area, motivated by the
lack of real world traffic matrices available, due to the proprietary
nature of most traffic data. Publicly available data are often obtained
from networks operated and maintained by research institutions and
universities, such as G\'{E}ANT \cite{GEANT} and Abilene
\cite{Abilene}, which are limited in scope and is certainly biased
towards research and educational networks.
Thus, network operations stand to gain much from artificially
synthesised traffic matrices, via a combination of good models.

Synthesis is not demanding in some ways. Traffic matrices are usually
relatively small compared to other types of traffic data when
measured at a reasonable level of aggregation and time scale. However,
in other ways these matrices are quite challenging. For instance:
\begin{enumerate}

\item we have few sets of traffic matrix data, and even fewer that are
  public, and somehow need to use these to estimate properties of
  these complex, high-dimensional objects;

\item there is a real relationship between topology and traffic
  (although we would like a traffic matrix to be invariant to the
  topology, there are clear cases where, particularly IE matrices are
  not);

\item traffic matrices come in a wide variety of types (at different
  levels of aggregation, for particular applications and so on) and it
  is unlikely that one model fits all; and

\item there are a number of conflicting goals in synthesis, \eg~to
  generate variability, but well ``matched'' to real traffic
  matrices. 
\end{enumerate}

The major use of synthesis is in simulation. Artificial traffic matrices may 
be used in capacity planning to stress
test network topologies to see if they stand up to heavy loads without
ending up with congestion. Monte Carlo-type simulations may be used to
produce estimates of the behaviour of networks. Synthesised traffic
matrices provide an avenue to explore the limitations of a protocol in
a controlled environment before running it on an actual network. A
good model simplifies these simulations, as parameters of the model
can be tuned to generate a variety of scenarios for testing protocols.

In many cases a traffic matrix is enough for a simulation, but in others, 
we need to translate this into packets (or at least connections). The analogue in
transportation modelling is often called a {\em micro-simulation}
model \cite{Algers97Micro,Hoogendoorn01Vehicular}. 
Here, the problem becomes one of taking a {\em demand} matrix
(remember, most of the work here is related to traffic, not demand), and
translating this into carried load. We know how to do that (using
simulation tools such as \verb|ns|) but doing it efficiently is
difficult. One paper \cite{sommers11:_effic} starts to tackle this
problem, but as in the work of transportation modelling, there is
considerable scope for advanced scalable micro-simulation of traffic.

Another use for synthetic traffic matrices is in the further task of
synthetic topology generation, but we shall leave discussion of this
topic to Chapter 7 of this book.

Unfortunately, there is a dearth of work on traffic matrix synthesis,
apart from \cite{Nucci05TMSynth,Roughan05GravSynth} and a brief
mention regarding synthesis of matrices from the independent
connections model
\cite{Erramilli06IndepConn,Erramilli06IndepConnTech}. The problem of
synthesising traffic matrices is the inverse of the inference
problem. In synthesis, the topology of the network matters, as the
generated entries of the artificial traffic matrix must not exceed the
link capacity it is mapped to. Some models, such as the gravity model,
automatically satisfy bandwidth constraints naturally
\cite{Roughan05GravSynth}. However, if the generated entries do not
conform to these constraints, there are algorithms to solve these
problems \cite{Nucci05TMSynth}, albeit with added computational 
complexity.

Computational complexity of the model
is the other important issue, dependent on the number of
parameters. Hence, there is an inherent tradeoff between the
descriptive power of the model and the ease of synthesising traffic
matrices. A guideline is to preferably choose the model with as little
parameters as possible but enough proven descriptive power (focussed on
the traffic aspect the practitioner intends to capture) measured
via an information criterion, such as the AIC \cite{Akaike74AIC}.

We here describe the simple approach of \cite{Roughan05GravSynth}
motived by works such as \cite{FRT02}, in order to provide a starting
point for future work on such synthesis.  We start by taking $\bx^{\rm
  in}$ and $\bx^{\rm out}$ to be vectors of $N$ i.i.d.~exponential 
random variables with mean one.
The traffic matrix is then generated using \autoref{eq:network_grav_mtx}. We can
then adjust the total traffic to match the desired total by simple
scaling.  This method is extremely simple (an exponential distribution
has only one parameter to estimate), and we need generate only $2N$
random variables. Yet it matches observed statistics for both Abilene
and G\'{E}ANT data extremely well \cite{Alderson06Topology}.

Synthesis of traffic matrices is an open area, and is important to further 
explore due to the benefits of using traffic matrices in traffic planning and
engineering. There is considerable hope that progress can be made in terms
of generating synthetic matrices, both for green-fields network design
\cite{Kowalski95TeleModel}, and for simulation in general.

### 2013-07-22 Compiled/chapter8/future.tex
all files compile correctly

\clearpage
\section{Future Directions}
\label{sec:future}

There are some interesting tasks left for traffic matrix
research. The various algorithms and techniques described
here could be improved, though in many cases the improvements may be
relatively incremental given the success of existing approaches. More
interest may be found in extending the ideas and techniques used here
to new domains, and to evolving Internet traffic.

There are a few obvious cases (and no doubt many less obvious cases
that we have not thought of), for instance: multicast traffic has not,
to our knowledge, been studied in this way. Multicast is interesting
because it violates the {\em traffic conservation} assumption that
lies underneath many techniques for estimation and modelling of
traffic matrices. We could imagine modelling it by considering the
``flow'' to be the traffic on a multicast group, from say, one source,
to a set of destinations, and then stacking a vector with these. The
routing matrices now include elements for every link used (no longer
following a single path). The traffic ``matrix'' could then be a
column vector of the traffic on each of these flows. So the idea of
multicast traffic can fit into the structure we have talked about
here, but appropriate models for performing tasks such as inference
do not seem to exist.

It would also be very interesting to understand the way that CDNs are
affecting network traffic. A step in this direction, although not directly
on traffic, but more on the discovery of the content hosts is \cite{Ager11CDN}.
A CDN's typical goal is to bring content
closer to the user, thereby reducing network traffic. However, that
explicitly violates the ``friction free'' assumption in most gravity
models, and introduces distance as something to be modelled.

That leads naturally to the consideration of global traffic. Almost all
studies of traffic have concentrated on a single network no larger
than the national scale. That may still be very large -- for example,
several studies looked at Tier 1 providers in the USA, which for some
time dominated Internet traffic. However, although large, it was still
relatively homogeneous traffic between people speaking much the same
language(s) from place to place in the network. When we consider the
Internet globally, we may see that there are language or cultural
clusters where large groupings of traffic are focussed.

On a large scale, time zones also play a significant role. Traffic
patterns show strong cyclic behaviour based on user activity, but such
activity is strongly dependent on the local time zone. If traffic is
flowing from user to user, then this can result in strong apparent
locality effects, simply because people in the same time zone are more
likely to be awake at the same time \cite{gerber:03}.

While language and cultural focussing may be geographic in nature, it
might also be considered per network, and that leads to another topic
of some interest. Very few papers have tried to consider inter-AS
(also known as inter-domain) traffic in any detail. Exceptions are
Chang~\etal\cite{Chang05EmpiricalAS} (which presents suggestions for
estimating traffic based on models derived from business models and resulting
usage); Bharti~\etal\cite{Bharti10Invisible} (which considered
inference of hidden elements of this matrix using a subset of data),
Feldman~\etal\cite{Feldmann04Web} (which aimed to estimate a global
traffic matrix, but only in the limited domain of WWW traffic), and
Labovitz~\etal\cite{labovitz10:_inter_inter_domain_traff} (which
looked at inter-domain traffic from 110 network operators over a
two-year period, though not in the form of a matrix). Study of the
Internet's global traffic matrix is made difficult by the sheer scale
of the project: Labovitz~\etal studied 110 network operators over a
two-year period\footnote{To put this in context, there are tens of
  thousands of ASes in the Internet.} and to do so, collected over 200
Exabytes of data. Many network operators do not collect or store data
of the type required for such a study, and many more regard it as
proprietary or covered by privacy legislation with provisions such
that no researcher is ever likely to see it. So we can see that study
of the inter-domain matrix is likely to be a long-term, and rather
challenging project.

In addition, we know that the {\em traffic profile} (or mix of
applications) has changed fairly rapidly over time. It is likely this
trend will continue, and there are bound to be effects on traffic
patterns as a result. Peer-2-peer traffic significantly altered
traffic patterns when it appeared because it was more symmetric than
traditional (at the time) WWW traffic. However, in addition,
peer-2-peer applications have the potential to exploit locality
information to download from sources closer to the destination. This
could potentially change the ``no friction'' assumption in much the
same way that CDNs can, though in the early days it did not appear to
be the case \cite{gerber:03}. An example traffic matrix (drawn from
\cite{gerber:03}) showing normalised\footnote{The elements have been
  normalised by dividing each row by the row-sum, so that each element
  actually represents the probability that a packet enters the network
  at a given region $i$ will depart the network at region $j$.}
traffic between regions in a cable-network operator is given in
\autoref{tab:cable_tm}. The major deviation, in this data, from a pure
gravity model seemed to be time-zone differences.

\begin{table}[!h]
  \centering
  \begin{tabular}{r|rrrrrrrr}
    From/To & R1 & R2 & R3 & R4 & R5 & R6 & R7 & R8 \\
    \hline
    R1 & -     & 0.180 & 0.140 & 0.126 & 0.174 & 0.128 & 0.124 & 0.127 \\
    R2 & 0.172 & -     & 0.141 & 0.126 & 0.190 & 0.132 & 0.118 & 0.120 \\
    R3 & 0.132 & 0.120 & -     & 0.189 & 0.135 & 0.145 & 0.139 & 0.140 \\
    R4 & 0.107 & 0.111 & 0.182 & -     & 0.124 & 0.163 & 0.155 & 0.158 \\
    R5 & 0.161 & 0.180 & 0.136 & 0.132 & -     & 0.135 & 0.127 & 0.129 \\
    R6 & 0.107 & 0.108 & 0.145 & 0.155 & 0.125 & -     & 0.187 & 0.173 \\
    R7 & 0.107 & 0.106 & 0.137 & 0.157 & 0.127 & 0.182 & -     & 0.184 \\
    R8 & 0.109 & 0.111 & 0.127 & 0.161 & 0.128 & 0.178 & 0.185 & - \\
  \end{tabular}
  \caption{Normalised inter-regional traffic matrix from \cite{gerber:03}.}
  \label{tab:cable_tm}
\end{table}

% T = [
%  [ 0     , 0.180 , 0.140 , 0.126 , 0.174 , 0.128 , 0.124 , 0.127 ]
%  [ 0.172 , 0     , 0.141 , 0.126 , 0.190 , 0.132 , 0.118 , 0.120 ]
%  [ 0.132 , 0.120 , 0     , 0.189 , 0.135 , 0.145 , 0.139 , 0.140 ]
%  [ 0.107 , 0.111 , 0.182 , 0     , 0.124 , 0.163 , 0.155 , 0.158 ]
%  [ 0.161 , 0.180 , 0.136 , 0.132 , 0     , 0.135 , 0.127 , 0.129 ]
%  [ 0.107 , 0.108 , 0.145 , 0.155 , 0.125 , 0     , 0.187 , 0.173 ]
%  [ 0.107 , 0.106 , 0.137 , 0.157 , 0.127 , 0.182 , 0     , 0.184 ]
%  [ 0.109 , 0.111 , 0.127 , 0.161 , 0.128 , 0.178 , 0.185 , 0 ]
% ]

New traffic classes may change traffic matrices in the future, and
modelling these will be interesting. On the other side, applications
such as anomaly detection are likely to remain interesting due to
their immediate benefits to operators, but the most overlooked task is
traffic matrix synthesis. As mentioned in the previous section, the
lack of real world traffic matrix datasets motivates the use of
artificial traffic matrices to provide some degree of approximation in
network traffic planning, provisioning and engineering tasks.  It is
an important area to concentrate on, considering its usefulness to
network operators.

%Synthesis means generating artificial traffic matrices, typically for
%use in simulations. There are only, to our knowledge, two papers
%\cite{Nucci05TMSynth,Roughan05GravSynth} on synthesising Internet
%traffic matrices, but there are already quite a few where synthetic
%traffic matrices were used, and this demand for such matrices will
%continue.
%
%Synthesis is not demanding in some ways. Traffic matrices are usually
%relatively small (compared to other types of traffic data), when
%measured at a reasonable level of aggregation and time scale. However,
%in other ways these matrices are quite challenging. For instance:
%\begin{itemize}
%
%\item we have few sets of traffic matrix data, and even fewer that are
%  public, and somehow need to use these to estimate properties of
%  these complex, high-dimensional objects;
%
%\item there is a real relationship between topology and traffic
%  (although we would like a traffic matrix to be invariant to the
%  topology, there are clear cases where, particularly IE matrices are
%  not);
%
%\item traffic matrices come in a wide variety of types (at different
%  levels of aggregation, for particular applications and so on) and it
%  is unlikely that one model fits all; and 
%
%\item there are a number of conflicting goals in synthesis, \eg~to
%  generate variability, but well ``matched'' to real traffic
%  matrices. 
%
%\end{itemize}
%However, there is considerable hope that progress can be made in terms
%of generating synthetic matrices, both for green-fields network design
%\cite{Kowalski95TeleModel}, and for simulation in general. 
%
%The major use of synthesis is in simulation. In many cases a traffic
%matrix is enough for a simulation, but in others, we need to translate
%this into packets (or at least connections). The analogue in
%transportation modelling is often called a {\em micro-simulation}
%model. Here, the problem becomes one of taking a {\em demand} matrix
%(remember, most of the work here is related to traffic, not demand), and
%translating this into carried load. We know how to do that (using
%simulation tools such as \verb|ns|) but doing it efficiently is
%difficult. One paper \cite{sommers11:_effic} starts to tackle this
%problem, but as in the work of transportation modelling, there is
%considerable scope for advanced scalable micro-simulation of traffic.
%
%Another use for synthetic traffic matrices is in the further task of
%synthetic topology generation, but we shall leave discussion of this
%topic to another section of this book.

### 2013-07-22 Compiled/chapter8/introduction.tex
all files compile correctly

\section{Introduction}
\label{sec:introduction}

In the era of the telephone voice traffic dominated physical
telecommunication lines. With the birth of the Internet, and its
subsequent adoption as a key means of communication, data traffic has
taken over (though some of this data traffic is actually re-badged
voice traffic using Voice over IP).  With the advent of new
applications such as video streaming, and the rapid growth of data
traffic from mobile devices, we are witnessing a global data
explosion.

Given the ever increasing importance of the Internet, knowledge of its
traffic has become increasingly critical.  The Internet, however, is
just an all-encompassing term to describe the vast global collection
of networks, and so it largely falls on individual network providers
to determine their own traffic.  This knowledge is vital for continued
operations because it allows network operators to perform important
tasks such as providing enough network capacity to carry the current
traffic, as well as to predict and prepare for future trends. Traffic
data is also important in network maintenance, which is necessary if
services and content are to be provided to customers with minimal
interruption.

The focus of this chapter is on the {\em traffic matrix}, which, in a
nutshell, is an abstract representation of the traffic volume flowing
between sets of source and destination pairs. Each element in the
matrix denotes the amount of traffic between a source and destination
pair.  There are many variants: depending on the network layer under
study, sources and destinations could be routers or even whole
networks. And ``Amount'' is generally measured in the number of bytes
or packets, but could refer to other quantities such as connections.

Traffic matrices, as will be clearer below, are utilised for a
variety of network engineering goals, such as prediction of future
traffic trends, network optimisation, protocol design and anomaly
detection.  Considering the widespread use of these matrices, the first
objective of this chapter is to provide an entry level for graduate
students and new researchers to current research on traffic
matrices. To that end, the material is organised in a tutorial-like
fashion, so as to ensure key concepts can be easily understood.

% In
% addition, it is hoped this chapter would be a definitive reference for
% researchers and practitioners directly involved in using traffic
% matrices for network engineering applications.

\subsection{Motivation}

Why study Internet traffic matrices? Simply because their implications
for network operators are vast. If the traffic matrix of a network is
exactly known, then, armed with topology and routing information of
the network, the operator knows \emph{exactly} what is going on in the
network, rendering network management tasks relatively easy. If the
traffic matrix is completely unknown, then a network operator is
blind. Subtle faults may plague their network, substantially reducing
performance. Congestion may be rife, or sudden shifts in traffic may
cause transient traffic losses.

The issues are becoming more important.  The dominant philosophy in
the early days of the Internet was best effort delivery. Most
applications did not have high quality of service (QoS) requirements,
and had some tolerance to packet drops and link failures. In recent
years, however, the landscape of the Internet is changing quickly with
the introduction of streaming content such as video, high definition
television and Voice over IP (VoIP), which have more stringent QoS
requirements. For example, excessive packet drops and delays would
produce highly noticeable artefacts in streamed video. These changes
are driven primarily by user demands, with the introduction of new
applications, such as online social networking, entertainment services
and online multi-player gaming. Furthermore, with the increasing
computational power of mobile devices and increasing wireless speeds,
it is evident that a significant portion of future traffic will be
generated by these devices. These trends are making measurements more
and more critical.

For most operators, the state of their measurements is somewhere
between extremes. Most operators have some traffic data (if only link
counts). However, it is rare (in our experience) to find a network
operator who knows everything about their traffic. As such, one of the
tasks we shall consider here is how to measure these
matrices\footnote{This chapter is not really a primer on measurement
  tools as such, so much as the principles that underlie those
  tools. We will not tell the reader how to set up NetFlow on the reader's particular
  router, but instead we will aim to inform the reader what could be achieved
  with this type of flow-level measurements.}, but also how to obtain
estimates of traffic matrices when presented with incomplete data. The
traffic matrix {\em inference} or {\em completion} or {\em recovery}
problem is one of the major areas of research into these interesting
creatures, and it is intricately related to modelling the matrices,
both as measurements supply data to populate the models, and because
models are used to perform the inference.  Even those operators with
extensive measurements and exact knowledge of today's traffic matrix
may be interested in methods to predict their matrices for use in
future planning, and this can be seen as another form of matrix
completion.

Traffic matrices have many uses, apart from the simple fact that this
type of business intelligence is critical to understanding your
network. The more direct uses include network optimisation, anomaly
detection and protocol design.

There are three common optimisation problems on networks.  Capacity
planning is needed to ensure there is adequate bandwidth for users in
the present and future, but at minimal cost. There are two types of
network planning: evolutionary planning and green fields planning; see
\autoref{sec:applications} for details. Traffic engineering tasks
include day-to-day maintenance of the network as well as predicting
growth trends and anticipating traffic demands \cite{Mitra05StochasticTE,Buriol03GAOSPF,Murphy02TE,Fortz02OSPF,Fortz03OSPF,Nucci07IGP,Uhlig04Implications,Feamster03BGPTE,Rexford06Route}. Routing involves
organising traffic flow in the network, as well as modelling routing of 
large networks \cite{Quoitin05CBGP}. This includes functions such
as finding the shortest paths for flows but also, importantly, load
balancing to ensure links remain uncongested. In all these cases, the
traffic matrix is a key input to perform the tasks effectively and
efficiently.

Traffic matrices can also be used to detect sudden shifts in traffic
due to anomalies. Anomalies include sudden unexpected events, such as
network failures, or more malicious events, such as the September 11
World Trade Centre attack, worm infections and distributed denial of
service (DDoS) attacks \cite{Roughan02BRvariable}. Regardless, these
anomalies need to be detected so as to develop appropriate measures
against possible threats to the network.

Traffic matrices may also be used to conduct reliability analyses,
where the effect (on traffic) of network failures is considered. A
basic task in most network design is to create redundant paths to
carry traffic in case of failure, but if high reliability is required,
then an operator should also ensure that there is sufficient capacity
in the network to carry this traffic along its alternate paths.

Further, the performance of many network protocols depends on the
traffic they carry, and the cross-traffic which forms their
background. Design of new protocols therefore requires realistic
measurements, or models of such traffic.  Models can be used to test
protocols on artificially synthesised traffic. In this way, the
limitations of a protocol may be understood in a controlled
environment before running it on an actual network.

These issues will be examined in-depth in \autoref{sec:applications},
where algorithms utilising traffic matrices to perform these tasks
will be discussed.

%The
%information obtained by the operator is also useful in addressing customer queries on their bandwidth usage, how their
%traffic is routed (as well as related issues on privacy) or the overall quality of the service. Moreover, the network
%operator can use the same information for accounting purposes when billing customers, or modeling and predicting customer
%demands.

%Finally, the network operator requires the traffic matrix to rate the quality of the network. Generally, there is a 
%minimum standard to the network quality in terms of traffic loss, delay and jitter, as outlined in Service Level Agreement
%(SLA). Although more sophisticated SLAs may appear in the future, the traffic volume from a source to its destination is
%ultimately required by the network operator to measure quality.

\subsection{A primer on modelling}

Motivated by a lack of understanding of Internet traffic, and as a
response to the recent shifting landscape of traffic demands, various
traffic models have been developed over the last decade or so. There
are hundreds of papers describing data traffic modelling, however,
here we shall focus on one group of such models that are particularly
useful to operators: models for traffic matrices. There is no single
model that captures all observed properties of once and future traffic
matrices, and so here we examine the issues that go into deciding on
appropriate models for particular tasks.

% On one hand, the rapidly changing state of the Internet strongly
% motivates the collection of data and development of models of the
% traffic matrices of the networks that serve as the lynchpins of the
% Internet.

The ultimate goal is to use these models in various networking
tasks, and the ``essential'' qualities that the model must capture
depend critically on the task. For instance, traffic properties are
highly dependent on time scale: in very short time scales (seconds), 
the traffic volume distributions have been shown to exhibit
complex statistical behaviour such as high variability, long-range
dependence, and multi-fractal behaviour. On long time scales (hours to
days to weeks) traffic exhibits strong periodicities: there are
distinct diurnal and weekly cycles with traffic peaks occurring around
mid-day or in the evening and troughs in the early morning.  This
pattern correlates to user behaviour, where they access the Internet
during the day for work or school and sleep in the night. There is
also an observable long-term growth trend underlying the traffic
volume when measured over years, corresponding to increasing global
traffic demand. However, most traffic matrices are measured at some
intermediate time scale, often at 5 to 15 minute intervals.

The time scale is not just a property of the measurements though. The
tasks we wish to perform usually have an associated time scale as
well. In capacity planning, we are often concerned with a ``busy
hour'' -- certainly peak measures over some intermediate period are
important. However, anomaly detection needs to act at fine time
scales: minutes, or perhaps even seconds.

We also have to consider the {\em planning horizon} of a
task. Capacity planning may be aimed at determining the correct
network design six months in advance or more, whereas traffic
engineering is sometimes conducted on scales as short as a day. The
planning horizon determines how far in advance we must predict traffic
matrices.

% to predict future
% traffic matrices, and their impact on a network topology. The other
% significant gain is the ability to simulate traffic matrices for the
% purposes of testing traffic predictions, anomaly detection and routing
% weight changes. Considering the value of knowing the traffic matrix of
% a network, clearly, having a model that sufficiently describes the
% dynamics of the matrix, while balancing the number of parameters it
% has to enable tractable analysis and simulation, will be an important
% traffic engineering tool.

We can start to see that modelling traffic matrices is a challenging
task. Moreover, there are complexities layered upon complexities. The
Internet is designed in terms of layers, with different protocols
overlaid atop each other. Such a paradigm was adopted to ensure that
each layer works independently, in theory, of each other, although
there are some overlaps between these layers in practice.  The basic
properties of traffic matrices will depend on the network level, and
traffic can be measured between logical or physical
source/destinations, or at different levels of aggregation.

Measurements of networks clearly serve as the foundation in any model
development. The caveat, however, is that the measurements themselves
are subject to errors and inconsistency which may lead to an incorrect
model. Moreover, several hypotheses may fit a particular observation
of the network, leading to several possible models explaining the same
observation. To argue for the use of one model over another requires
additional knowledge from new data or from domain knowledge. And when
new information become available, there is a question of how to
incorporate it into the model. There are a variety of possible
approaches, all equally valid \cite{Alderson06Topology}.

To further compound the problem, underlying all Internet traffic is 
the fact that all traffic is driven by consumer demands. Unfortunately, 
human behaviour is inherently difficult to model let
alone understand. Furthermore, changing trends in network usage,
deployment of Content Distribution Networks (CDNs), and increasing mobile traffic
all have implications on traffic measurement. Although measurements could be
made extremely accurate, if one overlooks the costs involved, 
the observations themselves may be outdated very quickly. 
Therefore, a complicated model based on these measurements
may be accurate for today, but
fail in predicting traffic demands for the next few years or so, given
fluctuations in demand and unexpected changes in traffic patterns
\cite{Teixeira04PotatoSIG}.

%We can now see that a good model should capture the 
%essential characteristics of the underlying
%traffic, while being robust enough to cope with changes in traffic,
%for example, with regards to the time of the day, or the introduction
%of unexpected events, such as attacks on the network. Much can be
%gained from a good model, enabling the categorisation and analysis of
%the data in a systematic, and desirably, a simplified manner.

Furthermore, it is misleading to talk about a ``correct'' traffic
matrix model. As pointed out in \cite{Alderson06Topology}, just
because a model replicates some set of properties of the observed data
does not necessarily mean the model is ``correct''. At the least,
there is the dangerous possibility of over-fitting. After all, a
better fit is always achievable simply by adding more parameters to
the model.  Several information criteria address the over-fitting
problem, for instance the Akaike information criterion (AIC)
\cite{Akaike74AIC}, Bayesian information criterion (BIC)
\cite{Schwarz78BIC}, Minimum Message Length (MML) \cite{Wallace99MML}
or the Minimum Description Length (MDL) \cite{Rissanen83MDL}. While
these criteria are beyond the scope of our discussion, the basic
principle is to choose the simplest explanation (measured in some
information metric) amongst all competing explanations of the data.

% The issue highlights the
% difficulty of choosing the number of parameters in a model, as it
% requires several tradeoffs between simplicity and its connection to
% reality.

It is for these reasons models should be evaluated not just on their
accuracy in making predictions of particular statistics, but also
their simplicity, robustness and consistency in relation to the
realities of network operations. Model assumptions should ``make
sense'' to an operator as well as be empirically tested on various
datasets to understand their reliability and pitfalls, which is not to
say we cannot learn new ideas and principles from measurements. We just
need to keep in mind their scope of application, and the usefulness of
these principles in practice may be limited.  It is often preferable to have simple, robust models, in
preference to precise, but fragile ones.

At a fundamental level we need to accept that models are all
unrealistic in some way. A model describing the properties of a
smaller scale network, such as a Local Area Network (LAN), may be
unsuitable at the backbone network level. The underlying assumptions
of one may not hold in the other. Models are simplifications of the
glorious complexity that comprises humanity's primary means of
telecommunication. We must, instead, reread the adage, by George
E.~P.~Box: ``All models are wrong, but some are useful''.  Some models
have been more successfully used in real networks, and it is to these
that we shall devote the most time here. However, we shall endeavour to
cover the majority of simple models with the view that individuals
should use the best model {\em for their application} without fear or
favour.

No doubt, the murkiness and apparent self-contradictions of this
discussion have left our readers no wiser, as yet. Modelling is a
topic that could be discussed endlessly. It is our aim that through
consideration of the qualities of various traffic matrix models, we
shall not only inform about these particular models, but also bring
the reader to a new understanding of modelling to make these
issues a little less opaque.

%   The important principle is to know why they
% work, hopefully shedding light on the variability of traffic flow and
% properties.

\subsection{Chapter outline}

The theme of this chapter is to directly report on all key works in
the field. No attempt will be made to provide overt commentary on what
techniques or models are good or bad, as the objective of this chapter
is to present comprehensive summaries of existing work. It is
important to note that the techniques, algorithms and models presented
here were developed as tools to suit specific applications. Hence, it
is the onus of the practitioner to evaluate and decide which of these
are applicable to his or her situation. Whenever possible, the
strengths and weaknesses of the inference techniques and models are
discussed objectively, so as to minimise the problem of a practitioner
treating a particular tool as the silver hammer for all proverbial
nails.

The chapter is organised as follows. In \autoref{sec:tm}, an overview
is provided on traffic matrices: the basic definitions and some
illustrations to give a better handle on the topic.
\autoref{sec:measurements} discusses how data on traffic matrices are
collected in practice. \autoref{sec:models} discusses the various
models proposed in literature that aims to capture the statistical
properties of traffic matrices. \autoref{sec:applications} goes into a
more in-depth treatment of the applications of traffic matrices, in
particular, how traffic matrix models are used for inference, network
optimisation applications, anomaly detection and traffic matrix
synthesis. Not everything is known about these matrices, and
\autoref{sec:future} summarises some open questions 
and concludes by giving some thoughts on what the future holds for
traffic matrix research. The chapter is concluded in \autoref{sec:conclusion}.

### 2013-07-22 Compiled/chapter8/measurements.tex
all files compile correctly

\clearpage
\section{Measurements}
\label{sec:measurements}

In theory, it is possible to collect accurate data of Internet traffic
from a network. In reality, however, many issues confound such
measurements. Budget constraints driven by the cost of collection, or
the massive data storage facilities required due to the sheer amount
of data traversing a backbone network limit what can be achieved. And
even good measurement systems can suffer from errors and missing
data. Added to this, current operator practice rarely includes any
significant calibration of measurement apparatus, so often the degree
of accuracy of the measurements is unknown.

There are several well-known strategies for collecting traffic
measurements. A {\em packet trace} is a collection of packets headers
(perhaps with some payload) and timestamps.  A packet trace can be
collected through various means, for example, through hardware such as
a splitter placed strategically in optical fibre; adding a monitor
port on a router; or through software tools such as \texttt{tcpdump}
executed on several hosts in a shared network. An OD traffic matrix
can be constructed from such a trace by simple consideration of the IP
address in the packet headers (with the caveats mentioned earlier).

Such an approach is ideal in many respects: we have almost complete
information, and the matrices may be drawn at almost any time
resolution. However, collecting packet traces is expensive due to the
need for dedicated hardware, and the huge amount of storage required:
potentially, over a terabyte of data per hour on OC48 links (2.5 Gbps)
is needed. It is rare for any but the smallest network to be able to
completely instrument its network at this level of detail.

Fortunately, constructing a traffic matrix does not require such
detailed information. Perhaps the most common alternative is {\em
  flow-level aggregation} where packets are aggregated according to a
common flow key.  One popular definition of the key is the 5-tuple
comprised of the IP source and destination address, TCP/UDP source and
destination port numbers and protocol number. A series of packets
possessing a common key is called a \emph{flow}, and we maintain
simple statistics (byte and packet counters, and start and stop times)
for each flow. The aggregation of packets into flows reduces the
number of records needed to be stored by removing redundancies of data
from a packet trace. Flow-level collection is generally performed in
15 minute time bins\footnote{The issue of timing of flows is actually
  somewhat more complicated, but readers should refer to detailed
  descriptions of specific flow-capture protocols for information on
  their particular flow capture.} and is often an in-built function of
a router. The only additional infrastructure required is the Network
Management Station (NMS) and flow records themselves are usually
compressed by the router before being exported to the NMS. Despite
this reduction, flows arrive at a router at rapid rates and the
formation and storage of flow records at a router often burdens the
router's CPU.

To further reduce the number of flow records at a router, sampling
methods are employed. The most popular sampling method is packet
sampling, where incoming packets are sampled based on predetermined
sampling patterns, used, for instance by Cisco's NetFlow
\cite{NetFlow}. Such pseudo-random patterns have a similar effect to
independently picking incoming packets given sufficient mixing of
traffic.  The sampling rates can be adjusted depending on the capacity
of the incoming links with recommendations such as 1 in 256 packets
for OC192 links (10 Gbps). Higher capacity links require aggressive
data reduction, and so lower sampling rates are used.

Packet sampling reduces the number of flows significantly by omitting
many, but it is important to realise that although it may select
\emph{packets} in an unbiased way, it is not unbiased with respect to
\emph{flows}. Packet sampling has a strong bias towards long flows,
since it is more likely to have sampled packets from a long flow than
a short one. Furthermore, the sampled flow size is not the true size
of the flow and there are several works
\cite{Duffield05Sampled,Duffield05Smart} proposing methods to sample
and estimate the true size of a sampled flow. While the strong bias
may be a problem for some applications, there is usually no problem in
using sampled flows to form the IE traffic matrix.

In addition to packet sampling, we may also sample a set of flows, and
these sampled flows can then be used to create traffic matrices. The
resulting reduction in intermediate storage and processing can be
substantial, particularly if the sampling is done in a clever way
\cite{Duffield05Sampled,Duffield05Smart}.

It must be remembered that sampling is an inherently lossy process,
regardless of the underlying sampling method used.  The loss of
information translates into errors or noise in the data. The size of
these errors should, in best practice, be estimated for a given setup,
but most operators do not undertake such procedures due to the
difficulties involved in obtaining ground-truth data with which to
compare to the sampled data. In many cases it is simply assumed that
these errors, once the data is aggregated further into traffic
matrices, are negligible, but this assumption is rarely validated.

Furthermore, there is the problem of possible multiple counting of 
traffic flows from the aggregated records of flows. The problem 
arises if the aggregated flow records come from internal backbone
routers of the network, since a single flow may be recorded more than
once by several routers if it traverses several links. One way to
get around this relies on the placement of the measurement
points. One simply performs traffic measurements at the ingress points.
Specifically, the measurements are performed on the backbone routers 
connected to the access routers, or on dedicated devices placed at the 
links connecting the access routers to the backbone routers. 
In this way, the total incoming traffic of the network may be reliably
measured from the flow records. There is then no longer a problem of 
multiple counting of flows.

Trajectory sampling \cite{Duffield01Trajectory,Duffield04Unreliable} 
is another method. It exploits the pseudorandomness 
of a hash function to simulate random sampling of packets. However, if a flow 
was sampled at one router, it will be consistently sampled at every 
other router the flow traverses by the deterministic operation of hash 
functions. Since the tracking of flows is now feasible, it is then also 
straightforward to disambiguate their identities and eliminate multiple
counting. The downside is that a network operator must deploy 
trajectory sampling throughout the whole network, and configure
the hash function on each router each time before a measurement interval
begins.

A less costly alternative is easily obtainable link counts. A link
count, or link load measurement, gives the volume (in bytes or
packets) of traffic on a link during a particular time interval.  Link
counts are obtainable from measurement data by the Simple Network
Management Protocol (SNMP) \cite{SNMP}, defined as part of the IETF
standard and present on many Internet devices, including most
routers. SNMP data from a single router provides two measurements for
each interface, the incoming and outgoing byte counts. SNMP data is
obtained by an NMS by periodically polling requests through an
interface, typically UDP port 161. The polling period varies from 1
minute to several minutes, but the default seems to be 5 minutes. SNMP
data is highly susceptible to error, due to the following factors:
\begin{enumerate}
\item \textbf{missing link observations}: data is transmitted via
  unreliable UDP packets and there may be errors when copying the data
  into the observer's archive,

\item \textbf{incorrect link observations}: poor router vendor
  implementations causes inaccuracies, and

\item \textbf{sampling coarseness}: polling times are often inaccurate
  either due to poor NMS or SNMP agent implementations, high loads, or
  network delays.

\end{enumerate}
As with flow-level data, link count data should be calibrated, but
rarely is. There is only one experiment of which we are aware that
does so \cite{roughan10:_case_accur_snmp_measur}. The study showed
that in one network, SNMP errors were typically low (90\% of
measurements had an error of less than 1\%), but a small number of
measurements were very large, some as large as 100\%. This type of
heavy-tailed distribution causes problems for some estimation
approaches and should be considered in context.

Another drawback of SNMP data is that it only provides aggregate link
statistics, omitting details such as types of traffic on the link and
the traffic source and destination. Despite all these issues, SNMP
data is, at present, the easiest and perhaps most common way to obtain
large-scale traffic data.

The observed link counts provide some information about the traffic
matrix, but only in an indirect manner. Thus, the traffic matrix has
to be inferred through a process called \textit{tomography}. 
Network tomography was first introduced by Vardi
\cite{Vardi96Tomo}, with the inspiration taken from inference
techniques for medical tomography as both problems are similar in
nature. Vardi's work was subsequently expanded upon by Tebaldi and
West \cite{Tebaldi98Tomo} and Cao \etal\cite{Cao00Tomo}. Various
other works in network tomography measure other properties of the
network, such as link delays (see
\cite{tomo_CCLNY_2004,tomo_CHNY_2002} for an overview) via active
packet probing, but for traffic matrix estimation, we are only
concerned with the link count observations from SNMP data.

Mapping traffic to links requires topological and routing data in the
form of a \emph{routing matrix}. The routing matrix $\bA$ is defined
by
\begin{align*}
A_{i,r} = 
\begin{cases}
F_{i,r}, & \text{if traffic for $r$ traverses link $i$},\\
0, & \text{otherwise}.
\end{cases}
\end{align*}
where $F_{i,r} \in (0,1]$ is the fraction of traffic from
source-destination pair $r = (s,d)$ traversing link $i$. Fractional
values occur in cases when load balancing is used, but it has often
been assumed that $F_{i,r} = 1$, resulting in $A_{i,r} \in
\{0,1\}$. The size of the routing matrix depends on the network: with
$N$ nodes and $L$ links, the routing matrix has size $L \times N(N-1)$
(traffic from a node is assumed not rerouted to itself).

Information on the routing matrix can be obtained from several
different sources (router configuration files, traceroutes, or from
the routing protocols themselves), but the collection of such
information is not the topic of the chapter. The interested reader is
referred to \cite{Feldmann00Netscope,Feldmann01TMdemand}.

A common assumption is that the routing matrix remains stable during
the measurement interval, thus the temporal dependence is dropped,
\ie~$\bA(t) = \bA$ for all $t \in \cT$. However, changes in routing
may occur if there are link or router failures, necessitating traffic
reroutes. Generally, it is assumed the measurement interval is chosen
over a period of time (minutes to hours) when $\bA$ is stable enough
to be considered static, justified by observations in
\cite{Paxson97Routing}, however in at least one case it is proposed
that the changes be created, and
exploited~\cite{soule07:_estim_dynam_traff_matric_using} for traffic
matrix inference.

\begin{figure}[!thbp] 
  \begin{center}
    \hfill
    \begin{subfigure}[b]{\twoup}
      \centering
      \includegraphics[width=1\columnwidth]{tm_formulation.pdf}
      \caption{Link labels}
      \label{fig:tomo_a}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{\twoup}
      \centering
      \includegraphics[width=1\columnwidth]{tm_formulation_2.pdf}
      \caption{Traffic labels and routes}
      \label{fig:tomo_b}
    \end{subfigure}  
    \hspace*{3mm}

    \caption{A simplified network and traffic (the main
      simplifications are that we only consider a single router, and
      only consider unidirectional traffic, not bidirectional as in
      real networks).\label{fig:tomo}}
  \end{center}
\end{figure}

All the link counts may be grouped into an $L \times 1$ vector
$\by$. Then, based on link observations in one measurement interval,
the SNMP link counts may be expressed as 
\be 
  \by = \bA \bx,
  \label{eq:observation}
\ee 
where $\bx$ is the $N(N-1) \times 1$ vectorised version of the
traffic matrix $\bX$, with its columns stacked upon one another.

\autoref{fig:tomo} presents an example of \autoref{eq:observation}. It
shows how the traffic on a single link $y_1$, is built from the sum of
traffic routed across the link $x_1 + x_2$. We can see that by
stacking each of these equations we would get
\be
  \left(\begin{array}{l} y_1 \\ y_2 \\ y_3 \end{array} \right) =
  \left(\begin{array}{lll} 1 & 1 & 0 \\ 1 & 0 & 1 \\ 0 & 1 & 1 \end{array} \right)
  \left(\begin{array}{l} x_1 \\ x_2 \\ x_3 \end{array} \right),
\ee
which, written in matrix notation, is just
\autoref{eq:observation}. Note that in this case the routing matrix
$\bA$ is invertible, so the problem of inferring the traffic matrix from
link measurements is easy, but this is rarely the case. Usually, $L$
is usually much smaller than $N(N-1)$, and so the
problem is highly underconstrained.

There are two main assumptions implicit in this observation model. It
assumes the traffic matrix is stationary, \ie~its statistical
properties remain stable throughout the measurement interval and there
are no errors in the measurement. Stationarity is preserved by
choosing an appropriate measurement interval, say 1 hour (see
\autoref{sec:models}). Moreover, in reality, errors do occur and to
account for it, the model \autoref{eq:observation} is extended to 
\be 
\by = \bA \bx +\bz.
 \label{eq:observation_noise}
\ee

\noindent The second model is a simple noise additive model often used
to test the robustness of an inference technique. Each element of the
additive noise term $\bz$ is typically chosen to be independent and identically
distributed (i.i.d.) white
Gaussian noise with mean zero and variance $\sigma^2$. Often
$\sigma^2$ is kept small, as large values would result in some
elements of $\by$ violating the non-negativity constraint. It is
for this reason other distributions may be used, for example,
log-normal or gamma distributions. Additionally, due to the problem of
missing link information due to poor SNMP implementation, some of the
elements of $\by$ may be missing. Finally, if the given routing
matrix $\bA$ is incorrect, the observations $\by$ would
significantly depart from the true SNMP link counts. However, most
works assume an accurate $\bA$ because there are reliable methods for
obtaining routing information \cite{Shaikh02OSPF}.

There are usually many less links than the total number of IE pairs,
and so the inverse problem above is highly underconstrained.  Whether
noise is present or not there may be an infinite number of solutions
$\bx$ that fit the observations
\eqref{eq:observation}. \autoref{fig:tomo_hard} shows a picture of
such a network, where we only measure at the bottleneck. Now, even in
this very simple network the equations $y_1 = x_1 + x_2$ are
underconstrained. In order to make progress, some additional
information needs to be assumed, usually in the form of a traffic
matrix model, and we shall consider some of these in the following
section.

\begin{figure}[!thbp] 
  \begin{center}
    \includegraphics[width=0.6\columnwidth]{underdetermined.pdf}
    \caption{A harder inference problem where we only have one
      measurement, but two traffic elements to estimate. There is
      therefore a fundamental ambiguity in this problem.
      \label{fig:tomo_hard}}
  \end{center}
\end{figure}

Before moving on to the modelling aspects of traffic matrices, note that
there are other strategies for measurement. For instance, if MPLS is
being used, this creates a set of tables (in many implementations)
that can almost be read directly to obtain the matrix (\eg~see
\cite{blili08:_best_pract_deter_traff_matric}).  Alternatively, the
network operator may have more accurate local traffic matrices,
obtained through specific functionality at the routers. It is, in
principle, easy for a router to keep counts of its decisions
\cite{Varghese03Measure}, essentially amounting to a table of the
volumes of traffic between pairs of interfaces.  Locality here is
defined in the sense of the matrix's restriction to a single router --
we essentially see an IE traffic matrix of the single router's
interface.  These local matrices from all routers in the network can
be used to improve the estimation of the IE traffic matrix; see
\autoref{sec:applications}. On some special cases, such as a star
network, a single local matrix would be equivalent to the traffic
matrix, serving to highlight the information gain from local traffic
matrices. These matrices provide greater than a two-fold increase in
accuracy of the tomography estimation schemes by
\cite{Medina02TMdirections,Zhang03Fast}.

### 2013-07-22 Compiled/chapter8/models.tex
all files compile correctly

\clearpage
\section{Models}
\label{sec:models}

In this section, several canonical as well as recently proposed models
are presented. Modelling is divided into three categories: purely
temporal modelling, spatial modelling and spatio-temporal modelling.

\subsection{Temporal Modelling}
\label{ssec:temporal_modelling}

Purely temporal models only focus on the time series properties of the
traffic matrix. A key application of these models is in anomaly
detection, so as to be able to pinpoint the time and location of the
anomalous event. This is especially vital in detecting attacks on the
network or worm outbreaks. However, temporal behaviour is also
important in prediction, say for planning capacity of a future
network.

Before delving into the models, some basic issues regarding the
temporal properties of OD flows need to be understood. It is commonly
agreed that IP data traffic is rising exponentially, and has been for
more than a 
decade~\cite{claffy94:_track_long_term_growt_of_nsfnet,groschwitz94:_time_series_model_of_long,odlyzko03:_internet_growth,cisco_visual_networ_index2011}. There
was much early controversy about such growth estimates, because the
growth rate was vastly overestimated based on a small sample of
data. However, these days, exponential growth is considered the common
case (though with a much lower rate of growth), and is justified both
by data, and as a consequence of increasing computational and
networking speeds, governed by Moore's and Gilder's laws
respectively. As of the present, the introduction of mobile devices
and the rapid growth of traffic from these devices are set to further
increase data traffic in the years to come. Cisco (who admittedly have
a vested interest in a high forecast) estimate that ``Annual global IP
traffic will surpass the zettabyte threshold (1.3 zettabytes) by the
end of 2016''~\cite{cisco_visual_networ_index2011}.

Relatively few countries appear to monitor their national traffic, but
Australia is an exception. The Australian Bureau of Statistics have
collected and published traffic statistics for many
years~\cite{13:_abs_internet}. \autoref{fig:aust_traffic} shows the
growth of traffic in Australia from 2000 to 2012.

\begin{figure}[!b]
  \begin{center}
    \includegraphics[width=0.7\oneup]{australian_internet_usage_log.eps}
    \caption{Australian Internet traffic volumes based on data from
      the Australian Bureau of Statistics from 2000-2012. The
      dashed line shows a linear fit to the (log) data.  Note
      the log $y$-axis, so the plot shows quite a reasonable fit to
      exponential growth with a doubling period of 465 days. Over the
      same period the growth in broadband subscribers has been almost
      exactly linear (soon this trend must decrease as a large
      proportion of Australia's population are now connected), so most of
      the growth has come from growth in the amount downloaded per customer.
      \label{fig:aust_traffic}}
  \end{center}
\end{figure}

Regardless of your belief about the rate of growth, and/or the best
model (exponential is common, but linear and logistic models may also
be appropriate in many cases), any model of long-term traffic needs to
be able to incorporate such growth.

Second, most network traffic is human generated. Therefore, it stands
to reason that traffic is influenced by human activity in a 24 hour
cycle. In fact, distinct diurnal patterns have been observed, with
peak traffic occurring around mid-day or in the evening and dips
during the night, for example see \autoref{fig:abilene_2004_b}. This
correlates to the daily schedule of an average human being, where
mid-day traffic is generated for work or school purposes, while the
lack of traffic during the night correlates to sleeping periods.  Peak
traffic rate is also noticeably less on the weekends.  The regularity
of this behaviour can be quite strong as shown in the figure where two
successive weeks' data are overlaid so one can see how closely they
match. Most traffic measurements see some measure of cyclic behaviour,
with the degree of regularity determined by the type of traffic (is it
made up of many individuals or a few large flows) and the scale.

% \begin{figure}[thbp] 
%   \begin{center}
%     \hfill
%     \begin{subfigure}[b]{\twoup}
%       \centering
%       \Diurnal
%       \caption{Traffic rate over the span of a single week starting
%         from May 7th, 2001, with the traffic rate of
%         the following week overlaid.}
%       \label{fig:cycle_week}
%     \end{subfigure}
%     \hfill
%     \begin{subfigure}[b]{\twoup}
%       \centering
%       \DiurnalZoom
%       \caption{Traffic rate of a single day,  May 8th
%         2001, on the same ISP, with traffic on  
%         May 15th, 2001 overlaid on top.}
%       \label{fig:cycle_day}
%     \end{subfigure}  
%     \hfill\mbox{ } \\

%     \caption{Observations of the cyclicality of the total traffic rate of a large PoP.\label{fig:cycle}}
%   \end{center}
% \end{figure}

Third, and leading from the last point, the traffic volume itself is
dependent on the measurement period and the aggregation level of
traffic. At very short scales, in milli-seconds or seconds, the
traffic distribution is highly variable, and shows strong
dependencies, making use of such measures statistically non-trivial,
even if such measurements were easy to collect.  A common paradigm is
to consider a measurement interval of minutes to an hour, where
measurement is easy. Also important is the fact that at these times
scales \emph{stationarity}\footnote{Stationarity refers to the concept
  that the statistics of the traffic (for instance, the mean and
  variance, but in general including all statistics) are constant with
  respect to the time at which they are measured. Wide-sense
  stationarity is when only the mean and variance of the traffic
  satisfies the stationarity property.  In Internet traffic data it is
  only ever approximately true.  Moreover, it is hard to test for
  stationarity when traffic has long-term correlations, and so we can
  only ever talk about the degree of stationarity.} of the traffic
volume distribution is often a reasonable assumption, though we can
see the limits of this in \autoref{fig:abilene_2004_b} (stationarity clearly
doesn't hold for more than a couple of hours), however it was shown
that {\em cyclo-stationarity} holds to a large extent
\cite{Soule04TMLargest}.

Fourth, there are natural variations in traffic over time, and these
are often modelled as a random (or stochastic) process. This random
process could have all sorts of features, but there are some basics
that should be observed by all reasonable models. For instance,
network operators aggregate traffic from multiple sources, which is
known as \emph{multiplexing}. Multiplexing is used to boost the
efficiency of the links in a network by ``smoothing'' out variations
in traffic. The apparent smoothness is a result of decreases in the
relative variance, as predicted by the central limit
theorem~\cite{Cao01Nonstationarity}.  The more OD flows multiplexed on a
link, the higher the efficiency and smoothing effect, provided the
aggregated bandwidth does not exceed link capacity. Thus, any model
for the large traffic rates in a network must be consistent under
multiplexing, for example, when the number of flows being multiplexed
is increased the relative variance should decrease in a predictable
manner. Furthermore, the statistical properties of the aggregated
traffic must also be consistent with the statistical assumptions of
the traffic from a single user.

Finally, although rare, sometimes there may be sudden ``spikes'' in
traffic.  Such a component may arise from unusual traffic behaviour,
such as DDoS attacks, worm propagation or Border Gateway Protocol (BGP) 
routing instability from misconfiguration. Flash crowds are also an example of this behaviour,
which happens when there is a significant jump in the number of
clients to a particular web server or CDN. 
Extreme unforeseen events, such as the September 11 attacks on
the World Trade Centre in 2001 may instead cause a significant drop of
traffic rates. In any case, a massive shift in traffic rates would be
of interest to a network operator.

One temporal model of OD flows traversing backbone routers was
proposed by Roughan \etal~\cite{Roughan02BRvariable} by generalising
the Norros model \cite{Norros94Storage}, originally used for modelling
LAN traffic. Each OD flow is assumed to be generated from an
independent source. The model is characterised by the following
components (at time $t$):
\begin{enumerate}
\item $L(t)$, the long term traffic trend,
\item $S(t)$, the seasonal (cyclical) component, 
\item $W(t)$, random fluctuations, and
\item $I(t)$, the anomaly component.
\end{enumerate}
These components correspond to the observations of traffic described earlier.

The long term trend, $L(t)$, depends on the observed underlying
traffic growth in the data. This factor captures the overall growth of traffic over
a long time period. An exponential growth model for instance,
could be found by fitting $L(t) = A \exp(ct)$ to the data, with the
parameters $A$ and $c$ easily estimated via log-linear regression as
in \autoref{fig:aust_traffic}.

The component $S(t)$ may be any periodic function, such that
$S(t+kT_s) = S(t)$ for all integers $k$, where the period $T_s$ would
typically be 24 hours, or one week.

The component $W(t)$ is assumed to be a stochastic process with
zero mean and unit variance, to model the small fluctuating component of 
observed traffic. Finally, $I(t)$ captures the large variability of traffic from
anomalies, say, a massive upsurge or downsurge in traffic. These events were 
captured in an individual component to separate their influence from 
traffic under the network's normal operating conditions.

Let $x(t)$ denote the volume of an OD flow at time $t$. The model
takes the following form, 
\be 
  x(t) = m(t) + \sqrt{a m(t)} W(t) + I(t),
  \label{eq:temporal_model}
\ee

\noindent where $m(t) = S(t)\,L(t)$ is the mean of the OD flow,
assumed to be a product of the seasonal and long term trends, and $a$
is the \emph{peakedness} of the traffic. The average is modelled in
such a way because as large OD flows have a larger range of variation
in the size of their cycles. The parameter $a$ controls the smoothness
of the OD flow's volume in a way that is consistent given multiplexing
of aggregated flows.

\autoref{fig:decomposition} shows the decomposition in action on the
set of Abilene data shown in \autoref{fig:abilene_2004_b}, extending
the analysis beyond into the section of missing data. The estimated
long-term trend $\hat{L}(t)$ is not shown as it is almost constant
over this period, but we show the estimated mean $\hat{m}(t)=
\hat{S}(t)\, \hat{L}(t)$ (derived using a 5 term seasonal moving
average, further smoothed with a 13 term moving average), shown in
green. Normal bounds on this, obtained by modelling the random
fluctuations $W(t)$ as a Gaussian process are shown as dashed lines,
and where the traffic falls outside these bounds, we have indicated an
anomaly.

We have deliberately looked at this period, which is followed by
period of missing data to illustrate one of the advantages of this
approach, which is the ability to fill in missing gaps in the data,
though more sophisticated approaches to solve that problem also
exist, \eg~see \cite{Zhang09TMCS}.

\begin{figure}[htp]
  \begin{center}
    \includegraphics[width=\oneup]{Abilene_2004_decomp_1.eps}
    \caption{Results of decomposing traffic into components. The blue
      curve shows the original Abilene traffic from
      \autoref{fig:abilene_2004_b} (note that the data is missing for
      the third week). The solid green line shows $\hat{m}(t)=
      \hat{S}(t)\, \hat{L}(t)$, the estimated  mean, and the dashed
      green lines show the bounds of normal variation around this
      mean. The red asterisks indicate the anomalies $I(t)$.
      \label{fig:decomposition}}
  \end{center}
\end{figure}

Another feature of this model is the preservation of the properties of the
model through a linear combination, advantageous when looking at the
aggregated behaviour of the OD flows. Consider $K$ aggregated OD
flows, then 
\[
  x_{\text{agg}}(t) = \sum_{i=1}^K m_i(t) + \sum_{i=1}^K \sqrt{a_i m_i(t)} W_i(t) + \sum_{i=1}^K I_i(t). 
\]

\noindent The mean of $x_{\text{agg}}(t)$ is simply $m_{\text{agg}}(t)
= \sum_{i=1}^K m_i(t)$, and the peakedness is the weighted average of
the component peakedness, $a_{\text{agg}} =
\tfrac{1}{m_{\text{agg}}(t)} \sum_{i=1}^K a_i m_i(t)$. The linearity
properties allow $x_{\text{agg}}(t)$ to be expressed in the same form
as \autoref{eq:temporal_model} with the new parameters
$m_{\text{agg}}(t)$ and $a_{\text{agg}}$. The linearity property
enables the consistent computation of the variances of the aggregated
traffic, which is useful for network planning and analysis. Besides
this, \cite{Roughan02BRvariable} demonstrated the ease of estimating
the parameters of model \autoref{eq:temporal_model}, via simple
estimators and filtering.

The cyclical nature of the aggregated OD flows is also amenable to
Fourier analysis. The Fourier transform decomposes a periodic signal
into a weighted sum of sinusoids with distinct frequencies and
phases. It would be reasonable to assume the observed cycles can be
represented by a small number of Fourier coefficients (this does not
require that the signal be exactly sinusoidal as any periodic signal
can be approximated by a limited number of sinusoids). Indeed, it has
been demonstrated this is the case with the traffic volumes
\cite{Soule04TMLargest,eriksson10:_basis}, where as little as just
5 Fourier coefficients were needed to achieve low error in fitting
large OD flows of a Tier 1 network, demonstrating the relatively few
significant frequencies present in a diurnal cycle of the OD
flows.

\autoref{fig:fourier} shows a similar analysis of Abilene data from
\autoref{fig:abilene_2004_b}. It shows a simple example of the
excellent degree of approximation to traffic we can obtain using only
a very small number of Fourier coefficients corresponding to daily
periods. 
\begin{figure}[thbp] 
  \begin{center}
    \mbox{ }\hfill
    \begin{subfigure}[b]{\twoup}
      \centering
      \includegraphics[width=\textwidth]{Abilene_2004_fourier_sprectra.eps}
      \vspace{-5mm}
      \caption{Fourier spectra (technically a periodogram) of the
        Abilene data.}
      \label{fig:fourier_a}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.47\textwidth}
      \centering
      \includegraphics[width=\textwidth]{Abilene_2004_fourier_cum_power.eps}
      \vspace{-5mm}
      \caption{Cumulative power captured by Fourier components
        (sorted in descending order).}
      \label{fig:fourier_b}
    \end{subfigure}  
    \hfill \mbox{ }

      \vspace{5mm}
    \begin{subfigure}[b]{\oneup}
      \centering
      \includegraphics[width=\textwidth]{Abilene_2004_fourier_model_10.eps}
       \vspace{-9mm}
     \caption{Fourier approximation to the data using 10 (complex) coefficients.}
      \label{fig:fourier_c} 
    \end{subfigure}

      \vspace{5mm}
    \begin{subfigure}[b]{\oneup}
      \centering
      \includegraphics[width=\textwidth]{Abilene_2004_fourier_model_100.eps}
      \vspace{-9mm}
      \caption{Fourier approximation to the data using 100 (complex) coefficients.}
      \label{fig:fourier_d}
    \end{subfigure}

    \caption{Fourier analysis of Abilene data shown in \autoref{fig:abilene_2004_b}. Blue curves represent the
    Fourier approximation of the traffic, indicated by the red curves.\label{fig:fourier}}
  \end{center}
\end{figure}         
\autoref{fig:fourier_a} shows the periodogram of the data
(the absolute magnitude of the Discrete Fourier Transform (DFT)). We
can see that only a few peaks (with frequencies of 1, 2, 7, and 14
cycles per week, corresponding to daily and weekly cycles and their
harmonics) are large enough to matter for gross features.
\autoref{fig:fourier_b} shows the cumulative power contained in the
largest of the Fourier coefficients, clearly showing that a small
number of these represent the power of the signal
well. \autoref{fig:fourier_c} shows an approximation of the original
signal using only 10 coefficients. We can see that the cyclic
components of the data are represented well, though the actual signal
varies around that noisily. Including additional components provides a
better approximation (in the sense of fitting the data more
accurately), but we can see that we are simply recreating the
noise. There is a clear tradeoff here between noise and signal, with
no absolute standard for the correct choice, but the underlying
promise of Fourier analysis is obvious.

The choice of time interval to use in Fourier analysis/approximation
is interesting. A longer interval provides more data, and hence better
estimates if the data is truly {\em cyclostationary}\footnote{A
  cyclostationary process can be thought of as one whose component
  processes formed from times embedded at multiples of the fundamental
  period from stationary sequences.}. However, as we discussed earlier, 
  there are noticeable
% http://en.wikipedia.org/wiki/Cyclostationary_process
trends in the overall volume, and so it is reasonable to assume that
there will sometimes be significant shifts in the pattern (with
respect to time of day or time of week). In these cases, extending the
length of the dataset can confuse the statistical variability with
non-stationary effects, resulting in biased estimates. The best
tradeoff appears to depend on the dataset, but periods of perhaps a
month seem to work reasonably well for estimating weekly cycles.

However, there are alternative tools for such analysis, designed to
provide some flexibility in the tradeoff between time and
frequency. The simplest is perhaps the Short-Time Fourier Transform,
in which the signal is {\em windowed}, and the standard DFT is applied
to the windows of data to produce a {\em spectrogram}. The technique
has been applied in \autoref{fig:spectrogam} to a longer (6
weeks) sequence of the Abilene data. This segment of the data is more
challenging, as it contains many anomalies, \eg~the frequent spikes
in the traffic, or the drop in traffic on the Independence Day holiday,
observed on July 5th in 2004\footnote{Independence Day is actually 
celebrated on July 4th, but in 2004, the day falls on a Sunday. Thus, the following
day was declared a public holiday as well.}. \autoref{fig:spectrogam_a} shows the
set of data under consideration, along with its approximation using 10
coefficients. \autoref{fig:spectrogam_b} shows a spectrogram with
windows of length 1 week, and here we can still clearly see the weekly
and daily cycles appearing in most of the
windows. \autoref{fig:spectrogam_c} shows a spectrogram using 1 day
windows, with overlapping windows to smooth the results. The resulting
picture is poor at showing the cyclical behaviour of the traffic, but
clearly highlights the large anomalous spikes in the traffic in second
week of the data.

\begin{figure}[thbp] 
  \begin{center}
    \begin{subfigure}[b]{\oneup}
      \centering
      \includegraphics[width=\textwidth]{Abilene_2004_spectrogram_input.eps}
       \vspace{-9mm}
     \caption{Six weeks of Abilene data.}
      \label{fig:spectrogam_a} 
    \end{subfigure}

      \vspace{5mm}
    \begin{subfigure}[b]{\oneup}
      \centering
      \includegraphics[width=\textwidth]{Abilene_2004_fourier_spectrogram_week.eps}
       \vspace{-5mm}
       \caption{Spectrogram with weekly windows. Brighter colours
         indicate more power. Again we see strong power at 1 and 7
         cycles per week, though the strength of these varies per
         week. For instance, in week 5 (when the Independence Day
         holiday was held) there was a week day, whose traffic more
         closely resembled weekend traffic, breaking the weekly cycle,
         and pushing more power into the daily cycle.}
      \label{fig:spectrogam_b} 
    \end{subfigure}

      \vspace{5mm}
    \begin{subfigure}[b]{\oneup}
      \centering
      \includegraphics[width=\textwidth]{Abilene_2004_fourier_spectrogram_day.eps}
      \vspace{-5mm}
      \caption{Spectrogram with daily windows. Note that the
        time-resolution in this case is actually poorer than the image
      would suggest, accounting for the poor resolution of the daily
      and weekly cycles in this figure. However, the large anomalous
      spikes of traffic in the second week stand out clearly in this
      view, as they spread power across a range of frequencies.}
      \label{fig:spectrogam_c}
    \end{subfigure}

    \caption{Short-Time Fourier analysis of Abilene data.\label{fig:spectrogam}}
  \end{center}
\end{figure}

An even more powerful set of techniques that have been applied to the
temporal analysis of traffic are the Wavelet
transforms~\cite{Veitch97Wavelet,Barford02Anomaly,Papagiannaki05Long},
which provide a more flexible set of time/frequency
tradeoffs. Wavelets have also been applied to spatial analysis of
traffic matrices 
\cite{CrovellaKolaczyk03,CoatesCompressedNetworkMonitoring,wang10:_wavel_based_traff_matrix_model,rincon08:_dw,roughan_mra_09},
which we will consider in a moment.  However, wavelets are a
relatively complicated set of techniques, and it is outside the scope
of this chapter to provide an introduction to that material. See for
instance \cite{mallet:_book} for more information.

Principal components analysis (PCA) has also been employed to quantify
the temporal correlations of the traffic matrix. If $\cX$ is a matrix
where the rows represent a measurement (for instance an OD flow) and
columns represented traffic volumes at time $t$, then temporal PCA
decomposes the matrix $\cX^\T \cX$ into its corresponding components
of eigenvalues and eigenvectors. Often, each column is \emph{centred}, simply by
subtracting the mean vector $\bar\bx$, the average of all columns, from each 
column in $\cX$. In what follows, $\cX$ is assumed to be centred.

The matrix $\cX^\T \cX$ is \emph{positive
semidefinite}. Visualising this geometrically, if the columns of the matrix is reinterpreted
as a set of points, then they trace out an ellipsoid. Alternatively, $\cX^\T \cX$
may be viewed as the \emph{empirical covariance matrix} of the columns of
$\cX$, in effect computing temporal correlations in traffic.

PCA is used to find the directions of greatest variance of $\cX^\T \cX$
by decomposing $\cX^\T\cX = \bW \bD \bW^\T$, where $\bW$ is an orthonormal matrix 
containing the eigenvectors of $\cX^\T\cX$ and $\bD$ the diagonal 
matrix containing the eigenvalues of $\cX^\T\cX$. The eigenvectors
are known collectively as the \emph{principal axes}. The eigenvectors are ordered in a 
non-decreasing order with respect to their associated eigenvalues, starting from the
eigenvector associated with the largest eigenvalue to the smallest. An equivalent view
is that PCA essentially performs a Singular Value Decomposition (SVD) of the matrix
$\cX$, by computing only its right singular basis $\bW$.

Thus, every row of $\cX$ can be expressed as $\bx_k =  \baa^\T_k \bW^\T$, 
\ie~a linear combination of a coefficient vector
$\baa_k$, called the \emph{principal components}. Here, $\bW$ is equivalent to a linear
transform, post-multiplied to the data. Intuitively, if the size of the set of
principal axes with large principal components are small, then this is evidence
there are high temporal correlations between the traffic flows. In practice, it is
common to focus on the few largest principal axes for the purpose of data reduction.
Basically, this means choosing the first few significant columns of $\bW$ to approximate each
$\bx_k$ with $\tilde\bx_k$ such that $\|\bx_k - \tilde \bx_k\|_2 < \epsilon$, for some 
small error $\epsilon > 0$.

As an aside, PCA may be performed on $\cX\cX^\T$, in effect computing the spatial
correlations of $\cX$ instead. Here, we have $\cX\cX^\T = \bV \tilde\bD \bV^\T$, with
each column $\bar\bx_k = \bV \tilde \baa_k$, equivalent to $\bV$ pre-multiplied
with the data. Spatial PCA was used in the context of anomaly detection
\cite{Lakhina04Anomaly,Lakhina04Diagnosing,Lakhina05Mining} but there are problems
with this approach. These discussions are deferred to \autoref{sec:applications}.

\begin{figure}[thbp] 
  \begin{center}
     \includegraphics[width=0.7\textwidth]{PCA_example.eps}
    \caption{Principal components analysis of the empirical covariance matrix of a two dimensional 
    data matrix of 1000 centred points $\cX$. Here, ``PC 1''  and ``PC 2'' are the principal 
    components. Note the elliptical shape of the contours of the
    density with semi-major and semi-minor axis given by the first and
    second principal components, respectively.  \label{fig:pca}}
  \end{center}
 \end{figure}

\autoref{fig:pca} demonstrates an example of PCA performed on the covariance matrix of 
1000 two dimensional data points with zero mean, \ie~$\cX$ has 2 rows and 1000 columns. 
The matrix $\cX^\T \cX$ formed by the data points vaguely resembles an ellipse. Here, there
are two principal components, denoted by  ``PC 1''  and ``PC 2'', with the higher variance
captured by PC 1. This is clear from the way the points on the figure
are distributed. The key point to take away is that both components capture the direction
of highest variance and are orthogonal to each other.  Moreover, the principal components matrix 
$\bW$ has PC 1 and PC 2 as its first and second columns respectively. Each data point 
can be expressed as linear combination of these two components. The concept is easily
extended beyond two dimensions to the larger dimensions typically encountered with
traffic matrices.

PCA was performed by Lakhina \etal on empirical data from two backbone networks show
that OD flows are a combination of no more than 35 ``eigenflows'' (the principal axes), and
in fact, fewer than this in general \cite{Lakhina04TrafficStruct}, out of over 100 OD flows. 
These eigenflows belonged to one of three categories, depending on their properties:
\begin{enumerate}
\item \textbf{deterministic or $d$-eigenflow}: generally the
  significant diurnal component of the largest OD flows. Although
  present in smaller OD flows, these eigenflows are less
  significant. These eigenflows have a cyclo-stationary property and
  suggests that these eigenflows may be approximated by a small number
  of Fourier coefficients. These eigenflows account for the majority
  of the total traffic of the OD flow.
\item \textbf{spike or $s$-eigenflow}: medium sized eigenflows with a
  spikiness behaviour in time, with values ranging up to 5 standard
  deviations from the mean of the OD flow. This suggests these
  contributions come from bursty processes and may be modelled by a
  wideband Fourier process.
\item \textbf{noise or $n$-eigenflow}: small eigenflows behaving like
  stationary additive white Gaussian noise. These eigenflows have
  small energy and their contribution to overall traffic is
  negligible. The majority of eigenflows from Lakhina {\em et al.}'s datasets
  belong to this category.
\end{enumerate}
There are several eigenflows belonging to two or more categories, but
these eigenflows are rare \cite{Lakhina04TrafficStruct}.  For the most
part, these categories are very distinct for almost all
eigenflows. The low number of eigenflows compared to the dimension of
the traffic matrices under study suggests low intrinsic dimensionality
of traffic matrices, although the upper bound of 35 eigenflows indicates that the OD traffic
is only ``approximately'' low rank. The results show a power law-type distribution
of the principal components \cite{Lakhina04TrafficStruct}. The decay of the
distribution varies depending on the ISP, with some distributions
exhibiting a very fast decay and some much slower decay.

In many senses PCA confirms the previous analysis
and modelling, but it is interesting because its approach simply looks
for correlations across different sets of measurements, and uses a
different set of assumptions from, for instance, Fourier analysis
which can be performed on a single time series.

Finally, it is important to note that the full data needs to be
available (no missing entries in $\cX$) in order to perform
PCA. Furthermore, the basic flavour of PCA as described above is not a
robust method, since it is an entirely data driven method and is
therefore sensitive to outliers~\cite{Ringberg07PCA}. Robust variants
have been proposed but they necessarily complicate the basic version
of PCA presented here, since these modifications entail constructing
methods to identify and exclude outliers. Despite these disadvantages,
in its purest form, PCA is a useful tool to learn the temporal
structure of traffic flows.

\subsection{Spatial Modelling}

Spatial models only focus on the properties of traffic between source
and destination pairs, typically within a single measurement interval,
without regard to how the traffic changes in time. The models
presented here are the gravity model and its generalisations, the
discrete choices model, the independent connections model and low rank
spatial model. However, we shall start with the simplest test models. For ease of
exposition in this section, the set of sources and destinations are
assumed to be sets of PoP ingress and egress nodes, denoted by $\cI$
and $\cE$ respectively. The set $\Omega$ represents the set of all
nodes in the network, \ie~$\Omega = \cI \cup \cE$.

\subsubsection{Simple {\em test} models}

We must remember that the purpose of models is not always to
``realistically'' represent a network's traffic. Their purpose is to
provide inputs to other tasks. One common task is to assess the
sensitivity of a network to different types of traffic, and to that
end, engineers can consider the effect of various artificial {\em test}
models.

Three such are the uniform traffic model, peak load model, and
focussed overload model. They are extremely simple:
\begin{description}
\item[uniform] this simple model assigns the same value to all
  traffic matrix elements. It is used to provide a base load in some
  experiments, or to see the behaviour of a network under one extreme
  (the most uniform extreme) of traffic \cite[Chapter 4.5.1]{Cahn98WANDesign}.

\item[peak load] this model is equally simple, and equally extreme. It
  has zero for all loads except one OD flow. It simulates the opposite
  extreme where the aim is to see the effect of one dominant flow.

\item[focussed overload] this type of traffic matrix simulates the effect of a
  {\em focussed overload}, or {\em flash crowd}\footnote{See also the
    slashdot effect.}, where many users become interested in one
  location or resource and the traffic to this single location from
  all other sources is the dominant effect in the network. As a
  result, the focussed overload can be represented by a matrix with all
  elements zero, except for one row. We can likewise represent a
  focussed traffic load arising from a single point (say as response
  traffic to a focussed set of queries) by a matrix with a single
  non-zero column.
\end{description}
The advantage of each of the models lies in its simplicity. The
simplicity means that the effect of the traffic is easy to interpret,
and thus gain insights from these models where a more complex
model would perhaps confound us with multiple potential causes for
some results. For instance, in each of the above models we can
gradually increase the traffic to see when capacity bounds are
reached, and where those bounds would be reached in order to identify
potential bottlenecks in a network.

Other test models based on classical distributions such as the Poisson
and Gaussian distributions were proposed by Vardi \cite{Vardi96Tomo},
and Tebaldi and West \cite{Tebaldi98Tomo} and Cao \etal 
\cite{Cao00Tomo}, respectively. Their well-known properties make it easy
to analyse results and provide insights, at the cost of a departure from 
real traffic properties.

\subsubsection{Gravity model}

The gravity model is perhaps the next simplest type of model, but it
has a great deal to offer. Here, traffic from the source to the destination
are modelled as a random process. In its simplest form it assumes, any
packet originating a source to a destination nodes are \emph{independent}
of other packets. 
Depending on context, this could be the origin and destination, or ingress
and egress nodes respectively.  Consequently, the traffic between 
two nodes is \emph{proportional} to
the total traffic from the source node to the destination node. The
gravity model is amongst one of the most well-studied
models and is considered a canonical first generation model.

The name of the model derives from Newton's model of gravitation,
where the gravitational force is proportional to the product of the
mass of two objects divided by the distance between them squared. The
general formulation of the gravity model is defined by two forces: the
\emph{repulsive} force (factor) $R_i$, associated with ``leaving''
from $i$ and the \emph{attractive} force (factor) $A_j$, associated
with ``going'' into $j$. Its general form is described by the
following equation: \be X_{i,j} = \frac{R_i \,A_j}{f_{i,j}},
\label{eq:grav_model}
\ee where $f_{i,j}$ represents the \emph{friction factor}, which
describes the weakening of the forces (akin to distance in Newton's
model), depending on the physical structure of the modelled
phenomenon. The model has been used extensively in various fields, for
instance the modelling of street traffic~\cite{potts72}.

In the context of Internet traffic matrix modelling, the friction
factors have typically been taken to be constant. That is, distance is
assumed to have little effect on network traffic. That certainly
seemed to be true even at a fairly large scale in the past, but it is
unknown to what extent the deployment of CDNs (Content Distribution
Networks) over the last few years has changed distance dependence, 
since CDNs locate traffic closer to the end user to avoid paying for
transit costs across the network, or
how inter-country matrices are affected by distance (for instance
through language barriers). Where distance is ignored, equation
\autoref{eq:grav_model} becomes 
\be X_{i,j} =
\frac{X_i^{\text{in}}\, X_j^{\text{out}}}{X^{\text{total}}},
\label{eq:network_grav}
\ee
where $X_i^{\text{in}}$ is the total traffic entering the
network through $i$, $X_j^{\text{out}}$ is the total traffic exiting
the network through $j$ and $X^{\text{total}}$ is the total traffic
across the network \cite{Zhang03Fast}. The model can be expressed
succinctly as the single rank matrix 
\be 
\bX =
\frac{\bx^{\text{in}}\, \left.\bx^{\text{out}}\right.^\T}{X^{\text{total}}}.
\label{eq:network_grav_mtx}
\ee
The popularity of the model stems from the ease of
estimating the $X_i^{\text{in}}$ and $X_j^{\text{out}}$ for each node
pair $(i,j)$, and especially at the PoP or backbone level, since the
level of traffic aggregation mitigates errors in the estimation of
these quantities from sampled traffic.

The gravity model only captures the spatial structure of the
traffic. The key assumption of the gravity model is the independence
between each source $i$ and destination $j$. Coupled with the
assumption that none of the nodes act as a source or sink of traffic
(\ie~that traffic is conserved in the network) $X^{\text{total}} =
\sum_{k \in \cI} X_k^{\text{in}} = \sum_{\ell \in \cE}
X_\ell^{\text{out}}$. Under normal operating conditions in most
backbone routers, where congestion is kept to a minimum, the
conservation assumption appears reasonable. With this assumption,
\be
X_{i,j} = X^{\text{total}}\, p_i^{\text{in}}\, p_j^{\text{out}},
\label{eq:network_grav_normalised}
\ee
where
\ben
p_i^{\text{in}} 
    = \frac{X_i^{\text{in}}}{\sum_{k \in \cI} X_k^{\text{in}} },
\quad \mbox{ and } \quad 
p_j^{\text{out}} 
    = \frac{X_j^{\text{out}}}{\sum_{\ell \in \cE} X_\ell^{\text{out}}},
\een
are the proportions of traffic entering the ingress and
exiting the egress nodes respectively, called \textit{fanouts}. The
formulation \autoref{eq:network_grav_normalised} is known as the
\emph{fanout} formulation because it describes how a packet entering
via node $i$ is distributed to several nodes $j \in \cE$. Fanout has
been demonstrated to be close to a constant over several measurement
intervals, compared to the traffic matrix \cite{Medina02TMdirections},
suggesting the fanout may be a better alternative to measure and use
in, for instance, anomaly detection, than the raw traffic volumes.

Observe the implication of independence between the source and
destination in \autoref{eq:network_grav_normalised}: $\Pr(\cI,\cE) =
p_\cI^{\text{in}} p_\cE^{\text{out}}$. An immediate consequence is
$\Pr(\cE\,|\,\cI) = P_d(\cE)$, where $P_d(\cE)$ is the marginal
distribution of the traffic demand distribution at the
destinations. The assumption of independence between the source and
destination leads to two important properties of the gravity model
making it well suited to traffic matrix modelling.

\begin{thm}[Independence]
  Independence between the source and destination traffic holds for
  any randomly chosen submatrix of the model.
  \label{thm:independence}
\end{thm}
\begin{proof}
  The independence property implies $\Pr(s,d) = p_s^{\text{in}}\,
  p_d^{\text{out}}$, holding for every $s \in \cI$ and $d \in \cE$.
  This condition would also hold for a subsample of locations in $\cI$
  and $\cE$.
\end{proof}

\begin{thm}[Aggregation]
An aggregate of the gravity model is itself also a gravity model. 
\label{thm:aggregation}
\end{thm}
\begin{proof}
  Let all nodes be partitioned into $N$ subsets $\{\cS_1,\cS_2,
  \cdots,\cS_N\}$, with $\cS_i \cap \cS_j = \emptyset$ for $i \ne j$
  and $\cup_{i=1}^N \cS_i = \Omega$. The aggregated traffic matrix is
  defined as \be X_{\cS_i,\cS_j} = \sum_{i \in \cS_i} \sum_{j \in
    \cS_j} X_{i,j}.
\label{eq:aggregated}
\ee
The independence condition implies 
\be
X_{i,j} = \frac{X_{i,\Omega} \, X_{\Omega,j}}{X^{\text{total}}}.
\label{eq:independence_model}
\ee
Substituting \autoref{eq:independence_model} into \autoref{eq:aggregated}, 
\begin{align*}
X_{\cS_i,\cS_j} &= \sum_{i \in \cS_i} \sum_{j \in \cS_j} \frac{X_{i,\Omega} \,X_{\Omega,j}}{X^{\text{total}}}
= \frac{1}{X^{\text{total}}} \sum_{i \in \cS_i} X_{i,\Omega} \sum_{j \in \cS_j} X_{\Omega,j}\\
&=  \frac{X_{\cS_i,\Omega} \, X_{\Omega,\cS_j}}{X^{\text{total}}}.
\end{align*}
which is also a gravity model.
\end{proof}

These are not just theoretical results. Any model should be consistent
in the sense that if the data to which it applies is viewed in a
different way (for instance by sampling or aggregation) then the model
should still apply (though its parameter values may change). It seems
like an obvious requirement, and yet there are many models to which it
does not apply.

The utility of the gravity model is not just restricted to network
measurement. It is used in various areas: teletraffic modelling
\cite{Kowalski95TeleModel,Lam97Teletraffic}, economy and trade
\cite{Pyhnen63Trade,Tinbergen62Econ}, epidemiology
\cite{Ferrari06Pollinator,Murray77Measles,Xia04Measles}, sociology
\cite{Stewart48Socio}, the retail industry, specifically Reilly's law
of retail gravitation
\cite{Converse49RetGravNew,Jung59RetailGravTrue,Reynolds53RetailGrav},
and in vehicular traffic modelling \cite{Erlander90GravTransport}. More
advanced discussion on the gravity model (albeit with an economics
flavour) is found in \cite{Sen95Gravity}.

The gravity model can be interpreted in terms of the \emph{principle
of maximum entropy}. Entropy here is the Shannon entropy from
information theory parlance \cite{Cover06InfoTheory}. The principle is
closely related to Occam's Razor, essentially choosing the
parsimonious explanation of the data amongst competing
explanations. With little information regarding the traffic matrix
besides the total traffic information, it turns out that the best one
can do, according to the principle, is to describe the observations
with a model promoting independence and symmetry, consistent with
known constraints.  In this way, the model enjoys robustness compared
to other models, as the gravity model seeks to minimise deviation from
what has already been observed.

The model, however, is not without its drawbacks. The main critique
against the gravity model is in its main assumption: the independence
of the ingress and egress nodes\footnote{The difference between OD and
  IE traffic matrices becomes critical here.}. It has been pointed out
in several papers \cite{Erramilli06IndepConn} that this assumption
does not hold true. Most traffic between node pairs are determined by
connections, for example TCP initiated sessions, so there exist
dependencies between node pairs. The second is the violation of the
conservation of traffic assumption, for \eg~when there is high
congestion, causing packets to be dropped from router queues.

Actual traffic matrices are generally asymmetric, violating the main assumption
of gravity models. For example, forward traffic volumes of a source-destination
pair of nodes do not typically match up with the volume of reverse
traffic.  Even if the OD traffic matrix matches the gravity model
well, the corresponding IE traffic matrix may be vastly different, due to hot
potato routing \cite{Teixeira04PotatoSIG}.

Hot potato routing is implemented as a part of the BGP 
decision process \cite{Rekhter1995BGP}. BGP allows 
network operators to choose the egress points of traffic at the 
prefix level. The decision may also vary across a network so 
that traffic at different points  can end up being routed to 
different egress points. The idea of hot potato routing comes 
from its namesake: traffic is the ``hot
potato'' in this case and the network tries to get rid of the ``hot
potato'' as quickly as possible to avoid costs of transiting it over 
long distances. Therefore, traffic is sent on the shortest external 
route connecting an ingress to egress point. BGP provides less
control over ingress points and this is what leads to the 
fundamental asymmetry in the IE traffic matrix.

An example of hot potato routing is in \autoref{fig:hot_potato}.
Here, there is a clear asymmetry since the paths taken by traffic flows from
Perth to Sydney differ from Sydney to Perth. 
To further understand hot potato routing, we refer the reader to 
 \cite{Caesar05BGP} for a basic understanding
of BGP routing policy.

\begin{figure}
  \hfill \HotPotatoIn \hfill \HotPotatoOut \hfill \mbox{ } \\
  \caption{Traffic flow between two ASes, one in Perth and the other in
    Sydney. Note the asymmetry in traffic: due to the action of hot
    potato routing, the path taken by a traffic flow from Perth to
    Sydney differs from the reverse path, since by BGP's implementation, 
    the closest external link of an AS is always chosen to route traffic out from the AS.}
  \label{fig:hot_potato}
\end{figure}

Thus, although the source-destination independence assumption may hold
for OD traffic matrices, it may not necessarily hold for IE traffic
matrices, due to distortion by inter-domain routing. Consider a simple
toy example of a network in \autoref{fig:gravity_example} (originally
from \cite{Alderson06Topology}). The ASes A, B and C are assumed to be
connected, with A having three routers: 1, 2 and 3. The inter-domain
routing protocol between these ASes uses hot potato routing, seeking
the shortest path between these ASes.

Suppose $X^{\text{total}} = 9$. Consider an OD traffic matrix with the form of a gravity model, with even spread of traffic over each 
internal router 1, 2 and 3, with $\bx^{\text{in}} = \bx^{\text{out}} = \bx$. The OD traffic matrix has the form $\bX_{\text{OD}} = \bx \bx^
\T/X^{\text{total}}$, with $\bx = (1,1,1,3,3)^\T$, and written explicitly as
\be
\bX_{\text{OD}}  = 
\begin{array}{lll}
       & \;\;1 \;\;\;\;\;\;\;  2 \;\;\;\;\;\;\; 3 \;\;\;\;\;\;\; \text{B} \;\;\;\;\;\;\; \text{C} \\
          \begin{array}{lll}
            1 \\
            2 \\
            3 \\
            \text{B} \\
            \text{C} \\
          \end{array} 
       & 
       \hspace{-4mm}
          \left( 
          \begin{array}{lllll}
         1/9 & 1/9 & 1/9 & 1/3 & 1/3 \\
         1/9 & 1/9 & 1/9 & 1/3 & 1/3 \\
         1/9 & 1/9 & 1/9 & 1/3 & 1/3 \\
         1/3 & 1/3 & 1/3 & 1   & 1   \\
         1/3 & 1/3 & 1/3 & 1   & 1   \\
          \end{array} 
          \right)
      \end{array}.
\ee
By \autoref{thm:aggregation}, the gravity model for the aggregated OD matrix, comprising OD traffic volumes between ASes A, 
B and C, is given by
\be
\bX'_{\text{OD}} =
 \begin{array}{lll}
       &   \text{A} \;\;\; \text{B} \;\;\; \text{C} \\
          \begin{array}{lll}
            \text{A} \\
            \text{B} \\
            \text{C} \\
          \end{array}
       &
       \hspace{-4mm}
          \left(
          \begin{array}{lll}
            1 & 1 & 1 \\
            1 & 1 & 1 \\
            1 & 1 & 1 \\
          \end{array}
          \right)
      \end{array}
\label{eq:OD_aggregate},
\ee
simply by summing the traffic in the internal nodes. In this case, $\bX'_{\text{OD}}  = \bx\bx^\T/X^{\text{total}}$, 
with $\bx = (3,3,3)^\T$, still a gravity model.

\begin{figure}[p]
\GravityEx
\caption{Example toy network with three ASes: A, B and C are all assumed to be peers. The routers 1, 2 and 3 are internal to A.}
\label{fig:gravity_example}
\end{figure}

\begin{figure}[p] 
  \begin{center}
    \begin{subfigure}[b]{\twoup}
      \centering
      \GravityInternal 
      \caption{Internal traffic within A.}
      \label{fig:traffic_routing_a}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{\twoup}
      \centering
      \GravityIn
      \caption{Incoming traffic to A.}
      \label{fig:traffic_routing_b}
    \end{subfigure}

    \begin{subfigure}[b]{\twoup}
      \centering 
      \GravityOut
      \caption{Outgoing traffic from A.} 
      \label{fig:traffic_routing_c}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{\twoup}
      \centering
      \GravityExternal
      \caption{Traffic external to A.}
      \label{fig:traffic_routing_d}
    \end{subfigure}

    \caption{Traffic flows within the network of
      \autoref{fig:gravity_example}, classified into four
      components.\label{fig:traffic_routing}}
  \end{center}
\end{figure}

In order to construct the IE traffic matrix, the ingress and egress points of the network in A needs to be determined. The following
assumptions are made:
\begin{enumerate}
\item A, B and C are peers,
\item the shortest AS path protocol is used for inter-domain routing, 
\item hot potato routing is used internally by A, and
\item the Interior Gateway Protocol (IGP) weights are all equal.
\end{enumerate}
Suppose ingress and egress points are defined by the following routing table ($*$ represents a wildcard character)
 \begin{center}
\begin{tabular}{|c|c|c|}
\hline
  \textbf{Origin router} & \textbf{Destination}  & \textbf{Egress router} \\
  \hline
  1 & B & 2  \\
  1 & C & 3  \\
  2 & * & 2 \\
  3 & * & 3 \\
  \hline
\end{tabular}
\end{center}
The path for each traffic flow in the network, therefore, differs depending on its source 
and destination.

All traffic flows between the PoPs may be decomposed into four components: internal traffic within A, traffic 
departing A, traffic coming into A and traffic external to A, shown in \autoref{fig:traffic_routing}. 
The internal traffic of A (\autoref{fig:traffic_routing_a}) is just the top-left $3 \times 3$ submatrix of $\bX_{\text{OD}}$, which is
\be
 \bX_{\rm internal} =  \begin{array}{lll}
       & \;\;\;  1 \;\;\;\;\;\;\;  2 \;\;\;\;\;\;\; 3  \\
          \begin{array}{lll}
            1 \\
            2 \\
            3 \\
          \end{array} 
       & 
       \hspace{-4mm}
          \left( 
          \begin{array}{lllll}
         1/9 & 1/9 & 1/9 \\
         1/9 & 1/9 & 1/9  \\
         1/9 & 1/9 & 1/9  \\
          \end{array} 
          \right)
      \end{array}
\ee
Traffic bound for A, as seen in \autoref{fig:traffic_routing_b} to be specifically for router 1 in this instance, from its peers has entry 
points controlled by B and C, given the above routing table. Hence, from A's point of view, the traffic behaves as if the traffic 
randomly distributed across ingress links. Assuming the traffic is evenly spread, the traffic matrix is 
\be
 \bX_{\rm arriving} =  \begin{array}{lll}
       & \;\;\; 1 \;\;\;\;\;\;\;  2 \;\;\;\;\;\;\; 3  \\
          \begin{array}{lll}
            1 \\
            2 \\
            3 \\
          \end{array} 
       & 
       \hspace{-4mm}
          \left( 
          \begin{array}{lllll}
         0 & 0 & 0  \\
         1/3 & 1/3 & 1/3  \\
         1/3 & 1/3 & 1/3  \\
          \end{array} 
          \right)
      \end{array}
\ee
Traffic departing from A, seen in \autoref{fig:traffic_routing_c} as originating from router 1, and routed by hot
potato routing, is described by
\be
\bX_{\rm departing} =  \begin{array}{lll}
       & \; 1 \;\;\;\;\;\;  2 \;\;\;\;\;\;\; 3  \\
          \begin{array}{lll}
            1 \\
            2 \\
            3 \\
          \end{array} 
       & 
       \hspace{-4mm}
          \left( 
          \begin{array}{lllll}
         0 & 1/3 & 1/3  \\
         0 & 2/3 & 0  \\
         0 & 0   & 2/3  \\
          \end{array} 
          \right)
      \end{array}
\ee
Since A does not provide transit for B and C, traffic external to A, \ie~between B and C, should not appear on A, the traffic will 
remain unseen by A (\autoref{fig:traffic_routing_d}). Thus, the total IE traffic matrix is the sum of the component traffic above, so 
that the entry and exit points match, and is given by
\be
\bX_{\text{IE}} =
 \begin{array}{lll}
          \left(
          \begin{array}{lll}
         1/9 & 4/9  & 4/9 \\
         4/9 & 10/9 & 4/9 \\
         4/9 & 4/9  & 10/9 \\
          \end{array}
          \right)
      \end{array}.
\ee

\noindent The matrix $\bX_{\text{IE}}$ is not equal to $\bX'_{\text{OD}}$ in
\autoref{eq:OD_aggregate}, simply due to traffic asymmetry resulting
from hot potato routing. Moreover, the assumption of the conservation
of traffic no longer holds, since the total traffic of $\bX'_{\text{IE}}$ is
not equal to $X^{\text{total}}$. The diagonal terms, for example, are
much larger than in $\bX'_{\text{OD}}$.  This example demonstrates that even
if the OD traffic matrix is generated from the gravity model, the IE
traffic matrix does not necessarily have a structure that conforms to the
gravity model.

For large backbone networks where large aggregates of traffic are
observed, the gravity model performs admirably, as evident
from the results of \cite{Zhang03Fast} and its use in AT\&T's backbone
network for traffic engineering purposes. On smaller, local area networks, however,
its effectiveness is limited. The friction factor $f_{i,j}$ may not necessarily be
constant in actual traffic matrices, possibly due to different time
zones \cite{Roughan10Robust}, especially for a global spanning
network, language barriers, or the increased deployment of CDNs\footnote{The
deployment of CDNs exacerbates this effect since they are located close to
the end user so as to avoid having to pay for their traffic transiting 
other networks.}. There may also a distance dependency present 
between ingress and egress points \cite{Alderson06Topology}.

The gravity model by itself incurs significant estimation error as the
estimates obtained typically do not match the observed link
counts. Due to violations of these assumptions, the gravity model
turns out to be inaccurate when used in traffic matrix estimation. For
example, it was reported to have $\pm 39\%$ accuracy when used in
estimating traffic matrices \cite{Roughan05GravSynth}.

Despite the flaws mentioned, the gravity model was reported to be a
good initial estimate to more sophisticated methods.  The model was
paired with SNMP link measurements to develop the so-called
\emph{tomogravity} technique \cite{Zhang03Fast}. The gravity model is
also surprisingly useful in the synthesis of traffic matrices. When
proposed as a method for synthesising traffic matrices by Roughan
\cite{Roughan05GravSynth}, the gravity model serves as an excellent
first order model for generating the cumulative distribution function
of the traffic demands, closely mimicking the statistical properties
of actual traffic matrices. While the basic gravity model may not
necessarily be an optimal model, it is a simple and good first order
model for estimation and synthesis purposes, and it can be improved to
take into account the factors described above.

\subsubsection{Generalised gravity model}

In order to improve the efficacy of the basic gravity model and to
address its deficiencies, a generalisation of the gravity model was
developed \cite{Zhang03InfoSIGCOMM,Zhang05InfoTh}. In a nutshell, the
assumption of independent ingress and egress nodes was relaxed by
dividing traffic into several classes of ingress and egress nodes,
evident from the example in the previous section.  Independence only
applies to traffic belonging within a certain class, effectively
enforcing a \emph{conditional independence} criterion. Such an
assumption is closer to actual conditions between ingress-egress pairs
in a network.

In particular, the model now accounts for asymmetry of the IE traffic
matrix. To account for the effect from hot potato routing, traffic is
separated into classes based on peering and access links. Consider
again the network in \autoref{fig:gravity_example}. From the figure,
two classes can be defined: internal and external classes. There are
then four types of source-destination links (see
\autoref{fig:traffic_routing}): \textit{internal to internal, internal
  to external, external to internal} and \textit{external to
  external}.

In the generalised gravity model, independence between nodes are only
assumed between the internal to internal class and the external to
external class. Thus, routers 1,2 and 3 in AS A are independent to
each other, and so are ASes A, B and C to one another, but not traffic
from 2 to B, for instance.

Thus, in the generalised gravity model, a modification is made by
ensuring the independence assumption still holds, but only when
conditioned within each traffic class. In terms of probabilities,
traffic is \textit{conditionally independent}, as formulated below for
the joint fanout distribution of the sets of access nodes of the
network of interest $\cA$ and peering nodes $\cP$ respectively:
\be p_{S,D}(s,d) =
\begin{cases}
\frac{p_S(s)}{p_S(\cA)} \frac{p_D(d)}{p_D(\cA) } ( 1 - p_S(\cP) - p_D(\cP) ), &  \text{ for } s \in \cA, d \in \cA, \\
p_S(s) \frac{p_D(d)}{p_D(\cA)},  & \text{ for } s \in \cP, d \in \cA, \\
\frac{p_S(s) }{p_S(\cA)} p_D(d),   & \text{ for } s \in \cA, d \in \cP, \\
0, & \text{ for } s \in \cP, d \in \cP. 	
\end{cases}
\label{eq:conditional_independence}
\ee 
The four probabilities corresponds to the four cases in
\autoref{fig:traffic_routing}. In particular, as per intuition,
peering traffic is set to zero, since this class does not transit the
network of interest.

The stratification of traffic into several classes results in an
improved model. Its performance in the traffic matrix estimation
results is significantly better than the basic gravity model
\cite{Zhang03InfoSIGCOMM}. Further stratification beyond separating
peering and access nodes is possible. For example, the origin of the
traffic, whether from a fixed location or mobile device, or the
destination of the traffic, depending on application profiles, may be
defined as new classes in the model. However, further classification in
this manner is only possible with more side information available.

The generalised gravity model is superior to the basic model, but
gravity models in general have been somewhat tarnished by the same
brush. Most works benchmarking the performance of various models, for
instance, for estimation, compare against only the simple gravity
model, but make confusing statements that could lead one to believe
that all such models are faulty. In fact, the generalised gravity
model is vastly superior, but rarely used outside of the company at
which it was first developed --- AT\&T.  The chief reason is that the
model requires additional topological and routing data, and for the
external traffic flows to be mapped using this data \cite{Zhang05InfoTh}. This is a
non-trivial task. In addition, in many external studies researchers
have not had access to, in particular, knowledge on access and
peering links in the network under study. Network operators
are not open to releasing information on their networks to the
public, however, the Abilene dataset, used in \cite{Zhang05InfoTh},
and the G\'EANT dataset \cite{GEANT} are
publicly available, and contains enough information to make such
comparisons.

%%%% Abiline dataset where we did provide this!!!

\subsubsection{Discrete choice models}

Another proposed model is the \textit{choice model}, introduced by
Medina \etal~\cite{Medina02TMdirections}. The basis of the discrete
choice model (DCM) is the theory of choice models for decision
behaviour, originally developed in psychology, and later expanded upon
by researchers in other fields, more recently in economics, by Daniel
McFadden, for which he won the Nobel Prize in Economics in 2000 (see
for example, \cite{McFadden78Choice}).

Choice models are popular in econometric applications as the model is
used to describe a simplified underlying mechanism of rational
decision behaviour. It has been used for transportation analysis,
econometrics, marketing and consumer theory. The main inspiration for
its use in Internet modelling comes from \cite{Swait59Choice}, where a
choice model is used in the context of modelling the behaviour of
travellers between the cities of Maceio and Sao Paulo, two cities in
Brazil, as it parallels traffic traversing PoPs.

The choice model is defined by four elements:
\begin{enumerate}
\item the decision makers,
\item the set of alternatives (choices),
\item the attributes of the decision maker and the set of alternatives, and
\item the decision rules.
\end{enumerate}
All these elements play a key role in ultimately determining the decision process. The \emph{decision makers} represent the 
agents making the decisions on which choices to go for. The \emph{set of alternatives} characterise the set of possible 
actions the agents can choose. Each decision maker executes several choices based on its own inherent properties, or
\emph{attributes}, as well as the attributes of the set of alternatives. These attributes predispose a decision maker to
certain alternatives. Finally, the \emph{decision rules} determine how choices are made. How good a choice is, is measured
by a standard based on a set of criteria. The rules establish constraints on the choices of the
decision makers, enforcing consistency in the entire system. All four elements of the model aim to capture how
agents would naturally decide on several differing choices in a system, in a rational and consistent way, based on a set of rules.

In the context of network traffic modelling, there are two interdependent factors influencing choices.
First, the network users' behaviours determine much of how traffic flows are generated, as discussed in  
\autoref{ssec:temporal_modelling}. Second, the network design and configuration plays a very important role in how traffic flows 
are delivered on
the network. Routing protocols, policies, QoS as determined by the network operator and the geographical local
of routers and PoPs determine how traffic is transported within the network and between networks. One could visualise this as
a two level process: users generate the traffic flows, whereupon the flows are routed through the network, based on the network's 
design and policies, to the flows' destinations.

All four elements have direct analogues in the context of network traffic modelling. The decision makers are the set of ingress 
nodes, aggregating all information about the users' behaviours and network design and policies. The set of alternatives are the
set of egress nodes, which aggregates the information about the users connected to these nodes. Thus, each decision maker $i$
has a choice set $\cC \subseteq \cE$. Each node $i$, a decision maker, is modelled by the equation, for all $j \in \cC$,
\be
U^i_j = V^i_j + \varepsilon^i_j,
\label{eq:utility_node}
\ee
where $U^i_j$ denotes the utility between the node pair $i$ and $j$, $V^i_j$ aggregates the information from the user 
behaviour and network design, which is deterministic, and $\varepsilon^i_j$ is a random component to account for missing
information from unknown factors. The term $V^i_j$ can be thought of accounting for the level of attractivity
of a destination node $j$. In \cite{Medina02TMdirections}, the authors proposed $M$-attributes per decision maker-choice
pair, such that
\be
V^i_j = \sum_{m=1}^M \mu_m \omega^i_j(m) + \gamma_j,
\label{eq:two_factor}
\ee
where $\omega^i_j(m)$ denotes the $m$-th attribute, $\mu_m$ are weights to account for the relative importance of the $m$-th 
attribute, and $\gamma_j$ is a scaling term for other factors for attractivity, besides all $M$ attributes. An attribute
$\omega^i_j(m)$ could be the size of the destination node PoP, since a large egress PoP is more likely to have traffic
exiting from it, or the number of peering links the destination node $j$ has.

Based on the above, Medina \etal \cite{Medina02TMdirections} proposed a traffic matrix model that assumes the decomposition
\be
X_{i,j} = O_i \alpha_{i,j}.
\label{eq:choice_model}
\ee
The parameters $O_i$ and $\alpha_{i,j},\,\forall j$ denote the total outgoing traffic volume from node $i$ and the 
\emph{fanout} of node $i$ respectively. For each $i$, $\sum_{j}
\alpha_{i,j} = 1$. The total traffic from a node $O_i$ is known from SNMP data. Observe 
that the traffic matrix is now parameterised by the fanout
distribution which has a direct analogy in the gravity model. In inference applications, it is the fanout distribution 
being estimated, thus indirectly inferring the traffic matrix, rather than directly estimating the traffic demands. Fanouts have been 
shown to be generally stable over a measurement period (several hours), compared to traffic demands \cite{Gunnar04TMLargeIP}, 
which is advantageous in traffic matrix estimation, since the stability contributes to more accurate inference.

The fanout distribution is determined by a decision rule. In \cite{Medina02TMdirections}, a utility maximisation criterion 
was used,
\be
\alpha_{i,j} =  \Pr\left(U^i_j = \max_{k\in\cC}\  \{U^i_k\}\right).
\label{eq:decision_rule}
\ee
Now, $\alpha_{i,j}$ is a random quantity as it depends on $\epsilon_{i,j}$, as observed from equation \autoref{eq:utility_node}. 
A natural starting point is to assume $\varepsilon^i_j$ is i.i.d.~Gaussian distributed with mean 0 and variance 1. This
transforms \autoref{eq:decision_rule} to the well-known multiple normal probability unit or m-probit model
\cite{McCullagh89GenLin}. However, there is no closed form for \autoref{eq:decision_rule} under this assumption. Instead, by
assuming $\varepsilon^i_j$ is i.i.d.~distributed following the Gumbel distribution, the m-probit model can be approximated,
with  \autoref{eq:decision_rule} now having a closed form. This model is popularly called the multiple logistic probability
unit or m-logit model \cite{McCullagh89GenLin}. The closed form is simply
\be
\alpha_{i,j} = \frac{\exp(V^i_j)}{\sum_{k \in \cC} \exp(V^i_k)},
\label{eq:logit}
\ee
implying that
\be
X_{i,j} = O_i \frac{\exp(V^i_j)}{\sum_{k \in \cC} \exp(V^i_k)}.
\label{eq:choice_model_logit}
\ee
The difficulty lies in determining what attributes should be included. The authors considered two models which they empirically 
validated:
\begin{enumerate}
\item $V^i_j = \mu_1 \omega_j(1) + \gamma_j$, where $\omega_j(1)$ denotes the total incoming bytes to an egress PoP $j$, and
\item $V^i_j = \mu_1 \omega_j(1) +\mu_2 \omega^i(2) + \gamma_j$, where in addition, $\omega^i(2)$ denotes the total bytes 
leaving the ingress PoP $i$.
\end{enumerate}
In general, the second model is more accurate, owing to the additional attribute, but it is not known if it is just a case of overfitting 
or the new parameter is truly useful.

The choice model is a variation of the gravity model. In particular, looking back at 
equation \autoref{eq:choice_model}, the total traffic outflowing from ingress $i$ may be regarded as the \emph{repulsion
factor}, while the parameters $\alpha_{i,j}$ combining both the \emph{attractiveness factor} and the \emph{friction factor}. A
quick comparison of the choice model to \autoref{eq:network_grav_normalised} highlights the strong link between both models. 
The choice model, however, has a larger number of parameters to account for the attributes of the decision maker and set of 
alternatives.

\subsubsection{Independent connections model}

The independent connections model (ICM) was introduced in
\cite{Erramilli06IndepConn,Erramilli06IndepConnTech}. Unlike the
gravity model, this model discards the assumption of independence
between the ingress and egress nodes, and instead focuses on the
\textit{connections} between nodes. More specifically, the model
differentiates between \textit{initiators}, nodes that initiate a
traffic connection, such as a TCP connection, and \textit{responders},
the nodes that accept these connections. The independence assumption
comes in by assuming that each initiator and responder are
independent, in effect, resulting in independent {\em connections}.

The inspiration for the ICM comes from
traffic characterisation studies, specifically on TCP behaviour. TCP
creates two-way connections in response to a SYN packet, the packet used
to initialise a connection. Although it is
common for the majority of traffic to flow in one direction, 
there is also a smaller reverse flow. Common examples include an HTTP
query, which involves query packets flowing in one direction, and a
much larger set of data flowing in the other as a response, or an FTP transaction
which may involve mainly data flow in one direction, but the forward
packets require acknowledgement packets in the reverse direction.  Therefore, 
the model uses the notion of a connection: a two-way exchange of
packets between an \emph{initiator} and a \emph{responder}, corresponding to
the ingress and egress nodes, without necessarily assuming symmetry of the
two-way traffic.

Three parameters were defined as a product of these studies.
The first parameter, the forward traffic proportion $f_{i,j}$ is the
normalised proportion of forward traffic from a connection between
ingress $i$ to egress $j$, measured in packets or bytes and $0 \le
f_{i,j} \le 1$, $\forall i \in \cI$ and $j \in \cE$. The second
parameter $A_i$ describes the activity level of the users at $i$ (the
$A$ stands for `activity').  Finally, some nodes may be chosen for
connection more than others, and thus, $P_j$ (stands for `preference')
denotes the preference for node $j$.

The main assumption of the model
is that the probability that a connection responder belongs to node
$j$ depends on $j$ only. The values of $P_j$ for $j \in \cE$ are
unnormalised. They are divided by the sum $\sum_{k \in \Omega} P_k$ in
order to treat them as the probability a node $j$ is a connection
responder. The parameters $A_i$ and $P_j$ were shown to be
uncorrelated on empirical data, providing some evidence these
parameters describe two very different underlying quantities.

The model is expressed by
\be
X_{i,j} = \frac{f_{i,j}\, A_i\, P_j}{\sum_{k \in \Omega} P_k} 
+ \frac{(1-f_{j,i}) \, A_j\, P_i}{\sum_{k \in \Omega} P_k}.
\label{eq:independent_conn}
\ee

\noindent The first term captures the forward traffic of the
connection between initiator $i$ and responder $j$ while the second
term its reverse traffic, generated by the users from $i$ and $j$
respectively. The model may be viewed as a weighted sum of two gravity
models, with one gravity model characterising the forward traffic,
while the other the reverse traffic. Thus potential asymmetries in
traffic can be accounted for.

The model is sufficiently flexible to accommodate variations. For example, the \emph{simple IC model} modifies one parameter of 
model \autoref{eq:independent_conn} by setting $f_{i,j} = f$, where $f$ is a constant as it has been observed that $f$ is fairly 
stable 
from week to week (at least on the Abilene dataset \cite{Erramilli06IndepConnTech}) simplifying the model considerably. Another 
variation, the \emph{time-varying IC model} includes temporal variation of the parameters, \ie
\ben
X_{i,j}(t) = \frac{f(t)\, A_i(t)\, P_j(t)}{\sum_{k \in \Omega} P_k(t)} + \frac{(1-f(t)) \, A_j(t)\, P_i(t)}{\sum_{k 
\in \Omega} P_k(t)},
\een
and the \emph{stable-$fP$ IC model} removes the time dependency of $f$ and the preferences $\{P_j\}_{j \in \cI}$, while 
the \emph{stable-$f$ IC model} only removes the temporal dependence of $f$. These variations allows trade-offs between the 
degrees of freedom of the model and computational complexity, especially when used for the synthesis or inference of traffic 
matrices. With less parameters, which was shown to be less than the basic gravity model, the model is easier to compute.

The parameters $\{A_i\}_{i\in\cI}$ and $\{P_j\}_{j\in\cE}$ were validated on actual data. Activity levels $\{A_i\}_{i\in\Omega}$ 
possess diurnal patterns, corresponding to user access patterns, and a periodic pattern on a weekly timescale. In particular, 
activity levels are higher on weekdays compared to the weekend, matching observations such as \autoref{fig:abilene_2004_b}. There is 
also a more prominent periodic pattern when considering larger nodes, as this effect is due to aggregation,
as it captures the users with higher activity levels. These observations are consistent with the temporal properties 
discussed of traffic matrices discussed in \autoref{sec:tm}. The model was shown to be effective in estimating
traffic matrices, improving over the basic gravity model by $20\%  -25\%$ for the G\'{E}ANT dataset \cite{GEANT} 
and almost $10\%$ for the Totem dataset \cite{Erramilli06IndepConn,Erramilli06IndepConnTech}. 
These results and observations show that average user behaviour is largely stable and predictable, a great boon to traffic 
modelling development.

In some ways, the ICM is similar to the DCM, in that both models include parameters to describe the
underlying user behaviour, unlike the basic gravity model. For example, both models have a parameter to quantify the level of
attractiveness of one node (connection) to another. The DCM is also able to incorporate features of the ICM as well. The differences end there, however. For one, the ICM has a slightly richer description of the flow connections between nodes, such as the forward and reverse traffic flows
between nodes, whereas the DCM aggregates the information in a single parameter. The ICM seeks 
to capture the behaviour of each connection made, rather than merely model the relationship between nodes, emphasising a 
different focus compared to the DCM. Thus, the ICM may account for hot potato routing and other
asymmetries in traffic flow.

\subsubsection{Low-rank spatial models}

The very noticeable feature of both DCM and ICMs is that they can
better represent traffic matrices, but are more highly
parameterised. It is, in general, possible to fit a data set more
accurately when more parameters are available, but this presents a
difficulty -- does one accept the more complex, more highly
parameterised model, or the simpler, perhaps more robust model?

In the previous cases, this was an ``all or none'' decision (at least
we had to decide on the type of model we used, of not the exact number
of choices involved), whereas the gravity model is fixed in its
parameterisation. However, there are concepts that easily extend the
gravity model.

The low-rank model somewhat new, made popular by its use in matrix
completion problems \cite
{CandesPlan09MtxNoise,CandesRecht08MtxComp,CandesTao09MtxComp,Recht07MtxComp}. Low
rank models assume the traffic matrix is well-represented by the low
rank approximation 
\be X_r = \sum_{i=1}^r \sigma_i^2 \bu_i \bv_i^\T,
\label{eq:low_rank1}
\ee

\noindent where $\sigma_i$ denotes the $i$-th singular value, with all
singular values arranged in order of descending order, \ie~$\sigma_1
\ge \sigma_2 \ge \cdots \ge\sigma_r$. The famous Eckhart-Young theorem
\cite[Theorem 4.32, p.~70]{Stewart98MtxVol1} states this is the best
rank-$r$ approximation, in the sense of the Frobenius norm\footnote{The 
Frobenius norm is the Euclidean norm applied to matrices \ie~
$\|\bX\|_F = \sqrt{\sum_{i,j} x^2_{i,j}}$.}, of a
matrix $\bA$ given by retaining the largest $r$ singular values of
its Singular Value Decomposition (SVD). The theorem, however, assumes
that the target matrix for approximation is already known. In low-rank
matrix recovery, however, the target matrix is unknown.

In the context of traffic matrix modelling, low-rank models are a
relatively recent introduction, beginning with work in
\cite{Zhang09TMCS}, although this model was spatio-temporal
(see further below). Low rank purely spatial models were proposed
and used to good effect by Bharti \etal \cite{Bharti10Invisible}.
The choice of a low rank model has strong empirical backing by the
earlier results of PCA applied to traffic data, for instance in
\cite{Lakhina04TrafficStruct,Zhang05Anomography}.

In essence, we can see \autoref{eq:low_rank1} as expressing a traffic
matrix as a weighted sum of gravity models, \ie~each single rank
component looks exactly the same as that expressed in
\autoref{eq:network_grav_mtx}. It seems a logical approach simply
because the Internet is not a homogenous entity. In particular there
are many types of applications running across the network: from
interactive session, to voice, to HTTP, to streamed video. We might
imagine that a class of traffic, say streaming video, satisfies the
gravity law, but with different row and column sums to, say, voice
traffic. Given this, it seems that a weighted sum of gravity matrices
is a natural extension.

Previous models actually turn out to be special cases of this low-rank
model. The gravity and discrete choice model are spatial rank-1
models. The generalised gravity model and the ICM are 
spatial rank-2 models, the latter of which can be observed
from the summation of the forward and reverse traffic contributions in
equation \autoref{eq:independent_conn}. The low-rank model may be
viewed as a general model for providing a fundamental framework for
further model development.  We shall consider this idea in more detail
below in the context of spatio-temporal modelling.

\subsection{Spatio-Temporal modelling}

Spatio-temporal models aim to describe spatial and temporal structure
jointly. Considering the rich structure traffic matrices have both
spatially and temporally, these models would be more sophisticated
than their purely temporal or spatial counterparts. There has been
relatively little work performed on this type of modelling as yet, but
there is considerable suggestion that it will be fruitful in the
future.

\subsubsection{Low-rank spatio-temporal models}

The main idea in spatio-temporal modelling of traffic matrices so far
has been to exploit the low-rank models mentioned above, but in this
context to apply it to the stacked representation of a series of
traffic matrices denoted here by $\cX$.

Low-rank models assume the traffic matrix is well-represented by the
low-rank approximation 
\be \cX_r = \sum_{i=1}^r \bar\sigma_i^2 \bu_i
\bv_i^\T,
\label{eq:low_rank}
\ee

\noindent where $\bar\sigma_i$ denotes the $i$-th singular value. 
As before, all singular values are arranged in a descending 
order. Note the difference between model \autoref{eq:low_rank} and the above is the 
low rank assumption also applies to the temporal structure of the traffic matrix.

In the context of traffic matrix modelling, low-rank models are a
relatively recent introduction, beginning with work in
\cite{Zhang09TMCS}. Besides spatial correlations (exploited by the
models proposed previously), traffic matrices are known to exhibit
temporal correlations, resulting in a low-rank structure both
spatially and temporally, justifying the rationale behind the
model. The objective of the work is to approximate the time series of
traffic matrices $\cX$ by a rank-$r$ model $\cX_r$.  The model
proposed here is \emph{spatio-temporal}, in contrast to the models
discussed previously, which are only spatial in nature.

Simply insisting on low rank, however, is missing another important point,
which is that matrices also exhibit locality, \ie~elements that are
close in time (where this might mean time of day, not absolute time),
or space exhibit strong correlation.  It turns out that the model
\autoref{eq:low_rank} is greatly enhanced with additional simple
constraints on the temporal and spatial structure to reflect the
smoothness property of Internet traffic, under normal operating
conditions.

The low-rank construction also proved relatively easy to use in
practical applications such as matrix completion, and
\cite{Zhang09TMCS} showed that it could be used to do matrix inference
from link data, impute missing data (from as little as a few
percent of extant values), or be used to predict matrices into the
future.

Despite the demonstration of its effectiveness in traffic matrix
estimation, low-rank models are still not well understood. Unlike the
previous models, where the parameters are, by design, quantitative
measures of an underlying network property, low rankedness (in
spatio-temporal matrices) does not correspond to any particular
network aspect, such as user behaviour. It is just a measure of
the spatio-temporal correlation between traffic flows. It does,
however, hint that OD traffic flows are \emph{clustered}, if one considers 
the allocation of IP prefixes. A better interpretation is
necessary to understand the properties of the model, and work in
\cite{Bharti10Invisible,Lakhina04TrafficStruct} may provide clues in
the right direction.

Furthermore, in the recovery of traffic matrices using the low-rank
model, theoretical work on the minimum number of measurements required
for recovery under structured losses of rows and columns of $\cX$,
which occurs frequently in the networking context, is left open. At
present, the current focus is on random erasures of the elements of
$\cX$
\cite{CandesPlan09MtxNoise,CandesRecht08MtxComp,CandesTao09MtxComp,Recht07MtxComp}.
Overcoming structured losses is far more important than random
erasures, as such a scenario is frequently encountered in real
networks. For example, a router failure may result in missing data for
an entire row of the traffic matrix. The results of \cite{Zhang09TMCS}
show much promise, as the method is largely immune to structured
losses. The challenge now is to construct a theory as to why this is so and
as to what extent structured losses may be recovered.

Low-rank models hold much promise for the development of more
sophisticated models. More work is required to understand the
spatio-temporal properties of the traffic matrix, but as preliminary
results indicate, there is a potentially rich structure to exploit.

\subsubsection{Tensors and hyper-matrices}

A time series of purely spatial traffic matrices is simply a
3-dimensional array, which is sometimes also called a hyper-matrix. 
Such a representation would be a more natural representation
as it would theoretically preserve spatio-temporal properties better
than the stacked matrix, as well as track the evolution of the traffic
demands throughout the measurement interval. A tensor representation of
traffic is even better as it is invariant to changes of basis, unlike hyper-matrices.
The difficulty, however, is identifying the type of decomposition of the tensor that would
produce low-rank structures, or a beneficial, exploitable
structure. There are many proposed methods for tensor decomposition,
but the two most popular are the Canonical Polyadic (CP) or PARAFAC
decomposition and the Tucker or multilinear decomposition
\cite{Kolda09Tensor}. Tensor decomposition requires a large number of
computations, which may be an obstacle to its adoption in traffic
matrix recovery.  At present the one work exploiting the tensor structure
of network traffic to impute missing entries of network traffic tensor
is found in \cite{Acar10Tensor}.

### 2013-07-22 Compiled/chapter8/resources.tex
all files compile correctly

% \section{Resources}
% \label{sec:resources}

\clearpage
\section{Conclusion}
\label{sec:conclusion}

This chapter has been aimed at introducing the reader to the
state-of-the-art in Internet traffic matrix measurement, modelling and
applications. It is not a complete survey of all research about
Internet traffic matrices as such a survey would necessarily
consume a much larger amount of space, and be less digestible, and we
apologise to those whose work has not been referenced.

The chapter has also aimed to clarify a set of common terminology in a
field which has occasionally been confounded by ambiguous or confusing
terms. We have discussed how difficult it is a task to measure and model traffic matrices,
due to the intricate complexities of the layering of network protocols. Moreover, these 
complexities also contribute to challenges in determining the focus a model should take.

Models presented here were classified in three categories, depending on the traffic properties they
aim to capture: purely temporal, purely spatial and
spatio-temporal models. Several examples of popular models 
of varying complexity were discussed: Fourier and wavelet models, PCA, the modified Norros
model, uniform and Gaussian priors, the gravity model and its variants, the independent connections model,
the discrete choice model and low rank models. There are  
many more models, as the list of models presented here is not meant to 
be exhaustive. All models have their strengths and weaknesses, and they may only be
used in circumstances where the assumptions from which they derive from hold. Furthermore,
practitioners should always keep in mind the particular application the model is being used for.

Several applications of traffic matrices and their models were discussed, 
including their recovery, network optimisation and engineering, anomaly detection 
and synthesis. These tasks emphasise the practical usefulness of traffic matrices 
to a network operator.

There are several more open questions, considering the shift in current traffic
trends. Multicast traffic has not been studied in-depth. Content delivery networks
have a profound impact on traffic behaviour, but to our knowledge, has not garnered much
attention with regard to studying their traffic behaviour. Other areas to understand
further are the global traffic properties of traffic matrices, application profiles of the
traffic and traffic matrix synthesis.

Understanding and modelling traffic matrices is a difficult problem, exacerbated by the lack of
publicly available datasets. It is our aim to also provide, as an adjunct to this chapter, links
to the most commonly used datasets in this domain, and code to perform
some of the commonest tasks. In this way, we hope to provide a firm
foundation for future work in the area, and to help those who just
want to use traffic matrices in their research.

### 2013-07-22 Compiled/chapter8/tm.tex
all files compile correctly

\clearpage
\section{Definitions and Notation}
\label{sec:tm}

In this section, a formal approach to networks and traffic matrices is
defined.  Traffic originates from a \emph{source} and is delivered to
a \emph{destination} (or several destinations). The traffic traverses
a set of \emph{links} between some set of sources and destinations.
The links connecting these define the \emph{topology} of the network,
and the paths chosen by traffic flows determine the {\em
  routing}. Traffic may be split across multiple paths by load
balancing, or may keep to a single path.  Often, sources and
destinations are identified with network devices such as switches or
routers, but they can also refer to a location in a logical space
attached to the network, for instance IP addresses of prefix blocks.

Let $\Omega$ denote the non-empty set of all sources and destinations
in a network and let $|\Omega| = N$.  The sources and destinations
often have a physical, geographic location, and so we regard their
indices as {\em spatial variables}, even when they are actually
logical entities, such as Autonomous Systems (ASes), or cannot be
directly identified as network devices, as with IP addresses.

\begin{figure}
  \begin{center}
    \begin{minipage}[m]{0.64\columnwidth}
      \vspace{0pt}
      \includegraphics[width=0.9\columnwidth]{traffic_matrices.pdf}      
    \end{minipage}
    \hspace{-10mm}
    \begin{minipage}[m]{0.3\columnwidth}
      \vspace{0pt}
      $$ X = \displaystyle
              \left( \begin{array}{lllll}
                x_{AA} & {\color{red} x_{AB}} & x_{AC} & \cdots \\
                x_{BA} & x_{BB} & x_{BC} & \cdots \\
                \vdots & \vdots & \vdots & \ddots \\
	      \end{array} \right)
    $$
  \end{minipage}
  \caption{An example of a traffic matrix. Note that often the
    diagonal elements, $x_{AA}, x_{BB}, \ldots$ are zero as this
    traffic does not cross the network, however, in almost as many
    cases it is non-zero because a ``node'' in the graph represents an
    aggregation of devices such as a PoP or an AS. In these cases we
    often do wish to measure these diagonals, even though they may not
    cross the logical links pictured, because they affect traffic
    engineering within the PoP, for example.
  \label{fig:tm} }
  \end{center}
\end{figure}

A traffic matrix is naturally represented by a three dimensional,
nonnegative hyper-matrix $\bX(t)$, with $i,j$-th entry $X_{i,j}(t)$.
Each entry represents the traffic volume, or demand, measured in terms
of bytes or packets, from source $i$ to destination $j$ in time
interval $\lbrack t, t+\Delta t) \subset \cT$, the full measurement
interval being denoted by $\cT$. As an aside, a matrix representation is
useful for the representation of other aspects of the network, for
instance, delay, jitter, loss, bottleneck-bandwidth and distance
\cite{Medina02Taxonomy}, but throughout the chapter, traffic will be
the focus. Whenever the context is clear, for example, when
considering only the spatial structure of the matrix, the time index
$t$ is dropped. An abstract example of the traffic matrix is presented
in \autoref{fig:tm}.

A closely related concept is the {\em demand matrix}, distinct from
the traffic matrix because the former is {\em offered load}, and the
latter {\em carried load} \cite{Feldmann01TMdemand}. 
They may be the same, but may differ where
congestion limits the carried traffic, or rate limiting is used on
some traffic streams. In general we cannot measure offered traffic,
only carried traffic, and so almost all empirical research has
concentrated on traffic matrices, but it is important to note that
many of the assumptions of traffic matrix models are actually
motivated by intuition about demand matrices, and these may not apply
where the two differ substantially\footnote{Many works make no
  distinction between demand and traffic matrices, leading to
  confusion. Here, we shall try to keep the two distinct: when we say
  traffic matrix, we are referring to carried traffic.}.

% The spatial variable $i$ and $j$
% are generally associated with the address structure of the Internet,
% such as Internet Protocol (IP) addresses, or prefixes of these
% addresses.

Large-scale, real-time monitoring of traffic is intractable at
present, thus limiting measurements to the average traffic in a
discrete time interval. Shorter time intervals \ie~small $\Delta t$,
benefit anomaly detection applications, the tradeoff being a
possibility of uncertainty from traffic burstiness at shorter time
scales combined with larger potential measurement or sampling
errors\footnote{Traffic is typically sampled at the backbone network
  level to cope with the tremendous volumes of data that could be
  collected.}. Longer time intervals result in ``smoother'' traffic, 
  averaging out measurement errors. However, this smooths out
real variability in the traffic as well, and can result in meaningless
estimates in the presence of strong non-stationarity. Hence, the
choice of $\Delta t$ depends on the application and available
measurements. Common choices range from 5 minutes to an hour. Further
discussion on the temporal properties of traffic flows is found in
\autoref{sec:models}.

There are two popular definitions of traffic matrices: the
origin--destination (OD) matrix and the ingress--egress (IE) matrix.
\begin{enumerate}

\item \textbf{OD traffic matrix}: this matrix measures traffic from
  true source to destination, \ie~the point that generates a packet
  to the point that receives it. In the Internet, it is perhaps most
  reasonably defined in terms of Internet Protocol (IP) addresses.
  However, if $\Omega$ is defined over the entire IPv4 address space of
  $2^{32}$ addresses (with even more for IPv6), this poses storage and
  computational problems. Moreover, the matrix would be very
  sparse. What's more, protocols such as NAT, HTTP proxies and
  firewalls may obscure the true IP address mappings. One way to
  overcome some of these deficiencies is to aggregate the traffic
  matrix into blocks of IP addresses, frequently using routing
  prefixes. Bharti \etal \cite{Bharti10Invisible} defined the idea of 
  \textit{atomic aggregation}
  to partially circumvent the problem of size of an OD matrix. 
  If the logical indexing chosen is \textit{atomic}, then a 
  non-zero element of an OD traffic matrix implies that all 
  flows between the particular source and destination 
  pair is fully visible to the network operator (see \cite{Bharti10Invisible}
  for details).

\item \textbf{IE traffic matrix}: any single network operator sees
  only a small proportion of the Internet OD matrix. Thus this matrix
  is not just unknown, but unmeasurable (by a single
  operator). Instead, many operators find that using their edge
  routers (or even the edge links) as sources and destinations results
  in a local traffic matrix of great use. We call this the IE traffic
  matrix as the set $\Omega$ includes ingress, \ie~traffic going into
  the network, and egress, \ie~traffic flowing out of the network,
  points as proxies for sources and destinations. A single ingress or
  egress ``node'' may denote a router, a collection of physically
  co-located routers called a Point of Presence (PoP), or some other
  abstract collection of traffic ingress/egress points depending on
  the level of coarseness required in the modelling process. The PoP
  level convention is often adopted as it provides a simple
  visualisation of the network its operators.
\end{enumerate}

IE traffic matrices can be obtained in a number of ways. They can be
formed from OD traffic matrices simply by mapping IP prefixes to
ingress/egress locations in the network, but this assumes knowledge of
all flows traversing the ingress/egress nodes.  Traffic at egress
nodes may be inferred from router data (see the next section) and
measurements of ingress traffic, but typically, the converse is
difficult. Likewise, it is usually difficult to form an OD matrix from
IE matrices.

Consequently, the IE traffic matrix is frequently adopted for network
optimisation applications as it is more practical to measure, and
because in aggregating traffic the OD flows are ``bundled'' together
into locally meaningful groupings. A network may carry flows between
billions of IP address pairs and millions of prefix pairs, but only
thousands of router pairs, and hundreds of PoP pairs.  In this way,
the IE matrix is a more compact representation, but more importantly,
the aggregation of the traffic into large bundles results in a
smoothing effect on the data, reducing the number of independent
parameters that may have to be estimated. At the PoP level, the
aggregation of flows results in averaging out sampling error (similar
to the choice of $\Delta t$ above). This is highly beneficial for
numerical iterative algorithms used to estimate the traffic matrices,
as aggregation leads to better conditioning of the traffic matrix. The
trade-off, unfortunately, is the loss of fine grained data, as one can
no longer observe IP-level flow data, or examine application profiles.

Another consideration worthy of concern in applying traffic matrices
is {\em invariance}.  A good representation of the traffic has
to be invariant to other network aspects, such as routing and the
network topology. For instance, if the traffic matrix for a network
changes in response to changes in link placement, then the matrix is
not terribly useful for network design. IE matrices are subject, for
example, to large changes due to routing shifts, and this means that
they are less useful to operators compared to OD matrices. However,
the practicalities of measurements mean that IE matrices may be all
that is measurable.

% That leads naturally to our next topic: measurement of traffic
% matrices.

% Another example was discussed in \cite{Alderson06Topology} using the G\'{E}ANT \cite{GEANT} network for illustration, where a 
% regional network uses the G\'{E}ANT router to transit traffic within the regional network itself. The result is a net increase of the traffic 
% demands at the nodes being used for this purpose, although these demands themselves do not traverse G\'{E}ANT. Also, distortions 
% may occur in the form of hot potato routing \cite{Paxson97Routing}; see Section \ref{sec:models}. Despite 
% these issues, due to the practicality of the IE traffic matrix, focus will generally be on the IE traffic matrix, unless specified otherwise.

\subsection{Example}

As an example we present the Abilene (Internet2) data from 2004 that
was used in several studies
\cite{Zhang05Anomography,Roughan05GravSynth,Alderson06Topology}.  The
backbone network is located in North America (see
\autoref{fig:abilene_2004_map}).  The traffic data contains averages
over 5 minute intervals, from March 1st to September 11th, 2004
(though there are some missing periods).  The dataset is an
Ingress-Egress, router-router\footnote{In Abilene, at the time, the
  routers almost corresponded to PoPs.} traffic
matrix. \autoref{tab:traffic_matrix} shows one example traffic matrix
from that period, along with row and column sums.

\input{eg_tm.tex}

We can immediately see that the matrix is not symmetric, and that
there are quite a range of values. Also notable, the diagonals are not
zero, as there is some measured traffic that enters Abilene, and then
exits at the same PoP.

A timeseries view of the data is seen in 
\autoref{fig:abilene_2004}. The top figure shows the total traffic
during each time interval for the whole dataset, and the bottom figure
shows the first two weeks of the data, overlaid so as to illustrate
the cyclic nature of the data. We can clearly pick out peak traffic
hours, and weekdays from the weekend. However, the variation between
peak and off-peak loads is perhaps smaller than is typical in
commercial environments.

\begin{figure}[thbp] 
  \begin{center}

    \begin{subfigure}[b]{\oneup}
      \centering
      \includegraphics[width=\textwidth]{abilene_map.eps}
      \vspace{-5mm}
      \caption{Map of Abilene PoPs. \autoref{tab:traffic_matrix} is a
        router to router traffic matrix, but there is one router per
        PoP except in Atlanta, where there is a second router. The
        router numbers in the table are alphabetical.}
      \label{fig:abilene_2004_map}
    \end{subfigure}

    \vspace{3mm}
    \begin{subfigure}[b]{\oneup}
      \centering
      \includegraphics[width=\textwidth]{Abilene_2004_totals.eps}
      \vspace{-9mm}
      \caption{Abilene 5 minute totals of the traffic matrix from
        March 1st to September 11th, 2004.}
      \label{fig:abilene_2004_a}
    \end{subfigure}

    \vspace{3mm}
    \begin{subfigure}[b]{\oneup}
      \centering
      \includegraphics[width=\textwidth]{Abilene_2004_totals_short.eps}
      \vspace{-8mm}
      \caption{The weeks starting March 1st (blue), and March 8th (red).}
      \label{fig:abilene_2004_b} 
    \end{subfigure}

    \vspace{2mm}
    \caption{Abilene, 2004\label{fig:abilene_2004}}
  \end{center}
\end{figure}

We have endeavoured to put this data into an easily read format, and
placed it onto a web server at 
\url{http://www.maths.adelaide.edu.au/matthew.roughan/traffic_matrices.html},
where we will also place other TM datasets as far as possible, for
comparison studies, or as test datasets for education.

The data illustrates a number of properties of the data, perhaps most notably
the fact that many datasets have anomalous ``spikes'' (these are often
real, but may also be artefacts in the data), and periods of missing
data. The question of how these arise then naturally leads to our next
chapter on measuring traffic matrices.

### 2013-07-22 Compiled/chapter9/Content/conclusion.tex
all files compile correctly

\section{Conclusion: Integrating Spantaneous Wireless Networks in the IP Architecture\label{sec:conclusion}}
%\section{Conclusion\label{sec:conclusion}}

This chapter reviewed recent trends towards more collaborative network layer paradigms, accommodating spontaneous wireless networks. The thread followed throughout the chapter is the compatibility, in practice, with standard IP protocols currently at work in today's Internet. Indeed, absent such compatibility, slim are the chances that a given solution would actually be deployed and have a concrete impact. If one cannot just ``reboot'' the Internet to accommodate a convenient fresh start, one can nevertheless drive a continuous evolution of the Internet towards what is needed to allow seamless spontaneous wireless networking. In other words, research in this domain has to not only discover an alternative state in which things would work better, but also discover smooth transitions towards this alternative state, starting from the state we are currently in. The IETF is one of the important venues where such transitions are discussed, evaluated and designed. This chapter thus focused on standards developed by the IETF, which are relevant for spontaneous wireless networks. \ \\ \ \\
%
In principle, spontaneous wireless networking is IP-disruptive: the way in which communication is performed in spontaneous wireless networks challenges some of the fundamental assumptions underlying traditional computer networking and the legacy IP networking architecture. The first part of this chapter has focused on identifying and discussing the impact of spontaneous wireless networking paradigms on layer 3, and has studied an alternative architectural model that could integrate spontaneous wireless networks in the IP networking architecture. \ \\ \ \\
%
Due to their harsh characteristics, spontaneous wireless networks cannot be efficiently managed by standard protocols at layer 3 and above. In particular, legacy routing and flooding mechanisms are unsuitable to efficiently track low bandwidth, asymmetric, time-variant and lossy communication channels, between devices that may be mobile and thus create even more instability in the network topology. The second part of this chapter reviews various advanced techniques have been recently developed in order to accommodate these demanding characteristics: efficient flooding, non-trivial link metrics, neighborhood discovery, jittering techniques, duplicate detection mechanisms. These techniques are employed by several routing protocols developed by the IETF, mainly targeting Mobile Ad Hoc Networks (MANETs) and Low-Power Lossy Networks (LLNs), two categories of spontaneous wireless networks.\ \\ \ \\
%
Taking a step back, it is perhaps worthy to observe that there are essentially four categories of solutions to deal with IP-disruptive characteristics \cite{EB-HABILITATION2013}: \ \\ \ \\
%
{\bf Adaptation layer developments}. This type of solution proposes to design intermediate layers, which interface between two of the legacy layers, {\em i.e.} from bottom up: (1) the physical layer, (2) the MAC layer, (3) the network layer, (4) the transport layer and (5) the application layer. Such approaches enable interoperability with legacy software by providing a black-box which emulates an appropriate behavior, compatible with upper layers, operating on top of disruptive lower layers. The system that results from such an approach is thus significantly more complex than the legacy system, in that it introduces a whole new ``world'' of protocols in addition to the legacy protocols. However, this approach can be effective in practice: a current example is 6LoWPAN \cite{6LOWPAN-WG}, which designed a series of mechanisms at layer 2.5 ({\em i.e.} sitting between layer 2 and 3), enabling the operation of standard IP protocols at layers 3 and above on the IEEE 802.15.4 MAC layer.\ \\ \ \\
%
{\bf Intra-layer optimizations}. This type of solution proposes to modify or replace specific protocols currently in use within a legacy layer, to cope with IP-disruptive characteristics from lower layers. Most of the efforts that are mentioned in this chapter fall in this category. There are however limits to what one can achieve when taking this approach: it is unlikely that one can achieve game-changing innovation if one is allowed to replace only a single, small part of the whole system. Yet other types of solutions have thus been proposed, described in the following. \ \\ \ \\
%
{\bf Cross-layer optimizations}. This type of solution proposes to partially or totally abolish the distinction between two or more legacy layers, to produce a new system that performs significantly better, thanks to new protocols that can leverage cross-layer information to better cope with IP-disruptive lower layers. One example of such an approach is the XPRESS cross-layer stack \cite{laufer11}, which collapses transport, network, and MAC layers and uses backpressure to provide better performance in wireless mesh networks. Cross-layer approaches are probably the most disruptive type of approaches, as their deployability and interoperability with standard legacy software is in general difficult to assess if at all possible -- lack of interoperability is often the price to pay for radical performance improvements. There is however yet another type of solution proposing drastic changes while maintaining interoperability with legacy layers, as described below.\ \\ \ \\
%
{\bf Top layer developments}. This type of solution aims at building a radically new system sitting on top of the legacy protocol stack, at the application layer. Essentially, such an approach considers the Internet as a black box providing a service equivalent to a cable connecting source(s) and destination(s), and provides novel mechanisms efficiently using this cable to cope with IP-disruptive characteristics. One example of such construction is the experimental Delay Tolerant Networking (DTN) architecture developed by the Internet Research Task Force (IRTF) \cite{DTNRG} \cite{ietf:rfc5050} \cite{ietf:rfc5326}, including a specific routing protocol targeting DTNs \cite{ietf:rfc6693}.\ \\ \ \\
%
%
It can be anticipated that innovative networking paradigms will continue to appear in the future, providing improvements at the price of IP-disruptive characteristics. However, in order to deploy or advance towards these new paradigms, one of the above approaches will have to be employed.

### 2013-07-22 Compiled/chapter9/Content/flood_rtg.tex
all files compile correctly

%\section{Flooding and Routing Mechanisms over Wireless Ad Hoc Networks}
\section{Flooding and Routing in Spontaneous Wireless Networks}
\label{s:flood_rtg}

% Note: At some point it should be clear that we are assuming a SINGLE wireless interface per router.
% Note: Provide consistent notation message/packet throughout the chapter.

As described in section \ref{s:comm_wless}, there are important differences in the way that spontaneous wireless networks enable communication between nodes, with respect to the classic fixed/wired networks. These differences have a significant impact in the mechanisms and protocols used in wireless multi-hop scenarios to disseminate information through the network ({\em flooding}) and find and maintain paths between pairs of computers in the network ({\em routing}). This section examines several mechanisms that are used in different routing protocols, discusses the issues and problems that these mechanisms have when operating in spontaneous wireless networks, and describes some techniques to fix or overcome these issues.

Section \ref{ss:nd} explores the use of neighbor discovery procedures in spontaneous wireless networks. Section \ref{ss:flood} describes techniques to perform efficient flooding over such networks. Finally, section \ref{ss:metrics} presents the problem of estimating link costs and using them to to identify ``good routes" over the network.

\subsection{Neighborhood Discovery}
\label{ss:nd}

In many routing and flooding protocols, routers need to be aware of their own neighborhood. This is particularly important (although not only) in spontaneous wireless networks, in which the neighborhood may change frequently during network operation. Routers acquire knowledge about their neighborhood by way of a {\bf neighborhood discovery} mechanism.

\begin{definition}[Neighborhood Discovery]
(ND) is the process whereby each router advertises all the routers to which direct communication is possible ({\em i.e.}, the routers to which there are network links) about its presence in the network. This way, routers receiving such advertisements from other (neighboring) routers gain insight on their own neighborhood.
\end{definition}

%\begin{remark}
%About "neighbor".
%Messages in which a router advertises its presence to directly-reachable routers are typically denominated {\bf Hello messages}.
%\end{remark}

\paragraph{1-hop and 2-hop neighborhood}

Depending on the information advertised by ND messages, receiving routers learn different aspects about their neighborhood. If messages only advertise the presence of the originating router, the receiving router will acquire information about the routers that maintain links to itself. If links are bi-directional (as IP links in standard IP networks), this is sufficient for enabling bi-directional communication between routers: a router receiving a ND message from a neighbor can exchange packets in both directions with it. This is the case of traditional neighbor discovery protocols for Internet, such as the \textit{Neighbor Discovery Protocol} (NDP) for IPv6 \cite{rfc4861}, which assumes that all the links are bi-directional and is used to actively keep track of which neighbors are reachable. \ \\ \ \\
%
%Neighborhood discovery is the process whereby each router discovers the routers which are in direct communication range of itself (1-hop neighbors), and detects with which of these it can establish bi-directional communication. For traditional neighbor discovery protocol for Internet, such as \textit{Neighbor Discovery in IPv6} \cite{rfc4861}, assumes that all the links are bi-directional, and is used to actively keep track of which neighbors are reachable. 
%
In spontaneous wireless networks, bi-directional communication availability cannot be inferred from the reception of an ND advertisement, given the fact that asymmetric links are possible (section \ref{ss:issues}). This is taken into account in ND protocols for spontaneous wireless networks. In these protocols, ND advertisements (typically denominated {\bf Hello messages}), contain not only the id of the originating router, but also the list of its current neighbors ({\em i.e.}, routers from which the originating router has received Hello messages). This enables every router in the network to detect the (1-hop) neighbors with which bi-directional communication is possible, and identify the routers that belong to its 2-hop neighborhood -- that is, the set of routers that are 2-hop neighbors or ``neighbors of its neighbors''.

\begin{itemize}
\item {\bf Example} Assume that router $A$ receives a Hello from a neighbor $B$, in which $B$ indicates to have recently received a Hello from $A$; then $A$ learns that link $A$-$B$ is symmetric. As $B$ lists identifiers of all its 1-hop neighbors in its Hello, $A$ learns its 2-hop neighbors through this process.
\end{itemize}

Together with 1-hop neighbors, additional information may be included in Hello messages -- in particular, the cost of the links towards the listed neighbors, when metrics other than hop-count (see section \ref{ss:metrics}) is used. Exchange of Hello messages is typically done periodically, although some events may trigger non-periodic Hellos ({\em e.g.}, changes in the topology). \ \\ \ \\ %({\em e.g.} ETX \cite{etx}, based on the number of expected transmissions to send successfully a packet over a link, or EDR \cite{edr}, that estimates the data rate at which packets can be sent) is used.
%
The {\em Mobile Ad Hoc Network Neighborhood Discovery Protocol} (NHDP, RFC 6130) \cite{NHDP-RFC6130} is the main ND protocol for spontaneous wireless networks. It is used as auxiliary protocol by other routing protocols that need neighborhood information to take their decisions, such as OLSRv2 (see section \ref{sec:olsr}). In NHDP, Hello messages are exchanged periodically and they contain the id of the originating router and the list of its 1-hop neighbors. The IETF also standardized an optimization of NDP for IPv6 over Low-Power Wireless Personal Area Networks (6LoWPANs). This optimization, specified in RFC 6775, adapts the operation of NDP to the lossy conditions of communications and the low-power device constraints of LoWPANs. This is done, for instance, by avoiding unsollicited messages (such as periodic router announcements), reducing the use of multicast for address resolution, limiting the duplicate address detection checks, or enabling better compression algorithms \cite{rfc6775}. \ \\ \ \\
%
%lossy, low-power, low-bit-rate, short-range; with many nodes saving energy with long sleep periods. Multicast as used in IPv6 Neighbor Discovery (ND) [RFC4861] is not desirable in such a wireless low-power and lossy network. Moreover, LoWPAN links are asymmetric and non-transitive in nature. 
%
Other routing protocols include their own Hello mechanism. This is the case of AODV \cite{AODV-RFC3561} (see section \ref{ss:aodv}) or OSPF and its MANET extensions \cite{rfc5449,rfc5614,rfc5820} (see section \ref{ss:ospfmanet}). In the last case, some approches have been explored in order to avoid redundant notifications and hence reduce control traffic by only reporting changes in the neighborhood occurred since the last Hello transmission: this principle leads to the {\em incremental Hellos} mechanism used in the Overlapping Relays extension of OSPF (OR/SP \cite{rfc5820}) and the {\em differential Hellos} mechanism used in the MANET Designated Routers extension of OSPF (OSPF-MDR \cite{rfc5614}). However, experiments show that the potential benefits (mostly, saved amount of traffic) of these two mechanisms are not significant, in particular when compared with the additional complexity they introduce in the corresponding protocols \cite{aircc}.

%\paragraph{ND Protocols and Optimizations for Wireless Multi-Hop Ad hoc Networks}

%The Mobile Ad Hoc Network Neighborhood Discovery Protocol (NHDP) \cite{NHDP-RFC6130} is the main ND protocol for wireless multi-hop ad hoc networks. It is used as auxiliary protocol by other routing protocols that need neighborhood information to take their decisions, such as OLSRv2. In NHDP, periodic Hello messages are exchanged locally so that each router can determine its 1-hop and symmetric 2-hop neighbors. The Hello message carries identifier of the originator, and its 1-hop neighbor information. Through Hello message exchanges, the router can determine the status of the link -- and in particular, detect whether bi-directional communication is possible: assume that router $A$ receives a Hello from a neighbor $B$, in which $B$ indicates to have recently received a Hello from $A$; then $A$ learns that link $A$-$B$ is symmetric. As $B$ lists identifiers of all its 1-hop neighbors in its Hello, $A$ learns its 2-hop neighbors through this process.

%Other routing protocols include their own Hello mechanism. This is the case of AODV \cite{AODV-RFC3561} (see section \ref{ss:aodv}) or OSPF and its MANET extensions \cite{rfc5449,rfc5614,rfc5820} (see section \ref{ss:ospfmanet}). In the last case, some optimizations have been explored in order to avoid redundant notifications by only reporting changes in the neighborhood occurred since the last Hello transmission: this principle leads to the  {\em incremental Hellos} mechanism used in the Overlapping Relays extension of OSPF (OR/SP \cite{rfc5820}) and the {\em differential Hellos} mechanism used in the MANET Designated Routers extension of OSPF (OSPF-MDR \cite{rfc5614}). However, experiments show that the potential benefits of these two mechanisms are not very relevant, in particular when compared with the additional complexity they introduce in the corresponding protocols \cite{aircc}.

 %Each router sends Hellos, listing the identifiers of all the routers from which it has recently received a Hello, as well as the ``status'' of the link (heard, veriﬁed bi-directional).

%The purpose of the Hello message is to broadcast the local node topology information to the one-hop neighbors to serve the task such as link sensing and neighbor detection.

\subsection{Flooding}
\label{ss:flood}
%
Flooding is the process through which information originated in one router is disseminated across the network, so that it can be received by every other router in the network. \ \\ \ \\
%
The most obvious procedure to perform flooding from a router in a conventional IP network consists of the {\bf pure flooding} procedure:

\begin{enumerate}
\item The source router sends the message through all its network interfaces.
\item Every router that receives the message {\em for the first time} retransmits it over all the network interfaces {\em except the one over which it was received}.
\end{enumerate}

The fact that each router retransmits only {\em once} ensures that the process terminates in a finite number of steps. The fact that {\em all} routers receiving the message retransmit it ensures that the message is received --if there are no packet losses-- by every router in the network at least once. \ \\ \ \\
%
In a spontaneous wireless network, as routers communicate with {\em all} their wireless neighbors by way of a {\em single} wireless interface (see section \ref{s:comm_wless}), the straightforward usage of this mechanism implies that the source router broadcasts the message to be flooded and the neighboring routers rebroadcast it over the same interface it was received. It is known \cite{broadcaststorm} that such a naive approach is not efficient and does not scale in a wireless multi-hop scenario. Three reasons can be highlighted:

\begin{enumerate}[a)]
\item excessive retransmissions that reduce the available bandwidth, 
\item systematic packet collisions due to concurrent transmissions of wireless interfaces (partly) sharing the same wireless channel, and 
\item duplicate packets reception due to the fact that the packet is received and retransmitted over the same interface (and, therefore, transmitted twice in the intersection between the coverage area of the sender and the receiver interface). 
\end{enumerate}

\begin{remark}
Although the three effects are closely inter-related, and all are due to the bandwidth scarcity and the semi-broadcast properties of wireless communication detailed in section \ref{s:comm_wless}, it is important to point out that they constitute {\em different} effects; solving one of them does not necessarily solve the others. 
\end{remark}

\paragraph{Excessive Number of Retransmissions and Efficient Flooding}

In a spontaneous wireless network, a single transmission from a wireless interface is received by the wireless interfaces of all the neighbors within its coverage area. If all routers retransmit the same message as they receive it, this is likely to cause a significant number of {\em redundant transmissions} -- {\em i.e.}, transmissions that do not bring new information for {\em any} of the interfaces receiving them. \ \\ \ \\
%
Consider the situation in Figure \ref{f:excessive}, in which node $A$ (in the center) floods a message to all its neighbors, and they in turn retransmit the same message so that it is received by all the 2-hop neighbors of $A$. In theory, every 2-hop neighbor has received the message -- possibly several times. The redundant retransmissions do not bring new information, but increase the probability of collisions. This effect becomes more relevant in a context of bandwidth scarcity and high network router density. %, as node $C_2$ which receives it from $B_1$, $B_2$ and $B_3$. Some retransmissions of $A$'s 1-hop neighbors are {\em redundant}, for instance $B_2$: if $B_2$ had not retransmitted the message, all the 2-hop neighbors that received $B_2$'s retransmission ($C_1$, $C_2$ and $C_3$) would have received the message via $B_1$ and $B_3$. In practice, the situation is worse: as several transmissions are being performed at the same time by close nodes that share (at least partly) the same wireless channel, the probability of packet collisions is high: several 2-hop neighbors "theoretically" covered by different 1-hop neighbors may be in practice unable to receive and decode successfully the message from $A$, due to collisions between retransmitters \cite{broadcaststorm}.

\begin{figure}[h]
\centering
\includegraphics[width=0.4\textwidth]{Figures/purebc-crop.pdf}
\caption{Classical flooding in spontaneous wireless networks.}
\label{f:excessive}
\end{figure}

{\bf Efficient flooding} techniques explore different strategies to reduce the number of flooding retransmissions (and therefore, to decrease the amount of traffic overhead involved), while preserving as much as possible the ability of the flooding procedure to reach all (or most of) the routers in the spontaneous wireless network. \ \\ \ \\
%
For every efficient routing technique, a set of the routers that receive a message (typically, not all of them) are allowed to retransmit it. If efficient flooding reaches all routers in the network, the set of routers allowed to retransmit a message is a {\em Dominating Set} (DS) in the network graph. As only one router originates and originally sends the message, and every other forwarding router has previously received it via flooding, the set of routers and the wireless links between them usually form a {\bf Connected Dominating Set} (CDS). Given a graph $G=(V,E)$ representing a spontaneous wireless network, where $V$ is the set of vertices (representing network routers) and $E$ is the set of edges (representing network links), a Connected Dominating Set of $G$ is a subset of vertices $D \subseteq V$ with two properties:

%, assuming that the message is originally sent by only one router and every other forwarding router has received the message previously

\begin{enumerate}
\item {\em Connection}. $D$ induces a connected subgraph of $G$, that is, any node in $D$ can reach any other node in $D$ by a path that stays entirely within $D$. 
	\begin{align*}
	\forall x, y \in V &,& \exists p=(x, p_1, p_2, ..., p_n, y) \subseteq D \wedge \bar{xp_1}, \bar{p_1p_2}, ..., \bar{p_ny} \in E
	\end{align*}
\item {\em Domination}. $D$ is a dominating set of $G$, meaning that every vertex in $G$ either belongs to $D$ or is adjacent to a vertex in $D$.
	\begin{align*} 
	\forall v \in V &,& v \in D \vee ( \exists w \in D : \bar{uw} \in E )
	\end{align*}
\end{enumerate}

%The rules used for determining which routers are selected as {\em flooding relays} lead to different techniques \cite{broadcaststorm}.

Figure \ref{f:ex_cds} displays an example of Connected Dominating Set over a network graph.

\begin{figure}[htb]	% H-must be here or [htb]
\centering
\includegraphics[width=0.6\textwidth]{Figures/excds-crop.pdf} %	** if .eps or .pdf, don't need extension}
\caption{Example of CDS (thick edges) over a network graph of 30 nodes (light edges represent communication between nodes).}
\label{f:ex_cds}
\end{figure}

The notion of CDS is useful for efficient flooding purposes: several efficient flooding techniques rely on the construction and maintenance of Connected Dominating Sets of forwarding nodes, over which packets are flooded through the network. One of the main techniques based on this principle is the {\em Multi-Point Relaying} (MPR) technique.

\begin{itemize}
\item {\bf Multi-Point Relays} \cite{mpr-hicss2002} is an algorithm through which a node selects a subset of its 1-hop neighbors (\emph{multi-point relays}) such that each 2-hop neighbor is reachable through (at least) one of the selected 1-hop neighbors ({\em MPR coverage criterion}). MPR selection requires that the selecting node knows the 2-hop neighbors that will be covered by its MPRs. By using MPR, the retransmission of flooding traffic can be significantly reduced, as shown in Figure \ref{f:mpr}, compared to classical flooding in Figure \ref{f:excessive}.

\begin{figure}[h]
\centering
\includegraphics[width=0.65\textwidth]{Figures/mpr-crop.pdf}
\caption{Efficient flooding with Multi-Point Relays.}
\label{f:mpr}
\end{figure}

	\begin{remark}
	Note that the MPR coverage criterion does not guarantee by itself that the set of nodes selected as MPRs form a Connected Dominating Set \cite{softcom10}. The set of MPRs is by definition a dominating set, as every node is either a MPR for a neighbor, or is adjacent to its own MPR. As the heuristic for MPR selection is relative to the source, however, the subgraph resulting from MPRs and the MPR links (links connecting nodes with their relays) between them {\em is not necessarily connected}. This can be easily fixed by adding an arbitrary router and all its links to the subgraph, as proved in Cordero (2010) \cite{softcom10}.
	\end{remark}

%\begin{itemize}
%\item We present MPR: MPR, by which each router is able to, efficiently, conduct network-wide broadcasts. Each router designates, from among its bi-directional neighbors, a subset (MPR set) such that a message transmitted by the router and relayed by the MPR set is received by all its 2-hop neighbors
%\end{itemize}

Multipoint relays of a router can be used to perform efficient flooding, but the principle can be used for performing other networking operations. The MPR selection algorithm can be slightly modified so that the overlay that includes all links between routers and its (modified) multi-point relays is sufficient to compute shortest paths over the underlying network \cite{softcom10}. This result has been exploited in some extensions of OSPF for MANETs, as shown in section \ref{s:wospf}.

\end{itemize}

%The idea behind different kind of efficient flooding techniques is to find a reduced CDS. One of these techniques is the {\bf Multi-Point Relaying} (MPR) technique.

\paragraph{Systematic Packet Collisions and Jittering Techniques}

Consider the spontaneous wireless network of Figure \ref{f:syspcol}, in which router $A$ floods a message through the network. The broadcast transmission of $A$ is received at the same time by $B$ and $C$, which retransmit the message towards $E$ and $D$ (in the case of $B$) and $D$ and $F$ (in the case of $C$). Then, concurrent retransmissions from $B$ and $C$ cause a systematic packet collision from $D$'s perspective.

\begin{remark}
In this example, the collision could not be detected with any CSMA\footnote{Carrier Sense Multiple Access. CSMA is a medium access control (MAC) protocol for wireless networks in which nodes sense the medium before transmitting, and only transmit if the sensed medium is idle, that is, if the node does not detect any ongoing transmission within its reception range. See {\em e.g.} Tanenbaum {\em et al.} \cite{tanenbaum} for reference.} layer 2 mechanism neither by $B$ nor by $E$, due to the fact that $B$ and $E$ are not neighbors to each other. $B$ is a {\em hidden node} for $E$ (and vice versa).
\end{remark}

\begin{remark}
Note that this problem cannot be addressed only by way of {\em efficient flooding} approaches: none of the retransmissions by $B$ and $E$ are redundant, so none of them could be avoided without leaving nodes uncovered ($C$ if $B$ does not retransmit, $F$ if $E$ does not retransmit). 
\end{remark}

\begin{figure}
\centering
\includegraphics[width=0.2\textwidth]{Figures/abcdef-crop.pdf}
\caption{An example of spontaneous wireless network. {\em (Broken lines denote direct communication)}}
\label{f:syspcol}
\end{figure}

The fact that flooded messages are forwarded simultaneously by wireless interfaces receiving them through the same wireless shared medium may cause packet collisions during the flooding procedure (depending on the network topology at the time of flooding). Unlike other packet transmissions for which the collision probability may vary depending on the traffic pattern, these flooding collisions are {\em systematic} and will occur, for a given network topology and flooding algorithm, any time that the source node floods a new message (in the example, any time $A$ floods a message). \ \\ \ \\
%
This effect could be alleviated by allowing routers to wait a random amount of time (denominated {\bf jitter}) before retrasmitting a flooded message, in order to reduce the probability of concurrent transmissions by neighboring wireless interfaces. This technique is known as {\bf jittering}, and has been standardized by the IETF in RFC 5148 \cite{jitter-RFC5148}. The recommendation from RFC 5148 is that delays are selected following an uniform distribution between $0$ and a maximum jitter value, $MAX\_JITTER$. Figure \ref{f:jitter} illustrates the effect of jittering techniques in the network example of Figure \ref{f:syspcol}. In the example, node \textit{A} is flooding a packet to all the other nodes. When node \textit{B} and \textit{C} receive the packet from \textit{A}, instead of retransmitting the packet immediately, they wait a random delay. In this way, simultaneous transmission of $B$ and $C$ (which can cause collision at $D$ in this case) can be avoided. \ \\ \ \\
%
\begin{figure}[htb]
\centering
\includegraphics[width=0.6\textwidth]{Figures/jitter4flooding-crop.pdf}
\caption{Use of jitter for flooding. Node \textit{A} is flooding a packet in a network. Node \textit{B} and \textit{C} wait a random delay before the packet is retransmitted. The dashed overlapping arrows represent the packet collision that would occur if no jitter were used.}
\label{f:jitter}
\end{figure}
%
Although jittering can be theorically implemented at different layers of the protocol stack, it has been shown that its use in layers upper than L3 brings little benefit \cite{jitter-broadcast}. As the problem of systematic collisions affects every L3 routing protocol using wireless flooding, jittering techniques can be implemented by different protocols. The addition of random delays in flooded packets impacts differently in proactive and reactive protocols, given the different use of flooding in both routing strategies. Other jittering effects are due to the specifities of the techniques employed in each case. In proactive protocols where jitter is used (as described in RFC 5148 \cite{jitter-RFC5148}, which recommends to introduce random delays and piggyback all pending messages when a transmission is scheduled), such as OLSR or the OSPF extensions for MANETs, jittering leads to longer LSA messages -- this may cause additional packet collisions, if jitter values are not configured properly \cite{INFOCOM-jitter,jitter-wpc}. In reactive protocols such as AODV, where jitter is used for Route Request (RREQ) flooding, the addition of random delays may lead to suboptimal path selections, which can be minimized by adapting the random distribution used for determining jitter values \cite{Infocom12, wiopt}.

\paragraph{Duplicate Packets and Detection Techniques}

The reception of duplicate packets is a common situation in wireless flooding, due to the fact that flooded messages are retransmitted by forwarding nodes over the same wireless interface in which they were received. Consider the situation of Figure \ref{f:dpd}: router \textit{N2} is retransmitting a broadcast packet received from router \textit{N1} on the same interface as the one over which it was received, so as to ensure receipt also by router \textit{N3}, causing router \textit{N1} to receive the packet a second time.

\begin{figure}[ht]
\centering
\includegraphics[width=0.4\textwidth]{Figures/dpd.pdf}
\caption{The need for duplicate detection: retransmission over the same interface as a packet was received.}
\label{f:dpd}
\end{figure}

Depending of the protocol and the use it makes of flooding and of flooded packets, the way to detect duplicate packets might be different. In link-state routing protocols such as OLSR or OSPF, for instance, flooded messages are link-state advertisements (LSAs) that are stored locally before being retransmitted, in case they bring fresh topology information; in this case, a duplicate LSA can be easily recognized by checking whether the LSA is already installed in the local Link-State Database (LSDB). If flooded messages are not stored locally, the protocol needs to store state for every forwarded message in order to detect a duplicate -- this is, for instance, the strategy of the Simplified Multicast Forwarding (SMF) protocol \cite{rfc6621}. In case a received message was already received and forwarded, it is dropped.

%\subsection{Synchronization}
\subsection{Link Metrics}
\label{ss:metrics}

\glslink{metric}{Metrics} are used to evaluate the cost of a link or a path (set of links), so that a routing protocol is able to determine whether a path or a link should be preferred over another. \ \\ \ \\
%
The simplest link \glslink{metric}{metric} is the {\bf hop-count}: a link has \glslink{metric}{metric} 1 if it is available, 0 otherwise. Typically, in the early versions of routing protocols for spontaneous wireless networks (e.g., AODV, OLSRv1 for mobile ad hoc networks), only the hop-count \glslink{metric}{metric} is used. This way, the \glslink{metric}{metric} or cost for a path is equal to the number of hops involved. \ \\ \ \\
%
When a route between two hosts is being calculated under the hop-count metric, paths with less number of hops are preferred to paths with more number of hops. However, using only minimum hop routes in spontaneous wireless networks may result in suboptimal routing in practice, as the minimum-hop routes are not necessarily the best ones \cite{couto03, wang95}. \ \\ \ \\
%
Figure \ref{f:metric} give an example showing the limit of hop-count metric. The minimum hop route from node \textit{A} to \textit{B} is $\{A,X,B\}$. However, the links $\bar{AX}$ and $\bar{XB}$ are poor with high loss rate (but still able to deliver packets), and $\bar{AY}$, $\bar{YZ}$, $\bar{ZB}$ are reliable links. In this case, $\{A,Y,Z,B\}$ is preferred to the route with minimum hop count. \ \\ \ \\
%
\begin{figure}
\centering
\includegraphics[width=0.25\textwidth]{Figures/axb-crop.pdf}
\caption{An example of different link metric. $\bar{AX}$, $\bar{XB}$ are unreliable links; $\bar{AY}$, $\bar{YZ}$, $\bar{ZB}$ are reliable links.}
\label{f:metric}
\end{figure}
%
Because of the limitation of hop-count metric, new metrics need to be defined. The link metrics used in spontaneous wireless network are expected to have following properties \cite{metrics}:

\begin{itemize}
\item {\bf Dimensionless}. The \glslink{metric}{metric} may correspond to specific physical information, but this knowledge is not used by the routing protocol. 
\item {\bf Additiveness}. The \glslink{metric}{metric} of a route is the sum of the metrics of the links forming that route. It also requires a \glslink{metric}{metric} where a low value of a link \glslink{metric}{metric} indicates a "good" link a high value of a link \glslink{metric}{metric} indicates a ``bad" link. 
\item {\bf Directionality}. The \glslink{metric}{metric} from a router $A$ to router $B$ does not need be the same as the \glslink{metric}{metric} from $B$ to $A$. This is a direct consequence of the fact that wireless links are not bi-directional. 
\end{itemize}

The kind of \glslink{metric}{metric} used in a network depends on the link/physical layer protocol used, the type of information that is available from lower layers, the application requirements, etc. Some examples of link \glslink{metric}{metric} include delay, packet loss probability (Expected Transmission Count, ETX \cite{etx}, that estimates the average number of transmissions before success over a link), queue length (at the receiver) or data rates (Expected Data Rate, EDR \cite{edr}; this \glslink{metric}{metric} is not additive, thus a mapping that inverts its ordering must be applied).

%%\begin{itemize}
%%\item Delay. 
%%\item Packet loss probability. 
%%\item Queue length. 
%%\item Data rates. It is not additive, thus a mapping that inverts its ordering must be applied. 
%%\end{itemize}

### 2013-07-22 Compiled/chapter9/Content/glossary.tex
all files compile correctly

\newglossaryentry{link}
{
  name=Network link between nodes,
  description={There is a {\em link} between two nodes $A$ and $B$, denoted by $A \longrightarrow B$, if and only if $A$ is able to transmit data to $B$ and $B$ is able to receive such data, without intervention of any other node. Nodes are connected to these network links by way of \glslink{interface}{network interfaces}.}
}

\newglossaryentry{interface}
{
  name=Network interface,
  description={A {\em network interface} of a node is a device that provides access from that node to a network link through an underlying physical communication channel.}
}

\newglossaryentry{neighbor}
{
  name=Neighbor,
  description={Two nodes are {\em neighbors} if they can directly communicate. More generally, two nodes are {\em i-hop neighbors} if they can communicate in $i$ hops.}
}

\newglossaryentry{routing-table}
{
  name=Routing table,
  description={The {\em routing table of a router} is a local database that maps a destination in the network to the network interface through which a packet sent to that destination should be forwarded, and indicates the cost of sending the packet through the corresponding link, according to the \glslink{metric} in use. Information in a routing table is collected and distributed by way of a \glslink{routing-protocol}{routing protocol}.}
}

\newglossaryentry{routing-protocol}
{
  name=Routing protocol,
  description={A {\em routing protocol} is a set of procedures performed over the network in order to collect routes and maintain the routing tables of the routers in the network. These procedures enable nodes to transmit and successfully deliver packets to desired destinations in the network.}
}

\newglossaryentry{computer-network}
{
  name=Computer network,
  description={A {\em computer network} is a set of network links and the computers (hosts and routers) attached to any of these links.}
}

\newglossaryentry{host}
{
  name=Host,
  description={A {\em host} (or {\em end system}) is a node in the network able to be source or final destination or network traffic, but is not able to forward packets from one link to another.}
}

\newglossaryentry{end-system}
{
  name=End system,
  description={See {\em host}.}
}

\newglossaryentry{router}
{
  name=Router,
  description={A {\em router} (or {\em intermediate system}) is a node that is able to forward packets from one network link to another. Routers take forwarding decisions based on the information they store in the local \glslink{routing-table}{routing table}.}
}

\newglossaryentry{intermediate-system}
{
  name=Intermediate system,
  description={See {\em router}.}
}

\newglossaryentry{metric}
{
  name=Link metric,
  description={Under a \glslink{routing-protocol}{routing protocol}, a {\em link metric} is a map that matches every \glslink{link}{link} in the network with an estimation of the cost of sending packets over that link. The most trivial and most widely used link metric in wireless multi-hop networks is {\bf hop-count}: all available links are assigned a value $1$. Examples of other link metrics are the Expected Transmission Count (ETX) \cite{etx}, based on packet loss probability; the Expected Data Rate (EDR) or other estimations based on delay or available bandwidth.}
}

\newglossaryentry{ip-link}
{
  name=IP link,
  description={Two \glslink{interface}{network interfaces}, $x$ and $y$, are connected to the same {\em IP link} when they can exchange packets in an IP network without requiring that any router forwards them, that is, when packets sent from one interface are received in the other with the same TTL/hop-limit value. This relationship is denoted as $x \sim_{IP} y$.}
}

\newglossaryentry{autonomous-system}
{
  name=Autonomous System,
  description={``An \emph{Autonomous System (AS)} is a connected group of one or more IP prefixes [internetwork] run by one or more network operators which has a SINGLE and CLEARLY DEFINED routing policy" \cite{rfc1930}, the term ``routing policy" denoting the way that routing information is exchanged between (but not within) Autonomous Systems. In the interior of an AS, ``routers may use one or more interior routing protocols, and sometimes several sets of metrics" \cite{rfc1812}.}
}

%\newglossaryentry{neighbor-discovery}
%{
%  name=Neighborhood discovery,
%  description={{\em Neighborhood discovery} (ND) is the process whereby each router advertises all the routers to which direct communication is possible ({\em i.e.}, the routers to which there are network links) about its presence in the network. This way, routers receiving such advertisements from other (neighboring) routers gain insight on their own neighborhood.}
%}

### 2013-07-22 Compiled/chapter9/Content/introduction.tex
all files compile correctly

\section{Introduction}
%
Since the end of the 20th century, wireless networking is experiencing explosive growth, driven by the popularity of wireless telephony on one hand, and by the development of wireless computer networks on the other hand. Both trends are currently merging into a single attempt: enabling massive wireless Internet access. This phenomenon was inspired by Norman Abramson's pioneer work on packet radio networks \cite{ABRAMSON-70} in the 1970s, and made possible by the authorization of wireless spectrum use for civil telecommunication purposes, in the 1980s\footnote{ISM (Industrial, Scientific, Medical) bands, released in 1985 by US Federal Communications Commission (FCC) for unlicensed use.}. At first, this deregulation encouraged the democratization of wireless telephony, in the 1990s, thanks to the availability of cheaper, more efficient hardware stemming from Cold War military industry efforts. Since 2000, the introduction of new wireless communication standards using the spectrum authorized for civil use has also fueled the development of wireless computer networks and wireless Internet access.
%
\subsection{Managed Wireless Networks}
%
Wireless Internet access is nowadays mostly provided via link layer technologies such as Wifi (IEEE 802.11 infrastructure mode standards \cite{IEEE-802-11}), WiMAX\footnote{Worldwide Interoperability for Microwave Access.} (IEEE 802.16 \cite{IEEE-802-16}), UMTS\footnote{Universal Mobile Telecommunications System.} or LTE\footnote{Long Term Evolution.} (3GPP standards \cite{3GPP}), on user terminals such as smartphones, tablets, laptops, {\em etc}. Such technologies have in common a communication model that is similar to the local wired network model: user terminals (hereafter denominated \glslink{host}{\em hosts}) access the Internet through a dedicated, authoritative infrastructure device (hereafter denominated \glslink{router}{\em router}). In that sense user terminals are competing ``consumers" of the same networking resource, which consists locally in access to the \glslink{router}{router} granting internetwork (Internet) connectivity. \glslink{router}{Routers}, on the other hand, are ``providers" of the networking resource, and collaborate with one another to provide this resource, \emph{i.e.} internetwork connectivity. This similarity enables IPv4 and IPv6 protocol suites to run quite naturally over such wireless access networks, although IP protocols were in fact designed for wired networks at a time when massive use of wireless Internet access was not yet envisioned. \ \\ \ \\
%
The basic mechanisms provided by IEEE 802.11 infrastructure mode, WiMAX, UMTS or LTE thus provide communication capabilities over a single wireless hop, between a user terminal and an infrastructure access point. 
Some extensions of these basic mechanisms provide direct device-to-device communication (as the Wifi ad hoc mode) or even multi-hop wireless communication through relays planned in advance ({\em e.g.} with LTE or WiMAX). However, these wireless networks all have in common their \emph{managed} nature: they depend entirely on an infrastructure planned and deployed in advance, controlled by an operator. This chapter does not focus on such networks.

\subsection{Spontaneous Wireless Networks}
\label{ss:spontaneous}
%
Although so far not as successful as managed wireless networking, an alternative type of wireless networks has also emerged since 2000: \emph{spontaneous wireless networks}. Inspired by the Push-To-Talk concept used in walkie-talkies (portable half-duplex radio transceivers developed during the Second World War), spontaneous wireless networks depart from the traditional distinction between \glslink{router}{routers} and \glslink{host}{hosts}, whereby each user terminal (hereafter, {\em node}) may behave as a \glslink{router}{router} and a \glslink{host}{host} simultaneously. In spontaneous wireless networks, user terminals are thus ``prosumers" ({\em i.e.} both producers and consumers) of networking resources instead of mere consumers. Terminals self-organize to provide multi-hop wireless communications among themselves, with or without help/control from infrastructure devices. Each node may thus simultaneously originate/receive traffic (role of a \glslink{host}{host}), as well as forward traffic on behalf of other terminals (role of a \glslink{router}{router}). \ \\ \ \\ 
%
Popular examples of spontaneous wireless networks include mobile ad hoc networks, wireless mesh networks, wireless sensor or actuator networks, wireless smart meter networks, vehicular networks, opportunistic wireless networks or delay tolerant networks. Spontaneous wireless networks are considered as interesting solutions to extend and offload managed wireless networks hampered by increasingly heavy smartphone data communications \cite{NY-TIMES}. They can also increase the resilience of the network in scenarios where infrastructure is not usable, due to a disaster, to the military situation or to the political situation, for instance \cite{COMMOTION}. In addition, spontaneous wireless networking is an effective way to extend the reach of wireless Internet access, without costly additional infrastructure deployment \cite{OLPC}. \ \\ \ \\ 
%
Popular link layer technologies providing device-to-device communication in spontaneous networks include so far IEEE 802.11 ad hoc mode \cite{IEEE-802-11} and IEEE 802.15.4 \cite{IEEE-802-15.4}. However, in order to provide multi-hop communication in spontaneous wireless networks, additional techniques have to be employed on top of such link layer technologies, and that is the subject of this chapter. The focus is put on the use of standard IP protocols to enable multi-hop wireless communications in spontaneous wireless networks -- in order for these networks to effectively blend in the Internet, where appropriate.%\ \\ \ \\
%
\paragraph{Handling heterogeneity at layer 3} Since the early days of computer networking and the first steps of today's Internet, the diversity of networking technologies has been handled exclusively at the physical and the link layers (layers 1 and 2 OSI). The internetworking layer (layer 3) has been conceived as a ``convergence layer" in which a single protocol (the Internet Protocol, IP) runs unchanged on top of heterogeneous interconnected networks, as it can be observed in Figure \ref{f:conv} \cite{miya}.

\begin{figure}[ht]	% H-must be here or [htb]
\centering
\includegraphics[width=0.7\textwidth]{Figures/protostack-crop.pdf} %	** if .eps or .pdf, don't need extension}
\caption{OSI reference model and IP networking architecture \cite{miya}.}
\label{f:conv}
\end{figure}
%
The development of wireless technology entails however substantial changes in the way that networks are usually represented and conceived. Characteristics of spontaneous wireless networks cannot be handled exclusively at lower layers of communication, as they challenge some of the key assumptions of the IP-based networking architecture. They need thus to be taken into account at layer 3. As more flexible wireless networks are deployed and get increasingly interconnected and integrated with other networks --or in the Internet--, the use of IP over these networks need thus to be adapted or reconsidered. The {\bf first contribution} of the chapter is a review of these considerations, as it elaborates on how the IP-based network architecture is challenged by spontaneous wireless networks.
%
\subsection{Mobile Ad hoc and Low-Power Lossy Networks}
\label{ss:manet_lln}
%
IP protocols are developed, standardized and maintained by the Internet Engineering Task Force (IETF \cite{IETF}). Most of the IETF's protocol design and standardization activities have so far focused on two categories of spontaneous wireless networks: Mobile Ad hoc Networks (MANETs) and Low-Power Lossy Networks (LLNs).

% ({\em e.g.}, a \glslink{router}{router} with multiple \glslink{host}{hosts} and wireless communications devices)--herein simply referred to as `nodes'--

\paragraph{Mobile Ad hoc Networks (MANETs)} According to the IETF's terminology (defined in RFC 2501 \cite{rfc2501}), a MANET consists in a set of ``mobile platforms (..) --herein simply referred to as `nodes'-- (..) which are free to move about arbitrarily. The nodes may be located in or on airplanes, ships, trucks, cars, perhaps even on people or very small devices, and there may be multiple \glslink{host}{hosts} per \glslink{router}{router}. A MANET is an autonomous system of mobile nodes. The system may operate in isolation, or may have gateways to and interface with a fixed network" \cite{rfc2501}. Note that this definition {\em allows} \glslink{router}{router} mobility, but it is {\em not restricted} to mobile networks; the term includes all wireless multi-hop ad hoc networks, regardless of whether they are static or not.

\paragraph{Low-Power Lossy Networks (LLNs)} According to the IETF's terminology (defined in {\tt draft-ietf-roll-terminology-12}\footnote{An Internet-Draft in the Last Call for becoming an RFC, at the the time of writing this chapter.} \cite{draft_lln}), LLNs are ``typically composed of many embedded devices with limited power, memory, and processing resources interconnected by a variety of links, such as IEEE 802.15.4, LowPower WiFi" \cite{draft_lln}. LLNs are thus a more specific case of MANETs (as defined in the previous paragraph), in which \glslink{router}{routers} typically operate with constraints on processing power, memory, and energy (battery power). Their interconnections are characterized by high loss rates, low data rates, and link instability.  LLNs are comprised of anything from a few dozen to thousands of \glslink{router}{routers}.  Supported traffic flows include point-to-point (between devices inside the LLN), point-to-multipoint (from a central control point to a subset of devices inside the LLN), and multipoint-to-point (from devices inside the LLN towards a central control point). Alternative, but similar terminology is employed in {\tt draft-ietf-lwig-terminology} \cite{bormann13}, which defines the terms ``constrained nodes" and ``constrained networks" with various classes of constraints. \\

Concrete examples of MANETs and LLNs include the following three use cases, selected only for illustrative purposes, to hint at the wide heterogeneity of features, requirements and user expectations that one must address in spontaneous wireless networking.

\paragraph{Vehicular Ad hoc Networks (VANETs)} Communication in VANETs is enabled between moving vehicles in urban scenarios or roadways, (possibly) with fixed devices installed in Roadside Units (RSUs) along the road/street. The combination of vehicles and RSUs forms a mobile, highly dynamic ad hoc network. Devices participating in vehicular networks (either inside vehicles or in RSUs) have neither significant energy constraints nor severe computational limitations, but those installed in vehicles are not, in general, cooperative and willing to dedicate resources to others' communication. Research in these networks has typically focused on safety applications, such as distribution along the highway of information about traffic-related events -- \emph{e.g.}, jams or accidents \cite{bc_vanet}. Other purposes could be also considered, such as dissemination of service availability along the highway (gas stations, tolls, accommodation, etc.). As such, a VANET is a category of MANET.

\paragraph{Community Wireless Mesh Networks} These are cooperative, non-commercial networking projects in which users join and contribute to the deployment of the network, in particular by sharing resources and allowing the use of their devices as networking relays. Several initiatives have flourished in the last years, such as Spain's {\em Guifi} \cite{GUIFI}, mostly deployed over the Eastern coast of Spain (Catalonia and Valencia) but also present in many other parts of the country. Other examples include Germany's {\em Freifunk} \cite{FREIFUNK} and Austria's {\em Funkfeuer} \cite{FUNKFEUER}. Some of them cover large geographical areas and contain thousands of nodes\footnote{{\em Guifi.net}, for instance, claims 31865 nodes, 20425 of them being ``operating nodes" (last query to {\tt http://www.guifi.net} on April 10th, 2013).}. These networks are typically static, most of the links are wireless links operating in free (unlicensed) frequency bands. Their topology and capacity evolve dynamically, in an unplanned manner, subject to events such as the ingress and egress of users, the subsequent availability of new links and resources or the upgrade of a particular networking region. These networks enable free communication among their users, but they can also provide access to the Internet if there are gateways available. As such, a community wireless mesh networks is also a category of MANET.

\paragraph{Wireless Sensor Networks (WSNs)} WSNs are collections of sensors intended to measure one or several properties of the environment in which they are deployed. Communication facilities required by such networks need to include, at least, the transmission of collected information from the sensors to a gateway or central server that stores and eventually process it, and the transmission of information (\emph{e.g.}, configuration instructions or measurement schedules) from the server to one or more sensors. There is a broad range of information that may be collected and exchanged through WSNs, some examples including climate studies, bird observation, power monitoring in buildings or tracking of patients' health parameters with body sensors. Properties of a WSN may vary depending on the purposes of the sensor deployment, but there are some usual constraints. Sensors are often battery driven, the lifetime of the sensor is limited by the battery lifetime. Protocols for enabling communication within WSNs must therefore be designed with energy consumption and energy-efficiency in mind. As such, a wireless sensor network is a category of LLN.\\

Despite their heterogeneity, these use cases -- and other applications of spontaneous wireless networks -- have common characteristics, including bandwidth scarcity and need for self-organization. These characteristics both require the use of efficient, highly decentralized routing and flooding mechanisms, able to react quickly to topology changes without overloading the network. This chapter thus focuses more specifically  on IP protocols that enable routing and flooding in MANETs and LLNs. While plenty of protocols have been proposed in the literature, only few have been effectively implemented, standardized and used in real-world deployments. The {\bf second contribution} of the chapter consists in:

\begin{enumerate}[(1)] 
\item an analysis of the main implications of wireless mesh characteristics on the task of flooding and routing typically implemented in upper-layer protocols; and 
\item a description and discussion of the key mechanisms and operation of the main protocols deployed so far and standardized at the IETF for routing in MANETs, in LLNs and in heterogeneous wired/wireless inter-networks.
\end{enumerate}

\subsection{Reader's Guide}

Reader is assumed to be familiar with the main concepts of computer networking and the TCP/IP network reference model, whose terminology is used in this chapter. Interested readers are referred to the book of Tanenbaum {\em et al.} \cite{tanenbaum} for details; the glossary at the end of the chapter displays standard definitions of the basic networking concepts. Basic knowledge of the Internet Protocol (IP) operation, addressing model, routing and Internet architecture is also preferable, but not necessary. These elements are briefly overviewed in section \ref{s:fundamentals}, in order to better highlight the issues that arise with the traditional IP model in spontaneous wireless networks, addressed in section \ref{s:comm_wless}. This section describes the conditions under which wireless communication occurs, and examines their impact on the communication performance and the architecture of spontaneous wireless networks. In particular, the section explains the non-suitability of the conventional IP networking model for spontaneous wireless networks, and discusses an alternative model. \ \\ \ \\ 
%
The rest of the chapter focuses on the mechanisms and protocols that have been designed to handle flooding and routing in spontaneous wireless networks, paying a particular attention to the efforts deployed at the IETF. Section \ref{s:flood_rtg} motivates and presents the mechanisms, and section \ref{s:specific} describes the routing and flooding protocols that have been specifically designed in the IETF to operate on MANETs and LLNs. Section \ref{s:wospf} focuses on the problem of extending legacy Internet routing protocols so that they can efficiently operate on hybrid (wired/wireless) inter-networks. Finally, section \ref{sec:conclusion} concludes the chapter.

### 2013-07-22 Compiled/chapter9/Content/wospf.tex
all files compile correctly

\section{Routing in Wired/Wireless Internetworks with OSPF}
\label{s:wospf}
%
Protocols reviewed in section \ref{s:specific} have been specifically designed for spontaneous wireless networks. However, the increasing deployment of wireless technologies and the integration of different sorts of flexible networks with the Internet is leading to more complex inter-networks, neither purely wired networks nor purely spontaneous wireless networks, resulting from the interconnection of wireless mesh networks with fixed, wired networking infrastructure, inside Internet's Autonomous Systems. Figure \ref{f:cpd_as} shows schematically a {\em compound Autonomous System}, in which fixed and wireless mesh networks are interconnected in the same routing domain. \ \\ \ \\
%
\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth]{Figures/cpd_as-crop.pdf}
\caption{A compound Autonomous System.}
\label{f:cpd_as}
\end{figure}
%
In these scenarios, a classic IGP (in IP networks, typically OSPF) is used for routing in the fixed network inside the AS. Rather than using an additional protocol for routing in the wireless mesh network inside the AS, it makes sense to explore approaches {\em extending} the protocol already used in the AS, so that it can take into consideration the issues described in section \ref{s:comm_wless}, run efficiently over wireless dynamic networks, handle the heterogeneity of the hybrid internetwork and thus perform routing over the whole compound AS. Extension for hybrid internetworks of a protocol already in use can significantly reduce the transition costs (technical implementation, engineer training...), as only minor changes, or no changes at all, will be needed in the networks using the original protocol. It may be also benefial in terms of networking management complexity and routing performance, as a single (extended) routing protocol is more bearable than several protocols running in different parts of the internetwork. In the latter case, route distribution between different protocols operating at wireless and wired networks needs to be performed in specific {\em hybrid routers} (see Figure \ref{f:cpd_as}); this adds another layer of networking complexity and is likely to cause routing suboptimality. The advantages of extending a protocol in use come, however, at the expense of increasing the complexity and narrowing the space for optimization in the extended protocol, which needs to cope efficiently with a broader range of networking scenarios.\ \\ \ \\
%
This section reviews the IETF extensions of the Open Shortest Path First (OSPF) protocol for MANETs. Section \ref{ss:ospf} shortly reviews the basics of OSPF, the main IGP for IP networks and a major representative of the link-state routing family, and indicates the reasons that prevent OSPF to be used ``as-is'' in wireless multi-hop ad hoc networks. Section \ref{ss:ospfmanet} describes the main elements of the three extensions for MANETs standardized by the IETF: Multi-Point Relays (MPR-OSPF, specified in RFC 5449), MANET Designated Routers (OSPF-MDR, specified in RFC 5614) and Overlapping Relays (OR/SP, specified in RFC 5820).
%
\subsection{Open Shortest Path First Protocol (OSPF)}
\label{ss:ospf}
%
OSPF \cite{rfc2328,rfc5340} is a link-state routing protocol for IP networks. Each router maintains a local {\bf Link State Database} (LSDB), representing the full network topology. The protocol ensures that each router has the same LSDB and, thus, the exact same view of the network topology. Paths to every possible destination are derived from the {\bf Shortest Path Tree} (SPT) that every router computes, by way of Dijkstra's algorithm \cite{dijkstra59}. \ \\ \ \\
%
Routers acquire information about their 2-hop (bi-directional) neighborhood and advertise their own presence and their 1-hop neighbors by periodically exchanging {\bf Hello} messages with all their neighbors, in the way described in section \ref{ss:nd}. \ \\ \ \\
%
Topology information is also disseminated through the network by way of {\bf Link State Advertisements} (LSAs). Each such LSA lists mainly the current adjacencies of the router which generated the LSA. The local LSDB stored by a router contains the most recent LSAs received from every other router in the network. \ \\ \ \\
%
Each router synchronizes its LSDB with a subset of its bidirectional neighbors. Synchronization between two neighboring routers is performed on a master-slave basis, by exchanging summaries of all LSAs in their LSDB, and then allowing each router to request retransmission of missing or locally outdated link-state advertisements. Links between a router and its synchronized neighbors are called {\bf adjacencies}. The set of adjancencies is expected to form a network-wide connected backbone, connecting all routers in the network, in order to ensure paths can be computed correctly.\ \\ \ \\
%
Finally, routers also acquire remote topology information by receiving LSAs. LSAs are flooded through the entire network in reliable fashion (explicit acknowledgements and retransmissions) via the backbone formed by adjacencies. Thus, any router which has formed adjacencies must advertise this periodically by way of constructing an LSA and performing LSA flooding.
%
\begin{equation}
\textrm{SPT links} \subset \textrm{Adjacent links} \subset \textrm{Bi-directional links}
\end{equation}
%
Remote topology information is then used for the construction of the Shortest Path Tree: each router computes the shortest paths based on the information contained in the set of received LSAs. \ \\ \ \\
%
This operation implies that OSPF exchanges control traffic and performs routing according to two principles:
%
\begin{enumerate}
\item Data traffic is routed to the corresponding destination through links contained in the Shortest Path Tree.
\item Data and control and traffic (LSAs and acknowledgements) is sent over adjacent (synchronized) links.
\end{enumerate}
%
\paragraph{Interface Types} 
%
Rules for flooding and adjacency handling vary for the different \emph{interface types} supported by OSPF. Four main interface types are specified in RFC 2328 \cite{rfc2328}:
%
\begin{itemize}
\item \emph{Point-to-point interfaces} are those connected to point-to-point links. Such a link only permits communicating with a single (neighboring) interface.
\item \emph{Broadcast interfaces} participate in a broadcast link, in which any interface can directly communicate with any other interface. A classic example of broadcast link is Ethernet.
\item \emph{Non-Broadcast Multiple Access (NBMA) interfaces}, for non-broadcast networks ({\em i.e.}, networks supporting more than two routers, but without broadcast capability) in which each pair of interfaces can communicate directly. This interface type may be used with X.25 and ATM networks with Switched Virtual Circuits (SVC).
\item \emph{Point-to-multipoint interfaces}, for those non-broadcast networks in which direct communication between any pair of interface is not guaranteed. This may be the case, for instance, in Frame Relay networks using only Permanent Virtual Circuits (PVC), if not every pair of routers have a PVC between them. 
\end{itemize}
%
OSPF only provides support for the two first interface types. In the NBMA and point-to-multipoint cases, OSPF {\em emulates} the behavior of a broadcast link and point-to-point links, repectively. For NBMA networks, LSA flooding and LSDB synchronization are handled by way of {\bf Designated Routers} (DRs). A Designated Router (as well as a Backup Designated Router, BDR, expected to become DR in case of DR's failure) is elected from among routers whose interfaces are connected to the same link. DRs (and BDRs) form adjacencies with all the routers connected to the same link, and the Designated Router becomes responsible for flooding of LSAs, originated by routers on that link. A router point-to-multipoint link, in turn, is handled as a set of independent point-to-point links, one per neighboring router with which direct communication is available.
%
\subsection{MANET Extensions: A Wireless Interface for OSPF}
\label{ss:ospfmanet}
%
Standard interface types for non-broadcast networks (point-to-multipoint and NBMA) are not adapted for operation in a wireless multi-hop ad hoc network. As discussed in section \ref{ss:issues}, routers in a wireless multi-hop network may not agree on which routers are connected to a given link. This implies that the DR-based mechanisms of NBMA cannot be directly used in wireless multi-hop ad hoc networks. DR election may be inconsistent between different routers, causing flooding to disfunction and, possibly even preventing the protocol from converging. The use of the point-to-point interface, in turn, does not scale in these dynamic networks: point-to-point emulation for every pair of interfaces directly reachable to each other causes an excessive control traffic overhead, even for relatively small networks, as shown experimentally in Henderson {\em et al.} \cite{henderson03}. This fact has led the research and industrial OSPF community to develop a new interface type to support the characteristics of wireless multi-hop ad hoc networks.\ \\ \ \\
%
This new interface type needs to optimize the operation of (1) describing local topology in LSAs, (2) performing LSA flooding and (3) establishing and maintaining adjacencies in the context of wireless communication. Different approaches have been explored at the IETF, which have led to three different extensions of OSPF, consisting of three different interfaces for wireless multi-hop networks (or MANETs, in IETF's terminology).
%
\paragraph{Multi-Point Relays} MPR-OSPF \cite{rfc5449} use Multi-Point Relays (MPR \cite{mpr-hicss2002}, see section \ref{ss:flood}) to optimize topology description, LSA flooding and LSDB synchronization. Nodes select MPRs from among their bidirectional neighbors in order to provide 2-hop coverage, and use them to disseminate their LSAs. A router becomes adjacent to both neighbors which it has selected as multi-point relays (MPRs) and neighbors which have selected the router as their multi-point relay (MPR selectors). Each router advertises in its LSAs its own MPRs and MPR selectors; consequently, the Shortest Path Tree is constructed over the set of adjacencies. 
%
\paragraph{Overlapping Relays \& Smart Peering} The Overlapping Relays / Smart Peering (OR/SP) extension of OSPF \cite{rfc5820} floods LSAs via MPR as in MPR-OSPF, where the multi-point relays selected among the adjacent (synchronized) neighbors of the electing router. Adjacencies are selected following the Smart Peering (SP) rule, in which a neighbor becomes adjacent if it is not already reachable through the computing router's current Shortest Path Tree. The SP criterion reduces dramatically the number of synchronized links in the network. LSAs list adjacent neighbors, and may also list additional bidirectional neighbors (so-called \emph{unsynchronized adjacencies}). The SPT is thus constructed over adjacencies and a subset of bidirectional neighbors. 
%
\paragraph{MANET Designated Routers} OSPF-MDR \cite{rfc5614} relies on two Connected Dominating Sets (CDS): the MANET Designated Routers (MDR) backbone and the Backup MDRs (BMDR) backbone. Both extend the standard OSPF (for NBMA networks) notions of ``Designated Routers" and ``Backup Designated Routers" to MANETs. This implies that routers behave differently depending on their role. MDRs are the only nodes allowed to flood LSAs. Every non-MDR router becomes adjacent at least to the closest MDR, and MDRs must become adjacent to other MDRs. LSAs list a configurable subset of links of the originator, which must at least include the adjacent neighbors. The SPT is thus constructed over adjacencies and a subset of bidirectional neighbors. \ \\ \ \\
%
Compatibility with the OSPF routing philosophy detailed in section \ref{ss:ospf} varies significantly depending on the considered OSPF extension. MPR-OSPF is designed to preserve the two principles in OSPF routing: shortest, synchronized paths for data traffic and synchronized links for control traffic. Under the Overlapping Relays extension, data traffic paths are synchronized, but they are not necessarily optimal, as routers only synchronize a small fraction of their available links. Although providing several configuration parameters to tune the protocol's performance, the MANET Designated Routers (OSPF-MDR) also try to minimize the control traffic by reducing the number of synchronized links, even when this may lead to path suboptimality for data traffic.
%
\paragraph{Preserving OSPF routing principles} Performed experiments suggest that extensions providing (theoretical) shortest paths for data traffic achieve a better performance than those neglecting shortest paths or allowing suboptimal routing in a wireless multi-hop network \cite{aircc, porto}. Further analysis showed that preserving the second principle (all traffic is sent over synchronized links) in OSPF over mobile ad hoc networks, in the way that MPR-OSPF does, requires a significant amount of overhead due to the LSDB exchange between routers becoming {\em adjacent} (synchronized), and does not bring substantial benefit, due to short lifetime of several synchronized links in a wireless multi-hop dynamic network \cite{hicss2011}.
%
\paragraph{Why maintain LSDB synchronization in Extended OSPF ?} LSDB synchronization between two routers proves useful in classic (wired) Internet internetworks, but is an expensive operation to perform in a dynamic (wireless multi-hop or mobile ad hoc) network. This is the reason why other link-state protocols such as OLSR, designed specifically for wireless mesh and mobile ad hoc networks, does not provide any mechanism for synchronizing the LSDBs of neighboring routers: topology information is only disseminated through the network by way of LSA flooding (see section \ref{sec:olsr}). In the case of extended OSPF, there are two reasons for maintaining the notion of LSDB synchronization: 
%
\begin{enumerate}
\item {\bf OSPF backwards compatibility}. In standard OSPF \cite{rfc2328,rfc5340}, the notion of {\em adjacency} is essential in the protocol's architecture and the router's operation, regardless of the specific types used for running OSPF in the router's interfaces.
\item {\bf Routing in heterogeneous internetworks}. Unlike OLSR, extended OSPF is expected to run over hybrid internetworks (or compound Autonomous Systems, see Figure \ref{f:cpd_as}), that is, internetworks in which wired networks handled by standard OSPF interface types coexist and are interconnected with wireless multi-hop networks using the adapted wireless interface of (extended) OSPF. In these scenarios, in which some nodes (with wireless interfaces) are exposed to frequent disconnections from the network (meaning that their LSDBs may be no longer updated for a while) and others maintain stable links with their neighbors (those with wired interfaces), the fact that every router is expected to synchronize its LSDB with {\em at least} one of its neighbors provides an upper bound for the maximum time that a router $A$ (in the wireless region of the internetwork) stays {\em disconnected} (that is, unaware of its local topology) from another router $B$ (in the wired region of the internetwork) after missing an LSA flooded by router $B$. This becomes an issue as the time between consecutive LSA flooding processes from $B$ is typically high -- as wired links are stable and thus require less frequent updates about their state than wireless ones.
\end{enumerate}
%
\paragraph{Further Extensions: adapted LSDB synchronization, MPR+SP and SLOT-OSPF} In this context, some additional approaches can be explored beyond the three standardized extensions of OSPF. Clausen {\em et al.} \cite{secon2004}, for example, propose a LSDB synchronization process based on the periodic broadcasting of {\em signatures} of the LSDB by every router to its neighborhood. These signatures allow neighbors of the originator to detect topology inconsistencies with its own LSDB, and request unicast retransmission of the corresponding LSAs. This turns the standard OSPF synchronization mechanism, based on a router-to-router LSDB exchange, to a router-to-neighborhood mechanism that takes advantage from the semibroadcast nature of communication in a wireless multi-hop network.

Without modifying the standard OSPF adjacency-forming process, LSDB synchronization can be kept, but the number of adjacencies per router should be reduced as much as possible, given the high cost of synchronization in terms of overhead and its small benefit in a dynamic network, with short-lived links. Data traffic should be sent over shortest paths (that is, optimal paths over the network, according to the available LSDB information and the metric in use), but these paths do not need to be synchronized. This leads to combine in the same OSPF extension the mechanisms to provide shortest paths (MPR selection for topology description) and the mechanisms reducing the most the number of adjacencies to be established per router ({\em e.g.}, the Smart Peering rule used in the Overlapping Relays extension). The resulting extension, denominated MPR+SP, presents a better routing performance than extensions MPR-OSPF and OR/SP in which it is based, as shown in Cordero {\em et al.} \cite{hicss2011}. Similarly, extension SLOT-OSPF \cite{mass2010} using the Relative Neighbor Graph (RNG \cite{toussaint80}) for establishing adjacencies and MPR selection for computing shortest paths, also achieves better results in terms of delivery ratio and control traffic overhead than the standard extension to which it compares. In both cases (MPR+SP and SLOT-OSPF) shortest path computation, for which a comprehensive view of the network topology (with most of the links) is required, is splitted from the adjacency-forming criterion, which aims to reduce as much as possible the number of LSDB synchronizations to be performed. This split enables a further optimization of the protocol routing performance.

### 2013-07-22 originals/chapter1/ebook-sec1.tex
all files compile correctly

% This work is licensed under
% http://creativecommon.org/licenses/by/3.0/
\section{Introduction}
\label{sec:sec1}

The Internet is increasingly mobile.  
Users access Internet services from mobile devices that move from one
wireless access point to another, or switch between WiFi and cellular 
network connectivity.  
Ubiquitous computing relies on sensors and actuators attached to vehicles,
portable objects, and animals as well as people.
Applications 
provide customers with application-level identities that can be
used to reach them at whichever device they are currently using.
Increasingly, software runs on virtual machines that can migrate from one
physical server, or even one data center, to another.

We define network
mobility as the capability that allows a communicating entity
to continue to communicate over a network, despite the fact that its
location at (or binding to) a lower-level communicating entity is changing.
Further, we focus on so-called ``seamless'' mobility, in which the
high-level entity's communication channels are preserved throughout
the change.

It is important to note that this definition applies at all conceptual
levels.
A person can have an identifier as a communicating entity, and can be
mobile by moving from one networked device to another.
A device such as a
cellphone can have an identifier, and can be mobile by moving from one
network attachment point to another.
An interface on a device, such as an Ethernet interface, can have an
identifier and be mobile by moving within or between local area networks.

In recent years, many mechanisms have emerged for supporting 
mobility.
A recent survey~\cite{rfc6301} cites 22 Internet protocols dating from 1991
to 2009, including most prominently Mobile IPv4, Mobile IPv6,
MSM-IP, HIP, MOBIKE,
Cellular IP, HAWAII,
ILNP, and LISP Mobile Node.
In other contexts mobility is supported by:
\begin{itemize}
\item
Ethernet LANs and VLANs, which allow an interface to retain its IP address 
as the host moves within the LAN;

\item 
the scalable ``flat'' routing architectures 
SEATTLE \cite{seattle}, PortLand~\cite{portland},
VL2~\cite{vl2}, NVP~\cite{nvp}, and
Rbridges/TRILL~\cite{rbridges,rfc5556}
that also naturally support routing to an interface that
retains its addresses as the host move;

\item
injecting the IP address of a mobile machine into existing routing protocols
such as OSPF and BGP~\cite{connexion,smog};

\item
the General Packet Radio Service (GPRS) Tunneling Protocol (GTP),
which supports mobility in most cellular networks;

\item
other research proposals including
TCP Migrate \cite{tcp-migrate}, Serval \cite{serval}, 
and the Internet Indirection Infrastructure~\cite{i3};

\item
application-level protocols such as 
the Session Initiation Protocol (SIP)~\cite{mobilesip}.
\end{itemize}

\noindent
Each well-known proposal tends to spawn a family of variants, so
the total number is probably in the hundreds and growing.

These various mechanisms operate at different levels, and make
different assumptions about naming, routing, session
protocols, scale, security, and the cooperation of
remote endpoints or multiple administrative domains.  
Because the
community lacks a common framework for describing and comparing
mobility mechanisms, their relationships are poorly understood.
Comparisons tend to be based on superficial characteristics rather
than inherent ones.  
Quantitative comparison must be based on
labor-intensive prototyping and measurement or simulation.

This book chapter has two goals:
\begin{itemize}
\item
To describe and compare existing proposals for implementing mobility.
\item
To map out a design space in which
new mobility mechanisms can be discovered,
evaluated, and exploited.
\end{itemize}

\noindent
To achieve these goals,
we explain mobility in a new way.
We begin by defining a common framework for describing network
architectures.
This framework is
called the {\it geomorphic view} of networking, and is
introduced in
Section~\ref{sec:sec2}.

The geomorphic view has been developed to be simple, modular,
comprehensive, and formalizable.
It supports the first goal by providing a precise, unique description
of each mobility mechanism
that omits inessential detail while exposing subtle differences
and important engineering trade-offs.

The common framework supports the second goal in several ways.
It allows us to generalize over the implementations of mobility,
showing that they are all instances of two major patterns.
It also allows us to understand how 
different instances of the mobility patterns at different places in a
network
architecture can be composed, generating a potentially large design
space to be explored.

Section~\ref{sec:sec3} of this chapter introduces the two major
patterns---\emph{dynamic-routing mobility}
and \emph{session-location mobility}---for implementing mobility.
If they are both implemented within an IP layer, the difference centers
on whether a mobile machine
retains its IP address when
it moves, or changes its IP address and updates its correspondents.
Each pattern has a completely different set of subsidiary design
decisions and resource costs.

The next sections 
use the patterns to describe and compare many of the most
important proposals for mobility.  
Sections~\ref{sec:sec4} and
\ref{sec:sec5} compare different ways to implement
dynamic-routing mobility,
with a non-hierarchical name space ({\it e.g.,}
MAC addresses in a local area
network) or a hierarchical name
space ({\it e.g.,} IP addresses in the wide
area), respectively.  
Section~\ref{sec:sec6} compares four prominent protocols for
session-location mobility at different stages in the IETF
standardization process.

Section~\ref{sec:sec7} of the chapter turns to
a more systematic exploration of the
design space, first showing that there may be some freedom concerning
where in an architecture a particular kind of mobility is handled.
The section also discusses composition of mobility
mechanisms.
As an example, we illustrate how Mobile IPv6 is a
composition of both the dynamic-routing mobility and session-location
mobility design patterns.

Finally, Section~\ref{sec:sec8} surveys several topics closely
related to mobility, including multihoming, anycast services, site
mobility, incremental deployment of mobility protocols, and security
issues for mobility.  
The chapter ends with a brief
conclusion outlining several more advanced areas of study.

### 2013-07-22 originals/chapter1/ebook-sec2.tex
all files compile correctly

% This work is licensed under
% http://creativecommon.org/licenses/by/3.0/
\section{The geomorphic view of networking}
\label{sec:sec2}

The geomorphic
view of networking was originally inspired by the work of Day \cite{pna},
although we have made many changes and additions in both content and
presentation.
In this common framework for describing networks,
the module is a
{\it layer}, and a network architecture is a hierarchy of layers.

\subsection{Comparison with the Internet and OSI models}
Layers may seem familiar and obvious because both the classic Internet
architecture \cite{philo} and the
OSI reference model \cite{OSI}
also describe network architecture as a hierarchy of layers.
However,
our concept of a layer is very different.
As a preview of this section, our layer hierarchies differ from these
earlier ones in at least four ways:
\begin{itemize}
\item
The classic Internet architecture and the OSI reference model both have
a fixed number of levels.
In a geomorphic layer hierarchy, there can be any number of levels.
\item
In the earlier models, there is only one layer on each level, so there
is no distinction between layer and level.
In a geomorphic hierarchy, there can be multiple layers on the same
level.
\item
In the earlier models, each layer has a specific function that is
distinct from the functions of other layers.
In the geomorphic view
each layer is a microcosm of networking, containing all of the basic
components and functions in some form.
In different layer instances there are different versions of these
basic ingredients, used at
different levels, with different scopes, and for
different purposes.
\item
Most people interpret the earlier models as describing the data
plane of networking only.
The control plane is seen as separate and not modularized in the
same way.
In the geomorphic view, each layer---being a microcosm of
networking---has a data plane and a control plane.
Layers decompose both planes into modules.
\end{itemize}
Figure~\ref{fig:geoname} illustrates these differences, and also
shows how the ``geomorphic'' view got its name.
The complex arrangement of layers, with overlapping, abutting, and
bulging shapes, can resemble the complex arrangement of layers in the
earth's crust.

\begin{figure}
\centering
\includegraphics[scale=0.45]{figures/geoname.pdf}
\caption{Arrangement of layers in the classic Internet architecture
(left), the geomorphic view (middle), and the earth's crust (right).}
\label{fig:geoname}
\end{figure}

\subsection{Components of a layer}
\label{sec:layercomponents}

A layer has {\it members}, each of which has a unique and persistent
{\it name} within the layer.
For example, Figure~\ref{fig:layer} is a snapshot of a layer with five
members, each having a capital letter as a name.
In general a member is a concurrent process, {\it i.e.,} a locus
of state and control with the potential for autonomous action.

The members of a layer communicate with each other through {\it links},
shown by lines in Figure~\ref{fig:layer}.
A link is a communication channel.

\begin{figure}
\centering
\includegraphics[scale=0.80]{figures/littlelayer.pdf}
\caption{Members and links of a layer.}
\label{fig:layer}
\end{figure}

One of the two primary functions of a layer is to enable members to send
messages to each other.
This function is accomplished by a {\it forwarding protocol},
which runs in all members and has operations for sending and
receiving messages over the links.

In general, a layer does not have a link between each pair of members.
Such a layer needs {\it routes}
indicating how one member can reach another through links and
intermediate members.
For example, ({\it A, B, D, E} ) is a route from {\it A} to {\it E}.
If {\it B} receives a message that is destined
for {\it E}, its forwarding protocol uses the route information
to forward the message to {\it D} on its way to {\it E}.

The {\it routes} are shared state of the layer, and a simple 
geomorphic description need say no more about them. 
To provide more realistic detail, in a real layer the {\it routes}
information is often distributed over forwarding tables found in
the individual members.

The other primary function of a layer is to implement enriched
end-to-end
communication services on top of its bare message transmission.
This function is carried out by a {\it session protocol}.
The forwarding protocol can be unreliable, especially if links are
dynamic and the current routes are obsolete.
A session protocol can provide services including reliability,
FIFO delivery,
and quality-of-service guarantees.
Figure~\ref{fig:impl} shows a session between endpoints {\it a}
and {\it e} of the lower layer.

A {\it channel} is an instance of a communication
service.
Both links and sessions are channels.
A layer can implement its own links internally, and a
layer can implement its sessions for the benefit of its own
members.

Most commonly, however, a link in one layer is
implemented by a session in another layer,
as shown in Figure~\ref{fig:impl}, placing the other
layer lower in the {\it ``uses'' hierarchy}.
If an underlay (lower layer) is implementing a link for an overlay
(higher layer), then the basic attributes of the channel must be
stored in the states of both layers.
In the overlay, the channel object is one of its {\it links}.
In the underlay, the channel object is one of its {\it sessions}.
There must be two names for the sets of channels of interest to a layer,
because a typical layer both uses {\it links} and 
implements {\it sessions}.

For a link in an overlay to be implemented by a session in an underlay,
both endpoint {\it machines} must have members in both layers,
as shown in Figure~\ref{fig:impl}.
The boundary of a
{\it machine} is the boundary of an operating system that provides fast,
reliable communication between members of different layers on the machine.
This fast, reliable operating-system
communication is the foundation on which networked
communication is built.\footnote{Although layer members have been
described as concurrent processes, they are not usually ``processes''
as defined by the operating system; processes in an operating system
have many more properties and associations than layer members do.  
A virtual machine can be regarded as a {\it machine}, in which case
communication through the hypervisor and soft switch of the physical
machine is regarded as networked communication.}

\begin{figure}
\centering
\includegraphics[scale=0.80]{figures/impl.pdf}
\caption{Implementation of a link in an overlay by a session in an
underlay.}
\label{fig:impl}
\end{figure}

The relation between an overlay member and an underlay member on the
same machine is called {\it registration}.
Registrations must be stored in the state of both layers.
In the overlay a registration is recorded as an {\it attachment}, 
which says that the overlay member is attached to the network through
a particular lower layer.
In the underlay a registration is recorded as a
{\it location}, 
which says that a particular member of a particular overlay is attached to
the network at a particular member (its location) of this layer.

The session protocol creates and maintains {\it sessions} data in its
layer, and uses {\it locations} data.
For example, in Figure~\ref{fig:impl}, {\it A} sent a request to {\it a}
for a session with {\it E}.
To create this session, {\it a} learned from its layer's {\it locations}
that {\it E} is currently located at {\it e}.
Messages sent from {\it A} to {\it E} through the link in the overlay
travel through {\it a, b, d,} and {\it e}; the first and last steps
uses operating-system communication, while the middle three steps use
networked communication.

All the major components of a layer are shown in
Figure~\ref{fig:table}.
The forwarding and session protocols perform the two primary functions
of the layer.
These protocols and their operations are collectively known as the
``data plane'' of the layer.
The network's data plane also includes the inter-layer interfaces
through which the endpoints of an implemented link transfer messages to 
and from the
implementing session.

There are six major state components, all of which 
can be dynamic.
We have seen that the session protocol creates and maintains
{\it sessions};
the other five are created and maintained by their own maintenance
algorithms.
The state and algorithms are collectively known as the ``control plane''
of the layer.
Note that the network's control plane also includes the inter-layer
interfaces through which the control algorithms communicate.

\begin{figure}
\centering
\includegraphics[scale=1.00]{figures/table.pdf}
\caption{Major components of a layer.  Arrows show which protocol
or algorithm writes a state component.}
\label{fig:table}
\end{figure}

\subsection{Layers within a network architecture}

Figure~\ref{fig:scope} shows a geomorphic view of the classic
Internet architecture.
The {\it scope} of a layer is its set of potential members.
For example, 
at the top level of the hierarchy, there are two application layers.
The scope of each layer is the set of potential processes running
software for that application.
These layers are pictured as overlapping because the horizontal
dimension is an approximation of geographical space, and both
applications can have members world-wide.
In particular, the registration lines in the diagram show that each
application has a member on one particular machine.

\begin{figure}
\centering
\includegraphics[scale=0.80]{figures/newip.pdf}
\caption{Geomorphic view of the classic Internet architecture.
Internet links are labeled with the LAN that implements them.}
\label{fig:scope}
\end{figure}

In the middle level of the hierarchy there is a single layer called
the ``Internet core.''
Its members are the IP interfaces of networked machines.
In this layer, IP (the ``network layer'' of the classic Internet
architecture) is the forwarding protocol, and TCP and UDP
(the ``transport layer'' of the classic Internet architecture)
are variants of the session protocol.

At the bottom level of the hierarchy there are local area networks
(LANs) with local scopes.
Each LAN member is an interface appropriate to the type of LAN.
For example, for an Ethernet LAN, the members are the Ethernet interfaces
of machines.
Figure~\ref{fig:scope} illustrates the point, made in the preview
of this section, that in the geomorphic view there can be multiple layers
at one level of the ``uses'' hierarchy.

Note that every member of the Internet core is attached to a member
of a layer at the bottom level.
Note especially that for two members of the Internet core layer to
be linked, both of those members must be attached to the same layer
at a lower level, so that the lower layer can implement the link.
This observation is important for understanding mobility.
A gateway in the Internet core layer is attached to multiple LANs, 
so it can forward messages from one LAN to another.

Because layers instantiated at different levels have different purposes,
they have different versions of the common components enumerated in
Figure~\ref{fig:table}.
For one
example, the best-known routing algorithms are found in the Internet
core, where their purpose is reachability.
Now consider a middleware layer, above the Internet core,
offering cloud services and other facilities for enterprise computing.
To provide security, this layer might have routing that ensures
that all messages
to a particular destination pass
through a particular filtering server.
Thus this layer has its own routing (control plane),
separate from Internet routing.
One of the major purposes of its routing is enterprise-specific
security.

This example
illustrates the points, made in the preview
of this section, that in the geomorphic view the number of levels
is not fixed (the middleware layer need not be present for the Internet
to work),
and that each layer can contain its own version of any basic
function or component of networking (such as routing).
In some layers, where a particular function or component is not
needed, its presence is vestigial.

For another example of a basic function with different forms in
different layers, low-level layers such as Ethernet
LANs provide
broadcast as a communication service.
In geomorphic terms, channels (links and sessions) can be multi-point
as well as point-to-point.
The main
services provided by the Internet core are point-to-point, while
an application layer might implement its own multi-party communication
service.\footnote{For simplicity, in the remainder of this chapter,
all communication channels are assumed to be point-to-point.
This is sufficient for a study of mobility.}

Today's Internet is host to many customized architectures running
simultaneously \cite{roscoe,spatscheck}.
Middleware is an important part of the ecosystem, while cloud services
and virtual private networks
add extra layers to the classic Internet architecture.
It is self-evident that fixed layer structures cannot describe
these architectures adequately.
The geomorphic view is intended not only to describe them,
but also to generate a design space including many others not yet
explored.

\subsection{Layers and mobility}
\label{sec:layers.and.mobility}

If asked to define network mobility, most people would say something
like, ``A mobile device continues to have network connectivity as it
moves geographically.''
For a simple Internet example, we can imagine a laptop that detaches
from one edge subnetwork, where it has one IP address, and re-attaches
to another edge subnetwork, where it has another IP address.

No layering is required to understand this scenario.
At the same time, technologically
the scenario is indistinguishable from a scenario
in which one laptop is tossed into a deep lake and another 
laptop is purchased new.

Clearly mobility is more than this.
As the mobile device detaches and re-attaches, we expect it to retain some
identity and credentials, so that it can be reached in some of the
same ways as
before, and has some of the same rights and capabilities as before.
The identity that is preserved is its membership in some layer, which
must not change.
What does change
is the attachment of this identifying
process to some process in some lower layer.

%JEN: references a figure in section 7.  intentional? move figure here?
The left column of Figure~\ref{fig:lift} (which appears later, in 
Section~\ref{sec:fig:lift}) shows the two forms that
this change of attachment can take.
In the top picture, a process {\it m} changes its registration from
one member of a lower layer to another member of the same layer.
The name {\it m} might be an application name, and {\it a1} and {\it a2}
might be IP addresses in an Internet core layer.
In the bottom picture, {\it m} changes its registration from
a member of one lower layer to a member of another lower layer.
Here {\it m} might be an IP address, and {\it a1} and {\it a2}
might be members of two different LANs.

This shows that layering is an intrinsic part of the study of
mobility, because it explains
what stays the same and what changes.
It will help us understand how a person can
call a friend's cellphone, even
though that friend has traveled hundreds of miles since the last call.

Even this is not sufficient to explain, however, how a person can talk
to a friend's cellphone {\it while the friend is traveling hundreds of
miles.}
To explain this aspect of mobility
it is necessary to focus on the communication channel
that is being preserved across mobility events.
As presented in Section~\ref{sec:layercomponents}, a communication
channel is most often used in one layer, where it is called a link,
and implemented in a lower layer, where it is called a session.
Here is another place where layering is intrinsic to the study of
mobility, explaining that the layer that benefits from mobility is
usually not the layer that has the responsibility of implementing it.

To summarize, there are two relationships on layer pairs that are
important in mobility.
There is a dynamic
registration relationship between an overlay with a mobile member
and the
underlays to which that member of the overlay is attached over time.
There is an implementation relationship between an overlay with a link
and the underlay whose session implements that link.
Two overlay/underlay pairs---in any given instance of mobility, they
must be the same overlay and underlay, right?

Wrong.  
Section~\ref{sec:sec3} will show that there are two patterns for mobility.
In one pattern the layer pairs coincide, and in the other 
they are different.
People are often confused by mobility because it is often over-simplified.
Mobility is easy to over-simplify when one is not explicit about the
layers involved.

\subsection{Mobility in the wild}

It might be said that the problem with mobility is not too few proposals,
but too many.
As mentioned in the introduction, the
total number is probably in the hundreds and growing.

In this chapter, the geomorphic view will provide a descriptive 
framework that imposes some order on this chaotic design space.
This works because every mobility proposal has a unique description
in terms of layers in the geomorphic view.

Unique description is achieved only because the geomorphic view is
precisely defined and precisely used.
For one example, in the geomorphic view there is one name space per
layer.
If any proposal has two different names for the same machine,
one higher-level and one lower-level, then those names must be in
the name spaces of two different layers.
For another example, in the geomorphic view there is no tunneling.
Tunneling is evidence that there are two distinct layers: a higher
layer in which the ``tunnel'' is a link, and a lower layer that
implements the link.

As a result, a geomorphic description of a network architecture might have
more layers than a different description, and some components of some
layers might be vestigial.
This is a cost, but in return we get many benefits, even beyond the
benefits of having a unique and comparable description of each proposal:
\begin{itemize}
\item
Each layer is simpler, with a minimum of {\it ad hoc} complications.
\item
Proposals that might seem very diverse fall into a few recognizable
patterns that apply at any level of the network stack.
\item
We can identify opportunities for re-use of formal models, 
formal analysis, and implementation code.
\item
Mobility is not the only networking challenge.
If other complex mechanisms are also described in terms of the geomorphic
view, we can make sure that they interact correctly.
\end{itemize}
Furthermore, redundancies in a description or model can be
removed by optimization in an implementation phase.
The trick is to understand and analyze the model first, then use the
analysis to determine which optimizations are safe.

This approach leads to differences from other literature on
mobility.
As exemplified by
\cite{akyildiz}, it is common for mobility proposals to be
classified according to the layer of the classic Internet architecture
where they are implemented.
In contrast, we emphasize that each specific proposal is an instance
of a general pattern, and that the general pattern can be used at any
level of a network architecture.
Comparisons between ideas are less subjective, because they are based
on a common framework that exposes real similarities and differences,
even when obscured by incidentals of language and application.

### 2013-07-22 originals/chapter1/ebook-sec4.tex
all files compile correctly

% This work is licensed under
% http://creativecommon.org/licenses/by/3.0/
\section{Examples of dynamic-routing mobility in non-hierarchical layers}
\label{sec:sec4}

Dynamic-routing mobility is often used in LANs, which have smaller
scopes and can function without a hierarchical name space.  
%
% JEN: Pamela, check this
%
These LANs handle mobility naturally as part of the normal routing
function, since end-points retain their addresses as they move and
routing does not rely on location-dependent addressing.

\subsection{Wired Ethernet LANs}

An Ethernet LAN is a single layer.
Its member processes are the Ethernet representatives of hosts (endpoints)
and switches (routers),
and its names are MAC addresses.
It has no pre-attachment requirements or configuration for hosts,
which makes it ``plug and play.''

The LAN offers both broadcast and point-to-point services to
higher layers.
In this brief section we do not consider these communication services
further, so
there will be no discussion of the layer's sessions or locations.
Also, for simplicity, we will not extend the modeling into lower levels,
so links in the Ethernet layer are primitives.

An Ethernet layer has two kinds of links.
There are point-to-point links between 
switches, each of which is basically a wire
between two machines.
There are also shared media or buses.
A bus delivers each message
to every machine on the bus, and is used to connect a switch to a
set of hosts.
Either kind of link
can be identified at each switch that
uses it by the port on the switch's machine to which it is attached.

The inter-switch links of the layer must form a bidirectional
spanning tree (see Figure~\ref{fig:ethlinks}). 
Otherwise, when flooding is used (see below), 
the network could be overwhelmed by
messages traveling on cycles.
There are usually more physical links than needed for the spanning tree,
but the extras
can only be used when other links fail and the spanning tree
is recomputed.

\begin{figure}
\centering
\includegraphics[scale=1.00]{figures/ethlinks.pdf}
\caption{Inter-switch links of an Ethernet LAN layer (left) and an overlay
layer (right).
The Ethernet links are physical, while the overlay links are virtual.}
\label{fig:ethlinks}
\end{figure}

Each switch has a forwarding table containing {\it (MAC address, port)}
pairs.
The port identifies the link on which the switch should forward messages
destined for the MAC address.  
Each switch's table is sparse and is populated lazily by a routing
algorithm called ``MAC learning.''
Upon receiving a message with a source MAC 
address that is not in its forwarding
table, the switch adds to its table the MAC address and the link
on which the message was received.

The forwarding algorithm of a switch is similarly simple.
Upon receiving a message not destined for itself, the switch looks for the
destination MAC address in its own forwarding table.
If it finds an entry, it forwards the message on the designated link.
If it does not find an entry, it ``floods'' by forwarding the message
on every link except the one on which it was received.

These mechanisms implement dynamic-routing mobility as an
aspect of normal operation rather than as a special case.
When a host moves within the layer, it changes the link through which
it is attached to the layer.
As soon as it sends messages, new routes to it begin to propagate through
the layer.
Obsolete forwarding-table entries
are removed when their time-to-live expires.
Missing table entries are handled by flooding.
Note that an entry might also be removed from a forwarding table because
the table is full and space for a newer entry is needed.

\subsection{Ethernet overlays}

Several recent
designs~\cite{seattle,vl2,nvp} avoid flooding by forming
an overlay topology that interconnects all of the edge switches, as
shown on the right side of Figure~\ref{fig:ethlinks}.  While the
inter-switch links of an Ethernet are physical and form a spanning
tree, the inter-switch links of an overlay network are virtual and
fully connect the switches.

The virtual links are communication services implemented by a second,
lower layer.  For example, Figure~\ref{fig:seattle} shows the path of
a message from host {\it Hv} to host {\it Hz} (the lower-case letters
stand for their MAC addresses).  On each hop, the path is labeled with
the source name above and the destination name below.  The virtual hop
between switches {\it Sw} and {\it Sy} in the overlay layer is
implemented in the underlay, where the message is encapsulated in a
message with source {\it w} and destination {\it y}.

How are the virtual links in the overlay implemented by the underlay?
The members of the underlay layer are the switches
only, not the hosts.  Each switch's name is the MAC address of its
machine, just as in the overlay, so there is no need for a 
{\it locations} state component to map one name to another.  
The members of the underlay are stable and
stationary.  Routing is static except for failures, and the forwarding
tables are fully populated.  Because there is no flooding, there is no
need to restrict the links to a spanning tree, and all of the physical
links between switches can be fully utilized.  
The underlay can run an efficient routing protocol, such as a link-state
protocol, to compute a shortest path from one edge switch to
another.

\begin{figure}
\centering
\includegraphics[scale=1.00]{figures/seattle.pdf}
\caption{The path of a message through two layers and three
switches.}
\label{fig:seattle}
\end{figure}

Routing in the overlay is unusual compared to routing in general,
because every edge switch is directly linked to every other edge switch.
This means that an inter-switch route to a host
can be identified simply by the
MAC address of the host's edge switch, and is exactly the same
no matter which switch needs the route!
Thus inter-switch routing is a mapping that is {\it global}
within the layer.
%
% JEN: Pamela, check this
%
Note that, despite the use of a single global directory, the mapping
performed is indeed part of \emph{routing} within the layer (i.e., the
{\it routes} mapping), \emph{not} a {\it locations} mapping between
two layers.  The underlay layer in Figure~\ref{fig:seattle} exists to
make the \emph{routing} between the edge switches more scalable, not
to implement the \emph{link} between the two end-points.

As with Ethernet LANs, each switch has a routing table that is
populated lazily ({\it e.g.}, through MAC learning).
The difference lies in what happens when a switch needs a route to
an unknown destination.
Rather than flooding, it looks the route up in a global routing
directory.

When a host moves, the directory is
updated with the new route to the host.  The exact update mechanism
differs from one overlay design to another, depending on whether
mobility is planned ({\it e.g.,} 
virtual-machine migration in a data center)
or unplanned ({\it e.g.,} a mobile device moving within a campus).  In a
data center, a central controller that triggers virtual-machine
migration can also update the directory with the new
route~\cite{vl2,nvp}.  If the directory cannot be informed in advance
that a host is moving, the new local switch can learn that a new
device has connected and subsequently update the
directory~\cite{seattle}.

The routing directory in an overlay is an important part of its design, 
and may have
many features to make both queries and updates fast and efficient.
Different designs have different directory structures.  VL2~\cite{vl2}
and NVP~\cite{nvp} run a centralized directory on a collection of
server machines.  In these designs, if the ingress switch {\it Sw}
does not know the route for host {\it Hz}, {\it Sw} queries a
directory server to learn the route {\it Sy}.  In contrast,
SEATTLE~\cite{seattle} implements the directory as a ``one-hop
Distributed Hash Table''~\cite{gupta} running directly on the
switches.  
If the ingress switch {\it Sw} does not know the route for
host {\it Hz}, {\it Sw} computes the hash of the {\it Hz}'s MAC
address and forwards the message over a single overlay link to the
switch responsible for this hashed value.  This switch, in turn,
forwards the message to {\it Hz}'s local switch {\it Sy} and informs
switch {\it Sw} of the route for {\it Hz} so that future messages flow
directly from {\it Sw} to {\it Sy}.

To improve the speed of mobile handoff, the ingress switch {\it Sw}
can receive an update when a host moves to a new location.  To perform
these updates, the directory could maintain information about all
ingress switches that recently queried the directory for a route to
{\it Hz}.  However, this can require the directory to maintain a large
amount of state.  Instead, when a host moves, the directory can update
the mobile host's old local switch.  Upon receiving a message for the
mobile host, the old local switch can both forward the message to the
mobile host and send an immediate notification about the new route to
the sending switch~\cite{seattle}.  This reactive invalidation of
stale routes obviates the need for the directory to maintain
information about which ingress switches sent queries for {\it Hz},
while still ensuring rapid invalidation of stale routes.

In addition to SEATTLE, VL2, and NVP, several other designs adopt
certain aspects of the overlay solution.  
The early work on
Rbridges~\cite{rbridges}, and the resulting TRILL~\cite{rfc5556}
standard at the IETF, also forms an Ethernet overlay with
shortest-path routing in the underlay.  
However, instead of having an
explicit directory service, TRILL relies on flooding to reach hosts
with unknown routes.  
Rather than flooding on all normal overlay links, TRILL floods
on a special multicast link in the overlay.
This special link is implemented in the underlay by a multicast tree
formed on the underlay topology.

Like VL2 and NVP, the PortLand~\cite{portland} design
has a set of directory servers that allow ingress switches to learn
the route to a destination host.  Instead of encapsulating a message,
PortLand assigns each edge switch a block of host ``pseudo-MAC
addresses'' and rewrites the host MAC addresses at the edge.  To
enable the use of hierarchical pseudo-MAC addresses, PortLand is
restricted to the tree topologies common in data-center networks.
Table~\ref{tab:overlays} summarizes the structural characteristics
of all five designs.

\begin{table}
\begin{center}
\begin{tabular}{|l|l|l|} \hline
\multicolumn{1}{|c|}{\bf Protocol} &
 \multicolumn{1}{c|}{\bf Routing Directory} &
 \multicolumn{1}{c|}{\bf Encapsulation} \\ \hline
SEATTLE &
 one-hop DHT on the switches &
 simple encapsulation \\
VL2 &
 directory servers &
 simple encapsulation \\
NVP &
  directory servers &
  simple encapsulation\\
Rbridges/TRILL &
  none, flooding on multicast tree &
  simple encapsulation\\
PortLand &
 directory servers &
 none, MAC rewriting\\
\hline 
\end{tabular}
\end{center}
\caption{Ethernet overlay designs for dynamic-routing mobility.}
\label{tab:overlays}
\end{table}

\subsection{Comparative resource costs}
\label{sec:props}

Concerning storage costs, both Ethernet LANs and overlay designs
incur the costs of the
forwarding tables in switches.
These costs are kept moderate by the fact that the tables are sparsely
populated.
Because there is no aggregation of names or table entries, the costs
of densely populated tables would be too great.
In addition to the forwarding tables, the overlay designs incur a storage cost
for the routing directory, which maintains global state for the layer.

Concerning update costs, both approaches incur negligible costs for 
populating forwarding tables lazily through MAC learning.
The biggest update cost is the cost of Ethernet flooding.
The cost of flooding, in terms of bandwidth, grows quadratically with the
size of the network---which makes it a potential scalability problem.
Whether it becomes an actual problem or
or not depends on its frequency, which
depends on both the frequency of moves and the number of correspondents
that a mobile host tends to have.
SEATTLE, VL2, NVP, and PortLand have no flooding cost, though
they do have the additional cost of updating the directory.

Mobility in the overlay designs incurs no path cost.
The path cost of Ethernet mobility is significant, because the spanning
tree (which is necessitated by flooding) forces paths to be longer and
forces some physical links to go unused.

We can measure handoff latency from the instant when the mobile host
re-attaches to the network and informs its local switch (before that
time no mobility mechanism can take effect).
The following scenarios assume that a correspondent switch is sending
a steady stream of messages to a mobile host.
They describe the elapse of time before messages sent by the
correspondent switch (CS) are forwarded to the mobile host at its new
attachment.

The Ethernet scenario:
\begin{enumerate}
\item
The time-to-live of the CS's route to the mobile host expires, if it has
not already.
\item
CS receives the next message from the correspondent host and floods it.
\item
After a round trip to the mobile host, CS learns the new route.
\end{enumerate}
After Step 3, messages sent by CS are forwarded to the mobile host at
its new attachment.

In the directory-based overlay solutions ({\it i.e.,}
SEATTLE, VL2, NVP, and PortLand):
\begin{enumerate}
\item
The directory receives an update of the mobile host's new local switch.
\item
The directory informs the mobile host's old local switch of the new route.
\item
The next message arrives at the mobile host's old local switch, and
is forwarded on the new route.
\item
The mobile host's old switch also informs the CS of the new route.
\end{enumerate}
At Step 3 and after, messages sent by CS are forwarded to the mobile
host at its new attachment.
If Step 1 of the Ethernet scenario takes time, then the handoff latency
of the overlay designs will be smaller than the Ethernet's.

In addition to resource costs,
security and privacy are ever-present concerns.
In Section~\ref{sec:majordifferences} we noted that DRM usually
has minimal security problems because only routers participate
in routing.
Ethernet flooding is an exception to this rule because 
it allows hosts to play a role in routing.
Malicious
hosts can force flooding by filling up the network's forwarding tables.
(This would be accomplished by sending many messages from spoofed
source MAC addresses.)
Severe flooding can cause denial of service.
Also, the
malicious hosts will receive all the flooded packets, which
may contain private information that they wish to see.

### 2013-07-22 originals/chapter1/ebook.tex
all files compile correctly

% This work is licensed under
% http://creativecommon.org/licenses/by/3.0/
\documentclass{article}
\usepackage{graphicx}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{color}
\usepackage{url}
\hyphenation{wrap-around}

%this is a nice-looking bullet list:
%\begin{list}{$\bullet$}{
%                      \setlength{\topsep}{0in}
%                      \setlength{\itemsep}{0in}
%                      \setlength{\parsep}{0in}
%                      \setlength{\leftmargin}{.13in}
%                     }
%\item
%\end{list}

%this is a two-column figure:
%(for a one-column figure, remove *)
%figure must be in ???.pdf
%\begin{figure*}[hbt]
%\begin{center}
%\includegraphics[scale=1.00]{???}
%\end{center}
%\caption{???}
%\label{fig:???}
%\end{figure*}

%this is a two-column, floating table:
%(for a one-column table, remove *)
%(for a non-floating table, remove table environment)
%\begin{table*}
%\begin{tabular}
%\end{tabular}
%\end{table*}

%JEN: for new text added by jen.  can simply remove \textcolor{red} to
%get rid of red color, or remove the \jen{} where it appears in the text.
\newcommand{\jen}[1]{\textcolor{red}{#1}}

\title{The Design Space of Network Mobility}

\author{
Pamela Zave \\
AT\&T Laboratories---Research \\
Florham Park, New Jersey, USA \\
{\tt pamela@research.att.com}
\and
Jennifer Rexford \\
Princeton University \\
Princeton, New Jersey, USA \\
{\tt jrex@cs.princeton.edu}
}

\begin{document}
\thispagestyle{empty}
\maketitle

\begin{abstract}
While the Internet is increasingly mobile, seamless mobility is
difficult to implement at Internet scale.  Over the years, standards
bodies and the research community have introduced a large and
confusing collection of mobility proposals that are difficult to
compare.  In this tutorial, we present these mobility proposals
in a uniform framework, called the geomorphic view of networking.
The geomorphic view shows that there are two distinct patterns
for implementing mobility, each with its own range of 
design choices and cost-benefit trade-offs.
We use these patterns to classify and explain
a representative sample of mobility
mechanisms, abstractly yet precisely.
The patterns also serve as a basis for evaluating properties of
these mechanisms such as resource costs and scalability,
and for considering composition of mobility mechanisms.
\end{abstract}

\input{ebook-sec1}

\input{ebook-sec2}

\input{ebook-sec3}

\input{ebook-sec4}

\input{ebook-sec5}

\input{ebook-sec6}

\input{ebook-sec7}

\input{ebook-sec8}

\input{ebook-sec9}

\bibliographystyle{IEEEtran}
\bibliography{ebook}

\end{document}

### 2013-07-22 Compiled/chapter4/sections/Challenges.tex
all chapters compile independently

\section{Challenges in Content Delivery}\label{sec:challenges}

The challenges that CDIs and P2P systems are faced with are based on the fact
that they are unaware of the underlying network infrastructure and its
conditions. In the best case, they can try to detect and infer the topology and
state of the ISP's network through measurements, but even with large scale
measurements, it is a difficult task, especially if accuracy is necessary.
Furthermore, when it comes to short-term congestion and/or avoiding network
bottlenecks, measurements are of no use. In the following we describe the
challenges those systems face in more detail.

\subsection{Content Delivery Infrastructures (CDIs)}\label{sec:CDIs-and-Challenges}

From the viewpoint of the end-users and ISPs, the redirection schemes employed
by existing CDIs have three major limitations:

\paragraph{Network Bottlenecks.} Despite the traffic flow optimization
performed by CDIs, the assignment of end-user requests to servers by CDIs may
still result in sub-optimal content delivery performance for the end-users.
This is a consequence of the limited information CDIs have about the network
conditions between the end-user and their servers. Tracking the ever changing
conditions in networks, \ie through active measurements and end-user reports,
incurs an overhead for the CDI without a guarantee of performance improvements
for the end-user.  Without sufficient information about the network paths
between the CDI servers and the end-user, any assignment performed by the CDI
may lead to additional load on existing network bottlenecks, or to the creation
of new bottlenecks.

\paragraph{User Mis-location.} DNS requests received by the CDI DNS servers
originate from the DNS resolver of the end-user, not from the end-user itself.
The assignment is therefore based on the assumption that end-users are close to
their DNS resolvers. Recent studies have shown that in many cases this
assumption does not hold~\cite {Precise:Mao2002,DNS-IMC-2010}. As a result, the
end-user is mis-located and the server assignment is not optimal. As a response
to this issue, DNS extensions have been proposed to include the end-user IP
information~ \cite{DNS-extension-IP-client}.  \paragraph{Content Delivery Cost}
Finally, CDIs strive to minimize the overall cost of delivering huge amounts of
content to end-users. To that end, their assignment strategy is mainly driven
by economic aspects.  While a CDI will always try to assign users in such a way
that the server can deliver reasonable performance, this can again result in
end-users not being directed to the server able to deliver best performance.

\subsection{Peer-to-Peer Networks (P2P)}\label{sec:P2P-and-Challenges}

P2P traffic often starves other applications like Web traffic of
bandwidth~\cite{swxzz-hptp-07}.  This is because most P2P systems rely on
application layer routing based on an overlay topology on top of the Internet,
which is largely independent of the Internet routing and
topology~\cite{abfw-mendg-04}. This can result in a situation where a node in
Frankfurt downloads a large content file from a node in Sydney, while the same
information may be available at a node in Berlin.  As a result P2P systems use
more network resources due to traffic crossing the underlying network multiple
times. For more details and information on P2P systems, see
Section~\ref{sec:P2P}.

\subsection{Internet Service Providers (ISPs)}\label{sec:ISPs-and-Challenges}

ISPs face several challenges regarding the operation of their network
infrastructure. With the emergence of Content, and especially distributed
content delivery, be it from CDIs or P2P networks, these operational challenges
have increased manifold.

\paragraph{Network Provisioning.} Provisioning and operation a network means
running the infrastructure at its highest efficiency. To ensure this, new
cables as well as the peering points with other networks need to be established
and/or upgraded. However, with the emergence of CDIs and P2P networks, the
network provisioning has become more complicated, since the network loads tend
to shift depending on the content that is currently transported while the
direct peering might not be effective anymore.

\paragraph{Volatile Content Traffic.} CDIs and P2P networks strive to optimize
their own operational overhead, possibly at the expense of the underlying
infrastructure. In terms of CDIs, this means that a CDI chooses the best server
based on its own criteria, not knowing what parts of the networks
infrastructure is being used. Especially with globally deployed CDIs it becomes
increasingly difficult for ISPs to predict what CDI is causing what traffic
from where based on past behavior. This has a direct implication on the traffic
engineering of the network, as this is usually based on traffic predictions
from past network traffic patterns.

\paragraph{Customer Satisfaction.} Regardless of the increased difficulty with
network provisioning and traffic engineering, end-users are demanding more and
larger content. This, coupled with the dominant form of flat rates for customer
subscriptions, increases the pressure on ISPs to delay network upgrades as long
as possible to keep prices competitive. But letting links run full increases
packet loss. This, in turn, drastically reduces the quality of experience of
the end-users. This, in turn, encourages end-users to switch their
subscriptions.

\subsection{Summary}

In summary, we identify the following challenges in todays content delivery:

\begin{itemize*}

\item The ISP has limited ability to manage its traffic and therefore incurs
  potentially increased costs, e.g., for its interdomain traffic, as well as
  for its inability to do traffic engineering on its internal network while
  having to offer competitive subscriptions to its end-users.

\item The P2P system has limited ability to pick an optimal overlay topology
  and therefore provide optimal performance to its users, as it has no prior
  knowledge of the underlying Internet topology. It therefore has to either
  disregard or reverse engineer it.

\item The CDI has limited ability to pick the optimal server and therefore
  provide optimal performance to its users, as it has to infer the network
  topology as well as the dynamic network conditions. Moreover, it has limited
  knowledge about the location of the user as it only knows the IP address of
  the DNS resolver.

\item The different systems try to measure the path performance independently.

\end{itemize*}

### 2013-07-22 Compiled/chapter4/sections/Collaboration-Future.tex
all chapters compile independently

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Future of Collaboration}\label{sec:future}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\padis and \cate are designed to enable cooperation between CDI and ISPs for
the already deployed servers. Recent advances in virtualization offer CDIs
additional degree of freedom to scale-up or shrink the footprint on demand.
This can be done either by jointly deploying and operating new servers with the
ISPs.  In this section we formally introduce the design of on-demand services
motivated by the recent announcement of major ISPs to support generic
hardware-network appliances, also referred to as microdatacenters, and offer
them to application, service, and content providers. We also provide the design
and implementation of \Netpaas, a system to orchestrate the deployment of
on-demand services inside microdatacenters, by utilizing the view of the ISP
about the network and additional computation and storage resources inside the
network.

\subsection{The New Cloud: Microdatacenters Deep Inside the Network}\label{sec:generic-appliances}

Applications are increasingly relying on direct interactions with end-users and
are very sensitive to delay~\cite{ImprovingPerformanceInternet2009}. Indeed,
transaction delay is critical for online
businesses~\cite{amazon-study-about-transaction-speed}.  Network delay and loss
are important contributors.  Today, large-scale service deployments are
restricted by limited locations in the network, \eg datacenters, peering
locations, or IXPs. These locations are not necessarily ideal~\cite{CloudCmp}.
We point out that \emph{selection of service location is critical and currently
not flexible enough}.  Services should be located close enough to, in terms of
network distance, the clients.  Since client demands are volatile and change
across time, CDIs need agility~\cite{EmbrassingleDistributed}.  They can
improve their service quality by quickly allocating, de-allocating, and
migrating resources on-demand where and when they are needed. Indeed, since
delay and packet loss are among the critical metrics, the service may need to
be deployed deep inside the network, as many ISPs do for IPTV services. This
option is not yet available for non-ISP content delivery Infrastructures, \eg
for cloud services.

Currently, most services and networks are run by independent entities with
different and often conflicting objectives. Lack of information about the other
entity leads to suboptimal performance and resource allocation for both the CDI
and the ISP. For example, CDIs implement sophisticated methods to infer network
conditions to improve perceived end-user experience~\cite{Akamai-Network}, \eg
active measurements within the ISPs. Yet, the information gleaned from these
measurements is already available with far greater precision to the ISP. On the
other hand, ISPs continuously upgrade their infrastructures without being able
to efficiently engineer the CDI traffic flows~\cite{PADIS2010}. Today,
cooperation and/or partnership between providers is limited to, \eg peering or
lately direct interconnections with content delivery Infrastructures. This
level of cooperation is too narrow to reduce operational costs, improve
end-user experience, circumvent bottlenecks, handle flash crowds, and adapt to
changing network conditions and user demands. This has led to initial
discussions on how to improve communication between the various entities, \eg
within the IETF ALTO and CDNi working groups.

\subsubsection{The ISPs Proposal}\label{NFV}

To overcome the above mentioned obstacles in service deployment and operation,
major ISPs, including AT\&T, Verizon, Deutsche Telekom, Telefonica, NTT, have proposed
the use of cloud resources consisting of general purpose appliances that are
co-located at network aggregation points inside the ISP. With the convergence
of computing, storage, and communications, the acceptance of cloud services,
and the ever increasing demand for popular services, ISPs are moving towards
deploying general-purpose computing and storage infrastructures in their points
of presences (PoPs). Henceforth, we refer to these as \emph{microdatacenters}.
The description of the functionality of these microdatacenters is provided in a
white paper~\cite{NFV} that appeared in the SDN and OpenFlow World Congress in
October 2012 and signed by 13 of the largest ISPs. Microdatacenters can be also
the technical solution needed to materialize recent alliances of major CDIs,
such as Akamai with large ISPs in the area of content
delivery~\cite{content-delivery-alliance1,content-delivery-alliance2,content-delivery-alliance3}.
We notice that Software Defined Networks (SDNs) is another alternative to
redirect traffic or perform traffic engineering when applied within an ISP or
between and ISP and a CDN in cooperation.  The comparison of the two
approaches, NFV and SND, is out of the scope of this chapter and we refer the
users to the related literature on SDN
\eg~\cite{Ethane,NOX,OpenFlow,SDN-Internet-Infrastructure,B4}.

\begin{figure} \begin{center}
\includegraphics[width=0.9\linewidth]{figures/ISP+Content.eps} \end{center}
\caption{\MDCs in an ISP with \Netpaas enabled} \label{fig:story} \end{figure}

Figure~\ref{fig:story} illustrates the basic idea. The ISP can offer
\emph{slices} within its microdatacenters, that can be leased by the
CDIs---using our proposed mechanism---based on their needs. This approach
leverages recent advances in virtualization technology, and flexible billing
models, such as pay-as-you-go, to provide cost-efficient and scalable service
deployment, enabling unprecedented flexibility.  Moreover, the diversity of
available service locations within the network can be used to improve end-user
experience and makes it possible to launch even more demanding applications,
such as interactive ones. On-demand service enables CDIs to rely on a fixed
infrastructure deployment for their baseline operation and then scale it up by
dynamically allocating resources closer to end-users. It also lowers the burden
of entrance in the service market for smaller CDIs who might rely exclusively
on the \onservice at first.

\subsubsection{Microdatacenter
Specifications}\label{sec:microdatacenters-specification}

Microdatacenters consist of one or more racks of off-the-shelf hardware
deployed in general purpose rack space at network aggregation points.
State-of-the-art solutions have been proposed by the VMware/Cisco/EMC VCE
consortium \cite{VCE}, and are also offered by other vendors, such as NetApp
and Dell.  These solutions are general-purpose and provide a shared
infrastructure for a large range of applications. They typically consist of two
basic components: hardware and management software.

\begin{description*} \item [Hardware:] Typical microdatacenters include
\emph{storage}, \emph{computing}, \emph{memory}, and \emph{network access}
components.  Storage consists of tens of Terabytes with an ultra-fast
controller providing I/O throughput in the order of hundreds of Gbps. The
storage component is connected to the Internet through multi-Gbps interfaces
and to the computing component with Gigabit Ethernet switches.  Typically, a
rack includes up to 40 physical multi-core blade servers as well as two routers
and two switches in mesh configuration, for redundancy and load balancing.
\item [Management Software:] Each vendor offers a set of management tools not
only for administering the components but also to create resource slices and to
delegate the operation of the slices to external entities. This can be done
per-server or via hardware supported virtualization\footnote{ For example,
para-virtualization~\cite{virtualization} presents the VM with an abstraction
that is similar but not identical to the underlying hardware.}. The management
software is also responsible for storage allocation and handling network
resources, including IP address space. In addition, the management tools come
with a monitoring interface that allows the ISP to monitor the utilization of
the overall microdatacenter as well as the information for each slice that can
be shared with the external entity.  \end{description*}

An ISP can allocate resource slices consisting of computing, storage, memory,
and network access in a microdatacenter and then delegate the operation of the
slice to a CDI. This is what we refer to as the \emph{ISPs cloud service} which
is realized via resource slices in microdatacenters throughout the ISPs
infrastructure.

\subsubsection{Microdatacenter Network Footprint}\label{sec:microdatacenters-footptint}

Most ISPs' networks consist of an access network to provide Internet access to
DSL and/or cable customers, as well as an aggregation network for business
and/or VPN customers. Routers at this level are often referred to as {\em edge
routers}.  The access and aggregation networks are then connected to the ISP's
backbone which consists of {\em core routers}.  {\it Border routers} are core
routers that are used to connect either to other networks or to co-location
centers.  Opportunities to deploy microdatacenters exist at each level: edge,
core, or border router locations.

The advantage of deploying service infrastructure only at the core router
locations is that there are a few large and well established locations. This is
also a disadvantage as location diversity might be limited. Location diversity
is highest at the edge router locations.  However, it might not always be
possible to deploy a microdatacenter, i.e., due to limited space and/or power
at the facilities, or due to cost. These locations, however, minimize the
distance to the customers.  Border router locations are often a subset of core
routers, hence they inherit the same advantages and disadvantages.

The advantage of using an ISP cloud service vs.\ a public cloud
service for a CDI is the chance to minimize the distance to the
end-user. \ondemandbase allows the CDI to control the
location of the slices and ensures that there are no major network
bottlenecks.

\subsection{\OnService Design}\label{sec:service-design}

An {\it \onservice} is a service of the ISP (see Figure~\ref{fig:story}) that
enables CDIs to use a hosting infrastructure that scales according to end-user
demands, so as to minimize its capital expenditures and operating costs, as
well as the distance between its hosting infrastructure and the source of the
demand.  Moreover, it offers an interface that enables the CDI to map user
requests to appropriate slices in order to maximize slice utilization and
minimize the distance between the end-user and the slices.

{\bf Definition 1: ISP \OnService.}  The ISP \onservice is a service offered by
the ISP and uses as its base unit of resource allocation the notion of a
microdatacenter \emph{slice}.  It is the ISP's task to allocate/de-allocate the
slices since it operates the microdatacenter.  The CDI requests slices based on
its clients demand. When the slice is allocated to the CDI, the service can be
installed on the slice. From that point on, the CDI fully controls the
operation of the service installed in the microdatacenter. Negotiation about
slices are done via the \emph{\onservice interface} through which CDI demands
are matched to the ISPs resources. How to map demands to resources in an
efficient manner is the task of the ISP and part of the \emph{\onservice
realization}.  In addition, the interface allows for access to the billing
information.  Moreover, the \emph{\onservice interface} enables the mapping of
user requests to appropriate slices.

The above mentioned use of microdatacenters is in-line with the available
primitives of private and public clouds operated in large-scale datacenters,
\eg~\cite{amazon,azure}.

\subsubsection{Microdatacenter Slice}\label{sec:Slice}

Based on our description of microdatacenters in the previous sections, we define
a slice as follows.

{\bf Definition 2: Slice.} The slice of a microdatacenter is a set of
physical or virtualized resources of a specific capacity, for each of the
resources of the microdatacenter. The slice is delegated to the service
provider that can install and operate its service using the resources of
the slice.

For example, a slice can be a 1-core server with 2 GB RAM, 30 GB storage, a
1 Gbps Internet access bandwidth, 2 public IPs---an actual physical resource.
Alternatively, it can be a VServer with 2GB and 1 Gbps Internet
access bandwidth, 1 public IP, and a pre-installed OS---a virtual machine
of a specific type.  With the current management and virtualization tools
available from microdatacenter vendors, it is possible to
allocate/deallocate slices on-demand in with unprecedented degree of
freedom, \eg~\cite{aboveclouds} and references within.

\subsubsection{\OnService Realization}\label{sec:Realization}

Based on the above specification of \onservice, the ISP has to implement two
functions to offer its \onserviceemph: \emph{mapping of service pro\-vi\-der
demands to slices} and \emph{assigning users to slices}.

Note, the time scales at which these two services are expected to be used
differ significantly. The first one allows the service provider to flexibly
allocate and de-allocate its slices based on its forecast of demands, in those
locations where it wants them.  We foresee that requests for slices are not
issued individually but rather collectively on a time scale of tens of minutes
or hours.

The CDI provides the ISP with a set of demands for slice resources, predicted
demand locations, desired slice locations, as well as optimization criteria.
The ISP then has to map the demands to its microdatacenter resources.  We
expect that the major degree of freedom that the ISP uses to jointly optimize
performance is the desired slice location. We refer to this optimization
problem as the \sliceallocation problem. If the desired slice locations are
fully specified or the predicted demand locations are missing, the
\sliceallocation problem becomes trivial and the ISP only grants or denies the
request.

At the second time scale, the ISP can help the CDI in assigning users to
slices. Since the service is offered at multiple locations, a good assignment
of users to slices impacts not only the load on the network but also the
network delay and packet loss, which are key contributors to the user
experience. Jointly optimizing this mapping is therefore of interest to both
the ISP and the CDI. The CDI can query the ISP for each request on where to map
it, based on the current set of slice assignments and service loads. The ISP
then uses its network information to propose possible slices. We refer to this
problem as the \sliceassignment problem, see \cite{Cate-CCR}.

Another degree of freedom \onservice offers to the CDI is auto-scaling.  While
it is quite feasible to dimension applications, flash-crowds or device failures
are hard to predict. To this end, a CDI may allow \onservice to create replicas
if its monitoring indicates that the capacity of the service at a given
location is or will be exceeded.  To realize this service, the ISP needs to
constantly monitor resource availability and if necessary migrate or suggest
the creation of additional slices. Moreover, it has to allow the CDI to monitor
the utilization of its slices.

\subsubsection{Service Interfaces}\label{sec:service-interfaces}

The ISP offers four interfaces to the content delivery Infrastructures:

\begin{description*}

\item [Resource discovery:] Using this interface the CDI requests information
about resources, e.g., about available locations for slices and if in principle
slices are available at those locations at what price.

\item [Slice allocation:] Using this interface the CDI requests slice
allocation within a certain cost limit.

\item [User-slice assignment:] Using this interface the CDI requests
recommendations for user demand to slice mapping.

\item [Monitoring and billing:] Using this interface the CDI monitors the
status and cost of its slices.

\end{description*}

In Section~\ref{sec:future-netpaas} we give specific examples of how these
service interfaces can be used by a CDI and ISP to cooperate in order to
improve their services.

\subsubsection{Billing}\label{sec:billing}

It is important for the CDI to minimize and track the cost of its use of
\onservice.  Depending on the scale of the services, the service provider has
to pay the usual price or negotiate bilateral agreements with the ISP. Using
the resource discovery interface, it estimates the cost of slice allocation at
possible locations. Using the slice allocation interface, it can bound the
total cost of the request.

We expect that the billing of a slice allocated via \onservice follows that of
large-scale datacenters.  This means that there is an installation cost and a
usage cost.  The installation cost applies to a single slice in a
microdatacenter and is charged only once or over long time intervals, \eg
hours, and is fixed. The installation cost typically increases if additional
licenses have to be leased, \eg software licenses. The installation cost can
depend on the location of the microdatacenter that hosts the slice or the
time-of-day.

The usage cost follows a pay-as-you-go billing model and charges for the usage
of different resources assigned to a slice.  The billing among different
resources in the same slice can be quite diverse. The slice can use expensive
resources such as bandwidth or cheaper ones such as CPU.

For example, a slice may have a $\$ 0.01$ per hour installation cost and a
usage cost that depends on its use of various resources, \eg $\$ 0.02$ per real
CPU usage per hour, $\$ 0.001$ per GByte stored per hour, and $\$ 0.001$ per
Gbps outgoing traffic per hour.  If the slice is idle, then only the
installation cost is charged. Note, that if the slice is used for a short
period within the allocation time, \eg a few minutes, then the charge may apply
to the minimum billing granularity.

To minimize the cost of deploying an on-demand service, the CDI can change its
total slice demands as well as its slice specifications dynamically. Moreover,
it can relax the slice specifications to reduce overall cost of its service
deployment.

\subsection{Network Platform as a Service (NetPaaS)}\label{sec:future-netpaas}

Next, we discuss the prototype system that has been proposed to materialize the
On-demand service, Network Platform as a Service (NetPaaS). NetPaaS leverages
the view of PaDIS and also utilize the knowledge about the status of the
microdatacenters within the network. NetPaaS is also able to map the requests
of CDIs to available microdatacenters to better match the demand with the
resources inside the network.  The granularity at which they are exchanged via
the service interface.  We also outline several possible protocols for the
service interfaces. We focus on resource discovery, slice allocation, and
user-slice assignment. We do not discuss monitoring and billing because they
can be realized today using techniques similar to those in use by current
public clouds, \eg~\cite{aboveclouds}. Due to space limitations, we refer the
reader to \cite{NetPaaS} for a formalization of NetPaaS, as well as an
evaluation on a CDI use case.

Recall that our assumption that the time scales at which the two principle
components of \onservice operate are different. On the one hand, resource
discovery and slice allocation are expected to be done on time scales of tens
of minutes, hours, or even days. On the other hand, user-slice assignment
potentially happens on a per user request basis. Accordingly, the protocols
differ. We propose to use out-of-band protocols for the first two service
interfaces and in-band protocols for the third one.

\subsubsection{Resource Discovery}\label{sec:Resource-Discovery}

The purpose of resource discovery is to provide the CDI with the ability to
gather information about the resources offered by the ISP. Accordingly, we
have two message types: CDI\_Discovery\_Request and ISP\_Discovery\_Re\-sponse.

\begin{description*}

\item [CDI\_Discovery\_Request:] Is issued either without and with arguments.
In the first case the response is the set of resources that are offered. In the
second case the responds contains details about the resources named in the
argument.

\item [ISP\_Discovery\_Response:] List of available resource or details about
the resources specified in the argument.

\end{description*}

So far we have not outlined at what granularity and specificity the resources
are requested. This depends on the agreements between the CDI and the ISP.  For
example, the ISP may have no problem revealing its microdatacenter locations to
a major CDI. However, it may not want to share this information with an
untrusted CDI that wants to run a single slice. For the latter, the region in
which the microdatacenter is located might well suffice.

With regards to granularity, the ISP can specify which type of servers it is
offering in each microdatacenter region, as is common for public cloud
services~\cite{amazon}, unless another agreement is in place that enables
access to more specific information.  With regards to the availability and/or
price, the ISP can either return a base price, including installation and usage
cost, to indicate that resources are available or offer an auction-based
system. In the latter case, the discovery request returns information about
past prices.

\subsubsection{Slice Allocation}\label{sec:system-slice-allocation}

\begin{figure}
    \begin{center}
    \includegraphics[width=0.6\linewidth]{figures/Resource_Negotiation.eps}
    \end{center}
    \vspace*{-1em}
    \caption{Slice allocation message exchange.}
    \label{fig:ResourceAllocation}
\end{figure}

Slice allocation enables the CDI and ISP to cooperate for allocating slices in
microdatacenters close to the end-user that are able to satisfy the demands.
We envision five message types: CDI\_Demand\_Request, ISP\_Demand\_Response,
CDI\_ Slice\_Request, ISP\_Slice\_Response, CDI\_Slice\_Commit. The first two
message types enable the cooperation between the CDI and the ISP to allocate
slices at appropriate microdatacenter by utilizing network information, see
Figure~\ref{fig:ResourceAllocation}. The last three messages enable a three way
handshake between the CDI and the ISP to verify that the ISP is able to provide
a specific slice for the CDI.

\begin{description*}

\item [CDI\_Demand\_Request:] Is submitted by the CDI to the ISP and contains a
summary of the hardware resources the CDI wants together with optimization
criteria, constraints, and a demand forecast, \eg per region or per network
prefix.  Possible optimization criteria are to minimize network distance or the
cost. The constrains include: number of locations, minimum resources per slice,
etc.

\item [ISP\_Demand\_Response:] The ISP returns a set of proposed slice
configurations and their price. It computes these by solving the
\sliceallocation problem.

\item [CDI\_Slice\_Request:] The CDI either selects, based on its criteria, a
set of proposed slices as returned by the ISP\_De\-mand\_Response, or it
completes a specification of a slice set request using information it retrieved
via the resource discovery interface. In addition, the request contains a
maximum cost.

\item [ISP\_Slice\_Response:] Upon receiving a CDI\_Slice\_Request, the ISP
checks if it can offer the set of slices at the requested cost. This involves
solving another version of the \sliceallocation problem. If possible, the ISP
returns the price otherwise it declines the request.  At this step, the ISP
reserves the requested resources to guarantee their availability. Note, the ISP
does not have to return precise slice definitions, \eg instead of returning
that slice x should be located in microdatacenter~y attached to router~z it
only returns slice~x should be located in region~xyz.

\item [CDI\_Slice\_Commit:] This step confirms CDI\_Slice\_Requests. Upon
receiving the commit from the CDI, the ISP allocates the slices and delegates
their control to the CDI.

\end{description*}

Now we discuss different ways in which a CDI and an ISP can cooperate using the
above messages. These ways differ in which information is shared and with whom.

\begin{description*}

\item [Minimum information exchange:] The CDI uses the information from the
ISP\_Demand\_Response for queries via CDI\_Demand\_Request with a uniform
distributed demand vector. The responses include slice candidates with servers
having specified hardware profiles and in specific regions. Then, the CDI
scales the suggested slices according to its demand locations and uses the
CDI\_Slice\_Request message to check if the ISP can offer it and at what price.
Once it has found a satisfactory configuration it can use the
CDI\_Slice\_Commit message to request the necessary slices.

\item [Partnership 1:] The CDI uses CDI\_Demand\_Request with a scaled demand
CDI selects one of these and uses the ISP\_Slice\_Request message so that the
ISP can reserve the resources. Upon successful reservation, the
CDI\_Slice\_Commit message confirms the allocation of the slices.

\item [Partnership 2:] The CDI uses the CDI\_Demand\_Request without a demand
vector but with specific resource requests. The ISP response specifies
candidate microdatacenters with available resources.  Then, the CDI uses its
version of the \sliceallocation problem to find possible slice sets at a subset
of these locations. Then, the CDI uses the ISP\_Slice\_Request message to see
if the ISP can offer it and at what price. Once it finds a satisfactory
configuration it uses the CDI\_Slice\_Commit message to allocate the slices.

\end{description*}

The first scenario corresponds to the minimum information that has to be
exchanged in order to reach a consensus on the locations and specification of
the slices. The latter two summarize different forms of possible cooperations
that can be agreed upon in bilateral agreements.

So far, we have assumed that there are no preallocated slices. However, this is
typically not the case, and the actual task is to augment a preexisting set of
slices in such a way as to best serve the predicted demand for the next time
period. To enable this, another argument to each message request can be
provided, indicating a set of already allocated resources and a penalty value
for deallocating slices in one location and allocating them in another.  This
penalty is needed as part of the optimization problem.  Basically, it indicates
up to which point it is preferable to keep a suboptimal location because of
stability of resource allocation vs.\ when to migrate the service to a new
location.  Based on the returned information, the CDI has the option of either
moving slices using VM migration or to de-allocate and allocate new slices.

The ISP microdatacenter can offer VM migration and/or
consolidation\footnote{Here, consolidation corresponds to moving multiple VMs
with minimal resource requirements to the same physical machine to keep a base
service.} with keeping the IP addresses only within the same microdatacenter
location. Across microdatacenters it may only offer migration with tunnels
which requires the CDI to temporarily operate two slices at both locations.
However, the old one is a good candidate for consolidation so that it is
possible to reduce the allocated resources to a minimum within a
microdatacenter once all new requests are served by the newly allocated slices.
Thus, if an ISP offers service consolidation, one option for CDIs that want to
use diverse sets of microdatacenters is to always keep a minimal slice active
at each location and expand or shrink it according to the demand.

\subsubsection{User-Slice Assignment}\label{sec:system-user-assignment}

\begin{figure}[tbp]
    \begin{center}
    \includegraphics[width=0.8\linewidth]{figures/Mapping_System.eps}
    \end{center}
    \vspace*{-1em}
    \caption{User-slice assignment schematic.}
    \label{fig:UserSliceAllocation}
\end{figure}

The purpose of the user-slice assignment interface is to enable small time
scale interactions between the CDI and the ISP, to ensure that end-user demand
is mapped to the appropriate slice.  Therefore, the interface has to be
integrated into the process used by the CDI to map user requests to CDI
servers.  Next, we first review how this is currently realized using DNS. Then,
we discuss options on how the user-slice assignment interface can be
integrated, see Figure~\ref{fig:UserSliceAllocation}.

{\bf CDI: User request to CDI server mapping.} Before an end-user issues a
request for the CDI service, \eg downloading some content or watching a video,
it issues a DNS request to map the hostname to an IP address of the server that
hosts the service.  This DNS request is sent to the ISPs DNS resolver or an
alternative DNS resolver. This resolver contacts the authoritative DNS server
of the CDI service, since caching is typically restricted by small TTL's. The
authoritative DNS server uses a CDI service, the CDI mapping system, to select
a CDI server from which to satisfy the future requests of this end-user. The
CDI mapping system performs a CDI specific optimization.  This optimization may
consider the load on the CDI servers, the network location of the CDI server as
well as the requesting DNS server, the price of network access at the CDI
server, etc. Note, the CDI's authoritative DNS name servers usually does not
have access to the IP address of the end-user as the request is forwarded via
the DNS resolver unless the eDNS~\cite{DNS-extension-IP-client} standard is
used.  With eDNS, the client IP address or the IP prefix can be added to the
DNS request sent by the DNS resolver to the authoritative DNS name server. In
addition, the CDI can use HTTP redirection to further load balance within its
infrastructure.

{\bf User-Slice Assignment: Option 1.} Considering the process outlined above,
one possible way to use the user-slice assignment interface is within the
optimization that the CDI mapping system performs.  For this case, we envision
two message types: CDI\_User-Slice\_Assign\_Re\-quest and
ICDI\_User-Slice\_Assign\_Response which correspond to steps 3 and 4 in
Figure~\ref{fig:UserSliceAllocation}.

\begin{description*}

\item [CDI\_User-Slice\_Assign\_Request:] Issued by the CDI's DNS server to the
  ISP. It contains the client IP address as well as slice locations within or
  outside of the ISP.

\item [ISP\_User-Slice\_Assign\_Response:] The ISP responses with a ranking of
  the slice locations.

\end{description*}

The previous two messages enable the CDI to consider information from the ISP,
conveyed via the ranking, for its mapping. This is equivalent to the
functionality and protocols proposed by the IETF ALTO working group.  However,
we envision a more light weight implementation. We propose to not rely on XML
for encoding each request if the granularity of the requests is on a per
connection level. If the CDI uses a coarser granularity such as subnet or
region, the efficiency of the message exchange is even less critical.\\

{\bf User-Slice Assignment: Option 2.}
Another way to integrate the user-slice assignment interface within
the above process is by pro-actively sharing information between the CDI
and the ISP.  For this case, we again envision two message types:
ISP\_User-Slice\_Assign\_Pro\-po\-sal and CDI\_User-Slice\_Assign\_Ack.

\begin{description*}

\item [ISP\_User-Slice\_Assign\_Proposal:] Is sent by the ISP to the CDI
  mapping system. It contains a set of IP prefixes each with an associated
  ranking of the different microdatacenter locations.

\item [CDI\_User-Slice\_Assign\_Ack:] The CDI either acknowledges or rejects
  the proposal.

\end{description*}

This again enables the CDI to include information from the ISP, conveyed via
the ranking, in its mapping.  For example, one can use BGP to send such
messages---a mechanism Akamai already utilizes to aid in mapping users to
its clusters~\cite{Akamai-Network}.

\subsection{Summary}

Motivated by the high requirements newly deployed services put on the ISPs
networks, we formally introduced the design of on-demand services to relieve
both the CDI and the ISP.  We also provided the design and implementation of
\Netpaas, a system to orchestrate the deployment of on-demand services using
the ISPs view of the network and available resources. In~\cite{NetPaaS} we
evaluate \Netpaas using operational traces from the biggest commercial CDN and
a European Tier-1 ISP. We quantify the benefits of \Netpaas by utilizing
different optimization goals \eg delay reduction or reduced link utilization.
Our results show that \Netpaas yields encouraging results for a joint
deployment of new services that significantly improves the end-users
performance while reducing the network utilization for the ISP and offering
agile service deployment to the CDI. For the analysis and performance evaluation of
on-demand server placement algorithms in wide-area networks we refer the reader
to~\cite{ToN-dFL} and \cite{SimultaneousSourceLocation}.

### 2013-07-22 Compiled/chapter4/sections/Collaboration-Use-Cases.tex
all chapters compile independently

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Collaboration Use Cases: P2P and TE}\label{sec:Oracle-and-TE}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The growth of demand for content is motivating collaboration between ISPs and
applications. In this chapter we review to use cases: P2P and Traffic Engineering.

\subsection{Use Case: P2P}\label{sec:oracle}

Recall, P2P systems are self-organizing systems of autonomous entities, called
peers, that cooperate for common goals. These common goals range from sharing
of resources, e.g., music and video files, processing power, or storage
space~\cite{P2P-LNCS-05} to collaborative routing as in Skype and P2P-TV.  A
fundamental characteristic of these systems is their distributed designs and
resources.

Advantages of P2P systems include elimination of bottlenecks and single
points-of-failure within the system, increased processing power, high
availability/redundancy, and little or no dependence on any particular central
entity.  However, P2P systems are plagued by some fundamental issues, such as
overlay/underlay topological and routing mismatch~\cite{sa-oidronl-06},
inefficiencies in locating and retrieving resources, and scalability and
performance issues caused by uncontrolled traffic swamps~\cite{P2P-LNCS-05}.

Several of these drawbacks can be addressed by collaboration between the P2P
overlay and the Internet routing underlay. To overcome these limits each ISP
can offer the ``oracle'' service as introduced in
Section~\ref{sec:p2p-collaboration} to the P2P users which explicitly helps P2P
users to choose ``good'' neighbors.  The P2P user can supply its ISP's oracle
with a list of possible P2P neighbors, during bootstrapping and/or content
exchange.  The ISP's oracle then returns a ranked list to the querying user,
according to its preference (\eg AS-hop distance) and knowledge of the ISP
topology and traffic volume, while at the same time keeping the interest of the
P2P user in mind.  We  show that in principle, P2P systems as well as the ISPs
profit from the use of the oracle even when only considering the AS-distance
for ranking nodes~\cite{afs-cispp2pcip-ccr07}, because the overlay topology is now
localized and respects the underlying Internet topology, and the P2P user
profits from the ISP's knowledge.

To study the impact of biased neighbor selection on a real P2P network that
implements its own routing, we run extensive simulations of the Gnutella
protocol. We show that in contrast to the unmodified P2P system, the ISP-aided
localized P2P system shows consistent improvements in the observed end-user
experience, measured in terms of content download times, network locality of
query responses and desired content, and quality of query responses. A
significantly large portion of P2P traffic remains local to the ISP network,
and ISPs notice a substantial reduction in overall P2P traffic. This can lead
to immense cost savings for the ISPs~\cite{Cachelogic}. The oracle consistently
shows performance gains even across different topologies under a broad range of
user behavior scenarios. For a more detailed analysis of the P2P oracle
service, see~\cite{afs-cispp2pcip-ccr07, af-ispabqsp2pt-eccs07,afk-icsecispp2p-icin07}.

\subsubsection{Influence on P2P Topology}\label{sec:Influence-Oracle-on-P2P-Topology}

To explore the influence of consulting the oracle on the network topology we
visualize, in Figure~\ref{fig:oracle:gnutella_topology}~\cite{tashev-thesis},
the Gnutella overlay topology. At a particular instant in time, we sample the
Gnutella overlay topology, display all the online nodes in the graph, and join
two nodes with an edge if there exists a Gnutella peering between them at this
point of time. The resulting graph structures are displayed in
Figure~\ref{fig:oracle:gnutella_topology}. For our simulations we consider $5$
different topologies: Germany, USA, World1, World2 and World3, each modeled
after their respective AS topologies (World1-3 differ in the size of the ASes).
We can easily observe that the Gnutella topology in the biased case is well
correlated with the Internet AS topology, where the nodes within an AS form a
dense cluster, with only a few connections going to nodes in other ASes. This
is in stark contrast to the unbiased Gnutella graph, where no such property can
be observed. Multiple runs of the above experiments, using the different
topologies yield similar results.

\begin{figure}[tbp]
   \begin{subfigure}[]{0.4\linewidth}
     {\includegraphics[angle=-90,width=1\linewidth]{figures/snapshot_nond.eps}}
   \caption{Unbiased Gnutella}%
   \end{subfigure}
\hfill
   \begin{subfigure}[]{0.45\linewidth}
     {\includegraphics[angle=-90,width=1\linewidth]{figures/snapshot_nd.eps}}
   \caption{Gnutella with Oracle}%
   \end{subfigure}   \caption{Visualization of Gnutella overlay topology}
   \label{fig:oracle:gnutella_topology}
\end{figure}

\subsubsection{Benefits of Collaboration}\label{sec:Benefits-Collaboration-P2P-results}

\noindent\textbf{Mean AS distance:} \\
The benefits of using an oracle for biasing the neighborhood in Gnutella are visible in
Figure~\ref{fig:oracle:gnutella_metric:2a}, which shows the average AS distance (in the underlay)
between any two connected Gnutella nodes. The AS distance is obtained as follows. We map
each Gnutella node's IP address to its parent AS, and for each overlay edge, we find the
network distance in AS hops between the two end-nodes.  We observe that the least amount
of decrease in the average AS distance occurs from $1.93$ to $0.8$ at $1000$ seconds, and
the maximum decrease from $1.94$ to $0.25$ happens at $5000$ seconds. Given that the AS
diameter remains constant at $4$ hops, the average decrease of $1.45$ in the AS distance
is significant. Besides, as the average AS distance in the case of oracle list size of
$1000$ is $0.45$, a value less than $1$, it implies that most of the Gnutella peerings are
indeed within the ASes, i.e., they are not crossing AS boundaries. This can be a major
relief for ISPs, as they do not incur any additional cost for traffic within their
domains. Also traffic that does not leave the network is easier to manage. Moreover, P2P
traffic will not encounter inter-ISP bottlenecks.

\begin{figure}[tbp]
	\begin{subfigure}[]{0.32\linewidth}
     {\includegraphics[width=1\textwidth]{figures/asdist.eps}}
    \caption{Mean AS distance in Underlay\label{fig:oracle:gnutella_metric:2a}}
    \end{subfigure}
 \hfill
    \begin{subfigure}[]{0.32\linewidth}
    {\includegraphics[width=1\textwidth]{figures/intraAS-leaf.eps}}
    \caption{Intra-AS peerings (\%) for Leaf nodes\label{fig:oracle:gnutella_metric:2b}}
    \end{subfigure}
 \hfill
    \begin{subfigure}[]{0.32\linewidth}
    {\includegraphics[width=1\textwidth]{figures/intraAS-up.eps}}
    \caption{Intra-AS peerings (\%) for Ultrapeers\label{fig:oracle:gnutella_metric:2c}}
    \end{subfigure}
    \caption{Metrics for Gnutella simulations}
   \label{fig:oracle:gnutella_metric}
\end{figure}

\vspace{0.75\baselineskip}
\noindent\textbf{Intra-AS P2P connections:} \\
The above observations on AS distance are even better understood from the plots in
Figures~\ref{fig:oracle:gnutella_metric:2b} and \ref{fig:oracle:gnutella_metric:2c}, where we show the total number of intra-AS
P2P connections in the Gnutella network as a percentage of the total number of intra- and
inter-AS P2P connections, for both leafs and ultrapeers.

In Figure~\ref{fig:oracle:gnutella_metric:2b}, we observe that in the case of leaf nodes, taking
the average over the $10$ time points, the percentage of intra-AS P2P connections
increases from $14.6$\% in unbiased case to $47.88$\% in the case of oracle with list size
$100$. For oracle with list size $1000$, we note an average of $82.22$\% intra-AS P2P
connections.

In Figure~\ref{fig:oracle:gnutella_metric:2c}, we observe similar results for ultrapeers. The
percentage of intra-AS P2P connections increases from an average value of $14.54$\% in the
unbiased case to $38.04$\% in the case of oracle with list size $100$, and further to
$74.95$\% in case of oracle with list size $1000$.

The percentage increase in intra-AS P2P connections is larger for leaf nodes as compared
to ultrapeers, a welcome development.  One needs a certain number of inter-AS connections,
to maintain network connectivity and to be able to search for file content that may not be
available within an AS. However, as leaf nodes typically have poor connectivity to the
Internet, and have lower uptimes, it is reasonable to have leaf nodes keep most of their
peerings within their AS, while allowing the ultrapeers to have slightly more connections
outside their ASes.

For the impact of Oracle on download time under different topologies we refer
the reader to~\cite{improving-oracle-GI-2008}. For the impact of Oracle-like
localization techniques on the inter-AS traffic flow of the BitTorrent P2P
system we refer the reader to~\cite{DeepDiving,LocalityP2P,BlindMiceBT,Impact-Profitability}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Use Case: Traffic Engineering}\label{sec:te-cate}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The growth of demand for content and the resulting deployment of content
delivery infrastructures pose new challenges to CDIs and to ISPs. For CDIs, the
cost of deploying and maintaining such a massive infrastructure has
significantly increased during the last years~\cite{AkamaiCutting:2009} and the
revenue from delivering traffic to end-users has decreased due to the intense
competition. Furthermore, CDIs struggle to engineer and manage their
infrastructures, replicate content based on end-user demand, and assign users
to appropriate servers.

The latter is challenging as end-user to server assignment is based on
inaccurate end-user location
information~\cite{Precise:Mao2002,DNS-extension-IP-client}, and inferring the
network conditions within an ISP without direct information from the network is
difficult. Moreover, due to highly distributed server deployment and adaptive
server assignment, the traffic injected by CDIs is volatile. For example, if
one of its locations is overloaded, a CDI will re-assign end-users to other
locations, resulting in large traffic shifts in the ISP network within minutes.
Current traffic engineering by ISP networks adapts the routing and operates on
time scales of several hours, and is therefore too slow to react to rapid
traffic changes caused by CDIs.

The pressure for cost reduction and customer satisfaction that both CDIs and
ISPs are confronted with, coupled with the opportunity that distributed server
infrastructures offer, motivate us to propose a new tool in the traffic
engineering landscape. We introduce \emph{Content-aware Traffic Engineering}
(\cate). \cate leverages the location diversity offered by CDIs and, through
this, it allows to adapt to traffic demand shifts. In fact, \cate relies on the
observation that by selecting an appropriate server among those available to
deliver the content, the path of the traffic in the network can be influenced
in a desired way. Figure~\ref{fig:flowSelection} illustrates the basic concept
of \cate. The content requested by the client is in principle available from
three servers (A, B, and C) in the network. However, the client only connects
to one of the network locations. Today, the decision of where the client will
connect to is solely done by the CDI and is partially based on measurements
and/or inference of network information and end-user location. With \cate the
decision on end-user to server assignment can be done jointly between the CDI
and ISP.

\begin{figure}[tbp]
  \center\includegraphics[width=0.8\linewidth]{figures/trafficShift-concept.eps}
  \caption{By choosing a CDI server for a client with the help of CaTE, traffic
  engineering goals and accurate end-user server assignment become possible.}
  \label{fig:flowSelection}
  \vspace{-1.5em}
\end{figure}

%%%%
\subsubsection{The CaTE Approach}\label{sec:cate}
%%%%

\cate complements existing traffic engineering solutions
\cite{ietf-alto-protocol,TECDN,CooperativeISPCDN,BeyondMLU,CooperativeISPCDN:Workshop,p4p}
by focusing on traffic demands rather than routing.  Let {\bf y} be the vector
of traffic counts on links and {\bf x} the vector of traffic counts in
origin-destination (OD) flows in the ISP network.  Then {\bf y}=$A${\bf x},
where $A$ is the routing matrix. $A_{ij}=1$ if the OD flow $i$ traverses link
$j$ and $0$ otherwise.  Traditional traffic engineering is the process of
adjusting $A$, given the OD flows {\bf x}, so as to influence the link traffic
{\bf y} in a desirable way.  In \cate, we revisit traffic engineering by
focusing on traffic demands rather than routing changes. Content-aware Traffic
Engineering (\cate) is thus the process of adjusting the traffic demand vector
{\bf x}, without changing the routing matrix $A$, so as to change the link traffic {\bf y}
in a desirable way.

\cate offers additional traffic engineering capabilities to both ISPs and CDNs
to better manage the volatility of content demand in small time scales.
Traditional traffic engineering
\cite{ietf-alto-protocol,TECDN,CooperativeISPCDN,BeyondMLU,CooperativeISPCDN:Workshop,p4p}
relies on changes of routing weights that take place in the time scale of hours
\cite{FT01}.  On the contrary, in \cate, the redirection of end-users to servers can take place
per request or within the TTL of a DNS query that is typically tens of seconds
in large CDNs \cite{PADIS2010}.  Thanks to the online recommendations by ISP
networks, CDNs gain the ability to better assign end-users to servers and
better amortize the deployment and maintenance cost of their infrastructure.
Network bottlenecks are also circumvented and thus the ISP operation is
improved.  Furthermore, the burden of measuring and inferring network topology,
and the state of the network, both challenging problems, is removed from the
CDNs. Moreover, in \cite[Sections 4 and 5]{CaTE-TR} we show that the online
\cate decisions on the end-user to server assignment leads to optimal traffic
assignment within the network under a number of different metrics.  The
advantage is that now the problem of assigning traffic to links reduces to a
fractional solution (on the contrary, assigning routing weights to links is
NP-hard).  In short, all involved parties, including the end-users, benefit
from \cate, creating a win-win situation for everyone.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{A Prototype to Support CaTE}\label{sec:system}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\cate relies on a close collaboration between CDN and ISP in small time scales
(seconds or per request).  To achieve this goal, network information has to be
collected and processed by the ISP.  Candidate CDN servers have to be
communicated to the ISP and ranked based on a commonly agreed criteria, \eg to
optimize the delay between the end-user and the CDN server.  Today, there is no
system to support the above operations.  This motivate us to design, implement
and evaluate a novel and scalable system that can support \cate.  In this
section we describe the architecture and deployment of our working prototype to
enable \cate.  We start by presenting our prototype in
Section~\ref{sec:architecture}.  We then  comment on its operation and
deployment within the ISP, its interaction with a CDN, and its performance that
is beyond the state-of-the-art \cite{ietf-alto-protocol}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\ \\\noindent\textbf{Architecture:}\label{sec:architecture}\\\noindent
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The \cate system is installed in an ISP and interacts with the existing CDN
server selector. The main tasks of the \cate system are to: (1) maintain an
up-to-date annotated map of the ISP network and its properties, (2) produce
preference rankings based on the paths between end-users and candidate servers,
and (3) communicate with the CDN server selection system to influence the
assignment of end-user to servers.  To this end, we propose an architecture
that comprises a {\it Network Monitoring} component, a {\it Query Processing}
component and a {\it communication interface} between an ISP and a CDN.  For an
overview of the architecture see Figure~\ref{fig:CaTE-architecture}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\ \\\noindent\textbf{Network Monitoring:}\label{sec:Network-monitoring}\\\noindent
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The network monitoring component gathers information about the topology and the
state of the network from several sources to maintain an up-to-date view of the
network.  The network monitoring component consists of the following
subcomponents:

The {\bf Topology Information} component gathers detailed information about the
basic network topology, \ie routers and links, as well as annotations such as
link utilization, router load, and topological changes.  An Interior Gateway
Protocol (IGP) listener provides up-to-date link-state (\ie IS-IS, OSPF)
information.  Information about routers and links is retrieved, thus, the
network topology can be extracted.  The nominal link delay, \ie the latency on
a link without queuing, can be found through the link length and physical
technology.  The link utilization and other metrics can be retrieved via SNMP
from the routers or an SNMP aggregator.

The {\bf Connectivity Information} component uses routing information to
calculate the paths that traffic takes through the network. Finding the path of
egress traffic can be done by using a Border Gateway Protocol (BGP) listener.
Ingress points of traffic into the ISP network can be found by utilizing
Netflow data. This allows for complete forward and reverse path mapping inside
the ISP.  Furthermore, the system can map customers as well as CDN
infrastructures into the network map by finding the routers that announce the
address space associated with them. In total, this allows for a complete path
map between any two points in the ISP network.  Finally, our system has access
to an uplink database that provides information about the connectivity
statistics of end-users.

\begin{figure}[tbp]
    \center \includegraphics[width=3.2in]{figures/CaTE_Schematic.eps}
  \vspace{-0.1in}
    \caption{CaTE System architecture and flow of messages.}
    \label{fig:CaTE-architecture}
    \vspace{-0.2in}
\end{figure}

The {\bf Network Map Database} component processes the information collected by
the {\it Topology} and {\it Connectivity Information} components to build an
annotated network map of the ISP network tailored towards fast lookup on path
properties.  It uses a layer of indirection to keep the more volatile
information learned from BGP separate from the slower changing topological
information.  This allows address space to be quickly reassigned without any
reprocessing of routing or path information.  It also enables pre-calculation
of path properties for all paths that yields a constant database lookup
complexity independent of path length and network architecture.  If topology
changes, \eg IGP weights change or a link fails, the {\it Topology Information}
component immediately updates the database which only recalculates the
properties of the affected paths.  Having ISP-centric information ready for
fast access in a database ensures timely responses and high query throughput.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\ \\\noindent\textbf{Query Processing:}\label{sec:Query-Processing}\\\noindent
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The {\bf Query Processing} component receives a description of a request for
content from the CDN, which specifies the end-user making the request and a
list of candidate CDN servers.  It then uses information from the {\it Network
  Map Database} and a selected ranking function to rank the candidate servers.
This component consists of the following subcomponents:

The {\bf Query Processor} receives the query from the CDN.  First, the query
processor maps each source-destination (server to end-user) pair to a path in
the network.  In most cases, the end-user is seen as the ISP DNS resolver,
unless both ISP and CDN support the client IP eDNS
extension~\cite{DNS-extension-IP-client}.  Once the path is found, the
properties of the path are retrieved. Next, the pairs are run individually
through the location ranker subcomponent (see below) to get a preference value.
Finally, the list is sorted by preference values, the values are stripped from
the list, and it is sent back to the CDN.

The {\bf Location Ranker} component computes the preference value for
individual source-destination pairs based on the source-destina\-tion path
properties and an appropriate function.  Which function to use depends on (a)
the CDN, (b) what metrics the CDN asked for and (c) the optimization goal of
the ISP.  The preference value for each source-destination pair is then handed
back to the Query Processor.  Multiple such optimization functions being
defined upon the collaboration agreement and subsequently selected individually
in each ranking request. For example, a function might be the minimization of
end-user and server delay.  In~\cite{CaTE-TR} we evaluate \cate
with multiple ranking functions for different optimization goals.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\ \\\noindent\textbf{Communication Interfaces:}\label{sec:Communication-interfaces}\\\noindent
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
When a CDN receives a content request, the {\it Server Selector} needs to
choose a content server to fulfill this request. We propose that the server
selector sends the list of eligible content servers along with the source of
the query and an optimization goal to the ISP's \cate system to obtain
additional guidance about the underlying network.  If the guidance is at
granularity of a single DNS request, we propose a DNS-like protocol using UDP
to prevent extra overhead for connection management.  If the granularity is at
a coarser level, \ie seconds or even minutes, we rely on TCP.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Privacy and Performance}\label{sec:discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
During the exchange of messages, none of the parties is revealing any sensitive
operational information.  CDNs only reveal the candidate servers that can
respond to a given request without any additional operational information (\eg
CDN server load, cost of delivery or any reason why a server is chosen). The
set of candidate servers can be updated per request or within a TTL that is
typically in the order of a tens of seconds in popular CDNs \cite{PADIS2010}.
On the other side, the ISP does not reveal any operational information or the
preference weights it uses for the ranking. In fact, the ISP only re-orders a
list of candidate servers provided by the CDN.  This approach differs
significantly from \cite{ietf-alto-protocol,p4p}, where partial or complete
ISP network information, routing weights, or ranking scores are publicly
available.  We argue that an important aspect to improve content delivery is to
rely on up-to-date information during server selection of the CDN.  This also
eliminates the need of CDNs to perform active measurements to infer the
conditions within the ISP that can add overhead to CDN operation and may be
inaccurate.  With \cate, the final decision is still made by the CDN, yet it is
augmented with up-to-date network guidance from the ISP.

To improve the performance of our system, we do not rely on XML-based network
maps as proposed in~\cite{ietf-alto-protocol}, but on light protocols that are
close to DNS in design.  This design choice is important as topology
information in large networks (in the order of multiple MBytes). Transferring
this information periodically to many end-users is likely to be challenging. In
a single instance of our system, we manage to reply to up to $90,000$
queries/sec when 50 candidate servers supplied by the CDN. At this level, the
performance of our system is comparable to that of current DNS servers, such as
BIND. However, the number of replies drops to around $15,000$ per second when
considering $350$ candidate servers. The additional response time when our
system is used is around $1$ ms when the number of candidate servers is $50$
and around $4$ ms when considering $350$ candidate servers. This overhead is
small compared to the DNS resolution time~\cite{DNS-IMC-2010}. The performance
was achieved on a commodity dual-quad core server with 32 GB of RAM and 1GBit
Ethernet interfaces. Furthermore, running additional servers does not require
any synchronization between them.  Thus, multiple servers can be located in
different places inside the network.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\ \\\noindent\textbf{Deployment:}\label{sec:deployment:}\\\noindent
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Deploying the system inside the ISP network does not require any change in the
network configuration or ISP DNS operation.  Our system solely relies on
protocol listeners and access to ISP network information. Moreover, no
installation of special software is required by end-users.  The \cate system
adds minimal overhead to ISPs and CDNs.  It only requires the installation of a
server in both sides to facilitate communication between them.

Typically, an ISP operates a number of DNS resolvers to better balance the load
of DNS requests and to locate DNS servers closer to end-users. To this end, we
envision that the ISP's \cate servers can be co-located with DNS resolvers in
order to scale in the same fashion as DNS.  \cate servers can also be located
close to peering points in order to reduce the latency between the CDN and an
instance of the system. Synchronization of multiple \cate instances is not
necessary as they are aware of the state of the same network.  We concluded
that this is the best deployment strategy, other possible deployment strategies
we have considered are presented in \cite{CaTE-TR}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\ \\\noindent\textbf{Operation:}\label{sec:operation:}\\\noindent
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We now describe the operation of our working prototype and its interaction with
the CDN.  In Figure~\ref{fig:CaTE-architecture} we illustrate the basic system
architecture to support \cate including the flow of information when the \cate
system is used. When a DNS request is submitted by an end-user to the ISP DNS
resolvers~{\it(1)} there are a number of recursive steps~{\it(2)} until the
authoritative DNS server is found~{\it(3)}.  Then, the ISP DNS resolver
contacts the authoritative DNS server~{\it(4)}.  There, the request is handed
to the content request processor operated by the CDN query processing
component~{\it(5)}. The content request processor has access to full
information about the status of the CDN.  Based on the operational status of
the CDN servers, the server selection system \cite{Akamai-Network} is
responsible for choosing eligible content servers~{\it(6)}.  In the end, a
preference list of content servers is generated. At this point, the CDN server
selector sends the list of eligible content servers~{\it(A)} along with user
information, such as the IP of the DNS resolvers or client and an optimization
metric to ISP.  The query processor of the ISP system ranks the list using the
location ranker {\it(B)}.  After all the elements have been processed, the
query processor has an annotated list with preferences for the ISP~{\it(C)}.
The query processor sorts the list by the preference values, strips the values
and sends the list back to the CDN~{\it(D)}. The CDN server selector
incorporates the feedback, selects the best content server(s) and hand them
back to the content request processor {\it(7)}.  Then, the answer travels the
path back to the client, \i.e. from the CDN's authoritative DNS server~{\it(8)}
via the ISP DNS resolver {\it(9)} to the end-user {\it(10)}.  Finally, the
end-user contacts the selected server {\it(11)} and downloads the content
{\it(12)}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Modeling CaTE}\label{sec:model-CaTE}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Next, we formalize \cate and discuss how it relates to traditional traffic
engineering and multipath routing.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\ \\\noindent\textbf{Architecture:}\label{sec:Network-Model-Architecture}\\\noindent
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We model the network as a directed graph $G(V,E)$ where $V$ is the set of nodes
and $E$ is the set of links. An origin-destination (OD) flow $f_{od}$ consists
of all traffic entering the network at a given point $o \in V$ (origin) and
exiting the network at some point $d \in V$ (destination).  The traffic on a
link is the superposition of all OD flows that traverse the link.

The relationship between link and OD flow traffic is expressed by the routing
matrix $A$. The matrix $A$ has size $|E|$ $\times$ $|V|^2$. Each element of
matrix $A$ has a boolean value.  $A_{ml}=1$ if OD flow $m$ traverses link $l$,
and $0$ otherwise. The routing matrix $A$ can be derived from routing
protocols, e.g., OSPF, ISIS, BGP. Typically, $A$ is very sparse since each OD
flow traverses only a very small number of links. Let {\bf y} be a vector of
size $|E|$ with traffic counts on links and {\bf x} a vector of size $|V|^2$
with traffic counts in OD flows, then {\bf y}=$A${\bf x}. Note, {\bf x} is the
vector representation of the traffic matrix.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallskip
\noindent{\bf Traditional Traffic Engineering:}\label{sec:TE-Definitions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In its broadest sense, traffic engineering encompasses the application of
technology and scientific principles to the measurement, characterization,
modeling, and control of Internet traffic~\cite{Awduche_OverviewTE:2002}.
Traditionally, traffic engineering reduces to controlling and optimizing the
routing function and to steering traffic through the network in the most
effective way.  Translated into the above matrix form, traffic engineering is
the process of adjusting $A$, given the OD flows {\bf x}, so as to influence
the link traffic {\bf y} in a desirable way, as coined
in~\cite{LakhinaSigmetrics2004}. The above definition assumes that the OD flow
vector {\bf x} is known. For instance, direct observations can be obtained, \eg
with Netflow data~\cite{Cisco_Netflow:99,TrafficDemand:ToN2001}.

\begin{figure}[tbp]
\center\includegraphics[width=0.90\linewidth]{figures/trafficShift-illustration.eps}
\caption{Content-aware Traffic Engineering Process}
\label{fig:Content-aware-illustration2}
\vspace{-1.5em}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallskip
\noindent {\bf Terminology:}\label{sec:Terminology}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We denote as {\it flow} an OD flow between two routers in the network.  We call
a flow {\it splittable} if arbitrarily small pieces of the flow can be assigned
to other flows.  This is not to be confused with end-to-end sessions, \ie TCP
connections, which are \emph{un-splittable}.  The assumption that flows are
splittable is reasonable, as the percentage of traffic of a single end-to-end
session is small compared to that of a flow between routers. Let $C$ be the set
of nominal capacities of the links in the network $G$. We denote as {\it link
utilization} the fraction of the link capacity that is used by flows. We denote
as {\it flow utilization} the maximum link utilization among all links that a
flow traverses. We introduce the terms of {\it traffic consumer} and {\it
traffic producer} which refer to the aggregated demand of users attached to a
router, and the CDIs that are responsible for the traffic respectively. We
refer to the different alternatives from which content can be supplied by a
given CDI as \emph{network locations} that host servers.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\ \\\noindent\textbf{Definition of CaTE:}\label{sec:CaTE-Definitions}\\\noindent
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We revisit traffic engineering by focusing on the traffic demands rather than
changing the routing.

\noindent \textbf{Definition 1: Content-aware Traffic Engineering(\cate)} is
the process of adjusting the traffic demand vector {\bf x}, given a routing
matrix $A$, so as to change the link traffic {\bf y}.

Not all the traffic can be adjusted arbitrarily. Only traffic for which
location diversity is available can be adjusted by \cate.  Therefore, {\bf
  x}={\bf x$_r$}+{\bf x$_s$} where {\bf x$_r$} denotes the content demands that
can be adjusted and {\bf x$_s$} denotes the content demands that can not be
adjusted as there is only a single location in the network where the content
can be downloaded from.  The amount of traffic that can be adjusted depends on
the diversity of locations from which the content can be obtained.  We can
rewrite the relation between traffic counts on links and traffic counts in
flows as follows: {\bf y}=$A$({\bf x$_s$} $+$ {\bf x$_r$}).  \cate adjusts the
traffic on each link of the network by adjusting the content demands {\bf
  x$_r$}: {\bf y$_r$}=$A${\bf x$_r$}. Applying \cate means adjusting the
content demand to satisfy a traffic engineering goal.

\noindent \textbf{Definition 2: Optimal Traffic Matrix} is the new traffic
matrix, {\bf x}$^*$, after applying \cate, given a network topology $G$, a
routing matrix $A$ and an initial traffic matrix {\bf x}.

Figure~\ref{fig:Content-aware-illustration2} illustrates the \cate process.  A
content consumer requests content that three different servers can deliver. Let
us assume that, without \cate, the CDI redirects the clients to servers B and
C.  Unfortunately, the resulting traffic crosses a highly-utilized link. With
\cate, content can also be downloaded from server A, thus, the traffic within
the network is better balanced as the highly utilized link is circumvented.

Minimizing the maximum utilization across all links in a network is a popular
traffic engineering goal~\cite{FT00,FT01,ImprovingPerformanceInternet2009}.  It
potentially improves the quality of experience and postpones the need for
capacity increase. \cate mitigates bottlenecks and minimizes the maximum link
utilization by re-assigning parts of the traffic traversing heavily loaded
paths. Thus it redirects traffic to other, less utilized paths.  Later in this
chapter, we will elaborate in Section~\ref{sec:impact}, different metrics such
as path length or network delay can also be used in \cate.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\ \\\noindent\textbf{CaTE and Traditional TE:}\label{sec:CaTE-TE}\\\noindent
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\cate is complementary to routing-based traffic engineering as it does not
modify the routing. Routing-based traffic engineering adjusts routing weights
to adapt to traffic matrix changes. To avoid micro-loops during IGP
convergence~\cite{transient-IGP}, it is common practice to only adjust a small
number of routing weights~\cite{FT01}. To limit the number of changes in
routing weights, routing-based traffic engineering relies on traffic matrices
computed over long time periods and offline estimation of the routing weights.
Therefore, routing-based traffic engineering operates on time scales of hours,
which can be too slow to react to rapid change of traffic demands.  \cate
complements routing-based traffic engineering and can influence flows at
shorter time scales by assigning clients to servers on a per request
basis. Thus, \cate influences the traffic within a network online in a
fine-grained fashion.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\ \\\noindent\textbf{CaTE and Multipath Routing:}\label{sec:CaTE-Multipath}\\\noindent
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Multipath routing helps end-hosts to increase and control their upload
capacity~\cite{PathSelection:CACM2011}. It can be used to minimize transit
costs~\cite{Optimizing:Goldenberg2004}. Multipath also enables ASes to
dynamically distribute the load inside networks in the presence of volatile and
hard to predict traffic demand
changes~\cite{TrafficDemand:ToN2001,MATE2001,TeXCP,Replex}.  This is a
significant advantage, as routing-based traffic engineering can be too slow to
react to phenomena such as flash crowds. Multipath takes advantage of the
diversity of paths to better distribute traffic.

\cate also leverages the path diversity, and can be advantageously combined
with multipath to further improve traffic engineering and end-user
performance. One of the advantages of \cate is its limited investments in
hardware deployed within an ISP. It can be realized with no change to routers,
contrary to some of the previous multipath
proposals~\cite{TeXCP,MATE2001,Replex}.  The overhead of \cate is also limited
as no state about individual TCP connections needs to be maintained, contrary
to multipath~\cite{TeXCP,MATE2001,Replex}.  In contrast
to~\cite{MATE2001,TeXCP}, \cate is not restricted to MPLS-like solutions and is
easily deployable in today's networks.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\ \\\noindent\textbf{CaTE and Oscillations:}\label{sec:CaTE-Oscilations}\\\noindent
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Theoretical results~\cite{AdaptiveRouting2005,Convergence:Fischer2006} have shown that
load balancing algorithms can take advantage of multipath while provably avoiding traffic
oscillations. In addition, their convergence is fast. Building on these theoretical results, Fischer
et al. proposed REPLEX~\cite{Replex}, a dynamic traffic engineering algorithm that exploits
the fact that there are multiple paths to a destination. It dynamically changes the traffic load
routed on each path. Extensive simulations show that REPLEX leads to fast convergence, without
oscillations, even when there is lag between consecutive updates about the state of the network.
\cate is derived from the same principles and thus inherits all the above-mentioned desired properties.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Potential of Collaboration}\label{sec:impact}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In this section, we quantify the potential benefits of \cate when deployed
within an European Tier-1 ISP using operational data.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\noindent\textbf{Experimental Setting:}\label{sec:Experimental-Setting}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
To evaluate \cate, an understanding of the studied ISP network is necessary,
including its topological properties and their implications on the flow of
traffic.  Indeed, the topological properties of the ISP network influence the
availability of disjoint paths, which are key to benefit from the
load-balancing ability of \cate.  Because \cate influences traffic aggregates
inside the ISP network at the granularity of requests directed to CDIs,
fine-grained traffic statistics are necessary.  Traffic counts per-OD flow,
often used in the literature, are too coarse an input for \cate.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\noindent\textbf{Data from a Large European ISP:}\label{sec:Residential-Traces}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
To build fine-grained traffic demands, we rely on anonymized packet-level
traces of residential DSL connections from a large European Tier-1 ISP,
henceforth called \emph{ISP1}. For ISP1, we have the complete annotated
router-level topology including the router locations as well as all public and
private peerings. ISP1 contains more than $650$ routers and $30$ peering points
all over the world.  Using the same monitoring infrastructure as in
Section~\ref{sec:traces}, we collect a $10$ days long trace of HTTP and DNS
traffic starting on May 7, 2010. We observe 720 million DNS messages as well as
more than 1 billion HTTP requests involving about 1.4 million unique hostnames,
representing more than 35 TBytes of data. We note that more than 65\% of the
traffic volume is due to HTTP.

A large fraction of the traffic in the Internet is due to large CDIs, including
CDNs, hyper-giants, and OCHs, as reported in earlier
studies~\cite{TrafficTypesGrowth:2011,arbor,PADIS2010}. In
Figure~\ref{fig:CDF-Traffic-CPs}, we plot the cumulative fraction of HTTP
traffic volume as a function of the CDIs that originate the traffic. For this,
we regard a CDI as a organizational unit where all servers from the distributed
infrastructure serve the same content, such as Akamai or Google.  We rank the
CDIs by decreasing traffic volume observed in our trace. Note that the x-axis
uses a logarithmic scale. The top 10 CDIs are responsible for around 40\% of
the HTTP traffic volume and the top 100 CDIs for close to 70\% of the HTTP
traffic volume. The marginal increase of traffic is diminishing when increasing
the number of CDIs. This shows that collaborating directly with a small number
of large CDIs, can yield significant savings.

In Figure~\ref{fig:CPs-Total-Traffic} we plot the traffic of the top 1, 10, 100
CDIs by volume as well as the total traffic over time normalized to the peak
traffic in our dataset.  For illustrative purposes, we show the evolution
across the first $60$ hours of our trace. A strong diurnal pattern of traffic
activity is observed. We again observe that a small number of CDIs are
responsible for about half of the traffic. Similar observations are made for
the rest of the trace.

\begin{figure}[tbp]
\center\includegraphics[width=1\linewidth]{figures/providersVsVolume.eps}
\caption{CDF of traffic volume of CDIs in ISP1.}
\label{fig:CDF-Traffic-CPs}
\vspace{-1.5em}
\end{figure}

\begin{figure}[tbp]
\center\includegraphics[width=1\linewidth]{figures/Volume_timeseries.eps}
\caption{Normalized traffic for top CDIs by volume in ISP1.}
\label{fig:CPs-Total-Traffic}
\vspace{-1.5em}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\noindent\textbf{Understanding the Location Diversity of CDIs:}\label{sec:Traffic-by-Content-Provider}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
To achieve traffic engineering goals, it is crucial to also understand the
location diversity of the top CDIs, as \cate relies on the fact that the same
content is available at multiple locations.  Traffic originated from multiple
network locations by a given CDI is seen by \cate as a single atomic traffic
aggregate to be engineered. Furthermore, as routing in the Internet works per
prefix, we assume that the granularity of subnets is the finest at which \cate
should engineer the traffic demand. Thus, we differentiate candidate locations
of CDIs by their subnets and quantify the location diversity of CDIs through the
number of subnets from which content can be obtained.

We examine the amount of location diversity offered by CDIs based on traces from
ISP1.  To identify the subnets of individual CDIs, we rely on a similar
methodology to the one from Poese \etal~\cite{PADIS2010}. Our granularity is
comparable to their "infrastructure redirection
aggregation". Figure~\ref{fig:VolumeChoices} shows the cumulative fraction of
HTTP traffic as a function of the number of subnets (logarithmic scale) from
which a given content can be obtained, over the entire $10$ days of the
trace. We observe that more than $50\%$ of the HTTP traffic can be delivered
from at least $8$ different subnets, and more than $60\%$ of the HTTP traffic
from more than $3$ locations. These results confirm the observations made
in~\cite{PADIS2010}.

\begin{figure}[tbp]
\center\includegraphics[width=1\linewidth]{figures/subnetVsVolumeBucket.eps}
\caption{Subnet diversity from which content is available.}
\label{fig:VolumeChoices}
\vspace{-1.5em}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\noindent\textbf{Dynamics in Location Diversity:}\label{sec:Server-Dynamics}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
So far the location diversity of CDIs has been evaluated irrespective of time.
To complement the finding, we turn our attention to the location diversity
exposed by CDIs at small time-scales, \ie in the order of minutes.  To this
end, we split the original trace into $10$ minutes bins.
Figure~\ref{fig:TemporalEffect} shows the evolution of the number of exposed
subnets of five of the top 10 CDIs by volume.  Note that the diversity exposed
by some CDIs exhibits explicit time of day patterns, while others do not. This
can be due to the structural setup or the type of content served by the CDI.
The exposed location diversity patterns, \ie flat or diurnal, are
representative for all CDIs with a major traffic share in our trace. We
conclude that a significant location diversity is exposed by popular CDIs at
any point in time, and is quite extensive during the peak hour.

\begin{figure}[tbp]
\center
\includegraphics[width=1\linewidth]{figures/SubnetChoice_r-2.eps}
\caption{Evolution over time of number of subnets for selected CDIs in the top 10 CDIs.}
\label{fig:TemporalEffect}
\vspace{-1.5em}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\noindent\textbf{Content Demand Generation:}\label{sec:TM}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The location diversity is not a mere observation about CDIs deployment. It
requires to revisit the mapping between a given content demand and the realized
traffic matrix. Given the location diversity for content, multiple traffic
matrices can be realized from a given content demand. The standard view of the
OD flows therefore provides an incomplete picture of the options available for
\cate.

As an input for \cate, we introduce an abstraction of the demand that reflects
the available location diversity.  We rely on the notion of \emph{potential
  vectors}, that were denoted as $x_{r}$ in Section~\ref{sec:CaTE-Definitions}.
To generate the potential vector for a given CDI, the amount of traffic this CDI
originates as well as the potential ingress points need to be known. Combining
all potential vectors and $x_{s}$, we synthesize a network-wide content demand
matrix for each time bin, by scaling the traffic demand to match the network
utilization of ISP1.  For our evaluation, we use the series of content demand
matrices over a period of $10$ days. The content demands are based exclusively
on the HTTP traffic of our trace.

\subsection{Summary}

In this section we presented the potential of the \emph{Oracle} and
\emph{Content-aware Traffic Engineering} (\cate), two collaborative approach to
improve content delivery while achieving traffic engineering goals.  Both
leverage location diversity offered by P2P systems or CDIs. Moreover, \cate
enables dynamic adaption to traffic demand shifts. With \cate/the oracle the
decision on end-user to server/peer assignment can be done jointly between the
CDI/P2P system and the ISP.  Our analysis of operational data from a European
Tier-1 ISP has shown ample opportunities for \cate to improve content delivery
as it is done today.

Through extensive experiments, we show that both P2P users and ISPs benefit
from ISP-aided P2P locality, measured in terms of improved content download
times, increased network locality of query responses and desired content, and
overall reduction in P2P traffic. For a more detailed analysis of the possible
improvements and additional background information on the parameter set and
resulting improvements, we refer the reader to~\cite{afs-cispp2pcip-ccr07,A-IANSPS-08}.

In~\cite{Cate-CCR, CaTE-TR, Sigmetrics2012} we quantify the benefits of \cate
and consider one of the most popular traffic engineering goals, namely
minimizing the maximum utilization of the links in the
network~\cite{FT00,FT01}.  Our evaluation shows that \cate yields encouraging
results, even when only a few large CDIs are collaborating with an ISP. In fact,
even metrics that are not directly related to the optimization function of
\cate are improved. Besides significant improvements for the operation of ISP
networks, the end-users to also benefit from these gains. This can be
attributed to the decrease of delay as well as the reduced link
utilization. In~\cite{CaTE-TR} we also consider other network metrics such as
path length or path delay and the effect of other network topologies.  We also
outline how \cate can aid in the deployment of popular large scale
applications, \eg NetFlix, by selecting strategic locations for caches and
specific optimization goals to support their operation. With this we conclude
the section on collaborative traffic engineering and continue to elaborate on
our idea for ``in-network server deployment'' in the next chapter.

### 2013-07-22 Compiled/chapter4/sections/Conclusion.tex
all chapters compile independently

\section{Conclusion}\label{sec:conclusion}

People value the Internet for the content and the applications it makes
available~\cite{CCN}. For example, the demand for online entertainment and web
browsing has exceeded 70\% of the peak downstream traffic in the United
States~\cite{sandvine}. Recent traffic
studies~\cite{TrafficTypesGrowth:2011,arbor,PADIS2010} show that a large
fraction of Internet traffic is originated by a small number of Content
Distribution Infrastructures (CDIs). Major CDIs include highly popular
rich-media sites such as YouTube and Netflix, One-Click Hosters (OCHs), \eg
RapidShare, Content Delivery Networks (CDNs) such as Akamai, and hyper-giants,
\eg Google and Yahoo!. Gerber and Doverspike~\cite{TrafficTypesGrowth:2011}
report that a few CDIs account for more than half of the traffic in a US-based
Tier-1 carrier.

To cope with the increasing demand for content, CDIs deploy massively
distributed infrastructures~\cite{ImprovingPerformanceInternet2009} to
replicate content and make it accessible from different locations in the
Internet~\cite{CDNsec2009,Cartography}.  Not all CDIs are built upon the same
philosophy, design, or technology. For example, a CDI can be operated
independently by deploying caches in different networks, by renting space in
datacenters, or by building datacenters. Furthermore, some CDIs are operated by
ISPs, some by Content Producers, or in the case of Peer-to-Peer networks, by
self-organized end users. Accordingly, we give an overview of the spectrum of
CDI solutions.

CDIs often struggle in mapping users to the best server, regardless of whether
the best server is the closest, the one with the most capacity, or the one
providing the lowest delay.  The inability of CDIs to map end users to the
right server stems from the fact that CDIs have limited visibility into ISP
networks, \ie a CDI has incomplete knowledge of an ISP's set up, operation, and
current state.  Thus, in this book chapter, we propose viewing the challenges
that CDIs and ISPs face as an opportunity: \emph{to collaborate}.  We point out
the opportunities and incentives for all parties---CDIs, ISPs and end
users---to get involved. This collaboration may ultimately lead to major
changes in the way that content is distributed across the Internet.

Accordingly, we review the proposed enablers and building blocks for
collaboration ranging from the P2P oracle service, P4P, Ono, and PaDIS, to the
IETF activities~\cite{afs-cispp2pcip-ccr07,p4p,taming,PADIS2010,ALTO-WG,CDNi}. To
illustrate the benefits of collaboration between applications and networks, we
provide two use-cases: P2P and traffic engineering. The main take away is that
substantial benefits for all involved parties are obtainable.

Upcoming trends include virtualization and the Cloud. These trends offer new
ways of collaborative deployment of content delivery infrastructure if combined
with the proposed enablers for collaboration.  Accordingly, we propose Network
Platform as a Service (\Netpaas), which allows CDIs and ISPs to cooperate not
only on user assignment, but on dynamically deploying and removing servers and
thus scaling content delivery infrastructure on demand.

We believe that ISP-CDN collaboration and \Netpaas can play a significant role
in the future content delivery ecosystem.  Most of the collaboration enablers,
however, have not yet been deployed in the wild, and therefore only the future
will tell if the Internet takes advantage of these opportunities.

### 2013-07-22 Compiled/chapter4/sections/Content-Delivery-Landscape.tex
all chapters compile independently

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Content Delivery: The Landscape}\label{sec:content-delivery}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Internet traffic grows at a rate of approximately 30\% per
year~\cite{cisco-study} and is dominated by the delivery of content to end
users~\cite{IXPSIGCOMM2012,TrafficTypesGrowth:2011,arbor,PADIS2010}. To cope
with the increasing demand for content, and to support the level of reliability
and scalability required by commercial-grade applications, Content Distribution
Infrastructures (CDIs) have emerged. In general terms, CDIs are overlays built
on top of existing network infrastructures that aim to accelerate the delivery
of content to end-users. CDIs include, but are not limited to, Content
Distribution Networks (CDNs), such as Akamai and Google, Video Streaming
Portals (VSP) such as YouTube, One-Click-Hosters (OCH) like Rapidshare and
MegaUpload. However, a CDI does not necessarily produce the content that it
delivers. Thus, we define a Content Producer (CP) as the entity that generates
content. In some cases, \eg Google and YouTube, the CP and CDI can be the same
entity. In other instances, for example Akamai and Limelight, the CDI only
delivers what a CP pays for.

But not all CDIs are built upon the same philosophy, designs and technology.
For example, a CDI can be operated independently by deploying caches in
different networks, by renting space in datacenters or by building its own
datacenters. Furthermore, some CDIs are operated by ISPs, by Content Producers,
or in the case of Peer-to-Peer networks, by self-organized end-users. To
summarize the spectrum of CDI solutions, Figure~\ref{fig:cdn-spectrum} provides
an overview of different CDI solutions. They are aligned by their architectures
according to which parties are involved.

\begin{figure}[tbp]
    \begin{center}
    \includegraphics[width=0.7\linewidth]{figures/spectrum.eps}
    \end{center}
    \caption{Spectrum of content delivery solutions and involvement of stake-holders in the content delivery.}
    \label{fig:cdn-spectrum}
\end{figure}

\subsection{Independent Content Distribution}

Independent CDIs are usually referred to as Content Delivery Networks (CDNs).
They have a strong customer base of content producers and are responsible for
delivering the content of their customers to end-users around the world.
Today, they are, by traffic volume as well as hosted content, the largest
players on the Internet. In general, there are four main components to
independent CDN architectures: a server deployment, a strategy for replicating
content on servers, a mechanism for directing users to servers, and a system
for collecting and processing server logs.

For server deployment, three main approaches
exist~\cite{ImprovingPerformanceInternet2009}: centralized, datacenter based
and distributed infrastructures:

\textbf{Central Location}: This approach is used by small CDNs, One-Click
Hosters, and applications running in public clouds. Centralized hosting takes
advantage of (a) the economies of scale that a single location offers~
\cite{aboveclouds}, (b) the flexibility that multihoming
offers~\cite{Optimizing:Goldenberg2004}, and (c) the connectivity opportunities
that IXPs offer~\cite{IXPSIGCOMM2012}. The disadvantages of centralized hosting
are the potential for a single point of failure, and the limited ability to
ensure low latency to users located in different networks around the
world~\cite {CloudCmp}.

\textbf{Datacenter Based}: This approach deploys in several large data centers.
It again leverages economies of scale while improving reliability and creating
a larger footprint with further reach. However, by utilizing multiple
datacenters, new challenges regarding the content distribution, synchronization
and delivery arise. For example, the datacenter delivering content to an
end-user cannot be statically configured anymore, but the selection needs to
take the location of the end-user into account. This approach is used by CDNs
such as Limelight, EdgeCast and BitGravity. Many cloud providers also use this
approach, including Amazon CloudFront and Microsoft Azure.

\textbf{Distributed Infrastructures}: This approach consists of a highly
distributed infrastructure deployment, potentially deep inside third-party
networks. Here, the large number of servers scattered across numerous networks
offer high availability and replication of content while being very close to
end-users. Furthermore, this type can balance traffic across locations, best
react to flash crowds by dynamic server assignments, and deliver content with
improved latency. However, with the highly distributed infrastructures, the
challenges of assigning users to the right server location increase many-fold.
Also, with deep deployment datacenters are usually not available anymore,
leading to the question where to deploy how many servers. Today, Akamai is only
one independent CDN that uses this approach on a global scale.

\vspace{0.5em} CDNs with more than one location typically follow a pull
strategy~\cite{Akamai-Network} for content distribution and replication. Thus,
content requests can be directed to servers that do not have the required
object cached.  When a requested object is not at the selected server,
neighboring servers in the same cluster or region are asked. If the object is
not available at neighboring servers, the origin or root server responsible for
the object is contacted to retrieve the content. A requested object that is
fetched from a remote server is saved locally and then delivered to the end
user. To keep the copies of the object fresh, a TTL value is assigned to it.
When the TTL value expires, the object is removed. For scalability reasons, any
server of the CDN or within a region can respond to the request of an end
user~\cite{CDNsec2009}.

A special case of the independent CDI category are free CDNs such as
Coral~\cite{CoralCDN}, which follow a similar architectural design. In these
CDNs, server resources are offered by end-users or non-profit organizations.

\subsection{ISP-operated CDIs} The potential for generating revenue from
content delivery has motivated a number of ISPs to build and operate their own
Content Distribution Infrastructures. For example, large ISPs such as AT\&T and
Verizon have built their own CDNs along the same general architectural
principles as independent CDIs. However, due to the limitations arising from
being restricted to one network, these CDNs are not deployed in a distributed
fashion across multiple networks and thus are not globally operating solutions.
To overcome this issue, the CDNi group at the IETF~\cite {CDNi} is discussing
how to interconnect these CDNs to boost their efficiency and coverage. The
content provider are third parties, applications and services offered by the
ISP. Other ISPs with large footprints, such as Level3 and Telefonica~ \cite
{interdatacenter,ToN-DTB}, have also built CDNs in order to efficiently
transfer content across the globe and offer improved services to their end
users.

\subsection{Emerging Trends in CDI Architectures} Economics, especially cost
reduction, is the key driving force behind emerging CDI architectures. The
content delivery market has become highly competitive. While the demand for
content delivery services is rising and the cost of bandwidth is decreasing,
the profit margins of storage and processing \cite{aboveclouds} are dwindling,
increasing the pressure on CDIs to reduce costs. At the same time, more parties
are entering the market in new ways, looking to capture a slice of the revenue.
However, today's traditional CDI deployments lack agility to combat these
effects.  Contracts for server deployments last for months or years and the
available locations are typically limited to datacenters. The time required to
install a new server today is in the order of weeks or months. Such timescales
are too large to react to changes in demand. CDIs are therefore looking for new
ways to expand or shrink their capacity, on demand, and especially at low cost.

\subsubsection{Hybrid Content Distribution} In a hybrid CDI, end-users download
client software that assists with content distribution. As in P2P file-sharing
systems, the content is broken into pieces and offered by both other users who
have installed the client software as well as by the CDI's servers. The client
software contacts dedicated CDI servers, called control plane servers, which
schedule which parts of the content are to be downloaded from what peers.
Criteria for selecting peers include AS-level proximity as well as the
availability of the content. If no close peers are found, or if the download
process from other peers significantly slows the content delivery process, the
traditional CDI servers take over the content delivery job entirely. Akamai
already offers NetSession~\cite{HybridCDN-Bruce}, a hybrid CDI solution for
delivering very large files such as software updates at lower cost to its
customers.  Xunlei~\cite{Xunlei}, an application aggregator with high
penetration in China, follows a similar paradigm.  It is used to download
various types of files including videos, executables, and even emails, and
supports popular protocols such as HTTP, FTP, and RTSP.  Xunlei maintains its
own trackers and servers. A study of hybrid CDIs~\cite{HybridCDN-Ross} showed
that up to 80\% of content delivery traffic can be outsourced from server-based
delivery to end-users, without significant degradation in total download time.

\subsubsection{Licensed CDNs} Licensed CDNs have been proposed to combine the
benefits of the large content-provider customer base of an independent CDI with
the end-user base of an ISP~\cite{LicensedCDN}.  A licensed CDN is a
partnership between an independent CDI and an ISP. The CDI licenses the content
delivery software that runs on servers to the ISP while the ISP owns and
operates the servers. The servers deliver content to the end-users and report
logging information back to the CDI.  The revenue derived from content
producers is then shared between the two parties. Thus, a CDI can expand its
footprint deep inside an ISP network without investing in hardware, incurring
lower operating costs.  The ISP benefits from not having to invest in
developing the software for a reliable and scalable content distribution. More
importantly, a licensed CDN also alleviates the ISP's need to negotiate
directly with content producers, which might be challenging, given an ISPs
limited footprint.

\subsubsection{Application-based CDIs} Recently, large application and content
producers have rolled out their own CDIs, hosted in multiple large data
centers. Some popular applications generate so much traffic that the content
producers can better amortize delivery costs by doing content distribution
themselves.  Google is one such example.  It has deployed a number of data
centers and interconnected them with high speed backbone networks.  Google
connects its datacenters to a large number of ISPs via IXPs and also via
private peering.  Google has also launched the Google Global Cache
(GGC)~\cite{GoogleCache}, which can be installed inside ISP networks.  The GGC
reduces the transit cost of small ISPs and those that are located in areas with
limited connectivity, \eg Africa. The GGC servers are given for free to the
ISPs which install and maintain them.  GGC also allows an ISP to advertise
through BGP the prefixes of users that each GGC server should serve.  As
another example, Netflix, which is responsible for around 30\% of the traffic
in North America at certain times, is also rolling out its own CDI. The Netflix
system is called Open Connect Network~\cite{NetflixCDN}. Netflix offers an
interface where ISPs can advertise, via BGP, their preferences as to which
subnets are served by which Open Connect Network servers.

\subsubsection{Meta-CDIs} Today, content producers contract with multiple CDIs
to deliver their content.  To optimize for cost and
performance~\cite{Optimizing:SIGCOMM2012}, meta-CDIs act as brokers to help
with CDI selection. These brokers collect performance metrics from a number of
end-users and try to estimate the best CDI, based on the server that a user is
assigned. To this end, the brokers place a small file on a number of CDIs. Then
they embed the request for this file in popular websites' source code, in the
form of a javascript. When users visit these sites, they report back statistics
based on the servers that each CDI assigned the users. The broker then
recommends CDIs for a given source of demand taking also into consideration the
cost of delivery. Cedexis is one of these brokers for web browsing.  Another
broker for video streaming is Conviva~\cite{Conviva2011}. These brokers may
compensate when a CDI does not assign a user to the optimal server (which a
recent study~\cite{PADIS2010} has shown sometimes occurs) by selecting a
different CDI.

\subsubsection{CDI Federations} To avoid the cost of providing a global
footprint and perhaps to allow for a single negotiating unit with content
providers, federations of CDIs have been proposed. In this architecture,
smaller CDIs, perhaps operated by ISPs, join together to form a larger
federated CDI.  A CDI belonging to the federation can replicate content to a
partner CDI in a location where it has no footprint.  The CDI reduces its
transit costs because it only has to send the object once to satisfy the demand
for users in that location. Overall, cost may be reduced due to distance-based
pricing~\cite{HowManyTiers}. The IETF CDNi working group \cite{CDNi} works on
CDI federation.

### 2013-07-22 Compiled/chapter4/sections/Content-Delivery-Overview.tex
all chapters compile independently

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Content Delivery: An Overview}\label{sec:overview}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

While content may be seen as king not all content is equally popular among
users. Indeed, content popularity often follows ``Zipf's law''.  If the
popularity of elements as function of the rank is consistent with a power-law
distribution it is referred to as Zipf's-like (see~\cite{Zipf49,Mitzenmacher03}
and references therein).  The rank is determined by the number of occurrence of
an element, where a low rank index refers to a popular element.  Not
surprisingly Zipf's law does not only apply to the popularity of content but
also quite a number of different quantities in Internet traffic, including the
popularity of Web pages~\cite{breslau:99,SerpanosKarakostasWolf:00}, traffic
demands~\cite {Fang_Peterson:1999, feldmann:00,
ZhangBreslauPaxsonShenker2002,Wallerich:2005, BrownleeClaffy02}, as well as
interdomain Web traffic demands~\cite{Feldmann:Kammenhuber:2004}.  Thus, while
some content can be served by a single server most content, namely the popular
content, can only be served if it is highly replicated across multiple servers.
Thus, one of the main challenges in content delivery is {\bf{server
selection}}. Server selection means identifying a specific server from which
the request for content by a user is satisfied.

Content delivery and the network infrastructure interact mostly through content
source selection, often called server selection. Here, it does not matter
whether the source is a server pushing content through HTTP or from a peer in a
P2P network. In the case of HTTP, the domain name system (DNS) is the preferred
mechanism for performing server selection. In the case of P2P, peer selection
strategies drive where the content is obtained from and how, e.g., when the
content is cut into chunks.

To direct users to appropriate servers, CDIs rely extensively on the Domain
Name System (DNS).  We describe this and other server selection mechanisms in
detail later in this section.  The CDI chooses a server based on several
metrics. Criteria for server selection include the IP address of the end-user's
DNS resolver, the availability of the server, the proximity of the server to
the resolver, and the monetary cost of delivering the content. Note that the
server selection does not know the client IP address or network location, it
only knows the IP address of the DNS resolver the end-user contacted. A recent
study~\cite {DNS-IMC-2010} showed that sometimes the end-user is not close to
the resolver. To improve the mapping of end-users to servers, the client-IP
eDNS extension~\cite {DNS-extension-IP-client} has been recently proposed.

In P2P systems peers can choose among all other peers to download content from
but only if the have the desired content. Thus the problem of getting content
in a P2P system is actually two-fold: first the user needs to find the content
and once it knows of possible peers it can download the content from, it needs
to connect to some of them to get the desired content.  In P2P systems the
content lookup is realized in many different ways. Some P2P network, called
structured P2P, implement a distributed lookup system most often referred to as
distributed hash table (DHT). Other P2P systems, called unstructured P2P, like
Gnutella, flood search request into the network.  Some systems rely on a
partial centralized infrastructure to obtain content information. We discuss
the different approaches in P2P systems in more detail in
section~\ref{sec:P2P}.

Before we can discuss all the various options on how content delivery can be
improved in the current Internet we give a short overview how a typical Content
Distribution Network operates.

\subsection{Content Delivery Networks} \label{sec:content_delivery}

\begin{figure} \begin{center}
\includegraphics[width=0.7\linewidth]{figures/flows-example.eps} 
\end{center}
\caption{Example of CDI deployment and traffic flows (Web traffic demands).}
\label{fig:aka:cdn} \end{figure}

\begin{figure} 
\begin{center}
\includegraphics[width=0.6\linewidth]{figures/page.eps}
\vspace*{-.7cm}
\end{center}
\caption{Example Web page with some CDI content.}
\label{fig:aka:cdn_page}
\end{figure}

Recall content is king in the current Internet and content is typically first
placed on the Web site of the content producer, the original Web servers.
Content Delivery Infrastructures (CDIs) (see,
\eg~\cite{cdn:2002,akamai:2002,gadde01web,bent02whole,johnson00measured,bala:imw01,gribble:osdi02})
are designed to reduce the load on origin servers and at the same time improve
performance for the user.  Most CDIs have a large set of servers deployed
throughout the Internet and cache the content of the original publisher at
these servers.  Therefore another view of CDIs is that they provide reverse
proxy services for content providers, the publishers.  In order to take
advantage of their distributed infrastructure, requests for data are redirected
to the ``closest'' cache server. Intelligent redirection can reduce network
latency and load (and therefore network congestion) improving response time.
CDIs differ in their approach to redirecting traffic. Some (such as
Akamai~\cite{Akamai-Network}), use DNS to translate the hostname of a page
request into the IP address of an appropriate server. This translation may
consider the location of the client, the location of the server, the
connectivity of the client to the server, the load on the server, and other
performance and cost based criteria.

An example that shows how the CDI infrastructure is embedded in the Internet
architecture is shown in Figure~\ref{fig:aka:cdn}. Recall, the Internet is
divided into a collection of autonomous systems (ASes).  Each AS is managed by
an Internet Service Provider (ISP), who operates a backbone network that
provides connectivity to clients and to other ISPs.  Figure~\ref{fig:aka:cdn}
shows four ASes, numbered 1--4, whose backbones consist of three routers each,
two Web site publishers, \url{home.ex} and \url{adserver.ex}, and two sets of
clients. The publisher \url{home.ex} is connected to AS\,3 while the publisher
\url{adserver.ex} is connected to AS\,2. A set of clients is connected to
AS\,1, another to AS\,4.

The location of the CDI's servers differ from CDI to CDI and depends on
contractual agreements between the CDI and the individual ISPs.  In some
instances, the CDI servers are deployed within the data centers of the ISP and
therefore belong to the same AS, like AS\,1,\,2,\,4 in
Figure~\ref{fig:aka:cdn}.  Clients of the ISP (end-users) are typically served
by these servers in the same AS.  With other ISPs, the CDI may have a private
peering agreement that allows the CDI to serve requests from the ISPs clients
via a direct connection between the CDI and the AS.  The CDI may also co-locate
servers with the ISP's clients, \eg on university campuses.  With other ISPs
there may be no relationship with the CDI, and the traffic to the ISP's clients
is routed via another AS.

Let us consider the steps that are necessary to download the Web page shown in
Figure~\ref{fig:aka:cdn_page}. This page consists of one main page located at
\url{home.ex/index.htm} and four embedded objects. The publisher responsible
for \url{home.ex} has decided to use the services of a CDI, \url{cdi.ex}. One
object (\url{ex2.gif}) of the sample page is located on the same server as the
page itself (\url{index.htm}); another object (\url{ex3.gif}) is served by a
company providing dynamic advertisements, \url{adserver.ex}; and objects
\url{ex1.gif} and \url{ex4.jpg} are hosted by the CDI.

If a specific client from client set~A in Figure~\ref{fig:aka:cdn} accesses the
Web page, publisher \url{home.ex}  serves the bytes for the main page and one
embedded object, publisher \url{adserver.ex} serves the bytes for the object
located on its servers, and the ``nearest'' CDI server serves the two
CDI-located objects---in this case, they will be served from AS\,1.  In
contrast, if a specific client from client set~B accesses the page, the two CDI
objects are delivered from a different CDI server, namely the one in AS\,4.
Keep in mind that it is the objective of the CDI to direct the client to a CDI
server that is close to the client.

To complete the picture one question remains. How does the CDI choose the
``nearest'' server to deliver the content from? Today's CDI landscape relies
mainly on three techniques to assign end-users to servers.

\begin{enumerate*}

  \item IP-Anycast

  \item DNS based redirection

  \item HTTP redirection

\end{enumerate*}

While all techniques help the CDIs to assign end-users to their servers, all of
them have different drawbacks. In the following we will explain how the
different techniques work and what those drawbacks are:

\vspace{1em} \noindent \textbf{IP-Anycast.} IP Anycast is a routing technique
used to send IP packets to the topologically closest member of a group of
potential CDI servers. IP Anycast is usually realized by announcing the
destination address from multiple locations in a network or on the Internet.
Since the same IP address is available at multiple locations, the routing
process selects the shortest route for the destination according to its
configuration. Simply speaking, each router in a network selects one of the
locations the Anycasted IP is announced from based on the used routing metrics
(\eg path length or routing weights) and configures a route towards it. Note
that, if a network learns of an Anycasted IP address from different sources, it
does not necessarily direct all its traffic to one of its locations. Its
routing can decide to send packets from region A in the network to location A'
while region B gets a route to location B'. This means that the entire server
selection of a CDI becomes trivial as it is now a part of the routing process.
This means that the CDI loses control of how the users are mapped to the server
because the network calculates the routing based on its own metrics. Another
issue is that the routing in a network is optimized based on the ISPs criteria
which might not be the same as the CDIs or even contrary. Thus the ``nearest''
server might not be the best one the CDI could offer.

\vspace{1em} \noindent \textbf{DNS-based redirection.} Today most CDIs rely on
the Domain Name System (DNS) to direct users to appropriate servers.  When
requesting content, the end-user typically asks a DNS resolver, \eg the
resolver of its ISP, for the resolution of a domain name. The resolver then
asks the authoritative server for the domain.  This can be the CDI's
authoritative server, or the the content provider's authoritative server, which
then delegates to the CDI's authoritative server.  At this point the CDI
selects the server for this request based on where the request comes from. But
the request does not come directly from the end-user but from its DNS resolver!
Thus the CDI can only select a server based on the IP address of the end-user's
DNS resolver. To improve the mapping of end-users to servers, the client-IP
eDNS extension~\cite{DNS-extension-IP-client} has been recently proposed.
Criteria for server selection include the availability of the server, the
proximity of the server to the resolver, and the monetary cost of delivering
the content. For proximity estimations the CDIs rely heavily on network
measurements~\cite{Akamai-Network} and geolocation information~\cite{MaxMind}
to figure out which of their servers is close by and has the best network path
performance.  A recent study~\cite{DNS-IMC-2010} showed that sometimes the end
user is not close to the resolver and another study points out that geolocation
databases can not be relied upon~\cite{PUKDG-IGDU-11}. Thus the proximity
estimations for the ``nearest'' CDI server highly depend on the quality and
precision of network measurements and a proper DNS deployment of the ISPs. For
an excellent survey on DNS-based Server Selections in CDNs, we refer the reader
to~\cite{dns-redirection}.

\vspace{1em} \noindent \textbf{HTTP redirection.} The Hypertext Transfer
Protocol (HTTP) is today's de-facto standard to transport content in the
Internet (see section~\ref{sec:traffic-diversity}). The protocol incorporates a
mechanism to redirect users at the application level at least since it was
standardized as version 1.0 in 1996~\cite{RFC1945}. By sending an appropriate
HTTP status code (HTTP status codes 3XX) the web server can tell the connected
user that a requested object is available from another URL, which can also
point to another server. This allows a CDI to redirect an end-user to another
server. Reasons for this might include limited server capacities, poor transfer
performance or when another server is closer to the end-user, \eg a client from
the US connecting to a server in Europe although the CDI has servers in the US.
The HTTP redirection mechanism has some important benefits over the DNS based
approach. First, the CDI directly communicates with the end-user and thus knows
the exact destination it sends the traffic to (opposed to the assumption that
the DNS resolver is ``close''). Yet it still has to estimate the proximity of
the end-user using the same methodologies as described in the DNS based case.
Second, the CDI already knows which object the end-user requests and can use
this information for its decision. It allows a CDI to direct a user towards a
server where the content object is already available to improve its cache hit
rate. Other important informations includes the size and type of the object.
This allows the CDI to optimize the server selection based on the requirements
to transfer the object \eg for delay sensitive ones like streaming video or
more throughput oriented ones like huge software patches. Yet this improvement
comes at a price as the user has to establish a new connection to another
server. This includes another DNS lookup to get the servers IP address
as well as the whole TCP setup including performance critical phases like slow
start. This can repeat itself multiple times before an appropriate server is
found, which delays the object delivery even further.

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Peer-to-Peer Networks}\label{sec:P2P}
%%%%%%%%%%%%%%%%%%%%%%%

Peer-to-peer (P2P) is a distributed system architecture in which all
participants, the so called peers, are equally privileged users of the system.
A P2P system forms an overlay network on top of existing communication networks
(\eg the Internet). All participating peers of the P2P system are the nodes of
the overlay network graph, while the connections between them are the edges. It
is possible to extend this definition of edges in the overlay network graph to
all known peers, in contrast to all connected peers.  Based on how peers
connect to each other and thus build the overlay network, we can classify P2P
systems into two basic categories:

\textbf{Unstructured}: The P2P system does not impose any structure on the
overlay network. The peers connect to each other in an arbitrary fashion. Most
often peers are chosen randomly. Content lookups are flooded to the network
(\eg Gnutella), resulting in limited scalability, or not offered at all (\eg
plain BitTorrent).

\textbf{Structured}: Peers organize themselves following certain criteria and
algorithms. The resulting overlay network graphs have specific topologies and
properties that usually offer better scalability and faster lookups than
unstructured P2P systems (\eg Kademlia, BitTorrent DHT).

The overlay network is mainly used for indexing content and peer discovery
while the actual content is transferred directly between peers. Thus the
connection between the individual peers has significant impact on both the
direct content transfers as well as the performance of the resulting overlay
network. This has been shown in previous studies and multiple solutions have
been
proposed~\cite{p4p,taming,DraftingAkamai:SIGCOMM2006,afs-cispp2pcip-ccr07,ietf-alto-protocol,CDNi}
which are described in detail in section~\ref{sec:Opportunities}.

Applications of P2P systems in content delivery range from time insensitive
applications like file sharing, software delivery or patch distribution to very
time sensitive ones like streaming TV or on demand video delivery.

\noindent\textbf{Peer-to-Peer systems} To construct an overlay topology,
unstructured P2P networks usually employ an arbitrary neighbor selection
procedure~\cite{P2P-LNCS-05}. This can result in a situation where a node in
Frankfurt downloads a large content file from a node in Sydney, while the same
information may be available at a node in Berlin. While structured P2P systems
follow certain rules and algorithms, the information available to them either
has to be inferred by measurements~\cite{rhks-taocss-02} or rely on publicly
available information such as routing information~\cite {Routeviews}. Both
options are much less precise and up-to-date compared to the information
information an ISP has readily at hand.  It has been shown that P2P traffic
often crosses network boundaries multiple
times~\cite{abfw-mendg-04,krp-sifpacd-05}. This is not necessarily optimal as
most network bottlenecks in the Internet are assumed to be either in the access
network or on the links between ISPs, but rarely in the backbones of the
ISPs~\cite{ass-eewib-03}. Besides, studies have shown that the desired content
is often available ``in the proximity'' of interested
users~\cite{krp-sifpacd-05,rsr-oletgo-06}. This is due to content language and
geographical regions of interest. P2P networks benefit from increasing their
traffic locality, as shown by Bindal~et.~al~\cite{b-itlbbns-06} for the case of
BitTorrent.

P2P systems usually implement their own routing~\cite{abkm-ron-01} in the
overlay topology. Routing on such an overlay topology is no longer done on a
per-prefix basis, but rather on a query or key basis. In unstructured P2P
networks, queries are disseminated, e.g., via flooding~\cite{Gnutellav0.6} or
random walks, while structured P2P networks often use DHT-based routing systems
to locate data~\cite{P2P-LNCS-05}. Answers can either be sent directly using
the underlay routing~\cite{P2P-LNCS-05} or through the overlay network by
retracing the query path~ \cite{Gnutellav0.6}. By routing through the overlay
of P2P nodes, P2P systems hope to use paths with better performance than those
available via the Internet native routing~\cite{abkm-ron-01,sch-eeips-99}.
However, the benefits of redirecting traffic on an alternative path, e.g., one
with larger available bandwidth or lower delay, are not necessarily obvious.
While the performance of the P2P system may temporarily improve, the available
bandwidth of the newly chosen path may deteriorate due to the traffic added to
this path. The ISP has then to redirect some traffic so that other applications
using this path can receive enough bandwidth. In other words, P2P systems
reinvent and re-implement a routing system whose dynamics should be able to
explicitly interact with the dynamics of native Internet
routing~\cite{ktci-cithon-04,sa-oidronl-06}.  While a routing underlay as
proposed by Nakao et al.~\cite{npb-ruon-03} can reduce the work duplication, it
cannot by itself overcome the problems created by the interaction. Consider a
situation where a P2P system imposes a lot of traffic load on an ISP network.
This may cause the ISP to change some routing metrics and therefore some paths
(at the native routing layer) in order to improve its network utilization. This
can however cause a change of routes at the application layer by the P2P
system, which may again trigger a response by the ISP, and so on.

%%%%%%%%%%%%%%%%%%%%%%%
\noindent\textbf{P2P today.}\label{sec:P2P-today}
%%%%%%%%%%%%%%%%%%%%%%%
The P2P paradigm has been very successful in delivering content to end-users.
BitTorrent~\cite {BTCohen} is the prime example, used mainly for file sharing.
Other examples include more time sensitive applications such as video
streaming~\cite{Conviva2011,Conviva2012,Video-IMC2012}. Despite the varying
(and perhaps declining) share of P2P traffic in different regions of the
world~\cite{OnDominantCharacteristics2009}, P2P traffic still constitutes a
significant fraction of the total Internet traffic. P2P systems have been shown
to scale application capacity well during flash crowds~\cite{BTCapacity}.
However, the strength of P2P systems, \ie anybody can share anything over this
technology, also turns out to be a weakness when it comes to content
availability. In fact, mostly popular content is available on P2P networks,
while older content disappears as users' interest in it declines. In the
example of BitTorrent, this leads to torrents missing pieces,
in which case a download can never be completed. In case of video streaming,
the video might simply no longer be available or the number of available peers
is too low to sustain the required video bit-rate, resulting in gaps or
stuttering of the video stream.

### 2013-07-22 Compiled/chapter4/sections/Incentives.tex
all chapters compile independently

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Incentives for Collaboration}\label{sec:Incentives}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

ISPs are in a unique position to help CDIs and P2P systems to improve content
delivery.  Specifically, ISPs have the knowledge about the state of the
underlying network topology and the status of individual links that CDIs are
lacking. This information not only helps CDIs in their user-to-server mapping,
but also reduces the need for CDIs to perform large-scale active measurements
and topology discovery \cite{sureroute}.  It also enables CDIs to better
amortize their existing infrastructure, offer better quality of experience to
their users, and plan their infrastructure expansion more efficiently. On the
other side, ISPs are not just selflessly giving up their network information.
Offering their intimate knowledge of the network to CDIs puts ISPs in the
position that they can also actively guide the CDIs. This allows ISPs to gain
unprecedented influence on CDI traffic.

The opportunity for ISPs to coordinate with CDIs is technically possible
thanks to the decoupling of server selection from content delivery. In general,
any end-user requesting content from a CDI first does a mapping request,
usually through the Domain Name System (DNS). During this request, the CDI
needs to locate the network position of the end-user and assign a server
capable of delivering the content, preferably close to the end-user.  ISPs have
this information ready at their fingertips, but are currently not able to
communicate their knowledge to CDIs. Furthermore, ISPs solve the challenge of
predicting CDI traffic, which is very difficult due to the lack of information
on the CDI mapping strategy regarding the end-users to servers assignment. In
order to reap the benefits of the other's knowledge, both parties require
incentives to work together.

\subsection{Incentives for CDIs}

The CDIs' market requires them to enable new applications while reducing
their operational costs and improve end-user experience~\cite{Akamai-Network}.
By cooperating with an ISP, a CDI improves the mapping of end-users to servers,
improves in the end-user experience, has accurate and up-to-date knowledge of
the networks and thus gains a competitive advantage. This is particularly
important for CDIs in light of the commoditization of the content delivery
market and the selection offered to end-users, for example through
meta-CDNs~\cite{Conviva2011}. The improved mapping also yields better
infrastructure amortization and, thanks to cooperation with ISPs, CDIs will no
longer have to perform and analyze voluminous measurements in order to infer
the network conditions or end-user locations.

To stimulate cooperation, ISPs can operate and provide their network
knowledge as a free service to CDIs or even offer discounts on peering or
hosting prices, \eg for early adopters and CDIs willing to cooperate. The loss
of peering or hosting revenue is amortized with the benefits of a lower network
utilization, reduced investments in network capacity expansion and by taking
back some control over traffic within the network. Ma et al.~\cite
{CooperativeSettlement:ToN} have developed a methodology to estimate the prices
in such a cooperative scheme by utilizing the Shapley settlement mechanism.
Cooperation can also act as an enabler for CDIs and ISPs to jointly launch new
applications in a cost-effective way, for example traffic-intensive
applications such as the delivery of high definition video on-demand, or
real-time applications such as online games.

\subsection{Incentives for ISPs}\label{sec:ISP-Incentive}

ISPs are interested in reducing their operational and infrastructure upgrade
costs, offering broadband services at competitive prices, and delivering the
best end-user experience possible. Due to network congestion during peak hour,
ISPs in North America have recently revisited the flat pricing model and some
have announced data caps to broadband services. A better management of traffic
in their networks allows them to offer higher data caps or even alleviate the
need to introduce them. From an ISP perspective, cooperation with a CDI offers
the possibility to do global traffic and peering management through an improved
awareness of traffic across the whole network. For example, peering agreements
with CDIs can offer cooperation in exchange for reduced costs to CDIs. This can
be an incentive for CDIs to peer with ISPs, and an additional revenue for an
ISP, as such reduced prices can attract additional peering customers.
Furthermore, collaboration with CDIs has the potential to reduce the
significant overhead due to the handling of customer complaints that often do
not stem from the operation of the ISP but the operation of CDIs~\cite
{ControllingDataCloud}. Through this, ISPs can identify and mitigate congestion
in content delivery, and react to short disturbances caused by an increased
demand of content from CDIs by communicating these incidents back directly to
the source.

\subsection{Effect on End-users}\label{sec:Users-Incentive}

Collaboration between ISPs and CDIs in content delivery empowers end-users to
obtain the best possible quality of experience. As such, this creates an
incentive for end-users to support the adoption of collaboration by both ISPs
and CDIs. For example, an ISP can offer more attractive products, \ie higher
bandwidth or lower prices, since it is able to better manage the traffic inside
its network.  Also, thanks to better traffic engineering, ISPs can increase
data caps on their broadband offers, making the ISP more attractive to
end-users. Moreover, CDIs that are willing to jointly deliver content can offer
better quality of experience to end-users. This can even be done through
premium services offered by the CDI to its customers. For example, CDIs
delivering streaming services can offer higher quality videos to end-users
thanks to better server assignment and network engineering.

### 2013-07-22 Compiled/chapter4/sections/Internet-Network-Infrastructures.tex
all chapters compile independently

%%%%%%%%%%%%%%%%%%%%%%%
\section{Internet Network Infrastructure}\label{sec:ISP}
%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure*}
  \begin{center}
  \includegraphics[width=\linewidth]{figures/Internet-Layout.eps}
  \end{center}
  \caption{Layout of the Internet Structure }
  \label{fig:InternetLayout}
\end{figure*}

The Internet Network Infrastructure is provided by a set of Internet Service
Providers (ISPs).  An ISP is, in general terms, an organization that provides
access to the Internet for its customers.  The Internet is structured by the
interconnection of multiple individual networks run by ISPs.  However, control
of an individual network remains solely with the ISP operating it. Figure~\ref
{fig:InternetLayout} shows how the Internet is structured today~\cite{arbor}.
Here, the ISPs run their own networks. This forces a clear distinction between
the individual network that an ISP runs and the global Internet as a network of
networks.  Also, from this, it can be deduced that nobody has control over the
Internet, but instead each ISP has only control over its own network and the
direct connections to other networks.

To be able to interconnect with other networks, an ISP needs to operate an
autonomous system (AS). An AS is an administrating entity, generally under the
control of one administrative domain. On the technical side, each AS is usually
managed by an Interior Gateway Protocol (IGP), \eg OSPF~\cite{RFC2328} or
ISIS~\cite{RFC1142} while the Border Gateway Protocol (BGP~\cite{RFC4271}) is
the de-facto standard for interconnecting different ASes. For more information
and additional details about the Internet topology, we'd like to refer the
reader to Chapter 7 of this book~\cite{ebook-chapter7-Redux}.

\subsection{Traffic Engineering in an AS}

\begin{figure} \begin{center}
\includegraphics[width=1\linewidth]{figures/Classic-TE.eps} 
\end{center} 
\caption{IGP based Traffic Management Example} 
\label{fig:SimplifiedTE} 
\end{figure}

The greatest challenge for an ISP is to keep its infrastructure operating
efficiently. This is especially hard, since the ISP itself controls neither the
behavior, nor the source nor destination of the majority of the traffic it
carries. The destination of the traffic is determined by the end-users the ISP
sells services to, while the source is usually operated by a Content Delivery
Infrastructure (CDI). The behavior is dictated through end-users requesting
content, and by the operational choices of the CDI. ISPs today tackle the
problem of network operation efficiency by performing Traffic Engineering (TE).
In its broadest sense, today's TE encompasses the application of technology and
scientific principles to the measurement, characterization, modeling, and
control of Internet traffic~\cite{Awduche_OverviewTE:2002}. Today, traffic
engineering reduces to controlling and optimizing the routing function and to
steering traffic on an Origin-Destination (OD) flow basis through the network
in the most effective way.

Traffic Engineering encompasses multiple steps in order to be performed
successfully. First, an ISP needs to record its traffic volume in terms of
Origin-Destination flows. This means keeping traffic statistics of how much
traffic flows from one router in the network to another. Once the OD flows have
been successfully recorded, TE uses this information to simulate the network
behavior with different IGP configurations. The goal of these simulations is to
find an IGP configuration that spreads the network load as evenly as possible.

Figure~\ref{fig:SimplifiedTE} shows an example of how an IGP configuration can
be used to engineer traffic. The labeled circles represent routers, while the
numbers in the squares represent the IGP-weight for the link. For ease of
presentation, the weights for each link are set to the same value for both
directions. An OD flow, which starts at one router and finishes at another,
takes the path through the network that yields the smallest sum over all
weights along the path. For example, in the starting configuration of the
network (Figure~\ref {fig:SimplifiedTE} (left)) the flow $IG$ does not take the
direct path $I \rightarrow H \rightarrow G$ two, since according to the IGP
weights, a more effective path exists. In fact, the path $I \rightarrow H
\rightarrow E \rightarrow D \rightarrow \ G$ has an accumulated weight of 4
instead of 5 (green path). All traffic at router I destined for router G takes
this path. Similarly, all traffic that originates from B and goes to G follows
the path $B \rightarrow E \rightarrow D \rightarrow G$ (blue path). Also, both
paths share links, leading to a possible overload situation. In order to solve
this problem, we choose to modify the link weight between the routers D and E.
By increasing the weight from 1 to 5 (marked red in the right network), the
blue as well as the green paths are shifted to the direct path. The change is
shown in Figure~\ref{fig:SimplifiedTE} (right).

This simple diagram allows for illustrating multiple caveats that IGP based
traffic engineering introduces. First, IGP-based traffic engineering affects
traffic on an OD-flow basis only. This means that the path from one router to
another can be changed, but the traffic on the OD flow cannot be split onto
multiple paths. Secondly, the change of one weight can affect multiple OD-flows
at the same time. Thus, the weights have to be changed very carefully. In the
worst case, it might not be possible to fully separate some OD-flows due to the
network layout.

One caveat is not immediately obvious but needs to be taken into account when
performing traffic engineering. While the link weights are usually known to all
routers, they are propagated by messages that routers exchange. This
propagation takes time, which can lead to short-term inconsistencies in the
view of a network. We again use Figure~ \ref{fig:SimplifiedTE} for illustrating
this. When the link weight is changed as described in the example explained
before, routers D and E update their routing. This has an immediate effect on
the traffic from B to G. With the update, the shortest path from router E to G
is now $E \rightarrow H \rightarrow G$. In accordance, E configures its routing
to send all traffic for G through H. However, H has not converged at this point
and still uses the old path ($H \rightarrow E \rightarrow D \rightarrow G$).
Thus, H still sends all traffic for G towards E. As long as H uses the outdated
IGP weight information, all traffic for G that reaches either E or H is sent
back and forth between the two routers. This forwarding, on the one hand,
likely overloads the link. On the other hand, most traffic that is affected by
this will be dropped due to its time-to-live (TTL) running out.

The work of Francois \etal~\cite{transient-IGP} shows that it is possible to
gradually change IGP weights by sequentially ordering changes. Accordingly,
routing loops like those in the example are avoided. However, these changes
still require time during which the network can be in a transient state with
overloaded links. Besides the challenges induced by optimizing the IGP, this
approach also assumes that traffic is predictable and stable over time. By
running simulations based on past traffic aggregates to engineer the routing
for the future, it is implicitly assumed that traffic patterns remain similar
over a longer period of time.

With the emergence of CDIs, however, traffic has become volatile in terms of
its origin. In fact, CDIs can shift massive amounts of traffic in a matter of
seconds from one server cluster to another. While this behavior is needed and
propagated by CDIs to cope with volatile demand surges, it is in stark contrast
to the ISP's traffic engineering, which assumes traffic behavior to be stable
for days, weeks or sometimes months.

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Domain Name System Basics}\label{sec:DNS}
%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[tbp] \begin{center}
\includegraphics[width=0.8\linewidth]{figures/DNSHierarchy.eps} 
\end{center}
\caption{Example DNS hierarchy} 
\label{fig:DNS-Hierarchy} 
\end{figure}

The Domain Name System (DNS) plays a major role in today's Internet
architecture and is an essential component of any Web based content delivery
architecture.  DNS relies on a distributed database with a hierarchical
structure. The root zone of the DNS system is centrally administered and serves
its {\em zone} information via a collection of {\em root servers}.  The root
servers delegate responsibility for specific parts (zones) of the hierarchy to
other {\em name servers}, which may in turn delegate the responsibility to
other name servers.  At the end, each site is responsible for its own {\em
domain} and maintains its own database containing its information and operates
an {\em authoritative} name server.

The whole DNS database is usually queried by end-hosts using a local name
server called {\em caching resolver}. If this name server receives a query for
a domain that it does not know about, it fetches this information from another
name server. If the server does not know how to contact the authoritative
server for a zone, it will query a root server\footnote{The first query can go
to some authoritative server below the root if there exists cached
information.}. The root server will \emph{refer} the resolver to another server
that is authoritative for the domain that is immediately below the root and of
which the zone is a part. The resolver will then query this server, and so
forth, stepping down the tree from the root to the desired zone.

To illustrate this process, Figure~\ref{fig:DNS-Hierarchy} show a sample DNS
hierarchy. In this case, the root of the DNS name space, denoted with a '.', is
hosted on two \emph{DNS root servers}. Both servers are under one
administrative control, and both can refer a request to any of the top level
domain servers. Here, three domains exist, \ie \textit{.com}, \textit {.net}
and \textit{.info}. Again, these name servers refer to the second level
domains. Since the domain name are concatenated together as the hierarchy is
traversed, the domains that are now possible are \textit {d1.com.}, \textit
{d2.com.}, \textit{d1.net.} and \textit{d3.info.}. At this point, the second
level domains d1.net. and d3.info have reached their authoritative resolver.
For example, a query to the name server of \textit{.d3} for
\textit{www.d3.info} is answered authoritatively from there. Note that the name
servers for the second level domains are operated by independent entities that
know nothing of each other. Thus, the database is distributed, while each party
is responsible for its own zone.  Finally, the name server of \textit{.d1.com.}
has a dual role. While it is referring the subdomains \textit{.sd1.d1.com.} and
\textit{.sd2.d1.com.} to other name servers, it also answers queries for other
names in its name space authoritatively.  This means that a query for
\textit{www.d1.com.} is directly answered, while a query for
\textit{www.sd1.d1.com} is referred to the name server responsible for
\textit{.sd1.d1.com}.

For efficiency reasons DNS relies heavily on caching
\cite{DNS-caching,DNS-IMC-2010}. All information that a name server delivers to
a resolver is cached for a duration specified in the time-to-live (TTL) field
of the {\em resource records} (RR). Caching today is usually also performed on
end-hosts by the operating system's {\em stub resolver}, as well as
applications, \eg web browsers.

\noindent\textbf{DNS Today.}\label{section:dns-today} When DNS was introduced in
1983, its sole purpose was to resolve host names into IP addresses in a more
scalable fashion than the until then used \texttt{hosts} file. Since then a
number of features and new uses have found their way into the now omnipresent
DNS. In addition to the increasing complexity within the DNS protocol itself
\cite {DNSComplexity2007}, new and oftentimes unforeseen (ab)uses have been
established. Paul Vixie gives an overview in \cite{WhatDNSIsNOT2009}. The most
important points of critique are as follows:

\begin{description} \item[CDI load balancing:] Content delivery infrastructures
set short TTLs on their DNS answers to allow for short reaction times to
shifting loads. Short TTLs impede on cacheability and therefore increase the
load on the whole DNS system. In addition, CDIs tailor their reply for the IP
address of the requesting resolver using the assumption that the DNS resolver
is close to the client originating the request. It has been shown in the past
that this assumption is quite often wrong~\cite
{Precise:Mao2002,dns-redirection,DNS-IMC-2010,DNS-extension-IP-client}.

\item[NXDOMAIN catcher:] Some ISPs and third party DNS providers mangle a
negative reply with the NXDOMAIN status code into a positive one with the IP
address of a search website under the control of the ISP. By hosting
advertisements along the search results it is easily possible to increase the
profit margin. While this may work to some degree for web browsing,
applications relying on proper delivery of NXDOMAIN records, \eg email, are
inevitably hampered.  \end{description}

A third-party ecosystem around DNS has evolved over the last couple of years.
Players like OpenDNS, AdvantageDNS, UltraDNS, and most recently Google offer
open resolvers to anyone with different feature sets. OpenDNS Basic does
NXDOMAIN catching but offers phishing and botnet protection for free.
Furthermore, OpenDNS increases the service level for payment between 5 dollars
a month up to several thousand dollars per year for business customers. When
Google Public DNS entered the market, their highest-valued goals were to
``speed up your browsing experience'' and to ``improve your security''. To
achieve both targets Google advertises an impressive list of optimizations and
fine tuning~\cite{googledns}, \eg prefetching, load balancing with shared
cache, validity checking, and nonce prepending. Google Public DNS also refrains
from utilizing NXDOMAIN to make profit. From an implementation perspective,
most if not all of the third-party resolvers host their DNS servers on multiple
sites around the globe and use anycast to guide DNS clients to the nearest
resolver.

In this open market space a user annoyed by his ISP's DNS can easily choose for
cost-free third-party service.  Tools such as namebench~\cite{namebench} might
help him in choosing a well-performing one. The irony however is that a user,
by choosing a different DNS than the one assigned by his ISP, will most likely
undermine the traffic matrix optimizations performed by CDIs and ISPs, and can
potentially even lower his quality of experience due to longer download
times~\cite{DNS-IMC-2010}.

### 2013-07-22 Compiled/chapter4/sections/Introduction.tex
all chapters compile independently

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}\label{sec:introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Recent traffic studies~\cite{TrafficTypesGrowth:2011,arbor,PADIS2010} show that
a large fraction of Internet traffic is due to content delivery and is
originated by a small number of Content Delivery Infrastructures (CDIs). Major
CDIs include highly popular rich-media sites like YouTube and Netflix,
One-Click Hosters (OCHs), \eg RapidShare~\cite{OCH-performance},
Content Delivery Networks (CDNs) such as Akamai and Limelight, and
hyper-giants, \eg Google, Yahoo!, and Microsoft.  Gerber and Doverspike~\cite
{TrafficTypesGrowth:2011} report that a few CDIs account for more than half of
the traffic of a US-based Tier-1 carrier.  Poese \etal~\cite{PADIS2010} report
a similar observation from the traffic of a European Tier-1 carrier. Labovitz
\etal~\cite{arbor} infer that more than 10\% of the total Internet inter-domain
traffic originates from Google, and Akamai claims to deliver more than 20\% of
the total Internet Web traffic~\cite{Akamai-Network}. Netflix alone, a company
that offers a high definition video video-on-demand streaming service, is
responsible for a significant fraction of the traffic in North America ISPs
during peak hours~\cite{sandvine,Netflix-CDN-presentation}.

To cope with the increasing demand for content, CDIs have deployed massively
distributed server infrastructures to replicate content and make it accessible
from different locations on the Internet \cite{CDNsec2009}. These
infrastructures have multiple choices as to how and where to place their
servers. As described in~\cite {ImprovingPerformanceInternet2009}, the main
approaches are (1) centralized hosting, (2) data center-based CDIs, (3)
edge-cache-based CDIs, and (4) peer-to-peer (P2P) networks. Approaches~2 and~3
scale content delivery by distributing the content onto dedicated
infrastructures.  These infrastructures can be composed of a few large data
centers, a large number of edge caches, or any combination thereof.

To complicate matters further, some of these infrastructures are entangled with
the very infrastructures that provide network connectivity to end-users. For
example, one of the largest players in content delivery, Akamai, operates more
than 120,000 servers in more than 2,000 locations across nearly 1,150 ISP
networks~\cite {Akamai-Network,Akamai-website}. Google is reported to operate
tens of data centers and front-end server clusters
worldwide~\cite{MovingBeyondE2E2009,Tariq:What-if,Google-Datacenters}.
Microsoft has deployed its content delivery infrastructure in 24 locations
around the world~\cite{azure}. Amazon maintains at least 5 large data centers
and caches in at least 21 locations around the world~\cite{amazon}. Limelight
operates thousands of servers in more than 22 delivery centers and connects
directly to 600 networks worldwide~\cite{LLNetworks}. Last but not least, P2P
networks rely on a huge number of end users to store, replicate, and distribute
content.

Despite the significant entanglement between the infrastructures that deliver
content and the network connectivity fabric, our knowledge of their
interactions is largely through the literature on network interconnections, \eg
see the recent book by W. Norton~\cite{peering-playbook}. Given the nature of
network interconnections, previous work has studied the interactions from an
economic perspective~\cite
{internet-economics,interactions-economics,content-oriented-economics}. The
limited knowledge available about the settlements between networks have led
researchers to try to reason about why peering choices are made~\cite
{ToPeerOrNot:Infocom2006} and what drives the evolution of the
Internet~\cite{evolution}.

Most of the literature has considered the interactions between content and the
network indirectly, \eg through peerings and traffic measurements, despite
recent changes in Internet traffic~\cite{TrafficTypesGrowth:2011,arbor} that
have shown the importance of content and applications.  The observed changes in
traffic, either through direct traffic measurements~\cite
{TrafficDemand:ToN2001,Feldmann:Kammenhuber:2004,uhlig-public,arbor,IXPSIGCOMM2012},
or through
inference~\cite{medina_tm:2002,zhang_tm:2003,zhang_link:2003,gunnar_estimation:2004,dina_distributed:2004,soule_tm:2005}
have repeatedly shown how volatile traffic can be. With the rise of
user-generated content and large shifts in content popularity, traffic
volatility has become especially relevant.

Handling changes in traffic has traditionally been done through traffic
engineering (TE). Initially, traffic engineering was used by large network
operators to optimize the utilization of their networks
\cite{Awduche_OverviewTE:2002}. The vast majority of the traffic engineering
literature has therefore focused on traffic engineering inside a single network
\cite{netscope,FT00,Xiao_Traffic:2000,Aukia_RATES:2000,MPLS:Infocom2000,FT01}.
In reality, most of the traffic in the Internet is exchanged between different
networks \cite{arbor}, and especially directly between data centers and
residential ISPs \cite{IXPSIGCOMM2012}. Organizations that originate a lot of
content, \eg Google, connect directly to a large number of other networks
\cite{arbor}, and need to optimize how content leaves their networks.
Organizations that provide Internet access to broadband or mobile users
typically wish to optimize how traffic enters their networks, as most users
still download more content than they upload. In between, transit ISPs try to
balance the load of the traffic exchanged between the networks they connect.

Traditional traffic engineering aims at reducing the likelihood that
bottlenecks arise inside a given network due to mismatches between network
provisioning and expected demand. Changes in network provisioning are slow,
taking place over time scales of weeks or months. Popular content, on the other
hand, generates bursts in demand over much smaller time scales, \eg hours or
minutes. Today's Internet requires much more reactive network control
techniques than those we have today, and these techniques must take content
delivery into consideration. A few steps have been made in this direction.
Indeed, collaborative approaches
\cite{TECDN,CooperativeSettlement:ToN,Cate-CCR} have been proposed to help deal
with the traffic generated by content delivery infrastructures. Even in the
case of P2P, portals have been proposed to allow P2P applications and users to
communicate with ISPs to receive updated views of their networks~\cite{p4p}. In
broad terms, all information CDIs are missing today for optimizing their
operations is available to ISPs. Combined with the already proposed schemes for
collaboration, it is surprising how little real collaboration is performed in
today's Internet between these parties.

In this chapter, we analyze the operation of CDIs as well as network operators.
The analysis demonstrates the potential for fruitful collaboration.  We argue
that for collaboration to become more common, it is important for every party
in the content delivery landscape, \ie the content delivery infrastructures,
the network operators, and the end users, to benefit.  Finally, we present, in
depth, two systems that have incentives for every party and that can readily be
used today.

### 2013-07-22 Compiled/chapter4/sections/Motivation.tex
all chapters compile independently

\section{Motivation}

The Internet is a hugely successful man-made artifact that has changed society
fundamentally.  Imagine the effect a prolonged outage of the Internet would
have: (1) Youngsters wouldn't know how to interact with their peers and how to
spend their leisure time as they increasingly rely on social networks, online
games, YouTube, and other online entertainment offerings.  (2) Manufacturing
would hit a roadblock as the communication paths within and between companies
increasingly rely on the Internet. (3) Control of critical infrastructures
would be hampered as it increasingly relies on the Internet for gathering input
data and propagating control information.

In becoming a hugely successful infrastructure, the usage of the Internet and
thus its structure has also undergone continuous changes. Usage has changed
from dominance by email and FTP in the early days, to the World Wide Web (WWW)
from 1995 to 2000, to peer-to-peer applications (P2P) from 2000 to 2007, back
to the WWW since 2007. These changes are driven in part by the Internet users'
interests as well as how content, including user generated content, is made
available.

When considering the current application mix and traffic streams in the
Internet, the latest buzz is that ``Content is King'' just as Bill
Gates~\cite{Gates-Content-King} predicted in his essay from 1996.  Hereby, the
term content has to be seen very broadly and encompasses everything from
commercially prepared content, \eg broadcast and interactive TV, news, and
software, to user-generated content, \eg videos uploaded to YouTube, and photos
uploaded to Flickr, to interactive activities, \eg online games. Or to quote
Bronfman~\cite{Bronfman-remarks}, the head of a major music producer and
distributor: ``What would the Internet be without `content'? It would be a
valueless collection of silent machines with gray screens. It would be the
electronic equivalent of a marine desert---lovely elements, nice colors, no
life.  It would be nothing.''

The idea of content delivery being the fundamental operation around which to
design future Internet architecture for comes as no surprise. In fact, the idea
of Content-Centric Networking (CCN)~\cite{CCN} is guided by this principle.
Change, however, takes time, and when hundreds of million of devices are
involved, change can only be made slowly. Before such a novel and radically
different architecture such as CCN is available or potentially deployable, the
Internet in its current state must cope with the challenge of delivering
ever-increasing amounts of content to Internet users.

Accordingly, it appears that solely providing connectivity to end users is no
longer sufficient for Internet Service Providers (ISPs).  Yet, connectivity is
a crucial ingredient and some authors, \eg Andrew
Odlyzko~\cite{odlyzko2001content} have opined that enabling communication is
the main task of the Internet network infrastructure. In his paper ``Content is
not king'' he claims that ``Content will have a place on the Internet, possibly
a substantial place. However, its place will likely be subordinate to that of
business and personal communication''.

At this point it is crucial to realize that the producers of content are
usually not the operators of today's Internet infrastructure. Nonetheless, both
content producers and network operators depend on each other. In fact, neither
the Internet infrastructure operators nor the content producers can be
successful without the other.  After all, the content producers want to ensure
that their content gets to Internet users with reasonable performance for which
they need to rely on the network infrastructure.  On the other hand, the
network infrastructure providers have to transport the content and manage the
infrastructure to satisfy the demand for content from their subscribers. It is
this symbiosis between the two parties that motivates our work collaboration
between content producers and network operators in delivering content.

\bigskip

\noindent\textbf{Outline:} We start this chapter with a short introduction in
Section~\ref{sec:introduction}.  Then, in Section~\ref{sec:ISP}, we set the
stage by providing an overview of today's Internet network infrastructure,
discussing how Internet Service Providers (ISPs) perform traffic engineering,
and reviewing the Domain Name System (DNS), an essential component of any
Web-based content-delivery architecture.   Next, we review current trends in
Internet traffic and the application mix as well as traffic dynamics in
Sections~\ref{sec:traffic} and~\ref{sec:traffic-diversity}.

We finish the overview with a brief summary on the background of content
delivery in Section~\ref{sec:overview}. Here, we assume that the reader is
familiar with the basic architecture of the Web. There are excellent text books
on this topic, \eg~\cite{web-book}.  Given that there are several approaches to
content delivery, we provide a general high level description of how different
Content Delivery Infrastructures work.  Since there are also many peer-to-peer
based content delivery systems we provide a short review of the basic P2P
architectures as well. For additional background on P2P we refer the reader to,
\eg~\cite{p2pbook1,p2pbook2}.

An overview of the current content delivery spectrum is presented in
Section~\ref{sec:content-delivery}. Here we discuss various types of Content
Delivery Infrastructures (CDIs) which range from Web-based Content Distribution
Networks (CDNs) to Hybrid CDNs to peer-to-peer (P2P) systems. Furthermore, in
Section~\ref{sec:challenges} we turn to the challenges that each party involved
in Internet content delivery faces separately today.

Finally, we turn to the state of the art of collaboration between networks and
content providers.  We start by outlining the collaboration incentives for each
member of the content delivery landscape in Section~\ref{sec:Incentives}. Next
we review the collaboration schemes that have been discussed in research as
well as at the Internet Engineering Task Force (IETF) in
Section~\ref{sec:Opportunities}. We briefly introduce the well-known approaches
and summarize their key functions. We then pick two collaboration schemes,
namely the P2P Oracle and the Provider-aided Distance Information System
(PaDIS) for a case study. In Section~\ref{sec:oracle} we discuss the P2P Oracle
with regards to its effect on the P2P system as well as on network operations.
Likewise, the second case study discusses the model of the Provider-aided
Distance Information System in Section~\ref{sec:te-cate}, including a large
scale analysis based on real traffic traces.  Section~\ref{sec:future} outlines
a possible future direction for collaboration between content providers and
network operators.  We conclude this part of the chapter in
Section~\ref{sec:conclusion}.

\bigskip

\noindent\textbf{Summary:} This chapter builds upon the student's
basic knowledge of how the Internet infrastructure operates, i.e., as a network
of networks. After reading this chapter the student should have a fundamental
understanding about how content distribution via the Internet works today, what
the challenges are, and which opportunities lie ahead. Moreover, the chapter
points out how all parties---including end users---can benefit from the
collaboration between ISPs and content providers. Indeed, simple, almost
intuitive, means will enable such collaboration.

### 2013-07-22 Compiled/chapter4/sections/Opportunities.tex
all chapters compile independently

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Opportunities for Collaboration}\label{sec:Opportunities}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

As pointed out ISPs are in a unique position to help CDIs and P2P systems to
improve content delivery since they have the knowledge about the state of the
underlying network topology, the status of individual links, as well as the
location of the user. In this section we first describe the high level concept
all existing solutions have in common and then continue by illustrating where
and why they differ in certain aspects.

The presented solutions include the original Oracle concept proposed by
Aggarwal et al~\cite{afs-cispp2pcip-ccr07}, P4P proposed by Xie et al.~\cite{p4p},
Ono proposed by Choffnes and Bustamante~\cite{taming} and PaDIS proposed by Poese
et al.~\cite{PaDIS-IC}. We also give an overview of the activities within the
IETF which have been fueled to some extend by the proposed systems discussed in
this section, namely ALTO and CDNi.

\subsection{Conceptual Design}\label{sec:Concept-Opportunties-Collaboration}

To overcome the challenges in Content Delivery, recall
section~\ref{sec:challenges}, various solutions have been proposed by the
research community. While they all differ in certain aspects, their basic idea
is the same: utilize available information about the network to make an
educated selection prior connecting to a service.  Following this idea, all of
the proposed solution employ the same basic conceptual design: the
\emph{management plane} is responsible for collecting up-to-date information
about the network while the \emph{control plane} acts as an interface to this
information for the application.\\

\noindent {\bf Management Plane: The Network Map.}
The systems management plane is responsible to collect up-to-date state network
information, such as network topology, routing information, link utilization
and other important metrics. This information is used to maintain an internal
map of the network representing the current state of the real network.  One
important aspect of this component is how the information about the network is
retrieved. The different implementations range from active measurements over
passive measurements to active participation in network management systems
(such as BGP). Another important aspect is the frequency in which the
information is collected. For certain information such as topology or routing
an immediate update is necessary to guarantee correct functioning of the
system, while others, such as link utilization or packet loss rates, only
degrade the quality of the system. Still other information, such as link
capacities or transmission delays, can be considered (semi-)static.  Last but
not least the systems differ in what information is necessary to be
operational and if additional information sources can be used to improve
accuracy.\\

\noindent {\bf Control Plane: The Information Interface.}
The control plane of the system is responsible for providing an interface to
the information of the management plane so that clients can make use of the
information. This can basically be seen as an interface or API that clients can
query to get information about the current network state. The various proposed
solutions differ mainly in which fashion and at which granularity the
information can be retrieved. There are two main competing approaches:
abstracted network maps and preference lists. The first one transforms the
available information from the management plane into an annotated
representation of nodes and edges. The big difference to the actual data of the
management plane is the aggregation level and the specific annotations. Clients
can then query the system to get an up-to-date abstract network map, which they
can use to decide which of the possible destination to connect to by
calculating the best candidates by themselves using their own optimization
target. The second one uses the information of the management plane to create a
ranked list of possible service destinations (read: IP addresses). The required
input includes the source, possible destinations and (if the system supports
multiple collaboration objectives) an optimization goal, \eg minimal delay. The
output consists of a re-ordered list of the possible destinations in regard to
the optimization goal, the first being the most and the last being the least
desirable destination.

Note that in both cases the client is in the position to select the final
destination, allowing to completely ignore the additional information. Another
important fact is that the client is not necessarily the end-user but might be
a service provider themselves. For instance a company providing content
delivery service (CDN) could make use of this service to improve its
user-to-server mapping accuracy or in case of the BitTorrent P2P system the
tracker could query the service prior returning an initial peer list to a
connected client.  While not strictly necessary, the two components are
usually implemented as separate entities within the system to allow better
scalability, information aggregation and/or anonymization without loosing
precision or multiple collaboration objectives. In addition to that, all
systems table important issues for any collaboration approach, such as privacy
information leakage or targeted objective(s).

The presentation of the following solutions will outline the specific
implementation and thus highlights the differences between the solutions.

\subsection{P2P Oracle Service}\label{sec:p2p-collaboration}

Aggarwal et al.~\cite{afs-cispp2pcip-ccr07} describe an \emph{oracle} service to
solve the mismatch between the overlay network and underlay routing network in
P2P content delivery. Instead of the P2P node choosing neighbors independently,
the ISP can offer a service, the \emph{oracle}, that ranks the potential
neighbors according to certain metrics: a client supplied peer list is
re-ordered based on coarse-grained distance metrics, \eg the number of AS
hops~\cite{routing-book-00}, the peer being inside/outside the AS or the
distance to the edge of the AS. This ranking can be seen as the ISP expressing
preference for certain P2P neighbors.  For peers inside the network additional
information can be used, such as access bandwidth, expected delay or link
congestion to further improve the traffic management.

\subsection{Proactive Network Provider Participation for P2P (P4P)}\label{sec:P4P}

The ``Proactive Network Provider Participation for P2P'' is another approach to
enable cooperative network information sharing between the network provider and
applications. The P4P architecture~\cite{p4p} introduces iTrackers as portals
operated by network providers that divides the traffic control responsibilities
between providers and applications. Each iTracker maintains an internal
representation of the network in the form of nodes and annotated edges.  A node
represents a set of clients that can be aggregated at different levels, \eg
certain locations (PoP) or network state (similar level of congestion).
Clients can query the iTracker to obtain the ``virtual'' cost for possible peer
candidates. This ``virtual'' cost allows the network operators to express any
kind of preferences and may be based on the provider's choice of metrics,
including utilization, transit costs, or geography. It also enables the client
to compare and choose the most suited peers to connect to.

\subsection{Ono - Travelocity-based Path Selection}\label{sec:Ono}

The Ono system~\cite{taming} by Choffnes and Bustamante is based on
``techniques for inferring and exploiting the network measurements performed by
CDNs for the purpose of locating and utilizing quality Internet paths without
performing extensive path probing or monitoring'' proposed by Su \etal
in~\cite{DraftingAkamai:SIGCOMM2006}. Based on their observations that CDN
redirection is driven primarily by latency~\cite{DraftingAkamai:SIGCOMM2006},
they formulate the following hypothesis: Peers that exhibit similar redirection
behavior of the same CDN are most likely close to each other, probably even in
the same AS.  For this each peer performs periodic DNS lookups on popular CDN
names and calculates how close other peers are by determining the cosine
similarity with their lookups. To share the lookup among the peers they use
either direct communication between Ono enabled peers or via distributed
storage solutions \eg DHT-based.  On the downside Ono relies on the precision
of the measurements that the CDNs perform and that their assignment strategy is
actually based mainly on delay. Should the CDNs change their strategy in that
regard Ono might yield wrong input for the biased peer selection the authors
envision.

When considering our design concept described above, Ono is a bit harder to fit
into the picture: Ono distributes the functionality of the management and
control planes among all participating peers.  Also, Ono does not try to
measure the network state directly, but infers it by observing Akamai's
user-to-server mapping behavior on a large scale and relies on Akamai doing the
actual measurements~\cite{Akamai-Network}, Thus the management plane of Ono
consists of recently resolved hostnames from many P2P clients. The quality of
other peers can then be assessed by the number of hostnames that resolve to the
same destination.  The control plane in Ono's case is a DHT, which allows
decentralized reads and writes of key-value pairs in a distributed manner, thus
giving access to the data of the management plane.

\subsection{Provider-aided Distance Information System (PaDIS)}\label{PaDIS-Opportunities-for-Collaboration}
\newcommand{\assignmentenabler}{{informed user-server as\-sign\-ment}\xspace}
\newcommand{\assignment}{{user-server as\-sign\-ment}\xspace}
\newcommand{\assignmentproblem}{{informed user-server as\-sign\-ment problem}\xspace}

In~\cite{PADIS2010,PaDIS-IC} Poese et al. propose a ``Provider-aided Information Systems
(PaDIS)'', a system to enable collaboration between network operators and
content delivery systems. The system enhances concept of the P2P Oracle to
include server based content delivery systems (\eg CDNs), to maintain an
up-to-date annotated map of the ISP network and its properties as well as the
state of ISP-operated servers that are open for rent. In addition, it provides
recommendations on possible locations for servers to better satisfy the demand
by the CDN and ISP traffic engineering goals.  In the management plane, it
gathers detailed information about the network topology, \ie routers and links,
annotations such as link utilization, router load as well as topological
changes. An Interior Gateway Protocol (IGP) listener provides up-to-date
information about routers and links.  Additional information, \eg link
utilization and other metrics can be retrieved via SNMP. A Border Gateway
Protocol (BGP) listener collects routing information to calculate the paths
that traffic takes through the network, including egress traffic. Ingress
points of traffic can be found by utilizing Netflow data. This allows for
complete forward and reverse path mapping inside the ISP and enables a complete
path map between any two points in the ISP network.  While PaDIS builds an
anotated map of the ISP network, it keeps the information acquired from other
components in separate data structures. This separation ensures that changes in
prefix assignments do not directly affect the routing in the annotated network
map.  Pre-calculating path properties for all paths, allow for constant lookup
speed independent of path length and network topology. On the control plane,
PaDIS makes use of the prefrence lists known from the P2P Oracle, but supports
multiple, individual optimization targets. Apart from basic default
optimizations (\eg low delay, high throughput), additional optimizations can be
negotiated between the network operator and the content delivery system.
For CDN-ISP collaboration opportunities when the ISP operates both the network and the CDN we
refer the reader to~\cite{CooperativeISPCDN,TECDN,N-CDN,CTTE2010}.

\subsection{Application-Layer Traffic Optimization (ALTO)}\label{sec:ALTO}

The research into P2P traffic localization has led the IETF to form a working
group for ``Application Layer Traffic Optimization (ALTO)''~\cite{ALTO-WG}.
The goal of the ALTO WG is to develop Internet standards that offer
``better-than-random'' peer selection by providing information about the
underlying network and to design a query-response protocol that the
applications can query for an optimized peer selection
strategy~\cite{ietf-alto-protocol}.  On the control plane, ALTO offers multiple
services to the Applications querying it, most notably are the Endpoint Cost
Service and the Map service. The Endpoint Cost Service allows the Application
the query the ALTO server for costs and rankings based on endpoints (usually IP
subnets) and use that information for an optimized peer selection process or to
pick the most suitable server of a CDI. The Network Map service makes use of
the fact that most endpoints are in fact rather close to each other and thus
can be aggregated into a single entity. The resulting set of entities is then
called an ALTO Network Map. The definition of proximity in that case depends on
the aggregation level, in one Map endpoints in the same IP subnet may be
considered close while in another all subnets attached to the same Point of
Presence (PoP) are close. In contrast to the Endpoint Cost Service the ALTO
Network Map is suitable when more Endpoints need to be considered and offers
better scalability, especially when coupled with caching techniques. Although
the ALTO WG statement is more P2P centric, the service is also suitable to
improve the connection to CDN servers.

### 2013-07-22 Compiled/chapter4/sections/Traffic-Trends-Overall.tex
all chapters compile independently

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Traffic Trends: Overall}\label{sec:traffic}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Before delving into the details of the collaborating opportunities for content
providers and infrastructures we embark on giving an overview of typical
characteristics of Internet traffic.  We start by summarizing previous studies
on how Internet traffic looks like. We consider four aspects: \first The
composition of the application mix, \second popular content-types, \third the
distribution of traffic over the course of a day, and \fourth the distribution
of connection sizes.

\subsection{Application Mix}\label{sec:related:appmix}

One constant in the Internet during the last 10 years has been its steady
growth by more than 50\perc each year~\cite{andrew03growth,telegeography2008}.
Initially, protocols such as FTP, SMTP, and NNTP were popular. Then, in about
1994, HTTP entered into the picture. Until 2000, P2P protocols such as Napster
and Gnutella became popular but were later overtaken by eDonkey and BitTorrent.
However, the traffic mix has undergone substantial changes. Therefore, we now
revisit previously reported results regarding the application mix of Internet
traffic. For this purpose we rely on various studies that report on the
application mix between 2007 and 2009 from different vantage points:

\begin{itemize*}

\item The study by Maier \etal~\cite{OnDominantCharacteristics2009}, which is
based on a subset of the traces studied in Section~\ref{sec:impact}. It was
presented at IMC\,'09.

\item Two studies by ipoque~\cite{ipoque09}, which report on different regions
in the world (Germany and Middle East).  These studies are available for
download after registration via a Web form.

\item The Arbor report~\cite{arbor} on the ATLAS Internet Observatory presented
at a recent NANOG\footnote{NANOG is the North American Network Operators
Group.} meeting.

\item The Sandvine report on ``Global Broadband Phenomena''~\cite{sandvine}.

\end{itemize*}

In order to compare the results we have to summarize and unify the traffic
categories as each study uses their own nomenclature (see
Figure~\ref{fig:related:appmix}). For this purpose we use the following seven
categories:

\begin{description}

\item [Web.] All HTTP traffic including One-Click-Hosters (OCHs or Direct
Download Providers) but excluding video and audio streaming over HTTP (\ie
Flash-Video).

\item [Streaming.] All types of streaming in the Internet including streaming
over HTTP, RTP, RTSP, RTMP, ShoutCast, etc.

\item [Usenet.] The article reading and posting system that evolved from UUnet
and which uses NNTP as protocol.

\item [BitTorrent/P2P.] The popular P2P-protocol BitTorrent and all other P2P
traffic that is not eDonkey. Note, that the P2P traffic that is not BitTorrent
or eDonkey only adds a tiny fraction.  Moreover, this category represents all
P2P traffic if the study no further subdivides P2P traffic. This is the case
for Arbor~\cite{arbor} and Sandvine~\cite{sandvine}. Note as well, that the
Arbor study~\cite{arbor} reports a table with traffic shares, stating 0.95\perc
for P2P. This table is annotated with the comment that P2P is more likely to
account for 18\perc based on payload inspection of a limited data subset.

\item [eDonkey.] Another P2P protocol, if reported.

\item [Other/known.] Other identified traffic, for details we refer to the
corresponding studies.

\item [Unclassified.] Traffic that has not been classified. Note, that the
Sandvine~\cite{sandvine} study does not mention unclassified traffic, which
either implies a minute fraction or that it is missing in the plot.

\end{description}

\begin{figure}[tbp]
\centering
\includegraphics[width=0.95\linewidth]{figures/appmix.eps}
\renewcommand{\capname}{Barplot of the application mix in the
Internet\xspace}
\caption[\capname]{\capname (unified categories) for different years,
different regions according to several
sources~\cite{OnDominantCharacteristics2009,ipoque09,arbor,sandvine}.
\capcomment (BitTorrent/P2P contains all P2P except eDonkey.)}
\label{fig:related:appmix}
\end{figure}

Looking at these statistics we find that all studies report a significant
fraction of Web traffic. Indeed, Web is dominant ($>$ 50\perc) in most studies,
followed by P2P and streaming. It is noteworthy that Usenet is responsible for
a non-negligible fraction in several studies. This is surprising and a good
example for the importance of revisiting the application mix periodically in
order to identify new trends.

In terms of P2P protocol distribution Figure~\ref{fig:related:appmix} shows
that BitTorrent is dominating and the shares of eDonkey are decreasing.  Thus,
we note that the results of Plissonneau \etal~\cite{plissonneau05} who observed
91\perc of the P2P traffic is due to eDonkey in 2004 are no longer applicable.
Indeed, the popularity among P2P protocols swapped in favor of BitTorrent. We
can also see a general trend: P2P is declining according to all studies. This
is also supported by the results of Anderson~\cite{p2pdrops}. He points out
that this decline comes with an increase in video streaming. Moreover, most of
the studies pointed out that currently One-Click-Hoster (\eg Rapidshare or
MegaUpload) are as important for file-sharing as P2P systems.

Of course there are also trends that do not impact the application mix, for
example Online Social Networks (OSNs) such as Facebook. This is due to the fact
that OSNs use HTTP and they do not transport large videos, but profile
elements. Nevertheless, OSNs are not unimportant given the huge number of OSN
users world-wide.

\subsection{Content-types in the Internet}\label{sec:related:ctypes}

Next, we turn to the popularity of content-types in the Internet. Again, we
leverage several data sources, namely Maier
\etal~\cite{OnDominantCharacteristics2009}, ipoque~\cite{ipoque09}, and Erman
\etal~\cite{network-caching}. Once more we unify the categories and present
results for contents transferred via BitTorrent, eDonkey, and HTTP. See
Figure~\ref{fig:related:ctypes} for a summary.

\begin{figure}[tbp]
\centering
\includegraphics[width=0.95\linewidth]{figures/ctypes.eps}
\renewcommand{\capname}{Barplot of content-type popularity in the
Internet\xspace}
\caption[\capname]{\capname (unified categories) for
different protocols, different regions according to several
sources~\cite{OnDominantCharacteristics2009,ipoque09,network-caching}.}
\label{fig:related:ctypes}
\end{figure}

We see that videos are the most popular content in P2P systems (BitTorrent and
eDonkey). Even in HTTP videos account for more traffic than any other category.
Although HTTP was designed to transfer Web pages (text, \eg HTML, XML, CSS,
JavaScript, and image files) these contribute less than a third of the total
HTTP volume.

Overall, a significant fraction of software and archives is noticeable.
According to Maier \etal~\cite{OnDominantCharacteristics2009} almost all videos
are in flash-video format and are served by video portals such as YouTube.
Similarly, almost all archives are served by One-Click-Hosters. This is
confirmed by the results of Erman \etal~\cite{network-caching}.

Shifts in the popularity of content-types can be another indicator of new
trends. For example, there have been almost no flash-videos before the
breakthrough of YouTube.

\subsection{Time-of-day Effects}\label{sec:related:tod}

In order to understand when people are active in the Internet we show
time-of-day usage plots of link utilization from Maier
\etal~\cite{OnDominantCharacteristics2009} in
Figure~\ref{fig:related:tod-maier}, and aggregated traffic volume
Sandvine~\cite{sandvine} in Figure~\ref{fig:related:tod-sandvine}.

\begin{figure}[tbp]
\centering
\includegraphics[angle=-90,width=0.85\linewidth]{figures/maier-tod.eps}
\renewcommand{\capname}{Timeseries of link utilization from Maier \etal}
\caption[\capname]{\capname~\cite{OnDominantCharacteristics2009}}
\label{fig:related:tod-maier}
\end{figure}

\begin{figure}[tbp]
\centering
\includegraphics[width=0.85\linewidth]{figures/sandvine-tod.eps}
\renewcommand{\capname}{Timeseries of traffic volume from Sandvine\xspace}
\caption[\capname]{\capname~\cite{sandvine}}
\label{fig:related:tod-sandvine}
\end{figure}

In general, we observe a peak utilization at prime-time around 8\,pm and a
daily low between 2\,am and 4\,am. As the data sets of all these studies are
primarily collected from residential networks, it not surprising that they all
show similar characteristics. The peak usage in the evening hours can easily be
explained by the fact that people are usually not at home during business
hours. Rising demands just before lunch and in the afternoon may be due to
children returning home from school.

### 2013-07-22 Compiled/chapter4/sections/Traffic-Trends-Server-Diversity.tex
all chapters compile independently

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Traffic Trends: Content Server Diversity}\label{sec:traffic-diversity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

So far we have highlighted that the Web and P2P protocols are responsible for a
major share of the Internet traffic. However, we have not yet explored if all
content is equally popular or if a few content providers dominate. This is the
goal of this section.

Our evaluation methodology relies on packet level traces from a large European
ISP. We analyze them towards identifying CDI infrastructures and their behavior
as seen by an ISP. Here, we find that CDIs rely on the domain Name System (DNS)
for their operation. Thus, we focus our analysis on the DNS infrastructure in
order to find the server deployment, mapping and operational behavior of CDIs.
Based on these observations, we develop classification methods to detect CDI
infrastructures and perform a first potential analysis on the impact of CDI
operation when basic ISP knowledge is available.

\subsection{Residential ISP Traces}\label{sec:traces} We base our study on
three sets of anonymized packet-level observations of residential DSL
connections collected at aggregation points within a large European ISP.  Our
monitor, using Endace monitoring cards, allows us to observe the traffic of
more than 20,000~DSL lines to the Internet.  The data anonymization,
classification, as well as application protocol specific header extraction and
anonymization is performed immediately on the secured measurement
infrastructure using the Bro NIDS~\cite{bro-paper} with dynamic protocol
detection (DPD)~\cite {dreger06dpd}.

We use an anonymized 24\,h packet trace collected in March 2010 (\martrace) for
detailed analysis of the protocol behavior. For studying longer term trends, we
used Bro's online analysis capabilities to collect an anonymized protocol
specific trace summary (\httplong) spanning 2 weeks. Additionally, we collected
an anonymized 5\,day DNS trace (\dnslong) in February 2010 to achieve a better
understanding of how hostnames are resolved by different sites. Due to the
amount of traffic at our vantage point and the resource intensive analysis, we
gathered the online trace summaries one at a time. Table \ref{tab:traces}
summarizes the characteristics of the traces, including their start, duration,
size, and protocol volume.  It is not possible to determine the exact
application mix for the protocol specific traces, as we only focus on the
specific protocol.  However, we use full traces to cross check the general
application mix evolution.

\begin{table*}
  \small
    \begin{center}
      \tabcolsep2mm
      \begin{tabular}{l|l|l@{ }l|r|r|r}

      Name      & Type     & \multicolumn{2}{|l|}{Start date} & Dur. & Size        & Application Volume \\
      \hline
      \martrace & packet   & 04 Mar'10    & 2am    & 24~h      & $>$5\,TB    & $>$ 3\,TB HTTP, $>$ 5\, GB DNS \\
      \hline
      \httplong & log file & 09 Sep'09    & 3am    & 14~d      & $>$ 200\,GB & corresponds to $>$ 40\,TB HTTP \\
      \dnslong  & packet   & 24 Feb'10    & 4pm    & 5~d       & $>$25\,GB   &  $>$ 25\,GB DNS \\
      \end{tabular}
    \end{center}
    \vspace*{-1\baselineskip}
    \caption{Summaries of anonymized traces from a European ISP}%
    \label{tab:traces}
\end{table*}

With regards to the application mix, recall Section~\ref{sec:traffic}, Maier et
al.~ \cite{OnDominantCharacteristics2009} find that HTTP, BitTorrent, and
eDonkey each contribute a significant amount of traffic, see Table
\ref{tab:traces}.  In \martrace HTTP alone contributes almost 60\perc of the
overall traffic at our vantage point, BitTorrent and eDonkey contribute more
than 10\perc. Recall that similar protocol distributions have been observed at
different times and at other locations of the same ISP, see
Figure~\ref{fig:related:appmix} summarizes the results. Note that almost all
streaming is done via the Web on top of HTTP.  Therefore, we conclude that
currently HTTP is the dominant service and P2P is still responsible for at
least 15\% of the traffic.

Analyzing \httplong, we find more than 1.2 billion HTTP requests, or 89 million
requests per day on average. This is consistent with 95 million requests in 24
hours in \martrace. The advantage of using click stream data from a large set
of residential users is their completeness. We are, e.g., not biased by the
content offered \first by a web service, \second whether sufficient users
installed measurement tools such as the \url{alexa.com} toolbar, or \third
whether users actually use some kind of Web proxy.

To identify the most popular web services, we focus on the most popular hosts.
As expected, the distribution of host popularity by volume as well as by number
of requests is highly skewed and is consistent with a Zipf-like distribution as
observed in other studies~\cite{OnDominantCharacteristics2009}. The top 10,000
hosts by volume and the top 10,000 hosts by number of requests together result
in roughly 17,500 hosts.  This indicates that on the one hand, some hosts that
are popular by volume may not be popular by number of requests and vice versa.
On the other hand, there are some hosts that are popular according to both
metrics.  The total activity by these hosts accounts for 88.5\perc of the
overall HTTP volume and more than 84\perc of the HTTP requests. Assuming that
the HTTP traffic volume accounts for roughly 60\perc of the total traffic,
similar to the observations made in September 2009~\cite{OnDominantCharacteristics2009,UGCcacheability} 
and in \martrace, more than 50\perc of the trace's total traffic is captured by
these hosts.

\begin{figure}[tbp]
  \centering
  \includegraphics[height=0.7\linewidth]{figures/dns-diversity.eps}
  \caption{DNS replies for two different sites hosted on a CDI, in two-hour bins}%
  \label{fig:dns_diversity}
\end{figure}

\subsection{Server Diversity and DNS Load Balancing}\label{sec:active_dns_measurement}

To better understand how HTTP requests are handled and assigned to servers, we
use \dnslong to analyze the 20 most heavily queried DNS names to identify
typical usage patterns. We consider only the most heavily used resolver.
Figure~\ref{fig:dns_diversity} shows two of the typical patterns for two of the
DNS names. It also shows how the resolved IP addresses change (y-axis) across
time (x-axis) for two hostnames; respectively a software site, labeled
Software1, and a media site, labeled Media1. The vertical lines annotate
midnight. If two IP addresses are plotted close to each other, this indicates
that the longest common prefix of the two addresses is close. We note that the
hostname of Software1 is mainly resolved to a single subnet, excepting a few
special cases. However, Media1 is load balanced across approximately 16
different sites. For Media1, there appears to be one main site which is almost
always available, while the remaining 15 are predominantly used during
afternoon and evening peak usage hours.

These results are promising, and show that individual sites do expose a certain
degree of server diversity to their users. While our trace (\httplong) includes
the queried hostnames, it does not include the resolved IP address, as a HTTP
request header contains the hostname but not the IP address of a server. To
verify the above behavior and get an up-to-date view of the DNS replies for the
hostnames of our trace, we used 3 hosts within the ISP to issue DNS queries to
the ISP's DNS resolver for all 17,500 hostnames repeatedly over a fourteen day
measurement period starting on Tue Apr 13th 2010. During these two weeks, we
received more than 16 million replies. Unless otherwise mentioned, we rely on
our active DNS measurements, with augmented statistics concerning volume and
requests from \httplong.

\subsection{Server Location Diversity}\label{sec:server_location_diversity}

Our analysis of hostnames and their assignment to servers in section
\ref{sec:active_dns_measurement} has shown that content can be served by
multiple servers in different locations. In fact, many domains use the service
of a \textit{Content~Delivery~Infrastructure} (CDI), which can be seen during
the DNS resolution progress: The original domain name is mapped to the domain
of a CDI, which then answers requests on behalf of the requested domain name
from one of its caches~\cite{DraftingAkamai:SIGCOMM2006}. Almost all CDIs rely
on a distributed infrastructure to handle the expected load, load spikes, flash
crowds, and special events. Additionally, this introduces needed redundancy and
fail over configurations in their services. Among the most studied CDI' are
Content Distribution Networks (CDNs), such as Akamai~\cite
{ImprovingPerformanceInternet2009, DraftingAkamai:SIGCOMM2006,
MeasuringAndEvaluating2008}, and Content Delivery Platforms (CDPs), such as
Google~\cite{MovingBeyondE2E2009} and their YouTube service~\cite {YouTube}.

\begin{figure}[tbp]
   \center
   \includegraphics[height=0.7\linewidth,angle=-90]{figures/avgResponseSize.ps}\\
   \includegraphics[height=0.7\linewidth,angle=-90]{figures/avgResponseSubnetSize.ps}

  \caption{CCDF of mean \# of IPs (top) and subnets (bottom) per DNS reply for the ISPs DNS resolver.}
  \label{fig:CDF-AVG-REPLY-IPs}
\end{figure}

The DNS server can choose to return one or more server IP addresses based on
the domain name in the request and the IP address of the requesting DNS
resolver. For example, it may use a geo-location database~\cite{MaxMind} to
localize the region of the DNS resolver, utilize BGP data to identify the ISP,
create a topology map derived via traceroutes, or any combination of these and
other topological and geographic localization techniques. A DNS server has, in
principle, two methods for load balancing across multiple servers:

\begin{description*}
\item[MultQuery:] Can return multiple IP addresses within a single DNS response
\item[CrossQuery:] Can return different IP addresses for repeated queries and
  thus perform DNS redirection.
\end{description*}

In our active DNS measurements, we found that often a mixture of MultQuery and
CrossQuery is being used in practice. Furthermore, we used the measurement
results to \first map hostnames to sets of IP addresses and \second check the
IP address diversity of these sets for a better understanding of server
diversity and their location. We achieved this by aggregating the returned IP
addresses into subnets based on BGP information obtained from within the
ISP. This allows for detailed information about the different locations within
the ISP, while giving an aggregated view of subnets reachable via peering
links.

Another issue stems from the fact that the IP address returned by the CDI
depends on the IP address of the ISP DNS
resolver~\cite{DNS-IMC-2010,dns-redirection,DraftingAkamai:SIGCOMM2006}. Due to
this, we used the DNS resolver of the ISP of our vantage point as well as
external DNS resolvers (see section \ref{sec:dns_resolvers}). The former
reflects the experience of most of the clients at our vantage point\footnote{We
verify using the traces that more than 95\perc of the clients use the ISP's DNS
resolver as their default one. }. The latter lets us discover additional
diversity as well as understand the preference of the CDI for this specific
ISP.

\begin{figure}[tbp]
  \center
   \includegraphics[height=0.7\linewidth,angle=-90]{figures/14day-returnedIPCDF-bytes.ps}
   \includegraphics[height=0.7\linewidth,angle=-90]{figures/14day-returnedIPCDF-hits.ps}
  \caption{CDF of \# of IPs for the ISP DNS resolver normalized by traffic
    volume  (top) and requests (bottom) including aggregation on
    domain levels. (Logarithmic x-axis.)}
  \label{fig:CDF-IPPAF-IPs}
\end{figure}

\begin{figure}[tbp]
  \center
   \includegraphics[height=0.7\linewidth,angle=-90]{figures/SubnetsByBytes.ps}
   \includegraphics[height=0.7\linewidth,angle=-90]{figures/SubnetsByHits.ps}
  \caption{CDF of \# of subnets for ISP DNS resolver normalized by traffic
    volume  (top) and by requests (bottom) including aggregation on
    domain levels. (Logarithmic x-axis.)}
  \label{fig:CDF-IPPAF-Subnets}
\end{figure}

\paragraph{Prevalence of MultQuery.} We start our analysis by checking the
prevalence of the first form of DNS based load balancing, MultQuery. Figure~
\ref{fig:CDF-AVG-REPLY-IPs} shows a CCDF plot of the average number of IP
addresses (top) and subnets (bottom) per DNS reply. In addition, we included
the same data normalized by traffic volume and number of requests.

A first observation is that the number of returned IP addresses per request is
rather small. The median is $1$, the average is $1.3$ and even the $0.9$
percentile is $2$. We note that even when an answer yields multiple IP
addresses, the majority of them are from the same subnet. Therefore, the
diversity decreases even further if we aggregate to subnets. From a network
perspective, this implies that there is not much choice, neither for the ISP
nor for the user, regarding where to download the content from. Both are
limited to the information provided by the DNS server. However, when we
normalize the hosts by their respective popularity, we see a significant
improvement. More than $29$\% of the volume and $19$\% of requests have a
choice among at least $2$ IP addresses.

\paragraph{Prevalence of CrossQuery.} Next, we check how prevalent CrossQuery,
the second form of DNS based load balancing is. Since CrossQuery returns
different IP addresses for repeated queries, its potential contribution to
server diversity can only be studied by aggregating across time. The lines
labeled \texttt{Full Domain Name} in Figures~\ref{fig:CDF-IPPAF-IPs} and~\ref
{fig:CDF-IPPAF-Subnets} capture this case.

We find that more than $50$\perc of the volume or requests can be served by
more than one IP address. Similarly, there is choice between at least two
subnets over $40$\perc of the time across both metrics, see Figure~\ref
{fig:CDF-IPPAF-Subnets}. This indicates that there is significant potential for
the ISP to bias the location preference of the CDI.

\begin{figure}[tbp]
  \center\includegraphics[height=0.7\linewidth,angle=-90]{figures/maxTTLCDF.ps}
  \caption{CDF of DNS TTL value by traffic volume and by number of requests.}
  \label{fig:CDF-IPPAF-TTL}
\end{figure}

\paragraph{Subdomain Aggregation.} Since some CDIs only use subdomains as hints
about the context of the requested URLs or the requested services, we
accumulate the answers further regarding the 2nd and 3rd part of the domain
names of the hosts, see Figures~\ref {fig:CDF-IPPAF-IPs}
and~\ref{fig:CDF-IPPAF-Subnets} at the respective data series called
\texttt{3rd Level Domain} and \texttt{2nd Level Domain}. For example, we might
accumulate the IP addresses from DNS replies for \url {dl1.example.org} and
\url{dl2.example.org} for the statistics on the 2nd level domain, but not the
third level domain.

This is a feasible approach, since many hosts respond to all requests that
belong to a subset of the subnets returned when accumulating by the
second-level domain of DNS resolver answer, including recursive requests and
redirections.  This behavior was verified with active measurements
in~\cite{PADIS2010}. We find that at least two major CDIs, a streaming provider
and a One-Click Hoster, serve requested content from servers that match in
their second level domain.

We note that the accumulation by third-level domain, and especially by second
level domain significantly increases the number of observed subnets per request
both normalized by requests as well as by volume.  The number of returned
subnets further increases when accumulating to the second-level domain of DNS
resolver answer. Studying our traces in more detail, we find that this is due
to the substantial traffic volume and number of requests that are served by
CDIs, some of which are highly distributed within ISPs or located in multihomed
datacenters or peer-exchange points.

\paragraph{Infrastructure Redirection Aggregation.}
Taking a closer look at the DNS replies~\cite{ietf-dns}, we find that some CDIs
use CNAME records to map queried hostname to an A record. These A records show
the same pattern as the hostnames in the previous section: the second level
domain is identical. Similar to the previous approach, we can aggregate by
these A records.

Turning our attention to the implications of the proposed aggregation schemes,
we notice the available diversity increases tremendously. More than 50\% of the
hits and 70\% of the bytes can be served by more than $20$ servers.  With
regards to subnets, the diversity decreases slightly. Nevertheless, more than
$5$ subnets are available for 45 \% of the hits and 55\% of the bytes.

If we consider aggregation periods in the order of tens of minutes, the numbers
do not decrease by much.  The reason that most of the diversity is observable
even over these short aggregation time periods, is that the typical TTL, see
Figure~\ref{fig:CDF-IPPAF-TTL}, is rather short with a mean of $2,100$ seconds
and an median of $300$ seconds normalized by volume. When weighted by requests,
the mean/median is $4,100$/$300$ seconds.

\paragraph{Alternative DNS Resolvers.}\label{sec:dns_resolvers}
So far we have only considered the effect of content diversity when the ISP DNS
resolver is used. To understand how much the DNS load balancing deployed by a
CDI is biased by the queried DNS resolver, we repeat the experiment from
Section~\ref{sec:active_dns_measurement} using two other DNS resolvers. In
particular, we pick the next most popular DNS resolvers found in our traces:
GoogleDNS and OpenDNS. Both are third-party resolvers with a global footprint
and utilize DNS anycast.

Comparing the results, we find that we attain more IP address diversity and
subnet diversity when using the ISP DNS resolver. This is mainly due to the
fact that CDIs select the supplied caches based on the source IP address of the
querying DNS resolver. Since the CDIs are no longer able to map the request to
the AS it originates from, but rather to AS the DNS resolver belongs to, the
server selection by the CDI cannot optimize for the location of the DNS client.

A possible solution to the problem is the EDNS-Client-Subnet
extension~\cite{EDNS}, an extension that utilizes the EDNS0 option field that
is used today by DNS Security Extensions (DNSSEC). A recent study~\cite{EDNS:IMC2012} showed that the
user-to-server allocation can be significantly improved as well as the
end-to-end performance for the client. On the other hand, this requires that
all the involved resolvers and authoritative servers in ISPs, CDNs, third
parties that maintain resolvers and authoritative servers, \eg GoogleDNS, OpenDNS, 
have to support EDNS-Client-Subnet extension.

\subsection{Impact on Traffic Localization}\label{sec:dns_localization}

\begin{table}
  \begin{center}
    \tabcolsep2mm{}
    \begin{tabular}{|c||c|c||c|c||c|c|}
      \hline      & \multicolumn{2}{c||}{ISP DNS} & \multicolumn{2}{c||}{OpenDNS} & \multicolumn{2}{c|}{GoogleDNS}\\
      \hline Metric  & observed  & potential & observed & potential   & observed  & potential   \\ \hline
      \hline IPs     & 12.3\perc & 24.2\perc & 5.8\perc  & 16.0\perc  & 6.0\perc  & 9.7\perc    \\
      \hline requests& 14.9\perc & 33.2\perc & 4.7\perc  & 18.8\perc  & 4.8\perc  & 6.4\perc    \\
      \hline volume  & 23.4\perc & 50.0\perc & 12.0\perc & 27.7\perc  & 12.3\perc & 13.4\perc   \\
      \hline
    \end{tabular}
  \end{center}

  \caption{Traffic localization within the network by different DNS resolvers normalized by number of requests and
  traffic volume together with the potentially available fraction of localized traffic.}
  \label{tab:TraceLocallizedTraffic}
\end{table}

Analyzing the three active DNS measurements from the ISP, OpenDNS as well as
Google DNS resolver, we find that a significant part of the requests that could
have been in principle served by sources within the ISP are directed towards
servers that are outside of the ISP.  However, before tackling this issue, we
need to understand what fraction of the traffic may be served by IP addresses
within the ISP's network and what fraction is served by IP addresses outside of
the AS. To this end, we analyze each of the three active DNS traces
separately. For each trace, we start by classifying all DNS replies regarding
the \texttt{redirection} aggregation described in Section~
\ref{sec:server_location_diversity} and account the volume (or hits) evenly to
each of the IP addresses.  Next, we classify the IP addresses in two groups -
inside and outside of the ISP network.  Table~\ref {tab:TraceLocallizedTraffic}
summarizes the results of this aggregation regarding the traffic and hits that
were kept inside the ISP's network in the columns labeled \texttt{observed}.

Turning to the results, we find that there is hardly any difference between
those clients that use the external DNS resolvers, \ie GoogleDNS or OpenDNS. Of
the returned IP addresses, less than $6$\perc are within the AS. When weighted
by number of requests, this does not change much. However, when normalizing by
volume, about $12$\perc of the traffic stays within the AS. In contrast,
clients that use the ISP's DNS resolver fare better: almost a quarter of the
traffic volume is served from servers within the AS. Normalized by requests, we
see a three fold increase, and normalized by hits or volume, roughly a two fold
increase over using external DNS resolvers.  Among the reasons for the ``bad''
performance of external DNS resolvers is that some CDIs may always return IP
addresses outside the ISP, despite the fact that many of its servers are
deployed within the ISP. The reason behind this is that the CDIs cannot map the
DNS resolver to the AS anymore, and thus are unaware of the origin of the
request. This explains the substantial difference and highlights on the one
hand the effectiveness of the CDI optimization, but also points out its
limits. As such, it is not surprising that there are efforts under way within
the IETF to include the source IP addresses of the DNS client in the DNS
requests~\cite{DNS-extension-IP-client}.

However, one can ask if the CDI utilizes the full potential of traffic
localization on an AS level. For this, we check the potential of traffic
localization, by changing the volume (or hit) distribution from even to greedy.
Thus, as soon as we observe at least one IP address inside the ISP's network,
we count all traffic for the entire aggregation to be
internal. Table~\ref{tab:TraceLocallizedTraffic} shows the results in the
columns labeled \texttt {potential} for all three DNS traces. Note the
substantial differences. Our results indicate that a gain of more than a factor
of two can be achieved. Furthermore, up to 50\perc of the traffic can be
delivered from servers within the ISP rather than only 23.4\perc. This may not
only in itself result in a substantial reduction of costs for the ISP, but it
also points out the potential of collaboration between CDIs and ISPs. While the
increase is noticeable it is nowhere near that of the ISP's DNS resolver. The
potential benefit when relying on GoogleDNS is rather small. A deeper study on
our results unveils that content served by highly distributed and redundant
infrastructures can be localized the most.

\subsection{Summary}\label{sec:diversity_summary}
We find that HTTP is again the dominant traffic source, while the prevalence of
P2P traffic decreases. Since most CDIs rely on distributed infrastructure, we
not only observe significant server location diversity but also significant
path diversity for accessing HTTP based content. Indeed, there is the potential
to bias roughly half of the overall traffic by redirecting queries to different
content servers.

More precisely, we estimate that around $70$\perc of the HTTP traffic in a big
European ISP can be redirected when taking advantage of the diversity due to
MultQuery, CrossQuery and hostname aggregation.  Furthermore, we show that
current CDI optimizations that approximate the location of end-users based on
the location of the local DNS resolvers are more effective than those based on
the location of third-party resolvers. Finally, we show that the traffic
localization potential within the above mentioned ISP is very high especially
when the ISP DNS resolver is utilized.

### 2013-07-22 Compiled/chapter5/conclusion.tex
all chapters compile independently

%\ed{A one page conclusion.} 
%Something on evolution of transport
%Something on Middleboxes
%Something on MPTCP
%Something on Minion
%One paragraph on the future

The Transport Layer in the Internet evolved for nearly two decades,
but it has been stuck for over a decade now.
A proliferation of middleboxes in the Internet,
devices in the network that look past the IP header,
has shifted the waist of the Internet hourglass 
upward from IP
to include UDP and TCP,
the legacy workhorses of the Internet.
While popular for many
different reasons, 
middleboxes thus deviate from the Internet’s
end-to-end design, creating large deployment
“black-holes”---singularities where legacy transports get through, but
any new transport technology or protocol fails,
severely limiting transport protocol evolution.
The fallout of this ossification
is that new transport protocols,
such as SCTP and DCCP,
that were developed to offer much needed 
richer end-to-end services to applications,
have had trouble getting deployed since they require changes 
to extant middleboxes.

Multipath TCP is perhaps the most significant change to TCP in the past twenty
years. It allows existing TCP applications to achieve better
performance and robustness over today's networks, and it has been
standardized at the IETF. The Linux kernel implementation shows that
these benefits can be obtained in practice.  However, as with any
change to TCP, the deployment bar for Multipath TCP is very high: only
time will tell whether the benefits it brings will outweigh the added
complexity it brings in the end-host stacks.

The design of Multipath TCP has been a lengthy, painful process that took around five
years. Most of the difficulty came from the need to support existing middlebox
behaviors, while offering the exact same service to applications as TCP. Although
the design space seemed wide open in the beginning, in the end we were \emph{just} able
to evolve TCP this way: for many of the design choices there was only one viable option
that could be used. When the next major TCP extension is designed in a network with even more
middleboxes, will we, as a community, be as lucky?

A pragmatic answer to the inability to deploy new transport protocols
is Minion. It allows deploying new transport services 
by being backward compatible with middleboxes by encapsulating new protocols inside TCP.
Minion demonstrates that it is possible to obtain
unordered delivery and multistreaming
from wire-compatible TCP and TLS streams
with surprisingly small changes to TCP stacks
and application-level code.
Minion offers a path toward
the performance benefits of unordered delivery,
which we expect to be useful to applications
that use TCP for a variety of pragmatic reasons.

%Minion is just the latest in an ever-growing encapsulation war that 
%has engulfed end-hosts and network providers for years on end. 
Early in the Internet's history,
all IP packets could travel freely through the Internet, as IP was the narrow waist of
the protocol stack. 
Eventually, 
apps started using UDP and TCP exclusively,
and some, such as Skype, used them adaptively,
probably due to security concerns
in addition to the increasing proliferation of middleboxes that allowed only UDP and TCP through.
As a result, UDP and TCP over IP were then perceived to constitute the new waist of the Internet.
(We'll note that HTTP has also recently
been suggested as the new waist~\cite{popa10http}.) 
%UDP is not that great getting through newer firewalls either, so TCP is the new narrow
%waist, or even HTTP as recently proposed.

Our observation is that
whatever the new waist is,
middleboxes will embrace it and optimize for it: 
if MPTCP and/or Minion become popular, 
it is likely that middleboxes will be devised that understand these protocols
to optimize for the most successful use-case of these protocols
and to help protect any vulnerable applications using them.
One immediate answer from an application would be to
use the encrypted communication proposed in Minion---%
but actively hiding information from a network operator
can potentially encourage the network operator
to embed middleboxes that intercept encrypted connections,
effectively mounting man-in-the-middle attacks
to control traffic over their network,
as is already being done in several corporate firewalls~\cite{marko10using}.
%a potential serious security breach, and many ISPs
%may either drop traffic or require their customers to ``install'' keys in the 
%browser that effectively allow middleboxes to become a man-in-the-middle of the TLS connection. 
To bypass these middleboxes,
new applications may encapsulate their data {\em even} deeper,
leading to a vicious circle resembling an ``arms race'' for control
over network use.
% them looking even deeper in the packets. As we can see, 
%this is a vicious circle that no-one can ever win.

This ``arms race'' is a symptom of a fundamental tussle between
end-hosts and the network: end-hosts will always want to deploy new
applications and services, while the network will always want to allow and optimize only
existing ones~\cite{tussle}.
To break out of this vicious circle,
we propose that
end-hosts and the network must co-operate,
and that they must build cooperation into their protocols.
Designing and providing protocols and incentives for this cooperation may hold the key to creating a
truly evolvable transport (and Internet) architecture.

### 2013-07-22 Compiled/chapter5/exercises.tex
all chapters compile independently

This appendix contains a few exercises on the evolution of transport protocols and their interactions with middleboxes.

\subsection{Transport protocols}

\begin{enumerate}

\item TCP provides a reliable transport service. Assuming that you control the two endpoints of a connection, how would you modify the TCP protocol to provide an unreliable service ? Explore two variants of such a transport service :
  \begin{itemize}
  \item an unreliable bytestream where bytes can be corrupted but where there are no losses
  \item an unreliable bytestream that prevents data corruption but can deliver holes in the bytestream
  \end{itemize}

\item Same question as above, but for SCTP.

\item TCP provides a connection-oriented bytestream service. How would you modify TCP to support a message-oriented service. Consider two variants of this service :
 \begin{itemize}
    \item A connection-oriented message-mode service that supports only small messages, i.e. all messages are smaller than one segment.
    \item A connection-oriented message-mode service that supports any message length. 
 \end{itemize}

\item The large windows extension for TCP defined in \cite{rfc1323} uses the \texttt{WScale} option to negotiate a scaling factor which is valid for the entire duration of the connection. Propose another method to transport a larger window by using a new type of option inside each segment. What are the advantages/drawbacks of this approach compared to \cite{rfc1323} assuming that there are no middleboxes ?

\item One of the issues that limits the extensibility of TCP is the limited amount of space to encode TCP options. This limited space is particularly penalizing in the SYN segment. Explore two possible ways of reducing the consumption of precious TCP option-space
\begin{itemize}
\item Define a new option that packs several related options in a smaller option. For example try to combine SACK with Timestamp and Window scale in a small option for the SYN segment. 
\item Define a compression scheme that allows to pack TCP options in fewer bytes. The utilisation of this compression scheme would of course need to be negotiated during the three way handshake.
\end{itemize}

\end{enumerate}

\subsection{Middleboxes}

Middleboxes may perform various changes and checks on the packets that they process.  Testing  real middleboxes can be difficult because it involves installing complex and sometimes costly devices. However, getting an understanding of the interactions between middleboxes and transport protocols can be useful for protocol designers.

A first approach to understand the impact of middleboxes on transport protocols is to emulate the interference caused by middleboxes.This can be performed by using \texttt{click} \cite{Kohler_click:2000} elements that emulate the operation of middleboxes \cite{Hesmans_click:2013} : 
\begin{itemize}
\item \texttt{ChangeSeqElement} changes the sequence number in the TCP header of processed segments to model a firewall that randomises sequence numbers
\item \texttt{RemoveTCPOptionElement} selectively removes a chosen option from processed TCP segments
\item \texttt{SegSplitElement} selectively splits a TCP segment in two different segments and copies the options in one or both segments
\item \texttt{SegCoalElement} selectively coalesces consecutive segments and uses the TCP option from the first/second segment for the coalesced one
\end{itemize}

Using some of these \texttt{click} elements, perform the following tests with one TCP implementation.

\begin{enumerate}

\item Using a TCP implementation that supports the timestamp option defined in \cite{rfc1323} evaluate the effect of removing this option in the \texttt{SYN}, \texttt{SYN+ACK} or regular TCP segments with the \texttt{RemoveTCPOptionElement} \texttt{click} element.

\item Using a TCP implementation that supports the selective acknowledgement option defined in \cite{rfc2018} predict the effect randomizing the sequence number in the TCP header without updating anything in this option as done by some firewalls. Use the \texttt{ChangeSeqElement} \texttt{click} element to experimentally verify your answer. Instead of using random sequence numbers, evaluate the impact of logarithmically increasing/decreasing the sequence numbers (i.e. +10, +100, +1000, +1000, \ldots)

\item Recent TCP implementations support the large windows extension defined in \cite{rfc1323}. This extension uses the \texttt{WScale} option in the \texttt{SYN} and \texttt{SYN+ACK} segments. Evaluate the impact of removing this option in one of these segments with the \texttt{RemoveTCPOptionElement} element. For the experiments, try to force the utilisation of a large receive window by configuring your TCP stack.

\item Some middleboxes split or coalesce segments. Considering Multipath TCP, discuss the impact of splitting and coalescing segments on the correct operation of the protocol. Use the Multipath TCP implementation in the Linux kernel and the \texttt{SegCoalElement} and \texttt{SegSplitElement} \texttt{click} elements to experimentally verify your answer.

%\item Using the Multipath TCP implementation in the Linux kernel and one of the above mentioned \texttt{click} elements,

\item The extensibility of SCTP depends on the utilisation of chunks. Consider an SCTP-aware middlebox that recognizes the standard SCTP chunks but drops the new ones. Consider for example the partial-reliability extension defined in \cite{rfc3758}. Develop a \texttt{click} element that allows to selectively remove a chunk from processed segments and evaluate experimentally its impact on SCTP.

\end{enumerate}

Another way to evaluate middleboxes is to try to infer their presence in a network by sending probe packets. This is the approach used by Michio Honda and his colleagues in \cite{honda2011still}. However, the TCPExposure software requires the utilisation of a special server and thus only allows to probe the path towards this particular server. An alternative is to use \texttt{tracebox} \cite{Detal_Tracebox:2013}. \texttt{tracebox} is an extension to the popular \texttt{traceroute} tool that allows to detect middleboxes on (almost) any path. \texttt{tracebox} sends TCP and UDP segments inside IP packets that have different Time-To-Live values like \texttt{traceroute}. When an IPv4 router receives an IPv4 packet whose TTL is going to expire, it returns an ICMPv4 \emph{Time Exceeded} packet that contains the offending packet. Older routers return in the ICMP the IP header of the original packet and the first 64 bits of the payload of this packet. When the packet contains a TCP segment, these first 64 bits correspond to the source and destination ports and the sequence number. However, recent measurements show that a large fraction of IP routers in the Internet, notably in the core, comply with \cite{rfc1812} and thus return the complete original packet. \texttt{tracebox} compares the packet returned inside the ICMP message with the original one to detect any modification performed by middleboxes. All the packets sent and received by \texttt{tracebox} are recorded as a libpcap file that can be easily processed by using \texttt{tcpdump} or \texttt{wireshark}.

\begin{enumerate}

\item Use \texttt{tracebox} to detect whether the TCP sequence numbers of the segments that your host sends are modified by intermediate firewalls or proxies.

\item Use \texttt{tracebox} behind a Network Address Translator to see whether \texttt{tracebox} is able to detect the modifications performed by the NAT. Try with TCP, UDP and regular IP packets to see whether the results vary with the protocol. Analyse the collected packet traces.

\item Some firewalls and middleboxes change the \texttt{MSS} option in the \texttt{SYN} segments that they process. Can you explain a possible reason for this change ? Use \texttt{tracebox} to verify whether there is a middlebox that performs this change inside your network.

\item Use \texttt{tracebox} to detect whether the middleboxes that are deployed in your network allow new TCP options, such as the ones used by Multipath TCP, to pass through.

\item Extend \texttt{tracebox} so that it supports the transmission of SCTP segments containing various types of chunks.

\end{enumerate}

\subsection{Multipath TCP}

Although Multipath TCP is a relatively young extension to TCP, it is already possible to perform interesting experiments and simulations with it. The following resources can be useful to experiment with Multipath TCP :

\begin{itemize}
\item \url{http://www.multipath-tcp.org} provides an implementation of Multipath TCP in the Linux kernel with complete source code and binary packages. This implementation covers most of \cite{rfc6824} and supports the coupled congestion control \cite{rfc6356} and OLIA \cite{olia}. Mininet and netkit images containing the Multipath TCP kernel are available from the above website.
\item \url{http://caia.swin.edu.au/urp/newtcp/mptcp/} provides a kernel patch that enables Multipath TCP in the FreeBSD-10.x kernel. This implementation only supports a subset of \cite{rfc6824}
\item The \texttt{ns-3} network simulator\footnote{See \url{http://www.nsnam.org/}.} contains two forms of support for Multipath TCP. The first one is by using a Multipath TCP model\footnote{See \url{https://code.google.com/p/mptcp-ns3/}}. The second is by executing a modified Linux kernel inside \texttt{ns-3}  by using Direct Code Execution\footnote{See \url{http://www.nsnam.org/projects/direct-code-execution/}}.

\end{itemize}

Most of the exercises below can be performed by using one of the above mentioned simulators or implementation.

\begin{enumerate}

\item Several congestion control schemes have been proposed for Multipath TCP and some of them have been implemented. Compare the performance of the congestion control algorithms that your implementation supports.

\item The Multipath TCP congestion control scheme was designed to move traffic away from congested paths. TCP detects congestion through losses. Devise an experiment using one of the above mentioned simulators/implementation to analyse the performance of Multipath TCP when losses occur.

\item The non-standard \texttt{TCP\_INFO} socket option\cite{Pfeiffer_Measuring:2007} in the Linux kernel allows to collect information about any active TCP connection. Develop an application that uses \texttt{TCP\_INFO} to study the evolution of the Multipath TCP congestion windows.

\item Using the Multipath TCP  Mininet or netkit image, experiment with Multipath TCP's fallback mechanism by using \texttt{ftp} to transfer files through a NAT that includes an application level gateway. Collect the packet trace and verify that the fallback works correctly.

\end{enumerate}

%\subsection {Minion}

%\ed{Any suggestion from Jana ?}

### 2013-07-22 Compiled/chapter5/internet.tex
all chapters compile independently

The TCP/IP protocol suite was designed with the end-to-end principle \cite{saltzer1984end} in mind. TCP and SCTP are no exception to this rule. They both assume that the network contains relays that operate only at the physical, datalink and network layers of the reference models.

In such an end-to-end Internet, the payload of an IP packet is never modified inside the network and any transport protocol can be used above IPv4 or IPv6. Today, this behavior corresponds to some islands in the Internet like research backbones and some university networks. Measurements performed in enterprise, cellular and other types of commercial networks reveal that IP packets are processed differently in deployed networks \cite{honda2011still,wang2011untold}.

In addition to the classical repeaters, switches and routers, currently deployed networks contain various types of middleboxes \cite{rfc3234}.  Middleboxes were not part of the original TCP/IP architecture and they have evolved mainly during the last decade. A recent survey in enterprise networks reveals that such networks contain sometimes as many middleboxes as routers \cite{sherry2012making}.

A detailed survey of all possible middleboxes is outside the scope of this chapter, but it is useful to study the operation of some important types of middleboxes to understand their impact on transport protocols and how transport protocols have to cope with them.

\subsection{Firewalls}

Firewalls perform checks on the received packets and decide to accept or discard them based on configured security policies. Firewalls play an important role in delimiting network boundaries and controlling incoming and outgoing traffic in enterprise networks. In theory, firewalls should not directly affect transport protocols, but in practice, they may block the deployment of new protocols or extensions to existing ones. Firewalls can either filter packets on the basis of a white list, i.e. an explicit list of allowed communication flows, or a black list, i.e. an explicit list of all forbidden communication flows. Most enterprise firewalls use a white list approach. The network administrator defines a set of allowed communication flows, based on the high-level security policies of the enterprise and configures the low-level filtering rules of the firewall to implement these policies. With such a whitelist, all flows that have not been explicitly defined are forbidden and the firewall discards all packets that do not match an accepted communication flow. This unfortunately implies that a packet that contains a different \emph{Protocol} than the classical TCP, ICMP and UDP protocols will usually not be accepted by such a firewall. This is a major hurdle for the deployment of new transport protocols like SCTP.

Some firewalls can perform more detailed verification and maintain state for each established TCP connection. Some of these stateful firewalls are capable of verifying whether a packet that arrives for an accepted TCP connection contains a valid sequence number. For this, the firewall maintains state for each TCP connection that it accepts and when a new data packet arrives, it verifies that it belongs to an established connection and that its sequence number fits inside the advertised receive window. This verification is intended to protect the hosts that reside behind the firewall from packet injection attacks despite the fact that these hosts also need to perform the same verification.

Stateful firewalls may also limit the extensibility of protocols like TCP. To understand the problem, let us consider the large windows extension defined in \cite{rfc1323}. This extension fixes one limitation of the original TCP specification. The TCP header \cite{rfc793} includes a 16-bits field that encodes the receive window in the TCP header. A consequence of this choice is that the standard TCP cannot support a receive window larger than 64 KBytes. This is not large enough for high bandwidth networks. To allow hosts to use a larger window, \cite{rfc1323} changes the semantics of the \emph {receive window} field of the TCP header on a per-connection basis. \cite{rfc1323} defines the \texttt{WScale} TCP option that can only be used inside the \texttt{SYN} and \texttt{SYN+ACK} segments. This extension allows the communicating hosts to maintain their receive window as a 32 bits field. The \texttt{WScale} option contains as parameter the number of bits that will be used to shift the 32-bits window to the right before placing the lower 16 bits in the TCP header. This shift is used on a TCP connection provided that both the client and the server have included the \texttt{WScale} option in the \texttt{SYN} and \texttt{SYN+ACK} segments.

Unfortunately, a stateful firewall that does not understand the \texttt{WScale} option, may cause problems. Consider for example a client and a server that use a very large window. During the three-way handshake, they indicate with the \texttt{WScale} option that they will shift their window by $14$ bits to the right. When the connection starts, each host reserves $2^{17}$ bytes of memory for its receive window\footnote{It is common to start a TCP connection with a small receive window/buffer and automatically increase the buffer size during the transfer \cite{Semke:1998hf}.}. Given the negotiated shift, each host will send in the TCP header a window field set to \texttt{0000000000000100}. If the stateful firewall does understand the \texttt{WScale} option used in the \texttt{SYN} and \texttt{SYN+ACK} segments, it will assume a window of 4 bytes and will discard all received segments. Unfortunately, there are still today stateful firewalls\footnote{See e.g. \url{http://support.microsoft.com/kb/934430}} that do not understand this TCP option defined in 1992.

Stateful firewalls can perform more detailed verification of the packets exchanged during a TCP connection. For example, intrusion detection and intrusion prevention systems are often combined with traffic normalizers \cite{Vutukuru_TCP:2008,handley2001network}. A traffic normalizer is a middlebox that verifies that all packets obey the protocol specification. When used upstream of an intrusion detection system, a traffic normalizer can for example buffer the packets that are received out-of-order and forward them to the IDS once they are in-sequence.

\subsection{Network Address Translators}

A second, and widely deployed middlebox, is the Network Address Translator (NAT). The NAT was defined in \cite{rfc1631} and various extensions have been developed over the years. One of the initial motivation for NAT was to preserve public IP addresses and allow a group of users to share a single IP address. This enabled the deployment of many networks that use so-called \emph{private addresses} \cite{rfc1918} internally and rely on NATs to reach the global Internet. At some point, it was expected that the deployment of IPv6 would render NAT obsolete. This never happened and IPv6 deployment is still very slow \cite{Dhamdhere_IPv6:2012}. Furthermore, some network administrators have perceived several benefits with the deployment of NATs \cite{rfc2993} including : reducing the exposure of addresses, topology hiding, independence from providers, \ldots Some of these perceived advantages have caused the IETF to consider NAT for IPv6 as well, despite the available address space. Furthermore, the IETF, vendors and operators are considering the deployment of large scale Carrier Grade NATs (CGN) to continue to use IPv4 despite the depletion of the addressing space \cite{Perreault_CGN:2012} and also to ease the deployment of IPv6 \cite{rfc6264}. NATs will remain a key element of the deployed networks for the foreseeable future.

There are different types of NATs depending on the number of addresses that they support and how they maintain state. In this section, we concentrate on a simple but widely deployed NAT that serves a large number of users on a LAN and uses a single public IP address. In this case, the NAT needs to map several private IP addresses on a single public address. To perform this translation and still allow several internal hosts to communicate simultaneously, the NAT must understand the transport protocol that is used by the internal hosts. For TCP, the NAT needs to maintain a pool of TCP port numbers and use one of the available port as the source port for each new connection initiated by an internal host. Upon reception of a packet, the NAT needs to update the source and destination addresses, the source (or destination) port number and also the IP and TCP checksums. The NAT performs this modification transparently in both directions. It is important to note that a NAT can only change the header of the transport protocol that it supports. Most deployed NATs only support TCP \cite{rfc5382}, UDP \cite{rfc4787} and ICMP \cite{rfc5508}. Supporting another transport protocol on a NAT requires software changes \cite{Hayes_NAT:2008} and few NAT vendors implement those changes. This often forces users of new transport protocol to tunnel their protocol on top of UDP to traverse NATs and other middleboxes \cite{rfc6773,Tuexen_Encap:2013}. This limits the ability to innovate in the transport layer.

NAT can be used transparently by most Internet applications. Unfortunately, some applications cannot easily be used over NATs \cite{rfc3027}. The textbook example of this problem is the File Transfer Protocol (FTP) \cite{rfc959}. An FTP client uses two types of TCP connections : a control connection and data connections. The control connection is used to send commands to the server. One of these is the \texttt{PORT} command that allows to specify the IP address and the port numbers that will be used for the data connection to transfer a file. The parameters of the \texttt{PORT} command are sent using a special ASCII syntax \cite{rfc959}. To preserve the operation of the FTP protocol, a NAT can translate the IP addresses and ports that appear in the IP and TCP headers, but also as parameters of the \texttt{PORT} command exchanged over the data connection. Many deployed NATs include Application Level Gateways (ALG) \cite{rfc3027} that implement part of the application level protocol and modify the payload of segments. For FTP, after translation a \texttt{PORT} command may be longer or shorter than the original one. This implies that the FTP ALG needs to maintain state and will have to modify the sequence/acknowledgment numbers of all segments sent over a connection after having translated a \texttt{PORT} command. This is transparent for the FTP application, but has influenced the design of Multipath TCP although recent FTP implementations rarely use the \texttt{PORT} command\cite{rfc2428}.

\subsection{Proxies}

The last middlebox that we cover is the proxy. A proxy is a middlebox that resides on a path and terminates TCP connections. A proxy can be explicit or transparent. The SOCKS5 protocol \cite{rfc1928} is an example of the utilization of an explicit proxy. SOCKS5 proxies are often used in enterprise networks to authenticate the establishment of TCP connections. A classical example of transparent proxies are the HTTP proxies that are deployed in various commercial networks to cache and speedup HTTP requests. In this case, some routers in the network are configured to intercept the TCP connections and redirect them to a proxy server \cite{McLaggan_WCCP:2012}. This redirection is transparent for the application, but from a transport viewpoint, the proxy acts as a relay between the two communicating hosts and there are two different TCP connections\footnote{Some deployments use several proxies in cascade. This allows the utilization of compression techniques and other non-standard TCP extensions on the connection between two proxies.}. The first one is initiated by the client and terminates at the proxy. The second one is initiated by the proxy and terminates at the server. The data exchanged over the first connection is passed to the second one, but the TCP options are not necessarily preserved. In some deployments, the proxy can use different options than the client and/or the server.

\subsection{How prevalent are middleboxes?}

We've learnt that middleboxes of various kinds exist, but are they really deployed in today's networks? Answering
this question can guide the right way to develop new protocols and enhance existing ones.

Here we briefly review measurement studies that have attempted to paint an accurate image of middlebox
deployments in use. The conclusion is that middleboxes are widespread to the point where 
end-to-end paths without middleboxes have become exceptions, rather than the norm.

There are two broad types of middlebox studies.
The first type of study uses ground-truth topology information from providers and other organizations.
While accurate, these studies may overestimate the influence of middleboxes on end-to-end traffic because
certain behavior is only triggered in rare, corner cases (e.g. an intrusion prevention system may only affect
traffic carrying known worm signatures). These studies do not tell us exactly \emph{what operations} are applied
to packets by middleboxes.
One recent survey study has found that the 57 enterprise networks surveyed deploy as many 
middleboxes as routers \cite{sherry2012making}.

Active measurements probe end-to-end paths to trigger ``known'' behaviors of middleboxes.
Such measurements are very accurate, pinpointing exactly what middleboxes do in certain scenarios.
However, they offer a lower bound of middlebox deployments, as they may pass through other 
middleboxes without triggering them. Additionally, such studies are limited to
probing a full path (cannot probe path segments) thus cannot tell how many middleboxes are deployed 
on a path exhibiting middlebox behavior. The data surveyed below comes from such studies.

Network Address Translators are easy to test for: the traffic source needs to compare its local source address
with the source address of its packets reaching an external site (e.g. what's my IP). When the addresses differ,
a NAT has been deployed on path. Using this basic technique, existing studies have shown that NATs are deployed almost
universally by (or for) end-users, be they home or mobile:
\begin{itemize}
\item Most cellular providers use them to cope with address space shortage and to provide some level of security. A study surveying 107
cellular networks across the globe found that 82 of them used NATs \cite{wang2011untold}.
\item Home users receive a single public IP address (perhaps via DHCP) from their access providers, and deploy NATs to support
multiple devices. The typical middlebox here is the ``wireless router'', an access point that aggregates all home traffic 
onto the access link. A study using peer-to-peer clients found that only 12\% of the peers have a public IP address \cite{nat-study}.
\end{itemize}

Testing for stateless firewalls is equally simple: the client generates traffic using different transport protocols and port
numbers, and the server acks it back. As long as some tests work, the client knows the endpoint is reachable, and that the failed
tests are most likely due to firewall behavior. There is a chance that the failed tests are due to the stochastic packet loss
inherent in the Internet; that is why tests are interleaved and run multiple times.

Firewalls are equally widespread in the Internet, being deployed by most cellular operators \cite{netpiculet}. Home routers often act as
firewalls, blocking new protocols and even existing ones (e.g. UDP) \cite{nat-study}. Most servers also deploy firewalls to restrict
in-bound traffic \cite{sherry2012making}. Additionally, many modern operating systems come with ``default-on'' firewalls. For instance, 
the Windows 7 firewall explicitly asks users to allow incoming connections and to whitelist applications allowed to make outgoing 
connections.

Testing for explicit proxies can be done by comparing the segments that leave the source with the ones arriving at the destination.
A proxy will change sequence numbers, perhaps segment packets differently, modify the receive window, and so forth. Volunteers 
from various parts of the globe ran TCPExposure, a tool that aims to detect such proxies and other middlebox behavior 
\cite{honda2011still}. The tests cover 142 access networks (including cellular, DSL, public hotspots and offices) in 
24 countries. Proxying behavior was seen on 10\% of the tested paths\cite{honda2011still}.

Beyond basic reachability, middleboxes such as traffic normalizers and stateful firewalls have some expectations on the packets they see: what exactly do these middleboxes do, and how widespread are they? The same study sheds some light into
this matter:
\begin{itemize}
\item Middlebox behavior depends on the ports used. Most middleboxes are active on port 80 (HTTP traffic).
\item A third of paths keep TCP flow state and use it to actively correct acknowledgments for data the middlebox has not
seen. To probe this behavior, TCPExposure sends a few TCP segments leaving a gap in the sequence number, while the server
acknowledgment also covers the gap. 
\item 14\% of paths remove unknown options from SYN packets. These paths will not allow TCP extensions to be deployed.
\item 18\% of paths modify sequence numbers; of these, a third seem to be proxies, and the other are most likely
firewalls than randomize the initial sequence number to protect vulnerable end hosts against in-window injection attacks.
\end{itemize}

\subsection{The Internet is Ossified}

Deploying a new IP protocol requires a lot of investment to change all the deployed hardware. Experience
with IPv6 after 15 years since it has been standardized is not encouraging: only a minute fraction of the Internet has
migrated to v6, and there are no signs of it becoming ``the Internet'' anytime soon.

Changing IPv4 itself is in theory possible with IP options. Unfortunately, it has been known for a while now
that ``IP options are not an option'' \cite{fonseca2005ip}. This is because existing routers implement forwarding in hardware 
for efficiency reasons; packets carrying unknown IP options are treated as exceptions that are processed in software.
To avoid a denial-of-service on routers' CPU's, such packets are dropped by most routers.

In a nutshell, we can't really touch IP - but have we also lost our ability to change transport protocols as well? 
The high level picture emerging from existing middlebox studies is that of a network that is highly tailored to today's traffic
to the point it is ossified: \emph{changing existing transport protocols is challenging as it needs to carefully consider
middlebox interactions}. Further, \emph{deploying new transport protocols natively is almost impossible}.

Luckily, changing transport protocols is still possible, albeit great care must be taken when doing so. 
Firewalls will block \emph{any} traffic they do not understand, so deploying new protocols
must necessarily use existing ones just to get through the network. This observation has recently lead to the development
of Minion, a container protocol that enables basic connectivity above TCP while avoiding its in-order, reliable bytestream
semantics at the expense of slightly increased bandwidth usage. We discuss Minion in Section \ref{section:minion}.

Even changing TCP is very difficult. The semantics of TCP are embedded in the network fabric, and new extensions
must function within the confines of these semantics, or they will fail. In section \ref{section:mptcp} we discuss Multipath TCP, 
another recent extension to TCP, that was designed explicitly to be middlebox compatible.

### 2013-07-22 Compiled/chapter5/intro.tex
all chapters compile independently

The first computer networks often used ad-hoc and proprietary
protocols to interconnect different hosts. During the 1970s and 1980s,
the architecture of many of these networks evolved towards a layered
architecture. The two most popular ones are the seven-layer OSI
reference model \cite{zimmermann1980osi} and the five-layer Internet
architecture \cite{clark1988design}. In these architectures, the
transport layer plays a key role. It enables applications to reliably exchange data.
A transport protocol can be characterized by the
service that it provides to the upper layer (usually the
application). Several transport services have been defined:

\begin{itemize}
\item a connectionless service
\item a connection-oriented bytestream service
\item a connection-oriented message-oriented service
\item a message-oriented request-response service
\item an unreliable delivery service for multimedia applications
\end{itemize}

The connectionless service is the simplest service that can be
provided by a transport layer protocol. The User Datagram Protocol
(UDP) \cite{rfc768} is an example of a protocol that provides this
service.

Over the years, 
the connection-oriented bytestream service has proven to be the 
transport layer service used by most applications. 
This service is currently provided by the
Transmission Control Protocol (TCP) \cite{rfc793} in the Internet.
TCP
is the dominant transport protocol in today's Internet, but other
protocols have provided similar services \cite{Iren_Transport:1999}.

Several transport protocols have been designed to support multimedia
applications. The Real-Time Transport protocol (RTP) \cite{rfc3550},
provides many features required by multimedia applications. Some of
the functions provided by RTP are part of the transport layer while
others correspond to the presentation layer of the OSI reference
model. The Datagram Congestion Control Protocol (DCCP) \cite{rfc4340}
is another protocol that provides functions suitable for applications
that do not require a fully reliable service.

The rest of this chapter is organized as follows. We first describe
the main services provided by the transport layer in
section~\ref{section:service}. This will enable us to look back at the
evolution of the reliable Internet transport protocols during the past 
decades. Section~\ref{section:today} then describes the organization
of today's Internet, the important role played by various types of
middleboxes, and the constraints that these middleboxes impose of the evolution of
transport protocols. Finally, we describe the design of 
two recent TCP extensions,
both of which
evolve the transport layer of the Internet
while remaining backward compatible with middleboxes.
Multipath TCP, described in
section~\ref{section:mptcp},
enables transmission of data segments within a transport connection
over multiple network paths.
Minion,
described in section~\ref{section:minion},
extends TCP and SSL/TLS~\cite{rfc5246} 
to provide richer services to the application---%
unordered message delivery and multi-streaming---%
without changing the protocols' wire-format.

\section{Providing the transport service}\label{section:service}
%==============================

As explained earlier, several services have been defined in the
transport layer. In this section, we first review the connectionless
service. Then we present in more detail how TCP and SCTP provide a
connection-oriented service and highlight the recent evolution of the
key functions of these protocols. This section concludes with a
discussion of the request-response service.

\subsection{Providing the connectionless service}
%---------------------------------------------

To provide a connectionless service, the transport layer mainly needs
to provide some multiplexing on top of the underlying network
layer. In UDP, this multiplexing is achieved by using port
numbers. The 8 byte UDP header contains the source and destination
port numbers that identify the applications that exchange data
on the two communicating hosts. In addition to these port numbers, the
UDP header contains a checksum that optionally covers the payload, the
UDP header and a part of the IP header. When UDP is used above IPv4, the
checksum is optional \cite{rfc791}. The sending application decides whether the UDP
payload will be protected by a checksum. If not, the checksum field is
set to zero. When UDP is used above IPv6, the checksum is mandatory
and cannot be disabled by the sender.

UDP has barely changed since the publication of \cite{rfc791}. The only significant modification has been the UDP-lite protocol \cite{rfc3828}. UDP-lite was designed for applications that could benefit from the delivery of possibly corrupted data. For this, UDP-lite allows the application to specify which part of the payload must be covered by a checksum. The UDP-lite header includes a checksum coverage field that indicates the part of the payload that is covered by the checksum.

\subsection{Providing the connection-oriented service}
%------------------------------------------------------

The connection-oriented service is both more complex and also more frequently used. TCP and SCTP are examples of current Internet protocols that provide this service. Older protocols like TP4 or XTP \cite{strayer1992xtp} also provide a connection-oriented service.

The connection-oriented service can be divided in three phases :
\begin{itemize}
\item the establishment of the connection
\item the data transfer
\item the release of the connection
\end{itemize}

\subsubsection{Connection establishment}

The first objective of the transport layer is to multiplex connections
initiated by different applications. This requires the ability to
unambiguously identify different connections on the same host. TCP
uses four fields that are present in the IP and TCP headers to uniquely
identify a connection:
\begin{itemize}
\item the source IP address
\item the destination IP address
\item the source port
\item the destination port
\end{itemize}

The source and destination addresses are the network layer addresses (e.g. IPv4 or IPv6 in the case of TCP) that have been allocated to the communicating hosts. When a connection is established by a client, the destination port is usually a well-known port number that is bound to the server application. On the other hand, the source port is often chosen randomly by the client \cite{rfc6056}. This random selection of the source port by the client has some security implications as discussed in \cite{Allman:2009:CSE:1517480.1517483}. Since a TCP connection is identified unambiguously by using this four-tuple, a client can establish multiple connections to the same server by using different source ports on each of these connections.

% three-way handshake

The classical way of establishing a connection between two transport
entities is the three-way handshake which is used by TCP
\cite{rfc793}. This three handshake was mainly designed to deal with host crashes. It assumes that the underlying network is able to guarantee that a packet will never remain inside the network for longer than the Maximum Segment Lifetime (MSL)\footnote{The default MSL duration is 2 minutes \cite{rfc793}.}. Furthermore, a host should not immediately reuse the same port number for subsequent connections to the same host. 
The TCP header contains flags that specify the role of
each segment. For example, the \texttt{ACK} flag indicates that the
segment contains a valid acknowledgment number while the \texttt{SYN}
flag is used during the three-way handshake. To establish a
connection, the original TCP specification \cite{rfc793} required the
client to send a TCP segment with the \texttt{SYN} flag sent, including an
initial sequence number extracted from a clock. According to \cite{rfc793},
this clock had to be implemented by using a 32-bit counter 
incremented at least once every 4 microseconds and after each TCP
connection establishment attempt. While this solution was sufficient
to deal with random host crashes, it was not acceptable from a
security viewpoint \cite{rfc6528}. When a clock is used to generate the initial
sequence number for each TCP connection, an attacker that wishes to
inject segments inside an established connection could easily guess
the sequence number to be used. To solve this problem, modern TCP
implementations generate a random initial sequence number
\cite{rfc6528}.

An example of the TCP three-way handshake is presented in figure~\ref{fig:tcp-establish}. The client sends a segment with the \texttt{SYN} flag set (also called a \texttt{SYN} segment). The server replies with a segment that has both the \texttt{SYN} and the \texttt{ACK} flags set (a \texttt{SYN+ACK} segment). This \texttt{SYN+ACK} segment contains contains a random sequence number chosen by the server and its acknowledgment number is set to the value of the initial sequence number chosen by the client incremented by one\footnote{TCP's acknowledgment number always contains the next expected sequence number and the \texttt{SYN} flag consumes one sequence number.}. The client replies to the \texttt{SYN+ACK} segment with an \texttt{ACK} segment that acknowledges the received \texttt{SYN+ACK} segment. This concludes the three-way handshake and the TCP connection is established.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/mptcp-ebook/Diapositive05.pdf}
\end{center}
\caption{TCP three-way handshake}\label{fig:tcp-establish}
\end{figure}

The duration of the three-way handshake is important for applications that exchange small amount of data such as requests for small web objects. It can become longer if losses occur because TCP can only rely on its retransmission timer to recover from the loss of a \texttt{SYN} or \texttt{SYN+ACK} segment. When a client sends a \texttt{SYN} segment to a server, it can only rely on the initial value of its retransmission timer to recover from losses\footnote{If the client has sent packets earlier to the same server, it might have stored some information from the previous connection \cite{rfc2140,Balakrishnan:1999:ICM:316188.316220} and use this information to bootstrap its initial timer. Recent Linux TCP/IP stacks preserve some state variables between connections.}. Most TCP/IP stacks have used an initial retransmission timer set to 3 seconds \cite{rfc1122}. This conservative value was chosen in the 1980s and confirmed in the early 2000s \cite{rfc2988}. However, this default implies that with many TCP/IP stacks, the loss of any of the first two segments of a three-way handshake will cause a delay of 3 seconds on a connection that may normally be shorter than that. Measurements conducted on large web farms showed that this initial timer had a severe impact on the performance perceived by the end users \cite{Chu_Tuning:2009}. This convinced the IETF to decrease the recommended initial value for the retransmission timer to 1 second \cite{rfc6298}. Some researchers proposed to decrease even more the value of the retransmission timer, notably in datacenter environments \cite{Vasudevan:2009:SEF:1592568.1592604}.

Another utilization of the three-way handshake is to negotiate options that are applicable for this connection. TCP was designed to be extensible. Although it does not carry a version field, in contrast with IP for example, TCP supports the utilization of options to both negotiate parameters and extend the protocol. 
TCP options in the \texttt{SYN} segment allow to negotiate the utilization of a particular TCP extension. To enable a particular extension, the client places the corresponding option inside the \texttt{SYN} segment. If the server replies with a similar option in the \texttt{SYN+ACK} segment, the extension is enabled. Otherwise, the extension is disabled on this particular connection. This is illustrated in figure~\ref{fig:tcp-establish}.

Each TCP option is encoded by using a Type-Length-Value (TLV) format, which enables a receiver to silently discard the options that it does not understand. Unfortunately, there is a limit to the maximum number of TCP options that can be placed inside the TCP header. This limit comes from the \texttt{Data Offset} field of the TCP header that indicates the position of the first byte of the payload measured as an integer number of four bytes word starting from the beginning of the TCP header. Since this field is encoded in four bits, the TCP header cannot be longer than 60 bytes, including all options. This size was considered to be large enough by the designers of the TCP protocol, but is becoming a severe limitation to the extensibility of TCP.

A last point to note about the three-way handshake is that the first TCP implementations created state upon reception of a \texttt{SYN} segment. Many of these implementations also used a small queue to store the TCP connections that had received a \texttt{SYN} segment but not yet the third \texttt{ACK}. For a normal TCP connection, the delay between the reception of a \texttt{SYN} segment and the reception of the third \texttt{ACK} is equivalent to a round-trip-time, usually much less than a second. For this reason, most early TCP/IP implementations chose a small fixed size for this queue. Once the queue was full, these implementations dropped all incoming \texttt{SYN} segments. This fixed-sized queue was exploited by attackers to cause denial of service attacks. They sent a stream of spoofed \texttt{SYN} segment\footnote{An IP packet is said to be spoofed if it contains a source address which is different from the IP address of the sending host. Several techniques can be used by network operators to prevent such attacks \cite{rfc2827}, but measurements show that they are not always deployed \cite{Beverly:2009:UED:1644893.1644936}.} to a server. Once the queue was full, the server stopped accepting \texttt{SYN} segments from legitimate clients \cite{rfc4987}. To solve this problem, recent TCP/IP stacks try to avoid maintaining state upon reception of a \texttt{SYN} segment. This solution is often called \emph{syn cookies}.

The principles behind \emph{syn cookies} are simple. To accept a TCP connection without maintaining state upon reception of the \texttt{SYN} segment, the server must be able to check the validity of the third \texttt{ACK} by using only the information stored inside this \texttt{ACK}. A simple way to do this is to compute the initial sequence number used by the server from a hash that includes the source and destination addresses and ports and some random secret known only by the server. The low order bits of this hash are then sent as the initial sequence number of the returned \texttt{SYN+ACK} segment. When the third \texttt{ACK} comes back, the server can check the validity of the acknowledgment number by recomputing its initial sequence number by using the same hash \cite{rfc4987}. Recent TCP/IP stacks use more complex techniques to deal notably with the options that are placed inside the \texttt{SYN} and need to be recovered from the information contained in the third \texttt{ACK} that usually does not contain any option.

At this stage, it is interesting to look at the connection establishment scheme used by the SCTP protocol \cite{rfc4960}. SCTP was designed more than two decades after TCP and thus has benefited from several of the lessons learned from the experience with TCP. A first difference between TCP and SCTP are the segments that these protocols use. The SCTP header format is both simpler and more extensible than the TCP header.

The first four fields of the SCTP header (Source and Destination ports, Verification tag and Checksum) are present in all SCTP segments. The source and destination ports play the same role as in TCP. The verification tag is a random number chosen when the SCTP connection is created and placed in all subsequent segments. This verification tag is used to prevent some forms of packet spoofing attacks \cite{rfc4960}. This is an improvement compared to TCP where the validation of a received segment must be performed by checking the sequence numbers, acknowledgment numbers and other fields of the header \cite{Gont_TCPSecurity:2012}. The SCTP checksum is a 32 bits CRC that provides stronger error detection properties than the Internet checksum used by TCP \cite{Stone:2000:CTC:347059.347561}. Each SCTP segment can contain a variable number of chunks and there is no apriori limit to the number of chunks that appear inside a segment, except that a segment should not be longer than the maximum packet length of the underlying network layer.

The SCTP connection establishment uses several of these chunks to specify the values of some parameters that are exchanged. A detailed discussion of all these chunks is outside the scope of this document and may be found in \cite{rfc4960}. The SCTP four-way handshake uses four segments as shown in figure~\ref{fig:sctp-4way}. The first segment contains the \texttt{INIT} chunk. To establish an SCTP connection with a server, the client first creates some local state for this connection. The most important parameter of the \texttt{INIT} chunk is the \emph{Initiation tag}. This value is a random number that is used to identify the connection on the client host for its entire lifetime. This \emph{Initiation tag} is placed as the \emph{Verification tag} in all segments sent by the server. This is an important change compared to TCP where only the source and destination ports are used to identify a given connection. The \texttt{INIT} chunk may also contain the other addresses owned by the client. The server responds by sending an \texttt{INIT-ACK} chunk. This chunk also contains an \emph{Initiation tag} chosen by the server and a copy of the \emph{Initiation tag} chosen by the client. The \texttt{INIT} and \texttt{INIT-ACK} chunks also contain an initial sequence number. A key difference between TCP's three-way handshake and SCTP's four-way handshake is that an SCTP server does not create any state when receiving an \texttt{INIT} chunk. For this, the server places inside the \texttt{INIT-ACK} reply a \emph{State cookie} chunk. This \emph{State cookie} is an opaque block of data that contains information from the \texttt{INIT} and \texttt{INIT-ACK} chunks that the server would have had stored locally, some lifetime information and a signature. The format of the \emph{State cookie} is flexible and the server could in theory place almost any information inside this chunk. The only requirement is that the \emph{State cookie}  must be echoed back by the client to confirm the establishment of the connection. Upon reception of the \texttt{COOKIE-ECHO} chunk, the server verifies the signature of the \emph{State cookie}. The client may provide some user data and an initial sequence number inside the \texttt{COOKIE-ECHO} chunk. The server then responds with a \texttt{COOKIE-ACK} chunk that acknowledges the \texttt{COOKIE-ECHO} chunk. The SCTP connection between the client and the server is now established. This four-way handshake is both more secure and more flexible than the three-way handshake used by TCP.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/mptcp-ebook/Diapositive06.pdf}
\end{center}
\caption{The four-way handshake used by SCTP}\label{fig:sctp-4way}
\end{figure}

\subsubsection{Data transfer}

Before looking at the techniques that are used by transport protocols to transfer data, it is useful to look at their service models. TCP has the simplest service model. Once a TCP connection has been established, two bytestreams are available. The first bytestream allows the client to send data to the server and the second bytestream provides data transfer in the opposite direction. TCP guarantees the reliable delivery of the data during the lifetime of the TCP connection provided that it is gracefully released.

SCTP provides a slightly different service model \cite{rfc3286}. Once an SCTP connection has been established, the communicating hosts can access two or more message streams. A message stream is a stream of variable length messages. Each message is composed of an integer number of bytes. The connection-oriented service provided by SCTP preserves the message boundaries. This implies that if an application sends a message of \emph{N} bytes, the receiving application will also receive it as a single message of \emph{N} bytes. This is in contrast with TCP that only supports a bytestream. Furthermore, SCTP allows the applications to use multiple streams to exchange data. The number of streams that are supported on a given connection is negotiated during connection establishment. When multiple streams have been negotiated, each application can send data over any of these streams and SCTP will deliver the data from the different streams independently without any head-of-line blocking. 
%TCP provides some support for urgent data which could be considered as a restricted form of out-of-band capability. However, this part of the TCP specification has not been implemented in the same way on all TCP stacks \cite{rfc6093}.

While most usages of SCTP may assume an in-order delivery of the
data, SCTP supports unordered delivery of messages at the receiver. 
Another extension to SCTP \cite{rfc3758} supports partially-reliable delivery. With this
extension,  an SCTP sender can be instructed to ``expire'' data based on 
one of several events,
such as a timeout,
the sender can signal the SCTP receiver to move on without waiting for 
the ``expired'' data.
%a receiver may accept gaps inside the received data and
%still acknowledge the corresponding segments to the sender. 
This
partially reliable service could be useful to provide timed delivery
for example. With this service, there is an upper limit on the time
required to deliver a message to the receiver. If the
transport layer cannot deliver the data within the specified delay,
the data is discarded by the sender without causing any stall in the stream.

To provide a reliable delivery of the data, transport protocols rely on various mechanisms that have been well studied and discussed in the literature : sequence numbers, acknowledgments, windows, checksums and retransmission techniques. A detailed explanation of these techniques may be found in standard textbooks \cite{Bonaventure_CNP3:2012,ross2003computer,Fall_TCP:2011}. We assume that the reader is familiar with them and discuss only some recent changes.

TCP tries to pack as much data as possible inside each segment \cite{rfc793}. Recent TCP stacks combine this technique with Path MTU discovery to detect the MTU to be used over a given path \cite{rfc4821}. SCTP uses a more complex but also more flexible strategy to build its segments. It also relies on Path MTU Discovery to detect the MTU on each path. SCTP then places various chunks inside each segment. The control chunks, that are required for the correct operation of the protocol, are placed first. Data chunks are then added. SCTP can split a message in several chunks before transmission and also the bundling of different data chunks inside the same segment.

% packet nagle in TCP, bundling in SCTP for lagre and small

Acknowledgments allow the receiver to inform the sender of the correct reception of data. TCP initially relied exclusively on cumulative acknowledgments. Each TCP segment contains an acknowledgment number that indicates the next sequence number that is expected by the receiver. Selective acknowledgments were added later as an extension to TCP \cite{rfc2018}. A selective acknowledgment can be sent by a receiver when there are gaps in the received data. A selective acknowledgment is simply a sequence of pairs of sequence numbers, each pair indicating the beginning and the end of a received block of data. SCTP also supports cumulative and selective acknowledgments. Selective acknowledgments are an integral part of SCTP and not an extension which is negotiated at the beginning of the connection. In SCTP, selective acknowledgments are encoded as a control chunk that may be placed inside any segment. In TCP, selective acknowledgments are encoded as TCP options. Unfortunately, given the utilization of the TCP options (notably the timestamp option \cite{rfc1323}) and the limited space for options inside a TCP segment, a TCP segment cannot report more than three blocks of data. This adds some complexity to the handling and utilization of selective acknowledgments by TCP.

Current TCP and SCTP stacks try to detect segment losses as quickly as possible. For this, they implement various heuristics that allow to retransmit a segment once several duplicate acknowledgments have been received \cite{Fall_TCP:2011}. Selective acknowledgment also aid to improve the retransmission heuristics. If these heuristics fail, both protocols rely on a retransmission timer whose value is fixed in function of the round-trip time measured over the connection \cite{rfc6298}.

Last but not least, the transport protocols on the Internet perform congestion control. The original TCP congestion control scheme was proposed in \cite{Jacobson_congestion:88}. Since then, it has evolved and various congestion control schemes have been proposed. Although the IETF recommends a single congestion control scheme \cite{rfc5681}, recent TCP stacks support different congestion control schemes and some allow the user to select the preferred one. A detailed discussion of the TCP congestion control schemes may be found in \cite{Afanasyev_Congestion:2012}. SCTP's congestion control scheme is largely similar to TCP's congestion control scheme.

Additional details about recent advances in SCTP may be found in \cite{Budzisz_Taxonomy:2012}. \cite{rfc4614} lists recent IETF documents that are relevant for TCP. \cite{Fall_TCP:2011} contains a detailed explanation of some of the recent changes to the TCP/IP protocol stack.

\subsubsection{Connection release}

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/mptcp-ebook/Diapositive08.pdf}
\end{center}
\caption{The four-way handshake used to close a TCP connection}\label{fig:tcp-fin}
\end{figure}

This phase occurs when either both hosts have exchanged all the required data or when one host needs to stop the connection for any reason (application request, lack of resources, \ldots). TCP supports two mechanisms to release a connection. The main one is the four-way handshake. This handshake uses the \texttt{FIN} flag in the TCP header. Each host can release its own direction of data transfer. When the application wishes to gracefully close a connection, it requests the TCP entity to send a \texttt{FIN} segment. This segment marks the end of the data transfer in the outgoing direction and the sequence number that corresponds to the \texttt{FIN} flag (which consumes one sequence number) is the last one to be sent over this connection. The outgoing stream is closed as soon as the sequence number corresponding to the \texttt{FIN} flag is acknowledged. The remote TCP entity can use the same technique to close the other direction \cite{rfc793}. This graceful connection release has one advantage and one drawback. On the positive side, TCP provides a reliable delivery of all the data provided that the connection is gracefully closed. On the negative side, the utilization of the graceful release forces the TCP entity that sent the last segment on a given connection to maintain state for some time. On busy servers, a large number of connections can remain for a long time \cite{Faber_TimeWait:1999}. To avoid maintaining such state after a connection has been closed, web servers and some browsers send a \texttt{RST} segment to abruptly close TCP connections. In this case, the underlying TCP connection is closed once all the data has been transferred. This is faster, but there is no guarantee about the reliable delivery of the data.

SCTP uses a different approach to terminante connections. When an application requests a shutdown of a connection, SCTP performs a three-way handshake. This handshake uses the \texttt{SHUTDOWN}, \texttt{SHUTDOWN-ACK} and \texttt{SHUTDOWN-COMPLETE} chunks. The \texttt{SHUTDOWN} chunk is sent once all outgoing data has been acknowledged. It contains the last cumulative sequence number. Upon reception of a \texttt{SHUTDOWN} chunk, an SCTP entity informs its application that it cannot accept anymore data over this connection. It then ensures that all outstanding data have been delivered correctly. At that point, it sends a \texttt{SHUTDOWN-ACK} to confirm the reception of the \texttt{SHUTDOWN} segment. The three-way handshake completes with the transmission of the \texttt{SHUTDOWN-COMPLETE} chunk \cite{rfc4960}.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/mptcp-ebook/Diapositive11.pdf}
\end{center}
\caption{The three-way handshake used to close an SCTP connection}\label{fig:sctp-close}
\end{figure}

SCTP also provides the equivalent to TCP's \texttt{RST} segment. The \texttt{ABORT} chunk can be used to refuse a connection, react to the reception of an invalid segment or immediately close a connection (e.g. due to lack of resources).

\subsection{Providing the request-response service}
%---------------------------------------------------

The request-response service has been a popular service since the 1980s.  At that time, many request-response applications were built above the connectionless service, typically UDP \cite{Birrell_RPC:1984}. A request-response application is very simple. The client sends a request to a server and blocks waiting for the response. The server processes the request and returns a response to the client. This paradigm is often called Remote Procedure Call (RPC) since often the client calls a procedure running on the server.

The first implementations of RPC relied almost exclusively on UDP to transmit the request and responses. In this case, the size of the requests and responses was often restricted to one MTU. In the 1980s and the beginning of the 1990s, UDP was a suitable protocol to transport RPCs because they were mainly used in Ethernet LANs. Few users were considering the utilization of RPC over the WAN. In such networks, CSMA/CD regulated the access to the LAN and there were almost no losses. Over the years, the introduction of Ethernet switches has both allowed Ethernet networks to grow in size but also implied a growing number of packet losses. Unfortunately, RPC running over UDP does not deal efficiently with packet losses because many implementation use large timeouts to recover for packet losses. TCP could deal with losses, but it was considered to be too costly for request-response applications. Before sending a request, the client must first initiate the connection. This requires a three-way handshake and thus ``wastes'' one round-trip-time. Then, TCP can transfer the request and receive the response over the established connection. Eventually, it performs a graceful shutdown of the connection. This connection release requires the exchange of four (small) segments, but also forces the client to remain in the \texttt{TIME\_WAIT} state for a duration of 240 seconds, which limits the number of connections (and thus RPCs) that it can establish with a given server.

TCP for Transactions or T/TCP \cite{rfc1644} was a first attempt to enable TCP to better support request/response applications. T/TCP solved the above problem by using three TCP options. These options were mainly used to allow each host to maintain an additional state variable, Connection Count (CC) that is incremented by one for every connection. This state variable is sent in the \texttt{SYN} segment and cached by the server. If a \texttt{SYN} received from a client contains a CC that is larger than the cached one, the new connection is immediately established and data can be exchanged directly (already in the \texttt{SYN}). Otherwise, a normal three-way handshake is used. The use of this state variable allowed T/TCP to reduce the duration of the \texttt{TIME\_WAIT} state. T/TCP used \texttt{SYN} and \texttt{FIN} flags in the segment sent by the client and returned by the server, which led to a two segment connection, the best solution from a delay viewpoint for RPC applications. Unfortunately, T/TCP was vulnerable to spoofing attacks \cite{deVivo_TTCP:1999}. An attacker could observe the Connection Count by capturing packets. Since the server only checked that the value of the CC state variable contained in a \texttt{SYN} segment was higher than the cached one, it was easy to inject new segments. Due to this security problem, T/TCP is now deprecated.

Improving the performance of TCP for request/response applications continued to be an objective for many researchers. However, recently the focus of the optimizations moved from the LANs that were typical for RPC applications to the global Internet. The motivation for several of the recent changes to the TCP protocol was the perceived performance of TCP with web search applications \cite{Chu_Tuning:2009}. A typical web search is also a very short TCP connection during which a small HTTP request and a small HTTP response are exchanged. A first change to TCP was the increase of the initial congestion window \cite{Chu_increasing:2013}. For many years, TCP used an initial window between 2 and 4 segments \cite{rfc3390}. This was smaller than the typical HTTP response from a web search engine \cite{Chu_Tuning:2009}. Recent TCP stacks use an initial congestion window of 10 segments \cite{Chu_increasing:2013}.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/mptcp-ebook/Diapositive12.pdf}
\end{center}
\caption{TCP fast open}\label{fig:tcp-fastopen}
\end{figure}

Another change that has been motivated by web search applications is the TCP Fast Open (TFO) extension \cite{radhakrishnan2011tcp}. This extension can be considered as a replacement for T/TCP. TCP fast open also enables a client to send data inside the \texttt{SYN} segment. TCP fast open relies on state sharing between the client and the server, but the state is more secure than the simple counter used by T/TCP. To enable the utilization of TCP fast open, the client must first obtain a \emph{cookie} from the server. This is done by sending a \texttt{SYN} segment with the TFO cookie request option. The server then generates a secure cookie by encrypting the IP address of the client with a local secret \cite{radhakrishnan2011tcp}. The encrypted information is returned inside a TFO cookie option in the \texttt{SYN+ACK} segment. The client caches the cookie and associates it with the server's IP address. The subsequent connections initiated by the client will benefit from TCP fast open. The client includes the cached cookie and optional data inside its \texttt{SYN} segment. The server can validate the segment by decrypting its cookie. If the cookie is valid, the server acknowledges the \texttt{SYN} and the data that it contains. Otherwise, the optional data is ignored and a normal TCP three-way handshake is used. This is illustrated in figure~\ref{fig:tcp-fastopen}.

### 2013-07-22 Compiled/chapter5/main.tex
all chapters compile independently

\newcommand{\CLASSINPUTinnersidemargin}{0.7in}
\newcommand{\CLASSINPUToutersidemargin}{0.7in}

\documentclass{article}

\usepackage{graphicx}
\usepackage{multirow}
\usepackage{array}
\usepackage{fancyvrb}
\usepackage{boxedminipage}
\usepackage{color}
\usepackage{paralist}
\usepackage{algorithmic}
\usepackage{slashbox}
\usepackage{xspace}
\usepackage{url}
\usepackage{subfig}
\usepackage{amsthm}
\newcommand{\ed}[1]{\textsf{\textcolor{red}{[#1]}}}
\newcommand{\remove}[1]{}
\usepackage{epstopdf}
\epstopdfsetup{update}

\begin{document}

\title{Recent advances in reliable transport protocols\footnote{
Parts of the text in this chapter have appeared in the following
 publications by the same author(s): \cite{login}, \cite{mptcp-cc},
 \cite{raiciu2012hard}, ~\cite{nowlan12fitting} 
and~\cite{iyengar10minion}.}
}

\author{}

\maketitle

\begin{abstract}
Transport protocols play a critical role in today's Internet. This
chapter first looks at the recent of the reliable transport
protocols. It then explains the growing impact of middleboxes on the
evolvability of these protocols. 
Two recent protocol extensions,
Multipath TCP and Minion,
which were both designed to extend the current Transport Layer 
in the Internet
are then described.
\end{abstract}

\section{Introduction}
%%%%%%%%%%%%%%%%%%

\input{intro.tex}

\section{Today's Internet}\label{section:today}
%%%%%%%%%%%%%%%%%%%%%

\input{internet.tex}

%\section{Structuring the transport layer}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\input{structure.tex}

\section{Multipath TCP}\label{section:mptcp}

%%%%%%%%%%%%%%%%%%%%

\input{mptcp.tex}

\section{Minion}\label{section:minion}
%%%%%%%%%%%%%

\input{minion.tex}

%\section{A glimpse into future transport protocols}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\input{future.tex}

\section{Conclusion}\label{section:conclusion}
%%%%%%%%%%%%%%%%%

\input{conclusion.tex}

\bibliographystyle{plain}
\bibliography{ebook,rfc}

\appendix

\section{Exercises}
%%%%%%%%%%%%%%%%

\input{exercises.tex}

\end{document}

### 2013-07-22 Compiled/chapter5/minion.tex
all chapters compile independently

\def\utcp{$u$TCP\xspace}
\def\utls{$u$TLS\xspace}
\def\ucobs{$u$COBS\xspace}

As the Internet has grown and evolved
over the past few decades,
congestion control algorithms and extensions such as MPTCP
continue to reflect an evolving TCP.
However,
a proliferation of middleboxes
such as 
Network Address Translators (NATs), 
Firewalls,
and 
Performance Enhancing Proxies (PEPs),
has arguably stretched the waist of the Internet hourglass
upwards from IP
to include
TCP and UDP~\cite{rosenberg08udp, ford08breaking, popa10http},
making it increasingly difficult
to deploy new transports
and to
use anything but TCP and UDP 
on the Internet.

TCP~\cite{rfc793} was originally designed to offer applications
a convenient, high-level communication abstraction
with semantics emulating Unix file I/O or pipes:
a reliable, ordered bytestream,
through an end-to-end channel (or connection).
As the Internet has evolved, however,
applications needed better abstractions
from the transport.
We start this section by examining
how TCP's role in the network has evolved
from a communication {\em abstraction} to a communication {\em substrate},
why its in-order delivery model makes TCP a poor substrate,
why other OS-level transports have failed to replace TCP in this role,
and why UDP is inadequate as the only alternative substrate to TCP.

\subsection{Rise of Application-Level Transports}

The transport layer's traditional role in a network stack
is to build high-level communication abstractions
convenient to applications,
atop the network layer's basic packet delivery service.
TCP's reliable, stream-oriented design~\cite{rfc793}
exemplified this principle,
by offering an inter-host communication abstraction
modeled on Unix pipes,
which were the standard {\em intra-host} communication abstraction
at the time of TCP's design.
The Unix tradition of
implementing TCP in the OS kernel
offered further convenience,
allowing much application code to ignore the difference between
an open disk file, an intra-host pipe, or an inter-host TCP socket.

\begin{figure}[tbp]
\centering
\includegraphics[width=0.8\textwidth]{figures/layers.eps}
\caption{Today's ``{\em de facto} transport layer''
	is effectively split between OS and application code. (taken from
    \cite{nowlan12fitting})}
\label{f:layers}
\end{figure}

Instead of building directly atop
traditional OS-level transports such as TCP or UDP, however,
today's applications
frequently introduce additional transport-like
protocol layers at user-level,
typically implemented via application-linked libraries.
Examples include the ubiquitous SSL/TLS~\cite{rfc5246},
media transports such as RTP~\cite{rfc3550},
and experimental multi-streaming transports such as
SST~\cite{ford07structured}, SPDY~\cite{spdy},
and \O{}MQ~\cite{zeromq}.
Applications increasingly use HTTP or HTTPS over TCP
as a substrate~\cite{popa10http};
this is also illustrated by 
the W3C's WebSocket interface~\cite{websocket},
which offers general bidirectional communication
between browser-based applications and Web servers atop HTTP and HTTPS.

In this increasingly common design pattern,
the ``transport layer'' as a whole has in effect become
a stack of protocols straddling the OS-application boundary.
Figure~\ref{f:layers} illustrates one example stack,
representing Google's experimental Chrome browser,
which inserts SPDY for multi-streaming and TLS for security
at application level,
atop the OS-level TCP.

One can debate whether a given application-level protocol
fits some definition of ``transport'' functionality.
The important point, however,
is that {\em today's applications no longer need, or expect,
the underlying OS to provide ``convenient'' communication abstractions:
an application simply links in libraries, frameworks,
or middleware offering the abstractions it desires.}
What today's applications
need from the OS is not convenience,
but {\em an efficient substrate}
atop which application-level libraries
can build the desired abstractions.

\subsection{TCP's Latency Tax}

While TCP has proven to be a popular
substrate for application-level transports,
using TCP in this role
converts its delivery model from a blessing into a curse.
Application-level transports are just as capable as the kernel
of sequencing and reassembling packets into
a logical data unit or ``frame''~\cite{clark90architectural}.
By delaying any segment's delivery to the application
until all prior segments are received and delivered, however,
TCP imposes a ``latency tax'' on all segments
arriving within one round-trip time (RTT) after any single lost segment.

This latency tax is a fundamental byproduct
of TCP's in-order delivery model,
and is irreducible,
in that an application-level transport
cannot ``claw back''
the time a potentially useful segment
has wasted in TCP's buffers.
The best the application can do is
simply to {\em expect} higher latencies to be common.
A conferencing application can use a longer jitter buffer,
for example,
at the cost of increasing user-perceptible lag.
Network hardware advances are unlikely to address this issue, 
since TCP's latency tax depends on RTT,
which is lower-bounded by the speed of light
for long-distance communications.

\subsection{Alternative OS-level Transports}
\label{sec:motiv-alts}

All standardized OS-level transports since TCP, including
	UDP~\cite{rfc768}, RDP~\cite{rfc908},
	DCCP~\cite{rfc4340}, and SCTP~\cite{rfc4960},
support out-of-order delivery.
The Internet's evolution has created strong barriers
against the widespread deployment of new transports
other than the original TCP and UDP, however.
These barriers are detailed elsewhere~\cite{
	rosenberg08udp,ford08breaking,popa10http},
but we summarize two key issues here.

First, 
adding or enhancing a ``native'' transport built atop IP
involves modifying popular OSes,
effectively increasing the bar for widespread deployment
and making it 
more difficult to evolve transport functionality
below the red line representing the OS API
in Figure~\ref{f:layers}.
Second, the Internet's original ``dumb network'' design,
in which routers that ``see'' only up to the IP layer,
has evolved into a ``smart network''
in which pervasive middleboxes
perform deep packet inspection and interposition
in transport and higher layers.
Firewalls tend to block ``anything unfamiliar'' for security reasons,
and Network Address Translators (NATs)
rewrite the port number in the  transport header,
making both incapable of allowing traffic from a new transport
without explicit support for that transport.
Any packet content
not protected by end-to-end security such as TLS---%
the yellow line in Figure~\ref{f:layers}---%
has become ``fair game''
for middleboxes to inspect and interpose on~\cite{
	reis08detecting},
making it more difficult to evolve transport functionality
anywhere below that line.

\subsection{Why Not UDP?}

As the only widely-supported transport
with out-of-order delivery,
UDP offers a natural
substrate for application-level transports.
Even applications otherwise well-suited to UDP's delivery model
often favor TCP as a substrate, however.

A recent study found over 70\% of streaming media using TCP~\cite{
	guo06delving},
and even latency-sensitive conferencing applications
such as Skype
often use TCP~\cite{baset06analysis}.
Firewalls can monitor the TCP state machine
to help thwart network attacks~\cite{handley01network},
but determining an application session's state atop UDP
requires application-specific analysis~\cite{northcutt05inside},
creating an incentive for firewalls to block UDP in general.

In general,
network middleboxes support UDP widely but not {\em universally}.
For this reason,
latency-sensitive applications
seeking maximal connectivity ``in the wild''
often fall back to TCP when UDP connectivity fails.
Skype~\cite{baset06analysis}
and Microsoft's Direct\-Access VPN~\cite{davies09directaccess},
for example,
support UDP but can masquerade
as HTTP or HTTPS streams atop TCP when required for connectivity.

TCP can offer performance advantages over UDP as well.
For applications requiring congestion control,
an OS-level implementation in TCP
may be more timing-accurate
than an application-level implementation in a UDP-based protocol,
because the OS kernel
can avoid the timing artifacts of 
system calls and process scheduling~\cite{zec02estimating}.
Hardware TCP offload engines
can optimize common-case efficiency
in end hosts~\cite{mogul03tcp},
and performance enhancing proxies
can optimize TCP throughput across diverse networks~\cite{
	rfc3234,cisco-rbscp}.
Since middleboxes can track TCP's state machine,
they impose much longer idle timeouts on open TCP connections---%
nominally two hours~\cite{rfc5382}---%
whereas UDP-based applications must send keepalives
every two minutes to keep an idle connection open~\cite{rfc4787},
draining power on mobile devices.

\subsection{Breaking Out Of the Transport Logjam}
Applications and application developers care most about 
services that the networking infrastructure offers to them
and not how packets look on the wire;
that is, they care about 
new transport {\em services}, not new transport {\em protocols}.
On the other hand,
middleboxes care most about how packets look on the wire,
and generally do not care about what services are offered to the applications;
that is, changing the transport protocol's bits on the wire
will require changing middleboxes to respond to these changes as well.

For application developers,
TCP versus UDP represents an ``all-or-nothing'' choice
on the spectrum of services applications need.
Applications desiring some but not all of TCP's services,
such as congestion control but unordered delivery,
must reimplement and tune all other services atop UDP
or suffer TCP's performance penalties.

Without dismissing UDP's usefulness 
as a truly ``least-common-denominator'' substrate,
we believe the factors above
suggest that TCP will also remain a popular substrate---%
even for latency-sensitive applications
that can benefit from out-of-order delivery---%
and that a deployable, backward-compatible workaround
to TCP's latency tax
can significantly benefit such applications.

Recognizing that the development of application-level transports
and the use of TCP as a substrate under them
is likely to continue and expand,
we now describe {\em Minion},
an architecture for efficient but backward-compatible
unordered delivery over extant transports, including TCP.
Minion consists of
{\em \utcp},
a small OS extension
adding basic unordered delivery primitives to TCP,
and two application-level protocols
implementing datagram-oriented delivery services
that function on either \utcp or unmodified TCP stacks.

While building a new transport on UDP 
or using alternative OS-level transports where available
are both perfectly reasonable design approaches,
Minion offers a uniform interface for applications to use
along with an alternative option
where the ubiquitous TCP protocol is adequate, but the sockets API is not;
where a new transport service
can be offered using TCP's bits on the wire.

\subsection{Minion Architecture Overview}

Minion is an architecture and protocol suite
designed to provide efficient unordered delivery built atop
existing transports.
Minion itself offers no high-level abstractions:
its goal is to serve
applications and higher application-level transports,
by acting as a ``packhorse''
carrying raw datagrams as reliably and efficiently as possible
across today's diverse and change-averse Internet.

\begin{figure}[tbp]
\centering
\includegraphics[width=0.6\textwidth]{figures/minionarch.eps}
\caption{Minion architecture (taken from~\cite{nowlan12fitting})}
\label{f:minionarch}
\end{figure}

Figure~\ref{f:minionarch} illustrates Minion's architecture.
Applications and higher application-level transports
link in and use Minion in the same way as they already use
existing application-level transports
such as DTLS~\cite{rfc4347},
the datagram-oriented analog of SSL/TLS~\cite{rfc5246}.
In contrast with DTLS's goal of layering security
atop datagram transports such as UDP or DCCP,
Minion's goal is to offer efficient datagram delivery
atop {\em any} available OS-level substrate, including TCP.

Minion consists of
several application-level transport protocols,
%in the spirit of Section~\ref{sec:motiv},
together with a set of optional enhancements
to end hosts' OS-level TCP implementations.

Minion's enhanced OS-level TCP stack,
called {\em \utcp} (``unordered TCP''),
includes sender- and receiver-side
API features supporting unordered delivery and prioritization,
as detailed later in this section.
These enhancements affect only the OS API
through which application-level transports such as Minion
interact with the TCP stack,
and make {\em no} changes to TCP's wire protocol.

Minion's application-level protocol suite currently consists
of the following main components:
\begin{itemize}
\item	{\bf \ucobs} is a protocol
	that implements a minimal unordered datagram delivery service
	atop either unmodified TCP or \utcp,
	using 
    {\em Consistent-Overhead Byte Stuffing}, or
    COBS encoding~\cite{cheshire97consistent}
	to facilitate out-of-order datagram delimiting
	and prioritized delivery,
	as described later in Section~\ref{sec:dgrams_utcp}.
\item	{\bf \utls} is a modification of
	the traditionally stream-oriented TLS~\cite{rfc5246},
	offering a secure, unordered datagram delivery service
	atop TCP or \utcp.
	The wire-encoding of \utls streams
	is designed to be indistinguishable in the network
	from conventional, encrypted TLS-over-TCP streams (e.g., HTTPS),
	offering a maximally conservative design point
	that makes no network-visible changes
	``below the yellow line''
	in Figure~\ref{f:minionarch}.
	Section~\ref{sec:dgrams_utcp} describes \utls.
\item	Minion adds shim layers
	atop OS-level datagram transports, such as UDP and DCCP,
	to offer applications a consistent API
	for unordered delivery across multiple OS-level transports.
	Since these shims are merely wrappers
	for OS transports already offering unordered delivery,
	this paper does not discuss them in detail.
\end{itemize}

Minion currently leaves to the application
the decision of {\em which} protocol to use for a given connection:
e.g., \ucobs or \utls atop TCP/\utcp,
or OS-level UDP or DCCP via Minion's shims.
Other ongoing work explores {\em negotiation protocols}
to explore the protocol configuration space dynamically,
optimizing protocol selection and configuration
for the application's needs and the network's 
constraints~\cite{ford09efficient}.
Many applications already incorporate simple negotiation schemes,
however---%
e.g., attempting a UDP connection first
and falling back to TCP if that fails---%
and adapting these mechanisms
to engage Minion's protocols according to
application-defined preferences and decision criteria
should be straightforward.

\subsection{Compatibility and Deployability}

Minion addresses the key barriers to transport evolution,
outlined in Section~\ref{sec:motiv-alts},
by creating a backward-compatible, incrementally deployable substrate
for new application-layer transports desiring unordered delivery.
Minion's deployability rests on the fact that it can,
when necessary,
avoid relying on changes either
``below the red line'' in the end hosts
(the OS API in Figure~\ref{f:minionarch}),
or
``below the yellow line'' in the network
(the end-to-end security layer in Figure~\ref{f:minionarch}).

While Minion's \ucobs and \utls protocols
offer maximum performance benefits from out-of-order delivery
when both endpoints include OS support for Minion's \utcp enhancements,
\ucobs and \utls still function and interoperate correctly
even if neither endpoint supports \utcp,
and the application need not know or care
whether the underlying OS supports \utcp.
If only one endpoint OS supports \utcp,
Minion still offers incremental performance benefits,
since \utcp's sender-side and receiver-side enhancements are independent.
A \ucobs or \utls connection atop a mixed TCP/\linebreak[0]\utcp endpoint-pair
benefits from \utcp's sender-side enhancements
for datagrams sent by the \utcp endpoint,
and the connection benefits from \utcp's receiver-side enhancements
for datagrams arriving at the \utcp host.

Addressing the challenge of network-compatibility
with middleboxes that filter new OS-level transports
and sometimes UDP,
Minion offers application-level transports
a continuum of substrates
representing different tradeoffs
between suitability to the application's needs
and compatibility with the network.

An application can use unordered OS-level transports
such as UDP, DCCP~\cite{rfc4340}, or SCTP~\cite{rfc4960},
for paths on which they operate,
but Minion offers an unordered delivery alternative
wire-compatible not only with TCP,
but with the ubiquitous TLS-over-TCP streams
on which HTTPS (and hence Web security and E-commerce) are based,
likely to operate in almost any network environment
purporting to offer ``Internet access.''

\subsection{Minion's Unordered Delivery using TCP and TLS}
We now briefly discuss Minion's true unordered delivery
over bytestream protocols;
we describe how Minion provides true unordered datagram
delivery without modifying the TCP and TLS wire-formats.

Minion enhances the OS's TCP stack
with API enhancements supporting unordered delivery
in both TCP's send and receive paths,
enabling applications to reduce transmission latency
at both the sender- and receiver-side end hosts
when both endpoints support \utcp.
Since \utcp makes no change to TCP's wire protocol,
two endpoints need not ``agree'' on whether to use \utcp:
one endpoint gains latency benefits from \utcp
even if the other endpoint does not support it.
Further, an OS may choose independently
whether to support the sender- and receiver-side enhancements,
and when available, applications can activate them independently.

\utcp does {\em not} seek to offer
``convenient'' or ``clean'' unordered delivery abstractions
directly at the OS API.
Instead, \utcp's design is motivated by the goals of
maintaining exact compatibility
with TCP's existing wire-visible protocol and behavior,
and facilitating deployability
by minimizing the extent and complexity
of changes to the OS's TCP stack.

\subsubsection{\utcp: Receiver-Side Modifications}

A conventional TCP receiver delivers data in-order 
to the receiving application,
holding back any data that is received out of order.
\utcp modifies the TCP receive path,
enabling a receiving application
to request immediate delivery of data 
that is received by \utcp,
both in order and out of order.

\utcp makes two modifications to a conventional TCP receiver.
First, whereas a conventional TCP stack
delivers received data to the application
only when prior gaps in the TCP sequence space are filled,
the \utcp receiver
makes data segments available to the application immediately upon receipt,
skipping TCP's usual reordering queue.
The data the \utcp stack delivers to the application
in successive application reads
may skip forward and backward in the transmitted byte stream,
and \utcp may even deliver portions
of the transmitted stream multiple times.
\utcp guarantees only that the data returned by each application read
corresponds to {\em some} contiguous sequence of bytes
in the sender's transmitted stream.

Second,
when servicing an application's read,
the \utcp receiver also delivers the logical offset
of the first returned byte
in the sender's original byte stream---information 
that a TCP receiver must maintain to arrange received segments in order.

\begin{figure}[htb]
\centering
\includegraphics[width=0.99\textwidth]{figures/utcpdelivery.eps}
\caption{
Delivery behavior of (a) standard TCP, and (b) \utcp,
upon receipt of in-order and out-of-order segments.
(taken from~\cite{nowlan12fitting})}
\label{f:utcpdelivery}
\end{figure}

Figure~\ref{f:utcpdelivery} illustrates \utcp's receive-side behavior,
in a simple scenario
where three TCP segments arrive in succession:
first an in-order segment, then an out-of-order segment,
and finally a segment filling the gap between the first two.
With \utcp,
the application receives each segment as soon as it arrives,
along with the sequence number information it needs to reconstruct
a complete internal view of
whichever fragments of the TCP stream have arrived.

\subsubsection{\utcp: Sender-Side Modifications}
While \utcp's receiver-side enhancements
address the ``latency tax''
on segments waiting in TCP's reordering buffer,
TCP's sender-side queue can also introduce latency,
as segments the application has already written to a TCP socket---%
and hence ``committed'' to the network---%
wait until TCP's flow and congestion control
allow their transmission.
Many applications can benefit from the ability to ``late-bind''
their decision on {\em what} to send until the last possible moment,
and also from being able to transmit a message of higher priority
that bypasses any lower priority messages in the sender-side queue.

A \utcp sender allows a sending application
to specify a tag with each application write,
which the \utcp sender currently
interprets as a priority level.
Instead of unconditionally placing the newly-written data
at the tail of the send queue as TCP normally would,
\utcp{} {\em inserts} the newly-written data into the send queue
just {\em before} any lower-priority data
in the send queue not yet transmitted.

With these modifications to a TCP stack,
none of which require changes to the TCP wire-format,
\utcp offers an interface which,
while not convenient for applications, 
is powerful.
In the next Section,
we discuss how we build a userspace library that uses this interface
that provides a simple unordered delivery service,
unordered delivery of encrypted messages,
and logically separate data streams 
within a single \utcp connection.

\subsubsection{Datagrams atop \utcp}
\label{sec:dgrams_utcp}

Applications built on datagram substrates such as UDP
generally assume the underlying layer preserves datagram boundaries.
TCP's stream-oriented semantics do not preserve
any application-relevant frame boundaries within a stream, however.
Both the TCP sender and network middleboxes can and do
coalesce TCP segments or re-segment TCP streams
in unpredictable ways~\cite{honda2011still}.

Atop \utcp,
a userspace library can reconstruct contiguous fragments 
in the received data stream
using the metadata sequence number information 
that \utcp passes along at the receiver.
However,
providing unordered message delivery service atop \utcp
requires delimiting application messages in the bytestream.
While record delimiting is commonly done
by application protocols such as HTTP, SIP, and many others,
a key property that we require to provide a true unordered delivery service
is that a receiver must be able to extract a given message 
independently of other messages.
That is, 
as soon as a complete message is received,
the message delimiting mechanism must
allow for extraction of the message from the bytestream fragment,
without relying on the receipt of earlier messages.

Minion implements self-delimiting messages in two ways:
\begin{enumerate}
    \item To encode application datagrams efficiently, 
    the userspace library employs 
    {\em Consistent-Overhead Byte Stuffing}, or
    COBS~\cite{cheshire97consistent}
    to delimit and extract messages.
    COBS is a binary encoding
    which eliminates {\em exactly} one byte value from a record's encoding
    with minimal bandwidth overhead.
    To encode an application record,
    COBS first scans the record for {\em runs}
    of contiguous marker-free data followed by exactly one marker byte.
    COBS then removes the trailing marker,
    instead {\em prepending} a non-marker byte indicating the run length.
    A special run-length value indicates a run of 254 bytes
    {\em not} followed by a marker in the original data,
    enabling COBS to divide arbitrary-length runs into 254-byte runs
    encoded into 255 bytes each,
    yielding a worst-case expansion of only 0.4\%.

    \item The userspace library 
    coaxes out-of-order delivery from
    the {\em existing} TCP-oriented TLS wire format,
    producing an encrypted datagram substrate
    indistinguishable on the wire from standard TLS connections.
    TLS~\cite{rfc5246} already breaks its communication into {\em records},
    encrypts and authenticates each record,
    and prepends a header
    for transmission on the underlying TCP stream.
    TLS was designed to decrypt records strictly in-order, however,
    creating challenges 
    which the userspace library overcomes~\cite{nowlan12fitting}.
    Run on port 443,
    our encrypted stream atop \utcp is indistinguishable from HTTPS---%
    regardless of whether the application actually uses HTTP headers,
    since the HTTP portion of HTTPS streams are TLS-encrypted anyway.
    Deployed this way,
    Minion effectively offers an end-to-end protected substrate
    in the ``HTTP as the new narrow waist'' philosophy~\cite{popa10http}.
\end{enumerate}

\subsection{Impact on Real Applications}

Minion's unordered delivery service 
benefits a number of applications;
we refer the reader to detailed discussion and experiments 
in~\cite{nowlan12fitting}.
Of these applications,
we briefly discuss how Minion fits within
ongoing efforts to develop a next-generation transport for the web,
such as SPDY\cite{spdy} and the Internet Engineering Task Force (IETF)'s 
HTTP/2.0 (httpbis) effort.
Developing a next-generation HTTP 
requires either submitting to TCP's latency tax
for backward compatibility,
as with SPDY's use of TLS/TCP,
or developing and deploying new transports atop UDP,
neither of which,
as we discussed earlier in this section,
is a satisfying alternative.

Minion bridges this gap 
and demonstrates that it is possible to obtain
unordered delivery from wire-compatible TCP and TLS streams
with surprisingly small changes to TCP stacks
and application-level code.
These protocols offer latency-sensitive applications
performance benefits comparable to UDP or DCCP,
with the compatibility benefits of TCP and TLS.
Without discounting the value of UDP and newer OS-level transports,
Minion offers a more conservative path toward
the performance benefits of unordered delivery,
which we expect to be useful to applications
that use TCP for a variety of pragmatic reasons.

%\subsection{Editorial Note}
%Some of the text and figures in this Section have been adapted from
%the following two publications:

### 2013-07-22 Compiled/chapter5/mptcp.tex
all chapters compile independently

%\ed{> Part IV describes in great detail Multipath TCP and TCP Minion,
%two recent significant evolutions of TCP that were built with
%deployability in mind. \textbf{Multipath TCP - Costin (10 pages)}}

Today's networks are multipath: mobile devices have multiple wireless
interfaces, datacenters have many redundant paths between servers and
multi-homing has become the norm for big server farms. Meanwhile, TCP
is essentially a single path protocol: when a TCP connection is
established, it is bound to the IP addresses of the two communicating
hosts. If one of these addresses changes, for whatever reason, the
connection fails. In fact a TCP connection cannot even be
load-balanced across more than one path within the network, because
this results in packet reordering and TCP misinterprets this
reordering as congestion and slows down.

This mismatch between today's multipath networks and TCP's single-path
design creates tangible problems. For instance, if a smartphone's WiFi interface
loses signal, the TCP connections associated with it stall - there is
no way to migrate them to other working interfaces, such as 3G. This
makes mobility a frustrating experience for users.  Modern datacenters
are another example: many paths are available between two endpoints,
and equal cost multipath routing randomly picks one for a particular TCP
connection.  This can cause collisions where multiple flows get placed
on the same link, hurting throughput - to such an extent that average 
throughput is halved in some scenarios.

Multipath TCP (MPTCP) \cite{rfc6284} is a major modification to TCP that allows multiple
paths to be used simultaneously by a single 
connection. Multipath TCP circumvents the issues above and several
others that affect TCP.  Changing TCP to use multiple paths is not a
new idea: it was originally proposed more than fifteen years ago by
Christian Huitema in the Internet Engineering Task Force (IETF) \cite{Huitema_Multi-homed:1995}, and
there have been a half-dozen more proposals since then to similar
effect. Multipath TCP draws on the experience gathered in previous
work, and goes further to solve issues of fairness when competing with
regular TCP and deployment issues due to middleboxes in today's
Internet.

\subsection{Overview of Multipath TCP}

The design of Multipath TCP has been influenced by many requirements,
but there are two that stand out: application compatibility and
network compatibility. Application compatibility implies that 
applications that today run over TCP should work without any change 
over Multipath TCP. Next, Multipath TCP must
operate over any Internet path where the TCP protocol operates.

As explained earlier, many paths on today's Internet include middleboxes that, unlike routers,
 know about the TCP connections they forward, and affect them in special ways. 
Designing TCP extensions that can safely traverse all these middleboxes has
proven to be challenging.

Multipath TCP allows multiple "subflows" to be combined to form a single MPTCP
session.  An MPTCP session starts with an initial subflow which is
very similar to a regular TCP connection, with a three way handshake.  After
the first MPTCP subflow is set up, additional subflows can be
established.  Each subflow also looks very similar to a
regular TCP connection, complete with three-way handshake and graceful tear-down,
but rather than being a separate connection it is bound into an
existing MPTCP session.  Data for the connection can
then be sent over any of the active subflows that has the capacity to
take it.

To examine Multipath TCP in more detail, let us consider a very simple
scenario with a smartphone client and a single-homed server. The
smartphone has two network interfaces: a WiFi interface and a 3G
interface; each has its own IP address.  The server, being
single-homed, has a single IP address. In this environment, Multipath
TCP would allow an application on the mobile device to use a single MPTCP
session that can use both the WiFi and the 3G interfaces to
communicate with the server.  The application opens a regular TCP socket, and
the kernel enables MPTCP by default if the remote end supports it, using both paths. 
The application does not need to concern itself with which radio interface is working best at any instant;
MPTCP handles that for it. In fact, Multipath TCP can work when
both endpoints are multihomed (in this case subflows are opened between
all pairs of "compatible" IP addresses), or even in the case when 
both endpoints are single homed (in this case different subflows will use
 different port numbers, and can be routed differently by multipath routing 
in the network).

\paragraph{Connection Setup.} Let us walk through the establishment of an MPTCP connection. 
Assume that the mobile device chooses its 3G interface to open the
connection. It first sends a \texttt{SYN} segment to the server. This segment
contains the \texttt{MP\_CAPABLE} TCP option indicating that the mobile device
supports Multipath TCP. This option also contains a key which is
chosen by the mobile device. The server replies with a \texttt{SYN+ACK} segment
containing the \texttt{MP\_CAPABLE} option and the key chosen by the server. The
mobile device completes the handshake by sending an \texttt{ACK} segment.

The initial MPTCP connection setup is shown graphically in the top part of Figure \ref{fig:mptcp_handshake},
where the segments are regular TCP segments carrying new multipath TCP-related options (shown in green).

\begin{figure}[t]
\centering
\includegraphics[width=0.9\textwidth]{figures/mptcp_handshake.png}
\caption{Multipath TCP handshake: multiple subflows can be added and removed after the initial
connection is setup and connection identifiers are exchanged.}
\label{fig:mptcp_handshake}
\end{figure}

At this point the Multipath TCP connection is established and the
client and server can exchange TCP segments via the 3G path. How
could the mobile device also send data through this Multipath TCP
session over its WiFi interface?

Naively, it could simply send some of the segments over the WiFi
interface.  However most ISPs will drop these packets, as they would
have the source address of the 3G interface.  Perhaps the client could
tell the server the IP address of the WiFi interface, and use that
when it sends over WiFi?  Unfortunately this will rarely work:
firewalls and similar stateful middleboxes on the WiFi path expect to
see a \texttt{SYN} segment before they see data segment.  The only solution that
will work reliably is to perform a regular three-way handshake on the WiFi path
before sending any packets that way, so this is what Multipath TCP
does.  This handshake carries the \texttt{MP\_JOIN} TCP option, providing
information to the server that can securely identify the
correct connection to associate this additional subflow with.  The
server replies with \texttt{MP\_JOIN} in the \texttt{SYN+ACK}, and the new subflow is
established (this is shown in the bottom part of Figure \ref{fig:mptcp_handshake}).

An important point about Multipath TCP, especially in the context of 
mobile devices, is that the set of subflows that are associated to a Multipath TCP 
connection is not fixed. Subflows can be dynamically added and removed from a 
Multipath TCP connection throughout its lifetime, without affecting the 
bytestream transported on behalf of the application. 
If the mobile device moves to another WiFi network, it will receive a new IP 
address. At that time, it will open a new subflow using its newly allocated address 
and tell the server that its old address is not usable anymore. The server will now 
send data towards the new address. These options allow mobile devices to easily move
through different wireless connections without breaking their
Multipath TCP connections \cite{Paasch_Mobile:2012}.

Multipath TCP also implements mechanisms that allows to inform the
remote host of the addition/removal of addresses even when an endpoint operates behind a NAT, or when a subflow using a
different address family is needed (e.g. IPv6). Endpoints can send
an \texttt{ADD\_ADDR} option that contains an address identifier together with an address.
The address identifier is unique at the sender, and allows it to identify its addresses
even when it is behind a NAT. Upon receiving an advertisement, the endpoint may initiate a new subflow to
the new address. An address withdrawal mechanism is also provided via the \texttt{REMOVE\_ADDR}
option that also carries an address identifier.

\paragraph{Data Transfer.} Assume now that two subflows have been established over WiFi and 3G:
the mobile device can send and receive data segments over both. Just like TCP, Multipath
TCP provides a bytestream service to the application.  In
fact, standard applications can function over MPTCP without being
aware of it - MPTCP provides the same socket interface as TCP.

Since the two paths will often have different delay characteristics,
the data segments sent over the two subflows will not be received in
order. Regular TCP uses the sequence number in the TCP header
to put data back into the original order. A simple solution for
Multipath TCP would be to just reuse this sequence number as
is.

Unfortunately, this simple solution would create problems with
some existing middleboxes such as firewalls. On each path, a middlebox
would only see half of the packets, so it would observe many gaps in
the TCP sequence space. Measurements indicate that some middleboxes
react in strange ways when faced with gaps in TCP sequence
numbers \cite{honda2011still}. Some discard the out-of-sequence segments while others try to
update the TCP acknowledgments in order to "recover" some of these
gaps. With such middleboxes on a path, Multipath TCP cannot safely
send TCP segments with gaps in the TCP sequence number space. On the
other hand, Multipath TCP also cannot send every data segment over all
subflows: that would be a waste of resources.

To deal with this problem, Multipath TCP uses its own sequence
numbering space.  Each segment sent by Multipath TCP contains two
sequence numbers: the subflow sequence number inside the regular TCP
header and an additional Data Sequence Number (DSN).

This solution ensures that the segments sent on any given
subflow have consecutive sequence numbers and do not upset
middleboxes.  Multipath TCP can then send some data sequence numbers
on one path and the remainder on the other path; the DSN will be used by the Multipath TCP
receiver to reorder the bytestream before it is given to the receiving
application.

Before we explain the way the Data Sequence Number is encoded, we first need to discuss two other
key parts of Multipath TCP  that are affected by the additional sequence number
space---flow control and acknowledgements.

\paragraph{Flow Control.}
TCP's receive window indicates the number of bytes beyond the sequence number from the acknowledgment field that the receiver 
can buffer. The sender is not permitted to send more than this amount of additional data.
Multipath TCP also needs to implement flow control, although segments
can arrive over multiple subflows. If we 
inherit TCP's interpretation of receive window, this would imply an MPTCP receiver maintains a pool of buffering 
per subflow, with receive window indicating per-subflow buffer occupancy. Unfortunately such an interpretation 
can lead to deadlocks:
\begin{enumerate}
\item The next segment that needs to be passed to the application was
  sent on subflow 1, but was lost.
\item In the meantime subflow 2 continues delivering data, and fills
  its receive window.
\item Subflow 1 fails silently.
\item The missing data needs to be re-sent on subflow 2, but there is
  no space left in the receive window, resulting in a deadlock.
\end{enumerate}

The correct solution is to generalize TCP's receive window semantics to MPTCP. For each connection \emph{a 
single receive buffer pool should be shared between all subflows}. The receive window then indicates 
the maximum data sequence number that can be sent rather than the maximum subflow sequence number. As 
a segment resent on a different subflow always occupies the same data sequence space, deadlocks cannot occur.

The problem for an MPTCP sender is that to calculate the highest data sequence number that can be sent, 
the receive window needs to be added to the highest data sequence number acknowledged. However the ACK 
field in the TCP header of an MPTCP subflow must, by necessity,
indicate only subflow sequence numbers to cope with middleboxes. 
Does MPTCP need to add an extra data acknowledgment field for the receive window to be interpreted correctly?

\paragraph{Acknowledgments.} 
The answer is positive: MPTCP needs and uses \emph{explicit
connection-level acknowledgments} or \emph{DATA\_ACKs}. The alternative is to infer connection-level
acknowledgments from subflow acknowledgments, by using a scoreboard maintained by the sender that maps
subflow sequence numbers to data sequence numbers. Unfortunately, MPTCP segments and their associated ACKs
will be reordered as they travel on different paths, making it impossible to correctly infer the connection-level
acknowledgments from subflow-level ones \cite{raiciu2012hard}.

\paragraph{Encoding.}
We have seen that in the forward path we need to encode a mapping of
subflow bytes into the data sequence space, and in the reverse path we
need to encode cumulative data acknowledgments.  There are two viable
choices for encoding this additional data:
\begin{itemize}
\item Send the additional data in TCP options.
\item Carry the additional data within the TCP payload, using a
  chunked or escaped encoding to separate control data from payload
  data.
\end{itemize}

For the forward path there aren't compelling arguments either
way, but the reverse path is a different matter.
Consider a hypothetical encoding that divides the payload into chunks
where each chunk has a TLV (type-length-value) header.  A data acknowledgment can then be
embedded into the payload using its own chunk type.  Under most
circumstances this works fine.  However, unlike TCP's pure ACK, anything
embedded in the payload must be treated as data.  In particular:
\begin{itemize}
\item It must be subject to flow control because the receiver must
  buffer data to decode the TLV encoding.

\item If lost, it must be retransmitted consistently, so that
  middleboxes can track sequence state correctly\footnote{TCP proxies re-send the original
    content they see a ``retransmission'' with different data. }
\item If packets before it are lost, it might be necessary to wait for retransmissions
  before the data can be parsed - causing head-of-line blocking. 
%\item The packet it is contained in may be split into two packets or
%  merged with other packets by middleboxes or TCP segmentation offload hardware.
\end{itemize}

%This last issue does not greatly affect MPTCP at present.  We observed
%3\% of paths coalesced segments (7\% on port 80) and 3\% of paths also
%split segments (7\% on port 80).  However almost all of these removed {\sc
%  Mp\_Capable} from the \SYNs, so MPTCP would not be enabled.  We did
%however observe one middlebox that passed {\sc Mp\_Capable} and
%resegmented data, but only when no new options were present in data
%packets - this might affect MPTCP if chunked payload encoding were used.

\begin{figure}[t]
\centering
\includegraphics[trim=1cm 17cm 1cm 1cm,clip=true,height=6cm]{figures/dataack_deadlock}
\caption{Flow Control on the path from C to S inadvertently stops the data flow from S to C}
\label{fig:dack_deadlock}
\end{figure}

Flow control presents the most obvious problem for the
chunked payload encoding.  Figure \ref{fig:dack_deadlock} provides an
example. Client C is pipelining requests to server S; meanwhile S's
application is busy sending the large response to the first request so
it isn't yet ready to read the subsequent requests.  At this point,
S's receive buffer fills up.

S sends segment 10, C receives it and wants to send the \texttt{DATA\_ACK}, but
cannot: flow control imposed by S's receive window stops him.
Because no \texttt{DATA\_ACK}s are received from C, S cannot free his send
buffer, so this fills up and blocks the sending application on S.  S's
application will only read when it has finished sending data to C, but
it cannot do so because its send buffer is full.  The send buffer can
only empty when S receives the \texttt{DATA\_ACK} from C, but C cannot send this 
until S's application reads.  This is a classic deadlock cycle.
As no \texttt{DATA\_ACK} is received, S will eventually time out the data it
sent to C and will retransmit it;  after many retransmits the whole connection will time out.

The conclusion is that \texttt{DATA\_ACK}s cannot be safely encoded in the
payload.  The only real alternative is to encode them in TCP options
which (on a pure ACK packet) are not subject to flow control.

\paragraph{The Data Sequence Mapping.} 
If MPTCP must use options to encode \texttt{DATA\_ACK}s, it is simplest to also
encode the mapping from subflow sequence numbers to data sequence
numbers in a TCP option.  This is the {\em data sequence
mapping} or DSM.

At first glance it seems the {\sc DSM} option simply needs to carry the data
sequence number corresponding to the start of the MPTCP segment.
Unfortunately middleboxes and interfaces that implement TSO or LRO make this far from
simple.

Middleboxes that re-segment data would cause a problem.
TCP Segmentation Offload (TSO) hardware in the
network interface card (NIC) also re-segments data and is commonly used to improve performance.
The basic idea is that the OS sends large segments and the NIC 
re-segments them to match the receiver's MSS.  What does TSO do 
with TCP options?  A test of 12 NICs supporting
TSO from four different vendors showed that all of them copy a TCP option sent
by the OS on a large segment into all the split segments \cite{raiciu2012hard}.

If MPTCP's DSM option only listed the data sequence number, TSO would
copy the same DSM to more than one segment, breaking the mapping.
Instead the DSM option must say precisely which subflow bytes map to
which data sequence numbers.  But this is further complicated by
middleboxes that modify the initial sequence number of TCP connections
and consequently rewrite all sequence numbers (many firewalls behave like this).
Instead, the {\sc DSM} option must map the offset from the
subflow's initial sequence number to the data sequence number, as the
offset is unaffected by sequence number rewriting.  The option must
also contain the length of the mapping.  This is robust - as long as
the option is received, it does not greatly matter which packet
carries it, so duplicate mappings caused by TSO are not a
problem.

\paragraph{Dealing with Content-Modifying Middleboxes.}
Multipath TCP and content-modifying middleboxes (such as application-level NATs, e.g. for FTP)
have the potential to interact badly.  In particular, due to FTP's ASCII
encoding, re-writing an IP address in the payload can necessitate
changing the length of the payload.  Subsequent sequence and ACK
numbers are then fixed up by the middlebox so they are consistent from
the point of view of the end systems.

Such length changes break the DSM option mapping - subflow bytes can
be mapped to the wrong place in the data stream.  They also
break every other possible mapping mechanism, including chunked
payloads.  There is no easy way to handle such middleboxes.

That is why MPTCP includes an optional checksum in
the DSM mapping to detect such content changes. If an MPTCP host
receives a segment with an invalid DSM checksum, it
rejects the segment and triggers a fallback process:  if any
other subflows exists, MPTCP terminates the subflow on which the
modification occurred;  if no other subflow exists, MPTCP drops back to
regular TCP behavior for the remainder of the connection, allowing the
middlebox to perform rewriting as it wishes. This fallback mechanism
preserves connectivity in the presence of middleboxes.

For efficiency reasons, MPTCP uses the same 16-bit ones complement checksum
used in the TCP header.  This allows the checksum over the payload to
be calculated only once.  The payload checksum is added to a checksum
of an MPTCP pseudo header covering the DSM mapping values and then
inserted into the DSM option.   The same payload checksum is added to the
checksum of the TCP pseudo-header and then used in the TCP checksum field.
MPTCP allows checksums to be disabled for high performance environments such as data-centers where
there is no chance of encountering such an application-level gateway.

The fall-back-to-TCP process, triggered by a checksum failure, can also
be triggered in other circumstances.  For example, if a routing
change moves an MPTCP subflow to a path where a middlebox
removes DSM options, this also triggers the fall-back procedure.

\begin{figure}[t]
\centering
\begin{verbatim}
                      1                   2                   3
  0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1
 +---------------+---------------+-------+----------------------+
 |     Kind      |    Length     |Subtype| (reserved) |F|m|M|a|A|
 +---------------+---------------+-------+----------------------+
 |           Data ACK (4 or 8 octets, depending on flags)       |
 +--------------------------------------------------------------+
 |   Data Sequence Number (4 or 8 octets, depending on flags)   |
 +--------------------------------------------------------------+
 |              Subflow Sequence Number (4 octets)              |
 +-------------------------------+------------------------------+
 |  Data-level Length (2 octets) |      Checksum (2 octets)     |
 +-------------------------------+------------------------------+
\end{verbatim}
\caption{The Data Sequence Signal option in MPTCP that carries the Data Sequence Mapping information,
the Data ACK, the Data FIN and connection-fall-back options. }
\label{fig:mptcp_dss}
\end{figure}

\paragraph{Connection Release.} Multipath TCP must allow a connection to survive
even though its subflows are coming and going. Subflows in MPTCP can be torn down
by means of a four-way handshake as regular TCP flows---this ensures MPTCP
allows middleboxes to clear their state when a subflow is not used anymore.

MPTCP uses an explicit four-way handshake for connection tear-down indicated by a \texttt{DATA\_FIN} option.
The \texttt{DATA\_FIN} is MPTCP's equivalent to TCP's \texttt{FIN}, and it occupies one byte in the data-sequence 
space. A \texttt{DATA\_ACK} will be used to acknowledge the receipt of the \texttt{DATA\_FIN}. MPTCP requires
that the segment(s) carrying a \texttt{DATA\_FIN} must also have the \texttt{FIN} flag set - this ensures
all subflows are also closed when the MPTCP connection is being closed.

For reference, we show the wire format of the option used by MPTCP for data exchange in Figure 
\ref{fig:mptcp_dss}. This option encodes the Data Sequence Mapping, Data ACK, Data FIN and 
the fall-back options. The flags specify which parts of the option are valid, and help reduce option
space usage.

\subsection{Congestion Control}

One of the most important components in TCP is its congestion controller which 
enables it to adapt its throughput dynamically in response to changing network 
conditions. To perform this functionality, each TCP sender maintains a congestion 
window $w$ which governs the amount of packets that the sender can send without 
waiting for an acknowledgment. The congestion window is updated dynamically 
according to the following rules:

\begin{itemize}
\item On each ACK, increase the window $w$ by $1/w$.
\item Each loss decrease the window $w$ by $w/2$.
\end{itemize}

TCP congestion control ensures fairness:  when multiple connections 
utilize the same congested link each of them will independently converge to the 
same average value of the congestion window.

What is the equivalent of TCP congestion control for multipath
transport? The obvious question to ask is why not just run regular TCP congestion
control on each subflow?
Consider the scenario in Fig. \ref{fig:1bottleneck}. If multipath
TCP ran regular TCP congestion control on both paths, then the
multipath flow would obtain twice as much throughput as the single
path flow (assuming all RTTs are equal). This is unfair. To solve
this problem, one solution is to try and detect shared bottlenecks
but that is unreliable; a better solution is be less aggressive on
each subflow (i.e. increase window slower) such that in aggregate the MPTCP connection 
is no more aggressive than a single regular TCP connection.

Before we describe solutions to MPTCP congestion control, let's discuss the three goals that
multipath congestion control must obey \cite{mptcp-cc}:
\begin{enumerate}
\item[\textbf{Fairness}] If several subflows of the same MPTCP connection
share a bottleneck link with other TCP connections, MPTCP should not
get more throughput than TCP. 
\item[\textbf{Deployability}] The performance of all the
Multipath TCP subflows together should be at least that of regular TCP
on any of the paths used by a Multipath TCP connection.  This ensures
that there is an incentive to deploy Multipath TCP. 
\item[\textbf{Efficiency}] A final, most important goal is that Multipath TCP should prefer efficient paths,
which means it should send more of its traffic on paths experiencing
less congestion.
\end{enumerate}

\begin{figure}[t]
\centering
\includegraphics*[width=0.45\columnwidth,bb=167 245 509 380]{figures/1bottleneck}
\caption{A scenario which shows the importance of weighting the aggressiveness of subflows (taken from \cite{mptcp-cc}).}
\label{fig:1bottleneck}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=0.5\textwidth]{figures/figure1}
\caption{Two links each with capacity 20~pkts/s. The top link is used by a single TCP connection,
and the bottom link is used by two TCP connections. A Multipath TCP connection uses
both links. Multipath TCP pushes most of its traffic onto less congested top link,
making the two links behave like a resource pool of capacity 40~pkts/s. Capacity is divided equally,
with each flow having throughput 10~pkts/s. (taken from \cite{mptcp-cc})}
\label{fig:mptcp1}
\end{figure}

Intuitively, this last goal ensures wide-area load balancing of traffic: 
when a multipath connection is using two paths loaded unevenly (such as 
Figure \ref{fig:mptcp1}), the multipath transport will prefer the unloaded path
and push most of its traffic there; this will decrease the load on the 
congested link and increase it on the less congested one.

If a large enough fraction of flows are multipath, congestion will spread 
out evenly across collections of links, creating "resource pools": links 
that act together as if they are a single, larger capacity link shared by 
all flows. This effect is called resource pooling \cite{pooling}.
Resource pooling brings two major benefits, discussed in the paragraphs below.

\paragraph{Increased Fairness.} 
Consider the example shown in Figure \ref{fig:mptcp1}: congestion
balancing ensures that all flows have \textbf{the same throughput}, making
the two links of 20~pkt/s act like a single pooled link with capacity 40~pkt/s 
shared fairly by the four flows. If more MPTCP flows would be added, the two links would still behave as a pool,
sharing capacity fairly among all flows. 
Conversely, if we remove the Multipath TCP flow, the links no longer form a pool, and the throughput allocation
is unfair as the TCP connection using the top path gets twice as much throughput as the TCP connections using the bottom path.

\paragraph{Increased Throughput.}
Consider the somewhat contrived scenario in Fig.\ref{fig:klingon},
and suppose that the three links each have capacity 12Mb/s.
If each flow split its traffic evenly
across its two paths subflow would get 4Mb/s 
hence each flow would get 8Mb/s. But if each flow used only 
the one-hop shortest path, it could get 12Mb/s: this is because
two-hop paths consume double the resources of one-hop paths,
and in a congested network it makes sense to only use the one-hop paths.

\begin{figure}[tb]
\centering
\includegraphics*[width=0.8\columnwidth]{figures/3pipes}
\caption{A scenario to illustrate the importance of choosing the
  less-congested path. (taken from \cite{mptcp-cc})}
\label{fig:klingon}
\end{figure}

In an idle network, however, using all available paths is much better: 
consider the case when only the blue connection is using the links. In this
case this connection would get 24Mb/s throughput; using the one hop path
alone would only provide 12Mb/s.

In summary, the endpoints need to be able to dynamically decide 
which paths to use based on conditions in the network. 
A solution has been devised in the theoretical literature on
congestion control, independently by Kelly and Voice \cite{kelly-voice} and
Han et al. \cite{towsley-cc}.  The core idea is that a multipath flow should
shift all its traffic onto the least-congested path.
In a situation like Fig. \ref{fig:klingon} the
two-hop paths will have higher drop probability than the one-hop
paths, so applying the core idea will yield the efficient
allocation.
Surprisingly it turns out that this can be achieved by doing independent
congestion control at endpoints.

\paragraph{Multipath TCP Congestion Control.}
The theoretical work on multipath congestion control \cite{kelly-voice,towsley-cc} assumes a rate-based
protocol, with exponential increases of the rate. TCP, in contrast, is a packet-based
protocol, sending $w$ packets every round-trip time (i.e. the rate is $w/RTT$); 
a new packet is sent only when an acknowledgment is received, confirming that an existing 
packet has left the network. This property is called ACK-clocking 
and is nice because it has good stability properties:
when congestion occurs round-trip times increase (due to buffering), which
\emph{automatically reduces the effective rate} \cite{Jacobson_congestion:88}.

Converting a theoretical rate-based exponential protocol to a practical 
packet-based protocol fair to TCP turned out to be more difficult than expected. 
There are two problems that appear \cite{mptcp-cc}: 
\begin{itemize}
\item When loss rates are equal on all paths, the theoretical algorithm will
place all of the window on one path or the other, not on both---this effect was termed 
``flappiness'' and it appears because of the discrete (stochastic) nature of packet losses which
are not captured by the differential equations used in theory.
\item The ideal algorithm always prefers paths with lower loss rate, but in practice these may have poor
performance. Consider a mobile phone with WiFi and 3G links: 3G links have very low loss rates and huge round-trip
times, resulting in poor throughput. WiFi is lossy, has shorter round-trip times and typically offers much better throughput.
In this common case, a ``perfect'' controller would place all traffic on the 3G path, violating the second goal (deployability).
\end{itemize}

The pragmatic choice is to sacrifice some load-balancing ability to ensure greater stability and
to offer incentives for deployment. This is what Multipath TCP congestion control does.

Multipath TCP congestion control is a series of simple changes to the standard TCP congestion control mechanism. Each subflow 
has its own congestion window, that is halved when packets are lost, as in standard TCP \cite{mptcp-cc}.

Congestion balancing is implemented in the increase phase of 
congestion control: here Multipath TCP will allow less congested subflows 
to increase proportionally more than congested ones. Finally, the total 
increase of Multipath TCP across all of its subflows is dynamically chosen 
in such a way that it achieves the first and second goals above.

The exact algorithm is described below and it satisfies the goals we've discussed:
\begin{itemize}
\item Upon ACK on subflow $r$, increase the window $w_r$ by
$\min(a/w_{total}),1/w_r)$.
\item Upon loss on subflow $r$, decrease the window $w_r$ by $w_r/2$.
\end{itemize}
Here
\begin{equation}
\label{eq:a}
a = w_{total} \frac{\max_r w_r/RTT_r^2}{(\sum_r
  w_r/RTT_r)^2},
\end{equation}
$w_r$ is the current window size on subflow $r$ and $w_{total}$ is the
sum of windows across all subflows.

The algorithm biases the increase towards uncongested paths: these will receive
more ACKs and will increase accordingly. However, MPTCP does keep some traffic
even on the highly congested paths; this ensures stability and allows it
to quickly detect when path conditions improve.

$a$ is a term that is computed dynamically upon each packet drop. Its purpose 
is to make sure that MPTCP gets at least as much throughput as TCP on the best path.
To achieve this goal, $a$ is computed by estimating how much TCP would get on each MPTCP path
(this is easy, as round-trip time and loss-rates estimates are known) and ensuring that MPTCP
in stable state gets at least that much. A detailed discussion on the design of the MPTCP congestion
control algorithm is provided in \cite{mptcp-cc}.

For example, in the three-path example above, the flow will put
45\% of its weight on each of the less congested path and 10\% on the
more congested path. This is intermediate between regular TCP
(33\% on each path) and a perfect load balancing algorithm (0\% on the more congested path)
that is impossible to implement in practice.

The window increase is capped at $1/w_r$, which ensures that
the multipath flow can take no more capacity on either path than a
single-path TCP flow would.

In setting the $a$ parameter, Multipath TCP congestion control uses subflow delays to compute the target rate.
Using delay for congestion control is not a novel idea: TCP Vegas \cite{vegas}, for instance, performs
congestion control only by tracking RTT values. Vegas treats increasing RTTs as a sign of congestion, instead
of relying on packet losses as regular TCP does. That is why Vegas loses out to regular TCP at shared bottlenecks,
and is probably the reason for its lack of adoption. MPTCP does not treat delay as congestion: it just uses it
to figure out the effective rate of a TCP connection on a given path. This allows MPTCP to compete fairly with TCP.

\paragraph{Alternative Congestion Controllers for Multipath TCP.} The standardized Multipath
TCP congestion control algorithm chooses a trade-off between load balancing, stability and
the ability to quickly detect available capacity. The biggest contribution of this
work is the clearly defined goals for what multipath congestion control should do,
and an instantiation that achieves (most of) the stated goals in practice.

This research area is relatively new, and it is likely that more work will lead to better 
algorithms---if not generally applicable, then at least tailored to some practical use-cases.
A new and interesting congestion controller called Opportunistic Linked Increases Algorithm (OLIA) 
has already been proposed \cite{olia} that offers better load balancing with seemingly few drawbacks.

We expect this area to be very active in the near future; of particular interest are designing multipath
versions of high-speed congestion control variants deployed in practice, such as Cubic or Compound TCP.

\subsection{Implementation and performance}

We now briefly cover two of the most compelling use cases for
Multipath TCP by showing a few evaluation results. We focus on mobile
devices and datacenters but note that Multipath TCP can also help in
other scenarios.  For example, multi-homed web-servers can perform
fine-grained load-balancing across their uplinks, while dual-stack
hosts can use both IPv4 and IPv6 within a single Multipath TCP
connection.

The full Multipath TCP protocol has been implemented in the Linux
kernel; its congestion controller has also been implemented in the ns2
and htsim network simulators. The results presented in here are from
the Linux kernel implementation \cite{raiciu2012hard}.

\begin{figure}[t]
\centering
\includegraphics[width=0.5\textwidth]{figures/figure2}
\caption{(Mobility) A mobile device is using both its WiFi and 3G interfaces, and then the WiFi
interface fails. We plot the instantaneous throughputs of Multipath TCP and application-layer
handover. (taken from \cite{Paasch_Mobile:2012})} 
\label{fig:mptcp2}
\end{figure}

The mobile measurements focus on a typical mode of operation where the device is 
connected to WiFi, the connection goes down and the phone switches to using 3G. 
The setup uses a Linux laptop connected to a WiFi and a 3G network, downloading a 
file using HTTP.

3G to Wifi handover is implemented in today's phones by changing the application to monitor
the network interfaces. When the app detects the loss of the interface, it creates a new TCP connection 
to the server and the connection can resume. This solution, while simple, is not applicable to all applications
because some of the bytes succesfully delivered by the old connection may be resent by the new one.

Applications that rely on HTTP GET requests (with no side effects) are, however, easy to change. The 
HTTP range header allows a client to resume the download of a file from a specified offset. This is the de-facto
standard for today's apps.

Figure \ref{fig:mptcp2} compares application layer-handover (with HTTP-range) against Multipath TCP.
The figure shows a 
smooth handover with Multipath TCP, as data keeps flowing despite the interface 
change. With application-layer handover there is a downtime of 4 seconds where the transfer 
stops-this is because it takes time for the application to detect the interface down 
event, and it takes time for 3G to ramp up.

Multipath TCP enables unmodified mobile applications to survive interface changes with
little disruption. Selected apps can be modified today to support handover, but their performance
is worse than with MPTCP. A more detailed discussion of the utilization of
Multipath TCP in WiFi/3G environments may be found in \cite{Paasch_Mobile:2012}.

\begin{figure}[t]
\centering
\includegraphics[width=\textwidth,trim=0 11cm 0 9cm,clip=true]{figures/throughput-subway}
\caption{(3G to WiFi Handover) A Linux distribution is downloaded by a mobile user while on the Bucharest subway. 3G coverage is ubiquitous. WiFi is available only in stations.}
\label{fig:throughput-subway}
\end{figure}

We also present a real mobility trace in Figure \ref{fig:throughput-subway}, where a mobile user downloads a Linux distro on his laptop
while travelling on the Bucharest underground. The laptop uses a 3G dongle to connect to the cellular network and its WiFi NIC to connect 
to access points available in stations.

The figure shows the download speeds of the two interfaces during a 30 minute underground ride. 3G throughput is stable: the average is 
2.3Mbps (shown with a dotted line on the graph), and the instantaneous throughput varies inversely proportional with the distance 
to the 3G cell. WiFi throughput is much more bursty: in some stations the throughput soars to 40Mbps, while in others it is zero, as 
the laptop doesn't manage to associate and obtain an IP address quickly enough. The average WiFi throughput (3.3Mbps) is higher 
than the average 3G throughput.

While MPTCP uses both interfaces in this experiment, using just WiFi when it is available may be preferable as it reduces 3G data bills
and reduces the load of the cellular network. This is easy to implement with MPTCP, as it supports a simple  prioritization mechanism 
that allows the client to inform the server to send data via preferred subflow(s) \footnote{MPTCP sends an MP\_PRIO option to 
inform the remote end about changes in priority for the subflows.}.

\begin{figure}[t]
\centering
\includegraphics[width=0.5\textwidth]{figures/figure3}
\caption{(Datacenter load-balancing) This graph compares standard TCP with MPTCP 
with two and four flows, when tested on an EC2 testbed with 40 instances. Each host 
uses iperf sequentially to all other hosts. We plot the performance of all flows (Y axis)
in increasing order of their throughputs (X axis). (taken from \cite{raiciu2011improving})}
\label{fig:mptcp3}
\end{figure}

\paragraph{Datacenters.} We also show results from running Multipath TCP in a different scenario: the EC2 datacenter.
Like most datacenters today, EC2 uses a redundant network topology where many paths are 
available between any pair of endpoints, and where connections are placed randomly 
onto available paths. In EC2, 40 machines (or instances) ran the 
Multipath TCP kernel. A simple experiment was run where every machine measured
the throughput sequentially to every other machine using first TCP, then Multipath 
TCP with two and with four subflows. Figure \ref{fig:mptcp3} shows the sorted throughputs measured over 
12 hours. The results show that Multipath TCP brings significant improvements compared 
to TCP in this scenario. Because the EC2 network is essentially a black-box , it is 
difficult to pinpoint the root cause for the improvements; however, a
detailed analysis of the cases where Multipath TCP can help and why in is presented in 
\cite{raiciu2011improving}.

\subsection{Impact of Multipath TCP}

MPTCP is deployable today, as it was designed and shown to work over existing networks and unmodified applications.
Whether MPTCP will be adopted or not is unclear at this point; only time will tell.
If MPTCP is adopted, however, it will likely affect the development of both the network and the applications running 
on top of it. Here we briefly speculate on what the impact might be. 
Multipath TCP embeds two key mechanisms: load balancing to avoid congested paths, and the ability 
to use different IP addresses in the same connection. Having these mechanisms implemented at the end-hosts
means other layers (e.g. the network) may become simpler, and more efficient.

For instance, MPTCP seamlessly uses available bandwidth even if it is short-lived (e.g. WiFi in underground stations), 
and provides robustness when a subset of paths fail by quickly resending data on working paths. This could take
some of the load imposed on BGP, that today must quickly reconverge in the case of a failure. With MPTCP at the endpoints
failures would be ``masked'' automatically, allowing BGP to react to failures slowly which will ensure there are no 
oscillations. However, one must ensure that the paths used by the end-systems are disjoint and that the failure
does not affect all of them.

Another example is layer 2 mobility, implemented in both WiFi and cellular networks, that aims to do a fast handover from 
one access point to the next, or from one cell to the next. With MPTCP as a transport protocol, a WiFi client could connect
simultaneously to two access points or cells (provided the L2 protocol allows it), and load balance traffic on the best 
link. MPTCP could replace fast handovers with slow handovers that are more robust. Of course, some changes may be needed 
at layer 2 to support such load balancing.

Network topologies may be designed differently if MPTCP is the transport protocol. An example is GRIN \cite{grin}, a work that
proposes to change existing datacenter networks by randomly interconnecting servers in the same rack directly, using their free 
NIC ports. This allows a server to opportunistically send traffic at speeds larger than its access link by having its idle neighbor 
relay some traffic on its behalf. With a minor topology change and MPTCP, GRIN manages to better utilize the datacenter network core.

The mechanisms that allow MPTCP to use different IPs in the same transport connection (i.e. the connection identifier) 
also allow connections to be migrated across physical machines with different IP addresses. One direct application is seamless virtual
machine migration: an MPTCP-enabled guest virtual machine can just resume its connections after it is migrated.

On the application side many optimizations are possible. Applications like video
and audio streaming only need a certain bitrate to ensure user satisfaction - they rarely fully utilize the wireless capacity.
Our mobility graph in the previous section shows that, in principle, WiFi has enough capacity to support these apps, even if it is 
only available in stations. A simple optimization would be to download as much of the video as possible while on WiFi, and only
use 3G when the playout buffer falls below a certain threshold. This would ensure a smooth viewing experience while pushing as 
much traffic as possible over WiFi.

The examples shown in this section are only meant to be illustrative, and their potential impact or even feasibility 
is not clear at this point. Further, the list is not meant to be exhaustive. 
We have only provided it to show that Multipath TCP benefits go beyond than
increasing throughput, and could shape both the lower and the upper layers of the protocol stack in the future.

### 2013-07-23 Compiled/chapter2/chapter2.tex
Full compiled

@@
-\newcommand{\CLASSINPUTinnersidemargin}{0.7in}
-\newcommand{\CLASSINPUToutersidemargin}{0.7in}
+%\newcommand{\CLASSINPUTinnersidemargin}{0.7in}
+%\newcommand{\CLASSINPUToutersidemargin}{0.7in}
 \documentclass{article}
@@
 Boston, MA 02215, USA}
-\date{\today}							% Activate to display a given date or no date
+\date{}							% Activate to display a given date or no date
 
@@
 
-\tableofcontents
+%\tableofcontents
 
@@
    \centering 
-   \includegraphics[width=4in]{gas-control-loop} 
+   \includegraphics[width=4in]{figures/gas-control-loop} 
    \caption{The gas control loop}
@@
    \centering
-   \includegraphics[width=4in]{net-control-loop.jpg} 
+   \includegraphics[width=4in]{figures/net-control-loop.jpg} 
    \caption{The network control loop}
@@
    \centering
-   \includegraphics[width=2in]{concave-utility.jpg} 
+   \includegraphics[width=2in]{figures/concave-utility.jpg} 
    \caption{Concave utility function}
@@
    \centering
-   \includegraphics[width=2in]{concave-function.jpg} 
+   \includegraphics[width=2in]{figures/concave-function.jpg} 
    \caption{Concave function. 
@@
    \centering
-   \includegraphics[width=5in]{net-model.jpg} 
+   \includegraphics[width=5in]{figures/net-model.jpg} 
    \caption{A network model}
@@
    \centering
-   \includegraphics[width=1.5in]{convex-set.jpg} 
+   \includegraphics[width=1.5in]{figures/convex-set.jpg} 
    \caption{A convex set.
@@
    \centering
-   \includegraphics[width=4in]{greedy-alloc.jpg} 
+   \includegraphics[width=4in]{figures/greedy-alloc.jpg} 
    \caption{Greedy network allocation}
@@
    \centering
-   \includegraphics[width=4in]{max-min-fair.jpg} 
+   \includegraphics[width=4in]{figures/max-min-fair.jpg} 
    \caption{Max-min fair capacity allocation}
@@
    \centering
-   \includegraphics[width=3in]{system-model.jpg} % requires the graphicx package
+   \includegraphics[width=3in]{figures/system-model.jpg} % requires the graphicx package
    \caption{Typical system model}
@@
    \centering
-   \includegraphics[width=4in]{tcp-red.jpg} % requires the graphicx package
+   \includegraphics[width=4in]{figures/tcp-red.jpg} % requires the graphicx package
    \caption{TCP Reno over RED feedback control system}
@@
    \centering
-   \includegraphics[width=2in]{red.jpg} % requires the graphicx package
+   \includegraphics[width=2in]{figures/red.jpg} % requires the graphicx package
    \caption{RED dropping (or marking) function}
@@
    \centering
-   \includegraphics[width=3in]{linearize.jpg} % requires the graphicx package
+   \includegraphics[width=3in]{figures/linearize.jpg} % requires the graphicx package
    \caption{Linearization}
@@
    \centering
-   \includegraphics[width=2.5in]{vegas-control.jpg} % requires the graphicx package
+   \includegraphics[width=2.5in]{figures/vegas-control.jpg} % requires the graphicx package
    \caption{Vegas-like system}
@@
    \centering
-   \includegraphics[width=4in]{vegas-BD.jpg} % requires the graphicx package
+   \includegraphics[width=4in]{figures/vegas-BD.jpg} % requires the graphicx package
    \caption{Block diagram of Vegas-like system}
@@
    \centering
-   \includegraphics[width=1.6in]{overdamped.jpg} 
+   \includegraphics[width=1.6in]{figures/overdamped.jpg} 
    \caption{Overdamped response}
@@
    \centering
-   \includegraphics[width=2.5in]{undamped.jpg} 
+   \includegraphics[width=2.5in]{figures/undamped.jpg} 
    \caption{Undamped response}
@@
    \centering
-   \includegraphics[width=4in]{pole-loc.jpg} % requires the graphicx package
+   \includegraphics[width=4in]{figures/pole-loc.jpg} % requires the graphicx package
    \caption{Time response depending on pole location}
@@
    \centering
-   \includegraphics[width=4in]{perf-spec.jpg} % requires the graphicx package
+   \includegraphics[width=4in]{figures/perf-spec.jpg} % requires the graphicx package
    \caption{Performance Metrics}
@@
    \centering
-   \includegraphics[width=4in]{cauchy.jpg} % requires the graphicx package
+   \includegraphics[width=4in]{figures/cauchy.jpg} % requires the graphicx package
    \caption{Cauchy's principle}
@@
    \centering
-   \includegraphics[width=3in]{closed-loop.jpg} % requires the graphicx package
+   \includegraphics[width=3in]{figures/closed-loop.jpg} % requires the graphicx package
    \caption{Typical closed-loop control system}
@@
    \centering
-   \includegraphics[width=3in]{s-contour.jpg} % requires the graphicx package
+   \includegraphics[width=3in]{figures/s-contour.jpg} % requires the graphicx package
    \caption{Contour around the unstable right-side of the s-plane}
@@
    \centering
-   \includegraphics[width=3in]{ex-nyquist.jpg} 
+   \includegraphics[width=3in]{figures/ex-nyquist.jpg} 
    \caption{Example showing the effect of feedback delay}
@@
    \centering
-   \includegraphics[width=5in]{block-routing.jpg} % requires the graphicx package
+   \includegraphics[width=5in]{figures/block-routing.jpg} % requires the graphicx package
    \caption{Block diagram with both flow and routing control}
@@
    \centering
-   \includegraphics[width=4in]{routing-conv.jpg} % requires the graphicx package
+   \includegraphics[width=4in]{figures/routing-conv.jpg} % requires the graphicx package
    \caption{Convergence}
@@
    \centering
-   \includegraphics[width=3in]{contractive.jpg} % requires the graphicx package
+   \includegraphics[width=3in]{figures/contractive.jpg} % requires the graphicx package
    \caption{Contractive mapping}
@@
    \centering
-   \includegraphics[width=3in]{non-contractive.jpg} % requires the graphicx package
+   \includegraphics[width=3in]{figures/non-contractive.jpg} % requires the graphicx package
    \caption{Divergent mapping}
@@
    \centering
-   \includegraphics[width=4in]{case1-block.jpg} 
+   \includegraphics[width=4in]{figures/case1-block.jpg} 
    \caption{Block diagram of the link sharing model}
@@
      \centering
-     \includegraphics[width=4in]{case2-tunnel.jpg} 
+     \includegraphics[width=4in]{figures/case2-tunnel.jpg} 
      \caption{Soft bandwidth-guaranteed tunnel}
@@
    \centering
-   \includegraphics[width=4in]{tunnel-block1} 
+   \includegraphics[width=4in]{figures/tunnel-block1} 
 \caption{Block diagram of the elastic-tunnel model} 
@@
 \begin{figure}[tbp]
-   \includegraphics[width=5in]{tunnel-block} 
+   \includegraphics[width=5in]{figures/tunnel-block} 
 \caption{Elastic-tunnel model (with feedback delay)} 
@@
   % Requires \usepackage{graphicx}
-  \includegraphics[width=0.7\textwidth]{blockDiagram}
+  \includegraphics[width=0.7\textwidth]{figures/blockDiagram}
   }

### 2013-07-23 Compiled/ebook.tex
top-level ebook

\documentclass{book}

%styles

\usepackage[final]{pdfpages}

% from chapter 1

%\usepackage{graphicx}
%\usepackage{latexsym}
%\usepackage{amsmath}
%\usepackage{color}
%\usepackage{url}
%\hyphenation{wrap-around}

\title{Recent Advances in Networking}

\begin{document}

\maketitle

%\frontmatter

\input{preface/preface.tex}

\newpage

\input{intro/intro.tex}
\newpage

\input{intro/board.tex}
\newpage

\input{intro/chapterlist.tex}

\newpage

%\mainmatter

\includepdf[pages=-]{chapter1/chapter1.pdf}
\includepdf[pages=-]{chapter2/chapter2.pdf}
\includepdf[pages=-]{chapter3/chapter3.pdf}
\includepdf[pages=-]{chapter4/chapter4.pdf}
\includepdf[pages=-]{chapter5/chapter5.pdf}
\includepdf[pages=-]{chapter6/chapter6.pdf}
\includepdf[pages=-]{chapter7/chapter7.pdf}
\includepdf[pages=-]{chapter8/chapter8.pdf}
\includepdf[pages=-]{chapter9/chapter9.pdf}

%\backmatter

\end{document}

### 2013-07-23 Compiled/intro/board.tex
Full compiled
fixed minor things

@@
-\documentclass[a4paper,16pt]{article} 
-\usepackage{a4wide,times,latexsym}
-\usepackage[compact,tiny]{titlesec}
-\usepackage{url}
-\usepackage{times}
-\usepackage{graphics}
-\usepackage{color}
+%\documentclass[a4paper,16pt]{article} 
+%\usepackage{a4wide,times,latexsym}
+%\usepackage[compact,tiny]{titlesec}
+%\usepackage{url}
+%\usepackage{times}
+%\usepackage{graphics}
+%\usepackage{color}
 %\usepackage[pdftex]{hyperref}
-\usepackage{multirow}
+%\usepackage{multirow}
 
-\pagestyle{empty}
-\begin{document}
+%\pagestyle{empty}
+%\begin{document}
 
@@
 Srinivasan Keshav &	University of Waterloo\\
-Jean-Yves Le Boudec	& $\acute{E}$cole polytechnique f$\acute{e}$d$\acute{e}$rale de Lausanne\\
+Jean-Yves Le Boudec	& $\acute{E}$cole polytechnique F$\acute{e}$d$\acute{e}$rale de Lausanne\\
 Jennifer Rexford &	Princeton University\\
@@
 
-\end{document}
+%\end{document}

### 2013-07-23 Compiled/intro/chapterlist.tex
Full compiled
fixes

@@
-\documentclass[a4paper,16pt]{article} 
-\usepackage{a4wide,times,latexsym}
-\usepackage[compact,tiny]{titlesec}
-\usepackage{url}
-\usepackage{times}
-\usepackage{graphics}
-\usepackage{color}
-%\usepackage[pdftex]{hyperref}
-\usepackage{multirow}
+%\documentclass[a4paper,16pt]{article} 
+%\usepackage{a4wide,times,latexsym}
+%\usepackage[compact,tiny]{titlesec}
+%\usepackage{url}
+%\usepackage{times}
+%\usepackage{graphics}
+%\usepackage{color}
+%%\usepackage[pdftex]{hyperref}
+%\usepackage{multirow}
 
-\pagestyle{empty}
-\begin{document}
+%\pagestyle{empty}
+%\begin{document}
 
@@
 
-\item \textbf{Collaboration Opportunities for Content Delivery and Network \\ Infrastructures} \\
+\item \textbf{Collaboration Opportunities for Content Delivery and Network Infrastructures} \\
 Benjamin Frank, Ingmar Poese, Georgios Smaragdakis, Anja Feldmann, Bruce M. Maggs, Steve Uhlig, Vinay Aggarwal, and Fabian Schneider

### 2013-07-24 Compiled/commonstyle.tex
common latex and bibtex styles for all chapters

\usepackage{fullpage} % use entire page layout
\usepackage{times}     % smaller and better fonts that the standard latex ones
\usepackage[letterpaper]{geometry} % force letter format
\usepackage[colorlinks=true,linkcolor=blue,urlcolor=blue]{hyperref}

### 2013-07-24 Compiled/chapter2/copyright.tex
added page numbers and initial support for copyright, urls are now clickable

\usepackage[absolute,showboxes]{textpos}

%set unit to be pagewidth and height, and increase inner margin of box
\setlength{\TPHorizModule}{\paperwidth}
\setlength{\TPVertModule}{\paperheight}
\TPMargin{5pt}

%define \copyrightstatement command for easier use
\newcommand{\copyrightstatement}{
    \begin{textblock}{0.7}(0.08,0.9)    % tweak here: {box width}(leftposition, rightposition)
         \noindent
         \footnotesize
         P. Zave, J. Rexford, ``The Design Space of Network
         Mobility'', in H. Haddadi, O. Bonaventure (Eds.), \emph{Recent
         Advances in Networking}, (2013), pp. 5-45. Licensed under a
         creative commons license and available from \url{http://www.sigcomm.org}
    \end{textblock}
}

### 2013-07-24 Compiled/ebook.tex
fixes
common latex and bibtex styles for all chapters
added page numbers and initial support for copyright, urls are now clickable

@@
 %styles
+\input{commonstyle}
+
+%\hypersetup{pdftitle={Recent Advances in Networking},\
+%           pdfauthor={Hamed Haddadi and Olivier Bonaventure (Editors)}}
 
 \usepackage[final]{pdfpages}
+
+%\includepdfset{pagecommand=\thispagestyle{plain}}
 
@@
 
+\author{
+Hamed Haddadi,  Olivier Bonaventure (Editors)\\
+}
+
 \begin{document}
 
+\date{August 2013}
+
 \maketitle
+\pagenumbering{roman}
 
@@
 
+\input{intro/intro.tex}
+\newpage
+
 \input{preface/preface.tex}
 
-\newpage
-
-\input{intro/intro.tex}
 \newpage
@@
 
+\pagenumbering{arabic}
+
 %\mainmatter
 
-\includepdf[pages=-]{chapter1/chapter1.pdf}
-\includepdf[pages=-]{chapter2/chapter2.pdf}
-\includepdf[pages=-]{chapter3/chapter3.pdf}
-\includepdf[pages=-]{chapter4/chapter4.pdf}
-\includepdf[pages=-]{chapter5/chapter5.pdf}
-\includepdf[pages=-]{chapter6/chapter6.pdf}
-\includepdf[pages=-]{chapter7/chapter7.pdf}
-\includepdf[pages=-]{chapter8/chapter8.pdf}
-\includepdf[pages=-]{chapter9/chapter9.pdf}
+\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter1/chapter1.pdf}
+\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter2/chapter2.pdf}
+\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter3/chapter3.pdf}
+\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter4/chapter4.pdf}
+\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter5/chapter5.pdf}
+\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter6/chapter6.pdf}
+\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter7/chapter7.pdf}
+\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter8/chapter8.pdf}
+\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter9/chapter9.pdf}
 

### 2013-07-24 Compiled/commonstyle.tex
added page numbers and initial support for copyright, urls are now clickable
new bibliography style with hyperlinks, based on http://www.math.tamu.edu/~comech/tools/bibtex-doi-eprint/

@@
 \usepackage[letterpaper]{geometry} % force letter format
-\usepackage[colorlinks=true,linkcolor=blue,urlcolor=blue]{hyperref}
+\usepackage[colorlinks=true,linkcolor=black,urlcolor=blue,anchorcolor=black,citecolor=black,backref=none]{hyperref}
+\usepackage[usenames,dvipsnames]{color}
+%% Disable backref and make \href colors more decent:
+\definecolor{MyDarkBlue}{rgb}{0,0.1,0.7}
+\hypersetup{pdfborder={0 0 0},colorlinks,breaklinks=true,
+  urlcolor={MyDarkBlue},citecolor={MyDarkBlue},linkcolor={MyDarkBlue} }

### 2013-07-25 Compiled/chapter5/main.tex
Full compiled
common latex and bibtex styles for all chapters
added page numbers and initial support for copyright, urls are now clickable
fixes

@@
-\newcommand{\CLASSINPUTinnersidemargin}{0.7in}
-\newcommand{\CLASSINPUToutersidemargin}{0.7in}
+%\newcommand{\CLASSINPUTinnersidemargin}{0.7in}
+%\newcommand{\CLASSINPUToutersidemargin}{0.7in}
 
 \documentclass{article}
+
+\input{../commonstyle}
+
+\input{copyright}
 
@@
 
-\begin{document}
-
-\title{Recent advances in reliable transport protocols\footnote{
+\title{Recent Advances in Reliable Transport Protocols\footnote{
 Parts of the text in this chapter have appeared in the following
@@
 
-\author{}
+\author{Costin Raiciu, Janardhan Iyengar, Olivier Bonaventure}
+
+\date{}
+
+\begin{document}
+
+\pagestyle{empty}
 
 \maketitle
+
+\thispagestyle{empty}
 
@@
 
-\bibliographystyle{plain}
+%\bibliographystyle{plain}
+
+\input{../commonbibstyle}
 \bibliography{ebook,rfc}

### 2013-07-25 Compiled/chapter5/internet.tex
bibliography chapter 5 with most hyperlinks

@@
 Furthermore, some network administrators have perceived several benefits with the deployment of NATs \cite{rfc2993} including : reducing the exposure of addresses, topology hiding, independence from providers, \ldots Some of these perceived advantages have caused the IETF to consider NAT for IPv6 as well, despite the available address space.
-Furthermore, the IETF, vendors and operators are considering the deployment of large scale Carrier Grade NATs (CGN) to continue to use IPv4 despite the depletion of the addressing space \cite{Perreault_CGN:2012} and also to ease the deployment of IPv6 \cite{rfc6264}.
+Furthermore, the IETF, vendors and operators are considering the deployment of large scale Carrier Grade NATs (CGN) to continue to use IPv4 despite the depletion of the addressing space \cite{rfc6888} and also to ease the deployment of IPv6 \cite{rfc6264}.
 NATs will remain a key element of the deployed networks for the foreseeable future.
@@
 
-Firewalls are equally widespread in the Internet, being deployed by most cellular operators \cite{netpiculet}. Home routers often act as
+Firewalls are equally widespread in the Internet, being deployed by most cellular operators \cite{wang2011untol}. Home routers often act as
 firewalls, blocking new protocols and even existing ones (e.g. UDP) \cite{nat-study}. Most servers also deploy firewalls to restrict

### 2013-07-25 Compiled/chapter1/copyright.tex
first pass of references on chapter 1

--- 2013-07-24 Compiled/chapter2/copyright.tex
@@
 \newcommand{\copyrightstatement}{
-    \begin{textblock}{0.7}(0.08,0.9)    % tweak here: {box width}(leftposition, rightposition)
+    \begin{textblock}{0.65}(0.17,0.9)    % tweak here: {box width}(hposition, vposition)
          \noindent
@@
          P. Zave, J. Rexford, ``The Design Space of Network
-         Mobility'', in H. Haddadi, O. Bonaventure (Eds.), \emph{Recent
-         Advances in Networking}, (2013), pp. 5-45. Licensed under a
-         creative commons license and available from \url{http://www.sigcomm.org}
+         Mobility'', in H. Haddadi, O. Bonaventure (Eds.), \href{http://www.sigcomm.org}{\emph{Recent
+         Advances in Networking}}, (2013), pp. xx-yy. Licensed under a
+         \href{http://creativecommons.org/licenses/by-sa/3.0/}{CC-BY-SA} license. 
     \end{textblock}

### 2013-07-26 Compiled/intro/intro.tex
fix
fixes
clickable references for chapters 1 and 2

@@
+\pagestyle{empty}
+
+{\Huge \textbf{Introduction}}
+
+\vspace{1in}
+
 Professors often rely on textbooks to teach undergraduate and graduate networking courses.
@@
 We are also grateful to the reviewers and the editorial board for spending many hours on each chapter to ensure a coherent level amongst all the chapters.
-We hope you enjoy reading this eBook and find it a useful resource.
+We hope you enjoy reading this eBook and find it a useful resource.\\
 
-Hamed Haddadi, Olivier Bonaventure
+Hamed Haddadi, Olivier Bonaventure \\
 
-August 2013
+\begin{figure}
+\begin{center}
+\includegraphics[width=3cm]{intro/sigcomm-logo-new}
+%\caption{\label{kmeans} {ACM SIGCOMM}}\vspace{-3mm}
+\end{center}
+\end{figure}
+%August 2013 \\
+\vspace{1in}
+Hamed Haddadi, Olivier Bonaventure, ``Recent Advances in Networking", Volume 1, ACM SIGCOMM eBook, August 2013.

### 2013-07-26 Compiled/chapter9/Content/flood_rtg.tex
DOIs updated in chapt 9

@@
 %
-The {\em Mobile Ad Hoc Network Neighborhood Discovery Protocol} (NHDP, RFC 6130) \cite{NHDP-RFC6130} is the main ND protocol for spontaneous wireless networks.
+The {\em Mobile Ad Hoc Network Neighborhood Discovery Protocol} (NHDP, RFC 6130) \cite{rfc6130} is the main ND protocol for spontaneous wireless networks.
 It is used as auxiliary protocol by other routing protocols that need neighborhood information to take their decisions, such as OLSRv2 (see section \ref{sec:olsr}).
@@
 Other routing protocols include their own Hello mechanism.
-This is the case of AODV \cite{AODV-RFC3561} (see section \ref{ss:aodv}) or OSPF and its MANET extensions \cite{rfc5449,rfc5614,rfc5820} (see section \ref{ss:ospfmanet}).
+This is the case of AODV \cite{rfc3561} (see section \ref{ss:aodv}) or OSPF and its MANET extensions \cite{rfc5449,rfc5614,rfc5820} (see section \ref{ss:ospfmanet}).
 In the last case, some approches have been explored in order to avoid redundant notifications and hence reduce control traffic by only reporting changes in the neighborhood occurred since the last Hello transmission: this principle leads to the {\em incremental Hellos} mechanism used in the Overlapping Relays extension of OSPF (OR/SP \cite{rfc5820}) and the {\em differential Hellos} mechanism used in the MANET Designated Routers extension of OSPF (OSPF-MDR \cite{rfc5614}).
@@
 This effect could be alleviated by allowing routers to wait a random amount of time (denominated {\bf jitter}) before retrasmitting a flooded message, in order to reduce the probability of concurrent transmissions by neighboring wireless interfaces.
-This technique is known as {\bf jittering}, and has been standardized by the IETF in RFC 5148 \cite{jitter-RFC5148}.
+This technique is known as {\bf jittering}, and has been standardized by the IETF in RFC 5148 \cite{rfc5148}.
 The recommendation from RFC 5148 is that delays are selected following an uniform distribution between $0$ and a maximum jitter value, $MAX\_JITTER$.
@@
 Other jittering effects are due to the specifities of the techniques employed in each case.
-In proactive protocols where jitter is used (as described in RFC 5148 \cite{jitter-RFC5148}, which recommends to introduce random delays and piggyback all pending messages when a transmission is scheduled), such as OLSR or the OSPF extensions for MANETs, jittering leads to longer LSA messages -- this may cause additional packet collisions, if jitter values are not configured properly \cite{INFOCOM-jitter,jitter-wpc}.
+In proactive protocols where jitter is used (as described in RFC 5148 \cite{rfc5148}, which recommends to introduce random delays and piggyback all pending messages when a transmission is scheduled), such as OLSR or the OSPF extensions for MANETs, jittering leads to longer LSA messages -- this may cause additional packet collisions, if jitter values are not configured properly \cite{INFOCOM-jitter,jitter-wpc}.
 In reactive protocols such as AODV, where jitter is used for Route Request (RREQ) flooding, the addition of random delays may lead to suboptimal path selections, which can be minimized by adapting the random distribution used for determining jitter values \cite{Infocom12, wiopt}.

### 2013-07-26 Compiled/chapter9/Content/fundamentals.tex
DOIs updated in chapt 9

\section{Fundamentals of IP Networking and Internet Routing}
\label{s:fundamentals}

This section introduces the main ideas and concepts that are used as a basis for traditional wired Internet. Subection \ref{ss:ip} presents the key elements of the IP networking model, including addressing, forwarding and the notion of IP link. Subection \ref{ss:routing} describes the most relevant routing techniques used in the Internet, and subsection \ref{ss:irarq} overviews the Internet routing architecture, based on the notion of Autonomous System. A certain familiarity with the basics of computer networking is assumed, so no details are provided. This section mainly follows the classic manuals of Tanenbaum {\em et al.} \cite{tanenbaum}, Comer \cite{comer} and Perlman \cite{interconnections}. Interested readers are referred to these resources for further explanations. \ \\ \ \\
% Assumptions:
% - TCP/IP stack
%
%This chapter focuses on communication in layer 3 (internetworking) for multi-hop IP networks (partly) composed of wireless links. Protocols and devices operating at this layer enable communication between hosts not belonging to the same link. 
%
\subsection{The IP Networking Model}
\label{ss:ip}
%
The Internet Protocol (IP) defines the key elements enabling communication in an IP network. This section presents the IP addressing mechanism, the notion of IP link and the routing rule used by router in IP networks -- the {\em longest prefix match} criterion.
%
\paragraph{Addressing}

In an IP network, every \glslink{interface}{\em network interface} is assigned at least one {\bf IP address} that identifies unambiguously the \glslink{interface}{interface} in the network. The IP address format varies depending on the protocol version (32 bits for IPv4, 128 bits for IPv6, see Figure \ref{f:ipa}), but three elements  can be distinguished.

\begin{itemize}
\item The {\bf host identifier} is the set of bits that identifies the \glslink{interface}{interface} in the network.
\item The {\bf network prefix} is the set of bits that identifies the network to which the \glslink{interface}{interface} is attached.
\item The {\bf network mask} allows to obtain the network prefix and the host identifier from the IP address. 
\end{itemize}

\begin{remark}
The IP address of a network \glslink{interface}{interface} is both an {\em identifier} and a {\em locator} of the interface: it indicates {\em who} is (unambiguosly in the internetwork) the attached \glslink{interface}{interface} and {\em where} is it attached (to which network).
\end{remark}

\begin{figure}[htb]
\centering
{\small a) IPv4 addressing example: $192.168.0.1/24$}
\begin{eqnarray*}
\textrm{Mask (24):} 		& \underbrace{{\tt 11111111}.{\tt 11111111}.{\tt 11111111}}_{\textrm{netmask (24 bits)}}.{\tt 00000000} \\
\textrm{IP address:} 	& \underbrace{{\tt 11000000}.{\tt 10101000}.{\tt 00000000}}_{\textrm{network prefix}}.\underbrace{{\tt 00000001}}_{\textrm{host identifier}}
\end{eqnarray*}
{\small b) IPv6 addressing example: $2001:0DB8:02DE::0E13/64$}
\begin{eqnarray*}
\textrm{Mask (64):} 			& \underbrace{{\tt FFFF}:{\tt FFFF}:{\tt FFFF}:{\tt FFFF}}_{\textrm{netmask (64 bits)}}:{\tt 0000}:{\tt 0000}:{\tt 0000}:{\tt 0E13} \\
\textrm{IP address:} 	& \underbrace{{\tt 2001}:{\tt 0DB8}:{\tt 02DE}:{\tt 0000}}_{\textrm{network prefix}}:\underbrace{{\tt 0000}:{\tt 0000}:{\tt 0000}:{\tt 0E13}}_{\textrm{host identifier}}
\end{eqnarray*}
\caption{IP address structure, for IPv4 and IPv6.}
\label{f:ipa}
\end{figure}
%
%\begin{remark}
%In the context of an IP network, a routing table is a local database that maps IP prefixes (from packet destinations) to the networking \glslink{interface}{interface} and the IP address of the next hop (to which the packet should be sent).
%\end{remark}
%
Based on the information contained in IP addresses from the destination field of the IP header, routers and hosts are able to take decisions upon reception of an IP packet. Trivially, a host receiving an IP packet will accept it only in case that the destination IP address is itself\footnote{Or the destination address is a broadcast address or a multicast address to which the host has suscribed.} and drop it otherwise. A router receiving an IP packet over an \glslink{interface}{interface} will compare the network prefix of the destination IP address with the prefix of its own interface: if it does not match, it may forward it through another interface, according to the IP forwarding rule (see below). In case of forwarding, the router decreases the {\em Time-To-Live} field (or {\em hop-limit} for IPv6), to indicate that the corresponding packet has traversed one (more) router in its path to its destination. This leads to the notion of \glslink{ip-link}{\em IP link} (see Figure \ref{f:iplink}).
%
\begin{definition}[IP Link]
Two network interfaces, $x$ and $y$, are connected to the same {\em IP link} when they can exchange packets in an IP network without requiring that any router forwards them, that is, when packets sent from one \glslink{interface}{interface} are received in the other with the same TTL/hop-limit value. This relationship is denoted as $x \sim_{IP} y$.
\end{definition}
%
\begin{itemize}
\item In these conditions, communication is performed in a single {\em IP hop}.
\end{itemize}

%Let $\sim_{IP}$ denote the IP link relationship, by which $x \sim_{IP} y$ if and only if network interfaces $x$ and $y$ belong to the same IP link, and let $a$, $b$ and $c$ be network interfaces. Definition~(\ref{d:ip_link}) implies the following properties of IP links:

\begin{remark}
Let $a$, $b$ and $c$ be network interfaces. The previous definition implies the following properties of IP links:

\begin{itemize}
\item \emph{Symmetry}: $a \sim_{IP} b \Longleftrightarrow b \sim_{IP} a$.
\item \emph{Transitivity}: $a \sim_{IP} b, b \sim_{IP} c \Longrightarrow a \sim_{IP} c$.
\end{itemize}
\end{remark}

Note that transitivity does {\em not} hold in terms of routers. The fact that a router $R_1$ and a router $R_2$ are connected to the same link, and $R_2$ and $R_3$ are connected to the same link, does not imply that $R_1$ and $R_2$ have a link in common: $R_2$ may be attached to two different links (one connecting with $R_1$ and another with $R_3$) by way of two different network interfaces.

\begin{figure}[htb]	% H-must be here or [htb]
\centering
\includegraphics[width=0.5\textwidth]{Figures/iplink-crop.pdf} %	** if .eps or .pdf, don't need extension}
\caption{An IP link {\tt p:} with network prefix $p$. \emph{IP addresses of nodes in this IP link have the structure {\tt p:i}$/[p]$, for $0 < i < 2^{[p]}$.}}
\label{f:iplink}
\end{figure}

\paragraph{Forwarding rule}

When source and destination of a packet do not belong to the same IP link, routers receiving the packet compare the IP address of the destination to the prefixes stored in the routing table, and forward the packet through the network \glslink{interface}{interface} corresponding to the prefix showing the {\em longest prefix match}, this is, the prefix in the routing table for which a bigger number of bits are coincident with those from the network prefix of the IP address of the packet destination. %This is the {\em closest} prefix towards the destination, according to the IP closeness notion and the IP ordering criterion over the addressing space induced by the definition of IP link.

%\begin{definition}[IP Partial Order]
%Given two IP addresses $IPA_1/m_1$ and $IPA_2/m_2$ ($m_i$ being the prefix length of IP address $IPA_i$), $IPA_1 \subseteq_{IP} IPA_2$ if and only if:
%\begin{enumerate}[(i)]
%\item $IPA_1 \otimes NM_{\textrm{ max}\{m_1, m_2\}} = IPA_2 \otimes NM_{\textrm{ max}\{m_1, m_2\}}$ 
%\item $m_1 \geq m_2$
%\end{enumerate}
%where $NM_k$ is the netmask of $k$ bits and $\otimes$ denotes the bitwise AND operation.
%\end{definition}

%\begin{remark}
%The relationship $\subseteq_{IP}$ satisfies trivially the axioms of partial order:
%	\begin{itemize}
%	\item \emph{Reflexivity}: $IP_a \subseteq_{IP} IP_a$.
%	\item \emph{Antisymmetry}: $IP_a \subseteq_{IP} IP_b, IP_b \subseteq_{IP} IP_a \Longrightarrow IP_a \sim_{IP} IP_b$, that is, $IP_a$ and $IP_b$ are in the same IP link.
%	\item \emph{Transitivity}: $IP_a \subseteq_{IP} IP_b, IP_b \subseteq_{IP} IP_c \Longrightarrow IP_a \subseteq_{IP} IP_c$.
%	\end{itemize}
%\end{remark}

%For routing of packets for which the IP link of the destination is not the same as the IP link of the source, the IP addressing model provides a simple rule for making forwarding decisions. Given the IP address of the packet’s destination, a router should forward the packet through the \glslink{interface}{interface} providing connection to the IP network closest to the destination, where the notion of {\em closeness} is as follows:

%\begin{definition}[IP Closeness]
%Given an IP address $IPA_d/m_d$ and two IP addresses $IPA_1/m_1$ and $IPA_2/m_2$, $IPA_1/m_1$ is \emph{IP-closer} to $IPA_d/m_d$ than to $IPA_2/m_2$ if:
%	\begin{itemize}
%	\item $IPA_d \subseteq_{IP} IPA_1$ and $IPA_d \not\subseteq_{IP} IPA_2$, or
%	\item $IPA_d \subseteq_{IP} IPA_1 \subseteq_{IP} IPA_2$ (which is equivalent to $|m_1| \geq |m_2|$, for the case $IPA_d \subseteq_{IP} IPA_{1,2}$).
%	\end{itemize}
%\end{definition}

%According to this decision criterion, routers select send a given IP packet to the \glslink{interface}{interface} whose IP address has the \emph{longest prefix match} with respect to the IP packet destination.

%\begin{remark}
%Note that the longest prefix match is only guaranteed to exist if and only if the router has a \emph{default route} ({\tt 0.0.0.0/0} in IPv4, {\tt ::/0} in IPv6).
%\end{remark}

\subsection{Main Routing Techniques}
\label{ss:routing}
%
Two types of routing techniques currently dominate \cite{interconnections}: {\em link-state} routing and {\em distance-vector} routing (with the variant of {\em path-vector} routing). The main protocols used historically and currently in the Internet are based on these techniques.
%
\paragraph{Link-State Routing} Routers advertise the status of their links (link-state) to the whole network. Link status may include information about the type of link (broadcast, point-to-point...), the link communication capabilities (one-directional, bi-directional, link cost) or the routers to which communication is available through this link. This way, every router in the network receives the link-state of other routers in the network, maintains information about the whole network topology and is therefore able to locally compute network-wide shortest paths, usually by way of Dijkstra's algorithm \cite{dijkstra59}. 
	\begin{itemize}
	\item Some examples of this approach are the Open Shortest Path First (OSPF, RFCs 2328 and 5340 \cite{rfc2328,rfc5340}) and the Intermediate System to Intermediate System (IS-IS, RFC 1142 \cite{rfc1142}) protocols, as well as the Optimized Link State Routing protocol (OLSR, RFC 3626 \cite{rfc3626}).
	\end{itemize}

\paragraph{Distance-Vector Routing} A router shares information from its routing table only with its neighbors, indicating \emph{distances} and next hops towards reachable destinations. Neighbor distance is defined according to the current \glslink{metric}{\em link metric}, which maps links between routers with estimations of the cost of sending packets through them, represented by scalar values. By receiving the routing tables of all its neighbors, which in turn have been shared with the neighbors of the neighbors, a router is able to identify, for each advertised destination, the neighbor that provides shortest distance and select it as next hop. Distance-vector protocols mostly use the distributed Bellman-Ford algorithm \cite{bellman58,ford56} to identify network-wide shortest paths. 
	\begin{itemize}
	\item The Routing Information Protocol (RIP, RFCs 1058 \cite{rfc1058}, 2080 \cite{rfc2080} and 2453 \cite{rfc2453}) is a prominent example of this family.
	\end{itemize}

\paragraph{Path-vector routing} It is based on the same principle as distance-vector routing, a router advertises to its neighbors the paths to all reachable destinations. Each path is described by indicating the routers that are traversed. This way, local distribution of locally maintained paths enables all routers in the network to build routes to all possible destinations. 
	\begin{itemize}
	\item The most prominent example of this family of protocols is the Border Gateway Protocol (BGP, RFC 1771 \cite{rfc1771}). \\
	\end{itemize}

% -------------
The link-state algorithm requires that every single router has storage and computational capacity to compute locally the shortest-path tree of the network, based on the information received from every other routers, and extract from that tree the next-hop towards every destination in the network. Distance-vector algorithms only require that each router updates the distance-vectors received from their neighbor to infer its own vector of distance vector and select its next-hops. \ \\ \ \\
%
Due to their computational simplicity, distance-vector protocols were used in the early stages of the Internet. They were gradually replaced by link-state protocols as the ARPANET grew bigger and more complex, due to problems such as the well-known count-to-infinity problem \cite{tanenbaum} (which appears in the original distance-vector algorithm, but does not appear on path-vector protocols). Poor scalability and slow convergence properties of distance-vector with respect to link-state algorithms were also major reasons to switch from one technique to the other \cite{interconnections}.

\begin{itemize}
\item The network reaction to a link failure illustrates the differences between link-state and distance-vector algorithms in terms of convergence. In distance-vector algorithms, once a router detects such a failure, it updates the cost of its route towards the lost neighbor and sends the new vector of distances to its neighbors. Neighbors receive this update and {\em recompute} the cost of the affected route, and then transmit in turn their new vectors. Propagation of topology changes is thus slower than in link-state algorithms, in which a router detecting the failure of the link towards one of its neighbors {\em floods} an updated topology description which is directly forwarded over the network, without delays caused by route re-computation in intermediate routers \cite{interconnections}.
\end{itemize}
% ------------------

Routing protocols for wired networks used to be {\em proactive} or table-driven, in which next hop to any possible destination is stored in a table.  With the emergence of wireless networks and, more generally, more dynamic networking architectures coping with more scarce (shared) bandwith, {\em reactive} routing protocols were then designed and deployed, in which routes were only computed upon request (on-demand).

\paragraph{Proactive routing} Routers collect and periodically disseminate topology information over the network; this enables them to maintain proactively (\emph{i.e.}, regardless on whether they are used) routes towards all destinations. This way, routers are able to forward packets at any time to any destination in the network.

\paragraph{Reactive routing} A router calculates a route to a destination only when it receives packets addressed to that destination and the routing table does not provide a next hop towards it. In this case, the router triggers a {\em route discovery} process by disseminating a Route Request (RREQ) packet through the network. The route discovery process terminates when the requested destination or another router knowing a valid route towards the destination reply to the requesting router.  
	\begin{itemize}
	\item Dynamic Source Routing (DSR, RFC 4728 \cite{rfc4728}) or Ad hoc On-Demand Distance Vector (AODV, RFC 3561 \cite{rfc3561}), both used in spontaneous wireless networks, are examples of reactive routing protocols. \\
	\end{itemize}

Proactive maintenance of next hops to every possible destination in the table requires a constant exchange of control traffic in the network, but enables routers to forward packets immediately after receiving them. Reactive protocols adapt the control traffic to the data traffic requirements: when there is no traffic to route, or the traffic follows known paths, a mostly negligible amount of control traffic (very low or even zero, depending on the protocol) is needed. When a router receives packets to be sent to a destination for which no route is known, the router needs to address a route discovery process over the network -- such a discovery process is costly in terms of overhead, and leads to significant delays in the forwarding.

\paragraph{Other routing approaches} Some other approaches have been explored for routing over spontaneous wireless networks. In some cases, they rely on additional assumptions about properties and capabilities of the involved devices. If nodes' position is available (for instance, by way of GPS), {\bf geographical routing} approaches are possible: in these protocols, a packet is forwarded to the relay getting closer to the final destination. The Greedy Perimeter Stateless Routing (GPSR) protocol \cite{gpsr} was the first protocol exploring this principle.

%\begin{tabular}{|c||c|c|}
%\hline
%    		& Proactive 		& Reactive  	\\ \hline\hline
%Link-State 	& OSPF, IS-IS, OLSR 	& 		\\ \hline
%Distance-Vector	& RPL, RIP, BGP 	& AODV, LOAD	\\ \hline
%\end{tabular}

%\subsubsection{Other Approaches}

%\begin{itemize}
%\item Opportunistic routing. ExOR.
%\item Source routing. DSR.
%\end{itemize}

\subsection{The Internet Routing Architecture}
\label{ss:irarq}

In terms of routing, the Internet is organized as a set of interconnected internetworks, denominated {\em Autonomous Systems} (see Figure \ref{f:ass}). The networks in each Autonomous System are under the same administrative control, and are assumed to perform routing inside the AS {\em autonomously} from the rest of networks in the Internet. The formal definition of an AS is as follows:

\begin{definition}[Autonomous System]
``An \emph{Autonomous System (AS)} is a connected group of one or more IP prefixes [internetwork] run by one or more network operators which has a SINGLE and CLEARLY DEFINED routing policy" \cite{rfc1930}, the term ``routing policy" denoting the way that routing information is exchanged between (but not within) Autonomous Systems. In the interior of an AS, ``routers may use one or more interior routing protocols, and sometimes several sets of \glslink{metric}{metrics}" \cite{rfc1812}.
\end{definition}

\begin{figure}[htb]	% H-must be here or [htb]
\centering
\includegraphics[width=0.8\textwidth]{Figures/ass-crop.pdf} %	** if .eps or .pdf, don't need extension}
\caption{Connection of different Autonomous Systems.}
\label{f:ass}
\end{figure}

The distinction between routing inside an Autonomous System (intra-AS or intra-domain routing) and routing between different ASes (inter-AS or inter-domain routing) leads to two different types of routing protocols:

\begin{enumerate}[(i)]
\item {\em Interior Gateway Protocols} (IGPs), for route discovery and maintenance within an Autonomous System. Intra-domain routing is mostly performed by way of link-state protocols; the most significant link-state routing protocol for TCP/IP networks in the Internet are the Open Shortest Path First protocol (OSPF, \cite{rfc2328, rfc5340}, described in section \ref{s:wospf}) and the Integrated IS-IS, an IP variant of the Intermediate Systems to Intermediate Systems (IS-IS) protocol (see \cite{rfc1195}).
\item {\em Exterior Gateway Protocols} (EGPs), for route acquisition and information exchange between different Autonomous Systems. The current standard protocol for inter-domain routing is the path-vector Border Gateway Protocol (BGP, \cite{rfc1771}).
\end{enumerate}

### 2013-07-26 Compiled/chapter9/Content/conclusion.tex
fix typo

@@
 Essentially, such an approach considers the Internet as a black box providing a service equivalent to a cable connecting source(s) and destination(s), and provides novel mechanisms efficiently using this cable to cope with IP-disruptive characteristics.
-One example of such construction is the experimental Delay Tolerant Networking (DTN) architecture developed by the Internet Research Task Force (IRTF) \cite{DTNRG} \cite{ietf:rfc5050} \cite{ietf:rfc5326}, including a specific routing protocol targeting DTNs \cite{ietf:rfc6693}.\ \\ \ \\
+One example of such construction is the experimental Delay Tolerant Networking (DTN) architecture developed by the Internet Research Task Force (IRTF) \cite{DTNRG} \cite{rfc5050} \cite{rfc5326}, including a specific routing protocol targeting DTNs \cite{rfc6693}.\ \\ \ \\
 %

### 2013-07-26 Compiled/chapter4/sections/Collaboration-Use-Cases.tex
Fix compilation of chapter4 with new pdf files for figures

@@
    \begin{subfigure}[]{0.4\linewidth}
-     {\includegraphics[angle=-90,width=1\linewidth]{figures/snapshot_nond.eps}}
+     {\includegraphics[angle=-90,width=1\linewidth]{figures-pdf/snapshot_nond}}
    \caption{Unbiased Gnutella}%
@@
    \begin{subfigure}[]{0.45\linewidth}
-     {\includegraphics[angle=-90,width=1\linewidth]{figures/snapshot_nd.eps}}
+     {\includegraphics[angle=-90,width=1\linewidth]{figures-pdf/snapshot_nd}}
    \caption{Gnutella with Oracle}%
@@
 	\begin{subfigure}[]{0.32\linewidth}
-     {\includegraphics[width=1\textwidth]{figures/asdist.eps}}
+     {\includegraphics[width=1\textwidth]{figures-pdf/asdist}}
     \caption{Mean AS distance in Underlay\label{fig:oracle:gnutella_metric:2a}}
@@
     \begin{subfigure}[]{0.32\linewidth}
-    {\includegraphics[width=1\textwidth]{figures/intraAS-leaf.eps}}
+    {\includegraphics[width=1\textwidth]{figures-pdf/intraAS-leaf}}
     \caption{Intra-AS peerings (\%) for Leaf nodes\label{fig:oracle:gnutella_metric:2b}}
@@
     \begin{subfigure}[]{0.32\linewidth}
-    {\includegraphics[width=1\textwidth]{figures/intraAS-up.eps}}
+    {\includegraphics[width=1\textwidth]{figures-pdf/intraAS-up}}
     \caption{Intra-AS peerings (\%) for Ultrapeers\label{fig:oracle:gnutella_metric:2c}}
@@
 \begin{figure}[tbp]
-  \center\includegraphics[width=0.8\linewidth]{figures/trafficShift-concept.eps}
+  \center\includegraphics[width=0.8\linewidth]{figures-pdf/trafficShift-concept}
   \caption{By choosing a CDI server for a client with the help of CaTE, traffic
@@
 \begin{figure}[tbp]
-    \center \includegraphics[width=3.2in]{figures/CaTE_Schematic.eps}
+    \center \includegraphics[width=3.2in]{figures-pdf/CaTE_Schematic}
   \vspace{-0.1in}
@@
 \begin{figure}[tbp]
-\center\includegraphics[width=0.90\linewidth]{figures/trafficShift-illustration.eps}
+\center\includegraphics[width=0.90\linewidth]{figures-pdf/trafficShift-illustration}
 \caption{Content-aware Traffic Engineering Process}
@@
 \begin{figure}[tbp]
-\center\includegraphics[width=1\linewidth]{figures/providersVsVolume.eps}
+\center\includegraphics[width=1\linewidth]{figures-pdf/providersVsVolume}
 \caption{CDF of traffic volume of CDIs in ISP1.}
@@
 \begin{figure}[tbp]
-\center\includegraphics[width=1\linewidth]{figures/Volume_timeseries.eps}
+\center\includegraphics[width=1\linewidth]{figures-pdf/Volume_timeseries}
 \caption{Normalized traffic for top CDIs by volume in ISP1.}
@@
 \begin{figure}[tbp]
-\center\includegraphics[width=1\linewidth]{figures/subnetVsVolumeBucket.eps}
+\center\includegraphics[width=1\linewidth]{figures-pdf/subnetVsVolumeBucket}
 \caption{Subnet diversity from which content is available.}
@@
 \center
-\includegraphics[width=1\linewidth]{figures/SubnetChoice_r-2.eps}
+\includegraphics[width=1\linewidth]{figures-pdf/SubnetChoice_r-2}
 \caption{Evolution over time of number of subnets for selected CDIs in the top 10 CDIs.}

### 2013-07-26 Compiled/chapter4/sections/Traffic-Trends-Server-Diversity.tex
Fix compilation of chapter4 with new pdf files for figures

@@
 
-\begin{figure}[tbp]
+\begin{figure}[htbp]
   \centering
-  \includegraphics[height=0.7\linewidth]{figures/dns-diversity.eps}
+  \includegraphics[height=0.7\linewidth]{figures-pdf/dns-diversity}
   \caption{DNS replies for two different sites hosted on a CDI, in two-hour bins}%
@@
 
-\begin{figure}[tbp]
+\begin{figure}[htbp]
    \center
-   \includegraphics[height=0.7\linewidth,angle=-90]{figures/avgResponseSize.ps}\\
-   \includegraphics[height=0.7\linewidth,angle=-90]{figures/avgResponseSubnetSize.ps}
+   \includegraphics[height=0.7\linewidth]{figures-pdf/avgResponseSize}\\
+   \includegraphics[height=0.7\linewidth]{figures-pdf/avgResponseSubnetSize}
 
@@
 
-\begin{figure}[tbp]
+\begin{figure}[htbp]
   \center
-   \includegraphics[height=0.7\linewidth,angle=-90]{figures/14day-returnedIPCDF-bytes.ps}
-   \includegraphics[height=0.7\linewidth,angle=-90]{figures/14day-returnedIPCDF-hits.ps}
+   \includegraphics[height=0.7\linewidth]{figures-pdf/14day-returnedIPCDF-bytes}
+   \includegraphics[height=0.7\linewidth]{figures-pdf/14day-returnedIPCDF-hits}
   \caption{CDF of \# of IPs for the ISP DNS resolver normalized by traffic
@@
 
-\begin{figure}[tbp]
+\begin{figure}[htbp]
   \center
-   \includegraphics[height=0.7\linewidth,angle=-90]{figures/SubnetsByBytes.ps}
-   \includegraphics[height=0.7\linewidth,angle=-90]{figures/SubnetsByHits.ps}
+   \includegraphics[height=0.7\linewidth]{figures-pdf/SubnetsByBytes}
+   \includegraphics[height=0.7\linewidth]{figures-pdf/SubnetsByHits}
   \caption{CDF of \# of subnets for ISP DNS resolver normalized by traffic
@@
 the ISP to bias the location preference of the CDI.
-
-\begin{figure}[tbp]
-  \center\includegraphics[height=0.7\linewidth,angle=-90]{figures/maxTTLCDF.ps}
+\clearpage
+\begin{figure}[htbp]
+  \center\includegraphics[height=0.7\linewidth]{figures-pdf/maxTTLCDF}
   \caption{CDF of DNS TTL value by traffic volume and by number of requests.}

### 2013-07-26 Compiled/chapter1/copyright.tex
copyright boxes in all chapters

@@
          P. Zave, J. Rexford, ``The Design Space of Network
-         Mobility'', in H. Haddadi, O. Bonaventure (Eds.), \href{http://www.sigcomm.org}{\emph{Recent
+         Mobility'', in H. Haddadi, O. Bonaventure (Eds.), \href{\ebookurl}{\emph{Recent
          Advances in Networking}}, (2013), pp. xx-yy. Licensed under a
-         \href{http://creativecommons.org/licenses/by-sa/3.0/}{CC-BY-SA} license. 
+         \href{http://creativecommons.org/licenses/by-sa/3.0/}{CC-BY-SA}
+         Creative Commons license. 
     \end{textblock}

### 2013-07-26 Compiled/chapter2/copyright.tex
copyright boxes in all chapters

@@
 \newcommand{\copyrightstatement}{
-    \begin{textblock}{0.7}(0.08,0.9)    % tweak here: {box width}(leftposition, rightposition)
+    \begin{textblock}{0.65}(0.17,0.9)    % tweak here: {box width}(hposition, vposition)
          \noindent
          \footnotesize
-         P. Zave, J. Rexford, ``The Design Space of Network
-         Mobility'', in H. Haddadi, O. Bonaventure (Eds.), \emph{Recent
-         Advances in Networking}, (2013), pp. 5-45. Licensed under a
-         creative commons license and available from \url{http://www.sigcomm.org}
+         I. Matta, ``Optimizing and Modeling Dynamics in Networks'', in H. Haddadi, O. Bonaventure (Eds.), \href{\ebookurl}{\emph{Recent
+         Advances in Networking}}, (2013), pp. xx-yy. Licensed under a
+         \href{http://creativecommons.org/licenses/by-sa/3.0/}{CC-BY-SA}
+         Creative Commons license. 
     \end{textblock}

### 2013-07-28 Compiled/chapter1/ebook.tex
Full compiled
common latex and bibtex styles for all chapters
added page numbers and initial support for copyright, urls are now clickable
fixes

--- 2013-07-22 originals/chapter1/ebook.tex
@@
 \documentclass{article}
+
+\input{../commonstyle}
+
+\input{copyright}
+
 \usepackage{graphicx}
@@
 \usepackage{url}
+\usepackage{hyperref}
+\usepackage{fullpage}
 \hyphenation{wrap-around}
@@
 \author{
-Pamela Zave \\
-AT\&T Laboratories---Research \\
-Florham Park, New Jersey, USA \\
-{\tt pamela@research.att.com}
+\href{http://www2.research.att.com/~pamela/}{Pamela Zave} \\
+%AT\&T Laboratories---Research \\
+%Florham Park, New Jersey, USA \\
+%{\tt pamela@research.att.com}
 \and
-Jennifer Rexford \\
-Princeton University \\
-Princeton, New Jersey, USA \\
-{\tt jrex@cs.princeton.edu}
+\href{http://www.cs.princeton.edu/~jrex/}{Jennifer Rexford} \\
+%Princeton University \\
+%Princeton, New Jersey, USA \\
+%{\tt jrex@cs.princeton.edu}
 }
 
+\date{}
 \begin{document}
+
+\copyrightstatement
+
+\pagestyle{empty}
+
+\maketitle
+
 \thispagestyle{empty}
-\maketitle
 
@@
 
-\bibliographystyle{IEEEtran}
+%\bibliographystyle{IEEEtran}
+\input{../commonbibstyle}
 \bibliography{ebook}

### 2013-07-28 Compiled/chapter6/main.tex
Full compiled
common latex and bibtex styles for all chapters
added page numbers and initial support for copyright, urls are now clickable
copyright boxes in all chapters
fixes

@@
 \documentclass{article}
-\usepackage{amsfonts,amsmath,amssymb}
+\input{../commonstyle}
+
+\input{copyright}
+
+\usepackage{amsmath,amssymb}
 \interdisplaylinepenalty=2500
@@
 \usepackage{xspace}
+\usepackage{hyperref}
 \usepackage{url}
@@
 
-\author{Luca Cittadini \and Giuseppe Di Battista \and Maurizio Patrignani}
-
-\date{~}
+\author{\href{http://www.dia.uniroma3.it/~compunet/www/view/person.php?id=cittadini}{Luca Cittadini} \and \href{http://www.dia.uniroma3.it/people/gdb/}{Giuseppe Di Battista} \and \href{http://www.dia.uniroma3.it/~patrigna/}{Maurizio Patrignani}}
+
+\date{}
+
+\copyrightstatement
+
+\pagestyle{empty}
 
 \maketitle
+
+\thispagestyle{empty}
 
@@
 of MPLS VPNs can be gained by reading Sections~\ref{se:details} and~\ref{se:advanced}.
-
-\newpage
-
-\tableofcontents
-
-\newpage
 
@@
 
-\bibliographystyle{plain}
+%\bibliographystyle{plain}
+
+\input{../commonbibstyle}
 \bibliography{bibliography}

### 2013-07-28 Compiled/ebook.tex
fixes
fix
added c4 figs

@@
 \usepackage[final]{pdfpages}
+\usepackage{url}
+\usepackage{hyperref}
+\usepackage{titlepic}
 
@@
 
-\title{Recent Advances in Networking}
+\title{\href{http://www.sigcomm.org/content/ebook}{Recent Advances in Networking}}
 
 \author{
-Hamed Haddadi,  Olivier Bonaventure (Editors)\\
+\href{http://www.eecs.qmul.ac.uk/~hamed/}{Hamed Haddadi},  \href{http://www.uclouvain.be/olivier.bonaventure}{Olivier Bonaventure} (Editors)\\
+\\
+\\
+\href{http://www.sigcomm.org/}{ACM SIGCOMM}
 }
+
+\titlepic{\includegraphics[width=3cm]{intro/sigcomm-logo-new}}
 
@@
 
-\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter1/chapter1.pdf}
+\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter7/chapter7.pdf}
+\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter5/chapter5.pdf}
+\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter8/chapter8.pdf}
 \includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter2/chapter2.pdf}
 \includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter3/chapter3.pdf}
-\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter4/chapter4.pdf}
-\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter5/chapter5.pdf}
 \includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter6/chapter6.pdf}
-\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter7/chapter7.pdf}
-\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter8/chapter8.pdf}
+\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter4/paper.pdf}
+\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter1/chapter1.pdf}
 \includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter9/chapter9.pdf}

### 2013-08-02 Compiled/intro/intro.tex
dd logo:
fix
added c4 figs
fixes

@@
 While there are many good introductory textbooks, there are very few books on advanced networking topics that could be suitable to graduate courses in networking.
-To fill this gap, SIGCOMM Education Committee has launched a community project to develop a high-quality, open-source, edited eBook on “Recent Advances in Networking”.
+To fill this gap, SIGCOMM Education Committee has launched a community project to develop a high-quality, open-source, edited eBook on ``Recent Advances in Networking".
 This eBook will be distributed online via the SIGCOMM website.
@@
 
-\begin{figure}
-\begin{center}
-\includegraphics[width=3cm]{intro/sigcomm-logo-new}
-%\caption{\label{kmeans} {ACM SIGCOMM}}\vspace{-3mm}
-\end{center}
-\end{figure}
+%\begin{figure}
+%\begin{center}
+%\includegraphics[width=3cm]{intro/sigcomm-logo-new}
+%%\caption{\label{kmeans} {ACM SIGCOMM}}\vspace{-3mm}
+%\end{center}
+%\end{figure}
 %August 2013 \\
 \vspace{1in}
-Hamed Haddadi, Olivier Bonaventure, ``Recent Advances in Networking", Volume 1, ACM SIGCOMM eBook, August 2013.
+Hamed Haddadi and Olivier Bonaventure (editors), ``Recent Advances in Networking", Volume 1, ACM SIGCOMM eBook, August 2013.

### 2013-08-06 Compiled/chapter7/router.tex
Full compiled
links for chapter 7

@@
-\clearpage
+%\clearpage
 \section{Router-level topology}
@@
 
-\clearpage
+%\clearpage
 {\bf Lesson 1:} {\em Due to the absence of accurate and rigorous methods for solving the IP alias

### 2013-08-07 Compiled/ebook.tex
added last page
added page numbers and fixed acents in intro

@@
 \author{
-\href{http://www.eecs.qmul.ac.uk/~hamed/}{Hamed Haddadi},  \href{http://www.uclouvain.be/olivier.bonaventure}{Olivier Bonaventure} (Editors)\\
+\href{http://www.eecs.qmul.ac.uk/~hamed/}{Hamed Haddadi},  \href{http://perso.uclouvain.be/olivier.bonaventure}{Olivier Bonaventure} (Editors)\\
 \\
@@
 \includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter6/chapter6.pdf}
-\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter4/paper.pdf}
+\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter4/chapter4.pdf}
 \includepdf[pages=-,pagecommand=\thispagestyle{plain}]{chapter1/chapter1.pdf}
@@
 
+\newpage
+
+\input{intro/lastpage.tex}
+
 \end{document}

### 2013-08-07 Compiled/intro/board.tex
fixes
added page numbers and fixed acents in intro

@@
 Ernst Biersack &	Eurecom \\
-Olivier Bonaventure &	Universit$\acute{e}$ catholique de Louvain \\
+Olivier Bonaventure &	Universit\'{e} catholique de Louvain \\
 Jon Crowcroft	&  University of Cambridge \\
@@
 Bruce Davie	& VMware \\
-Anja Feldmann	& Technische Universit$\ddot{a}$t Berlin\\
+Anja Feldmann	& Technische Universit\"{a}t Berlin\\
 Timothy Griffin	& University of Cambridge \\
@@
 Srinivasan Keshav &	University of Waterloo\\
-Jean-Yves Le Boudec	& $\acute{E}$cole polytechnique F$\acute{e}$d$\acute{e}$rale de Lausanne\\
+Jean-Yves Le Boudec	& \'{E}cole polytechnique F\'{e}d\'{e}rale de Lausanne\\
 Jennifer Rexford &	Princeton University\\
@@
 \begin{center}
-{\Huge \textbf{Review Board}}
+{\Large \textbf{Review Board}}
 
-\Large 
 \vspace{1in}
@@
 Saleem Bhatti	& University of St Andrews \\
-Olivier Bonaventure &	Universit$\acute{e}$ catholique de Louvain \\
+Olivier Bonaventure &	Universit\'{e} catholique de Louvain \\
 Gonzalo Camarillo  &  Ericsson Research   \\
@@
 Bruce Davie	& VMware \\
-Benoit Donnet  &   Universit$\acute{e}$ de Li$\grave{e}$ge  \\
+Benoit Donnet  &   Universit\'{e} de Li\`{e}ge  \\
 Damien Fay & Bournemouth University  \\

### 2013-08-07 Compiled/chapter4/sections/Collaboration-Use-Cases.tex
Added copyright information for reused figures

@@
    \caption{Gnutella with Oracle}%
-   \end{subfigure}   \caption{Visualization of Gnutella overlay topology}
+   \end{subfigure}   \caption{Visualization of Gnutella
+     overlay topology. Reprinted from \cite{afs-cispp2pcip-ccr07}. Included here by permission.}
    \label{fig:oracle:gnutella_topology}
@@
   \caption{By choosing a CDI server for a client with the help of CaTE, traffic
-  engineering goals and accurate end-user server assignment become possible.}
+  engineering goals and accurate end-user server assignment become
+  possible. Reprinted from \cite{Cate-CCR}. Included here by permission.}
   \label{fig:flowSelection}
@@
   \vspace{-0.1in}
-    \caption{CaTE System architecture and flow of messages.}
+    \caption{CaTE System architecture and flow of messages. Reprinted from \cite{Cate-CCR}. Included here by permission.}
     \label{fig:CaTE-architecture}
@@
 \center\includegraphics[width=0.90\linewidth]{figures-pdf/trafficShift-illustration}
-\caption{Content-aware Traffic Engineering Process}
+\caption{Content-aware Traffic Engineering Process. Reprinted from \cite{Cate-CCR}. Included here by permission.}
 \label{fig:Content-aware-illustration2}
@@
 \center\includegraphics[width=1\linewidth]{figures-pdf/providersVsVolume}
-\caption{CDF of traffic volume of CDIs in ISP1.}
+\caption{CDF of traffic volume of CDIs in ISP1. Reprinted from \cite{Cate-CCR}. Included here by permission.}
 \label{fig:CDF-Traffic-CPs}
@@
 \center\includegraphics[width=1\linewidth]{figures-pdf/Volume_timeseries}
-\caption{Normalized traffic for top CDIs by volume in ISP1.}
+\caption{Normalized traffic for top CDIs by volume in ISP1. Reprinted from \cite{Cate-CCR}. Included here by permission.}
 \label{fig:CPs-Total-Traffic}
@@
 \includegraphics[width=1\linewidth]{figures-pdf/SubnetChoice_r-2}
-\caption{Evolution over time of number of subnets for selected CDIs in the top 10 CDIs.}
+\caption{Evolution over time of number of subnets for selected CDIs in
+  the top 10 CDIs. Reprinted from \cite{Cate-CCR}. Included here by permission.}
 \label{fig:TemporalEffect}

### 2013-08-07 Compiled/chapter4/sections/Content-Delivery-Overview.tex
Fix compilation of chapter4 with new pdf files for figures
Added copyright information for reused figures

@@
 \begin{figure} \begin{center}
-\includegraphics[width=0.7\linewidth]{figures/flows-example.eps} 
+\includegraphics[width=0.7\linewidth]{figures-pdf/flows-example} 
 \end{center}
-\caption{Example of CDI deployment and traffic flows (Web traffic demands).}
+\caption{Example of CDI deployment and traffic flows (Web traffic
+  demands). Reprinted from \cite{inferring-web-demand}, \copyright ACM, 2004. Included here by permission.}
 \label{fig:aka:cdn} \end{figure}
@@
 \begin{center}
-\includegraphics[width=0.6\linewidth]{figures/page.eps}
+\includegraphics[width=0.6\linewidth]{figures-pdf/page}
 \vspace*{-.7cm}
 \end{center}
-\caption{Example Web page with some CDI content.}
+\caption{Example Web page with some CDI content. Reprinted from \cite{inferring-web-demand}, \copyright ACM, 2004. Included here by permission.}
 \label{fig:aka:cdn_page}

### 2013-08-07 Compiled/chapter4/sections/Traffic-Trends-Server-Diversity.tex
Added copyright information for reused figures

@@
   \includegraphics[height=0.7\linewidth]{figures-pdf/dns-diversity}
-  \caption{DNS replies for two different sites hosted on a CDI, in two-hour bins}%
+  \caption{DNS replies for two different sites hosted on a CDI, in two-hour bins.  Reprinted from \cite{PADIS2010}, \copyright ACM, 2010. Included here by permission.}%
   \label{fig:dns_diversity}
@@
 
-  \caption{CCDF of mean \# of IPs (top) and subnets (bottom) per DNS reply for the ISPs DNS resolver.}
+  \caption{CCDF of mean \# of IPs (top) and subnets (bottom) per DNS reply for the ISPs DNS resolver. Reprinted from \cite{PADIS2010}, \copyright ACM, 2010. Included here by permission.}
   \label{fig:CDF-AVG-REPLY-IPs}
@@
     volume  (top) and requests (bottom) including aggregation on
-    domain levels. (Logarithmic x-axis.)}
+    domain levels. (Logarithmic x-axis.) Reprinted from \cite{PADIS2010}, \copyright ACM, 2010. Included here by permission.}
   \label{fig:CDF-IPPAF-IPs}
@@
     volume  (top) and by requests (bottom) including aggregation on
-    domain levels. (Logarithmic x-axis.)}
+    domain levels. (Logarithmic x-axis.) Reprinted from \cite{PADIS2010}, \copyright ACM, 2010. Included here by permission.}
   \label{fig:CDF-IPPAF-Subnets}
@@
   \center\includegraphics[height=0.7\linewidth]{figures-pdf/maxTTLCDF}
-  \caption{CDF of DNS TTL value by traffic volume and by number of requests.}
+  \caption{CDF of DNS TTL value by traffic volume and by number of requests. Reprinted from \cite{PADIS2010}, \copyright ACM, 2010. Included here by permission.}
   \label{fig:CDF-IPPAF-TTL}

### 2013-08-08 Compiled/chapter9/Content/introduction.tex
fix typo in chapter9

@@
 
-\paragraph{Low-Power Lossy Networks (LLNs)} According to the IETF's terminology (defined in {\tt draft-ietf-roll-terminology-12}\footnote{An Internet-Draft in the Last Call for becoming an RFC, at the the time of writing this chapter.} \cite{draft_lln}), LLNs are ``typically composed of many embedded devices with limited power, memory, and processing resources interconnected by a variety of links, such as IEEE 802.15.4, LowPower WiFi" \cite{draft_lln}.
+\paragraph{Low-Power Lossy Networks (LLNs)} According to the IETF's terminology (defined in \cite{draft_lln}), LLNs are ``typically composed of many embedded devices with limited power, memory, and processing resources interconnected by a variety of links, such as IEEE 802.15.4, LowPower WiFi" \cite{draft_lln}.
 LLNs are thus a more specific case of MANETs (as defined in the previous paragraph), in which \glslink{router}{routers} typically operate with constraints on processing power, memory, and energy (battery power).

### 2013-08-08 Compiled/prepare_final.sh
added final chapters

@@ -1,0 +1,10 @@
+cp ebook.pdf chapters4web/SIGCOMMeBook2013v1.pdf
+cp chapter1/chapter1.pdf    chapters4web/SIGCOMMeBook2013v1_chapter8.pdf
+cp chapter2/chapter2.pdf    chapters4web/SIGCOMMeBook2013v1_chapter4.pdf
+cp chapter3/chapter3.pdf    chapters4web/SIGCOMMeBook2013v1_chapter5.pdf
+cp chapter4/chapter4.pdf    chapters4web/SIGCOMMeBook2013v1_chapter7.pdf
+cp chapter5/chapter5.pdf    chapters4web/SIGCOMMeBook2013v1_chapter2.pdf
+cp chapter6/chapter6.pdf    chapters4web/SIGCOMMeBook2013v1_chapter6.pdf
+cp chapter7/chapter7.pdf    chapters4web/SIGCOMMeBook2013v1_chapter1.pdf
+cp chapter8/chapter8.pdf    chapters4web/SIGCOMMeBook2013v1_chapter3.pdf
+cp chapter9/chapter9.pdf    chapters4web/SIGCOMMeBook2013v1_chapter9.pdf

### 2013-10-03 Compiled/chapter6/main.tex
fixes to c6

@@
 all} on router LER1. This command has the same effect of {\tt show ip bgp} but 
-it shows the routing entries related to IPv4 VPNs. In this case the output highlights 
-that LER1, in addition to its locally originated prefixes 10.0.0.0/24 with Route Distinguisher 100:1 and 193.1.1.0/24 with Route Distinguisher 100:2, knows two remote prefixes.
-Namely, it knows 200.1.6.0 with
-Route Distinguisher 100:11 and 193.1.1.0/24 with Route Distinguisher 100:22.
+it shows the routing entries related to IPv4 VPNs. In this case the output 
+highlights that LER1, in addition to its locally originated prefixes 10.0.0.0/24 
+with Route Distinguisher 100:11 and 193.1.1.0/24 with Route Distinguisher 
+100:22, knows two remote prefixes. Namely, it knows 200.1.6.0/24 with Route 
+Distinguisher 100:11 and 10.0.0.0/24 with Route Distinguisher 100:22.
 \end{shaded}
@@
 
- \item The P in the MPLS acronym stands for ``Private''.
+ \item The P in the VPN acronym stands for ``Private''.
 However, MPLS VPNs are only private at routing level: no authentication, confidentiality, or integrity is provided by the architecture.

### 2014-01-13 Compiled/intro/intro.tex
fixed typo

@@
 
-This eBook is composed of nine chapters chosen after a highly selective review process by the editorial board.
+The first edition of this eBook is composed of nine chapters chosen after a highly selective review process by the editorial board.
 The selected chapters cover advanced networking topics and accompanying teaching material (slides and exercises).
@@
 
-We wish to thank all there authors for providing such high quality chapters.
+We wish to thank all the authors for providing such high quality chapters.
 We are also grateful to the reviewers and the editorial board for spending many hours on each chapter to ensure a coherent level amongst all the chapters.
